### 2021-12-13 ms.Rnw
sgp

\documentclass[11pt,reqno,final]{amsart}
\usepackage{tikz}
\input{defs}
\usepackage[round,elide]{natbib}
\input{header}

\title{Structured Genealogy Processes}
\author[Lin]{Qianying Lin}
\author[King]{Aaron~A.~King}
\author[Ionides]{Edward~L.~Ionides}

\date{\today}

\hypersetup{pdftitle={Structured Genealogy Processes}}
\hypersetup{pdfauthor={Q.-Y. Lin, A.A. King, E.L. Ionides}}
\hypersetup{urlcolor=blue,citecolor=blue,linkcolor=blue,filecolor=blue}

<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
library(knitr)
prefix <- "strgp"
read_chunk("setup.R")
@
<<setup,include=FALSE,cache=FALSE,purl=FALSE>>=
@
<<packages,include=FALSE,cache=FALSE>>=
library(tidyverse)
library(ggtree)
library(pomp)
library(phylopomp)
library(cowplot)
stopifnot(getRversion() >= "4.0")
stopifnot(packageVersion("pomp")>="3.2")
stopifnot(packageVersion("phylopomp")>="0.0.14")
theme_set(theme_bw(base_family="serif"))
set.seed(1159254136)
@

\begin{document}
\maketitle

\begin{abstract}
  This is a followup paper on Markov Genealogy Process. We construct a continuous time Markov Process, called Structure Genealogy Process, to include structures in the focal population with a moderate adjustments in the setting of individuals, genealogies, and genealogy processes. We derive the exact expression of likelihood of a structured genealogy and develop simulation-based algorithms to conduct statistical inferences. Numerical simulation and real-world examples are included.
\end{abstract}

\section{Introduction}

\section{Mathematical settings}
\label{sec:settings}
\paragraph{Markov genealogy process}
A population process $\sg{X}_t\in\Xspace$ as a non-explosive Markov jump process is defined, where $t\in\Time$ indicates the time and the population is divided into $J\in\N$ classes, with initial-state distribution $p_0(x)$. We then define the jump rate functions $\jumprate_u(t,x)\in\rate$, where $u, x\in\Xspace$ are the jump event and states, respectively. Thus, the population process $\sg{X}_t$ is a stochastic process by defining a sequence of jumps, $\omega=\left(t_k, u_k, n_k\right)_{k=1}^K$, where $K\in\N\cup\{\infty\}$ is the total number of jumps and for the $k$-th jump, $t_k\in\Time$ is the time, $u_k\in\Uspace$ is the event type, and $n_{k}\in\N$ is the label of the individual who conducts $u_k$. The full construction can be referred to \citet{King2021}. We define the following functions on $\omega\in\Omega$:

\begin{equation}
  T_k(\omega)\coloneqq t_k,\qquad U_k(\omega)\coloneqq u_k,\qquad N_k(\omega)\coloneqq n_k,\qquad K(\omega)\coloneqq K
\end{equation}

and

\begin{equation}
  T(\omega)\coloneqq\left(T_k(\omega)\right)^{K(\omega)}_{k=0},\qquad U(\omega)\coloneqq\left(U_k(\omega)\right)^{K(\omega)}_{k=0},\qquad N(\omega,j)\coloneqq\left(N_k(\omega)\right)^{K(\omega)}_{k=0}.
\end{equation}

% \QYL{May not be necessary to repeat the GP paper.}
% 
% Obviously, there's a unique mapping from $\omega$ to the population process up to time $t\in\Time$. For $\omega\in\Omega$, we then can define the $t$-\emph{restriction} $\omega\vert_t$ and the sample path $\sg{X}_t(\omega)$ of $\omega$:
% 
% \begin{equation}
%   \omega\vert_t\coloneqq\max\{\omega'\in\Omega\,\vert\,\omega'\preceq\omega\text{ and }T_K(\omega')\le t\}
%   \quad\text{and}\quad
%   \sg{X}_t(\omega)=\sum_{k=0}^{K(\omega)}U_k(\omega)\indicator{\left[T_k(\omega),\infty\right)}(t),
% \end{equation}
% 
% where $\omega'\preceq\omega$ implies that $\omega$ is an extension of $\omega'$ and $\indicator{}$ is an indicator function.

\paragraph{Births}
Within a structured infective population, the parent and the child, at a given birth event, can be in different classes. That is, an individual in class $i$ can give birth to a newborn individual in Class $j$, where $i,j=1,\dots,J$. We then define the birth function $B_{ij}:\Uspace\to\N$ for any $i,j=1,\dots,J$. The set of birth events where an individual in Class $i$ gives birth to $n$ individuals in Class $j$ is $\Bset_{ij}\coloneqq\,B_{ij}^{-1}(\{1\})$. Trivially, the set of birth events in Class $j$ is $\Bset_j\coloneqq\bigcup_{i}\Bset_{ij}$.

\paragraph{Transitions}
Infective classes are interchangeable and we define the transition function from Class $i$ to $j$ for any $i\neq j$ and $i,j=1,\dots,J$ as $M_{ij}:\Uspace\to\N$ to represent the number of infectives transiting to Class $j$ from Class $i$. Thus the set or transition is $\Mset_{ij}\coloneqq\,M^{-1}_{ij}(\{1\})$ since we assume single transition at a single transition event. Similarly, we disallow coninciding transition events and other events.

\paragraph{Samples}
We assume samples are taken serially and no births or deaths are coincided. We then define the sampling function on the probability space of the population process $\sg{X}_t$: $G:\Uspace\to\N_0$ is the number of samples taken at event $u$, then the set of sampling events is $\Sset\coloneqq\,G^{-1}(\{1\})$. Of course, the non-coincidence assumption can be easily relaxed while it's not the focus in this paper.

\paragraph{Deaths and population size}
The population, the set of infected individuals in the context of epidemiology, is structured, categorized into interchangeable classes. We can then define population size function and death functions in Class $j$, respectively, as $I_{j}, D_{j}:\Xspace\to\N$ for any $j=1,\cdots,J$. 
\begin{equation}
  \jumprate_u(t,x) > 0 \Longrightarrow I_j(x+u)-I_j(x)=\sum_{i}\,B_{ij}(u)+\sum_{i}M_{ij}(u)-D_j(u)-\sum_{i}M_{ji}(u),
\end{equation}

for all $x, u\in\Xspace$. We disallow simultaneous death at a single death event at this stage and define the set of death events in Class $j$ as $\Dset_j\coloneqq\,D^{-1}(\{1\})$. We also insist, birth and death events don't coincide for any classes, $i,j$, that is,  $\Bset_{j}\cap\Dset_{i}=\emptyset$. Eventually, we can have the function for total population size by letting $I(x)\coloneqq\sum_{j=1}^J\,I_j(x)$ and it's trivial that

\begin{equation}
  \jumprate_u(t,x) > 0 \Longrightarrow I(x+u)-I(x)=\sum_{i,j}B_{i,j}(u)-\sum_{j}D_j(u),
\end{equation}

\section{Examples}

Some commonly used compartmental models in epidemiology are in fact structured, either explicitly or implicitly. Here, we demonstrate below how these representatives fit within the population processes we described previously.

\begin{figure}
\begin{minipage}{.5\textwidth}
    \begin{tikzpicture}[scale=1.5]
      \usetikzlibrary{arrows}
      \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
      \tikzstyle{det}=[color=black, thick, >=stealth]
      \tikzstyle{stoch}=[color=black, dashed, thick, >=stealth]
      \node (S) at (-2,0) {$S$};
      \node (E) at (-1,0) {$E$};
      \node (I) at (0,0) {$I$};
      \node (R) at (1,0) {$R$};
      \draw [det,->] (S) -- (E);
      \draw [det,->] (E) -- (I);
      \draw [det,->] (I) -- (R);
      \draw [det,->] (I) -- (0,.5) -- (-1.5,.5) -- (-1.5,0);
    \end{tikzpicture}
\end{minipage}%
\begin{minipage}{.5\textwidth}
    \begin{tikzpicture}[scale=1.5]
      \usetikzlibrary{arrows}
      \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
      \tikzstyle{det}=[color=black, thick, >=stealth]
      \tikzstyle{stoch}=[color=black, dashed, thick, >=stealth]
      \node (S) at (-2,0) {$S$};
      \node (I1) at (-.5,.7) {$I_1$};
      \node (I2) at (-.5,-.7) {$I_2$};
      \node (R) at (1,0) {$R$};
      \draw [det,->] (S) -- (I1);
      \draw [det,->] (S) -- (I2);
      \draw [det,->] (I1) -- (R);
      \draw [det,->] (I2) -- (R);
      \draw [det,->] (-.4,.5) -- (-.4,-.5);
      \draw [det,->] (-.6,-.5) -- (-.6,.5);
    \end{tikzpicture}
\end{minipage}
\caption{
  \label{fig:diagram}
}
\end{figure}

\paragraph{SEIR model}
The SEIR model is a simple extension of the most basic SIR model, by adding the latent infected compartment ``E'' into the system, where susceptible individual first become ``exposed'' when being infected, then turn into being infectious and eventually recover or get removed from the system. The infected population, which we are mostly interested in, is divided into two classes:
\begin{inparaenum}
  \item ``E'' represents those that are not able to spread the disease though infected and 
  \item ``I'' are those that are infectious. 
\end{inparaenum}
To fit within the aforementioned population process, we take $d=4$ so that the state vector is $\sg{X}=\left(s,e,i,g\right)$, where $s,e,i,g$ are the number of susceptibles, exposeds, infectives, and the cumulative number of genomic samples collected, respectively. We then summarize four types of jumps, with rate functions:

\begin{equation*}
  \jumprate_u =
  \begin{cases}
     b(t)\,s\,i, & u=\left(-1,1,0,0\right),\,s>0,\,i>0 \\
    \sigma\,e, & u=\left(0,-1,1,0\right),\,e>0,\\
    \gamma\,i, & u=\left(0,0,-1,0\right),\,i>0\\
    \psi(t,s,e,i,g), & u=\left(0,0,0,1\right),\, e+i>0\\
    0, & \text{otherwise.}
  \end{cases}
\end{equation*}

The above shows the rates of being infected, progression from exposed to infectious, recovery, and sampling, respectively. As in \citet{King2021}, the transmission $b(t)$ is time-dependent, and the sampling rate $\psi$ is any function, as long as the constraints in Section 2 are satisfied. Therefore, $I_1(\sg{X})=e$, $I_2(\sg{X})=i$, $I(\sg{X})=e+i$, $\Bset_{21}=\{\left(-1,1,0,0\right)\}$, $\Mset_{12}=\{\left(0,-1,1,0\right)\}$, $\Dset_2=\{\left(0,0,-1,0\right)\}$, $\Sset=\{\left(0,0,0,1\right)\}$, and $\Bset_{11}=\Bset_{12}=\Bset_{22}=\Mset_{21}=\Dset_1=\emptyset$.

\paragraph{SI$^2$R model}

We can customize the state and event vector to fit a complex system. In SI$^2$R model, we have two different while interchangeable classes of infectious individuals, the \emph{per capita} transmissibility therefore in the first class is $b_1(t)$ whilst in the second class is $b_2(t)$. Transition rates between Class $i$ and $j$ are also defined: Class $i$ infection turns into Class $j$ at rate $r_{ij}$ for $i,j=1,2$ and $i\neq j$. Furthermore, we also suppose that, once being infected, a susceptible individual becomes Class 1 infectious with probability $\rho$ while Class 2 infectious with probability $1-\rho$. To specify each event and the number of individuals involved, we define a relatively complex state vector by setting $d=10$ and $\sg{X}=\left(s,i_1,i_2,b_{11},b_{12},b_{21},b_{22},m_{12},m_{21},g\right)$, for $s$, $i_1$, and $i_2$ being the population size of the susceptible, the infectives in Class 1, and those in Class2, respectively, and $b_{11}$, $b_{12}$, $b_{21}$, $b_{22}$, $m_{12}$, and $m_{21}$ being the cumulative number of events of parents in Class 1 giving to children in Class 1 and 2, parents in Class 2 giving birth to children in Class 1 and 2, individuals transiting from Class 1 to Class 2, and vice versa, and sampling, respectively. In this case, we can summarize the event rates as:

\begin{equation*}
  \jumprate_u =
  \begin{cases}
                                          %s,i1,i2,b11,b12,b21,b22,m12,m21,g
     b_1(t)\,s\,i_1\,\rho,        & u=\left(-1,1,0,1,0,0,0,0,0,0\right), s>0, i_1>0 \\
     b_1(t)\,s\,i_1\,(1-\rho),    & u=\left(-1,0,1,0,1,0,0,0,0,0\right), s>0, i_1>0 \\
     b_2(t)\,s\,i_2\,\rho,        & u=\left(-1,1,0,0,0,1,0,0,0,0\right), s>0, i_2>0 \\
     b_2(t)\,s\,i_2\,(1-\rho),    & u=\left(-1,0,1,0,0,0,1,0,0,0\right), s>0, i_2>0 \\
     r_{12}\,i_1,                 & u=\left(0,-1,1,0,0,0,0,1,0,0\right), i_1>0 \\
     r_{21}\,i_2,                 & u=\left(0,1,-1,0,0,0,0,0,1,0\right), i_2>0 \\
     \gamma\,i_1,                 & u=\left(0,-1,0,0,0,0,0,0,0,0\right), i_1>0 \\
     \gamma\,i_2,                 & u=\left(0,0,-1,0,0,0,0,0,0,0\right), i_2>0 \\
     \psi(t,s,i_1,i_2,g),         & u=\left(0,0, 0,0,0,0,0,0,0,1\right), i_1+i_2>0\\
    0, & \text{otherwise.}
  \end{cases}
\end{equation*}

The nine non-zero rates above represent those of infection into Class 1 infected by individuals in Class 1, infection into Class 2 infected by individuals in Class 1, infection in Class 1 infected by individuals in Class 2, infection into Class 2 infected by individuals in Class 1, transition from Class 1 to Class 2, transition from Class 2 to Class 1, recovery from Class 1, recovery from Class 2, and sampling, respectively. We then summarize the population functions and sets as follows:

\begin{align*}
\begin{array}{lll}
  I_1(\sg{X})=i_1,                                  & I_2(\sg{X})=i_2,                              & I(\sg{X})=i_1+i_2, \\
  \Bset_{11}=\{\left(-1,1,0,1,0,0,0,0,0,0\right)\}, & \Bset_{12}=\{\left(-1,0,1,0,1,0,0,0,0,0\right)\}, & \\
  \Bset_{21}=\{\left(-1,1,0,0,0,1,0,0,0,0\right)\}, & \Bset_{22}=\{\left(-1,0,1,0,0,0,1,0,0,0\right)\}, & \\
  \Mset_{21}=\{\left(0,-1,1,0,0,0,0,1,0,0\right)\}, & \Mset_{12}=\{\left(0,1,-1,0,0,0,0,0,1,0\right)\}, & \\
  \Dset_1=\{\left(0,-1,0,0,0,0,0,0,0,0\right)\},    & \Dset_2=\{\left(0,0,-1,0,0,0,0,0,0,0\right)\},    & \\
  \Sset=\{\left(0,0, 0,0,0,0,0,0,0,1\right)\}. & &
\end{array}
\end{align*}

\paragraph{Migration}
One special case of the SI$^2$R model is that with one-time migration or introduction. In other words, an infectious individual from another population join our interested population at a certain point of time $\tau$ and then stay and spread the disease. In this case, if this specific individual is defined as Class 1 while other infectious individuals as Class 2, the transition rate $r_{12}$ is a Dirac delta function $\delta(t)$, where $\delta(t)=\infty$ for $t=\tau$ and $\delta(t)=0$ for $t\neq\tau$, and $r_{21}=0$ all the time.

\paragraph{Superspreading}
The definition of superspreading is ambiguous in epidemiology. Some researchers view it as a certain individual who has extremely strong transmissibility with high reproduction rate and call this individual \emph{superspreader}; others avoid to invidualize this phenomenon and consider it as an event where a cluster of infections appears in a single event within a small amount of time and define it as a \emph{superspreading event}. The first definition can be represented by the SI$^2$R model with two classes of infectious individuals: one with extremely strong transmissibility and the other with weak transmissibility. That is, $b_1(t)\ggg b_2(t)$ if Class 1 is the set of highly infective individuals and Class 2 is the set of ordinary infective. In terms of the second definition, the involvement of polytomies and multifurcating trees adds an extra layer of complexity in the population model, along with the issues of measurement error and model robustness, such that we would like to postpone the development of a new framework to a future study.

\section{Structure genealogy processes}

First, we define a global set $\Names$, which is a collection of infinite unique names. For any Class $j$ of infective population, we define an inventory $\sg{I}_t^j(\omega)=\mathrm{inven}\left(\omega\vert_t,j\right)$, a list to document all the extant infected individuals in Class $j$ at time $t$, by their names, where $\omega\vert_t$ is the sequence of jumps up to time $t$. On top of the global set $\Names$ and the $\mathrm{inven}$ procedure, we can then define six operators: 
\begin{enumerate}[(1)]
  \item $\mathrm{select}(\Names)$ returns a randomly selected name, call it $n$, from the set $\Names$, and update the global set $\Names=\Names\setminus\{n\}$;
  \item $\mathrm{initialize}(I)\coloneqq\bigcup_{i=1}^I\{\mathrm{select}(\Names)\}$;
  \item $\mathrm{add}(\sg{I})=\sg{I}\cup\{\mathrm{select}(\Names)\}$;
  \item $\mathrm{drop}(\sg{I},n)=\sg{I}\setminus\{n\}$, while update the global set $\Names=\Names\cup\{n\}$;
  \item $\mathrm{movein}(\sg{I},m)=\sg{I}\cup\{m\}$; and
  \item $\mathrm{moveout}(\sg{I},m)=\sg{I}\setminus\{m\}$.
\end{enumerate}

Now the deterministic and recursive procedure $\mathrm{inven}$ is described as follows, for $\omega$ is a finite jump sequence:

\begin{equation}
  \mathrm{inven}(\omega,j)\coloneqq
  \begin{cases}
    \mathrm{initialize}(I_j(\sg{X}_0(\omega)), & \text{if }K(\omega)=0; \\
    \mathrm{add}(\mathrm{inven}(\predec{\omega})), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Bset_j;\\
    \mathrm{drop}(\mathrm{inven}(\predec{\omega}),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Dset_j; \\
    \mathrm{movein}(\mathrm{inven}(\predec{\omega}),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ij}; \\
    \mathrm{moveout}(\mathrm{inven}(\predec{\omega}),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ji}; \\
    \mathrm{inven}(\predec{\omega}), & \text{otherwise}.
  \end{cases}
\end{equation}

Eventually, we can define $\mathrm{inven}(\omega)\coloneqq\{\mathrm{inven}(\omega,j)\}_{j=1}^J$ as a list of inventories.

\paragraph{Structured genealogies}
Types of nodes are defined in a structured genealogy: \emph{Leaves} are the extant members at current time and \emph{Internal Nodes} represent ancestors to a certain subset of extant members, which we further define \emph{Sampling Nodes} and \emph{Transition Nodes}, whose differences will be clear in a moment.

For the sake of distinction and visualization, we define a set of \emph{colored ball}s, $(f,n)\in\Colors\times\N$, where $n$ is the label and $f$ is the color of the ball, and $\Colors\coloneqq\{\green,\black,\blue,\red,\purple\}$. Therefore, we can define a genealogy \emph{node} as a quadruple $(n,t,w,z)$, where $n\in\N$ is the node's \emph{name}, $t\in\Time$ is its time, $w$ is the node's \emph{pocket} (\ie unordered pair of colored balls), and $z\in\N$ is the node's \emph{class}. For a node $\Node{p}$, we will use $n(\Node{p})$, $t(\Node{p})$, $w(\Node{p})$, and $z(\Node{p})$ to denote the name, time, pocket, and class of $\Node{p}$, respectively. We then assign black, green, blue, and purple balls to serve as pointers to the extant population at time $t$, to the internal nodes, to the samples, and to the transitions between classes, respectively.

Eventually, the current time $t$ and the sequence of nodes, $\Gen=\left(t,\left(\Node{p}_k\right)_{k=0}^{K-1}\right)$, explicitly define a structured genealogy $\Gen$, where $t\in\Time$, $K\in\N$, and node $\Node{p}_k$ are subjected to conditions described in \citet{King2021}. We then use $t(\Gen)$, $K(\Gen)$, and $\Node{p}_k(\Gen)$ to represent the time, the length (\ie the number of nodes), and the $k$-th node of the genealogy $\Gen$, respectively.

\paragraph{Effect of births, death, sampling and transistions}

Births, deaths, samples, and transitions change the topology of a structured genealogy. A death at time $t$ indicates the removal of a leaf in the genealogy and the dismiss of a node. The detailed definition can be found in \citet{King2021} and here we continue to use the $\mathrm{drop}(\Gen)$ to denote the resulting sequence of nodes.

A birth at time $t$ in a structured genealogy is similar to that in a unstructured one \citep{King2021}, while the new member introduced to the queue of sequence is in the same class as the parent. Let $\Ball{b}$ be the $n$-th black ball, the parent, and there exists a node $\Node{p}\in\Gen$ with $\Ball{b}\in w(\Node{p})=\{\Ball{b},\Ball{b}'\}$ where $\Ball{b}'$ is the other ball held by $\Node{p}$. We then can produce a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},z(\Node{p}))$, where $c=\mathrm{Select}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\black,c)$. By swapping balls between $\Node{p}$ and $\Node{p}'$, it ends up that $w(\Node{p})=\{\Ball{g},\Ball{b}'\}$ and $w(\Node{p}')=\{\Ball{b},\Ball{b}''\}$. We denote the resulting sequence of nodes as $\mathrm{add}(\seq{\Gen},n)$.

A sample at time $t$ results in the participation of a new node with a new blue ball. The procedure is similar to the previous sampling in \citet{King2021} and the birth mentioned above: the $n$-th black ball held by $\Node{p}$ (\ie $w(\Node{p})=\{\Ball{b},\Ball{b}'\}$) is selected, then a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},z(\Node{p}))$ is generated, where $c=\mathrm{Select}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\blue,q)$, at the end $\Node{p}$ exchanges $\Ball{b}$ for $\Ball{g}$ with $\Node{p}'$. Here, $q$ is the name of the new blue ball, which is the ordinal number of the transition, $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is blue}\}|$. Note that, same as that in a birth, $z(\Node{p}')=z(\Node{p})$. We also use the same notation $\mathrm{sample}(\Gen,n)$ as the resulting sequence of nodes.

A transition at time $t$ leads to the inclusion of a new node with a new purple ball. Let $\Ball{b}$ denotes the $n$-th black ball selected to indicate an extant in the population transits from his/her/their current class to a new one. Assuming that the unique node $\Node{p}\in\Gen$ baring this selected black ball (\ie $\Ball{b}\in w(\Node{p})=\{\Ball{b},\Ball{b}'\}$) is in Class $i$ (\ie $z(\Node{p})=i$) and it's transiting to Class $j$ where $j\neq i$. Now we initialize a new node by taking $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},j)$, where $c=\mathrm{Select}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\purple,q)$. Similar to other events, we swap $\Ball{b}$ for $\Ball{g}$ between nodes $\Node{p}$ and $\Node{p}'$ and insert $\Node{p}'$ at the last position of the node-sequence. Here, $q$ is the name of the new purple ball, which is the ordinal number of the transition, $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is purple}\}|$. We call this new node $\Node{p}'$ the \emph{Transition Nodes} in the genealogy and denote the resulting sequence of nodes by $\mathrm{transition}(\seq{\Gen},n,j)$.

\paragraph{Event times and structured genealogy process}

Given a genealogy $\Gen$, the set of \emph{genealogical event times}, $\Eset(\Gen)$, is the set of all node times.
Several of its subsets are of interest.
In particular, we define
\begin{equation*}
  \begin{aligned}
    \Eset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen\}, \\
    \Masterset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a green ball}\}, \\
    \Cset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains two green balls}\}, \\
    \Lset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains no green balls}\}, \\
    \Tset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a blue ball}\},\\
    \Zset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a purple ball}\},\\
    \Qset(\Gen) &\coloneqq \Masterset(\Gen) \cap \Tset(\Gen).
  \end{aligned}
\end{equation*}
With these definitions,
$\Masterset(\Gen)$ comprises the internal node times,
$\Cset(\Gen)$ is the set of branch times,
$\Lset(\Gen)$ contains the leaf times,
$\Tset(\Gen)$ holds the sample times,
$\Zset(\Gen)$ holds the transition times,
and $\Qset(\Gen)$ is the set of \emph{direct-descent times},
i.e., the times of samples that are themselves directly ancestral to other samples.

The structured genealogy evolves over time, so we can proceed to define the \emph{structured genealogy process} $\Gen_t$ now.

\section{Likelihood, filtering equation, and algorithms}

\section{Results}

\section{Discussion}

\bibliographystyle{preprint}
\bibliography{phylopomp}

\end{document}

### 2021-12-13 setup.R

@@ -1,0 +1,44 @@
+## ----setup--------------------------------------------------
+library(knitr)
+opts_chunk$set(
+  cache=TRUE,
+  cache.path=paste0("tmp/",prefix,"/"),
+  comment=NA,
+  echo=FALSE,
+  eval=TRUE,
+  include=TRUE,
+  dev="cairo_pdf",
+  dev.args=list(bg='transparent'),
+  dpi=300,
+  error=FALSE,
+  fig.align='center',
+  fig.dim=c(4,6.83),
+  fig.lp="fig:",
+  fig.path=paste0("tmp/",prefix,"/"),
+  fig.pos="h!",
+  fig.show='asis',
+  highlight=TRUE,
+  message=FALSE,
+  progress=TRUE,
+  prompt=FALSE,
+  purl=TRUE,
+  results="markup",
+  size='small',
+  strip.white=TRUE,
+  tidy=FALSE,
+  warning=FALSE
+  )
+
+options(
+  width=60,
+  keep.source=TRUE,
+  encoding="UTF-8"
+)
+
+registerS3method(
+  "knit_print",
+  "data.frame",
+  function (x, ...) {
+    print(x,row.names=FALSE)
+  }
+)

### 2021-12-17 setup.R
housekeeping work for the SGP paper

@@ -29,10 +29,17 @@ opts_chunk$set(
   warning=FALSE
   )
 
+knit_hooks$set(
+  document = function (x) {
+    sub('\\usepackage[]{color}','\\usepackage{xcolor}',x,fixed=TRUE)
+  }
+)
+
 options(
   width=60,
   keep.source=TRUE,
-  encoding="UTF-8"
+  encoding="UTF-8",
+  dplyr.summarise.inform=FALSE
 )
 
 registerS3method(

### 2022-01-05 ms.tex
sgp files

--- 2021-12-13 ms.Rnw
@@
-\documentclass[11pt,reqno,final]{amsart}
+\documentclass[11pt,reqno,final]{amsart}\usepackage[]{graphicx}\usepackage[]{color}
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage{tikz}
@@
 
-<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
-library(knitr)
-prefix <- "strgp"
-read_chunk("setup.R")
-@
-<<setup,include=FALSE,cache=FALSE,purl=FALSE>>=
-@
-<<packages,include=FALSE,cache=FALSE>>=
-library(tidyverse)
-library(ggtree)
-library(pomp)
-library(phylopomp)
-library(cowplot)
-stopifnot(getRversion() >= "4.0")
-stopifnot(packageVersion("pomp")>="3.2")
-stopifnot(packageVersion("phylopomp")>="0.0.14")
-theme_set(theme_bw(base_family="serif"))
-set.seed(1159254136)
-@
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}

### 2022-01-05 ms.Rnw
housekeeping work for the SGP paper
constellation diagram

@@
 % where $\omega'\preceq\omega$ implies that $\omega$ is an extension of $\omega'$ and $\indicator{}$ is an indicator function.
+
+\begin{figure}
+  \resizebox{0.6\linewidth}{!}{
+    \begin{tikzpicture}[scale=1.8]
+      \usetikzlibrary{arrows}
+      \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
+      \tikzstyle{det}=[color=black, thick, >=stealth]
+      \tikzstyle{stoch}=[color=black, dashed, thick, >=stealth]
+      \node (O) at (0,0) {$\Master_t$};
+      \node (X) at (180:2) {$\X_t$};
+      \node (H) at (210:2) {$\Hist_t$};
+      \node (I) at (240:2) {$\Inv_t$};
+      \node (G) at (270:2) {$\Gen_t$};
+      \node (P) at (300:2) {$\Prune_t$};
+      \node (V) at (330:2) {$\Vis_t$};
+      \node (W) at (360:2) {$\Wch_i$};
+      \draw [det,->] (O) -- (X);
+      \draw [det,->] (O) -- (H);
+      \draw [det,->] (H) -- (X);
+      \draw [det,->] (O) -- (I);
+      \draw [det,->] (O) -- (G);
+      \draw [det,->] (O) -- (P);
+      \draw [det,->] (O) -- (V);
+      \draw [det,->] (O) -- (W);
+      \draw [det,->] (V) -- (W);
+      \draw [det,->] (G) -- (I) node[midway,below,sloped] {$\mathrm{ext}$};
+      \draw [det,->] (G) -- (P) node[midway,below,sloped] {$\mathrm{prune}$};
+      \draw [det,->] (P) -- (V) node[midway,below,sloped] {$\mathrm{obs}$};
+      \draw [stoch,->] (H) -- (I);
+      \draw [stoch,->] (H) -- (G);
+      \draw [stoch,->] (H) -- (V);
+      \draw [stoch,->] (H) -- (W);
+    \end{tikzpicture}
+  }
+  \caption{
+    \label{fig:constellation}
+    Relations among the various Markov processes discussed in the paper.
+    Deterministic maps are indicated with solid arrows;
+    random maps are shown as dashed arrows.
+    All the maps shown commute.
+    $\X_t$ is the \emph{population process}, a model of the dynamics of some system, which we take as a starting point.
+    $\Hist_t$ is the \emph{history process}, which records the full history of $\X_t$.
+    $\Inv_t$ is the \emph{inventory process}: at each time $t$, $\Inv_t$ is an inventory of all extant individuals in the population, each of which has a globally unique name.
+    $\Gen_t$ is the \emph{genealogy process}, which captures the precise genealogical relationships among all individuals in $\Inv_t$, as well as among any samples that have been taken from the population.
+    $\Vis_t$ is the \emph{visible genealogy process}, which is $\Gen_t$ pruned so that only relationships among samples remain.
+    Finally $\Wch_i$ is the \emph{embedded chain of the visible genealogy process}, which is $\Vis_{s_i}$, $s_i$ being the time of the $i$-th sample.
+    All of these processes can be obtained via deterministic procedures applied to the \emph{master process} $\Master_t$, as described in the text.
+  }
+\end{figure}
 
@@
 
-\bibliographystyle{preprint}
+\bibliographystyle{../preprint}
 \bibliography{phylopomp}

### 2022-01-11 ms.Rnw
sgp ms updates

@@
 \begin{document}
+% \SweaveOpts{concordance=TRUE}
 \maketitle
@@
 \paragraph{Markov genealogy process}
-A population process $\sg{X}_t\in\Xspace$ as a non-explosive Markov jump process is defined, where $t\in\Time$ indicates the time and the population is divided into $J\in\N$ classes, with initial-state distribution $p_0(x)$.
+A population process $\sg{X}_t\in\Xspace$ as a non-explosive Markov jump process is defined, where $t\in\Time$ indicates the time and the population is divided into $J\in\N$ demes, with initial-state distribution $p_0(x)$.
 We then define the jump rate functions $\jumprate_u(t,x)\in\rate$, where $u, x\in\Xspace$ are the jump event and states, respectively.
@@
 \paragraph{Births}
-Within a structured infective population, the parent and the child, at a given birth event, can be in different classes.
-That is, an individual in class $i$ can give birth to a newborn individual in Class $j$, where $i,j=1,\dots,J$.
+Within a structured infective population, the parent and the child, at a given birth event, can be in different demes.
+That is, an individual in Deme $i$ can give birth to a newborn individual in Deme $j$, where $i,j=1,\dots,J$.
 We then define the birth function $B_{ij}:\Uspace\to\N$ for any $i,j=1,\dots,J$.
-The set of birth events where an individual in Class $i$ gives birth to $n$ individuals in Class $j$ is $\Bset_{ij}\coloneqq\,B_{ij}^{-1}(\{1\})$.
-Trivially, the set of birth events in Class $j$ is $\Bset_j\coloneqq\bigcup_{i}\Bset_{ij}$.
+The set of birth events where an individual in Deme $i$ gives birth to $n$ individuals in Deme $j$ is $\Bset_{ij}\coloneqq\,B_{ij}^{-1}(\{1\})$.
+Trivially, the set of birth events in Deme $j$ is $\Bset_j\coloneqq\bigcup_{i}\Bset_{ij}$.
 
 \paragraph{Transitions}
-Infective classes are interchangeable and we define the transition function from Class $i$ to $j$ for any $i\neq j$ and $i,j=1,\dots,J$ as $M_{ij}:\Uspace\to\N$ to represent the number of infectives transiting to Class $j$ from Class $i$.
+Infective demes are interchangeable and we define the transition function from Deme $i$ to $j$ for any $i\neq j$ and $i,j=1,\dots,J$ as $M_{ij}:\Uspace\to\N$ to represent the number of infectives transiting to Deme $j$ from Deme $i$.
 Thus the set or transition is $\Mset_{ij}\coloneqq\,M^{-1}_{ij}(\{1\})$ since we assume single transition at a single transition event.
@@
 \paragraph{Deaths and population size}
-The population, the set of infected individuals in the context of epidemiology, is structured, categorized into interchangeable classes.
-We can then define population size function and death functions in Class $j$, respectively, as $I_{j}, D_{j}:\Xspace\to\N$ for any $j=1,\cdots,J$.
+The population, the set of infected individuals in the context of epidemiology, is structured, categorized into interchangeable demes.
+We can then define population size function and death functions in Deme $j$, respectively, as $I_{j}, D_{j}:\Xspace\to\N$ for any $j=1,\cdots,J$.
 \begin{equation}
@@
 for all $x, u\in\Xspace$.
-We disallow simultaneous death at a single death event at this stage and define the set of death events in Class $j$ as $\Dset_j\coloneqq\,D^{-1}(\{1\})$.
-We also insist, birth and death events don't coincide for any classes, $i,j$, that is,  $\Bset_{j}\cap\Dset_{i}=\emptyset$.
+We disallow simultaneous death at a single death event at this stage and define the set of death events in Deme $j$ as $\Dset_j\coloneqq\,D^{-1}(\{1\})$.
+We also insist, birth and death events don't coincide for any demes, $i,j$, that is,  $\Bset_{j}\cap\Dset_{i}=\emptyset$.
 Eventually, we can have the function for total population size by letting $I(x)\coloneqq\sum_{j=1}^J\,I_j(x)$ and it's trivial that
@@
 The SEIR model is a simple extension of the most basic SIR model, by adding the latent infected compartment ``E'' into the system, where susceptible individual first become ``exposed'' when being infected, then turn into being infectious and eventually recover or get removed from the system.
-The infected population, which we are mostly interested in, is divided into two classes:
+The infected population, which we are mostly interested in, is divided into two demes:
 \begin{inparaenum}
@@
 We can customize the state and event vector to fit a complex system.
-In SI$^2$R model, we have two different while interchangeable classes of infectious individuals, the \emph{per capita} transmissibility therefore in the first class is $b_1(t)$ whilst in the second class is $b_2(t)$.
-Transition rates between Class $i$ and $j$ are also defined: Class $i$ infection turns into Class $j$ at rate $r_{ij}$ for $i,j=1,2$ and $i\neq j$.
-Furthermore, we also suppose that, once being infected, a susceptible individual becomes Class 1 infectious with probability $\rho$ while Class 2 infectious with probability $1-\rho$.
-To specify each event and the number of individuals involved, we define a relatively complex state vector by setting $d=10$ and $\sg{X}=\left(s,i_1,i_2,b_{11},b_{12},b_{21},b_{22},m_{12},m_{21},g\right)$, for $s$, $i_1$, and $i_2$ being the population size of the susceptible, the infectives in Class 1, and those in Class2, respectively, and $b_{11}$, $b_{12}$, $b_{21}$, $b_{22}$, $m_{12}$, and $m_{21}$ being the cumulative number of events of parents in Class 1 giving to children in Class 1 and 2, parents in Class 2 giving birth to children in Class 1 and 2, individuals transiting from Class 1 to Class 2, and vice versa, and sampling, respectively.
+In SI$^2$R model, we have two different while interchangeable demes of infectious individuals, the \emph{per capita} transmissibility therefore in the first deme is $b_1(t)$ whilst in the second deme is $b_2(t)$.
+Transition rates between Deme $i$ and $j$ are also defined: Deme $i$ infection turns into Deme $j$ at rate $r_{ij}$ for $i,j=1,2$ and $i\neq j$.
+Furthermore, we also suppose that, once being infected, a susceptible individual becomes Deme 1 infectious with probability $\rho$ while Deme 2 infectious with probability $1-\rho$.
+To specify each event and the number of individuals involved, we define a relatively complex state vector by setting $d=10$ and $\sg{X}=\left(s,i_1,i_2,b_{11},b_{12},b_{21},b_{22},m_{12},m_{21},g\right)$, for $s$, $i_1$, and $i_2$ being the population size of the susceptible, the infectives in Deme 1, and those in Deme 2, respectively, and $b_{11}$, $b_{12}$, $b_{21}$, $b_{22}$, $m_{12}$, and $m_{21}$ being the cumulative number of events of parents in Deme 1 giving to children in Deme 1 and 2, parents in Deme 2 giving birth to children in Deme 1 and 2, individuals transiting from Deme 1 to Deme 2, and vice versa, and sampling, respectively.
 In this case, we can summarize the event rates as:
@@
 
-The nine non-zero rates above represent those of infection into Class 1 infected by individuals in Class 1, infection into Class 2 infected by individuals in Class 1, infection in Class 1 infected by individuals in Class 2, infection into Class 2 infected by individuals in Class 1, transition from Class 1 to Class 2, transition from Class 2 to Class 1, recovery from Class 1, recovery from Class 2, and sampling, respectively.
+The nine non-zero rates above represent those of infection into Deme 1 infected by individuals in Deme 1, infection into Deme 2 infected by individuals in Deme 1, infection in Deme 1 infected by individuals in Deme 2, infection into Deme 2 infected by individuals in Deme 1, transition from Deme 1 to Deme 2, transition from Deme 2 to Deme 1, recovery from Deme 1, recovery from Deme 2, and sampling, respectively.
 We then summarize the population functions and sets as follows:
@@
 In other words, an infectious individual from another population join our interested population at a certain point of time $\tau$ and then stay and spread the disease.
-In this case, if this specific individual is defined as Class 1 while other infectious individuals as Class 2, the transition rate $r_{12}$ is a Dirac delta function $\delta(t)$, where $\delta(t)=\infty$ for $t=\tau$ and $\delta(t)=0$ for $t\neq\tau$, and $r_{21}=0$ all the time.
+In this case, if this specific individual is defined as Deme 1 while other infectious individuals as Deme 2, the transition rate $r_{12}$ is a Dirac delta function $\delta(t)$, where $\delta(t)=\infty$ for $t=\tau$ and $\delta(t)=0$ for $t\neq\tau$, and $r_{21}=0$ all the time.
 
@@
 Some researchers view it as a certain individual who has extremely strong transmissibility with high reproduction rate and call this individual \emph{superspreader}; others avoid to invidualize this phenomenon and consider it as an event where a cluster of infections appears in a single event within a small amount of time and define it as a \emph{superspreading event}.
-The first definition can be represented by the SI$^2$R model with two classes of infectious individuals: one with extremely strong transmissibility and the other with weak transmissibility.
-That is, $b_1(t)\ggg b_2(t)$ if Class 1 is the set of highly infective individuals and Class 2 is the set of ordinary infective.
+The first definition can be represented by the SI$^2$R model with two demes of infectious individuals: one with extremely strong transmissibility and the other with weak transmissibility.
+That is, $b_1(t)\ggg b_2(t)$ if Deme 1 is the set of highly infective individuals and Deme 2 is the set of ordinary infective.
 In terms of the second definition, the involvement of polytomies and multifurcating trees adds an extra layer of complexity in the population model, along with the issues of measurement error and model robustness, such that we would like to postpone the development of a new framework to a future study.
@@
 
-First, we define a global set $\Names$, which is a collection of infinite unique names.
-For any Class $j$ of infective population, we define an inventory $\sg{I}_t^j(\omega)=\mathrm{inven}\left(\omega\vert_t,j\right)$, a list to document all the extant infected individuals in Class $j$ at time $t$, by their names, where $\omega\vert_t$ is the sequence of jumps up to time $t$.
+\paragraph{Inventory process, deme process}
+First, we define a global set $\Names\coloneqq\N$, which is a collection of infinite unique names.
+For any Deme $j$ of infective population, we define an inventory $\Inv_t^j(\omega)=\mathrm{inven}\left(\omega\vert_t,j\right)$, a list to document all the extant infected individuals in Deme $j$ at time $t$, by their names, where $\omega\vert_t$ is the sequence of jumps up to time $t$.
 On top of the global set $\Names$ and the $\mathrm{inven}$ procedure, we can then define six operators:
 \begin{enumerate}[(1)]
-  \item $\mathrm{select}(\Names)$ returns a randomly selected name, call it $n$, from the set $\Names$, and update the global set $\Names=\Names\setminus\{n\}$;
-  \item $\mathrm{initialize}(I)\coloneqq\bigcup_{i=1}^I\{\mathrm{select}(\Names)\}$;
-  \item $\mathrm{add}(\sg{I})=\sg{I}\cup\{\mathrm{select}(\Names)\}$;
-  \item $\mathrm{drop}(\sg{I},n)=\sg{I}\setminus\{n\}$, while update the global set $\Names=\Names\cup\{n\}$;
-  \item $\mathrm{movein}(\sg{I},m)=\sg{I}\cup\{m\}$; and
-  \item $\mathrm{moveout}(\sg{I},m)=\sg{I}\setminus\{m\}$.
+  \item $\mathrm{next}(\Names)$ returns the minimum integer in $\Names$, call it $n$, and update the global set $\Names=\Names\setminus\{n\}$;
+  \item $\mathrm{initialize}(I)\coloneqq\bigcup_{i=1}^I\{\mathrm{next}(\Names)\}$;
+  \item $\mathrm{add}(\sg{I})=\sg{I}\cup\{\mathrm{next}(\Names)\}$;
+  \item $\mathrm{drop}(\sg{I},n)=\sg{I}\setminus\{n\}$; and
+  \item $\mathrm{movein}(\sg{I},n)=\sg{I}\cup\{n\}$.
 \end{enumerate}
@@
     \mathrm{initialize}(I_j(\sg{X}_0(\omega)), & \text{if }K(\omega)=0; \\
-    \mathrm{add}(\mathrm{inven}(\predec{\omega})), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Bset_j;\\
-    \mathrm{drop}(\mathrm{inven}(\predec{\omega}),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Dset_j; \\
-    \mathrm{movein}(\mathrm{inven}(\predec{\omega}),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ij}; \\
-    \mathrm{moveout}(\mathrm{inven}(\predec{\omega}),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ji}; \\
-    \mathrm{inven}(\predec{\omega}), & \text{otherwise}.
+    \mathrm{add}(\mathrm{inven}(\predec{\omega},j)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Bset_j;\\
+    \mathrm{drop}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Dset_j; \\
+    \mathrm{movein}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ij}; \\
+    \mathrm{drop}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ji}; \\
+    \mathrm{inven}(\predec{\omega},j), & \text{otherwise}.
   \end{cases}
@@
 
-Eventually, we can define $\mathrm{inven}(\omega)\coloneqq\{\mathrm{inven}(\omega,j)\}_{j=1}^J$ as a list of inventories.
+Eventually, we can define $\mathrm{inven}(\omega)\coloneqq\{\mathrm{inven}(\omega,j)\}_{j=1}^J$ as a list of inventories, such that $\Inv_t(\omega)\coloneqq\{\Inv_t^j(\omega)\}_{j=1}^J$.
+We can further define the deme process $\demefuc_t^j(\omega)\coloneqq\frac{\Inv_t^j(\omega)}{\Inv_t(\omega)}$ and $\demefuc_t(\omega)\coloneqq\{\demefuc_t^j(\omega)\}_{j=1}^J$.
 
@@
 For the sake of distinction and visualization, we define a set of \emph{colored ball}s, $(f,n)\in\Colors\times\N$, where $n$ is the label and $f$ is the color of the ball, and $\Colors\coloneqq\{\green,\black,\blue,\red,\purple\}$.
-Therefore, we can define a genealogy \emph{node} as a quadruple $(n,t,w,z)$, where $n\in\N$ is the node's \emph{name}, $t\in\Time$ is its time, $w$ is the node's \emph{pocket} (\ie unordered pair of colored balls), and $z\in\N$ is the node's \emph{class}.
-For a node $\Node{p}$, we will use $n(\Node{p})$, $t(\Node{p})$, $w(\Node{p})$, and $z(\Node{p})$ to denote the name, time, pocket, and class of $\Node{p}$, respectively.
-We then assign black, green, blue, and purple balls to serve as pointers to the extant population at time $t$, to the internal nodes, to the samples, and to the transitions between classes, respectively.
+Therefore, we can define a genealogy \emph{node} as a quadruple $(n,t,w,z)$, where $n\in\N$ is the node's \emph{name}, $t\in\Time$ is its time, $w$ is the node's \emph{pocket} (\ie unordered pair of colored balls), and $z\in\N$ is the node's \emph{deme}.
+For a node $\Node{p}$, we will use $n(\Node{p})$, $t(\Node{p})$, $w(\Node{p})$, and $z(\Node{p})$ to denote the name, time, pocket, and deme of $\Node{p}$, respectively.
+We then assign black, green, blue, and purple balls to serve as pointers to the extant population at time $t$, to the internal nodes, to the samples, and to the transitions between demes, respectively.
 
@@
 
-A birth at time $t$ in a structured genealogy is similar to that in a unstructured one \citep{King2021}, while the new member introduced to the queue of sequence is in the same class as the parent.
+A birth at time $t$ in a structured genealogy is similar to that in a unstructured one \citep{King2021}, while the new member introduced to the queue of sequence is in the same deme as the parent.
 Let $\Ball{b}$ be the $n$-th black ball, the parent, and there exists a node $\Node{p}\in\Gen$ with $\Ball{b}\in w(\Node{p})=\{\Ball{b},\Ball{b}'\}$ where $\Ball{b}'$ is the other ball held by $\Node{p}$.
-We then can produce a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},z(\Node{p}))$, where $c=\mathrm{Select}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\black,c)$.
+We then can produce a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},z(\Node{p}))$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\black,c)$.
 By swapping balls between $\Node{p}$ and $\Node{p}'$, it ends up that $w(\Node{p})=\{\Ball{g},\Ball{b}'\}$ and $w(\Node{p}')=\{\Ball{b},\Ball{b}''\}$.
@@
 A sample at time $t$ results in the participation of a new node with a new blue ball.
-The procedure is similar to the previous sampling in \citet{King2021} and the birth mentioned above: the $n$-th black ball held by $\Node{p}$ (\ie $w(\Node{p})=\{\Ball{b},\Ball{b}'\}$) is selected, then a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},z(\Node{p}))$ is generated, where $c=\mathrm{Select}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\blue,q)$, at the end $\Node{p}$ exchanges $\Ball{b}$ for $\Ball{g}$ with $\Node{p}'$.
+The procedure is similar to the previous sampling in \citet{King2021} and the birth mentioned above: the $n$-th black ball held by $\Node{p}$ (\ie $w(\Node{p})=\{\Ball{b},\Ball{b}'\}$) is selected, then a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},z(\Node{p}))$ is generated, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\blue,q)$, at the end $\Node{p}$ exchanges $\Ball{b}$ for $\Ball{g}$ with $\Node{p}'$.
 Here, $q$ is the name of the new blue ball, which is the ordinal number of the transition, $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is blue}\}|$.
@@
 A transition at time $t$ leads to the inclusion of a new node with a new purple ball.
-Let $\Ball{b}$ denotes the $n$-th black ball selected to indicate an extant in the population transits from his/her/their current class to a new one.
-Assuming that the unique node $\Node{p}\in\Gen$ baring this selected black ball (\ie $\Ball{b}\in w(\Node{p})=\{\Ball{b},\Ball{b}'\}$) is in Class $i$ (\ie $z(\Node{p})=i$) and it's transiting to Class $j$ where $j\neq i$.
-Now we initialize a new node by taking $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},j)$, where $c=\mathrm{Select}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\purple,q)$.
+Let $\Ball{b}$ denotes the $n$-th black ball selected to indicate an extant in the population transits from his/her/their current deme to a new one.
+Assuming that the unique node $\Node{p}\in\Gen$ baring this selected black ball (\ie $\Ball{b}\in w(\Node{p})=\{\Ball{b},\Ball{b}'\}$) is in Deme $i$ (\ie $z(\Node{p})=i$) and it's transiting to Deme $j$ where $j\neq i$.
+Now we initialize a new node by taking $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},j)$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\purple,q)$.
 Similar to other events, we swap $\Ball{b}$ for $\Ball{g}$ between nodes $\Node{p}$ and $\Node{p}'$ and insert $\Node{p}'$ at the last position of the node-sequence.
 Here, $q$ is the name of the new purple ball, which is the ordinal number of the transition, $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is purple}\}|$.
-We call this new node $\Node{p}'$ the \emph{Transition Nodes} in the genealogy and denote the resulting sequence of nodes by $\mathrm{transition}(\seq{\Gen},n,j)$.
+We call this new node $\Node{p}'$ the \emph{Transition Nodes} in the genealogy and denote the resulting sequence of nodes by $\mathrm{transition}(\seq{\Gen},n,i,j)$.
 
-\paragraph{Event times and structured genealogy process}
+\paragraph{Event times}
 
 Given a genealogy $\Gen$, the set of \emph{genealogical event times}, $\Eset(\Gen)$, is the set of all node times.
-Several of its subsets are of interest.
+Its subsets, including those for different event types and those for different demes, are of interest.
 In particular, we define
@@
   \begin{aligned}
-    \Eset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen\}, \\
-    \Masterset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a green ball}\}, \\
-    \Cset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains two green balls}\}, \\
-    \Lset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains no green balls}\}, \\
-    \Tset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a blue ball}\},\\
-    \Zset(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a purple ball}\},\\
-    \Qset(\Gen) &\coloneqq \Masterset(\Gen) \cap \Tset(\Gen).
+    \Eset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})\ \text{is}\ j\}, \\
+    \Masterset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a green ball}, z(\Node{p})\ \text{is}\ j\}, \\
+    \Cset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains two green balls}, z(\Node{p})\ \text{is}\ j\}, \\
+    \Lset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains no green balls}, z(\Node{p})\ \text{is}\ j\}, \\
+    \Tset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a blue ball}, z(\Node{p})\ \text{is}\ j\}, \\
+    \Zset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a purple ball}, z(\Node{p})\ \text{is}\ j\}, \\
+    \Qset^j(\Gen) &\coloneqq \Masterset^j(\Gen) \cap \Tset^j(\Gen).
   \end{aligned}
@@
 With these definitions,
-$\Masterset(\Gen)$ comprises the internal node times,
-$\Cset(\Gen)$ is the set of branch times,
-$\Lset(\Gen)$ contains the leaf times,
-$\Tset(\Gen)$ holds the sample times,
-$\Zset(\Gen)$ holds the transition times,
-and $\Qset(\Gen)$ is the set of \emph{direct-descent times},
-i.e., the times of samples that are themselves directly ancestral to other samples.
+$\Masterset^j(\Gen)$ comprises the internal node times of nodes that are derived from nodes in Deme $j$,
+$\Cset^j(\Gen)$ is the set of branch times of nodes that are produced by nodes in Deme $j$,
+$\Lset^j(\Gen)$ contains the leaf times where the tips are in Deme $j$,
+$\Tset^j(\Gen)$ holds the sample times for samples in Deme $j$,
+$\Zset^j(\Gen)$ holds the transition times derived from nodes in Deme $j$,
+and $\Qset^j(\Gen)$ is the set of \emph{direct-descent times} derived from samples in Deme $j$.
+Trivially, $\Masterset(\Gen) = \cup_j \Masterset^j(\Gen)$, $\Cset(\Gen) = \cup_j \Cset^j(\Gen)$, $\Lset(\Gen) = \cup_j\Lset^j(\Gen)$, $\Tset(\Gen)=\cup_j\Tset^j(\Gen)$, $\Zset(\Gen)=\cup_j\Zset^j(\Gen)$, and $\Qset(\Gen)=\cup_j\Qset^j(\Gen)$, representing all internal node times, branch times, leaf times, sample times, transistion times, and direct-descent times for $\Gen$, respectively.
 
+\paragraph{Structured genealogy process}
 The structured genealogy evolves over time, so we can proceed to define the \emph{structured genealogy process} $\Gen_t$ now.
+We define $\Gen_t(\omega) = \left(t,\mathrm{geneal}(\restrict{\omega}{t})\right)$, for the end time $t\in\Time$ and the sequence of jumps $\omega\in\Omega$.
+The operation $\mathrm{geneal}$ is defined recursively as follows for $\omega\in\mathring\Omega$:
+
+\fontsize{10pt}{12pt}\selectfont
+\begin{equation}\label{eq:Gdef}
+  \mathrm{geneal}(\omega)\coloneqq
+  \begin{cases}
+    \left(k,0,\{(\green,k),(\black,k)\},0\right)_{k=0}^{I(\X_0(\omega))}, &\text{if}\ K(\omega)=0;\\
+    \mathrm{add}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Bset;\\
+    \mathrm{drop}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Dset;\\
+    \mathrm{sample}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Sset;\\
+    \mathrm{transition}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega),i,j), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Mset_{ij};\\
+    \mathrm{geneal}(\predec{\omega}), &\text{otherwise}.
+  \end{cases}
+\end{equation}
+\normalsize
+
+A deterministic map, $\mathrm{ext}$, can be defined, such that $\Inv_t(\omega)=\mathrm{ext}(\Gen_t(\omega))$ for all $\omega\in\Omega$.
+
+\paragraph{Branch color}
+
+\paragraph{Pruned genealogy process, visible genealogy process}
+We divide the procedure of deriving a sampled genealogy from a full genealogy into two steps: \emph{pruning} and \emph{obscuring}.
+The former one indicates dropping all unsampled individuals (\ie black balls) in a genealogy and remaining the relationship between samples and transition records of samples, and the latter one represents further erasing all deme-transition records of samples and obtaining an ``obscured'' sampled genealogy.
+
+\paragraph{Node color}
+
+\paragraph{Lineage count}
+
+Given a genealogy, $\Vis$, at each time $t\in\Time$, there are a finite number, $\ell^j(t,\Vis)$, of lineages present for Deme $j$ in the genealogy at that time. Evidently, one has
+\begin{equation}\label{eq:elldef}
+  \ell^j(t,\Vis)\coloneqq \sum_{e\in\Cset^j(\Vis)} \indicator{\halfopen{e,\infty}}(t) - \sum_{e\in\Lset^j(\Vis)} \indicator{\halfopen{e,\infty}}(t)=\sum_{e\in\Masterset^j(\Vis)} \indicator{\halfopen{e,\infty}}(t) - \sum_{e\in\Tset^j(\Vis)} \indicator{\halfopen{e,\infty}}(t),
+\end{equation}
+and the lineage count for all demes $\ell(t,\Vis)=\sum_{j=1}^J\ell^j(t,\Vis)$. With this definition, note that both $\ell^j$ and $\ell$ are \cadlag functions of $t$.
 

### 2022-01-18 ms.Rnw
sgp update

@@
 Thus, the population process $\sg{X}_t$ is a stochastic process by defining a sequence of jumps, $\omega=\left(t_k, u_k, n_k\right)_{k=1}^K$, where $K\in\N\cup\{\infty\}$ is the total number of jumps and for the $k$-th jump, $t_k\in\Time$ is the time, $u_k\in\Uspace$ is the event type, and $n_{k}\in\N$ is the label of the individual who conducts $u_k$.
-The full construction can be referred to \citet{King2021}.
-We define the following functions on $\omega\in\Omega$:
+The full construction can be referred to \citet{King2021}. We define the following functions on $\omega\in\Omega$:
 
@@
       \tikzstyle{stoch}=[color=black, dashed, thick, >=stealth]
-      \node (O) at (0,0) {$\Master_t$};
+      \node (A) at (0,0) {$\Master_t$};
       \node (X) at (180:2) {$\X_t$};
-      \node (H) at (210:2) {$\Hist_t$};
-      \node (I) at (240:2) {$\Inv_t$};
-      \node (G) at (270:2) {$\Gen_t$};
-      \node (P) at (300:2) {$\Prune_t$};
-      \node (V) at (330:2) {$\Vis_t$};
-      \node (W) at (360:2) {$\Wch_i$};
-      \draw [det,->] (O) -- (X);
-      \draw [det,->] (O) -- (H);
+      \node (H) at (216:2) {$\Hist_t$};
+      \node (I) at (252:2) {$\Inv_t$};
+      \node (G) at (288:2) {$\Gen_t$};
+      \node (P) at (324:2) {$\Prune_t$};
+      \node (V) at (360:2) {$\Vis_t$};
+      \node (W) at (342:1.5) {$\Wch_i$};
+      \node (O) at (306:1.2) {$\Och_i$};
+      \draw [det,->] (A) -- (X);
+      \draw [det,->] (A) -- (H);
       \draw [det,->] (H) -- (X);
-      \draw [det,->] (O) -- (I);
-      \draw [det,->] (O) -- (G);
-      \draw [det,->] (O) -- (P);
-      \draw [det,->] (O) -- (V);
-      \draw [det,->] (O) -- (W);
+      \draw [det,->] (A) -- (I);
+      \draw [det,->] (A) -- (G);
+      \draw [det,->] (A) -- (V);
+      \draw [det,->] (A) -- (P);
+      \draw [det,->] (A) -- (O);
+      \draw [det,->] (P) -- (O);
+      \draw [det,->] (A) -- (W);
       \draw [det,->] (V) -- (W);
@@
       \draw [stoch,->] (H) -- (V);
+      \draw [stoch,->] (H) -- (P);
+      \draw [stoch,->] (H) -- (O);
       \draw [stoch,->] (H) -- (W);
@@
 
-for all $x, u\in\Xspace$.
-We disallow simultaneous death at a single death event at this stage and define the set of death events in Deme $j$ as $\Dset_j\coloneqq\,D^{-1}(\{1\})$.
+for all $x, u\in\Xspace$. We disallow simultaneous death at a single death event at this stage and define the set of death events in Deme $j$ as $\Dset_j\coloneqq\,D^{-1}(\{1\})$. 
 We also insist, birth and death events don't coincide for any demes, $i,j$, that is,  $\Bset_{j}\cap\Dset_{i}=\emptyset$. 
@@
 The definition of superspreading is ambiguous in epidemiology. 
-Some researchers view it as a certain individual who has extremely strong transmissibility with high reproduction rate and call this individual \emph{superspreader}; others avoid to invidualize this phenomenon and consider it as an event where a cluster of infections appears in a single event within a small amount of time and define it as a \emph{superspreading event}.
+Some researchers view it as a certain individual who has extremely strong transmissibility with high reproduction rate and call this individual \emph{superspreader}; 
+others avoid to invidualize this phenomenon and consider it as an event where a cluster of infections appears in a single event within a small amount of time and define it as a \emph{superspreading event}.
 The first definition can be represented by the SI$^2$R model with two demes of infectious individuals: one with extremely strong transmissibility and the other with weak transmissibility. 
@@
 
-\paragraph{Inventory process, deme process}
+\paragraph{Inventory process}
 First, we define a global set $\Names\coloneqq\N$, which is a collection of infinite unique names. 
@@
 Eventually, we can define $\mathrm{inven}(\omega)\coloneqq\{\mathrm{inven}(\omega,j)\}_{j=1}^J$ as a list of inventories, such that $\Inv_t(\omega)\coloneqq\{\Inv_t^j(\omega)\}_{j=1}^J$.
-We can further define the deme process $\demefuc_t^j(\omega)\coloneqq\frac{\Inv_t^j(\omega)}{\Inv_t(\omega)}$ and $\demefuc_t(\omega)\coloneqq\{\demefuc_t^j(\omega)\}_{j=1}^J$.
 
@@
 We then use $t(\Gen)$, $K(\Gen)$, and $\Node{p}_k(\Gen)$ to represent the time, the length (\ie the number of nodes), and the $k$-th node of the genealogy $\Gen$, respectively. 
+We define $\seq{\Gen}$ as the node sequence of genealogy $\Gen$, where $\Node{p}\in\Gen$ when $\Node{p}$ is one node in $\seq{\Gen}$.
 
 \paragraph{Effect of births, death, sampling and transistions}
+Births, deaths, samples, and transitions change the topology of a structured genealogy. 
+A death at time $t$ indicates the removal of a leaf in the genealogy and the dismiss of a node. The detailed definition can be found in \citet{King2021}.
 
-Births, deaths, samples, and transitions change the topology of a structured genealogy.
-A death at time $t$ indicates the removal of a leaf in the genealogy and the dismiss of a node.
-The detailed definition can be found in \citet{King2021} and here we continue to use the $\mathrm{drop}(\Gen)$ to denote the resulting sequence of nodes.
-
-A birth at time $t$ in a structured genealogy is similar to that in a unstructured one \citep{King2021}, while the new member introduced to the queue of sequence is in the same deme as the parent.
-Let $\Ball{b}$ be the $n$-th black ball, the parent, and there exists a node $\Node{p}\in\Gen$ with $\Ball{b}\in w(\Node{p})=\{\Ball{b},\Ball{b}'\}$ where $\Ball{b}'$ is the other ball held by $\Node{p}$.
-We then can produce a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},z(\Node{p}))$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\black,c)$.
+A birth at time $t$ in a structured genealogy is similar to that in a unstructured one \citep{King2021}: a black ball is selected as the parent and a new node introduced as the child, who is in the same deme as the parent.
+Let $\Ball{b}$ be randomly selected black ball, named \emph{n}, which is the parent, and there exists a node $\Node{p}\in\Gen$ holding $\Ball{b}\in w(\Node{p})=\{\Ball{b},\Ball{b}'\}$ where $\Ball{b}'$ is the other ball held by $\Node{p}$.
+Here, we assume the parent is from Deme $i$ (\ie $n\in\Inv_i$) and the child is born into Deme $j$. 
+We then can produce a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},i)$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, $\Ball{b}''=(\black,c)$, and update $\Inv_j=\Inv_j\cup\{c\}$. 
 By swapping balls between $\Node{p}$ and $\Node{p}'$, it ends up that $w(\Node{p})=\{\Ball{g},\Ball{b}'\}$ and $w(\Node{p}')=\{\Ball{b},\Ball{b}''\}$. 
@@
 
+A death at time $t$ is defined in the same way as that in \citet{King2021} and we continue to use the notation $\mathrm{drop}(\seq{\Gen},n)$ to denote the resulting sequence of nodes for a death, where an individual with name \emph{n} is selected randomly to be removed from the population.
+
 A sample at time $t$ results in the participation of a new node with a new blue ball. 
-The procedure is similar to the previous sampling in \citet{King2021} and the birth mentioned above: the $n$-th black ball held by $\Node{p}$ (\ie $w(\Node{p})=\{\Ball{b},\Ball{b}'\}$) is selected, then a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},z(\Node{p}))$ is generated, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\blue,q)$, at the end $\Node{p}$ exchanges $\Ball{b}$ for $\Ball{g}$ with $\Node{p}'$.
+The procedure is similar to the previous sampling in \citet{King2021} and the birth mentioned above: a black ball, named \emph{n}, where $n\in\Inv_j$, and held by $\Node{p}$ (\ie $w(\Node{p})=\{\Ball{b},\Ball{b}'\}$), is selected at random.
+Then a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},j)$ is generated, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\blue,q)$. 
+At the end, $\Node{p}$ exchanges $\Ball{b}$ for $\Ball{g}$ with $\Node{p}'$. 
 Here, $q$ is the name of the new blue ball, which is the ordinal number of the transition, $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is blue}\}|$. 
-Note that, same as that in a birth, $z(\Node{p}')=z(\Node{p})$.
-We also use the same notation $\mathrm{sample}(\Gen,n)$ as the resulting sequence of nodes.
+We use the notation $\mathrm{sample}(\seq{\Gen},n)$ as the resulting sequence of nodes.
 
 A transition at time $t$ leads to the inclusion of a new node with a new purple ball. 
-Let $\Ball{b}$ denotes the $n$-th black ball selected to indicate an extant in the population transits from his/her/their current deme to a new one.
-Assuming that the unique node $\Node{p}\in\Gen$ baring this selected black ball (\ie $\Ball{b}\in w(\Node{p})=\{\Ball{b},\Ball{b}'\}$) is in Deme $i$ (\ie $z(\Node{p})=i$) and it's transiting to Deme $j$ where $j\neq i$.
-Now we initialize a new node by taking $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},j)$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\purple,q)$.
-Similar to other events, we swap $\Ball{b}$ for $\Ball{g}$ between nodes $\Node{p}$ and $\Node{p}'$ and insert $\Node{p}'$ at the last position of the node-sequence.
-Here, $q$ is the name of the new purple ball, which is the ordinal number of the transition, $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is purple}\}|$.
-We call this new node $\Node{p}'$ the \emph{Transition Nodes} in the genealogy and denote the resulting sequence of nodes by $\mathrm{transition}(\seq{\Gen},n,i,j)$.
+Let $\Ball{b}$ denote the
+black ball with name \emph{n} and selected to indicate an extant in the population transits from his/her current
+deme to a new one. 
+Assuming this selected individual is from Deme $i$ (\ie $i\in\Inv_i$) and transiting to Deme $j$ ($j\neq i$), we then initialize a new node by taking $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},i)$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\purple,q)$.
+Note that, we update $\Inv_i=\Inv_i\setminus\{n\}$ and $\Inv_j=\Inv_j\cup\{n\}$. 
+Similar to other events, we therefore swap $\Ball{b}$ for $\Ball{g}$ between nodes $\Node{p}$ and $\Node{p}'$ and insert $\Node{p}'$ at the last position of the node-sequence. 
+Here, \emph{q} is the name of the new purple ball, which is the ordinal number of the transition,  $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is purple}\}|$. 
+We call this new node $\Node{p}'$ the \emph{Transition Nodes} in the genealogy and denote the resulting sequence of nodes by $\mathrm{transition}(\seq{\Gen},n)$.
 
@@
 We define $\Gen_t(\omega) = \left(t,\mathrm{geneal}(\restrict{\omega}{t})\right)$, for the end time $t\in\Time$ and the sequence of jumps $\omega\in\Omega$. 
-The operation $\mathrm{geneal}$ is defined recursively as follows for $\omega\in\mathring\Omega$:
+We first initialize the global set of names by letting $\Names=\N\setminus\{1,2,\dots,I(\X_0(\omega))\}$, then the operation $\mathrm{geneal}$ is defined recursively as follows for $\omega\in\mathring\Omega$:
 
@@
     \mathrm{sample}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Sset;\\
-    \mathrm{transition}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega),i,j), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Mset_{ij};\\
-    \mathrm{geneal}(\predec{\omega}), &\text{otherwise}.
+    \mathrm{transition}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Mset;\\
+    \mathrm{geneal}(\predec{\omega}), &\text{otherwise};
   \end{cases}
@@
 
+where $\Bset\coloneqq\bigcup_{i,j}\Bset_{ij}$, $\Dset\coloneqq\bigcup_j\Dset_j$, and $\Mset\coloneqq\bigcup_{i,j}\Mset_{ij}$. 
 A deterministic map, $\mathrm{ext}$, can be defined, such that $\Inv_t(\omega)=\mathrm{ext}(\Gen_t(\omega))$ for all $\omega\in\Omega$.
@@
 
-\paragraph{Pruned genealogy process, visible genealogy process}
-We divide the procedure of deriving a sampled genealogy from a full genealogy into two steps: \emph{pruning} and \emph{obscuring}.
-The former one indicates dropping all unsampled individuals (\ie black balls) in a genealogy and remaining the relationship between samples and transition records of samples, and the latter one represents further erasing all deme-transition records of samples and obtaining an ``obscured'' sampled genealogy.
+\paragraph{Pruned genealogy process}
+The sampled genealogy contains only relations among genomic samples. 
+In other words, information in a full genealogy about unsampled individuals and deme transitions are unobserved in a sampled genealogy, which we call it a visible genealogy. 
+We divide the procedure of deriving a visible genealogy from a full genealogy into two separate steps: \emph{pruning} and \emph{obscuring}. 
+The former drops all unsampled individuals (\ie black balls) in a genealogy and remains the relationship between samples and deme-transition records of samples, and the latter further erases all deme-transition records of samples and obtains an obscured genealogy.
+
+Given a genealogy $\Gen$, we define $\Inv=\mathrm{ext}(\Gen)$.
+%% Then. by definition, for each $c\in\Inv$, $(\black,c)\in w(\Gen)$.
+Let $\mathrm{prune}(\Gen)$ be the pruned genealogy derived by applying iterative $\mathrm{drop}$ operation to $\Gen$ for each name $c\in\Inv$.
+Specifically, suppose $\Inv^j=\{n_0^j,\dots,n_{m_j-1}^j\}$ is the inventory for Deme $j$ and $\Inv=\{\Inv^j\}_{j=1}^{J}$, where $J$ is the total number of demes. 
+Then we can define the set of all names in $\Inv$ as $\mathcal{N}\coloneqq\bigcup_{j=1}^J \Inv^j=\{c_0,\dots,c_{m-1}\}$, where $m=\sum_j m_j$.
+Let $\Node{P}_0=\seq\Gen$ and $\Node{P}_{k}=\mathrm{drop}(\Node{P}_{k-1},c_{k-1})$, for $k=1,\dots,m$.
+Then $\mathrm{prune}(\Gen)\coloneqq(t(\Gen),\Node{P}_{m})$.
+\Cref{fig:prune} illustrates the pruning procedure.
+
+For $\omega\in\Omega$, we define the \emph{pruned genealogy process},
+\begin{equation}
+  \Prune_t(\omega)\coloneqq\textrm{prune}(\Gen_t(\omega)).
+\end{equation}
+Note that $\Prune_t$ so defined is a itself a genealogy.
+\Cref{fig:prune}B depicts one example of $\Prune_t$ both graphically (as a tree) and diagrammatically (as a node-sequence).
+
+\paragraph{Obscured genealogy process}
+We first define an operation $\mathrm{reverse}$ on the sequence of nodes of a genealogy as follows: 
+\begin{inparaenum}[(1)]
+  \item a random purple ball $\Ball{b}$ with name $q$ is selected such that there exists a node $\Node{p}$, where $\Ball{b}\in w(\Node{p})=\{\Ball{b}, \Ball{b}'\}$ and $\Ball{b}'$ is the other ball held by $\Node{p}$;
+  \item there also exists a node $\Node{p}'$ such that $\Ball{g}=\left(\green,n(\Node{p})\right)\in w(\Node{p}')=\{\Ball{g},\Ball{b}''\}$ where $\Ball{b}''$ is the other ball held by $\Node{p}'$;
+  \item swapping balls between $\Node{p}$ and $\Node{p}'$ results in $w(\Node{p}')=\{\Ball{b}',\Ball{b}''\}$ and $w(\Node{p})=\{\Ball{g},\Ball{b}\}$;
+  \item we delete $\Node{p}$ from the node sequence.
+\end{inparaenum}
+The resulting sequence of nodes is denoted as $\mathrm{reverse}(\seq{\Gen},q)$.
+
+Suppose for a genealogy $\Gen$, the set of the serial number of the purple balls is $\{0,1,\dots,q\}$ and the list of inventories is $\Inv=\{\Inv^j\}_{j=1}^J$.
+$\mathrm{obs}(\Gen)$ applies the following steps on $\Gen$:
+\begin{inparaenum}[(a)]
+  \item let $\Node{P}_0=\seq{\Gen}$ and $\Node{P}_{k}=\mathrm{reverse}(\Node{P}_{k-1},k)$, for $k=1,\dots,q$; and
+  \item for $\Node{P}_q$, we update $\Inv=\bigcup_{j}\Inv^j$ and for each node $\Node{p}\in\seq{\Gen}$, $z(\Node{p})=0$.
+\end{inparaenum}
+
+For $\omega\in\Omega$, we define the \emph{obscured genealogy process},
+\begin{equation}
+  \Obscure_t(\omega)\coloneqq\textrm{obs}(\Gen_t(\omega)).
+\end{equation}
+Note that $\Obscure_t$ so defined is a itself a genealogy.
+\Cref{fig:obs}B depicts one example of $\Obscure_t$ both graphically (as a tree) and diagrammatically (as a node-sequence).
+
+\paragraph{Visible genealogy process}
 
 \paragraph{Node color}
+There are at most three kinds of nodes in a visible genealogy, distinguished by the contents of their pockets:
+\begin{inparaenum}[(a)]
+\item \emph{green nodes}, which have two green balls in their pockets;
+\item \emph{blue nodes}, which have one blue and one green ball;
+\item \emph{red nodes}, which have one red and one blue ball.
+\end{inparaenum}
+This distinction allows a compact graphical representation of the visible genealogy (\cref{fig:prune}C) and proves useful in deriving our main results as well.
+Green nodes correspond to \emph{coalescence points} (branch points) in the visible genealogy;
+red nodes, to leaves;
+blue nodes correspond to direct-descent events, i.e., samples which are directly ancestral to other samples.
+Thus, if $\Vis$ is a visible genealogy,
+$\Eset(\Vis)=\Cset(\Vis)\cup\Qset(\Vis)\cup\Lset(\Vis)$, where
+$\Cset(\Vis)$ is the set of times of the green nodes,
+$\Qset(\Vis)$ contains the times of the blue nodes, and
+$\Lset(\Vis)$ holds the times of the red nodes.
+Moreover, $\Tset(\Vis)=\Qset(\Vis)\cup\Lset(\Vis)$ is the set of sample times and $\Masterset(\Vis)=\Qset(\Vis)\cup\Cset(\Vis)$ comprises the \emph{attachment times},
+i.e., those points at which a sampled lineage attaches to the visible genealogy subtended by earlier samples.
 
-\paragraph{Lineage count}
+\paragraph{Deme proportion and lineage count}
+
+We can further define the deme proportion function $\demefuc(t,\omega)\coloneqq\{\demefuc^j(t,\omega)\}_{j=1}^J$, where $\demefuc^j(t,\omega)\coloneqq\frac{\vert\Inv_t^j(\omega)\vert}{\vert\Inv_t(\omega)\vert}$.
+
+\QYL{Lineage count function may not be a \cadlag}
 

### 2022-01-28 ms.Rnw
sgp results updated

@@
       \node (A) at (0,0) {$\Master_t$};
-      \node (X) at (180:2) {$\X_t$};
-      \node (H) at (216:2) {$\Hist_t$};
-      \node (I) at (252:2) {$\Inv_t$};
-      \node (G) at (288:2) {$\Gen_t$};
-      \node (P) at (324:2) {$\Prune_t$};
-      \node (V) at (360:2) {$\Vis_t$};
-      \node (W) at (342:1.5) {$\Wch_i$};
-      \node (O) at (306:1.2) {$\Och_i$};
-      \draw [det,->] (A) -- (X);
+      \node (H) at (30:1) {$\Hist_t$};
+      \node (X) at (30:2) {$\X_t$};
+      \node (G) at (-30:1) {$\Gen_t$};
+      \node (I) at (-30:2) {$\Inv_t$};
+      \node (P) at (0:1.732) {$\Prune_t$};
+      \node (O) at (1.732,.5) {$\Och_i$};
+      \node (V) at (0:2.732) {$\Vis_t$};
+      \node (W) at (2.732,.5) {$\Wch_i$};
       \draw [det,->] (A) -- (H);
       \draw [det,->] (H) -- (X);
-      \draw [det,->] (A) -- (I);
       \draw [det,->] (A) -- (G);
-      \draw [det,->] (A) -- (V);
-      \draw [det,->] (A) -- (P);
-      \draw [det,->] (A) -- (O);
-      \draw [det,->] (P) -- (O);
-      \draw [det,->] (A) -- (W);
-      \draw [det,->] (V) -- (W);
       \draw [det,->] (G) -- (I) node[midway,below,sloped] {$\mathrm{ext}$};
@@
       \draw [det,->] (P) -- (V) node[midway,below,sloped] {$\mathrm{obs}$};
-      \draw [stoch,->] (H) -- (I);
+      \draw [det,->] (P) -- (O);
+      \draw [det,->] (V) -- (W);
       \draw [stoch,->] (H) -- (G);
-      \draw [stoch,->] (H) -- (V);
-      \draw [stoch,->] (H) -- (P);
-      \draw [stoch,->] (H) -- (O);
-      \draw [stoch,->] (H) -- (W);
     \end{tikzpicture}
@@
 
-\paragraph{Transitions}
-Infective demes are interchangeable and we define the transition function from Deme $i$ to $j$ for any $i\neq j$ and $i,j=1,\dots,J$ as $M_{ij}:\Uspace\to\N$ to represent the number of infectives transiting to Deme $j$ from Deme $i$.
-Thus the set or transition is $\Mset_{ij}\coloneqq\,M^{-1}_{ij}(\{1\})$ since we assume single transition at a single transition event. 
-Similarly, we disallow coninciding transition events and other events.
+\paragraph{Migrations}
+Infective demes are interchangeable and we define the migration function from Deme $i$ to $j$ for any $i\neq j$ and $i,j=1,\dots,J$ as $M_{ij}:\Uspace\to\N$ to represent the number of infectives migrating to Deme $j$ from Deme $i$.
+Thus the set of migration is $\Mset_{ij}\coloneqq\,M^{-1}_{ij}(\{1\})$ since we assume single migration at a single migration event. 
+Similarly, we disallow coninciding migration events and other events.
 
@@
 The population, the set of infected individuals in the context of epidemiology, is structured, categorized into interchangeable demes. 
-We can then define population size function and death functions in Deme $j$, respectively, as $I_{j}, D_{j}:\Xspace\to\N$ for any $j=1,\cdots,J$. 
+We can then define population size function and death functions in Deme $j$, respectively, as $I^{j}, D_{j}:\Xspace\to\N$ for any $j=1,\cdots,J$. 
 \begin{equation}
-  \jumprate_u(t,x) > 0 \Longrightarrow I_j(x+u)-I_j(x)=\sum_{i}\,B_{ij}(u)+\sum_{i}M_{ij}(u)-D_j(u)-\sum_{i}M_{ji}(u),
+  \jumprate_u(t,x) > 0 \Longrightarrow I^j(x+u)-I^j(x)=\sum_{i}\,B_{ij}(u)+\sum_{i}M_{ij}(u)-D_j(u)-\sum_{i}M_{ji}(u),
 \end{equation}
@@
 We also insist, birth and death events don't coincide for any demes, $i,j$, that is,  $\Bset_{j}\cap\Dset_{i}=\emptyset$. 
-Eventually, we can have the function for total population size by letting $I(x)\coloneqq\sum_{j=1}^J\,I_j(x)$ and it's trivial that
+Eventually, we can have the function for total population size by letting $I(x)\coloneqq\sum_{j=1}^J\,I^j(x)$ and it's trivial that
 
@@
 In SI$^2$R model, we have two different while interchangeable demes of infectious individuals, the \emph{per capita} transmissibility therefore in the first deme is $b_1(t)$ whilst in the second deme is $b_2(t)$.
-Transition rates between Deme $i$ and $j$ are also defined: Deme $i$ infection turns into Deme $j$ at rate $r_{ij}$ for $i,j=1,2$ and $i\neq j$.
+Migration rates between Deme $i$ and $j$ are also defined: Deme $i$ infection turns into Deme $j$ at rate $r_{ij}$ for $i,j=1,2$ and $i\neq j$.
 Furthermore, we also suppose that, once being infected, a susceptible individual becomes Deme 1 infectious with probability $\rho$ while Deme 2 infectious with probability $1-\rho$.
-To specify each event and the number of individuals involved, we define a relatively complex state vector by setting $d=10$ and $\sg{X}=\left(s,i_1,i_2,b_{11},b_{12},b_{21},b_{22},m_{12},m_{21},g\right)$, for $s$, $i_1$, and $i_2$ being the population size of the susceptible, the infectives in Deme 1, and those in Deme 2, respectively, and $b_{11}$, $b_{12}$, $b_{21}$, $b_{22}$, $m_{12}$, and $m_{21}$ being the cumulative number of events of parents in Deme 1 giving to children in Deme 1 and 2, parents in Deme 2 giving birth to children in Deme 1 and 2, individuals transiting from Deme 1 to Deme 2, and vice versa, and sampling, respectively.
+To specify each event and the number of individuals involved, we define a relatively complex state vector by setting $d=10$ and $\sg{X}=\left(s,i_1,i_2,b_{11},b_{12},b_{21},b_{22},m_{12},m_{21},g\right)$, for $s$, $i_1$, and $i_2$ being the population size of the susceptible, the infectives in Deme 1, and those in Deme 2, respectively, and $b_{11}$, $b_{12}$, $b_{21}$, $b_{22}$, $m_{12}$, and $m_{21}$ being the cumulative number of events of parents in Deme 1 giving to children in Deme 1 and 2, parents in Deme 2 giving birth to children in Deme 1 and 2, individuals migrating from Deme 1 to Deme 2, and vice versa, and sampling, respectively.
 In this case, we can summarize the event rates as:
@@
 
-The nine non-zero rates above represent those of infection into Deme 1 infected by individuals in Deme 1, infection into Deme 2 infected by individuals in Deme 1, infection in Deme 1 infected by individuals in Deme 2, infection into Deme 2 infected by individuals in Deme 1, transition from Deme 1 to Deme 2, transition from Deme 2 to Deme 1, recovery from Deme 1, recovery from Deme 2, and sampling, respectively.
+The nine non-zero rates above represent those of infection into Deme 1 infected by individuals in Deme 1, infection into Deme 2 infected by individuals in Deme 1, infection in Deme 1 infected by individuals in Deme 2, infection into Deme 2 infected by individuals in Deme 1, migration from Deme 1 to Deme 2, migration from Deme 2 to Deme 1, recovery from Deme 1, recovery from Deme 2, and sampling, respectively.
 We then summarize the population functions and sets as follows:
@@
 In other words, an infectious individual from another population join our interested population at a certain point of time $\tau$ and then stay and spread the disease. 
-In this case, if this specific individual is defined as Deme 1 while other infectious individuals as Deme 2, the transition rate $r_{12}$ is a Dirac delta function $\delta(t)$, where $\delta(t)=\infty$ for $t=\tau$ and $\delta(t)=0$ for $t\neq\tau$, and $r_{21}=0$ all the time.
+In this case, if this specific individual is defined as Deme 1 while other infectious individuals as Deme 2, the migration rate $r_{12}$ is a Dirac delta function $\delta(t)$, where $\delta(t)=\infty$ for $t=\tau$ and $\delta(t)=0$ for $t\neq\tau$, and $r_{21}=0$ all the time.
 
@@
   \begin{cases}
-    \mathrm{initialize}(I_j(\sg{X}_0(\omega)), & \text{if }K(\omega)=0; \\
+    \mathrm{initialize}(I^j(\sg{X}_0(\omega)), & \text{if }K(\omega)=0; \\
     \mathrm{add}(\mathrm{inven}(\predec{\omega},j)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Bset_j;\\
@@
 \paragraph{Structured genealogies}
-Types of nodes are defined in a structured genealogy: \emph{Leaves} are the extant members at current time and \emph{Internal Nodes} represent ancestors to a certain subset of extant members, which we further define \emph{Sampling Nodes} and \emph{Transition Nodes}, whose differences will be clear in a moment.
+Types of nodes are defined in a structured genealogy: \emph{Leaves} are the extant members at current time and \emph{Internal Nodes} represent ancestors to a certain subset of extant members, which we further define \emph{Sampling Nodes} and \emph{Migration Nodes}, whose differences will be clear in a moment.
 
@@
 For a node $\Node{p}$, we will use $n(\Node{p})$, $t(\Node{p})$, $w(\Node{p})$, and $z(\Node{p})$ to denote the name, time, pocket, and deme of $\Node{p}$, respectively. 
-We then assign black, green, blue, and purple balls to serve as pointers to the extant population at time $t$, to the internal nodes, to the samples, and to the transitions between demes, respectively.
+We then assign black, green, blue, and purple balls to serve as pointers to the extant population at time $t$, to the internal nodes, to the samples, and to the migrations between demes, respectively.
 
@@
 
-\paragraph{Effect of births, death, sampling and transistions}
-Births, deaths, samples, and transitions change the topology of a structured genealogy. 
+\paragraph{Effect of births, death, sampling and migration}
+Births, deaths, samples, and migrations change the topology of a structured genealogy. 
 A death at time $t$ indicates the removal of a leaf in the genealogy and the dismiss of a node. The detailed definition can be found in \citet{King2021}.
@@
 Here, we assume the parent is from Deme $i$ (\ie $n\in\Inv_i$) and the child is born into Deme $j$. 
-We then can produce a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},i)$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, $\Ball{b}''=(\black,c)$, and update $\Inv_j=\Inv_j\cup\{c\}$. 
+We then can produce a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},i)$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, $\Ball{b}''=(\black,c)$.
 By swapping balls between $\Node{p}$ and $\Node{p}'$, it ends up that $w(\Node{p})=\{\Ball{g},\Ball{b}'\}$ and $w(\Node{p}')=\{\Ball{b},\Ball{b}''\}$. 
@@
 
-A death at time $t$ is defined in the same way as that in \citet{King2021} and we continue to use the notation $\mathrm{drop}(\seq{\Gen},n)$ to denote the resulting sequence of nodes for a death, where an individual with name \emph{n} is selected randomly to be removed from the population.
+% A death at time $t$ is defined in the same way as that in \citet{King2021} and we continue to use the notation $\mathrm{drop}(\seq{\Gen},n)$ to denote the resulting sequence of nodes for a death, where an individual with name \emph{n} is selected randomly to be removed from the population.
+
+A death at time $t$ is quite different from that defined in \citet{King2021}, while we continue to use the notation $\mathrm{drop}(\seq{\Gen},n)$ to denote the resulting sequence of nodes for a death, where an individual with name \emph{n} is selected randomly to be removed from the current population.
+When a death occurs, a black ball $\Ball{b}$ with name \emph{n} is selected arbitrarily, whose holder is node $\Node{p}$, \ie $\Ball{b}\in w(\Node{p})=\{\Ball{b},\Ball{b}'\}$. 
+There exists an unique $\green$ ball and an unique node $\Node{p}'$ such that $\Ball{g}=(\green, n(p))$ and $\Ball{g}\in w(\Node{p}')=\{\Ball{g},\Ball{b}''\}$. 
+The $\mathrm{drop}(\seq{\Gen},n)$ operation then proceeds differently, conditioned on the color and the deme of the other ball, $\Ball{b}'$, held by $\Node{p}$: 
+\begin{inparaenum}[(1)]
+  \item if $\Ball{b}'$ is $\green$, swap $\Ball{g}$ and $\Ball{b}'$ between $\Node{p}$ and $\Node{p}'$ then remove $\Node{p}$ from the sequence of nodes;
+  \item if $\Ball{b}'$ is $\black$, the above procedure is conducted when $\Ball{b}'$ and $\Node{p}$ are in the same deme, otherwise, replace $\Ball{b}$ by a new purple ball $\Ball{c}=\{\purple, q\}$ where $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is purple}\}|$;
+  \item if $\Ball{b}'$ is $\blue$, replace $\Ball{b}$ by a new red ball with the name of $\Ball{b}'$;
+  \item if $\Ball{b}'$ is $\purple$, change the deme of $\Ball{b}$ to that of $\Node{p}$, swap $\Ball{g}$ and $\Ball{b}$, and delete $\Node{p}$.
+\end{inparaenum}
+By doing aforementioned steps, we can keep track of the sampled individuals and the migration at a birth.
 
@@
 At the end, $\Node{p}$ exchanges $\Ball{b}$ for $\Ball{g}$ with $\Node{p}'$. 
-Here, $q$ is the name of the new blue ball, which is the ordinal number of the transition, $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is blue}\}|$. 
+Here, $q$ is the name of the new blue ball, which is the ordinal number of the migration, $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is blue}\}|$. 
 We use the notation $\mathrm{sample}(\seq{\Gen},n)$ as the resulting sequence of nodes.
 
-A transition at time $t$ leads to the inclusion of a new node with a new purple ball. 
+A migration at time $t$ leads to the inclusion of a new node with a new purple ball. 
 Let $\Ball{b}$ denote the
-black ball with name \emph{n} and selected to indicate an extant in the population transits from his/her current
+black ball with name \emph{n} and selected to indicate an extant in the population migrates from his/her current
 deme to a new one. 
-Assuming this selected individual is from Deme $i$ (\ie $i\in\Inv_i$) and transiting to Deme $j$ ($j\neq i$), we then initialize a new node by taking $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},i)$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\purple,q)$.
+Assuming this selected individual is from Deme $i$ (\ie $i\in\Inv_i$) and migrating to Deme $j$ ($j\neq i$), we then initialize a new node by taking $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},i)$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\purple,q)$.
 Note that, we update $\Inv_i=\Inv_i\setminus\{n\}$ and $\Inv_j=\Inv_j\cup\{n\}$. 
 Similar to other events, we therefore swap $\Ball{b}$ for $\Ball{g}$ between nodes $\Node{p}$ and $\Node{p}'$ and insert $\Node{p}'$ at the last position of the node-sequence. 
-Here, \emph{q} is the name of the new purple ball, which is the ordinal number of the transition,  $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is purple}\}|$. 
-We call this new node $\Node{p}'$ the \emph{Transition Nodes} in the genealogy and denote the resulting sequence of nodes by $\mathrm{transition}(\seq{\Gen},n)$.
+Here, \emph{q} is the name of the new purple ball, which is the ordinal number of the migration,  $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is purple}\}|$. 
+We call this new node $\Node{p}'$ the \emph{Migration Nodes} in the genealogy and denote the resulting sequence of nodes by $\mathrm{migrate}(\seq{\Gen},n)$.
 
-\paragraph{Event times}
+\paragraph{Genealogical event times}
+
+\QYL{the definition for sets of event times is problematic.}
 
@@
     \Eset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})\ \text{is}\ j\}, \\
-    \Masterset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a green ball}, z(\Node{p})\ \text{is}\ j\}, \\
-    \Cset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains two green balls}, z(\Node{p})\ \text{is}\ j\}, \\
-    \Lset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains no green balls}, z(\Node{p})\ \text{is}\ j\}, \\
-    \Tset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a blue ball}, z(\Node{p})\ \text{is}\ j\}, \\
-    \Zset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, w(\Node{p})\ \text{contains a purple ball}, z(\Node{p})\ \text{is}\ j\}, \\
+    \Masterset^j(\Gen) &\coloneqq , \\
+    \Cset^{j,j'}(\Gen) &\coloneqq , \\
+    \Lset^j(\Gen) &\coloneqq , \\
+    \Tset^j(\Gen) &\coloneqq , \\
+    \Zset^{j,j'}(\Gen) &\coloneqq , \\
     \Qset^j(\Gen) &\coloneqq \Masterset^j(\Gen) \cap \Tset^j(\Gen).
@@
 $\Masterset^j(\Gen)$ comprises the internal node times of nodes that are derived from nodes in Deme $j$,
-$\Cset^j(\Gen)$ is the set of branch times of nodes that are produced by nodes in Deme $j$,
+$\Cset^{j,j'}(\Gen)$ is the set of coalescent times of nodes in Deme $j$ and containing a ball in Deme $j'$,
 $\Lset^j(\Gen)$ contains the leaf times where the tips are in Deme $j$,
 $\Tset^j(\Gen)$ holds the sample times for samples in Deme $j$,
-$\Zset^j(\Gen)$ holds the transition times derived from nodes in Deme $j$,
+$\Zset^{j,j'}(\Gen)$ holds the deme-transition times derived from nodes in Deme $j$ and containing a ball in Deme $j'$,
 and $\Qset^j(\Gen)$ is the set of \emph{direct-descent times} derived from samples in Deme $j$.
-Trivially, $\Masterset(\Gen) = \cup_j \Masterset^j(\Gen)$, $\Cset(\Gen) = \cup_j \Cset^j(\Gen)$, $\Lset(\Gen) = \cup_j\Lset^j(\Gen)$, $\Tset(\Gen)=\cup_j\Tset^j(\Gen)$, $\Zset(\Gen)=\cup_j\Zset^j(\Gen)$, and $\Qset(\Gen)=\cup_j\Qset^j(\Gen)$, representing all internal node times, branch times, leaf times, sample times, transistion times, and direct-descent times for $\Gen$, respectively.
 
@@
     \mathrm{sample}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Sset;\\
-    \mathrm{transition}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Mset;\\
+    \mathrm{migrate}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Mset;\\
     \mathrm{geneal}(\predec{\omega}), &\text{otherwise};
@@
 
-\paragraph{Branch color}
-
-\paragraph{Pruned genealogy process}
+\paragraph{Visible genealogy process}
 The sampled genealogy contains only relations among genomic samples. 
-In other words, information in a full genealogy about unsampled individuals and deme transitions are unobserved in a sampled genealogy, which we call it a visible genealogy. 
+In other words, information in a full genealogy about unsampled individuals and deme-transitions are unobserved in a sampled genealogy, which we call it a visible genealogy. 
 We divide the procedure of deriving a visible genealogy from a full genealogy into two separate steps: \emph{pruning} and \emph{obscuring}. 
@@
 
-\paragraph{Obscured genealogy process}
-We first define an operation $\mathrm{reverse}$ on the sequence of nodes of a genealogy as follows: 
+Before we proceed to produce the visible genealogy, we first define an operation $\mathrm{reverse}$ on the sequence of nodes of a genealogy as follows: 
 \begin{inparaenum}[(1)]
@@
 
-Suppose for a genealogy $\Gen$, the set of the serial number of the purple balls is $\{0,1,\dots,q\}$ and the list of inventories is $\Inv=\{\Inv^j\}_{j=1}^J$.
-$\mathrm{obs}(\Gen)$ applies the following steps on $\Gen$:
+Suppose for a pruned genealogy $\Gen'$, the set of the serial number of the purple balls is $\{0,1,\dots,q\}$ and the list of inventories is $\Inv=\{\Inv^j\}_{j=1}^J$.
+$\mathrm{obs}(\Gen')$ applies the following steps on $\Gen'$:
 \begin{inparaenum}[(a)]
-  \item let $\Node{P}_0=\seq{\Gen}$ and $\Node{P}_{k}=\mathrm{reverse}(\Node{P}_{k-1},k)$, for $k=1,\dots,q$; and
+  \item let $\Node{P}_0=\seq{\Gen'}$ and $\Node{P}_{k}=\mathrm{reverse}(\Node{P}_{k-1},k)$, for $k=1,\dots,q$; and
   \item for $\Node{P}_q$, we update $\Inv=\bigcup_{j}\Inv^j$ and for each node $\Node{p}\in\seq{\Gen}$, $z(\Node{p})=0$.
@@
 
-For $\omega\in\Omega$, we define the \emph{obscured genealogy process},
+For $\omega\in\Omega$, we define the \emph{visible genealogy process},
 \begin{equation}
-  \Obscure_t(\omega)\coloneqq\textrm{obs}(\Gen_t(\omega)).
+  \Vis_t(\omega)\coloneqq\textrm{obs}(\textrm{prune}(\Gen_t(\omega))).
 \end{equation}
-Note that $\Obscure_t$ so defined is a itself a genealogy.
-\Cref{fig:obs}B depicts one example of $\Obscure_t$ both graphically (as a tree) and diagrammatically (as a node-sequence).
-
-\paragraph{Visible genealogy process}
+Note that $\Vis_t$ so defined is a itself a genealogy.
+\Cref{fig:obs}B depicts one example of $\Vis_t$ both graphically (as a tree) and diagrammatically (as a node-sequence).
 
 \paragraph{Node color}
-There are at most three kinds of nodes in a visible genealogy, distinguished by the contents of their pockets:
+In a visible genealogy, we can define four kinds of nodes based on the colors of balls in the pockets:
 \begin{inparaenum}[(a)]
-\item \emph{green nodes}, which have two green balls in their pockets;
-\item \emph{blue nodes}, which have one blue and one green ball;
-\item \emph{red nodes}, which have one red and one blue ball.
+  \item \emph{green nodes}, which have more than one green balls in their pockets;
+  \item \emph{blue nodes}, which have one blue;
+  \item \emph{red nodes}, which have one red;
+  \item \emph{purple nodes}, which have one purple ball.
 \end{inparaenum}
-This distinction allows a compact graphical representation of the visible genealogy (\cref{fig:prune}C) and proves useful in deriving our main results as well.
-Green nodes correspond to \emph{coalescence points} (branch points) in the visible genealogy;
-red nodes, to leaves;
-blue nodes correspond to direct-descent events, i.e., samples which are directly ancestral to other samples.
-Thus, if $\Vis$ is a visible genealogy,
-$\Eset(\Vis)=\Cset(\Vis)\cup\Qset(\Vis)\cup\Lset(\Vis)$, where
-$\Cset(\Vis)$ is the set of times of the green nodes,
-$\Qset(\Vis)$ contains the times of the blue nodes, and
-$\Lset(\Vis)$ holds the times of the red nodes.
-Moreover, $\Tset(\Vis)=\Qset(\Vis)\cup\Lset(\Vis)$ is the set of sample times and $\Masterset(\Vis)=\Qset(\Vis)\cup\Cset(\Vis)$ comprises the \emph{attachment times},
-i.e., those points at which a sampled lineage attaches to the visible genealogy subtended by earlier samples.
+In a \emph{compact} visible genealogy, green, blue, red, and purple nodes correspond to \emph{coalescent points}, leaves, direct-descent events, and deme-transition events, respectively.
+Thus, for a visible genealogy $\Vis$ and any deme $j$:
+
+\paragraph{Branch linetype}
 
@@
 
-\QYL{Lineage count function may not be a \cadlag}
+Given a pruned genealogy, $\Prune$, at each time $t\in\Time$, there are a finite number, $\ell^{j}(t, \Prune)$, of lineages in Deme $j$ present in the genealogy at that time.  Evidently, one has
 
-Given a genealogy, $\Vis$, at each time $t\in\Time$, there are a finite number, $\ell^j(t,\Vis)$, of lineages present for Deme $j$ in the genealogy at that time. Evidently, one has
-\begin{equation}\label{eq:elldef}
-  \ell^j(t,\Vis)\coloneqq \sum_{e\in\Cset^j(\Vis)} \indicator{\halfopen{e,\infty}}(t) - \sum_{e\in\Lset^j(\Vis)} \indicator{\halfopen{e,\infty}}(t)=\sum_{e\in\Masterset^j(\Vis)} \indicator{\halfopen{e,\infty}}(t) - \sum_{e\in\Tset^j(\Vis)} \indicator{\halfopen{e,\infty}}(t),
+\begin{equation}
+  \ell^j(t,\Prune)\coloneqq
 \end{equation}
-and the lineage count for all demes $\ell(t,\Vis)=\sum_{j=1}^J\ell^j(t,\Vis)$. With this definition, note that both $\ell^j$ and $\ell$ are \cadlag functions of $t$.
+
+\paragraph{Embedded chains}
+Due to discrete sample times, defined as $S_i$, we now proceed to consider the embedded chain for pruned genealogy process $\Prune_t$ and then for visible genealogy process $\Vis_t$. 
+For $\omega\in\Omega$, let $\Och_i(\omega)\coloneqq\Prune_{S_i(\omega)}(\omega)$ be the embedded chain of the pruned genealogy process, and $\Wch_i(\omega)\coloneqq\Vis_{S_i(\omega)}(\omega)$ be that of the visible genealogy process.
+At each sample time $S_i$, one extra lineage (\ie the $i$-th sample) is to be attached to the previous existing genealogy (\ie $\Och_i$ or $\Wch_i$) at a random attachment time, denoted by $A_i$. 
+Both $\Och_i$ and $\Wch_i$ are Markovian.
 
@@
 \section{Results}
+\label{sec:results}
+The derivation of the probability density of $\Wch_i$ given $\Hist_{S_i}$ can be divided into two steps:
+\begin{align}
+  P_{\Wch_i|\Hist}&=\sum_{\Och_i}\,P_{\Wch_i|\Och_i}P_{\Och_i|\Hist} \\
+                  &=\sum_{\Och_i}\,P_{\Wch_i|\Och_i}\left(P_{\Och_1|\Hist}\prod_{q=2}^{i}P_{\Och_q|\Och_{q-1}}\right) 
+\end{align}
+% Trivially, $P_{\Och_1|\Hist}=1$ and $P_{\Wch_1|\Och_1}=1$ such that $P_{\Wch_1|\Hist}=1$.
+
+We first compute $P_{\Och_q|\Och_{q-1},\Hist}$ by fixing $A_q=a_q$, $\Och_q=o_q$, $S_q=s_q$, and $\Hist_{s_q}=h=\left(s_q,\left(t_k,u_k\right)_{k=0}^K\right)$.
+For convenience, we write $x_k=\X(t_k)$, and define the sets
+$\Omega_q\coloneqq\{\omega\in\Omega\;|\;\Wch_{q}(\omega)=w_q,\Hist_{s_i}(\omega)=h\}\subset\Omega$ and
+$\Omega_{qk}\coloneqq N_k(\Omega_q)\subset\N$, for $q\le i$ and $k\le K$.
+
+\begin{equation*}
+  \begin{aligned}
+    % P_{\Och_q|\Och_{q-1},\Hist}(o_q|o_{q-1},h)&=\prod_{k=1}^K\sum_{n_k\in\Omega_{qk}}\auxmeas_{u_k,x_k-u_k}(n_k)=\prod_{k=1}^K\sum_{n_k\in\Omega_{qk}}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv^j_{t_{k-1}}(\Omega_q)}(n_k)}{I_j(x_k-u_k)}\right).
+    P_{\Och_q|\Och_{q-1},\Hist}(o_q|o_{q-1},h) &= \prod_{k=1}^K\sum_{\omega\in\Omega_q}\auxmeas_{u_k,x_k-u_k}(N_k(\omega)) = \prod_{k=1}^K\sum_{\omega\in\Omega_q}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}\right),
+  \end{aligned}
+\end{equation*}
+
+due to that $\auxmeas$ is uniform within any certain deme.
+
+We define $\phi_{qk}\coloneqq\sum_{\omega\in\Omega_q}\auxmeas_{u_k,x_k-u_k}(N_k(\omega)) = \prod_{k=1}^K\sum_{\omega\in\Omega_q}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}\right)$.
+
+Note that $\ell_{1k}=0$ for all $k$.
+
+All cases can be categorized into four groups based on the occurrence of population event within time $\halfopen{a_q,s_q}$: no population event or simultaneous events, a birth from Deme $j$ to $j'$, a sample of Deme $j'$, and a migration from Deme $j$ to $j'$ where $j\neq j'$.
+Within each group, we investigate $\phi_{qk}^{j''}$ by considering whether the corresponding population event was presented in the pruned genealogy and involving the $q$-th lineage sample, which was in Deme $j''$.
+For $q>1$, we then have:
+
+\begin{enumerate}[(a)]
+  \item If $t_k\notin\halfopen{a_q,s_q}$, or if $u_k\notin\Bset\cup\Sset\cup\Mset$, so we have $\phi_{qk}^{j''}=1$.
+  \item If $t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1})$, then again, $\phi_{qk}^{j''}=1$.
+  \item If $j''\notin\{j,j'\}$, then the new lineage sample was not involved in any population events involving Demes $j$ and $j'$ at time $t_k$ and $\phi_{qk}^{j''}=1$.
+  \item \label{item:M2Z} If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Mset^{j,j'}$, then there was a migration event at time $t_k$ and the $j$-th sample lineage was involved.
+  Whether this migration event was presented as a deme-transition event in $\Och_q$ depends on whether $j''=j'$.
+  The population size of Deme $j$ right before time $t_k$ was $I_{k-1}^j$, while that of Deme $j'$ right after $t_k$ was $I_k^{j'}$.
+  Lineages of $\Och_{q-1}$ were not involved in this migration event, so they were excluded.
+  Thus, at time $t_k$, one of the $I_{k-1}^{j}-\ell_{q,k-1}^{j}$ individuals was chosen to migrate and right after, among $I_{k}^{j'}-\ell_{qk}^{j'}$ unobserved individuals, the $j$-th sample was the exact one just migrated.
+  Therefore, $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
+  \item If $t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'}$, the logic is the same as that in case \ref{item:M2Z}, but the migration event was not presented in $\Och_q$.
+One can avoid the exact genealogical deme-transition induced by a migration at time $t_k$, with $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
+  \item \label{item:G2D} If $t_k=a_q\in\Lset^{j}(\Och_{q-1})$, and obviously, $u_k\in\Sset_j$.
+    A individual in Deme $j$ was sampled at $t_k$ and the $q$-th lineage sample was exactly the direct-descend of it, \ie a direct-descend event caused by the $q$-th lineage sample was presented in $\Och_q$ at time $t_k$.
+
+  \item
+  \item If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Bset^{j,j'}$, then there was a birth event at time $t_k$ and the $j$-th sample lineage was involved.
+    However, this birth event was observed as a deme-transition event (\ie $j\neq j'$) in $\Och_q$, instead of a coalescent event. 
+    This birth event resulted in a newly added node holding two black balls, which were separately from Deme $j$ and Deme $j'$.
+    Thus, the population size of Deme $j$ and $j'$, right after $t_k$, were $I_k^j$ and $I_k^{j'}$, respectively. 
+    There were in total $(I_k^{j}-\ell_{qk}^{j})(I_k^{j'}-\ell_{qk}^{j'})$ combinations of the black balls, because none of the lineages in $\Och_{q-1}$ was involved. 
+    Among all these combinations, the $j$-th sample lineage was exactly the black ball newly added to Deme $j'$ after the birth event.
+    Therefore, $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_k^{j}-\ell_{qk}^{j})(I_k^{j'}-\ell_{qk}^{j'})\right)$.
+  \item
+  \item
+\end{enumerate}
+
+To summarize, we have
+\begin{equation}\label{eq:qcases}
+  \phi_{qk}^{j''}=
+  \begin{cases}
+    1, &\text{if}\ t_k\notin\halfopen{a_q,s_q}\ \text{or}\ u_k\notin\Bset\cup\Sset\cup\Mset,\\
+
+    % deme-transition time can evolve to coalescent time, when the new sample is the parent of a "cross-deme" birth
+    1, &\text{if}\ t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1}),\\[12pt]
+    1, &\text{if}\ j''\notin\{j,j'\}\\[12pt]
+    \frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
+    1-\frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
+    \frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k=a_q\in\Lset^{j'}(\Och_{q-1}),\\[12pt]
+    1-\frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})\ \text{and}\ u_k\in\Sset^{j'},\\[12pt]
+    \frac{\delta_{j',j''}}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
+
+    % B2C: either j''=j' or j''=j, the numerator = 1 when j'!=j and =2 when j'=j.
+    \frac{1+\delta_{j,j'}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\\[12pt]
+    1-\frac{(1+\delta_{j,j'})\ell_{qk}^{j}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}-\frac{1}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
+    0, &\text{otherwise}.
+  \end{cases}
+\end{equation}
+
+We then can establish our first main result:
+
+\begin{thm}\label{thm:prunehthm1}
+  With the definitions as above, $P_{\Och_i|\Hist}(w|h) = \prod_{q=1}^{i}\prod_{k=1}^{K}\phi_{qk}$.
+\end{thm}
 

### 2022-02-01 ms.Rnw
geneal trans prob updates

@@
 All cases can be categorized into four groups based on the occurrence of population event within time $\halfopen{a_q,s_q}$: no population event or simultaneous events, a birth from Deme $j$ to $j'$, a sample of Deme $j'$, and a migration from Deme $j$ to $j'$ where $j\neq j'$.
-Within each group, we investigate $\phi_{qk}^{j''}$ by considering whether the corresponding population event was presented in the pruned genealogy and involving the $q$-th lineage sample, which was in Deme $j''$.
+Within each group, we investigate $\phi_{qk}^{j''}$ by considering whether the corresponding population event was presented in the pruned genealogy and involving the $q$-th lineage sample, which was in Deme $j''$ at time $t_k$.
 For $q>1$, we then have:
@@
 
-    % B2C: either j''=j' or j''=j, the numerator = 1 when j'!=j and =2 when j'=j.
-    \frac{1+\delta_{j,j'}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\\[12pt]
-    1-\frac{(1+\delta_{j,j'})\ell_{qk}^{j}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}-\frac{1}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
+    % B2C: either j''=j (in Z_{q-1} or j''=j' (not in Z_{q-1})
+    % the numerator = 1 when j'!=j and =2 when j'=j.
+    \frac{(1+\delta_{j,j'})}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\\[12pt]
+    1-\frac{(1+\delta_{j,j'})\ \ell_{qk}^{j}\ \delta_{j',j''}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}-\frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_{q-1})\cup\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
+    1-\frac{\delta_{j,j''}}{I_k^{j}I_k^{j'}-\ell_{qk}^j\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1})\setminus\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
     0, &\text{otherwise}.

### 2022-02-01 ms.Rnw
DSABNS, Bilbao, 2022-02-11

--- 2022-01-28 ms.Rnw
@@
   \resizebox{0.6\linewidth}{!}{
-    \begin{tikzpicture}[scale=1.8]
+    \begin{tikzpicture}[scale=2.5]
       \usetikzlibrary{arrows}
@@
 
-\section{Likelihood, filtering equation, and algorithms}
+\begin{lemma}\label{lemma:ell}
+  Define time intervals $\Delta_{i}^{j}$ and $\Delta_{i}^{j,j'}$ as follows:
+  \begin{equation*}
+    \Delta_{i}^{j}\coloneqq\begin{cases}
+      \halfopen{a_i^j,s_i^j}, & \Tset^j(\Och_i)\setminus\Tset^j(\Och_{i-1})\neq\emptyset, \\
+      \emptyset, &\textrm{otherwise},
+    \end{cases}
+    \quad\textrm{and}\quad
+    \Delta_{i}^{j,j'}\coloneqq\begin{cases}
+      \halfopen{z_i^{j,j'},s_i^j}, & \Zset^{j,j'}(\Och_i)\setminus\Zset^{j,j'}(\Och_{i-1})\neq\emptyset, \\
+      \emptyset, &\textrm{otherwise},
+    \end{cases}
+  \end{equation*}
+  where $s_i^j=\Tset^j(\Och_i)\setminus\Tset^j(\Och_{i-1})$, $a_i^j=\Masterset^j(\Och_i)\setminus\Masterset^j(\Och_{i-1})$, and $z_i^{j,j'}=\Zset^{j,j'}(\Och_i)\setminus\Zset^{j,j'}(\Och_{i-1})$ are the corresponding sample, attachment times, respectively, of the $i$-th sample lineage in Deme $j$, and the deme-transition times from Deme $j$ to $j'$, respectively.
+  Then, if it is understood that $\ell^j(t,\Och_0)=0$ for all $j$,
+  \begin{equation*}
+    \ell^j(t,\Och_i)=\ell^j(t,\Och_{i-1})+\indicator{\Delta_{i}^{j}}(t)-\sum_{j'}\left(\indicator{\Delta_{i}^{j,j'}}(t)-\indicator{\Delta_{i}^{j',j}}(t)\right), \qquad \forall i>0.
+  \end{equation*}
+  Trivially, $\ell(t,\Och_i)=\sum_j\ell^j(t,\Och_i)$.
+\end{lemma}
 
@@
 \begin{align}
+\begin{split}
   P_{\Wch_i|\Hist}&=\sum_{\Och_i}\,P_{\Wch_i|\Och_i}P_{\Och_i|\Hist} \\
                   &=\sum_{\Och_i}\,P_{\Wch_i|\Och_i}\left(P_{\Och_1|\Hist}\prod_{q=2}^{i}P_{\Och_q|\Och_{q-1}}\right) 
+\end{split}
 \end{align}
@@
 
-We define $\phi_{qk}\coloneqq\sum_{\omega\in\Omega_q}\auxmeas_{u_k,x_k-u_k}(N_k(\omega)) = \prod_{k=1}^K\sum_{\omega\in\Omega_q}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}\right)$.
+We define $\phi_{qk}^{j}\coloneqq\sum_{\omega\in\Omega_q}\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}$.
 
@@
 All cases can be categorized into four groups based on the occurrence of population event within time $\halfopen{a_q,s_q}$: no population event or simultaneous events, a birth from Deme $j$ to $j'$, a sample of Deme $j'$, and a migration from Deme $j$ to $j'$ where $j\neq j'$.
-Within each group, we investigate $\phi_{qk}^{j''}$ by considering whether the corresponding population event was presented in the pruned genealogy and involving the $q$-th lineage sample, which was in Deme $j''$.
+Within each group, we investigate $\phi_{qk}^{j''}$ by considering whether the corresponding population event was presented in the pruned genealogy and involving the $q$-th lineage sample at time $t_k$, given that the $q$-th lineage sample was in Deme $j''$ at time $t_k$.
 For $q>1$, we then have:
 
+% 1) No. of total events in the population
+% 2) No. of conditioned event in the population
+% 3) possibilities that presented in the genealogy
+% 4) probability of the exact sample involved
+
 \begin{enumerate}[(a)]
-  \item If $t_k\notin\halfopen{a_q,s_q}$, or if $u_k\notin\Bset\cup\Sset\cup\Mset$, so we have $\phi_{qk}^{j''}=1$.
-  \item If $t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1})$, then again, $\phi_{qk}^{j''}=1$.
-  \item If $j''\notin\{j,j'\}$, then the new lineage sample was not involved in any population events involving Demes $j$ and $j'$ at time $t_k$ and $\phi_{qk}^{j''}=1$.
-  \item \label{item:M2Z} If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Mset^{j,j'}$, then there was a migration event at time $t_k$ and the $j$-th sample lineage was involved.
-  Whether this migration event was presented as a deme-transition event in $\Och_q$ depends on whether $j''=j'$.
-  The population size of Deme $j$ right before time $t_k$ was $I_{k-1}^j$, while that of Deme $j'$ right after $t_k$ was $I_k^{j'}$.
-  Lineages of $\Och_{q-1}$ were not involved in this migration event, so they were excluded.
-  Thus, at time $t_k$, one of the $I_{k-1}^{j}-\ell_{q,k-1}^{j}$ individuals was chosen to migrate and right after, among $I_{k}^{j'}-\ell_{qk}^{j'}$ unobserved individuals, the $j$-th sample was the exact one just migrated.
-  Therefore, $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
-  \item If $t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'}$, the logic is the same as that in case \ref{item:M2Z}, but the migration event was not presented in $\Och_q$.
-One can avoid the exact genealogical deme-transition induced by a migration at time $t_k$, with $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
-  \item \label{item:G2D} If $t_k=a_q\in\Lset^{j}(\Och_{q-1})$, and obviously, $u_k\in\Sset_j$.
-    A individual in Deme $j$ was sampled at $t_k$ and the $q$-th lineage sample was exactly the direct-descend of it, \ie a direct-descend event caused by the $q$-th lineage sample was presented in $\Och_q$ at time $t_k$.
+  \item There was no population event within $\halfopen{a_q,s_q}$ or more than one population events occurring that involved the $q$-th sample:
+    \begin{enumerate}
+      \item if $t_k\notin\halfopen{a_q,s_q}$, or if $u_k\notin\Bset\cup\Sset\cup\Mset$, $\phi_{qk}^{j''}=1$;
+      \item if $t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1})$, $\phi_{qk}^{j''}=1$.
+    \end{enumerate}
 
-  \item
-  \item If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Bset^{j,j'}$, then there was a birth event at time $t_k$ and the $j$-th sample lineage was involved.
-    However, this birth event was observed as a deme-transition event (\ie $j\neq j'$) in $\Och_q$, instead of a coalescent event. 
-    This birth event resulted in a newly added node holding two black balls, which were separately from Deme $j$ and Deme $j'$.
-    Thus, the population size of Deme $j$ and $j'$, right after $t_k$, were $I_k^j$ and $I_k^{j'}$, respectively. 
-    There were in total $(I_k^{j}-\ell_{qk}^{j})(I_k^{j'}-\ell_{qk}^{j'})$ combinations of the black balls, because none of the lineages in $\Och_{q-1}$ was involved. 
-    Among all these combinations, the $j$-th sample lineage was exactly the black ball newly added to Deme $j'$ after the birth event.
-    Therefore, $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_k^{j}-\ell_{qk}^{j})(I_k^{j'}-\ell_{qk}^{j'})\right)$.
-  \item
-  \item
+  % M
+  \item There was a migration event, that is, $u_k\in\Mset^{j,j'}$ within $\halfopen{a_q,s_q}$, given that it was not presented in the pre-existing genealogy $\Och_{q-1}$, \ie $t_k\notin\Zset^{j,j'}(\Och_{q-1})$:
+    \begin{enumerate}
+      \item if this migration event exactly involved the $q$-th sample and was presented as a deme-transition event in $\Och_q$, \ie $t_k\in\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
+      \item otherwise, $t_k\notin\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
+    \end{enumerate}
+
+  % G
+  \item There was a sample event, that is $u_k\in\Sset^{j'}$ within $\halfopen{a_q,s_q}$:
+    \begin{enumerate}
+      \item if the lineage sampled at this event served as a direct ancestor to the $q$-th sample lineage, that is, $t_k=a_q\in\Lset^{j'}(\Och_{q-1})$, this indicates that among $I_k^{j'}-\ell_{qk}^{j'}$ unsampled individuals in deme $j'$ at $t_k$, the $q$-th sample lineage is the exact direct descent, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j'}-\ell_{qk}^{j'}\right)$;
+      \item otherwise, $t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})$, and $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left(I_k^{j'}-\ell_{qk}^{j'}\right)$.
+    \end{enumerate}
+
+  % B, j=j'
+  \item There was a birth event within the same deme, that is $u_k\in\Bset^{j,j'}$ and $\delta_{j,j'}=1$, but it was not shown in the genealogy yet because only one of the pre-existing $\ell_{qk}^{j'}$ lineages involved, \ie $t_k\notin\Cset^{j,j'}(\Och_{q-1})$:
+    \begin{enumerate}
+      % there were in total \binom{I}{2} combinations presented as births, while \binom{l}{2} is excluded because the birth was not presented in genealogy.
+      \item if this birth event was presented as a coalescent event in $\Och_q$, \ie $t_k\in\Cset^{j,j'}(\Och_q)$, with probability $\ell_{qk}^{j'}/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$, then the probability that the $q$-th sample attached to \emph{a certain} sample lineage in the genealogy is $1/\ell_{qk}^{j'}$, implying that the $q$-th sample lineage was exactly the other individual involved in the birth event, and eventually, $\phi_{qk}^{j''}=\left(\delta_{j,j'}\delta_{j',j''}\right)/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$;
+      \item otherwise, $t_k\notin\halfopen{a_q,s_q}$,  $\phi_{qk}^{j''}=1-\left(\delta_{j,j'}\delta_{j',j''}\ell_{qk}^{j'}\right)/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$.
+    \end{enumerate}
+
+  % B, j!=j'
+  \item There was a birth event between two different demes, that is $u_k\in\Bset^{j,j'}$ and $\delta_{j,j'}=0$:
+    \begin{enumerate}
+      % There were  l'*(I-l) possibilities of transitions
+      \item and given that it was presented as a deme-transition event in the pre-existing genealogy, that is, $t_k\in\Zset^{j,j'}(\Och_{q-1})$:
+        \begin{enumerate}
+          % the genealogy has the exact l' lineage as the child, and the sample is the exact 1 as the parent
+          % 1/[l'*(I-l)] * l' * 1/1
+          \item if the $q$-th sample was from deme $j$ and is exactly the parent in the birth, that is, $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j,j''}=1$, then $\phi_{qk}^{j''}=\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
+          \item otherwise, $t_k\in\Zset^{j,j'}(\Och_q)$,and $\phi_{qk}^{j''}=1-\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
+        \end{enumerate}
+      % there were in total I*I' in the birth, while l*l' is excluded since it was not presented as coalescent, and (I-l)*l' are excluded which as presented as transition.
+== I I' - l l' - (I-l) l' = I (I'-l')
+      \item or given that it was not presented in $\Och_{q-1}$, \ie $t_k\notin\Zset^{j,j'}(\Och_{q-1})\cup\Cset^{j,j'}(\Och_{q-1})$:
+        \begin{enumerate}
+          % the genealogy has l lineages as parent, and we choose one of them 1/l: 1/[I (I'-l')] * l * 1/l
+          \item if the $q$-th sample was from deme $j'$ and is coalesced at $t_k$, \ie $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j',j''}$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
+          % 
+          \item if the $q$-th sample was from deme $j'$ and is presented as a deme-transition event in $\Och_q$, \ie $t_k\in\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
+
+          \item otherwise, $t_k\notin\Cset^{j,j'}(\Och_q)\cup\Zset^{j,j'}(\Och_q)$, and $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left(I_{k}^{j'}-\ell_{qk}^{j'}\right)$;
+        \end{enumerate}
+    \end{enumerate}
 \end{enumerate}
@@
     1, &\text{if}\ t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1}),\\[12pt]
-    1, &\text{if}\ j''\notin\{j,j'\}\\[12pt]
     \frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
     1-\frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
-    \frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k=a_q\in\Lset^{j'}(\Och_{q-1}),\\[12pt]
+
+    % sample
+    \frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k=a_q\in\Lset^{j'}(\Och_{q-1})\ \text{and}\ u_k\in\Sset^{j'},\\[12pt]
     1-\frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})\ \text{and}\ u_k\in\Sset^{j'},\\[12pt]
-    \frac{\delta_{j',j''}}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
 
-    % B2C: either j''=j' or j''=j, the numerator = 1 when j'!=j and =2 when j'=j.
-    \frac{1+\delta_{j,j'}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\\[12pt]
-    1-\frac{(1+\delta_{j,j'})\ell_{qk}^{j}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}-\frac{1}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
+    % B, j=j'
+    \frac{\delta_{j,j'}\delta_{j',j''}}{\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j=j',\\[12pt]
+    1-\frac{\delta_{j,j'}\delta_{j',j''}\ell_{qk}^{j'}}{\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}}, &\text{if}\ t_k\in(a_q,s_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j=j',\\[12pt]
+
+    % B, j!=j'
+    % in Z
+    \frac{\delta_{j,j''}}{I_k^{j}-\ell_{qk}^{j}}, &\text{if}\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
+    1-\frac{\delta_{j,j''}}{I_k^{j}-\ell_{qk}^{j}}, &\text{if}\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
+    % not in Z
+    \frac{\delta_{j',j''}}{I_k^j(I_{k}^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\notin\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
+    \frac{\delta_{j',j''}}{I_k^j(I_{k}^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\notin\Zset^{j,j'}(\Och_{q-1}),\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
+    1-\frac{\delta_{j',j''}}{I_{k}^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\notin\Zset^{j,j'}(\Och_{q-1}),\ t_k\in(a_q,s_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
+
     0, &\text{otherwise}.
@@
 
+% To summarize, we have
+% \begin{equation}\label{eq:qcases}
+%   \phi_{qk}^{j''}=
+%   \begin{cases}
+%     1, &\text{if}\ t_k\notin\halfopen{a_q,s_q}\ \text{or}\ u_k\notin\Bset\cup\Sset\cup\Mset,\\
+%     
+%     % deme-transition time can evolve to coalescent time, when the new sample is the parent of a "cross-deme" birth
+%     1, &\text{if}\ t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1}),\\[12pt]
+%     \frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
+%     1-\frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
+%     
+%     % sample
+%     \frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k=a_q\in\Lset^{j'}(\Och_{q-1}),\\[12pt]
+%     1-\frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})\ \text{and}\ u_k\in\Sset^{j'},\\[12pt]
+%     \frac{\delta_{j',j''}}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1}),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
+%     
+%     % B2C: either j''=j (in Z_{q-1} or j''=j' (not in Z_{q-1})
+%     % the numerator = 1 when j'!=j and =2 when j'=j.
+%     \frac{\delta_{j,j''}+\delta_{j',j''}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\\[12pt]
+%     1-\frac{(\delta_{j,j''}+\delta_{j',j''})\ \ell_{qk}^{j}\ \delta_{j',j''}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}-\frac{(1-\delta_{j,j'})\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_{q-1})\cup\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
+%     1-\frac{\delta_{j,j''}}{I_k^{j}I_k^{j'}-\ell_{qk}^j\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
+%     0, &\text{otherwise}.
+%   \end{cases}
+% \end{equation}
+
 We then can establish our first main result:
@@
 \begin{thm}\label{thm:prunehthm1}
-  With the definitions as above, $P_{\Och_i|\Hist}(w|h) = \prod_{q=1}^{i}\prod_{k=1}^{K}\phi_{qk}$.
+  With the definitions as above, $P_{\Och_i|\Hist}(o|h) = \prod_{q=1}^{i}\prod_{k=1}^{K}\left(\sum_{j''=1}^{J}\phi_{qk}^{j''}\right)=\prod_{q=1}^{i}\prod_{k=1}^{K}\phi_{qk}^{j''_q}$, where $j''_q$ is the deme of the $q$-th sample lineage.
 \end{thm}
+We now exchange the order of multiplication over $q$ and $k$ then we can derive our \Cref{thm:prunethm2}.
+
+\begin{thm}\label{thm:prunethm2}
+  \begin{equation*}
+    P_{\Prune_t|\Hist_t}(\Prune_t|h)=.
+  \end{equation*}
+\end{thm}
+\begin{proof}
+  Let $\Prune_t=\Och_i$ for some $i$.
+
+  Last but not least, suppose $u_k\in\Bset^{j,j'}$.
+
+\end{proof}
 

### 2022-02-04 ms.Rnw
sgp updates

--- 2022-01-28 ms.Rnw
@@
 All cases can be categorized into four groups based on the occurrence of population event within time $\halfopen{a_q,s_q}$: no population event or simultaneous events, a birth from Deme $j$ to $j'$, a sample of Deme $j'$, and a migration from Deme $j$ to $j'$ where $j\neq j'$.
-Within each group, we investigate $\phi_{qk}^{j''}$ by considering whether the corresponding population event was presented in the pruned genealogy and involving the $q$-th lineage sample, which was in Deme $j''$.
+Within each group, we investigate $\phi_{qk}^{j''}$ by considering whether the corresponding population event was presented in the pruned genealogy and involving the $q$-th lineage sample at time $t_k$, given that the $q$-th lineage sample was in Deme $j''$ at time $t_k$.
 For $q>1$, we then have:
@@
   \item If $t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1})$, then again, $\phi_{qk}^{j''}=1$.
-  \item If $j''\notin\{j,j'\}$, then the new lineage sample was not involved in any population events involving Demes $j$ and $j'$ at time $t_k$ and $\phi_{qk}^{j''}=1$.
-  \item \label{item:M2Z} If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Mset^{j,j'}$, then there was a migration event at time $t_k$ and the $j$-th sample lineage was involved.
+  \item \label{item:M2Z} If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Mset^{j,j'}$, then there was a migration event at time $t_k$ and the $q$-th sample lineage was involved.
   Whether this migration event was presented as a deme-transition event in $\Och_q$ depends on whether $j''=j'$.
@@
   Lineages of $\Och_{q-1}$ were not involved in this migration event, so they were excluded.
-  Thus, at time $t_k$, one of the $I_{k-1}^{j}-\ell_{q,k-1}^{j}$ individuals was chosen to migrate and right after, among $I_{k}^{j'}-\ell_{qk}^{j'}$ unobserved individuals, the $j$-th sample was the exact one just migrated.
+  Thus, at time $t_k$, one of the $I_{k-1}^{j}-\ell_{q,k-1}^{j}$ individuals was chosen to migrate and right after, among $I_{k}^{j'}-\ell_{qk}^{j'}$ unobserved individuals, the $q$-th sample was the exact one just migrated.
   Therefore, $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
@@
   \item
-  \item If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Bset^{j,j'}$, then there was a birth event at time $t_k$ and the $j$-th sample lineage was involved.
+  \item If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Bset^{j,j'}$, then there was a birth event at time $t_k$ and the $q$-th sample lineage was involved.
     However, this birth event was observed as a deme-transition event (\ie $j\neq j'$) in $\Och_q$, instead of a coalescent event. 
@@
     There were in total $(I_k^{j}-\ell_{qk}^{j})(I_k^{j'}-\ell_{qk}^{j'})$ combinations of the black balls, because none of the lineages in $\Och_{q-1}$ was involved. 
-    Among all these combinations, the $j$-th sample lineage was exactly the black ball newly added to Deme $j'$ after the birth event.
+    Among all these combinations, the $q$-th sample lineage was exactly the black ball newly added to Deme $j'$ after the birth event.
     Therefore, $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_k^{j}-\ell_{qk}^{j})(I_k^{j'}-\ell_{qk}^{j'})\right)$.
@@
     1, &\text{if}\ t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1}),\\[12pt]
-    1, &\text{if}\ j''\notin\{j,j'\}\\[12pt]
     \frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
@@
     1-\frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})\ \text{and}\ u_k\in\Sset^{j'},\\[12pt]
-    \frac{\delta_{j',j''}}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
+    \frac{\delta_{j',j''}}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1}),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
 
-    % B2C: either j''=j' or j''=j, the numerator = 1 when j'!=j and =2 when j'=j.
-    \frac{1+\delta_{j,j'}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\\[12pt]
-    1-\frac{(1+\delta_{j,j'})\ell_{qk}^{j}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}-\frac{1}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
+    % B2C: either j''=j (in Z_{q-1} or j''=j' (not in Z_{q-1})
+    % the numerator = 1 when j'!=j and =2 when j'=j.
+    \frac{\delta_{j,j''}+\delta_{j',j''}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\\[12pt]
+    1-\frac{(\delta_{j,j''}+\delta_{j',j''})\ \ell_{qk}^{j}\ \delta_{j',j''}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}-\frac{(1-\delta_{j,j'})\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_{q-1})\cup\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
+    1-\frac{\delta_{j,j''}}{I_k^{j}I_k^{j'}-\ell_{qk}^j\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
     0, &\text{otherwise}.

### 2022-02-15 ms.Rnw
sgp probs update
updates in sgp thm 1 and thm2

--- 2022-02-01 ms.Rnw
@@
   \resizebox{0.6\linewidth}{!}{
-    \begin{tikzpicture}[scale=2.5]
+    \begin{tikzpicture}[scale=1.8]
       \usetikzlibrary{arrows}
@@
 Within a structured infective population, the parent and the child, at a given birth event, can be in different demes.
-That is, an individual in Deme $i$ can give birth to a newborn individual in Deme $j$, where $i,j=1,\dots,J$. 
-We then define the birth function $B_{ij}:\Uspace\to\N$ for any $i,j=1,\dots,J$. 
-The set of birth events where an individual in Deme $i$ gives birth to $n$ individuals in Deme $j$ is $\Bset_{ij}\coloneqq\,B_{ij}^{-1}(\{1\})$. 
-Trivially, the set of birth events in Deme $j$ is $\Bset_j\coloneqq\bigcup_{i}\Bset_{ij}$.
+That is, an individual in Deme $j$ can give birth to a newborn individual in Deme $j'$, where $j,j'=1,\dots,J$. 
+We then define the birth function $B_{j,j'}:\Uspace\to\N$ for any $j,j'=1,\dots,J$. 
+The set of birth events where an individual in Deme $j$ gives birth to $n$ individuals in Deme $j'$ is $\Bset^{j,j'}\coloneqq\,B_{j,j'}^{-1}(\{1\})$. 
+Trivially, the set of birth events in Deme $j'$ is $\Bset^{j'}\coloneqq\bigcup_{j}\Bset^{j,j}$.
 
 \paragraph{Migrations}
-Infective demes are interchangeable and we define the migration function from Deme $i$ to $j$ for any $i\neq j$ and $i,j=1,\dots,J$ as $M_{ij}:\Uspace\to\N$ to represent the number of infectives migrating to Deme $j$ from Deme $i$.
-Thus the set of migration is $\Mset_{ij}\coloneqq\,M^{-1}_{ij}(\{1\})$ since we assume single migration at a single migration event. 
+Infective demes are interchangeable and we define the migration function from Deme $j$ to $j'$ for any $j\neq j'$ and $j,j'=1,\dots,J$ as $M_{j,j'}:\Uspace\to\N$ to represent the number of infectives migrating to Deme $j'$ from Deme $j$.
+Thus the set of migration is $\Mset^{j,j'}\coloneqq\,M^{-1}_{j,j'}(\{1\})$ since we assume single migration at a single migration event. 
 Similarly, we disallow coninciding migration events and other events.
@@
 We assume samples are taken serially and no births or deaths are coincided. 
-We then define the sampling function on the probability space of the population process $\sg{X}_t$: $G:\Uspace\to\N_0$ is the number of samples taken at event $u$, then the set of sampling events is $\Sset\coloneqq\,G^{-1}(\{1\})$.
+We then define the sampling function on the probability space of the population process $\sg{X}_t$: $G_j:\Uspace\to\N_0$ is the number of samples from Deme $j$ taken at event $u$, then the set of sampling events is $\Sset^j\coloneqq\,G_j^{-1}(\{1\})$.
 Of course, the non-coincidence assumption can be easily relaxed while it's not the focus in this paper.
@@
 \begin{equation}
-  \jumprate_u(t,x) > 0 \Longrightarrow I^j(x+u)-I^j(x)=\sum_{i}\,B_{ij}(u)+\sum_{i}M_{ij}(u)-D_j(u)-\sum_{i}M_{ji}(u),
+  \jumprate_u(t,x) > 0 \Longrightarrow I^j(x+u)-I^j(x)=\sum_{j'}\,B_{j',j}(u)+\sum_{j'}M_{j',j}(u)-D_j(u)-\sum_{j'}M_{j,j'}(u),
 \end{equation}
 
-for all $x, u\in\Xspace$. We disallow simultaneous death at a single death event at this stage and define the set of death events in Deme $j$ as $\Dset_j\coloneqq\,D^{-1}(\{1\})$. 
-We also insist, birth and death events don't coincide for any demes, $i,j$, that is,  $\Bset_{j}\cap\Dset_{i}=\emptyset$. 
+for all $x, u\in\Xspace$. We disallow simultaneous death at a single death event at this stage and define the set of death events in Deme $j$ as $\Dset^j\coloneqq\,D_j^{-1}(\{1\})$. 
+We also insist, birth and death events don't coincide for any demes, $j,j'$, that is,  $\Bset_{j}\cap\Dset_{j'}=\emptyset$. 
 Eventually, we can have the function for total population size by letting $I(x)\coloneqq\sum_{j=1}^J\,I^j(x)$ and it's trivial that
@@
 \begin{equation}
-  \jumprate_u(t,x) > 0 \Longrightarrow I(x+u)-I(x)=\sum_{i,j}B_{i,j}(u)-\sum_{j}D_j(u),
+  \jumprate_u(t,x) > 0 \Longrightarrow I(x+u)-I(x)=\sum_{j,j'}B_{j,j'}(u)-\sum_{j}D_j(u),
 \end{equation}
@@
 As in \citet{King2021}, the transmission $b(t)$ is time-dependent, and the sampling rate $\psi$ is any function, as long as the constraints in Section 2 are satisfied. 
-Therefore, $I_1(\sg{X})=e$, $I_2(\sg{X})=i$, $I(\sg{X})=e+i$, $\Bset_{21}=\{\left(-1,1,0,0\right)\}$, $\Mset_{12}=\{\left(0,-1,1,0\right)\}$, $\Dset_2=\{\left(0,0,-1,0\right)\}$, $\Sset=\{\left(0,0,0,1\right)\}$, and $\Bset_{11}=\Bset_{12}=\Bset_{22}=\Mset_{21}=\Dset_1=\emptyset$.
+Therefore, $I_1(\sg{X})=e$, $I_2(\sg{X})=i$, $I(\sg{X})=e+i$, $\Bset^{2,1}=\{\left(-1,1,0,0\right)\}$, $\Mset^{1,2}=\{\left(0,-1,1,0\right)\}$, $\Dset^2=\{\left(0,0,-1,0\right)\}$, $\Sset^2=\{\left(0,0,0,1\right)\}$, and $\Bset^{1,1}=\Bset^{1,2}=\Bset^{2,2}=\Mset^{2,1}=\Dset^1=\Sset^1=\emptyset$.
 
@@
 
-\QYL{the definition for sets of event times is problematic.}
+% \QYL{the definition for sets of event times is problematic.}
 
@@
   \begin{aligned}
-    \Eset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})\ \text{is}\ j\}, \\
-    \Masterset^j(\Gen) &\coloneqq , \\
-    \Cset^{j,j'}(\Gen) &\coloneqq , \\
-    \Lset^j(\Gen) &\coloneqq , \\
-    \Tset^j(\Gen) &\coloneqq , \\
-    \Zset^{j,j'}(\Gen) &\coloneqq , \\
+    \Eset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j\}, \\
+    \Masterset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j, w(\Node{p})\ \text{contains one green ball but no purple balls}\}, \\
+    \Cset^{j,j'}(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j, w(\Node{p})\ \text{contains two green balls, one from Deme $j$ and the other from $j'$}\}, \\
+    \Lset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j, w(\Node{p})\ \text{contains no green balls}\}, \\
+    \Tset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j, w(\Node{p})\ \text{contains a blue ball}\}, \\
+    \Zset^{j,j'}(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j, w(\Node{p})\ \text{contains a purple ball and a green ball from Deme $j'$}\}, \\
     \Qset^j(\Gen) &\coloneqq \Masterset^j(\Gen) \cap \Tset^j(\Gen).
@@
 
-\begin{equation}
-  \ell^j(t,\Prune)\coloneqq
-\end{equation}
+\begin{align}
+\begin{split}
+  \ell^j(t,\Prune) &\coloneqq \sum_{j'}\sum_{e\in\Cset^{j',j}(\Prune)} \indicator{\halfopen{e,\infty}}(t)\;-\;\sum_{e\in\Lset^j(\Prune)} \indicator{\halfopen{e,\infty}}(t) \\
+  +&\;\sum_{j'}\left(\sum_{e\in\Zset^{j',j}(\Prune)}\indicator{\halfopen{e,\infty}}(t)\;-\;\sum_{e\in\Zset^{j,j'}(\Prune)}\indicator{\halfopen{e,\infty}}(t)\right).
+\end{split}
+\end{align}
 
@@
 
-We define $\phi_{qk}^{j}\coloneqq\sum_{\omega\in\Omega_q}\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}$.
-
-Note that $\ell_{1k}=0$ for all $k$.
+We define $\phi_{qk}^{j}\coloneqq\sum_{\omega\in\Omega_q}\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}$, 
+and let $I_k^j=I^j(x_k)$, $\ell_{qk}^{j}=\ell^j(t_k, \Och_{q-1})$.
+Note that $\ell^j_{1k}=0$ for all $k$ and $j$.
 
@@
     \begin{enumerate}
-      \item if this migration event exactly involved the $q$-th sample and was presented as a deme-transition event in $\Och_q$, \ie $t_k\in\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
-      \item otherwise, $t_k\notin\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
+      \item if this migration event had been presented in previous genealogy $\Och_{q-1}$, \ie $t_k\in\Zset^{j,j'}(\Och_{q-1})$, since no simultaneous migration events occurred at the same time and $t_k\in\Zset^{j,j'}{\Och_q}$ for sure, $\phi_{qk}^{j''}=1$;
+      \item if this migration event exactly involved the $q$-th sample and was presented as a deme-transition event in $\Och_q$, \ie $t_k\in\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_{k}^{j}-\ell_{q,k}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
+      \item otherwise, $t_k\notin\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left((I_{k}^{j}-\ell_{q,k}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
     \end{enumerate}
@@
     1, &\text{if}\ t_k\notin\halfopen{a_q,s_q}\ \text{or}\ u_k\notin\Bset\cup\Sset\cup\Mset,\\
+    1, &\text{if}\ t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1}),\\[12pt]
 
-    % deme-transition time can evolve to coalescent time, when the new sample is the parent of a "cross-deme" birth
-    1, &\text{if}\ t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1}),\\[12pt]
-    \frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
-    1-\frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
+    % M
+    1, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_{q}),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
+    \frac{\delta_{j',j''}}{(I_{k}^j-\ell_{q,k}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
+    1-\frac{\delta_{j',j''}}{(I_{k}^j-\ell_{q,k}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
 
-    % sample
+    % G
     \frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k=a_q\in\Lset^{j'}(\Och_{q-1})\ \text{and}\ u_k\in\Sset^{j'},\\[12pt]
@@
     \frac{\delta_{j,j''}}{I_k^{j}-\ell_{qk}^{j}}, &\text{if}\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
-    1-\frac{\delta_{j,j''}}{I_k^{j}-\ell_{qk}^{j}}, &\text{if}\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
+    1-\frac{\delta_{j,j''}}{I_k^{j}-\ell_{qk}^{j}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
     % not in Z
-    \frac{\delta_{j',j''}}{I_k^j(I_{k}^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\notin\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
+    \frac{\delta_{j',j''}}{I_k^j(I_{k}^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
     \frac{\delta_{j',j''}}{I_k^j(I_{k}^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\notin\Zset^{j,j'}(\Och_{q-1}),\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
-    1-\frac{\delta_{j',j''}}{I_{k}^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\notin\Zset^{j,j'}(\Och_{q-1}),\ t_k\in(a_q,s_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
+    1-\frac{\delta_{j',j''}}{I_{k}^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_{q-1}),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
 
@@
 \begin{thm}\label{thm:prunehthm1}
-  With the definitions as above, $P_{\Och_i|\Hist}(o|h) = \prod_{q=1}^{i}\prod_{k=1}^{K}\left(\sum_{j''=1}^{J}\phi_{qk}^{j''}\right)=\prod_{q=1}^{i}\prod_{k=1}^{K}\phi_{qk}^{j''_q}$, where $j''_q$ is the deme of the $q$-th sample lineage.
+  With the definitions as above, $P_{\Och_i|\Hist}(o|h) = \prod_{q=1}^{i}\prod_{k=1}^{K}\left(\sum_{j''=1}^{J}\phi_{qk}^{j''}\right)=\prod_{q=1}^{i}\prod_{k=1}^{K}\phi_{qk}^{j''_{qk}}$, where $j''_{qk}$ is the deme of the $q$-th sample lineage at time $t_k$.
 \end{thm}
@@
   Let $\Prune_t=\Och_i$ for some $i$.
+We define $\Phi_{k}\coloneqq\prod_{q=1}^{i}\phi_{qk}^{j''_{qk}}$ for each $k$ and as above, $I_k^{j}=I_j(x_k)$, $\ell_{qk}^{j}=\ell^j(t,\Och_{q-1})$.
+For the first two cases in \Cref{eq:qcases}, $\Phi_k=1$.
 
-  Last but not least, suppose $u_k\in\Bset^{j,j'}$.
+  Consider $u_k\in\Mset^{j,j'}$.
+Let $q^*=\{q\;\vert\;t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})\}$ and $\ell^*=\ell_{q^*k}^{j''_{q^*k}}$.
+We then define $Q_k=\{q\;\vert\;j''_{qk}=j'\ \text{and}\ q<q^*\}$ as an ordered set and denote $[Q_k]_m$ as the $m$-th element in $Q_k$, where, $m=1,\dots, \ell^*$.
+  \begin{enumerate}[(a)]
+    \item If $q<q^*$, then $t_k\notin\Zset^{j,j'}(\Och_q)$. We will have
+      \begin{equation*}
+        \Phi_k = \prod_{q\in Q_k} \left(1-\frac{1}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}\right).
+      \end{equation*}
+      Note that, $\{\ell_{qk}^{j'}\;\vert\;q\in Q_k\}=\{0,1,\dots,\ell^*-1\}$ then the previous equation becomes
+      \begin{equation*}
+        \Phi_k = \prod_{m=1}^{\ell*} \left(1-\frac{1}{\mathcal{Y}_m (I_k^{j'}-m+1)}\right) = \prod_{m=1}^{\ell*} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m)^{-1}}{I_k^{j'}-m+1},
+      \end{equation*}
+      where $\mathcal{Y}_m\coloneqq I_k^{j}-\ell_{[Q_k]_m k}^{j}$.
+  \end{enumerate}
 

### 2022-02-18 ms.Rnw
thm2 update

@@
 \begin{equation}
-  \jumprate_u(t,x) > 0 \Longrightarrow I^j(x+u)-I^j(x)=\sum_{j'}\,B_{j',j}(u)+\sum_{j'}M_{j',j}(u)-D_j(u)-\sum_{j'}M_{j,j'}(u),
+  \jumprate_u(t,x) > 0 \Longrightarrow I_j(x+u)-I_j(x)=\sum_{j'}\,B_{j',j}(u)+\sum_{j'}M_{j',j}(u)-D_j(u)-\sum_{j'}M_{j,j'}(u),
 \end{equation}
@@
 We also insist, birth and death events don't coincide for any demes, $j,j'$, that is,  $\Bset_{j}\cap\Dset_{j'}=\emptyset$. 
-Eventually, we can have the function for total population size by letting $I(x)\coloneqq\sum_{j=1}^J\,I^j(x)$ and it's trivial that
+Eventually, we can have the function for total population size by letting $I(x)\coloneqq\sum_{j=1}^J\,I_j(x)$ and it's trivial that
 
@@
   \begin{cases}
-    \mathrm{initialize}(I^j(\sg{X}_0(\omega)), & \text{if }K(\omega)=0; \\
+    \mathrm{initialize}(I_j(\sg{X}_0(\omega)), & \text{if }K(\omega)=0; \\
     \mathrm{add}(\mathrm{inven}(\predec{\omega},j)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Bset_j;\\
@@
 
-We can further define the deme proportion function $\demefuc(t,\omega)\coloneqq\{\demefuc^j(t,\omega)\}_{j=1}^J$, where $\demefuc^j(t,\omega)\coloneqq\frac{\vert\Inv_t^j(\omega)\vert}{\vert\Inv_t(\omega)\vert}$.
+We can further define the deme proportion function $\demefunc(t,\omega)\coloneqq\{\demefunc^j(t,\omega)\}_{j=1}^J$, where $\demefunc^j(t,\omega)\coloneqq\frac{\vert\Inv_t^j(\omega)\vert}{\vert\Inv_t(\omega)\vert}$.
 
@@
   \end{equation*}
-  where $s_i^j=\Tset^j(\Och_i)\setminus\Tset^j(\Och_{i-1})$, $a_i^j=\Masterset^j(\Och_i)\setminus\Masterset^j(\Och_{i-1})$, and $z_i^{j,j'}=\Zset^{j,j'}(\Och_i)\setminus\Zset^{j,j'}(\Och_{i-1})$ are the corresponding sample, attachment times, respectively, of the $i$-th sample lineage in Deme $j$, and the deme-transition times from Deme $j$ to $j'$, respectively.
+  where $s_i^j=\Tset^j(\Och_i)\setminus\Tset^j(\Och_{i-1})$, $a_i^j=\Masterset^j(\Och_i)\setminus\Masterset^j(\Och_{i-1})$, 
+  and $z_i^{j,j'}=\Zset^{j,j'}(\Och_i)\setminus\Zset^{j,j'}(\Och_{i-1})$ are the corresponding sample, attachment times, respectively, 
+  of the $i$-th sample lineage in Deme $j$, and the deme-transition times from Deme $j$ to $j'$, respectively.
   Then, if it is understood that $\ell^j(t,\Och_0)=0$ for all $j$,
@@
     % P_{\Och_q|\Och_{q-1},\Hist}(o_q|o_{q-1},h)&=\prod_{k=1}^K\sum_{n_k\in\Omega_{qk}}\auxmeas_{u_k,x_k-u_k}(n_k)=\prod_{k=1}^K\sum_{n_k\in\Omega_{qk}}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv^j_{t_{k-1}}(\Omega_q)}(n_k)}{I_j(x_k-u_k)}\right).
-    P_{\Och_q|\Och_{q-1},\Hist}(o_q|o_{q-1},h) &= \prod_{k=1}^K\sum_{\omega\in\Omega_q}\auxmeas_{u_k,x_k-u_k}(N_k(\omega)) = \prod_{k=1}^K\sum_{\omega\in\Omega_q}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}\right),
+    P_{\Och_q|\Och_{q-1},\Hist}(o_q|o_{q-1},h) &= \prod_{k=1}^K\sum_{\omega\in\Omega_q}\auxmeas_{u_k,x_k-u_k}(N_k(\omega)) = \prod_{k=1}^K\sum_{\omega\in\Omega_q}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I_j(x_k-u_k)}\right),
   \end{aligned}
@@
 
-We define $\phi_{qk}^{j}\coloneqq\sum_{\omega\in\Omega_q}\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}$, 
-and let $I_k^j=I^j(x_k)$, $\ell_{qk}^{j}=\ell^j(t_k, \Och_{q-1})$.
+We define $\phi_{qk}^{j}\coloneqq\sum_{\omega\in\Omega_q}\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I_j(x_k-u_k)}$, 
+and let $I_k^j=I_j(x_k)$, $\ell_{qk}^{j}=\ell^j(t_k, \Och_{q-1})$.
 Note that $\ell^j_{1k}=0$ for all $k$ and $j$.
 
-All cases can be categorized into four groups based on the occurrence of population event within time $\halfopen{a_q,s_q}$: no population event or simultaneous events, a birth from Deme $j$ to $j'$, a sample of Deme $j'$, and a migration from Deme $j$ to $j'$ where $j\neq j'$.
+All cases can be categorized into four groups based on the occurrence of population event within time $\halfopen{a_q,s_q}$: 
+no population event or simultaneous events, 
+a birth from Deme $j$ to $j'$, 
+a sample of Deme $j'$, 
+and a migration from Deme $j$ to $j'$ where $j\neq j'$. 
 Within each group, we investigate $\phi_{qk}^{j''}$ by considering whether the corresponding population event was presented in the pruned genealogy and involving the $q$-th lineage sample at time $t_k$, given that the $q$-th lineage sample was in Deme $j''$ at time $t_k$.
@@
   % M
-  \item There was a migration event, that is, $u_k\in\Mset^{j,j'}$ within $\halfopen{a_q,s_q}$, given that it was not presented in the pre-existing genealogy $\Och_{q-1}$, \ie $t_k\notin\Zset^{j,j'}(\Och_{q-1})$:
+  \item There was a migration event, that is, $u_k\in\Mset^{j,j'}$ within $\halfopen{a_q,s_q}$, 
+  given that it was not presented in the pre-existing genealogy $\Och_{q-1}$, 
+  \ie $t_k\notin\Zset^{j,j'}(\Och_{q-1})$: 
     \begin{enumerate}
-      \item if this migration event had been presented in previous genealogy $\Och_{q-1}$, \ie $t_k\in\Zset^{j,j'}(\Och_{q-1})$, since no simultaneous migration events occurred at the same time and $t_k\in\Zset^{j,j'}{\Och_q}$ for sure, $\phi_{qk}^{j''}=1$;
-      \item if this migration event exactly involved the $q$-th sample and was presented as a deme-transition event in $\Och_q$, \ie $t_k\in\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_{k}^{j}-\ell_{q,k}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
-      \item otherwise, $t_k\notin\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left((I_{k}^{j}-\ell_{q,k}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
+      \item if this migration event had been presented in previous genealogy $\Och_{q-1}$, 
+      \ie $t_k\in\Zset^{j,j'}(\Och_{q-1})$, 
+      since no simultaneous migration events occurred at the same time and $t_k\in\Zset^{j,j'}{\Och_q}$ for sure, $\phi_{qk}^{j''}=1$;
+      \item if this migration event exactly involved the $q$-th sample and was presented as a deme-transition event in $\Och_q$, 
+      \ie $t_k\in\Zset^{j,j'}(\Och_q)$, 
+      then $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_{k}^{j}-\ell_{q,k}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
+      \item otherwise, $t_k\notin\Zset^{j,j'}(\Och_q)$, 
+      then $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left((I_{k}^{j}-\ell_{q,k}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
     \end{enumerate}
@@
     \begin{enumerate}
-      \item if the lineage sampled at this event served as a direct ancestor to the $q$-th sample lineage, that is, $t_k=a_q\in\Lset^{j'}(\Och_{q-1})$, this indicates that among $I_k^{j'}-\ell_{qk}^{j'}$ unsampled individuals in deme $j'$ at $t_k$, the $q$-th sample lineage is the exact direct descent, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j'}-\ell_{qk}^{j'}\right)$;
-      \item otherwise, $t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})$, and $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left(I_k^{j'}-\ell_{qk}^{j'}\right)$.
+      \item if the lineage sampled at this event served as a direct ancestor to the $q$-th sample lineage, 
+      that is, $t_k=a_q\in\Lset^{j'}(\Och_{q-1})$, 
+      this indicates that among $I_k^{j'}-\ell_{qk}^{j'}$ unsampled individuals in deme $j'$ at $t_k$, 
+      the $q$-th sample lineage is the exact direct descent, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j'}-\ell_{qk}^{j'}\right)$;
+      \item otherwise, $t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})$, 
+      and $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left(I_k^{j'}-\ell_{qk}^{j'}\right)$.
     \end{enumerate}
@@
   % B, j=j'
-  \item There was a birth event within the same deme, that is $u_k\in\Bset^{j,j'}$ and $\delta_{j,j'}=1$, but it was not shown in the genealogy yet because only one of the pre-existing $\ell_{qk}^{j'}$ lineages involved, \ie $t_k\notin\Cset^{j,j'}(\Och_{q-1})$:
+  \item There was a birth event within the same deme, 
+  that is $u_k\in\Bset^{j,j'}$ and $\delta_{j,j'}=1$, 
+  but it was not shown in the genealogy yet because only one of the pre-existing $\ell_{qk}^{j'}$ lineages involved, 
+  \ie $t_k\notin\Cset^{j,j'}(\Och_{q-1})$:
     \begin{enumerate}
       % there were in total \binom{I}{2} combinations presented as births, while \binom{l}{2} is excluded because the birth was not presented in genealogy.
-      \item if this birth event was presented as a coalescent event in $\Och_q$, \ie $t_k\in\Cset^{j,j'}(\Och_q)$, with probability $\ell_{qk}^{j'}/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$, then the probability that the $q$-th sample attached to \emph{a certain} sample lineage in the genealogy is $1/\ell_{qk}^{j'}$, implying that the $q$-th sample lineage was exactly the other individual involved in the birth event, and eventually, $\phi_{qk}^{j''}=\left(\delta_{j,j'}\delta_{j',j''}\right)/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$;
-      \item otherwise, $t_k\notin\halfopen{a_q,s_q}$,  $\phi_{qk}^{j''}=1-\left(\delta_{j,j'}\delta_{j',j''}\ell_{qk}^{j'}\right)/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$.
+      \item if this birth event was presented as a coalescent event in $\Och_q$, 
+      \ie $t_k\in\Cset^{j,j'}(\Och_q)$,
+      with probability $\ell_{qk}^{j'}/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$, 
+      then the probability that the $q$-th sample attached to \emph{a certain} sample lineage in the genealogy is $1/\ell_{qk}^{j'}$, 
+      implying that the $q$-th sample lineage was exactly the other individual involved in the birth event, 
+      and eventually, $\phi_{qk}^{j''}=\left(\delta_{j,j'}\delta_{j',j''}\right)/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$;
+      \item otherwise, $t_k\notin\halfopen{a_q,s_q}$,  
+      $\phi_{qk}^{j''}=1-\left(\delta_{j,j'}\delta_{j',j''}\ell_{qk}^{j'}\right)/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$.
     \end{enumerate}
@@
   % B, j!=j'
-  \item There was a birth event between two different demes, that is $u_k\in\Bset^{j,j'}$ and $\delta_{j,j'}=0$:
+  \item There was a birth event between two different demes, 
+  that is $u_k\in\Bset^{j,j'}$ and $\delta_{j,j'}=0$:
     \begin{enumerate}
       % There were  l'*(I-l) possibilities of transitions
-      \item and given that it was presented as a deme-transition event in the pre-existing genealogy, that is, $t_k\in\Zset^{j,j'}(\Och_{q-1})$:
+      \item and given that it was presented as a deme-transition event in the pre-existing genealogy, 
+      that is, $t_k\in\Zset^{j,j'}(\Och_{q-1})$:
         \begin{enumerate}
@@
           % 1/[l'*(I-l)] * l' * 1/1
-          \item if the $q$-th sample was from deme $j$ and is exactly the parent in the birth, that is, $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j,j''}=1$, then $\phi_{qk}^{j''}=\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
-          \item otherwise, $t_k\in\Zset^{j,j'}(\Och_q)$,and $\phi_{qk}^{j''}=1-\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
+          \item if the $q$-th sample was from deme $j$ and is exactly the parent in the birth, 
+          that is, $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j,j''}=1$, then $\phi_{qk}^{j''}=\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
+          \item otherwise, 
+          $t_k\in\Zset^{j,j'}(\Och_q)$,and $\phi_{qk}^{j''}=1-\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
         \end{enumerate}
@@
 == I I' - l l' - (I-l) l' = I (I'-l')
-      \item or given that it was not presented in $\Och_{q-1}$, \ie $t_k\notin\Zset^{j,j'}(\Och_{q-1})\cup\Cset^{j,j'}(\Och_{q-1})$:
+      \item or given that it was not presented in $\Och_{q-1}$, 
+      \ie $t_k\notin\Zset^{j,j'}(\Och_{q-1})\cup\Cset^{j,j'}(\Och_{q-1})$:
         \begin{enumerate}
           % the genealogy has l lineages as parent, and we choose one of them 1/l: 1/[I (I'-l')] * l * 1/l
-          \item if the $q$-th sample was from deme $j'$ and is coalesced at $t_k$, \ie $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j',j''}$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
-          % 
-          \item if the $q$-th sample was from deme $j'$ and is presented as a deme-transition event in $\Och_q$, \ie $t_k\in\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
-
-          \item otherwise, $t_k\notin\Cset^{j,j'}(\Och_q)\cup\Zset^{j,j'}(\Och_q)$, and $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left(I_{k}^{j'}-\ell_{qk}^{j'}\right)$;
+          \item if the $q$-th sample was from deme $j'$ and is coalesced at $t_k$, 
+          \ie $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j',j''}$, 
+          then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
+          % I - l can be the parent, and we choose one of them 1/(I - l) as the sample: 1/[I (I'-l')] * (I-l) * 1/(I-l)
+          \item if the $q$-th sample was from deme $j'$ and is presented as a deme-transition event in $\Och_q$, 
+          \ie $t_k\in\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
+          \item otherwise, $t_k\notin\Cset^{j,j'}(\Och_q)\cup\Zset^{j,j'}(\Och_q)$, 
+          and $$\phi_{qk}^{j''}=1-\delta_{j',j''}/(I_{k}^{j'}-\ell_{qk}^{j'});$$
         \end{enumerate}
@@
     1, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_{q}),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
-    \frac{\delta_{j',j''}}{(I_{k}^j-\ell_{q,k}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
+    \frac{\delta_{j',j''}}{(I_{k}^j-\ell_{q,k}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1}),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
     1-\frac{\delta_{j',j''}}{(I_{k}^j-\ell_{q,k}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
@@
 \begin{thm}\label{thm:prunehthm1}
-  With the definitions as above, $P_{\Och_i|\Hist}(o|h) = \prod_{q=1}^{i}\prod_{k=1}^{K}\left(\sum_{j''=1}^{J}\phi_{qk}^{j''}\right)=\prod_{q=1}^{i}\prod_{k=1}^{K}\phi_{qk}^{j''_{qk}}$, where $j''_{qk}$ is the deme of the $q$-th sample lineage at time $t_k$.
+  With the definitions as above, 
+  $P_{\Och_i|\Hist}(o|h) = \prod_{q=1}^{i}\prod_{k=1}^{K}\left(\sum_{j''=1}^{J}\phi_{qk}^{j''}\right)=\prod_{q=1}^{i}\prod_{k=1}^{K}\phi_{qk}^{j''_{qk}}$, 
+  where $j''_{qk}$ is the deme of the $q$-th sample lineage at time $t_k$.
 \end{thm}
@@
   Let $\Prune_t=\Och_i$ for some $i$. 
-We define $\Phi_{k}\coloneqq\prod_{q=1}^{i}\phi_{qk}^{j''_{qk}}$ for each $k$ and as above, $I_k^{j}=I_j(x_k)$, $\ell_{qk}^{j}=\ell^j(t,\Och_{q-1})$.
+  We define $\Phi_{k}\coloneqq\prod_{q=1}^{i}\phi_{qk}^{j''_{qk}}$ for each $k$ and as above, 
+  $I_k^{j}=I_j(x_k)$, $\ell_{qk}^{j}=\ell^j(t,\Och_{q-1})$. 
   For the first two cases in \Cref{eq:qcases}, $\Phi_k=1$.
@@
   Consider $u_k\in\Mset^{j,j'}$.
-Let $q^*=\{q\;\vert\;t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})\}$ and $\ell^*=\ell_{q^*k}^{j''_{q^*k}}$.
-We then define $Q_k=\{q\;\vert\;j''_{qk}=j'\ \text{and}\ q<q^*\}$ as an ordered set and denote $[Q_k]_m$ as the $m$-th element in $Q_k$, where, $m=1,\dots, \ell^*$.
+Let $q^*=\{q\;\vert\;t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})\}$ and $\ell^*=\ell_{q^*k}^{j'}$.
+We then define $Q_k^{j'}=\{q\;\vert\;j''_{qk}=j'\ \text{and}\ q\leq q^*\}$ as an ordered set and denote $[Q_k^{j'}]_m$ as the $m$-th element in $Q_k^j$, where, $m=1,\dots, \ell^*$.
   \begin{enumerate}[(a)]
-    \item If $q<q^*$, then $t_k\notin\Zset^{j,j'}(\Och_q)$. We will have
+    \item \label{scn: notZ} If $q<q^*$, then $t_k\notin\Zset^{j,j'}(\Och_q)$. We will have
       \begin{equation*}
-        \Phi_k = \prod_{q\in Q_k} \left(1-\frac{1}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}\right).
+        \Phi_k = \prod_{q\in Q_k^{j'}} \left(1-\frac{1}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}\right).
       \end{equation*}
-      Note that, $\{\ell_{qk}^{j'}\;\vert\;q\in Q_k\}=\{0,1,\dots,\ell^*-1\}$ then the previous equation becomes
+      Note that, $\{\ell_{qk}^{j'}\;\vert\;q\in Q_k^{j'}\}=\{0,1,\dots,\ell^*\}$ then the previous equation becomes
       \begin{equation*}
-        \Phi_k = \prod_{m=1}^{\ell*} \left(1-\frac{1}{\mathcal{Y}_m (I_k^{j'}-m+1)}\right) = \prod_{m=1}^{\ell*} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m)^{-1}}{I_k^{j'}-m+1},
+        \Phi_k = \prod_{m=1}^{\ell^*} \left(1-\frac{1}{\mathcal{Y}_m^j (I_k^{j'}-m+1)}\right) = \prod_{m=1}^{\ell^*} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m^j)^{-1}}{I_k^{j'}-m+1},
       \end{equation*}
-      where $\mathcal{Y}_m\coloneqq I_k^{j}-\ell_{[Q_k]_m k}^{j}$.
+    where $\mathcal{Y}_m^j\coloneqq I_k^{j}-\ell_{[Q_k^{j'}]_m k}^{j}\in\Z$.
+
+    \item If $q\geq q^*$, then $t_k\in\Zset^{j,j'}$, and obviously, with $\mathcal{Y}_m^j$ defined as before,
+      \begin{equation*}
+        \Phi_k = \frac{1}{\mathcal{Y}_{\ell^*}^j (I_k^{j'}-\ell^*)}\, \prod_{m=1}^{\ell^*} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m^j)^{-1}}{I_k^{j'}-m+1}.
+      \end{equation*}
+  \end{enumerate}
+
+  Consider $u_k\in\Bset^{j,j'}$ and $j\neq j'$.
+Let $q^*=\{q\;\vert\;t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})\}$ and $q'=\{q\;\vert\;t_k=\Cset^{j,j'}(\Och_q)\setminus\Cset^{j,j'}(\Och_{q-1})\}$.
+  % first, get q* as the sample of deme-transition
+  % q' as the sample of coalescent
+  \begin{enumerate}[(a)]
+    % if not exists q* and q'
+    \item If $q^*=\emptyset$ and $q'=\emptyset$, then $t_k\notin\Zset^{j,j'}(\Och_i)\cup\Cset^{j,j'}(\Och_{i})$. So for any $q\leq i$, we have
+      \begin{equation*}
+        \Phi_k=\prod_{m=0}^{\ell_{ik}^{j'}}\left(1-\frac{1}{I_k^{j'}-m}\right)=1-\frac{\ell_{ik}^{j'}+1}{I_k^{j'}}.
+      \end{equation*}
+    % if not exists q* but q'
+    \item If $q^*\neq\emptyset$ but $q'=\emptyset$, then $t_k\in\Zset^{j,j'}(\Och_{i})$. Let $\ell^*=\ell_{q^*k}^{j'}$ and for $q^*<q\leq i$, we have
+      \begin{equation*}
+      \begin{split}
+        \Phi_k &= \prod_{m=0}^{\ell^*}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell^*)}\,\prod_{m=\ell_{q^*k}^{j}}^{\ell_{ik}^{j}}\left(1-\frac{1}{I_k^{j}-m}\right) \\
+        &= \frac{I_k^{j'}-\ell^*-1}{I_k^{j'}}\frac{1}{I_k^j(I_k^{j'}-\ell^*)}\frac{I_k^j-\ell_{ik}^{j}-1}{I_k^j-\ell_{q^*k}^{j}}\\
+        &= \frac{1}{\mathcal{Y}^{j}_{\ell^*}(I_k^{j'}-\ell^*)}\left(1-\frac{\ell_{ik}^{j}+1}{I_k^j}\right)\left(1-\frac{\ell^*+1}{I_k^{j'}}\right),
+      \end{split}
+      \end{equation*}
+      where $\mathcal{Y}^{j}_{\ell^*}$ is as defined in \ref{scn: notZ} when $u_k\in\Mset^{j,j'}$.
+    % if not exists q' but q*
+    \item If $q'\neq\emptyset$ but $q^*=\emptyset$, then $t_k\notin\Zset^{j,j'}(\Och_{q'-1})$ and $t_k\in\Cset^{j,j'}(\Och_i)$. Let $\ell'=\ell_{q'k}^{j'}$ and for $q'<q\leq i$, we have
+      \begin{equation*}
+        \begin{split}
+          \Phi_k &= \prod_{m=0}^{\ell'}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell')}=\left(1-\frac{\ell'+1}{I_{k}^{j'}}\right)\frac{1}{I_k^{j}(I_k^{j'}-\ell')}.
+        \end{split}
+      \end{equation*}
+    % if exists q* and q', and q' > q*
+    \item If $q^*\neq\emptyset$, $q'\neq\emptyset$, and obviously, $q^*<q'$, then $t_k\in\Zset^{j,j'}(\Och_{q^*})$, $t_k\notin\Zset^{j,j'}(\Och_{q'})$, and $t_k\in\Cset^{j,j'}(\Och_{q'})$.
+Let $\ell^*=\ell_{q^*k}^{j'}$ and $\hat{\ell}=\ell_{q'k}^{j}$.
+We have
+      \begin{equation*}
+      \begin{split}
+        \Phi_k &= \prod_{m=0}^{\ell^*}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell^*)}\prod_{m=\ell_{q^*k}^{j}}^{\hat{\ell}}\left(1-\frac{1}{I_k^{j}-m}\right)\,\frac{1}{I_k^{j}-\hat{\ell}} \\
+        &= \frac{1}{\mathcal{Y}_{\ell^*}^{j}(I_k^{j}-\hat{\ell})(I_k^{j'}-\ell^*)}\left(1-\frac{\hat{\ell}+1}{I_k^{j}}\right)\left(1-\frac{\ell^*+1}{I_k^{j'}}\right)
+      \end{split}
+      \end{equation*}
   \end{enumerate}

### 2022-02-25 ms.Rnw
update sgp thm2

@@
   Consider $u_k\in\Mset^{j,j'}$. 
-Let $q^*=\{q\;\vert\;t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})\}$ and $\ell^*=\ell_{q^*k}^{j'}$.
-We then define $Q_k^{j'}=\{q\;\vert\;j''_{qk}=j'\ \text{and}\ q\leq q^*\}$ as an ordered set and denote $[Q_k^{j'}]_m$ as the $m$-th element in $Q_k^j$, where, $m=1,\dots, \ell^*$.
+  Let $q^*$ be the sample satisfying $t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})$ for $q=1,\dots, i$.
+  If $\exists q^*$, we then define $Q_k^{j'}=\{q\;\vert\;j''_{qk}=j'\ \text{and}\ q\leq q^*\}$ as an ordered set; 
+  otherwise, $Q_k^{j'}=\{q\;\vert\;j''_{qk}=j'\}$.
+  We denote $[Q_k^{j'}]_m$ as the $m$-th element in $Q_k^j$ and $m=1,\dots, \ell^*+1$. 
+  Note that, $\ell^{*}=\ell^{j'}_{q^*k}$ if $\exists q^*$, otherwise, $\ell^*=\ell_{ik}^{j'}$.
   \begin{enumerate}[(a)]
-    \item \label{scn: notZ} If $q<q^*$, then $t_k\notin\Zset^{j,j'}(\Och_q)$. We will have
+    \item \label{scn: notZ} If $\nexists q^*$, then $t_k\notin\Zset^{j,j'}(\Och_i)$. We will have
       \begin{equation*}
@@
       \begin{equation*}
-        \Phi_k = \prod_{m=1}^{\ell^*} \left(1-\frac{1}{\mathcal{Y}_m^j (I_k^{j'}-m+1)}\right) = \prod_{m=1}^{\ell^*} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m^j)^{-1}}{I_k^{j'}-m+1},
+        \Phi_k = \prod_{m=1}^{\ell^*+1} \left(1-\frac{1}{\mathcal{Y}_m^j (I_k^{j'}-m+1)}\right) = \prod_{m=1}^{\ell_{ik}^{j'}+1} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m^j)^{-1}}{I_k^{j'}-m+1},
       \end{equation*}
@@
 
-    \item If $q\geq q^*$, then $t_k\in\Zset^{j,j'}$, and obviously, with $\mathcal{Y}_m^j$ defined as before,
+    \item If $\exists q^*$ and $i\geq q^*$, then $t_k\in\Zset^{j,j'}(\Och_i)$, and obviously, with $\mathcal{Y}_m^j$ defined as before,
       \begin{equation*}
-        \Phi_k = \frac{1}{\mathcal{Y}_{\ell^*}^j (I_k^{j'}-\ell^*)}\, \prod_{m=1}^{\ell^*} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m^j)^{-1}}{I_k^{j'}-m+1}.
+        \Phi_k = \frac{1}{\mathcal{Y}_{\ell^*+1}^j (I_k^{j'}-\ell^*)}\, \prod_{m=1}^{\ell^*} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m^j)^{-1}}{I_k^{j'}-m+1}.
+      \end{equation*}
+  \end{enumerate}
+
+  Consider $u_k\in\Sset^{j'}$. The results are similar to those in \citet{King2021}, that is
+  \begin{enumerate}[(a)]
+    \item If $t_k\in\Lset^{j'}(\Och_i)$, we have
+      \begin{equation*}
+        \Phi_k=1-\frac{\ell_{ik}^{j'}+1}{I_k^{j'}};
+      \end{equation*}
+    \item otherwise, if $t_k\in\Qset^{j'}(\Och_i)$,
+      \begin{equation*}
+        \Phi_k=\frac{1}{I_k^{j'}}.
+      \end{equation*}
+  \end{enumerate}
+
+  Consider $u_k\in\Bset^{j,j'}$ and $j=j'$. The results can be also referenced from \citet{King2021} as follows:
+  \begin{enumerate}[(a)]
+    \item If $t_k\notin\Cset^{j,j'}(\Och_i)$,
+      \begin{equation*}
+        \Phi_k=1-\frac{\binom{\ell_{ik}^{j'}+1}{2}}{\binom{I_k^{j'}}{2}};
+      \end{equation*}
+    \item otherwise, if $t_k\in\Cset^{j,j'}(\Och_i)$,
+      \begin{equation*}
+        \Phi_k=\frac{1}{\binom{I_k^{j'}}{2}}.
       \end{equation*}
@@
   Consider $u_k\in\Bset^{j,j'}$ and $j\neq j'$. 
-Let $q^*=\{q\;\vert\;t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})\}$ and $q'=\{q\;\vert\;t_k=\Cset^{j,j'}(\Och_q)\setminus\Cset^{j,j'}(\Och_{q-1})\}$.
+  Again, let $q^*$ be the sample satisfying $t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})\}$ and further, 
+  $q'$ be the sample satisfying $t_k=\Cset^{j,j'}(\Och_q)\setminus\Cset^{j,j'}(\Och_{q-1})$,
+  for $q=1,\dots, i$.
   % first, get q* as the sample of deme-transition
@@
     % if not exists q* and q'
-    \item If $q^*=\emptyset$ and $q'=\emptyset$, then $t_k\notin\Zset^{j,j'}(\Och_i)\cup\Cset^{j,j'}(\Och_{i})$. So for any $q\leq i$, we have
+    \item If $\nexists q^*$ and $\nexists q'$, then $t_k\notin\Zset^{j,j'}(\Och_i)\cup\Cset^{j,j'}(\Och_{i})$. We have
       \begin{equation*}
@@
     % if not exists q* but q'
-    \item If $q^*\neq\emptyset$ but $q'=\emptyset$, then $t_k\in\Zset^{j,j'}(\Och_{i})$. Let $\ell^*=\ell_{q^*k}^{j'}$ and for $q^*<q\leq i$, we have
+    \item If $\exists q^*$ but $\nexists q'$, then $t_k\in\Zset^{j,j'}(\Och_{i})$ and $\ell^*=\ell_{q^*k}^{j'}$. With the same definitions of $Q_k^{j'}$ and $\mathcal{Y}^{j}_{m}$ as before, we have
       \begin{equation*}
       \begin{split}
-        \Phi_k &= \prod_{m=0}^{\ell^*}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell^*)}\,\prod_{m=\ell_{q^*k}^{j}}^{\ell_{ik}^{j}}\left(1-\frac{1}{I_k^{j}-m}\right) \\
-        &= \frac{I_k^{j'}-\ell^*-1}{I_k^{j'}}\frac{1}{I_k^j(I_k^{j'}-\ell^*)}\frac{I_k^j-\ell_{ik}^{j}-1}{I_k^j-\ell_{q^*k}^{j}}\\
-        &= \frac{1}{\mathcal{Y}^{j}_{\ell^*}(I_k^{j'}-\ell^*)}\left(1-\frac{\ell_{ik}^{j}+1}{I_k^j}\right)\left(1-\frac{\ell^*+1}{I_k^{j'}}\right),
+        \Phi_k &= \prod_{m=0}^{\ell^*-1}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell^*)}\,\prod_{m=\ell_{q^*k}^{j}}^{\ell_{ik}^{j}}\left(1-\frac{1}{I_k^{j}-m}\right) \\
+        &= \frac{I_k^{j'}-\ell^*}{I_k^{j'}}\frac{1}{I_k^j(I_k^{j'}-\ell^*)}\frac{I_k^j-\ell_{ik}^{j}-1}{I_k^j-\ell_{q^*k}^{j}}\\
+        &= \frac{1}{\mathcal{Y}^{j}_{\ell^*+1}I_k^{j'}}\left(1-\frac{\ell_{ik}^{j}+1}{I_k^j}\right).
       \end{split}
       \end{equation*}
-      where $\mathcal{Y}^{j}_{\ell^*}$ is as defined in \ref{scn: notZ} when $u_k\in\Mset^{j,j'}$.
     % if not exists q' but q*
-    \item If $q'\neq\emptyset$ but $q^*=\emptyset$, then $t_k\notin\Zset^{j,j'}(\Och_{q'-1})$ and $t_k\in\Cset^{j,j'}(\Och_i)$. Let $\ell'=\ell_{q'k}^{j'}$ and for $q'<q\leq i$, we have
+    \item If $\nexists q^*$ but $\exists q'$, then $t_k\notin\Zset^{j,j'}(\Och_{q'-1})$ and $t_k\in\Cset^{j,j'}(\Och_i)$. Let $\ell'=\ell_{q'k}^{j'}$, we have
       \begin{equation*}
         \begin{split}
-          \Phi_k &= \prod_{m=0}^{\ell'}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell')}=\left(1-\frac{\ell'+1}{I_{k}^{j'}}\right)\frac{1}{I_k^{j}(I_k^{j'}-\ell')}.
+          \Phi_k &= \prod_{m=0}^{\ell'-1}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell')}=\frac{1}{I_k^{j}I_k^{j'}}.
         \end{split}
@@
     % if exists q* and q', and q' > q*
-    \item If $q^*\neq\emptyset$, $q'\neq\emptyset$, and obviously, $q^*<q'$, then $t_k\in\Zset^{j,j'}(\Och_{q^*})$, $t_k\notin\Zset^{j,j'}(\Och_{q'})$, and $t_k\in\Cset^{j,j'}(\Och_{q'})$.
+    \item If $\exists q^*$ and $\exists q'$, and obviously, $q^*<q'\leq i$, then $t_k\in\Zset^{j,j'}(\Och_{q^*})$, $t_k\notin\Zset^{j,j'}(\Och_{q'})$, and $t_k\in\Cset^{j,j'}(\Och_{q'})$.
 Let $\ell^*=\ell_{q^*k}^{j'}$ and $\hat{\ell}=\ell_{q'k}^{j}$.
@@
       \begin{split}
-        \Phi_k &= \prod_{m=0}^{\ell^*}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell^*)}\prod_{m=\ell_{q^*k}^{j}}^{\hat{\ell}}\left(1-\frac{1}{I_k^{j}-m}\right)\,\frac{1}{I_k^{j}-\hat{\ell}} \\
-        &= \frac{1}{\mathcal{Y}_{\ell^*}^{j}(I_k^{j}-\hat{\ell})(I_k^{j'}-\ell^*)}\left(1-\frac{\hat{\ell}+1}{I_k^{j}}\right)\left(1-\frac{\ell^*+1}{I_k^{j'}}\right)
+        \Phi_k &= \prod_{m=0}^{\ell^*-1}\left(1-\frac{1}{I_k^{j'}-m}\right)\,\frac{1}{I_k^{j}(I_k^{j'}-\ell^*)}\prod_{m=\ell_{q^*k}^{j}}^{\hat{\ell}-1}\left(1-\frac{1}{I_k^{j}-m}\right)\,\frac{1}{I_k^{j}-\hat{\ell}} \\
+        &= \frac{1}{\mathcal{Y}_{\ell^*+1}^{j}I_k^{j}I_k^{j'}}.
       \end{split}
@@
 
-\bibliographystyle{../preprint}
+\bibliographystyle{preprint}
 \bibliography{phylopomp}

### 2023-06-15 figs/markov_state.tex
major work

%% \documentclass[crop,tikz]{standalone}
%% \input{defs}
%% \begin{document}
\begin{tikzpicture}[scale=1]
  \usetikzlibrary{shapes,arrows,positioning}
  \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
  \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\large, fill=white, thick, minimum size=5em]
  \tikzstyle{trans}=[color=black, thick, >=stealth]
  \node[state] (s0) at (0,0) {$x=(S,E,I,R)$};
  \node[state] (s1) at (165:7.8) {$x'=(S-1,E+1,I,R)$};
  \node[state] (s2) at (195:7.1) {$x'=(S,E-1,I+1,R+1)$};
  \node[state] (s3) at (295:4) {$x'=(S,E,I-1,R+1)$};
  \node[state] (s4) at (355:9) {$x'=(S+1,E,I,R-1)$};
  \node[state] (s5) at (18:6) {$x'=(S,E,I,R)$};
  \draw[trans,->] (s0) -- (s1) node[midway,above,sloped] {transmission};
  \draw[trans,->] (s0) -- (s2) node[midway,above,sloped] {progression};
  \draw[trans,->] (s0) -- (s3) node[midway,above,sloped] {recovery};
  \draw[trans,->] (s0) -- (s4) node[midway,above,sloped] {loss of immunity};
  \draw[trans,->] (s0) -- (s5) node[midway,above,sloped] {sampling};
\end{tikzpicture}
%% \end{document}

### 2023-06-15 ms.Rnw
major work

--- 2022-02-15 ms.Rnw
@@
 \documentclass[11pt,reqno,final]{amsart}
-\usepackage{tikz}
-\input{defs}
 \usepackage[round,elide]{natbib}
 \input{header}
+\usepackage{tikz}
+\usetikzlibrary{arrows.meta,positioning,decorations,calc,math}
 
-\title{Structured Genealogy Processes}
+\title[Exact Phylodynamics]{Exact Phylodynamics via Structured Markov Genealogy Processes}
+\author[King]{Aaron~A.~King}
+\address{
+  A.~A.~King,
+  Department of Ecology \& Evolutionary Biology,
+  Center for the Study of Complex Systems, and
+  Department of Mathematics,
+  University of Michigan,
+  Ann Arbor, MI 48109 USA
+}
+\email{kingaa@umich.edu}
+\urladdr{\href{https://kinglab.eeb.lsa.umich.edu/}{https://kinglab.eeb.lsa.umich.edu/}}
 \author[Lin]{Qianying Lin}
-\author[King]{Aaron~A.~King}
+\address{
+  Q.-Y. Lin,
+  Theoretical Biology and Biophysics,
+  Los Alamos National Laboratory,
+  Los Alamos, NM XXXXX USA
+}
 \author[Ionides]{Edward~L.~Ionides}
-
+\address{
+  E.~L.~Ionides,
+  Department of Statistics
+  University of Michigan,
+  Ann Arbor, MI 48109 USA
+}
 \date{\today}
 
-\hypersetup{pdftitle={Structured Genealogy Processes}}
-\hypersetup{pdfauthor={Q.-Y. Lin, A.A. King, E.L. Ionides}}
+\hypersetup{pdftitle={Exact Phylodynamics via Structured Markov Genealogy Processes}}
+\hypersetup{pdfauthor={A.A. King, Q.-Y. Lin, E.L. Ionides}}
 \hypersetup{urlcolor=blue,citecolor=blue,linkcolor=blue,filecolor=blue}
@@
 <<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
-library(knitr)
-prefix <- "strgp"
-read_chunk("setup.R")
-@
-<<setup,include=FALSE,cache=FALSE,purl=FALSE>>=
+prefix <- "smgp"
+source("setup.R")
 @
@@
 library(pomp)
+library(cowplot)
 library(phylopomp)
-library(cowplot)
-stopifnot(getRversion() >= "4.0")
-stopifnot(packageVersion("pomp")>="3.2")
-stopifnot(packageVersion("phylopomp")>="0.0.14")
+stopifnot(getRversion() >= "4.3")
+stopifnot(packageVersion("pomp")>="5.1")
+stopifnot(packageVersion("phylopomp")>="0.9.2")
 theme_set(theme_bw(base_family="serif"))
@@
 \begin{document}
-% \SweaveOpts{concordance=TRUE}
+
 \maketitle
@@
 \begin{abstract}
-This is a followup paper on Markov Genealogy Process. 
-We construct a continuous time Markov Process, called Structure Genealogy Process, to include structures in the focal population with a moderate adjustments in the setting of individuals, genealogies, and genealogy processes.
-We derive the exact expression of likelihood of a structured genealogy and develop simulation-based algorithms to conduct statistical inferences. 
-Numerical simulation and real-world examples are included.
+
 \end{abstract}
@@
 
-\section{Mathematical settings}
-\label{sec:settings}
-\paragraph{Markov genealogy process}
-A population process $\sg{X}_t\in\Xspace$ as a non-explosive Markov jump process is defined, where $t\in\Time$ indicates the time and the population is divided into $J\in\N$ demes, with initial-state distribution $p_0(x)$.
-We then define the jump rate functions $\jumprate_u(t,x)\in\rate$, where $u, x\in\Xspace$ are the jump event and states, respectively. 
-Thus, the population process $\sg{X}_t$ is a stochastic process by defining a sequence of jumps, $\omega=\left(t_k, u_k, n_k\right)_{k=1}^K$, where $K\in\N\cup\{\infty\}$ is the total number of jumps and for the $k$-th jump, $t_k\in\Time$ is the time, $u_k\in\Uspace$ is the event type, and $n_{k}\in\N$ is the label of the individual who conducts $u_k$.
-The full construction can be referred to \citet{King2021}. We define the following functions on $\omega\in\Omega$:
+Problem of phylodynamics.
+Factorization of problem into two subproblems.
 
-\begin{equation}
-  T_k(\omega)\coloneqq t_k,\qquad U_k(\omega)\coloneqq u_k,\qquad N_k(\omega)\coloneqq n_k,\qquad K(\omega)\coloneqq K
-\end{equation}
+Relation to previous work.
+Existing methods \citep{Volz2009a,Stadler2010}.
+Large-population, small sample-size approximations.
 
-and
+Extension of previous results \citep{King2022}.
+Broader class of state-spaces.
+Accommodating discrete structure.
 
-\begin{equation}
-  T(\omega)\coloneqq\left(T_k(\omega)\right)^{K(\omega)}_{k=0},\qquad U(\omega)\coloneqq\left(U_k(\omega)\right)^{K(\omega)}_{k=0},\qquad N(\omega,j)\coloneqq\left(N_k(\omega)\right)^{K(\omega)}_{k=0}.
-\end{equation}
+Classes of Markov processes.
+Utility and flexibility of Markov assumptions.
 
-% \QYL{May not be necessary to repeat the GP paper.}
-% 
-% Obviously, there's a unique mapping from $\omega$ to the population process up to time $t\in\Time$.
-For $\omega\in\Omega$, we then can define the $t$-\emph{restriction} $\omega\vert_t$ and the sample path $\sg{X}_t(\omega)$ of $\omega$:
-% 
-% \begin{equation}
-%   \omega\vert_t\coloneqq\max\{\omega'\in\Omega\,\vert\,\omega'\preceq\omega\text{ and }T_K(\omega')\le t\}
-%   \quad\text{and}\quad
-%   \sg{X}_t(\omega)=\sum_{k=0}^{K(\omega)}U_k(\omega)\indicator{\left[T_k(\omega),\infty\right)}(t),
-% \end{equation}
-% 
-% where $\omega'\preceq\omega$ implies that $\omega$ is an extension of $\omega'$ and $\indicator{}$ is an indicator function.
+Population process induces Markov history and genealogy processes.
+Using these, we derive equations for the likelihood of a genealogy conditional on the history.
+We then integrate out the history to obtain nonlinear filtering equations, the solution of which yields the likelihood.
+These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likellihood.
+We demonstrate with several examples.
+
+\section{Population processes}
+
+Motivating examples: compartmental models.
+Wide variety of models.
+Linear chain trick.
+Migration, superspreading, competition between strains.
+
+Figure: example compartmental model with multiple demes.
 
 \begin{figure}
-  \resizebox{0.6\linewidth}{!}{
-    \begin{tikzpicture}[scale=1.8]
-      \usetikzlibrary{arrows}
-      \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-      \tikzstyle{det}=[color=black, thick, >=stealth]
-      \tikzstyle{stoch}=[color=black, dashed, thick, >=stealth]
-      \node (A) at (0,0) {$\Master_t$};
-      \node (H) at (30:1) {$\Hist_t$};
-      \node (X) at (30:2) {$\X_t$};
-      \node (G) at (-30:1) {$\Gen_t$};
-      \node (I) at (-30:2) {$\Inv_t$};
-      \node (P) at (0:1.732) {$\Prune_t$};
-      \node (O) at (1.732,.5) {$\Och_i$};
-      \node (V) at (0:2.732) {$\Vis_t$};
-      \node (W) at (2.732,.5) {$\Wch_i$};
-      \draw [det,->] (A) -- (H);
-      \draw [det,->] (H) -- (X);
-      \draw [det,->] (A) -- (G);
-      \draw [det,->] (G) -- (I) node[midway,below,sloped] {$\mathrm{ext}$};
-      \draw [det,->] (G) -- (P) node[midway,below,sloped] {$\mathrm{prune}$};
-      \draw [det,->] (P) -- (V) node[midway,below,sloped] {$\mathrm{obs}$};
-      \draw [det,->] (P) -- (O);
-      \draw [det,->] (V) -- (W);
-      \draw [stoch,->] (H) -- (G);
-    \end{tikzpicture}
-  }
+  \begin{center}
+    \resizebox{0.9\linewidth}{!}{\input{figs/model_diagrams}}
+  \end{center}
   \caption{
-    \label{fig:constellation}
-    Relations among the various Markov processes discussed in the paper.
-    Deterministic maps are indicated with solid arrows;
-    random maps are shown as dashed arrows.
-    All the maps shown commute.
-    $\X_t$ is the \emph{population process}, a model of the dynamics of some system, which we take as a starting point.
-    $\Hist_t$ is the \emph{history process}, which records the full history of $\X_t$.
-    $\Inv_t$ is the \emph{inventory process}: at each time $t$, $\Inv_t$ is an inventory of all extant individuals in the population, each of which has a globally unique name.
-    $\Gen_t$ is the \emph{genealogy process}, which captures the precise genealogical relationships among all individuals in $\Inv_t$, as well as among any samples that have been taken from the population.
-    $\Vis_t$ is the \emph{visible genealogy process}, which is $\Gen_t$ pruned so that only relationships among samples remain.
-    Finally $\Wch_i$ is the \emph{embedded chain of the visible genealogy process}, which is $\Vis_{s_i}$, $s_i$ being the time of the $i$-th sample.
-    All of these processes can be obtained via deterministic procedures applied to the \emph{master process} $\Master_t$, as described in the text.
+    Examples of compartmental models.
+    Demes are shaded.
+    \label{fig:examples}
   }
@@
 
-\paragraph{Births}
-Within a structured infective population, the parent and the child, at a given birth event, can be in different demes.
-That is, an individual in Deme $j$ can give birth to a newborn individual in Deme $j'$, where $j,j'=1,\dots,J$. 
-We then define the birth function $B_{j,j'}:\Uspace\to\N$ for any $j,j'=1,\dots,J$. 
-The set of birth events where an individual in Deme $j$ gives birth to $n$ individuals in Deme $j'$ is $\Bset^{j,j'}\coloneqq\,B_{j,j'}^{-1}(\{1\})$. 
-Trivially, the set of birth events in Deme $j'$ is $\Bset^{j'}\coloneqq\bigcup_{j}\Bset^{j,j}$.
+\begin{figure}
+  \begin{center}
+    \resizebox{0.9\linewidth}{!}{\input{figs/markov_state}}
+  \end{center}
+  \caption{
+    Markov state transition diagram for an SEIR model.
+    The state is characterized by four numbers, $S$, $E$, $I$, and $R$.
+    From a given state $x$, there are four possible jumps $x\mapsto{x'}$.
+    From the point of view of the induced genealogy process,
+    transmission is of birth type,
+    progression is of migration type,
+    and recovery is of death type, while loss of immunity is of neutral type.
+    In this example, when a sampling event occurs, the state does not change.
+    \label{fig:markov_state}
+  }
+\end{figure}
 
-\paragraph{Migrations}
-Infective demes are interchangeable and we define the migration function from Deme $j$ to $j'$ for any $j\neq j'$ and $j,j'=1,\dots,J$ as $M_{j,j'}:\Uspace\to\N$ to represent the number of infectives migrating to Deme $j'$ from Deme $j$.
-Thus the set of migration is $\Mset^{j,j'}\coloneqq\,M^{-1}_{j,j'}(\{1\})$ since we assume single migration at a single migration event. 
-Similarly, we disallow coninciding migration events and other events.
+\paragraph{Mathematical formalism}
 
-\paragraph{Samples}
-We assume samples are taken serially and no births or deaths are coincided. 
-We then define the sampling function on the probability space of the population process $\sg{X}_t$: $G_j:\Uspace\to\N_0$ is the number of samples from Deme $j$ taken at event $u$, then the set of sampling events is $\Sset^j\coloneqq\,G_j^{-1}(\{1\})$.
-Of course, the non-coincidence assumption can be easily relaxed while it's not the focus in this paper.
+Definition of state space.
+Jumps and jump rates.
+Demes.
 
-\paragraph{Deaths and population size}
-The population, the set of infected individuals in the context of epidemiology, is structured, categorized into interchangeable demes. 
-We can then define population size function and death functions in Deme $j$, respectively, as $I^{j}, D_{j}:\Xspace\to\N$ for any $j=1,\cdots,J$. 
+Kolmogorov equations.
+
+\subsection{Examples}
+
+\paragraph{SEIRS model}
+
+\paragraph{SIIR model}
+
+\paragraph{Linear birth-death model}
+
+\paragraph{Moran model and the Kingman coalescent}
+
+\section{History and genealogy processes}
+
+\subsection{Mathematical preliminaries}
+
+State space.
+Definition of $\Omega$ as space of \cadlag\ functions of time.
+Define everything in terms of \cadlag\ functions rather than sequences.
+Then the topology comes directly from the Skorokhod metric.
+Explain we use a base measure coming from unit-rate Poisson processes.
+
+Restrictions.
+Metric, Borel $\sigma$-algebra, base measure.
+
+Probability density.
+Master process.
+
+\paragraph{Markov jump processes}
+
+We will assume that our population process is a non-explosive Markov jump process $\X_t$ parameterized by time $t\in\Rp$, and taking values in some space $\Xspace$.
+In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but we dispense with that restriction here.
+The population process is completely specified by its initial-state distribution and its generator.
+In particular, we suppose that
+\begin{equation}\label{eq:ic}
+  \prob{\X_0=x}=p_0(x),
+\end{equation}
+for some choice of initial distribution, $p_0$.
+We suppose there are countably many types of jumps and denote the set of jump types by $\Jumps$.
+For each $u\in\Jumps$, there is a corresponding jump rate or hazard $\alpha_u$.
+We impose the further requirement that $\sum_{ux'}\alpha_u(t,x,x')<\infty$ for all $t,x$.
+We will always use a version of $\X_t$ in which the sample paths are right-continuous with left limits (\cadlag).
+
+The above may be compactly summarized by means of the corresponding Kolmogorov equations, as follows.
+Given $f\in L^{\infty}(\Xspace)$, let $F(s,x)\coloneq\expect{f(\X_t)\;|\;\X_s=x}$.
+It then follows that, for $s<t$ and $x\in\Xspace$,
+\begin{equation}\label{eq:gen}
+  \begin{gathered}
+    \frac{\partial{F}}{\partial{s}}(s,x)=-\int\alpha(s,x,x')\,\left[F(s,x')-F(s,x)\right]\,\dd{x'},
+    \qquad
+    F(t,x)=f(x).
+  \end{gathered}
+\end{equation}
+\Cref{eq:gen} is the so-called Kolmogorov backward equation.
+Its adjoint is the Kolmogorov forward equation (KFE, also called the \emph{master equation} in this context), which states that for $t\ge{0}$, $x\in\Xspace$,
+\begin{equation}\label[pluralequation]{eq:kfe}
+  \begin{gathered}
+    \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
+    \qquad
+    w(0,x) = p_0(x).
+  \end{gathered}
+\end{equation}
+If $w(t,x)$ satisfies \cref{eq:kfe}, then $w(t,x)=\prob{\X_t=x}$.
+
+\paragraph{Definitions}
+
+Our goal in this paper is to introduce a family of Markov processes induced by the jump process just described (\cref{fig:constellation}).
+While population processes of the kind described above can be constructed in a variety of time-honored ways \citep[\eg][]{Andersen1993,Kallenberg1997}, these classical approaches are not entirely sufficient for the tree-valued Markov processes we will erect on top of the population process.
+We therefore explicitly construct the probability space that underlies the stochastic processes we subsequently describe.
+Unavoidably, this leads to technicalities that are necessary for the firm establishment of the properties of these processes but that may distract from the overarching goals.
+Readers willing to stipulate these properties may skim the remainder of this section, in which we formally construct the probability space and make some definitions that will be needed in the sequel.
+
+Let us define a \emph{jump} to be an ordered quadruple $(t,u,x,n)\in\Rp\times\Jumps\times\Xspace\times\Zp$.
+We refer to $t$ as the \emph{time} of the jump;
+$u$ is the \emph{type} of the jump;
+$x$ is the state achieved at the end of the jump;
+$n$ is an \emph{auxiliary number} whose use will be made clear below.
+A \emph{jump sequence} is a countable sequence of jumps at increasing times.
+That is,
+\begin{equation*}
+  \omega=\left(t_k,u_k,x_k,n_k\right)_{k=0}^{K}
+\end{equation*}
+is a jump sequence if and only if $K\in\Zp\cup\{\infty\}$, $u_k\in\Jumps$, $x_k\in\Xspace$, $n_k\in\Zp$ for all $k$, and $0=t_0<t_1<t_2<\dots$.
+We will take our sample space, $\Omega$, to be the set of all jump sequences.
+For $\omega\in\Omega$ as above, we write
+\begin{equation}\label{eq:components1}
+  \begin{gathered}
+    T_k(\omega)\coloneq t_k, \qquad
+    U_k(\omega)\coloneq u_k, \qquad
+    X_k(\omega)\coloneq x_k, \qquad
+    N_k(\omega)\coloneq n_k, \qquad
+    K(\omega)\coloneq K
+  \end{gathered}
+\end{equation}
+and also
+\begin{equation}\label{eq:components2}
+  \begin{gathered}
+    T(\omega)\coloneq\left(T_k(\omega)\right)_{k=0}^{K(\omega)}, \qquad
+    U(\omega)\coloneq\left(U_k(\omega)\right)_{k=0}^{K(\omega)}, \\
+    X(\omega)\coloneq\left(X_k(\omega)\right)_{k=0}^{K(\omega)}, \qquad
+    N(\omega)\coloneq\left(N_k(\omega)\right)_{k=0}^{K(\omega)}.
+  \end{gathered}
+\end{equation}
+We will denote by $\mathring{\Omega}$ the set of all finite jump sequences, \ie $\mathring{\Omega}\coloneq\{\omega\in\Omega\;|\;K(\omega)<\infty\}$.
+
+Note that every element of $\Omega$ corresponds to a unique sample path of $\X_t$.
+In particular,
+\begin{equation}\label{eq:Xdef}
+  \X_t(\omega)=\sum_{k=0}^{K(\omega)}\!X_k(\omega)\,\indicator{\halfopen{T_k(\omega),\infty}}(t).
+\end{equation}
+Note that, for every $\omega\in\Omega$, the function $\X(\omega):\Rp\to\Xspace$ is thus \cadlag.
+It will be useful to define $\Xt$ to be the left-limit function of $\X$:
+\begin{equation*}
+  \Xt(\omega,t)\coloneq
+  \begin{cases}
+    \lim_{s\uparrow{t}}\X(\omega,t), & t>0,\\
+    \X(0), & t=0.\\
+  \end{cases}
+\end{equation*}
+
+The sample space $\Omega$ has a natural partial order.
+We write $\omega\preceq\omega'$ if $\omega'$ is an extension of $\omega$;
+that is, if $K(\omega)\le K(\omega')$, and $\left(T_k(\omega),U_k(\omega),X_k(\omega),N_k(\omega)\right)=\left(T_k(\omega'),U_k(\omega'),X_k(\omega'),N_k(\omega')\right)$ for $k=0,\dots,K(\omega)$.
+For $\omega\in\Omega$, the set
+$\left\{\omega'\in\Omega\;\vert\;\omega'\preceq\omega\right\}$
+is totally ordered.
+Moreover, for each $\omega\in\mathring{\Omega}$, there is a unique \emph{predecessor}, $\predec{\omega}$, such that $\predec{\omega}\preceq\omega$, $\predec{\omega}\ne\omega$, and, for all $\omega'$, $\predec{\omega}\preceq\omega'\preceq\omega$ implies that either $\omega'=\predec{\omega}$ or $\omega'=\omega$.
+
+In order to define probabilistic \emph{events}, it is necessary to define the $\sigma$-algebra of measurable subsets of $\Omega$.
+We first impose the condition that $\Xspace$ be a complete, separable, measurable metric space.
+We then follow \citet{King2022}, extending the Skorokhod metric \citep{Kallenberg1997,Ethier2009} to make $\Omega$ a complete, separable metric space (\ie a Polish space) and take our event space, $\Borel$, to be its Borel $\sigma$-algebra.
+
+We now turn to the probability measure, $\mathbb{P}$, on $\Omega\cong T(\Omega){\times}U(\Omega){\times}X(\Omega){\times}N(\Omega)$.
+We will specify this by giving its density.
+%% We will specify this by giving its density with respect to a base measure.
+%% A natural base measure on $T(\Omega)$ is the probability measure of the unit-rate Poisson point process \citep{Andersen1993,Kallenberg1997};
+%% the measure on $\Xspace$ extends naturally to a product measure on $X(\Omega)$;
+%% the counting measure serves for the discrete component $U(\Omega){\times}N(\Omega)$.
+%% Let $\pi$ denote the product of these three measures.
+%% We define a probability measure on $\Omega$ by specifying its density with respect to $\pi$.
+In particular, for each $u\in\Jumps$, $x,x'\in\Xspace$, we suppose $\auxmeas_{u,x,x'}$ is a given probability measure on $\Zp$;
+these will take specific forms below.
+For $t\in\Rp$ and $\omega\in\restrict{\Omega}{t}$, define the probability density function
+\begin{equation}\label{eq:probmeas}
+  P_t(\omega)\:\coloneq\:p_0(\X_0)\,\prod_{k=1}^{K}\left(\alpha_{U_k}(T_k,\Xt_{T_k},\X_{T_k})\,\auxmeas_{U_k,\Xt_{T_k},X_{T_k}}(N_k)\right)\exp{\left(-\int_0^t\sum_{ux'}{\alpha_u(s,\X_s,x')}\,\dd{s}\right)},
+\end{equation}
+\normalsize
+the latter sum being taken over $u\in\Jumps$, $x'\in\Xspace$.
+Here, for the sake of readability, we have suppressed the dependence of the random variables $\X_t$, $K$, $T_k$, $U_k$ and $N_k$ on $\omega$.
+%% It is readily verified that, for all $t$, $P_t$ is a probability density with respect to $\pi$.
+
+\paragraph{History process}
+
+Next among our constellation of related processes (\cref{fig:constellation}) is the \emph{history process}, which encapsulates the entire history of $\X$ up to time $t$.
+Specifically, for $\omega\in\Omega$, we define
+\begin{equation*}
+  \Hist_t(\omega) \coloneq \left(t,\left(T_k(\omega),U_k(\omega),X_k(\omega)\right)_{k=0}^{K(\restrict{\omega}{t})}\right),
+\end{equation*}
+where $T_k$, $U_k$, $X_k$, and $K$ are as in \cref{eq:components1}.
+Thus $\Hist_t$ contains exactly those elements of $\omega$ that are relevant to the history of the population process $\X_t$.
+It is trivial to verify that $\Hist_t$ is Markov and to compute its probability density.
+In particular, given any history $h_t=\left(t,\left(t_k,u_k,x_k\right)_{j=0}^{K}\right)$,
+the marginal density at $h_t$ is readily seen to be
+\begin{equation*}
+  \begin{aligned}
+    P_{\Hist_t}(h_t)\coloneq &p_0(x_0)\,\prod_{k=1}^{K}\left\{\alpha_{u_k}(t_k,x_{k-1},x_k)\,\exp{\left(-\int_{t_{k-1}}^{t_k}\sum_{ux'}{\alpha_u(s,x_{k-1},x')}\,\dd{s}\right)}\right\}\\
+    &\qquad\times\,\exp{\left(-\int_{t_{K}}^{t}\sum_{ux'}{\alpha_u(s,x_{K},x')}\,\dd{s}\right)}.
+  \end{aligned}
+\end{equation*}
+
+\paragraph{Inventory process}
+
+Our main interest is in tracking the genealogical relationships among the lineages in our demes.
+Accordingly, we will assign each individual lineage a unique number.
+let $\Inv_t:\Rp\to\Demesplus^{\Zp}$ be such that $\Inv_t(n)=i$ if and only if lineage $n$ is in deme $i$ at time $t$.
+If $t$ is before the birth or after the death of $n$, $\Inv_t(n)=\eth$.
+Note that $n_i(\X_t)=\left\vert\left\{n\;\vert\;\Inv_t(n)=i\right\}\right\vert$ for all $t\in\Rp$ and $i\in\Demes$.
+
+\paragraph{Genealogy process}
+
+\paragraph{Pruning}
+
+\paragraph{Filter equations}
+
+\begin{defn}
+  Suppose $X_t$ is a Markov process with Kolmogorov forward equation (KFE)
+  \begin{equation}\label{eq:kfe2}
+    \begin{gathered}
+      \frac{\partial{w}}{\partial{t}}=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
+      \qquad
+      w(0,x)=p_0(x).
+    \end{gathered}
+  \end{equation}
+  Suppose that $B(x,x')>0$ and $\lambda(t,x)>0$ are given functions, and $t_1<t_2<{\dots}<t_K$ is a given sequence of times.
+  We say that the equations
+  \begin{gather}
+    \frac{\partial{w}}{\partial{t}}=\int\!w(t,x')\,\alpha(t,x',x)\,B(x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'}-\lambda(t,x)\,w(t,x),\label{eq:filtereq1}\\
+    w(t_k,x)=\int\!\wt(t_k,x')\,\frac{\alpha(t,x',x)}{\mu}\,B(x',x)\,\dd{x'},\quad k=1,\dots,K,\label{eq:filtereq2}\\
+    w(0,x)=p_0(x).
+  \end{gather}
+  are the \emph{filter equations} for $X_t$ with \emph{boost} $B$, \emph{decay} $\lambda$, \emph{initial density} $p_0$, and \emph{observation times} $t_k$.
+\end{defn}
+
+The useful fact about filter equations is that they afford a means of computing the likelihood of a given sequence of events.
+This is facilitated by the following
+\begin{lemma}\label{lemma:filtereq}
+  \Cref{eq:filtereq1} is satisfied by $w(t,x)=\int{v\,U(t,x,v)\,\dd{v}}$, where $U$ satisfies the KFE
+  \begin{equation}\label{eq:filterlemma}
+    \begin{aligned}
+      \frac{\partial{U}}{\partial{t}}=&\int{U(t,x',v')\,\alpha(t,x',x)\,\delta(v,B(x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
+      &\qquad-\int{U(t,x,v)\,\alpha(t,x,x')\,\delta(v',B(x,x')\,v)\,\dd{x'}\,\dd{v'}}
+      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,U(t,x,v)\right].
+    \end{aligned}
+  \end{equation}
+  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
+\end{lemma}
+\begin{proof}
+  \begin{equation*}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}
+      =&\int{v\,\frac{\partial{U}}{\partial{t}}(t,x,v)\,\dd{v}}\\
+      =&\int{v\,U(t,x',v')\,\alpha(t,x',x)\,\delta(v,B(x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
+      &\qquad-\int{v\,U(t,x,v)\,\alpha(t,x,x')\,\delta(v',B(x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
+      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,U(t,x,v)\right]\,\dd{v}}.\\
+    \end{aligned}
+  \end{equation*}
+  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
+  \begin{equation*}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}
+      =&\int{v'\,U(t,x',v')\,\alpha(t,x',x)\,B(x',x)\,\dd{v'}\,\dd{x'}}
+      -\int{v\,U(t,x,v)\,\alpha(t,x,x')\,\dd{v}\,\dd{x'}}\\
+      &\qquad-\lambda(t,x)\,\int{v\,U(t,x,v)\,\dd{v}},
+    \end{aligned}
+  \end{equation*}
+  which is simplified to obtain \cref{eq:filtereq1}.
+\end{proof}
+
+We recognize in \cref{eq:filterlemma} the KFE of a certain process $(X_t,V_t)$.
+In particular, $X_t$ is a jump process with KF \cref{eq:kfe2}.
+The $V_t$ process has jumps wherever $X_t$ does, such that when $X_t$ jumps $x\mapsto{x'}$, $V_t$ jumps by a factor of $B(x,x')$.
+Between jumps, $V_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
+If we view $V_t$ as a weight, then \cref{lemma:filtereq} says that the weighted average of $X_t$ evolves according to \cref{eq:filtereq1}.
+
+\paragraph{Lineages, event-counter, deme-residence}
+
+Let $\Lin=\Lin(\Gen)$ denote the finite set of all samples represented in genealogy $\Gen$.
+Let $\prec2$ be any ordering of $\Lin$ that is compatible with ancestry.
+That is, if $j,j'\in\Lin$ are such that $j$ is ancestral to $j'$, then we must have $j\prec2{j'}$.
+Using this ordering, we can uniquely associate each point on a genealogical tree with the least of those lineages that descend from that point.
+In particular, any lineage $j\in\Lin$, corresponding to a sample taken at time $t^s_j$, can be traced backward from node to node until either it coalesces with some lesser lineage at some time $t^o_j>0$ or a root is reached (in which case, we define $t^o_j=0$).
+Each node encountered along the way represents a genealogical event from which $j$ emerges.
+Moreover, at each time $t\in\halfopen{t^o_j,t^s_j}$, lineage $j$ is in precisely one of the demes $D$.
+However, for $t\notin\halfopen{t^o_j,t^s_j}$, lineage $j$ does not exist.
+To express this, we say that lineage $j$ is in the \emph{undeme}, which we denote using the symbol $\eth$.
+We define $\Demesplus\coloneq\Demes\cup\{\eth\}$.
+
+It will be useful to define a function that captures all the relevant features of a pruned genealogy.
+Accordingly, let $\Yspace=(\Zp\times\Demesplus\times\Lin)^{\Lin}$ and define $y:\Rp\to\Yspace$ so that, for $j\in\Lin$ and $t\in\Rp$:
+\begin{compactenum}[(a)]
+\item $\ct(y_j(t))\in\Zp$ is a counting process which increases by 1 at each event along lineage $j$.
+\item $\deme(y_j(t))\in\Demesplus$ indicates in which deme lineage $j$ lies at time $t$.
+  In particular $\deme(y_j(t))=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
+\item $\anc(y_j(t))\in\Lin$ indicates the lineage in which the ancestors of lineage $j$ are found.
+  In particular, $\anc(y_j(t))=j$ for $t\ge{t^o_j}$, but $\anc(y_j(t))$ is well defined for all $j$ and $t$.
+\end{compactenum}
+One can verify that $y$ is \cadlag.
+Define $\yt(t)=\lim_{s\uparrow{t}}y(s)$.
+
+To visualize these functions, one can make a correspondence between demes and colors.
+Then a pruned genealogy is visualized as a tree with colored branches.
+Knowing the function $\deme(y)$ is equivalent to knowing the coloring, while $\ct(y)$ determines the locations of events in the genealogy and $\anc(y)$ determines the topology.
+Note in particular that $y_j(t)\ne\yt_j(t)$ if and only if $t$ is the time of an event from which lineage $j$ emerges.
+
+\paragraph{Lineage count, saturation}
+
+In the following, we will find that we need to count the deme-specific numbers of lineages present at a given time.
+Accordingly, for any $\Lin'\subseteq{\Lin}$, $\eta\in(\Zp\times\Demesplus\times\Lin')^{\Lin'}$, and $i\in\Demes$, let us define
+\begin{equation*}
+  \ell_i(\eta)\coloneq{\left\vert\left\{j\in\Lin'\;\vert\;\deme(\eta_j)=i\right\}\right\vert}\in\Zp, \qquad \ell(\eta)\coloneq(\ell_i(\eta))_{i\in\Demes}\in{\Zp^\Demes}.
+\end{equation*}
+Note that lineages $j$ for which $\eta_j=\eth$ are not counted.
+With this definition, it follows that $\ell_i(y(t))$ is the number of lineages in deme $i$ at time $t$.
+
+We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
+Accordingly, for $\Lin'\subseteq{\Lin}$, $\eta,\eta'\in(\Zp\times\Demesplus\times\Lin')^{\Lin'}$, and $i\in\Demes$, let us define
+\begin{equation*}
+  s_i(\eta,\eta')\coloneq{\left\vert\left\{j\in{\Lin'}\;\vert\;\deme(\eta_j)=i\ \ \&\ \ \ct(\eta'_j)=\ct(\eta_j)+1\right\}\right\vert}, \qquad s(\eta,\eta')\coloneq(s_i(\eta,\eta'))_{i\in\Demes}\in{\Zp^\Demes}.
+\end{equation*}
+With this definition, $s_i(\yt(t),y(t))$ is the number of lineages in deme $i$ that emerge from an event at time $t$ and that, if $t$ is not an event time, then $s(\yt(t),y(t))=0$.
+
+\subsection{Obscuration}
+
+\section{Results}
+
+\begin{defn}
+  For $n,r,\ell,n\in{\Zp^\Demes}$, define the multivariable binomial coefficient by
+  \begin{equation*}
+    \binom{n}{r}\coloneq\prod_{i\in\Demes}\binom{n_i}{r_i}.
+  \end{equation*}
+  Using this, we define the \emph{binomial ratio}
+  \begin{equation*}
+    \BoxProb{n}{\ell}{r}{s}\coloneq\frac{\binom{n-\ell}{r-s}}{\binom{n}{r}}\in{[0,1]}.
+  \end{equation*}
+\end{defn}
+
+In consequence of the Chu-Vandermonde identity, we have
 \begin{equation}
-  \jumprate_u(t,x) > 0 \Longrightarrow I^j(x+u)-I^j(x)=\sum_{j'}\,B_{j',j}(u)+\sum_{j'}M_{j',j}(u)-D_j(u)-\sum_{j'}M_{j,j'}(u),
+  \sum_{s\in\Zp^{\Demes}}\BoxProb{n}{\ell}{r}{s}\binom{\ell}{s}=1
 \end{equation}
 
-for all $x, u\in\Xspace$. We disallow simultaneous death at a single death event at this stage and define the set of death events in Deme $j$ as $\Dset^j\coloneqq\,D_j^{-1}(\{1\})$. 
-We also insist, birth and death events don't coincide for any demes, $j,j'$, that is,  $\Bset_{j}\cap\Dset_{j'}=\emptyset$. 
-Eventually, we can have the function for total population size by letting $I(x)\coloneqq\sum_{j=1}^J\,I^j(x)$ and it's trivial that
+\paragraph{Likelihood of a pruned genealogy}
 
+Now certain population-process events at certain times are incompatible with any given pruned genealogy.
+Example: a sample event where no sample is seen, or a migration event involving a change of deme not observed in the pruned genealogy.
+Define the function $Q_u(\eta,\eta')=1$ if a change $\eta\to\eta'$ at time $t$ is compatible with an event of type $u$ at that time, and $Q_u=0$ otherwise.
+In other words, $Q_u$ is the indicator function for the condition that there exist $\omega\in\Omega$, $t\in\Rp$, and $k\in\Zp$ such that $t=T_k(\omega)$, $u=U_k(\omega)$, $\yt(t,\omega)=\eta$, and $y(t,\omega)=\eta'$.
+Using this, we define
+\begin{equation*}
+  \phi_u(\xi,\eta,\eta')\coloneq\BoxProb{n(\xi)}{\ell(\eta')}{r_u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
+\end{equation*}
+
+\begin{thm}
+  \begin{equation*}
+    \prob{\Gen\vert\Hist}=\prod_{(u,t,x)\in\Hist}\phi_u(x,\yt(t),y(t)).
+  \end{equation*}
+\end{thm}
+
+\begin{figure}
+  \begin{center}
+    \input{figs/thm1diag}
+  \end{center}
+  \caption{\label{fig:thm1diag}}
+\end{figure}
+
+\begin{center}
+  \input{figs/triangle1}
+\end{center}
+
+\begin{equation*}
+  \frac{r!}{(r-s)!}\cdot\frac{(n-\ell)!}{n!}\cdot\frac{(n-r)!}{(n-r-\ell+s)!}
+  =\frac{r!(n-r)!}{n!}\cdot\frac{(n-\ell)!}{(n-\ell-r+s)!(r-s)!}
+  =\BoxProb{n}{\ell}{r}{s}
+\end{equation*}
+
+Suppose we have a pruned genealogy $\Prune^*$, defined on the time-interval $[0,T]$, with event times $0=t_0<t_1<\cdots<t_n=T$.
+From the theorem, we have, for $t\notin\event(\Prune^*)$,
 \begin{equation}
-  \jumprate_u(t,x) > 0 \Longrightarrow I(x+u)-I(x)=\sum_{j,j'}B_{j,j'}(u)-\sum_{j}D_j(u),
+  \frac{\partial{w}}{\partial{t}}=\sum_{u}\int\!w(t,x')\,\alpha_u(t,x',x)\,\phi_u\big(x,y(t),y(t)\big)\,\dd{x'}
+  -\sum_{u}\int\!w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}.
 \end{equation}
+At event times, $t\in\event(\Prune^*)$, one has
+\begin{equation}\label{eq:prune_filt_disc}
+  w(t,x)=\sum_{u}\int\!\wt(t,x')\,\frac{\alpha_u(t_k,x',x)}{\mu}\,\phi_u\big(x,\yt(t),y(t)\big)\,\dd{x'}\\,
+\end{equation}
+In addition, there are the initial and boundary conditions
+\begin{equation}
+  \begin{gathered}
+    w(0,x)=p_0(x), \qquad w(t,x)=0\quad\text{whenever}\quad n(x)<\ell(y(t)).
+  \end{gathered}
+\end{equation}
+
+\paragraph{Likelihood of an obscured genealogy}
+
+Let $\kappa^{\Vis}$ be the indicator function for the condition that a pruned genealogy is compatible with a given obscured genealogy.
+In particular, given an obscured genealogy $\Vis$, set $\kappa^{\Vis}_u(t,x,x',\eta,\eta')=\Indicator{\exists\Prune\ \st\ \kappa^{\Prune}_u(t,x,x',y,y')=1\ \&\ \obs(\Prune)=\Vis}$.
 
@@
 
-Some commonly used compartmental models in epidemiology are in fact structured, either explicitly or implicitly. 
-Here, we demonstrate below how these representatives fit within the population processes we described previously.
+\subsection{SEIRS}
 
-\begin{figure}
-\begin{minipage}{.5\textwidth}
-    \begin{tikzpicture}[scale=1.5]
-      \usetikzlibrary{arrows}
-      \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-      \tikzstyle{det}=[color=black, thick, >=stealth]
-      \tikzstyle{stoch}=[color=black, dashed, thick, >=stealth]
-      \node (S) at (-2,0) {$S$};
-      \node (E) at (-1,0) {$E$};
-      \node (I) at (0,0) {$I$};
-      \node (R) at (1,0) {$R$};
-      \draw [det,->] (S) -- (E);
-      \draw [det,->] (E) -- (I);
-      \draw [det,->] (I) -- (R);
-      \draw [det,->] (I) -- (0,.5) -- (-1.5,.5) -- (-1.5,0);
-    \end{tikzpicture}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
-    \begin{tikzpicture}[scale=1.5]
-      \usetikzlibrary{arrows}
-      \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-      \tikzstyle{det}=[color=black, thick, >=stealth]
-      \tikzstyle{stoch}=[color=black, dashed, thick, >=stealth]
-      \node (S) at (-2,0) {$S$};
-      \node (I1) at (-.5,.7) {$I_1$};
-      \node (I2) at (-.5,-.7) {$I_2$};
-      \node (R) at (1,0) {$R$};
-      \draw [det,->] (S) -- (I1);
-      \draw [det,->] (S) -- (I2);
-      \draw [det,->] (I1) -- (R);
-      \draw [det,->] (I2) -- (R);
-      \draw [det,->] (-.4,.5) -- (-.4,-.5);
-      \draw [det,->] (-.6,-.5) -- (-.6,.5);
-    \end{tikzpicture}
-\end{minipage}
-\caption{
-  \label{fig:diagram}
-}
-\end{figure}
+Jumps: $\Jumps=\{\mathrm{Inf},\mathrm{Prog},\mathrm{Recov},\mathrm{Wane},\mathrm{Birth},\mathrm{Death_S},\mathrm{Death_E},\mathrm{Death_I},\mathrm{Death_R},\mathrm{Sample}\}$.
 
-\paragraph{SEIR model}
-The SEIR model is a simple extension of the most basic SIR model, by adding the latent infected compartment ``E'' into the system, where susceptible individual first become ``exposed'' when being infected, then turn into being infectious and eventually recover or get removed from the system.
-The infected population, which we are mostly interested in, is divided into two demes:
-\begin{inparaenum}
-  \item ``E'' represents those that are not able to spread the disease though infected and 
-  \item ``I'' are those that are infectious. 
-\end{inparaenum}
-To fit within the aforementioned population process, we take $d=4$ so that the state vector is $\sg{X}=\left(s,e,i,g\right)$, where $s,e,i,g$ are the number of susceptibles, exposeds, infectives, and the cumulative number of genomic samples collected, respectively.
-We then summarize four types of jumps, with rate functions:
+Demes: $\Demes=\{\mathrm{E},\mathrm{I}\}$.
 
-\begin{equation*}
-  \jumprate_u =
-  \begin{cases}
-     b(t)\,s\,i, & u=\left(-1,1,0,0\right),\,s>0,\,i>0 \\
-    \sigma\,e, & u=\left(0,-1,1,0\right),\,e>0,\\
-    \gamma\,i, & u=\left(0,0,-1,0\right),\,i>0\\
-    \psi(t,s,e,i,g), & u=\left(0,0,0,1\right),\, e+i>0\\
-    0, & \text{otherwise.}
-  \end{cases}
-\end{equation*}
+Jump rates:
+\begin{itemize}
+\item $\alpha_{\mathrm{Inf}}(t,x,x')=\beta(t)\,\frac{x^{\mathrm{S}}x^{\mathrm{I}}}{N(t)}\,\Indicator{x'=x+(-1,1,0,0)}$
+\item $\alpha_{\mathrm{Prog}}(x,x')=\rho\,x^{\mathrm{E}}\,\Indicator{x'=x+(0,-1,1,0)}$
+\item $\alpha_{\mathrm{Recov}}(x,x')=\gamma\,x^{\mathrm{I}}\,\Indicator{x'=x+(0,0,-1,1)}$
+\item $\alpha_{\mathrm{Wane}}(x,x')=\upsilon\,x^{\mathrm{R}}\,\Indicator{x'=x+(1,0,0,-1)}$
+\item $\alpha_{\mathrm{Sample}}(t,x,x')=\psi\,x^I\,\Indicator{x'=x}$
+\item $\alpha_{\mathrm{Birth}}(t,x,x')=B(t)\,\Indicator{x'=x+(1,0,0,0)}$
+\item $\alpha_{\mathrm{Death}_k}(x,x')=\mu\,x^k\,\Indicator{x'^j=x^j-\delta_{jk}}$, $k\in\{\mathrm{S},\mathrm{E},\mathrm{I},\mathrm{R}\}$
+\end{itemize}
 
-The above shows the rates of being infected, progression from exposed to infectious, recovery, and sampling, respectively. 
-As in \citet{King2021}, the transmission $b(t)$ is time-dependent, and the sampling rate $\psi$ is any function, as long as the constraints in Section 2 are satisfied. 
-Therefore, $I_1(\sg{X})=e$, $I_2(\sg{X})=i$, $I(\sg{X})=e+i$, $\Bset^{2,1}=\{\left(-1,1,0,0\right)\}$, $\Mset^{1,2}=\{\left(0,-1,1,0\right)\}$, $\Dset^2=\{\left(0,0,-1,0\right)\}$, $\Sset^2=\{\left(0,0,0,1\right)\}$, and $\Bset^{1,1}=\Bset^{1,2}=\Bset^{2,2}=\Mset^{2,1}=\Dset^1=\Sset^1=\emptyset$.
-
-\paragraph{SI$^2$R model}
-
-We can customize the state and event vector to fit a complex system. 
-In SI$^2$R model, we have two different while interchangeable demes of infectious individuals, the \emph{per capita} transmissibility therefore in the first deme is $b_1(t)$ whilst in the second deme is $b_2(t)$.
-Migration rates between Deme $i$ and $j$ are also defined: Deme $i$ infection turns into Deme $j$ at rate $r_{ij}$ for $i,j=1,2$ and $i\neq j$.
-Furthermore, we also suppose that, once being infected, a susceptible individual becomes Deme 1 infectious with probability $\rho$ while Deme 2 infectious with probability $1-\rho$.
-To specify each event and the number of individuals involved, we define a relatively complex state vector by setting $d=10$ and $\sg{X}=\left(s,i_1,i_2,b_{11},b_{12},b_{21},b_{22},m_{12},m_{21},g\right)$, for $s$, $i_1$, and $i_2$ being the population size of the susceptible, the infectives in Deme 1, and those in Deme 2, respectively, and $b_{11}$, $b_{12}$, $b_{21}$, $b_{22}$, $m_{12}$, and $m_{21}$ being the cumulative number of events of parents in Deme 1 giving to children in Deme 1 and 2, parents in Deme 2 giving birth to children in Deme 1 and 2, individuals migrating from Deme 1 to Deme 2, and vice versa, and sampling, respectively.
-In this case, we can summarize the event rates as:
-
-\begin{equation*}
-  \jumprate_u =
-  \begin{cases}
-                                          %s,i1,i2,b11,b12,b21,b22,m12,m21,g
-     b_1(t)\,s\,i_1\,\rho,        & u=\left(-1,1,0,1,0,0,0,0,0,0\right), s>0, i_1>0 \\
-     b_1(t)\,s\,i_1\,(1-\rho),    & u=\left(-1,0,1,0,1,0,0,0,0,0\right), s>0, i_1>0 \\
-     b_2(t)\,s\,i_2\,\rho,        & u=\left(-1,1,0,0,0,1,0,0,0,0\right), s>0, i_2>0 \\
-     b_2(t)\,s\,i_2\,(1-\rho),    & u=\left(-1,0,1,0,0,0,1,0,0,0\right), s>0, i_2>0 \\
-     r_{12}\,i_1,                 & u=\left(0,-1,1,0,0,0,0,1,0,0\right), i_1>0 \\
-     r_{21}\,i_2,                 & u=\left(0,1,-1,0,0,0,0,0,1,0\right), i_2>0 \\
-     \gamma\,i_1,                 & u=\left(0,-1,0,0,0,0,0,0,0,0\right), i_1>0 \\
-     \gamma\,i_2,                 & u=\left(0,0,-1,0,0,0,0,0,0,0\right), i_2>0 \\
-     \psi(t,s,i_1,i_2,g),         & u=\left(0,0, 0,0,0,0,0,0,0,1\right), i_1+i_2>0\\
-    0, & \text{otherwise.}
-  \end{cases}
-\end{equation*}
-
-The nine non-zero rates above represent those of infection into Deme 1 infected by individuals in Deme 1, infection into Deme 2 infected by individuals in Deme 1, infection in Deme 1 infected by individuals in Deme 2, infection into Deme 2 infected by individuals in Deme 1, migration from Deme 1 to Deme 2, migration from Deme 2 to Deme 1, recovery from Deme 1, recovery from Deme 2, and sampling, respectively.
-We then summarize the population functions and sets as follows:
-
-\begin{align*}
-\begin{array}{lll}
-  I_1(\sg{X})=i_1,                                  & I_2(\sg{X})=i_2,                              & I(\sg{X})=i_1+i_2, \\
-  \Bset_{11}=\{\left(-1,1,0,1,0,0,0,0,0,0\right)\}, & \Bset_{12}=\{\left(-1,0,1,0,1,0,0,0,0,0\right)\}, & \\
-  \Bset_{21}=\{\left(-1,1,0,0,0,1,0,0,0,0\right)\}, & \Bset_{22}=\{\left(-1,0,1,0,0,0,1,0,0,0\right)\}, & \\
-  \Mset_{21}=\{\left(0,-1,1,0,0,0,0,1,0,0\right)\}, & \Mset_{12}=\{\left(0,1,-1,0,0,0,0,0,1,0\right)\}, & \\
-  \Dset_1=\{\left(0,-1,0,0,0,0,0,0,0,0\right)\},    & \Dset_2=\{\left(0,0,-1,0,0,0,0,0,0,0\right)\},    & \\
-  \Sset=\{\left(0,0, 0,0,0,0,0,0,0,1\right)\}. & &
-\end{array}
-\end{align*}
-
-\paragraph{Migration}
-One special case of the SI$^2$R model is that with one-time migration or introduction. 
-In other words, an infectious individual from another population join our interested population at a certain point of time $\tau$ and then stay and spread the disease. 
-In this case, if this specific individual is defined as Deme 1 while other infectious individuals as Deme 2, the migration rate $r_{12}$ is a Dirac delta function $\delta(t)$, where $\delta(t)=\infty$ for $t=\tau$ and $\delta(t)=0$ for $t\neq\tau$, and $r_{21}=0$ all the time.
-
-\paragraph{Superspreading}
-The definition of superspreading is ambiguous in epidemiology. 
-Some researchers view it as a certain individual who has extremely strong transmissibility with high reproduction rate and call this individual \emph{superspreader}; 
-others avoid to invidualize this phenomenon and consider it as an event where a cluster of infections appears in a single event within a small amount of time and define it as a \emph{superspreading event}.
-The first definition can be represented by the SI$^2$R model with two demes of infectious individuals: one with extremely strong transmissibility and the other with weak transmissibility. 
-That is, $b_1(t)\ggg b_2(t)$ if Deme 1 is the set of highly infective individuals and Deme 2 is the set of ordinary infective. 
-In terms of the second definition, the involvement of polytomies and multifurcating trees adds an extra layer of complexity in the population model, along with the issues of measurement error and model robustness, such that we would like to postpone the development of a new framework to a future study.
-
-\section{Structure genealogy processes}
-
-\paragraph{Inventory process}
-First, we define a global set $\Names\coloneqq\N$, which is a collection of infinite unique names. 
-For any Deme $j$ of infective population, we define an inventory $\Inv_t^j(\omega)=\mathrm{inven}\left(\omega\vert_t,j\right)$, a list to document all the extant infected individuals in Deme $j$ at time $t$, by their names, where $\omega\vert_t$ is the sequence of jumps up to time $t$.
-On top of the global set $\Names$ and the $\mathrm{inven}$ procedure, we can then define six operators: 
-\begin{enumerate}[(1)]
-  \item $\mathrm{next}(\Names)$ returns the minimum integer in $\Names$, call it $n$, and update the global set $\Names=\Names\setminus\{n\}$;
-  \item $\mathrm{initialize}(I)\coloneqq\bigcup_{i=1}^I\{\mathrm{next}(\Names)\}$;
-  \item $\mathrm{add}(\sg{I})=\sg{I}\cup\{\mathrm{next}(\Names)\}$;
-  \item $\mathrm{drop}(\sg{I},n)=\sg{I}\setminus\{n\}$; and
-  \item $\mathrm{movein}(\sg{I},n)=\sg{I}\cup\{n\}$.
-\end{enumerate}
-
-Now the deterministic and recursive procedure $\mathrm{inven}$ is described as follows, for $\omega$ is a finite jump sequence:
-
-\begin{equation}
-  \mathrm{inven}(\omega,j)\coloneqq
-  \begin{cases}
-    \mathrm{initialize}(I^j(\sg{X}_0(\omega)), & \text{if }K(\omega)=0; \\
-    \mathrm{add}(\mathrm{inven}(\predec{\omega},j)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Bset_j;\\
-    \mathrm{drop}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\Dset_j; \\
-    \mathrm{movein}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ij}; \\
-    \mathrm{drop}(\mathrm{inven}(\predec{\omega},j),N_{K(\omega)}(\omega)), & \text{if }K(\omega)>0\text{ and }U_K(\omega)\in\bigcup_i\Mset_{ji}; \\
-    \mathrm{inven}(\predec{\omega},j), & \text{otherwise}.
-  \end{cases}
-\end{equation}
-
-Eventually, we can define $\mathrm{inven}(\omega)\coloneqq\{\mathrm{inven}(\omega,j)\}_{j=1}^J$ as a list of inventories, such that $\Inv_t(\omega)\coloneqq\{\Inv_t^j(\omega)\}_{j=1}^J$.
-
-\paragraph{Structured genealogies}
-Types of nodes are defined in a structured genealogy: \emph{Leaves} are the extant members at current time and \emph{Internal Nodes} represent ancestors to a certain subset of extant members, which we further define \emph{Sampling Nodes} and \emph{Migration Nodes}, whose differences will be clear in a moment.
-
-For the sake of distinction and visualization, we define a set of \emph{colored ball}s, $(f,n)\in\Colors\times\N$, where $n$ is the label and $f$ is the color of the ball, and $\Colors\coloneqq\{\green,\black,\blue,\red,\purple\}$.
-Therefore, we can define a genealogy \emph{node} as a quadruple $(n,t,w,z)$, where $n\in\N$ is the node's \emph{name}, $t\in\Time$ is its time, $w$ is the node's \emph{pocket} (\ie unordered pair of colored balls), and $z\in\N$ is the node's \emph{deme}.
-For a node $\Node{p}$, we will use $n(\Node{p})$, $t(\Node{p})$, $w(\Node{p})$, and $z(\Node{p})$ to denote the name, time, pocket, and deme of $\Node{p}$, respectively. 
-We then assign black, green, blue, and purple balls to serve as pointers to the extant population at time $t$, to the internal nodes, to the samples, and to the migrations between demes, respectively.
-
-Eventually, the current time $t$ and the sequence of nodes, $\Gen=\left(t,\left(\Node{p}_k\right)_{k=0}^{K-1}\right)$, explicitly define a structured genealogy $\Gen$, where $t\in\Time$, $K\in\N$, and node $\Node{p}_k$ are subjected to conditions described in \citet{King2021}.
-We then use $t(\Gen)$, $K(\Gen)$, and $\Node{p}_k(\Gen)$ to represent the time, the length (\ie the number of nodes), and the $k$-th node of the genealogy $\Gen$, respectively. 
-We define $\seq{\Gen}$ as the node sequence of genealogy $\Gen$, where $\Node{p}\in\Gen$ when $\Node{p}$ is one node in $\seq{\Gen}$.
-
-\paragraph{Effect of births, death, sampling and migration}
-Births, deaths, samples, and migrations change the topology of a structured genealogy. 
-A death at time $t$ indicates the removal of a leaf in the genealogy and the dismiss of a node. The detailed definition can be found in \citet{King2021}.
-
-A birth at time $t$ in a structured genealogy is similar to that in a unstructured one \citep{King2021}: a black ball is selected as the parent and a new node introduced as the child, who is in the same deme as the parent.
-Let $\Ball{b}$ be randomly selected black ball, named \emph{n}, which is the parent, and there exists a node $\Node{p}\in\Gen$ holding $\Ball{b}\in w(\Node{p})=\{\Ball{b},\Ball{b}'\}$ where $\Ball{b}'$ is the other ball held by $\Node{p}$.
-Here, we assume the parent is from Deme $i$ (\ie $n\in\Inv_i$) and the child is born into Deme $j$. 
-We then can produce a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},i)$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, $\Ball{b}''=(\black,c)$.
-By swapping balls between $\Node{p}$ and $\Node{p}'$, it ends up that $w(\Node{p})=\{\Ball{g},\Ball{b}'\}$ and $w(\Node{p}')=\{\Ball{b},\Ball{b}''\}$. 
-We denote the resulting sequence of nodes as $\mathrm{add}(\seq{\Gen},n)$.
-
-% A death at time $t$ is defined in the same way as that in \citet{King2021} and we continue to use the notation $\mathrm{drop}(\seq{\Gen},n)$ to denote the resulting sequence of nodes for a death, where an individual with name \emph{n} is selected randomly to be removed from the population.
-
-A death at time $t$ is quite different from that defined in \citet{King2021}, while we continue to use the notation $\mathrm{drop}(\seq{\Gen},n)$ to denote the resulting sequence of nodes for a death, where an individual with name \emph{n} is selected randomly to be removed from the current population.
-When a death occurs, a black ball $\Ball{b}$ with name \emph{n} is selected arbitrarily, whose holder is node $\Node{p}$, \ie $\Ball{b}\in w(\Node{p})=\{\Ball{b},\Ball{b}'\}$. 
-There exists an unique $\green$ ball and an unique node $\Node{p}'$ such that $\Ball{g}=(\green, n(p))$ and $\Ball{g}\in w(\Node{p}')=\{\Ball{g},\Ball{b}''\}$. 
-The $\mathrm{drop}(\seq{\Gen},n)$ operation then proceeds differently, conditioned on the color and the deme of the other ball, $\Ball{b}'$, held by $\Node{p}$: 
-\begin{inparaenum}[(1)]
-  \item if $\Ball{b}'$ is $\green$, swap $\Ball{g}$ and $\Ball{b}'$ between $\Node{p}$ and $\Node{p}'$ then remove $\Node{p}$ from the sequence of nodes;
-  \item if $\Ball{b}'$ is $\black$, the above procedure is conducted when $\Ball{b}'$ and $\Node{p}$ are in the same deme, otherwise, replace $\Ball{b}$ by a new purple ball $\Ball{c}=\{\purple, q\}$ where $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is purple}\}|$;
-  \item if $\Ball{b}'$ is $\blue$, replace $\Ball{b}$ by a new red ball with the name of $\Ball{b}'$;
-  \item if $\Ball{b}'$ is $\purple$, change the deme of $\Ball{b}$ to that of $\Node{p}$, swap $\Ball{g}$ and $\Ball{b}$, and delete $\Node{p}$.
-\end{inparaenum}
-By doing aforementioned steps, we can keep track of the sampled individuals and the migration at a birth.
-
-A sample at time $t$ results in the participation of a new node with a new blue ball. 
-The procedure is similar to the previous sampling in \citet{King2021} and the birth mentioned above: a black ball, named \emph{n}, where $n\in\Inv_j$, and held by $\Node{p}$ (\ie $w(\Node{p})=\{\Ball{b},\Ball{b}'\}$), is selected at random.
-Then a new node $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},j)$ is generated, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\blue,q)$. 
-At the end, $\Node{p}$ exchanges $\Ball{b}$ for $\Ball{g}$ with $\Node{p}'$. 
-Here, $q$ is the name of the new blue ball, which is the ordinal number of the migration, $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is blue}\}|$. 
-We use the notation $\mathrm{sample}(\seq{\Gen},n)$ as the resulting sequence of nodes.
-
-A migration at time $t$ leads to the inclusion of a new node with a new purple ball. 
-Let $\Ball{b}$ denote the
-black ball with name \emph{n} and selected to indicate an extant in the population migrates from his/her current
-deme to a new one. 
-Assuming this selected individual is from Deme $i$ (\ie $i\in\Inv_i$) and migrating to Deme $j$ ($j\neq i$), we then initialize a new node by taking $\Node{p}'=(c,t,\{\Ball{g},\Ball{b}''\},i)$, where $c=\mathrm{next}(\Names)$, $\Ball{g}=(\green,c)$, and $\Ball{b}''=(\purple,q)$.
-Note that, we update $\Inv_i=\Inv_i\setminus\{n\}$ and $\Inv_j=\Inv_j\cup\{n\}$. 
-Similar to other events, we therefore swap $\Ball{b}$ for $\Ball{g}$ between nodes $\Node{p}$ and $\Node{p}'$ and insert $\Node{p}'$ at the last position of the node-sequence. 
-Here, \emph{q} is the name of the new purple ball, which is the ordinal number of the migration,  $q=|\{\Ball{b}\in w(\Gen)\;|\;\Ball{b}\ \text{is purple}\}|$. 
-We call this new node $\Node{p}'$ the \emph{Migration Nodes} in the genealogy and denote the resulting sequence of nodes by $\mathrm{migrate}(\seq{\Gen},n)$.
-
-\paragraph{Genealogical event times}
-
-% \QYL{the definition for sets of event times is problematic.}
-
-Given a genealogy $\Gen$, the set of \emph{genealogical event times}, $\Eset(\Gen)$, is the set of all node times.
-Its subsets, including those for different event types and those for different demes, are of interest.
-In particular, we define
-\begin{equation*}
-  \begin{aligned}
-    \Eset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j\}, \\
-    \Masterset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j, w(\Node{p})\ \text{contains one green ball but no purple balls}\}, \\
-    \Cset^{j,j'}(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j, w(\Node{p})\ \text{contains two green balls, one from Deme $j$ and the other from $j'$}\}, \\
-    \Lset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j, w(\Node{p})\ \text{contains no green balls}\}, \\
-    \Tset^j(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j, w(\Node{p})\ \text{contains a blue ball}\}, \\
-    \Zset^{j,j'}(\Gen) &\coloneqq \{t(\Node{p})\;|\;\Node{p}\in\Gen, z(\Node{p})=j, w(\Node{p})\ \text{contains a purple ball and a green ball from Deme $j'$}\}, \\
-    \Qset^j(\Gen) &\coloneqq \Masterset^j(\Gen) \cap \Tset^j(\Gen).
-  \end{aligned}
-\end{equation*}
-With these definitions,
-$\Masterset^j(\Gen)$ comprises the internal node times of nodes that are derived from nodes in Deme $j$,
-$\Cset^{j,j'}(\Gen)$ is the set of coalescent times of nodes in Deme $j$ and containing a ball in Deme $j'$,
-$\Lset^j(\Gen)$ contains the leaf times where the tips are in Deme $j$,
-$\Tset^j(\Gen)$ holds the sample times for samples in Deme $j$,
-$\Zset^{j,j'}(\Gen)$ holds the deme-transition times derived from nodes in Deme $j$ and containing a ball in Deme $j'$,
-and $\Qset^j(\Gen)$ is the set of \emph{direct-descent times} derived from samples in Deme $j$.
-
-\paragraph{Structured genealogy process}
-The structured genealogy evolves over time, so we can proceed to define the \emph{structured genealogy process} $\Gen_t$ now. 
-We define $\Gen_t(\omega) = \left(t,\mathrm{geneal}(\restrict{\omega}{t})\right)$, for the end time $t\in\Time$ and the sequence of jumps $\omega\in\Omega$. 
-We first initialize the global set of names by letting $\Names=\N\setminus\{1,2,\dots,I(\X_0(\omega))\}$, then the operation $\mathrm{geneal}$ is defined recursively as follows for $\omega\in\mathring\Omega$:
-
-\fontsize{10pt}{12pt}\selectfont
-\begin{equation}\label{eq:Gdef}
-  \mathrm{geneal}(\omega)\coloneqq
-  \begin{cases}
-    \left(k,0,\{(\green,k),(\black,k)\},0\right)_{k=0}^{I(\X_0(\omega))}, &\text{if}\ K(\omega)=0;\\
-    \mathrm{add}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Bset;\\
-    \mathrm{drop}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Dset;\\
-    \mathrm{sample}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Sset;\\
-    \mathrm{migrate}(\mathrm{geneal}(\predec{\omega}),N_{K(\omega)}(\omega)), &\text{if}\ K(\omega)>0\ \text{and}\ U_{K(\omega)}\in\Mset;\\
-    \mathrm{geneal}(\predec{\omega}), &\text{otherwise};
-  \end{cases}
-\end{equation}
-\normalsize
-
-where $\Bset\coloneqq\bigcup_{i,j}\Bset_{ij}$, $\Dset\coloneqq\bigcup_j\Dset_j$, and $\Mset\coloneqq\bigcup_{i,j}\Mset_{ij}$. 
-A deterministic map, $\mathrm{ext}$, can be defined, such that $\Inv_t(\omega)=\mathrm{ext}(\Gen_t(\omega))$ for all $\omega\in\Omega$.
-
-\paragraph{Visible genealogy process}
-The sampled genealogy contains only relations among genomic samples. 
-In other words, information in a full genealogy about unsampled individuals and deme-transitions are unobserved in a sampled genealogy, which we call it a visible genealogy. 
-We divide the procedure of deriving a visible genealogy from a full genealogy into two separate steps: \emph{pruning} and \emph{obscuring}. 
-The former drops all unsampled individuals (\ie black balls) in a genealogy and remains the relationship between samples and deme-transition records of samples, and the latter further erases all deme-transition records of samples and obtains an obscured genealogy.
-
-Given a genealogy $\Gen$, we define $\Inv=\mathrm{ext}(\Gen)$.
-%% Then. by definition, for each $c\in\Inv$, $(\black,c)\in w(\Gen)$.
-Let $\mathrm{prune}(\Gen)$ be the pruned genealogy derived by applying iterative $\mathrm{drop}$ operation to $\Gen$ for each name $c\in\Inv$.
-Specifically, suppose $\Inv^j=\{n_0^j,\dots,n_{m_j-1}^j\}$ is the inventory for Deme $j$ and $\Inv=\{\Inv^j\}_{j=1}^{J}$, where $J$ is the total number of demes. 
-Then we can define the set of all names in $\Inv$ as $\mathcal{N}\coloneqq\bigcup_{j=1}^J \Inv^j=\{c_0,\dots,c_{m-1}\}$, where $m=\sum_j m_j$.
-Let $\Node{P}_0=\seq\Gen$ and $\Node{P}_{k}=\mathrm{drop}(\Node{P}_{k-1},c_{k-1})$, for $k=1,\dots,m$.
-Then $\mathrm{prune}(\Gen)\coloneqq(t(\Gen),\Node{P}_{m})$.
-\Cref{fig:prune} illustrates the pruning procedure.
-
-For $\omega\in\Omega$, we define the \emph{pruned genealogy process},
-\begin{equation}
-  \Prune_t(\omega)\coloneqq\textrm{prune}(\Gen_t(\omega)).
-\end{equation}
-Note that $\Prune_t$ so defined is a itself a genealogy.
-\Cref{fig:prune}B depicts one example of $\Prune_t$ both graphically (as a tree) and diagrammatically (as a node-sequence).
-
-Before we proceed to produce the visible genealogy, we first define an operation $\mathrm{reverse}$ on the sequence of nodes of a genealogy as follows: 
-\begin{inparaenum}[(1)]
-  \item a random purple ball $\Ball{b}$ with name $q$ is selected such that there exists a node $\Node{p}$, where $\Ball{b}\in w(\Node{p})=\{\Ball{b}, \Ball{b}'\}$ and $\Ball{b}'$ is the other ball held by $\Node{p}$;
-  \item there also exists a node $\Node{p}'$ such that $\Ball{g}=\left(\green,n(\Node{p})\right)\in w(\Node{p}')=\{\Ball{g},\Ball{b}''\}$ where $\Ball{b}''$ is the other ball held by $\Node{p}'$;
-  \item swapping balls between $\Node{p}$ and $\Node{p}'$ results in $w(\Node{p}')=\{\Ball{b}',\Ball{b}''\}$ and $w(\Node{p})=\{\Ball{g},\Ball{b}\}$;
-  \item we delete $\Node{p}$ from the node sequence.
-\end{inparaenum}
-The resulting sequence of nodes is denoted as $\mathrm{reverse}(\seq{\Gen},q)$.
-
-Suppose for a pruned genealogy $\Gen'$, the set of the serial number of the purple balls is $\{0,1,\dots,q\}$ and the list of inventories is $\Inv=\{\Inv^j\}_{j=1}^J$.
-$\mathrm{obs}(\Gen')$ applies the following steps on $\Gen'$:
-\begin{inparaenum}[(a)]
-  \item let $\Node{P}_0=\seq{\Gen'}$ and $\Node{P}_{k}=\mathrm{reverse}(\Node{P}_{k-1},k)$, for $k=1,\dots,q$; and
-  \item for $\Node{P}_q$, we update $\Inv=\bigcup_{j}\Inv^j$ and for each node $\Node{p}\in\seq{\Gen}$, $z(\Node{p})=0$.
-\end{inparaenum}
-
-For $\omega\in\Omega$, we define the \emph{visible genealogy process},
-\begin{equation}
-  \Vis_t(\omega)\coloneqq\textrm{obs}(\textrm{prune}(\Gen_t(\omega))).
-\end{equation}
-Note that $\Vis_t$ so defined is a itself a genealogy.
-\Cref{fig:obs}B depicts one example of $\Vis_t$ both graphically (as a tree) and diagrammatically (as a node-sequence).
-
-\paragraph{Node color}
-In a visible genealogy, we can define four kinds of nodes based on the colors of balls in the pockets:
-\begin{inparaenum}[(a)]
-  \item \emph{green nodes}, which have more than one green balls in their pockets;
-  \item \emph{blue nodes}, which have one blue;
-  \item \emph{red nodes}, which have one red;
-  \item \emph{purple nodes}, which have one purple ball.
-\end{inparaenum}
-In a \emph{compact} visible genealogy, green, blue, red, and purple nodes correspond to \emph{coalescent points}, leaves, direct-descent events, and deme-transition events, respectively.
-Thus, for a visible genealogy $\Vis$ and any deme $j$:
-
-\paragraph{Branch linetype}
-
-\paragraph{Deme proportion and lineage count}
-
-We can further define the deme proportion function $\demefuc(t,\omega)\coloneqq\{\demefuc^j(t,\omega)\}_{j=1}^J$, where $\demefuc^j(t,\omega)\coloneqq\frac{\vert\Inv_t^j(\omega)\vert}{\vert\Inv_t(\omega)\vert}$.
-
-Given a pruned genealogy, $\Prune$, at each time $t\in\Time$, there are a finite number, $\ell^{j}(t, \Prune)$, of lineages in Deme $j$ present in the genealogy at that time.  Evidently, one has
-
-\begin{align}
-\begin{split}
-  \ell^j(t,\Prune) &\coloneqq \sum_{j'}\sum_{e\in\Cset^{j',j}(\Prune)} \indicator{\halfopen{e,\infty}}(t)\;-\;\sum_{e\in\Lset^j(\Prune)} \indicator{\halfopen{e,\infty}}(t) \\
-  +&\;\sum_{j'}\left(\sum_{e\in\Zset^{j',j}(\Prune)}\indicator{\halfopen{e,\infty}}(t)\;-\;\sum_{e\in\Zset^{j,j'}(\Prune)}\indicator{\halfopen{e,\infty}}(t)\right).
-\end{split}
-\end{align}
-
-\paragraph{Embedded chains}
-Due to discrete sample times, defined as $S_i$, we now proceed to consider the embedded chain for pruned genealogy process $\Prune_t$ and then for visible genealogy process $\Vis_t$. 
-For $\omega\in\Omega$, let $\Och_i(\omega)\coloneqq\Prune_{S_i(\omega)}(\omega)$ be the embedded chain of the pruned genealogy process, and $\Wch_i(\omega)\coloneqq\Vis_{S_i(\omega)}(\omega)$ be that of the visible genealogy process.
-At each sample time $S_i$, one extra lineage (\ie the $i$-th sample) is to be attached to the previous existing genealogy (\ie $\Och_i$ or $\Wch_i$) at a random attachment time, denoted by $A_i$. 
-Both $\Och_i$ and $\Wch_i$ are Markovian.
-
-\begin{lemma}\label{lemma:ell}
-  Define time intervals $\Delta_{i}^{j}$ and $\Delta_{i}^{j,j'}$ as follows:
-  \begin{equation*}
-    \Delta_{i}^{j}\coloneqq\begin{cases}
-      \halfopen{a_i^j,s_i^j}, & \Tset^j(\Och_i)\setminus\Tset^j(\Och_{i-1})\neq\emptyset, \\
-      \emptyset, &\textrm{otherwise},
-    \end{cases}
-    \quad\textrm{and}\quad
-    \Delta_{i}^{j,j'}\coloneqq\begin{cases}
-      \halfopen{z_i^{j,j'},s_i^j}, & \Zset^{j,j'}(\Och_i)\setminus\Zset^{j,j'}(\Och_{i-1})\neq\emptyset, \\
-      \emptyset, &\textrm{otherwise},
-    \end{cases}
-  \end{equation*}
-  where $s_i^j=\Tset^j(\Och_i)\setminus\Tset^j(\Och_{i-1})$, $a_i^j=\Masterset^j(\Och_i)\setminus\Masterset^j(\Och_{i-1})$, and $z_i^{j,j'}=\Zset^{j,j'}(\Och_i)\setminus\Zset^{j,j'}(\Och_{i-1})$ are the corresponding sample, attachment times, respectively, of the $i$-th sample lineage in Deme $j$, and the deme-transition times from Deme $j$ to $j'$, respectively.
-  Then, if it is understood that $\ell^j(t,\Och_0)=0$ for all $j$,
-  \begin{equation*}
-    \ell^j(t,\Och_i)=\ell^j(t,\Och_{i-1})+\indicator{\Delta_{i}^{j}}(t)-\sum_{j'}\left(\indicator{\Delta_{i}^{j,j'}}(t)-\indicator{\Delta_{i}^{j',j}}(t)\right), \qquad \forall i>0.
-  \end{equation*}
-  Trivially, $\ell(t,\Och_i)=\sum_j\ell^j(t,\Och_i)$.
-\end{lemma}
-
-\section{Results}
-\label{sec:results}
-The derivation of the probability density of $\Wch_i$ given $\Hist_{S_i}$ can be divided into two steps:
-\begin{align}
-\begin{split}
-  P_{\Wch_i|\Hist}&=\sum_{\Och_i}\,P_{\Wch_i|\Och_i}P_{\Och_i|\Hist} \\
-                  &=\sum_{\Och_i}\,P_{\Wch_i|\Och_i}\left(P_{\Och_1|\Hist}\prod_{q=2}^{i}P_{\Och_q|\Och_{q-1}}\right) 
-\end{split}
-\end{align}
-% Trivially, $P_{\Och_1|\Hist}=1$ and $P_{\Wch_1|\Och_1}=1$ such that $P_{\Wch_1|\Hist}=1$.
-
-We first compute $P_{\Och_q|\Och_{q-1},\Hist}$ by fixing $A_q=a_q$, $\Och_q=o_q$, $S_q=s_q$, and $\Hist_{s_q}=h=\left(s_q,\left(t_k,u_k\right)_{k=0}^K\right)$.
-For convenience, we write $x_k=\X(t_k)$, and define the sets
-$\Omega_q\coloneqq\{\omega\in\Omega\;|\;\Wch_{q}(\omega)=w_q,\Hist_{s_i}(\omega)=h\}\subset\Omega$ and
-$\Omega_{qk}\coloneqq N_k(\Omega_q)\subset\N$, for $q\le i$ and $k\le K$.
-
-\begin{equation*}
-  \begin{aligned}
-    % P_{\Och_q|\Och_{q-1},\Hist}(o_q|o_{q-1},h)&=\prod_{k=1}^K\sum_{n_k\in\Omega_{qk}}\auxmeas_{u_k,x_k-u_k}(n_k)=\prod_{k=1}^K\sum_{n_k\in\Omega_{qk}}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv^j_{t_{k-1}}(\Omega_q)}(n_k)}{I_j(x_k-u_k)}\right).
-    P_{\Och_q|\Och_{q-1},\Hist}(o_q|o_{q-1},h) &= \prod_{k=1}^K\sum_{\omega\in\Omega_q}\auxmeas_{u_k,x_k-u_k}(N_k(\omega)) = \prod_{k=1}^K\sum_{\omega\in\Omega_q}\left(\sum_{j=1}^J\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}\right),
-  \end{aligned}
-\end{equation*}
-
-due to that $\auxmeas$ is uniform within any certain deme.
-
-We define $\phi_{qk}^{j}\coloneqq\sum_{\omega\in\Omega_q}\dfrac{\indicator{\Inv_{t_{k-1}}^j(\omega)}(N_k(\omega))}{I^j(x_k-u_k)}$, 
-and let $I_k^j=I^j(x_k)$, $\ell_{qk}^{j}=\ell^j(t_k, \Och_{q-1})$.
-Note that $\ell^j_{1k}=0$ for all $k$ and $j$.
-
-All cases can be categorized into four groups based on the occurrence of population event within time $\halfopen{a_q,s_q}$: no population event or simultaneous events, a birth from Deme $j$ to $j'$, a sample of Deme $j'$, and a migration from Deme $j$ to $j'$ where $j\neq j'$.
-Within each group, we investigate $\phi_{qk}^{j''}$ by considering whether the corresponding population event was presented in the pruned genealogy and involving the $q$-th lineage sample at time $t_k$, given that the $q$-th lineage sample was in Deme $j''$ at time $t_k$.
-For $q>1$, we then have:
-
-% 1) No. of total events in the population
-% 2) No. of conditioned event in the population
-% 3) possibilities that presented in the genealogy
-% 4) probability of the exact sample involved
-
-\begin{enumerate}[(a)]
-  \item There was no population event within $\halfopen{a_q,s_q}$ or more than one population events occurring that involved the $q$-th sample:
-    \begin{enumerate}
-      \item if $t_k\notin\halfopen{a_q,s_q}$, or if $u_k\notin\Bset\cup\Sset\cup\Mset$, $\phi_{qk}^{j''}=1$;
-      \item if $t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1})$, $\phi_{qk}^{j''}=1$.
-    \end{enumerate}
-
-  % M
-  \item There was a migration event, that is, $u_k\in\Mset^{j,j'}$ within $\halfopen{a_q,s_q}$, given that it was not presented in the pre-existing genealogy $\Och_{q-1}$, \ie $t_k\notin\Zset^{j,j'}(\Och_{q-1})$:
-    \begin{enumerate}
-      \item if this migration event had been presented in previous genealogy $\Och_{q-1}$, \ie $t_k\in\Zset^{j,j'}(\Och_{q-1})$, since no simultaneous migration events occurred at the same time and $t_k\in\Zset^{j,j'}{\Och_q}$ for sure, $\phi_{qk}^{j''}=1$;
-      \item if this migration event exactly involved the $q$-th sample and was presented as a deme-transition event in $\Och_q$, \ie $t_k\in\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_{k}^{j}-\ell_{q,k}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
-      \item otherwise, $t_k\notin\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left((I_{k}^{j}-\ell_{q,k}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
-    \end{enumerate}
-
-  % G
-  \item There was a sample event, that is $u_k\in\Sset^{j'}$ within $\halfopen{a_q,s_q}$:
-    \begin{enumerate}
-      \item if the lineage sampled at this event served as a direct ancestor to the $q$-th sample lineage, that is, $t_k=a_q\in\Lset^{j'}(\Och_{q-1})$, this indicates that among $I_k^{j'}-\ell_{qk}^{j'}$ unsampled individuals in deme $j'$ at $t_k$, the $q$-th sample lineage is the exact direct descent, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j'}-\ell_{qk}^{j'}\right)$;
-      \item otherwise, $t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})$, and $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left(I_k^{j'}-\ell_{qk}^{j'}\right)$.
-    \end{enumerate}
-
-  % B, j=j'
-  \item There was a birth event within the same deme, that is $u_k\in\Bset^{j,j'}$ and $\delta_{j,j'}=1$, but it was not shown in the genealogy yet because only one of the pre-existing $\ell_{qk}^{j'}$ lineages involved, \ie $t_k\notin\Cset^{j,j'}(\Och_{q-1})$:
-    \begin{enumerate}
-      % there were in total \binom{I}{2} combinations presented as births, while \binom{l}{2} is excluded because the birth was not presented in genealogy.
-      \item if this birth event was presented as a coalescent event in $\Och_q$, \ie $t_k\in\Cset^{j,j'}(\Och_q)$, with probability $\ell_{qk}^{j'}/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$, then the probability that the $q$-th sample attached to \emph{a certain} sample lineage in the genealogy is $1/\ell_{qk}^{j'}$, implying that the $q$-th sample lineage was exactly the other individual involved in the birth event, and eventually, $\phi_{qk}^{j''}=\left(\delta_{j,j'}\delta_{j',j''}\right)/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$;
-      \item otherwise, $t_k\notin\halfopen{a_q,s_q}$,  $\phi_{qk}^{j''}=1-\left(\delta_{j,j'}\delta_{j',j''}\ell_{qk}^{j'}\right)/\left(\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}\right)$.
-    \end{enumerate}
-
-  % B, j!=j'
-  \item There was a birth event between two different demes, that is $u_k\in\Bset^{j,j'}$ and $\delta_{j,j'}=0$:
-    \begin{enumerate}
-      % There were  l'*(I-l) possibilities of transitions
-      \item and given that it was presented as a deme-transition event in the pre-existing genealogy, that is, $t_k\in\Zset^{j,j'}(\Och_{q-1})$:
-        \begin{enumerate}
-          % the genealogy has the exact l' lineage as the child, and the sample is the exact 1 as the parent
-          % 1/[l'*(I-l)] * l' * 1/1
-          \item if the $q$-th sample was from deme $j$ and is exactly the parent in the birth, that is, $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j,j''}=1$, then $\phi_{qk}^{j''}=\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
-          \item otherwise, $t_k\in\Zset^{j,j'}(\Och_q)$,and $\phi_{qk}^{j''}=1-\delta_{j,j''}/\left(I_k^{j}-\ell_{qk}^{j}\right)$;
-        \end{enumerate}
-      % there were in total I*I' in the birth, while l*l' is excluded since it was not presented as coalescent, and (I-l)*l' are excluded which as presented as transition.
-== I I' - l l' - (I-l) l' = I (I'-l')
-      \item or given that it was not presented in $\Och_{q-1}$, \ie $t_k\notin\Zset^{j,j'}(\Och_{q-1})\cup\Cset^{j,j'}(\Och_{q-1})$:
-        \begin{enumerate}
-          % the genealogy has l lineages as parent, and we choose one of them 1/l: 1/[I (I'-l')] * l * 1/l
-          \item if the $q$-th sample was from deme $j'$ and is coalesced at $t_k$, \ie $t_k\in\Cset^{j,j'}(\Och_q)$ and $\delta_{j',j''}$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
-          % 
-          \item if the $q$-th sample was from deme $j'$ and is presented as a deme-transition event in $\Och_q$, \ie $t_k\in\Zset^{j,j'}(\Och_q)$, then $\phi_{qk}^{j''}=\delta_{j',j''}/\left(I_k^{j}(I_{k}^{j'}-\ell_{qk}^{j'})\right)$;
-
-          \item otherwise, $t_k\notin\Cset^{j,j'}(\Och_q)\cup\Zset^{j,j'}(\Och_q)$, and $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left(I_{k}^{j'}-\ell_{qk}^{j'}\right)$;
-        \end{enumerate}
-    \end{enumerate}
-\end{enumerate}
-
-To summarize, we have
-\begin{equation}\label{eq:qcases}
-  \phi_{qk}^{j''}=
-  \begin{cases}
-    1, &\text{if}\ t_k\notin\halfopen{a_q,s_q}\ \text{or}\ u_k\notin\Bset\cup\Sset\cup\Mset,\\
-    1, &\text{if}\ t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1}),\\[12pt]
-
-    % M
-    1, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_{q}),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
-    \frac{\delta_{j',j''}}{(I_{k}^j-\ell_{q,k}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
-    1-\frac{\delta_{j',j''}}{(I_{k}^j-\ell_{q,k}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
-
-    % G
-    \frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k=a_q\in\Lset^{j'}(\Och_{q-1})\ \text{and}\ u_k\in\Sset^{j'},\\[12pt]
-    1-\frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})\ \text{and}\ u_k\in\Sset^{j'},\\[12pt]
-
-    % B, j=j'
-    \frac{\delta_{j,j'}\delta_{j',j''}}{\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j=j',\\[12pt]
-    1-\frac{\delta_{j,j'}\delta_{j',j''}\ell_{qk}^{j'}}{\binom{I_k^{j'}}{2}-\binom{\ell_{qk}^{j'}}{2}}, &\text{if}\ t_k\in(a_q,s_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j=j',\\[12pt]
-
-    % B, j!=j'
-    % in Z
-    \frac{\delta_{j,j''}}{I_k^{j}-\ell_{qk}^{j}}, &\text{if}\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
-    1-\frac{\delta_{j,j''}}{I_k^{j}-\ell_{qk}^{j}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
-    % not in Z
-    \frac{\delta_{j',j''}}{I_k^j(I_{k}^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_{q-1}),\ t_k\in\Zset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
-    \frac{\delta_{j',j''}}{I_k^j(I_{k}^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\notin\Zset^{j,j'}(\Och_{q-1}),\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
-    1-\frac{\delta_{j',j''}}{I_{k}^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_{q-1}),\ u_k\in\Bset^{j,j'},\ \text{and}\ j\neq j',\\[12pt]
-
-    0, &\text{otherwise}.
-  \end{cases}
-\end{equation}
-
-% To summarize, we have
-% \begin{equation}\label{eq:qcases}
-%   \phi_{qk}^{j''}=
-%   \begin{cases}
-%     1, &\text{if}\ t_k\notin\halfopen{a_q,s_q}\ \text{or}\ u_k\notin\Bset\cup\Sset\cup\Mset,\\
-%     
-%     % deme-transition time can evolve to coalescent time, when the new sample is the parent of a "cross-deme" birth
-%     1, &\text{if}\ t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1}),\\[12pt]
-%     \frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
-%     1-\frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
-%     
-%     % sample
-%     \frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k=a_q\in\Lset^{j'}(\Och_{q-1}),\\[12pt]
-%     1-\frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})\ \text{and}\ u_k\in\Sset^{j'},\\[12pt]
-%     \frac{\delta_{j',j''}}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1}),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
-%     
-%     % B2C: either j''=j (in Z_{q-1} or j''=j' (not in Z_{q-1})
-%     % the numerator = 1 when j'!=j and =2 when j'=j.
-%     \frac{\delta_{j,j''}+\delta_{j',j''}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\\[12pt]
-%     1-\frac{(\delta_{j,j''}+\delta_{j',j''})\ \ell_{qk}^{j}\ \delta_{j',j''}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}-\frac{(1-\delta_{j,j'})\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_{q-1})\cup\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
-%     1-\frac{\delta_{j,j''}}{I_k^{j}I_k^{j'}-\ell_{qk}^j\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
-%     0, &\text{otherwise}.
-%   \end{cases}
-% \end{equation}
-
-We then can establish our first main result:
-
-\begin{thm}\label{thm:prunehthm1}
-  With the definitions as above, $P_{\Och_i|\Hist}(o|h) = \prod_{q=1}^{i}\prod_{k=1}^{K}\left(\sum_{j''=1}^{J}\phi_{qk}^{j''}\right)=\prod_{q=1}^{i}\prod_{k=1}^{K}\phi_{qk}^{j''_{qk}}$, where $j''_{qk}$ is the deme of the $q$-th sample lineage at time $t_k$.
-\end{thm}
-We now exchange the order of multiplication over $q$ and $k$ then we can derive our \Cref{thm:prunethm2}.
-
-\begin{thm}\label{thm:prunethm2}
-  \begin{equation*}
-    P_{\Prune_t|\Hist_t}(\Prune_t|h)=.
-  \end{equation*}
-\end{thm}
-\begin{proof}
-  Let $\Prune_t=\Och_i$ for some $i$.
-We define $\Phi_{k}\coloneqq\prod_{q=1}^{i}\phi_{qk}^{j''_{qk}}$ for each $k$ and as above, $I_k^{j}=I_j(x_k)$, $\ell_{qk}^{j}=\ell^j(t,\Och_{q-1})$.
-For the first two cases in \Cref{eq:qcases}, $\Phi_k=1$.
-
-  Consider $u_k\in\Mset^{j,j'}$.
-Let $q^*=\{q\;\vert\;t_k=\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1})\}$ and $\ell^*=\ell_{q^*k}^{j''_{q^*k}}$.
-We then define $Q_k=\{q\;\vert\;j''_{qk}=j'\ \text{and}\ q<q^*\}$ as an ordered set and denote $[Q_k]_m$ as the $m$-th element in $Q_k$, where, $m=1,\dots, \ell^*$.
-  \begin{enumerate}[(a)]
-    \item If $q<q^*$, then $t_k\notin\Zset^{j,j'}(\Och_q)$. We will have
-      \begin{equation*}
-        \Phi_k = \prod_{q\in Q_k} \left(1-\frac{1}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}\right).
-      \end{equation*}
-      Note that, $\{\ell_{qk}^{j'}\;\vert\;q\in Q_k\}=\{0,1,\dots,\ell^*-1\}$ then the previous equation becomes
-      \begin{equation*}
-        \Phi_k = \prod_{m=1}^{\ell*} \left(1-\frac{1}{\mathcal{Y}_m (I_k^{j'}-m+1)}\right) = \prod_{m=1}^{\ell*} \frac{I_k^{j'}-m+1-(\mathcal{Y}_m)^{-1}}{I_k^{j'}-m+1},
-      \end{equation*}
-      where $\mathcal{Y}_m\coloneqq I_k^{j}-\ell_{[Q_k]_m k}^{j}$.
-  \end{enumerate}
-
-\end{proof}
+%% Nonlinear filtering equation:
+%% \begin{equation}
+%%   \begin{aligned}
+%%     \frac{\partial{w}}{\partial{t}}(t,x,y,z)=&B(t)\,\left[\sum_{y'z'}w\big(t,x-(1,0,0,0),y',z'\big)-w\big(t,x,y,z\big)\right]\\
+%%     &\qquad+\sum_{k\in\{\mathrm{S},\mathrm{E},\mathrm{I},\mathrm{R}\}}\mu\,(x^k+1)\,w(t,x+\delta_{k})-\mu\,x^k\,w(t,x)\\
+%%     &\qquad+\upsilon\,x^{\mathrm{R}}\,w(t,x-(1,0,0,-1))\\
+%%   \end{aligned}
+%% \end{equation}
 

### 2023-06-15 ms.R

@@ -1,0 +1,12 @@
+## ----packages,include=FALSE,cache=FALSE---------------------------------------
+library(tidyverse)
+library(ggtree)
+library(pomp)
+library(cowplot)
+library(phylopomp)
+stopifnot(getRversion() >= "4.3")
+stopifnot(packageVersion("pomp")>="5.1")
+stopifnot(packageVersion("phylopomp")>="0.9.2")
+theme_set(theme_bw(base_family="serif"))
+set.seed(1159254136)
+

### 2023-07-04 ms.Rnw
some progress...

@@
 We then integrate out the history to obtain nonlinear filtering equations, the solution of which yields the likelihood.
-These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likellihood.
+These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likelihood.
 We demonstrate with several examples.
 
-\section{Population processes}
+In the following, we show a Markov population process of the kind that is a staple in epidemiology induces a Markov process on the space of genealogies.
+We then show how one can comput the likelihood of a given genealogy.
+
+\section{From population processes to genealogy processes}
+
+\subsection{Population processes}
 
@@
 Migration, superspreading, competition between strains.
-
-Figure: example compartmental model with multiple demes.
 
@@
     Demes are shaded.
+    \AAK{[Perhaps another one or two examples here?]}
+    \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
     \label{fig:examples}
@@
 
+Another perspective on the Markov processes is to be had from its Markov state transition diagram (\cref{fig:markov_state}).
+
 \begin{figure}
   \begin{center}
-    \resizebox{0.9\linewidth}{!}{\input{figs/markov_state}}
+    \resizebox{0.9\linewidth}{!}{
+      \begin{tikzpicture}[scale=1]
+        \usetikzlibrary{shapes,arrows,positioning}
+        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
+        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\large, fill=white, thick, minimum size=5em]
+        \tikzstyle{trans}=[color=black, thick, >=stealth]
+        \node[state] (s0) at (0,0) {$x=(S,E,I,R)$};
+        \node[state] (s1) at (165:7.8) {$x'=(S-1,E+1,I,R)$};
+        \node[state] (s2) at (195:7.1) {$x'=(S,E-1,I+1,R+1)$};
+        \node[state] (s3) at (295:4) {$x'=(S,E,I-1,R+1)$};
+        \node[state] (s4) at (355:9) {$x'=(S+1,E,I,R-1)$};
+        \node[state] (s5) at (18:6) {$x'=(S,E,I,R)$};
+        \draw[trans,->] (s0) -- (s1) node[midway,above,sloped] {$\mathrm{Trans}$};
+        \draw[trans,->] (s0) -- (s2) node[midway,above,sloped] {$\mathrm{Prog}$};
+        \draw[trans,->] (s0) -- (s3) node[midway,above,sloped] {$\mathrm{Recov}$};
+        \draw[trans,->] (s0) -- (s4) node[midway,above,sloped] {$\mathrm{Wane}$};
+        \draw[trans,->] (s0) -- (s5) node[midway,above,sloped] {$\mathrm{Sample}$};
+        \node[font=\Large] (U) at ($(s0)+(-6,-5)$) {$\Jumps=\{\mathrm{Trans},\mathrm{Prog},\mathrm{Recov},\mathrm{Wane},\mathrm{Sample}\}$};
+      \end{tikzpicture}
+    }
   \end{center}
@@
     The state is characterized by four numbers, $S$, $E$, $I$, and $R$.
-    From a given state $x$, there are four possible jumps $x\mapsto{x'}$.
+    From a given state $x$, there are five possible kinds of events $x\mapsto{x'}$ as shown.
     From the point of view of the induced genealogy process,
-    transmission is of birth type,
-    progression is of migration type,
-    and recovery is of death type, while loss of immunity is of neutral type.
-    In this example, when a sampling event occurs, the state does not change.
+    $\mathrm{Trans}$ (transmission) is of birth type,
+    $\mathrm{Prog}$ (progression) is of migration type,
+    and $\mathrm{Recov}$ (recovery) is of death type,
+    while $\mathrm{Wane}$ (loss or waning of immunity) is of neutral type.
+    Note that, in this formulation, when a $\mathrm{Sample}$ (sampling) event occurs, the state does not change.
     \label{fig:markov_state}
@@
 
-\paragraph{Mathematical formalism}
+\paragraph{Mathematical notation}
 
-Definition of state space.
-Jumps and jump rates.
-Demes.
+Denote the underlying probability space by $(\Omega,\Borel,\mathbb{P})$.
+We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\coloneq\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$ and taking values in some space $\Xspace$.
+In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
+The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
+In particular, we suppose that
+\begin{equation}\label{eq:ic}
+  \Prob{\X_0=x}=p_0(x),
+\end{equation}
+for some choice of initial distribution, $p_0$.
+For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
+More precisely, the transition rates have the following properties:
+\begin{equation*}
+  \begin{gathered}
+    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
+  \end{gathered}
+\end{equation*}
+for all $t\in\Rp$ and $x,x'\in\Xspace$.
+Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
+Let $\N_t$ be the number of jumps that $\X$ has taken by time $t$.
+We assume that $\N_t$ is a simple counting process so that
+\begin{equation*}
+  \begin{gathered}
+    \Prob{\N_{t+\Delta}=n+1\;\vert\;\N_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),
+    \qquad
+    \Prob{\N_{t+\Delta}>n+1\;\vert\;\N_{t}=n}=o(\Delta),\\
+    \Prob{\X_{t+\Delta}\in\mathcal{E}\;\vert\;\X_{t}=x, \N_{t+\Delta}-\N_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}+\frac{o(\Delta)}{\Delta}.
+  \end{gathered}
+\end{equation*}
+\AAK{[Do we need the last term?]}
+We will further assume that $\X_t$ is non-explosive, \ie, that $\Prob{\N_t<\infty}=1$ for all $t$.
+\AAK{[Is this equivalent to non-explosivity? Or merely an implication?]}
 
-Kolmogorov equations.
+The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
+%\begin{equation}\label[pluralequation]{eq:kfe}
+\begin{equation}\label{eq:kfe}
+  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
+\end{equation}
+and if, moreover, $w(0,x) = p_0(x)$,
+then $\int_{\mathcal{E}}{w(t,x)\,\dd{x}}=\Prob{\X_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
+\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\X_t$.
+
+Without loss of generality, one can assume, as we do here, that the sample paths $t\mapsto\X_t(\omega)$ for $\omega\in\Omega$ are right-continuous with left limits (\cadlag).
+In fact, all of the processes we will describe in this paper will be taken to be \cadlag, and we will frequently need to refer to the left-limit.
+Accordingly, if $z=z(\omega,t)$ is any \cadlag\ random function, we define
+\begin{equation*}
+  \zt(\omega,t)\coloneq
+  \begin{cases}
+    \lim_{t'\uparrow{t}}z(\omega,t'), & t>0,\\
+    z(\omega,0), & t=0.\\
+  \end{cases}
+\end{equation*}
+Note that $\zt$ is thus left-continuous with right limits.
+
+\paragraph{Structured populations, demes}
+
+In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
+\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
+Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
+In particular, we suppose that there are a countable set of subpopulations that differ in their vital rates, but within each of which, individual lineages are statistically identical.
+We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
+
+For any $i\in\Demes$, we let $n_i(\X_t)$ denote the number of lineages present in deme $i$ at time $t$, \ie the \emph{occupancy} of deme $i$.
+Thus $n(\X_t)\in\Zp^{\Demes}$ is the vector of deme occupancies.
+
+\paragraph{Jump marks}
+
+In the following, it will be useful to break the jumps into distinct categories.
+For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
+\begin{equation*}
+  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
+\end{equation*}
+In \cref{fig:markov_state}, we use the marks to distinguish biologically distinct events.
+Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.
+
+Let us define the \emph{jump mark} process, $\U_t$, to be the mark of the latest jump as of time $t$.
+As usual, we take the sample paths, $t\mapsto{\U_t(\omega)}$ for $\omega\in\Omega$, to be \cadlag.
+Observe that $\U_t$ is a random, but not a Markov, process.
 
@@
 
-\section{History and genealogy processes}
-
-\subsection{Mathematical preliminaries}
-
-State space.
-Definition of $\Omega$ as space of \cadlag\ functions of time.
-Define everything in terms of \cadlag\ functions rather than sequences.
-Then the topology comes directly from the Skorokhod metric.
-Explain we use a base measure coming from unit-rate Poisson processes.
-
-Restrictions.
-Metric, Borel $\sigma$-algebra, base measure.
-
-Probability density.
-Master process.
-
-\paragraph{Markov jump processes}
-
-We will assume that our population process is a non-explosive Markov jump process $\X_t$ parameterized by time $t\in\Rp$, and taking values in some space $\Xspace$.
-In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but we dispense with that restriction here.
-The population process is completely specified by its initial-state distribution and its generator.
-In particular, we suppose that
-\begin{equation}\label{eq:ic}
-  \prob{\X_0=x}=p_0(x),
-\end{equation}
-for some choice of initial distribution, $p_0$.
-We suppose there are countably many types of jumps and denote the set of jump types by $\Jumps$.
-For each $u\in\Jumps$, there is a corresponding jump rate or hazard $\alpha_u$.
-We impose the further requirement that $\sum_{ux'}\alpha_u(t,x,x')<\infty$ for all $t,x$.
-We will always use a version of $\X_t$ in which the sample paths are right-continuous with left limits (\cadlag).
-
-The above may be compactly summarized by means of the corresponding Kolmogorov equations, as follows.
-Given $f\in L^{\infty}(\Xspace)$, let $F(s,x)\coloneq\expect{f(\X_t)\;|\;\X_s=x}$.
-It then follows that, for $s<t$ and $x\in\Xspace$,
-\begin{equation}\label{eq:gen}
-  \begin{gathered}
-    \frac{\partial{F}}{\partial{s}}(s,x)=-\int\alpha(s,x,x')\,\left[F(s,x')-F(s,x)\right]\,\dd{x'},
-    \qquad
-    F(t,x)=f(x).
-  \end{gathered}
-\end{equation}
-\Cref{eq:gen} is the so-called Kolmogorov backward equation.
-Its adjoint is the Kolmogorov forward equation (KFE, also called the \emph{master equation} in this context), which states that for $t\ge{0}$, $x\in\Xspace$,
-\begin{equation}\label[pluralequation]{eq:kfe}
-  \begin{gathered}
-    \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
-    \qquad
-    w(0,x) = p_0(x).
-  \end{gathered}
-\end{equation}
-If $w(t,x)$ satisfies \cref{eq:kfe}, then $w(t,x)=\prob{\X_t=x}$.
-
-\paragraph{Definitions}
-
-Our goal in this paper is to introduce a family of Markov processes induced by the jump process just described (\cref{fig:constellation}).
-While population processes of the kind described above can be constructed in a variety of time-honored ways \citep[\eg][]{Andersen1993,Kallenberg1997}, these classical approaches are not entirely sufficient for the tree-valued Markov processes we will erect on top of the population process.
-We therefore explicitly construct the probability space that underlies the stochastic processes we subsequently describe.
-Unavoidably, this leads to technicalities that are necessary for the firm establishment of the properties of these processes but that may distract from the overarching goals.
-Readers willing to stipulate these properties may skim the remainder of this section, in which we formally construct the probability space and make some definitions that will be needed in the sequel.
-
-Let us define a \emph{jump} to be an ordered quadruple $(t,u,x,n)\in\Rp\times\Jumps\times\Xspace\times\Zp$.
-We refer to $t$ as the \emph{time} of the jump;
-$u$ is the \emph{type} of the jump;
-$x$ is the state achieved at the end of the jump;
-$n$ is an \emph{auxiliary number} whose use will be made clear below.
-A \emph{jump sequence} is a countable sequence of jumps at increasing times.
-That is,
-\begin{equation*}
-  \omega=\left(t_k,u_k,x_k,n_k\right)_{k=0}^{K}
-\end{equation*}
-is a jump sequence if and only if $K\in\Zp\cup\{\infty\}$, $u_k\in\Jumps$, $x_k\in\Xspace$, $n_k\in\Zp$ for all $k$, and $0=t_0<t_1<t_2<\dots$.
-We will take our sample space, $\Omega$, to be the set of all jump sequences.
-For $\omega\in\Omega$ as above, we write
-\begin{equation}\label{eq:components1}
-  \begin{gathered}
-    T_k(\omega)\coloneq t_k, \qquad
-    U_k(\omega)\coloneq u_k, \qquad
-    X_k(\omega)\coloneq x_k, \qquad
-    N_k(\omega)\coloneq n_k, \qquad
-    K(\omega)\coloneq K
-  \end{gathered}
-\end{equation}
-and also
-\begin{equation}\label{eq:components2}
-  \begin{gathered}
-    T(\omega)\coloneq\left(T_k(\omega)\right)_{k=0}^{K(\omega)}, \qquad
-    U(\omega)\coloneq\left(U_k(\omega)\right)_{k=0}^{K(\omega)}, \\
-    X(\omega)\coloneq\left(X_k(\omega)\right)_{k=0}^{K(\omega)}, \qquad
-    N(\omega)\coloneq\left(N_k(\omega)\right)_{k=0}^{K(\omega)}.
-  \end{gathered}
-\end{equation}
-We will denote by $\mathring{\Omega}$ the set of all finite jump sequences, \ie $\mathring{\Omega}\coloneq\{\omega\in\Omega\;|\;K(\omega)<\infty\}$.
-
-Note that every element of $\Omega$ corresponds to a unique sample path of $\X_t$.
-In particular,
-\begin{equation}\label{eq:Xdef}
-  \X_t(\omega)=\sum_{k=0}^{K(\omega)}\!X_k(\omega)\,\indicator{\halfopen{T_k(\omega),\infty}}(t).
-\end{equation}
-Note that, for every $\omega\in\Omega$, the function $\X(\omega):\Rp\to\Xspace$ is thus \cadlag.
-It will be useful to define $\Xt$ to be the left-limit function of $\X$:
-\begin{equation*}
-  \Xt(\omega,t)\coloneq
-  \begin{cases}
-    \lim_{s\uparrow{t}}\X(\omega,t), & t>0,\\
-    \X(0), & t=0.\\
-  \end{cases}
-\end{equation*}
-
-The sample space $\Omega$ has a natural partial order.
-We write $\omega\preceq\omega'$ if $\omega'$ is an extension of $\omega$;
-that is, if $K(\omega)\le K(\omega')$, and $\left(T_k(\omega),U_k(\omega),X_k(\omega),N_k(\omega)\right)=\left(T_k(\omega'),U_k(\omega'),X_k(\omega'),N_k(\omega')\right)$ for $k=0,\dots,K(\omega)$.
-For $\omega\in\Omega$, the set
-$\left\{\omega'\in\Omega\;\vert\;\omega'\preceq\omega\right\}$
-is totally ordered.
-Moreover, for each $\omega\in\mathring{\Omega}$, there is a unique \emph{predecessor}, $\predec{\omega}$, such that $\predec{\omega}\preceq\omega$, $\predec{\omega}\ne\omega$, and, for all $\omega'$, $\predec{\omega}\preceq\omega'\preceq\omega$ implies that either $\omega'=\predec{\omega}$ or $\omega'=\omega$.
-
-In order to define probabilistic \emph{events}, it is necessary to define the $\sigma$-algebra of measurable subsets of $\Omega$.
-We first impose the condition that $\Xspace$ be a complete, separable, measurable metric space.
-We then follow \citet{King2022}, extending the Skorokhod metric \citep{Kallenberg1997,Ethier2009} to make $\Omega$ a complete, separable metric space (\ie a Polish space) and take our event space, $\Borel$, to be its Borel $\sigma$-algebra.
-
-We now turn to the probability measure, $\mathbb{P}$, on $\Omega\cong T(\Omega){\times}U(\Omega){\times}X(\Omega){\times}N(\Omega)$.
-We will specify this by giving its density.
-%% We will specify this by giving its density with respect to a base measure.
-%% A natural base measure on $T(\Omega)$ is the probability measure of the unit-rate Poisson point process \citep{Andersen1993,Kallenberg1997};
-%% the measure on $\Xspace$ extends naturally to a product measure on $X(\Omega)$;
-%% the counting measure serves for the discrete component $U(\Omega){\times}N(\Omega)$.
-%% Let $\pi$ denote the product of these three measures.
-%% We define a probability measure on $\Omega$ by specifying its density with respect to $\pi$.
-In particular, for each $u\in\Jumps$, $x,x'\in\Xspace$, we suppose $\auxmeas_{u,x,x'}$ is a given probability measure on $\Zp$;
-these will take specific forms below.
-For $t\in\Rp$ and $\omega\in\restrict{\Omega}{t}$, define the probability density function
-\begin{equation}\label{eq:probmeas}
-  P_t(\omega)\:\coloneq\:p_0(\X_0)\,\prod_{k=1}^{K}\left(\alpha_{U_k}(T_k,\Xt_{T_k},\X_{T_k})\,\auxmeas_{U_k,\Xt_{T_k},X_{T_k}}(N_k)\right)\exp{\left(-\int_0^t\sum_{ux'}{\alpha_u(s,\X_s,x')}\,\dd{s}\right)},
-\end{equation}
-\normalsize
-the latter sum being taken over $u\in\Jumps$, $x'\in\Xspace$.
-Here, for the sake of readability, we have suppressed the dependence of the random variables $\X_t$, $K$, $T_k$, $U_k$ and $N_k$ on $\omega$.
-%% It is readily verified that, for all $t$, $P_t$ is a probability density with respect to $\pi$.
+\subsection{The history, inventory, and genealogy processes}
 
@@
 
-Next among our constellation of related processes (\cref{fig:constellation}) is the \emph{history process}, which encapsulates the entire history of $\X$ up to time $t$.
-Specifically, for $\omega\in\Omega$, we define
+For $\omega\in\Omega$, $t\mapsto(\X_t(\omega),\U_t(\omega))$ is a \cadlag\ function of time.
+Let the \emph{history process}, $\Hist_t$, be the restriction of this random function to the interval $[0,t]$.
+Note that $\Hist_t$ is a Markov process.
+
+The non-explosiveness assumption implies that a.s., for every $t$, there is a finite, increasing sequence of random jump times $T_t\coloneq\left(t_k\right)_{k=1}^{K_t}$;
+the length $K_t$ of this sequence is itself random.
+However, conditional on $\Hist_t$, $T_t$ and $K_t$, together with the mark process $\U_t$, and the population process $\X_t$ are deterministic.
+Using these, one can write down an explicit probability measure for $\Hist_t$:
+\begin{equation}\label{eq:Hdens}
+  \begin{aligned}
+    \pi^{\Hist}(\dd{\Hist_t})
+    =&\pi^{\Hist}(\dd{T_t},\dd{\X_t},K_t,\U_t)\\
+    =&p_{0}(\X_{0})\,\dd{x_0}\,\prod_{k=1}^{K_{t}}{\left(\alpha_{\U_{t_k}}\left(t_k,\Xt_{t_k},\X_{t_k}\right)\,\dd{x_k}\,\dd{t_k}\right)}\\
+    &\qquad\times\exp{\left(-\sum_{u}{\int_{0}^{t}{\int{\alpha_u(t',\X_{t'},x')\dd{x'}}\dd{t'}}}\right)}.
+  \end{aligned}
+\end{equation}
+
+\paragraph{Inventories}
+
+Our goal in this paper is to probabilistically characterize how the genealogical relationships among lineages evolve through time.
+Accordingly, we develop some notation for this purpose.
+To begin with, we assign to each lineage a unique number $j\in\Zp$.
+This can be done in any fashion, so long as no two lineages ever receive the same number.
+For example, when a new lineage arises, we can assign it the smallest integer that has not yet been assigned.
+We will define the \emph{inventory process}, $\Inv_t$, so that, for every lineage $j\in\Zp$, $\Inv_t(j)$ is the deme in which $j$ is found.
+However, when $t$ is before the birth or after the death of lineage $j$, then clearly $\Inv_t(j)\notin\Demes$.
+We say in this case that lineage $j$ is in the \emph{underdeme}, which we denote using the symbol $\eth$, so that we can write $\Inv_t(j)=\eth$ and define $\Demesplus\coloneq\Demes\cup\{\eth\}$.
+With this definition, $\Inv:\Rp\times\Zp\to\Demesplus$.
+
+The birth and death times of lineage $j$ are therefore
 \begin{equation*}
-  \Hist_t(\omega) \coloneq \left(t,\left(T_k(\omega),U_k(\omega),X_k(\omega)\right)_{k=0}^{K(\restrict{\omega}{t})}\right),
+  \begin{gathered}
+    t^b_j=\min\{t|\Inv_t(j)\ne{\eth}\}\qquad
+    \text{and}\qquad
+    t^d_j=\sup\{t|\Inv_t(j)\ne{\eth}\},
+  \end{gathered}
 \end{equation*}
-where $T_k$, $U_k$, $X_k$, and $K$ are as in \cref{eq:components1}.
-Thus $\Hist_t$ contains exactly those elements of $\omega$ that are relevant to the history of the population process $\X_t$.
-It is trivial to verify that $\Hist_t$ is Markov and to compute its probability density.
-In particular, given any history $h_t=\left(t,\left(t_k,u_k,x_k\right)_{j=0}^{K}\right)$,
-the marginal density at $h_t$ is readily seen to be
-\begin{equation*}
-  \begin{aligned}
-    P_{\Hist_t}(h_t)\coloneq &p_0(x_0)\,\prod_{k=1}^{K}\left\{\alpha_{u_k}(t_k,x_{k-1},x_k)\,\exp{\left(-\int_{t_{k-1}}^{t_k}\sum_{ux'}{\alpha_u(s,x_{k-1},x')}\,\dd{s}\right)}\right\}\\
-    &\qquad\times\,\exp{\left(-\int_{t_{K}}^{t}\sum_{ux'}{\alpha_u(s,x_{K},x')}\,\dd{s}\right)}.
-  \end{aligned}
-\end{equation*}
+respectively.
+Observe that $n_i(\X_t)=\left\vert\left\{j\;\vert\;\Inv_t(j)=i\right\}\right\vert$ for all $t\in\Rp$ and $i\in\Demes$.
+Note also that $n$ does not count the inhabitants of the underdeme.
+
+\paragraph{Jump types}
+
+Different kinds of events that occur for the population process can have different kinds of effects on the inventory process, and indeed not every jump affects $\Inv_t$ at all.
+From the point of view of the inventory process, there are five distinct \emph{types} of jumps, which we enumerate here.
+\begin{enumerate}[1.]
+\item Birth-type events result in the branching of one or more new lineages, each from some existing lineage.
+  If $j$ is one of the new lineages, we use the expression $\Anc(j)$ to refer to its ancestor.
+  Examples of birth-type events include transmission events, speciations, and actual births.
+  It is not assumed that all new lineages arising from a birth event share the same ancestor.
+\item Death-type events result in the extinction of one or more lineages.
+  Examples include recovery, death of a host, and species extinctions.
+\item Migration-type events result in the movement of a lineage from one deme to another.
+  Spatial movements, changes in behavior, and progression of an infection can all be represented as migration-type events.
+\item Sample-type events result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
+\item Neutral-type events result in no change to any of the lineages.
+\end{enumerate}
+\Cref{fig:markov_state} depicts an example with all five of these types.
+It is not necessary that a jump fall into just one of these types.
+It is allowable, for instance to have compound jumps that fall into more than one category.
+For example, Sample-death-type events, in which a lineage is simultaneously sampled and removed, have been used, as have birth-death events in which one lineage reproduces at the same moment that another dies.
+The theory presented here places no restrictions on these events.
+
+However, we do impose the restriction that the \emph{production}, \ie the deme-specific number of lineages emerging from the event, be constant for all jumps of a given mark.
+To be precise, the \emph{production} is defined to be a function $r:\Jumps\times\Demes\to\Zp$, such that $r_u^i$ lineages of deme $i$ emerge from each event of mark $u$.
+We write $r_u=(r_u^i)_{i\in\Demes}$.
+Note that we do not count lineages that die as a result of the event.
+Also, it is important to note that the parent lineage or lineages, if they survive the event, are always counted among the emergent lineages.
+
+Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents, there is also assumed to be a deterministic function $Q_u$, for $u\in\Jumps$, (described below) that expresses these properties.
 
@@
 
-Our main interest is in tracking the genealogical relationships among the lineages in our demes.
-Accordingly, we will assign each individual lineage a unique number.
-let $\Inv_t:\Rp\to\Demesplus^{\Zp}$ be such that $\Inv_t(n)=i$ if and only if lineage $n$ is in deme $i$ at time $t$.
-If $t$ is before the birth or after the death of $n$, $\Inv_t(n)=\eth$.
-Note that $n_i(\X_t)=\left\vert\left\{n\;\vert\;\Inv_t(n)=i\right\}\right\vert$ for all $t\in\Rp$ and $i\in\Demes$.
+The structure of the state space for the inventory process, $\Inv_t$, has already been described.
+It remains to define its stochastic dynamics.
+The $\Inv_t$ process follows the population process $\X_t$ in that jumps in $\Inv_t$ do not occur except when jumps in $\X_t$ occur: $\predec{\Inv}_t\ne\Inv_t$ implies $\predec{\N}_t\ne\N_t$.
 
-\paragraph{Genealogy process}
+At jumps of birth type and mark $u$, one or more random parents are selected from the appropriate deme(s).
+The appropriate number of offspring are created in each deme.
+
+\paragraph{Genealogies}
+
+\citet{King2022}, showed how an unstructured population process induces a process on the space of genealogies.
+Although we now treat a more general case, the construction is much the same, so we abbreviate the presentation.
+Readers wishing for more detail should consult the earlier paper \citep{King2022}.
+
+Formally, we define a \emph{genealogy}, $\Gen$, to be a finite sequence of \emph{internal nodes}, together with a time.
+The time is denoted $T(\Gen)$ and the number of nodes is $K(\Gen)$.
+The $k$-th node is $\Node{p}_k(\Gen)$ and we write $\Node{p}\in\Gen$ if $\Node{p}$ is one of the nodes of $\Gen$.
+Each $\Node{p}\in\Gen$ has a creation-time, $T(\Node{p})$ and a parent, $\Anc(\Node{p})$.
+%% The nodes are ordered so that $k<k'$ implies that $T(\Node{p}_k(\Gen))\le{T(\Node{p}_{k'}(\Gen))}$.
+Root nodes are distinguished by being their own parents:
+$\Node{p}$ is a root if and only if $\Anc(\Node{p})=\Node{p}$.
+Every node also has one or more descendants called \emph{children}.
+There are three kinds of children:
+\begin{inparaenum}[(i)]
+\item \emph{Internal nodes}.
+  Children nodes must always be later than their parents:
+  $T(\Anc(\Node{p}))<T(\Node{p})$.
+\item \emph{Tip nodes}, which represent extant lineages.
+  In particular, every lineage alive at time $t$ is the child of some node.
+\item \emph{Sample nodes}.
+  When a sample is collected at time $t$, a new node, $\Node{p}$ is added with $T(\Node{p})=t$ and the sample as child.
+\end{inparaenum}
+
+\paragraph{Genealogy processes}
+
+The population process induces a stochastic process on the space of genealogies.
+In particular, at each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event:
+\begin{enumerate}[(a)]
+\item A birth-type event at time $t$ results in the creation of one new internal node for each parent lineage.
+  In particular, if $j$ is one of the parent lineages, and $\Node{p}$ is the new node, then $T(\Node{p})=t$, $\Anc(\Node{p})$ is the node in which $j$ lay prior to the event.
+  The children of $\Node{p}$ include all the new lineages that branched from $j$, as well as $j$ itself.
+\item In a death-type event, all the lineages $j$ that die are removed.
+  Nodes without children are then recursively removed.
+\item In a migration-type event, one node is added for each migrating lineage;
+  each one takes the migrating lineage as child.
+  The ancestor of the new node is the node in which the migrating lineage lay before the event.
+\item At a sample-type event, one new node is introduced for each sampled lineage.
+  Each one takes the sampled lineage as child, along with the sample node.
+  The ancestor of the node is that in which the sampled lineage lay before the vent.
+\item At a neutral-type event, no change is made to the genealogy.
+\item Finally, events of compound type are readily accommodated by combining the rules just stated.
+\end{enumerate}
+
+\begin{figure}
+  \caption{
+    Illustration of genealogy processes.
+    \AAK{[Similar to that of \citet{King2022} but with multiple demes represented.]}
+  }
+\end{figure}
 
 \paragraph{Pruning}
-
-\paragraph{Filter equations}
-
-\begin{defn}
-  Suppose $X_t$ is a Markov process with Kolmogorov forward equation (KFE)
-  \begin{equation}\label{eq:kfe2}
-    \begin{gathered}
-      \frac{\partial{w}}{\partial{t}}=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
-      \qquad
-      w(0,x)=p_0(x).
-    \end{gathered}
-  \end{equation}
-  Suppose that $B(x,x')>0$ and $\lambda(t,x)>0$ are given functions, and $t_1<t_2<{\dots}<t_K$ is a given sequence of times.
-  We say that the equations
-  \begin{gather}
-    \frac{\partial{w}}{\partial{t}}=\int\!w(t,x')\,\alpha(t,x',x)\,B(x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'}-\lambda(t,x)\,w(t,x),\label{eq:filtereq1}\\
-    w(t_k,x)=\int\!\wt(t_k,x')\,\frac{\alpha(t,x',x)}{\mu}\,B(x',x)\,\dd{x'},\quad k=1,\dots,K,\label{eq:filtereq2}\\
-    w(0,x)=p_0(x).
-  \end{gather}
-  are the \emph{filter equations} for $X_t$ with \emph{boost} $B$, \emph{decay} $\lambda$, \emph{initial density} $p_0$, and \emph{observation times} $t_k$.
-\end{defn}
-
-The useful fact about filter equations is that they afford a means of computing the likelihood of a given sequence of events.
-This is facilitated by the following
-\begin{lemma}\label{lemma:filtereq}
-  \Cref{eq:filtereq1} is satisfied by $w(t,x)=\int{v\,U(t,x,v)\,\dd{v}}$, where $U$ satisfies the KFE
-  \begin{equation}\label{eq:filterlemma}
-    \begin{aligned}
-      \frac{\partial{U}}{\partial{t}}=&\int{U(t,x',v')\,\alpha(t,x',x)\,\delta(v,B(x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int{U(t,x,v)\,\alpha(t,x,x')\,\delta(v',B(x,x')\,v)\,\dd{x'}\,\dd{v'}}
-      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,U(t,x,v)\right].
-    \end{aligned}
-  \end{equation}
-  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
-\end{lemma}
-\begin{proof}
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
-      =&\int{v\,\frac{\partial{U}}{\partial{t}}(t,x,v)\,\dd{v}}\\
-      =&\int{v\,U(t,x',v')\,\alpha(t,x',x)\,\delta(v,B(x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int{v\,U(t,x,v)\,\alpha(t,x,x')\,\delta(v',B(x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
-      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,U(t,x,v)\right]\,\dd{v}}.\\
-    \end{aligned}
-  \end{equation*}
-  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
-      =&\int{v'\,U(t,x',v')\,\alpha(t,x',x)\,B(x',x)\,\dd{v'}\,\dd{x'}}
-      -\int{v\,U(t,x,v)\,\alpha(t,x,x')\,\dd{v}\,\dd{x'}}\\
-      &\qquad-\lambda(t,x)\,\int{v\,U(t,x,v)\,\dd{v}},
-    \end{aligned}
-  \end{equation*}
-  which is simplified to obtain \cref{eq:filtereq1}.
-\end{proof}
-
-We recognize in \cref{eq:filterlemma} the KFE of a certain process $(X_t,V_t)$.
-In particular, $X_t$ is a jump process with KF \cref{eq:kfe2}.
-The $V_t$ process has jumps wherever $X_t$ does, such that when $X_t$ jumps $x\mapsto{x'}$, $V_t$ jumps by a factor of $B(x,x')$.
-Between jumps, $V_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
-If we view $V_t$ as a weight, then \cref{lemma:filtereq} says that the weighted average of $X_t$ evolves according to \cref{eq:filtereq1}.
 
@@
 Each node encountered along the way represents a genealogical event from which $j$ emerges.
-Moreover, at each time $t\in\halfopen{t^o_j,t^s_j}$, lineage $j$ is in precisely one of the demes $D$.
+Moreover, at each time $t\in\halfopen{t^o_j,t^s_j}$, lineage $j$ is in precisely one of the demes $\Demes$.
 However, for $t\notin\halfopen{t^o_j,t^s_j}$, lineage $j$ does not exist.
-To express this, we say that lineage $j$ is in the \emph{undeme}, which we denote using the symbol $\eth$.
+To express this, we again say that lineage $j$ is in the \emph{underdeme}, which we denote using the symbol $\eth$.
 We define $\Demesplus\coloneq\Demes\cup\{\eth\}$.
@@
 It will be useful to define a function that captures all the relevant features of a pruned genealogy.
-Accordingly, let $\Yspace=(\Zp\times\Demesplus\times\Lin)^{\Lin}$ and define $y:\Rp\to\Yspace$ so that, for $j\in\Lin$ and $t\in\Rp$:
+Accordingly, let $\Yspace=\Zp\times\Demesplus\times\Lin$ and define $y:\Rp\times\Lin\to\Yspace$ so that, for $j\in\Lin$ and $t\in\Rp$:
 \begin{compactenum}[(a)]
@@
   In particular, $\anc(y_j(t))=j$ for $t\ge{t^o_j}$, but $\anc(y_j(t))$ is well defined for all $j$ and $t$.
+  \AAK{[Can define this in a way similar to that of the original Kingman's coalescent\dots.]}
 \end{compactenum}
 One can verify that $y$ is \cadlag.
-Define $\yt(t)=\lim_{s\uparrow{t}}y(s)$.
+Define $\yt(t)=\lim_{t'\uparrow{t}}y(t')$.
 
@@
   \begin{equation*}
-    \prob{\Gen\vert\Hist}=\prod_{(u,t,x)\in\Hist}\phi_u(x,\yt(t),y(t)).
+    \Prob{\Prune_t\vert\Hist_t}=\prod_{k=1}^{K}{\phi_u\left(\X_{t_k},\yt(t_k),y(t_k)\right)}=\prod_{k=1}^{K}{\sum_{y_k}{\pi(y_{k-1},y_k)\,\frac{\phi_u\left(\X_{t_k},y_{k-1},y_k\right)}{\pi(y_{k-1},y_k)}\,\kappa^{\Prune_t}(t_k,y_{k-1},y_k)}},
   \end{equation*}
+  where $\kappa^{\Prune_t}(t,y,y')=\Indicator{\yt^{\Prune_t}(t)=y, y^{\Prune_t}(t)=y'}$.
+  \AAK{[Note that we can just sum over $\Prune_t$ and interchange sums and products (using conditional independence) to get Thm 2!]}
 \end{thm}
+\begin{proof}
+  Two approaches possible:
+  \begin{enumerate}
+  \item Conditional on $\Hist_t$, the likelihood of $\Prune_t$ is the product of $p_{jk}$, where $j$ ranges over lineages and $k$ over events.
+    Interchanging the order of integration yields the expression given.
+  \item Use the filter equations developed in the Appendix.
+  \end{enumerate}
+\end{proof}
 
@@
 
+\begin{thm}
+  State the filter equation definition.
+\end{thm}
+\begin{proof}
+  Two approaches possible:
+  \begin{enumerate}
+  \item Use the filter equations developed in the Appendix to compute $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}$.
+    In effect, we compute $\Prob{\Prune_t}$ and then sum over paintings $y$.
+  \item Change the order of the summation: $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}=\Expect{\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}\;\vert\;\Hist_t}}$.
+    That is, conditional on $\Hist_t$, the likelihood of $\Vis_t$ is sum over all $\Prune_t$ such that $\obs(\Prune_t)=\Vis_t$.
+    Therefore, any importance sampling scheme for $\Prune_t$ will do.
+    In particular, we choose a scheme that paints the tree forward in time, driven by a mimic of $\X_t$.
+    We show that this is equivalent to the specific filter equation.
+
+    It may be useful in this to first show that $\Prob{\Prune_t|\Hist_t}$ can be expressed as an expectation over $y$, where $y$ is constrained to be equal $y^{\Prune_t}$.
+    Then, show that the sum of the $\Prune_t$-constraint indicator functions is equal to the $\Vis_t$-constraint indicator function.
+  \end{enumerate}
+\end{proof}
+
 Let $\kappa^{\Vis}$ be the indicator function for the condition that a pruned genealogy is compatible with a given obscured genealogy.
@@
 
+\appendix
+
+\section{Filter equations}
+
+Explicit expressions for the probability densities that will arise in the following are not always available.
+Hence, we will develop some technology for manipulating them that avoids the need for explicit expressions.
+
+\begin{defn}
+  Suppose $X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
+  \begin{equation}\label{eq:kfe2}
+    \frac{\partial{w}}{\partial{t}}
+    =\int{w(t,x')\,\beta(t,x',x)\,\dd{x'}}
+    -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}},
+  \end{equation}
+  Suppose that $B(t,x,x')>0$ and $\lambda(t,x)\in\R$ are given functions.
+  We say that the equation
+  \begin{equation}\label{eq:filtereq}
+    \frac{\partial{w}}{\partial{t}}
+    =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
+    -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
+    -\lambda(t,x)\,w(t,x)
+  \end{equation}
+  is the \emph{filter equation} with \emph{driver} $X_t$, \emph{boost} $B$, and \emph{decay} $\lambda$.
+\end{defn}
+
+Filter equations afford a convenient means of computing the likelihood of a given sequence of events.
+This is facilitated by the following
+\begin{lemma}\label{lemma:filtereq}
+  \Cref{eq:filtereq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
+  \begin{equation}\label{eq:filterlemma}
+    \begin{aligned}
+      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
+      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
+      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right].
+    \end{aligned}
+  \end{equation}
+  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
+\end{lemma}
+\begin{proof}
+  \begin{equation*}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}
+      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
+      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
+      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
+      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
+    \end{aligned}
+  \end{equation*}
+  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
+  \begin{equation*}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}
+      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
+      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
+      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
+    \end{aligned}
+  \end{equation*}
+  which is simplified to obtain \cref{eq:filtereq}.
+\end{proof}
+
+We recognize in \cref{eq:filterlemma} the KFE of a certain process $(X_t,V_t)$.
+In particular, $X_t$ is the driver with KFE \cref{eq:kfe2}.
+The $V_t$ process has jumps wherever $X_t$ does, such that when $X_t$ jumps from $x$ to $x'$, $V_t$ jumps by the multiplicative factor $B(t,x,x')$.
+Between jumps, $V_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
+If we view $V_t$ as a weight, then \cref{lemma:filtereq} says that the $V_t$-weighted average of $X_t$ evolves according to \cref{eq:filtereq}.
+This motivates the following result, which effectively allows boosts of zero.
+
+\begin{prop}\label{prop:zeroboost}
+  Suppose $X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE as in \cref{eq:kfe2}.
+  Let $H_t$ be its history process.
+  That is, for $\omega\in\Omega$ and $t\in\Rp$, $H_{t}(\omega):[0,t]\to\Xspace$ such that, for $t'\in{[0,t]}$,
+  $H_{t}(\omega)(t')=X_{t'}(\omega)$.
+  Moreover, there are random variables $K\in\Zp$ and $t_k\in\halfclosed{0,t}$, $k=1,\dots,K$ such that $\predec{X}_t=X_t$ whenever $t\ne{t_k}$ for all $k$.
+  Suppose $F$ is a real-valued function of $H_t$ such that
+  \begin{equation*}
+    F(H_t)=\prod_{k}{Q\Big(t_k,\predec{H}(t_k),H(t_k)\Big)\,B\Big(t_k,\predec{H}(t_{k}),H(t_k)\Big)},
+  \end{equation*}
+  for some given measurable functions $B>{0}$, $Q\in\{0,1\}$.
+  Suppose $w$ satisfies the filter equation
+  \begin{equation*}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}(t,x)
+      =&\int{w(t,x')\,\beta(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
+      -\int{w(t,x)\,\beta(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
+      &\qquad-\int{w(t,x)\,\beta(t,x,x')\,\Big(1-Q(t,x,x')\Big)\,\dd{x'}}.
+    \end{aligned}
+  \end{equation*}
+  Then $\Expect{F(H_t)}=\int{w(t,x)\dd{x}}$.
+\end{prop}
+\begin{proof}
+  Apply \cref{lemma:filtereq} with the driver generated by the rate functions $\beta(t,x,x')\,Q(t,x,x')$.
+\end{proof}
+
+An important special case is that of a deterministic driving process.
+The following result is established by routine calculation.
+
+\begin{prop}
+  Suppose $X:[0,T]\to\Xspace$ is a deterministic, piecewise constant, \cadlag\ function.
+  Let $E$ be the set of its jump times.
+  Then the KFE for $X_t$ is \cref{eq:kfe2} with $\beta(t,x,x')=\sum_{e\in{E}}{\delta(t,e)\delta(x',X_e)}$.
+  With this driver, the filter equation (\cref{eq:filtereq}) becomes
+  \begin{equation*}
+    \begin{gathered}
+      \frac{\partial{w}}{\partial{t}}=
+      -\lambda(t,x)\,w(t,x)\quad\text{for}\quad{t\notin{E}},\\
+      w(e,x)=\delta(x,X_e)\,\int{\wt(e,x')\,B(e,x',X_e)\,\dd{x'}}\quad\text{for}\quad{e\in{E}}.
+    \end{gathered}
+  \end{equation*}
+\end{prop}
+
+The results so far allow us to compute expectations over random jumps and jump times.
+We will have occasion to compute marginal expectations in the situation where some jump times are known.
+The following result handles this case.
+\AAK{[Not sure if this is stated correctly.]}
+
+\begin{prop}
+  Suppose $X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE as in \cref{eq:kfe}.
+  Let $H_t$ be its history process and $E_t=\event(H_t)$, the set of its jump times to time $t$.
+  Suppose $0<t_1<\dots<t_N\le{t}$ are fixed times and set $O=\{t_1,\dots,t_N\}$.
+  Suppose $F$ is an $\Rp$-valued function of $H_t$ and $O$ such that
+  \begin{equation*}
+    F(H_t,O)=\prod_{e\in{E_t}}{Q\left(e,\predec{H}(e),H(e)\right)\,B\left(e,\predec{H}(e),H(e)\right)}
+    \times\prod_{e\in{O}}{R\left(e,\predec{H}(e),H(e)\right)\,C\left(e,\predec{H}(e),H(e)\right)},
+  \end{equation*}
+  for some given measurable functions $B,C>{0}$, $Q,R\in\{0,1\}$.
+  Suppose $w$ satisfies the filter equation
+  \begin{equation*}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}(t,x)
+      =&\int{w(t,x')\,\beta(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
+      -\int{w(t,x)\,\beta(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
+      &-\sum_{e\in{O}}{\int{w(t,x')\,\delta(t,e)\,\,R(e,x',x)\,C(e,x',x)\,\dd{x'}}}
+      -\sum_{e\in{O}}{\int{w(t,x)\,\delta(t,e)\,\,R(e,x,x')\,\dd{x'}}}\\
+      &\qquad-\int{w(t,x)\,\beta(t,x,x')\,\left(1-Q(t,x,x')\right)\,\dd{x'}}.
+    \end{aligned}
+  \end{equation*}
+  Then $\Expect{F(H_t,O)}=\int{w(t,x)\dd{x}}$.
+\end{prop}
+\begin{proof}
+  Apply \cref{lemma:filtereq} with the driver generated by $\beta(t,x,x')=\alpha(t,x,x')+\sum_{e\in{O}}{\delta(t,e)\,\pi(x,x')}$.
+\end{proof}
+
 \end{document}

### 2023-07-06 ms.Rnw
edits to last proposition

@@
 The following result handles this case.
-\AAK{[Not sure if this is stated correctly.]}
 
 \begin{prop}
-  Suppose $X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE as in \cref{eq:kfe}.
-  Let $H_t$ be its history process and $E_t=\event(H_t)$, the set of its jump times to time $t$.
-  Suppose $0<t_1<\dots<t_N\le{t}$ are fixed times and set $O=\{t_1,\dots,t_N\}$.
-  Suppose $F$ is an $\Rp$-valued function of $H_t$ and $O$ such that
+  Suppose $\X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE as in \cref{eq:kfe}.
+  Let $\Hist_t$ be its history process.
+  Suppose $0<t_1<\dots<t_N\le{t}$ are fixed times and set $E=\{t_1,\dots,t_N\}$.
+  Let $B>{0}$ and $Q\in\{0,1\}$ be given measurable functions.
+  Suppose $F_E$ is an $\Rp$-valued random variable such that
   \begin{equation*}
-    F(H_t,O)=\prod_{e\in{E_t}}{Q\left(e,\predec{H}(e),H(e)\right)\,B\left(e,\predec{H}(e),H(e)\right)}
-    \times\prod_{e\in{O}}{R\left(e,\predec{H}(e),H(e)\right)\,C\left(e,\predec{H}(e),H(e)\right)},
+    \Expect{F_E\;\vert\;\Hist_t}={\prod_{e\in{E}}{B(e,\Xt_e,\X_e)\,Q(e,\Xt_e,\X_e)}}.
   \end{equation*}
-  for some given measurable functions $B,C>{0}$, $Q,R\in\{0,1\}$.
-  Suppose $w$ satisfies the filter equation
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)
-      =&\int{w(t,x')\,\beta(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\beta(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
-      &-\sum_{e\in{O}}{\int{w(t,x')\,\delta(t,e)\,\,R(e,x',x)\,C(e,x',x)\,\dd{x'}}}
-      -\sum_{e\in{O}}{\int{w(t,x)\,\delta(t,e)\,\,R(e,x,x')\,\dd{x'}}}\\
-      &\qquad-\int{w(t,x)\,\beta(t,x,x')\,\left(1-Q(t,x,x')\right)\,\dd{x'}}.
-    \end{aligned}
-  \end{equation*}
-  Then $\Expect{F(H_t,O)}=\int{w(t,x)\dd{x}}$.
+  Let $\pi(x,x')$ be a given probability kernel and $Y_t$ be the Markov process generated by $\beta(t,x,x')=\left(\alpha(t,x,x')+\sum_{e\in{E}}{\delta(t,e)\,\pi(x,x')}\right)\,Q(t,x,x')$.
+  Suppose $w$ satisfies the filter equation with driver $Y_t$, boost $B$, and decay $\lambda=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\,\dd{x'}}$.
+  Then $\Expect{F_E}=\int{w(t,x)\,\dd{x}}$.
 \end{prop}
 \begin{proof}
-  Apply \cref{lemma:filtereq} with the driver generated by $\beta(t,x,x')=\alpha(t,x,x')+\sum_{e\in{O}}{\delta(t,e)\,\pi(x,x')}$.
+  Apply \cref{lemma:filtereq}.
 \end{proof}

### 2023-07-11 ms.Rnw
thms 1 & 2

@@
 
-Denote the underlying probability space by $(\Omega,\Borel,\mathbb{P})$.
-We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\coloneq\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$ and taking values in some space $\Xspace$.
+Denote the underlying probability space by $(\Omega,\Borel,\mathrm{Prob})$.
+We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\colonequals\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$ and taking values in some space $\Xspace$.
 In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
@@
 \begin{equation}\label{eq:ic}
-  \Prob{\X_0=x}=p_0(x),
+  \Prob{\X_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
 \end{equation}
-for some choice of initial distribution, $p_0$.
+for all measurable sets $\mathcal{E}\subseteq\Xspace$.
 For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
@@
 In fact, all of the processes we will describe in this paper will be taken to be \cadlag, and we will frequently need to refer to the left-limit.
-Accordingly, if $z=z(\omega,t)$ is any \cadlag\ random function, we define
+Accordingly, if $\rdm{Z}_t$ is any \cadlag\ random process, we define
 \begin{equation*}
-  \zt(\omega,t)\coloneq
+  \leftlim{\rdm{Z}}_t\colonequals
   \begin{cases}
-    \lim_{t'\uparrow{t}}z(\omega,t'), & t>0,\\
-    z(\omega,0), & t=0.\\
+    \lim_{t'\uparrow{t}}\rdm{Z}_t, & t>0,\\
+    \rdm{Z}_0, & t=0.\\
   \end{cases}
 \end{equation*}
-Note that $\zt$ is thus left-continuous with right limits.
+Note that $\leftlim{\rdm{Z}}_t$ is thus left-continuous with right limits.
 
@@
 
-The non-explosiveness assumption implies that a.s., for every $t$, there is a finite, increasing sequence of random jump times $T_t\coloneq\left(t_k\right)_{k=1}^{K_t}$;
-the length $K_t$ of this sequence is itself random.
-However, conditional on $\Hist_t$, $T_t$ and $K_t$, together with the mark process $\U_t$, and the population process $\X_t$ are deterministic.
-Using these, one can write down an explicit probability measure for $\Hist_t$:
+The non-explosiveness assumption implies that a.s., for every $t$, there is a finite, increasing sequence of random event times, which we denote by $\event(\Hist_t)\colonequals\left(\rdm{T}_k\right)_{k=1}^{\rdm{K}_t}$;
+the length, $\rdm{K}_t$, of this sequence is itself random.
+However, conditional on $\Hist_t$, $\rdm{T}_k$ and $\rdm{K}_t$, together with the mark process $\U_t$, and the population process $\X_t$ are deterministic.
+One can write down an explicit probability measure for $\Hist_t$ in terms of these:
 \begin{equation}\label{eq:Hdens}
@@
     \pi^{\Hist}(\dd{\Hist_t})
-    =&\pi^{\Hist}(\dd{T_t},\dd{\X_t},K_t,\U_t)\\
-    =&p_{0}(\X_{0})\,\dd{x_0}\,\prod_{k=1}^{K_{t}}{\left(\alpha_{\U_{t_k}}\left(t_k,\Xt_{t_k},\X_{t_k}\right)\,\dd{x_k}\,\dd{t_k}\right)}\\
-    &\qquad\times\exp{\left(-\sum_{u}{\int_{0}^{t}{\int{\alpha_u(t',\X_{t'},x')\dd{x'}}\dd{t'}}}\right)}.
+    =&p_{0}(X_{0})\,\dd{X_0}\,\prod_{k=1}^{K_{t}}{\left(\alpha_{U_{T_k}}\left(T_k,\leftlim{X}_{T_k},X_{T_k}\right)\,\dd{X_{T_k}}\,\dd{T_k}\right)}\\
+    &\qquad\times\exp{\left(-\sum_{u}{\int_{0}^{t}{\int{\alpha_u(t',\leftlim{X}_{t'},x')\,\dd{x'}}\,\dd{t'}}}\right)}.
   \end{aligned}
@@
 However, when $t$ is before the birth or after the death of lineage $j$, then clearly $\Inv_t(j)\notin\Demes$.
-We say in this case that lineage $j$ is in the \emph{underdeme}, which we denote using the symbol $\eth$, so that we can write $\Inv_t(j)=\eth$ and define $\Demesplus\coloneq\Demes\cup\{\eth\}$.
-With this definition, $\Inv:\Rp\times\Zp\to\Demesplus$.
+We say in this case that lineage $j$ is in the \emph{underdeme}, which we denote using the symbol $\eth$, so that we can write $\Inv_t(j)=\eth$.
+We define $\Demesplus\colonequals\Demes\cup\{\eth\}$ so that $\Inv:\Rp\times\Zp\to\Demesplus$.
 
@@
 Different kinds of events that occur for the population process can have different kinds of effects on the inventory process, and indeed not every jump affects $\Inv_t$ at all.
-From the point of view of the inventory process, there are five distinct \emph{types} of jumps, which we enumerate here.
-\begin{enumerate}[1.]
+From the point of view of the inventory process, there are five distinct \emph{pure types} of jumps, which we enumerate here.
+\begin{enumerate}[(a)]
 \item Birth-type events result in the branching of one or more new lineages, each from some existing lineage.
@@
 \item Migration-type events result in the movement of a lineage from one deme to another.
-  Spatial movements, changes in behavior, and progression of an infection can all be represented as migration-type events.
+  Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
 \item Sample-type events result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
@@
 \end{enumerate}
-\Cref{fig:markov_state} depicts an example with all five of these types.
+\Cref{fig:markov_state} depicts an example with all five of the pure types.
 It is not necessary that a jump fall into just one of these types.
 It is allowable, for instance to have compound jumps that fall into more than one category.
-For example, Sample-death-type events, in which a lineage is simultaneously sampled and removed, have been used, as have birth-death events in which one lineage reproduces at the same moment that another dies.
-The theory presented here places no restrictions on these events.
+For example, sample/death-type events, in which a lineage is simultaneously sampled and removed, have been used, as have birth/death events in which one lineage reproduces at the same moment that another dies.
+The theory presented here places no restrictions on the complexity of the events that can occur.
 
 However, we do impose the restriction that the \emph{production}, \ie the deme-specific number of lineages emerging from the event, be constant for all jumps of a given mark.
-To be precise, the \emph{production} is defined to be a function $r:\Jumps\times\Demes\to\Zp$, such that $r_u^i$ lineages of deme $i$ emerge from each event of mark $u$.
-We write $r_u=(r_u^i)_{i\in\Demes}$.
-Note that we do not count lineages that die as a result of the event.
-Also, it is important to note that the parent lineage or lineages, if they survive the event, are always counted among the emergent lineages.
+To be precise, the \emph{production} is defined to be a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
+We write $r^u=(r^u_i)_{i\in\Demes}$.
+Note that we lineages that die as a result of an event do not count in the production.
+Also, it is important to note that the parent lineage or lineages, if they survive the event, are always counted in the production.
 
-Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents, there is also assumed to be a deterministic function $Q_u$, for $u\in\Jumps$, (described below) that expresses these properties.
+Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents and demes, there is implicitly a deterministic indicator function $Q_u$, for $u\in\Jumps$, (described below) that captures these properties.
 
@@
 It remains to define its stochastic dynamics.
-The $\Inv_t$ process follows the population process $\X_t$ in that jumps in $\Inv_t$ do not occur except when jumps in $\X_t$ occur: $\predec{\Inv}_t\ne\Inv_t$ implies $\predec{\N}_t\ne\N_t$.
+The $\Inv_t$ process is driven by the population process $\X_t$ in that jumps in $\Inv_t$ do not occur except when jumps in $\X_t$ occur:
+$\leftlim{\Inv}_t\ne\Inv_t$ implies $\leftlim{\N}_t\ne\N_t$.
 
-At jumps of birth type and mark $u$, one or more random parents are selected from the appropriate deme(s).
-The appropriate number of offspring are created in each deme.
+At jumps of birth type and mark $u$, the appropriate number of random parents are selected from the appropriate deme(s) and each one sires the appropriate number of offspring in each deme.
+At jumps of death type, the appropriate number of lineages are selected from the appropriate demes and moved to the underdeme.
+At jumps of migration type, randomly selected lineages are moved between demes as appropriate.
+At sample-type and neutral-type jumps, no change occurs.
+Jumps of compound type are handled in the obvious way.
+
+Crucially, the assumption that the population process $\X_t$ is Markov empowers us to assume that the individual lineages within each deme are exchangeable with respect to the inventory process.
+Since the law governing $\X_t$ is independent of the identities of the individual lineages, and since the individual lineages are exchangeable by assumption, it follows that, conditional on $\Hist_t$, the jumps of the inventory are independent of one another.
 
@@
 Although we now treat a more general case, the construction is much the same, so we abbreviate the presentation.
-Readers wishing for more detail should consult the earlier paper \citep{King2022}.
+Readers wishing for more detail should consult the earlier paper.
 
-Formally, we define a \emph{genealogy}, $\Gen$, to be a finite sequence of \emph{internal nodes}, together with a time.
-The time is denoted $T(\Gen)$ and the number of nodes is $K(\Gen)$.
-The $k$-th node is $\Node{p}_k(\Gen)$ and we write $\Node{p}\in\Gen$ if $\Node{p}$ is one of the nodes of $\Gen$.
-Each $\Node{p}\in\Gen$ has a creation-time, $T(\Node{p})$ and a parent, $\Anc(\Node{p})$.
+For our purposes, a \emph{genealogy} is a labeled, time-calibrated tree.
+Its edges represent relationships of ancestry and descent among its nodes.
+There are three distinct kinds of nodes:
+\begin{inparaenum}[(i)]
+\item \emph{tip nodes}, which represent labeled extant lineages;
+\item \emph{internal nodes}, which represent ancestral events;
+\item \emph{sample nodes}, which represent labeled samples.
+\end{inparaenum}
+More formally, we can take a genealogy, $G$, to be a finite sequence of internal nodes, together with a time.
+Given $G$, the time is denoted $T(G)$; this is the time corresponding to the extant lineages.
+The number of nodes is $K(G)$.
+We order the nodes temporally, and denote the $k$-th node $\Node{p}_k(G)$ and we write $\Node{p}\in{G}$ if $\Node{p}$ is one of the nodes of $G$.
+Each $\Node{p}\in\Gen$ has a creation-time, $T(\Node{p})$, a parent, $\Anc(\Node{p})$, and a deme.
+We let $\event(G)$ denote the sequence of node times:
+$\event(G)=\left(\Node{p}_k(G)\right)_{k=1}^{K(G)}$, where $0=T(\Node{p}_1(G)){\le}T(\Node{p}_2(G)){\le}\dots{\le}T(\Node{p}_{K(G)}(G)){\le}T(G)$.
 %% The nodes are ordered so that $k<k'$ implies that $T(\Node{p}_k(\Gen))\le{T(\Node{p}_{k'}(\Gen))}$.
@@
 $\Node{p}$ is a root if and only if $\Anc(\Node{p})=\Node{p}$.
-Every node also has one or more descendants called \emph{children}.
-There are three kinds of children:
-\begin{inparaenum}[(i)]
-\item \emph{Internal nodes}.
-  Children nodes must always be later than their parents:
-  $T(\Anc(\Node{p}))<T(\Node{p})$.
-\item \emph{Tip nodes}, which represent extant lineages.
-  In particular, every lineage alive at time $t$ is the child of some node.
-\item \emph{Sample nodes}.
-  When a sample is collected at time $t$, a new node, $\Node{p}$ is added with $T(\Node{p})=t$ and the sample as child.
-\end{inparaenum}
+Every node also has one or more descendant nodes called \emph{children}.
 
@@
 
-The population process induces a stochastic process on the space of genealogies.
+The population and inventory processes together induce a stochastic process, $\Gen_t$, on the space of genealogies.
 In particular, at each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event:
@@
 \item A birth-type event at time $t$ results in the creation of one new internal node for each parent lineage.
-  In particular, if $j$ is one of the parent lineages, and $\Node{p}$ is the new node, then $T(\Node{p})=t$, $\Anc(\Node{p})$ is the node in which $j$ lay prior to the event.
-  The children of $\Node{p}$ include all the new lineages that branched from $j$, as well as $j$ itself.
+  In particular, if $j$ is one of the parent lineages, and $\Node{p}$ is the new node, then $T(\Node{p})=t$, $\Anc(\Node{p})$ is the node that was parent to $j$ prior to the event.
+  The children of $\Node{p}$ include one new tip node for each new lineage sired by $j$, as well as $j$ itself.
 \item In a death-type event, all the lineages $j$ that die are removed.
-  Nodes without children are then recursively removed.
+  Internal nodes without children are then recursively removed.
+  Sample nodes are never removed.
 \item In a migration-type event, one node is added for each migrating lineage;
   each one takes the migrating lineage as child.
-  The ancestor of the new node is the node in which the migrating lineage lay before the event.
-\item At a sample-type event, one new node is introduced for each sampled lineage.
-  Each one takes the sampled lineage as child, along with the sample node.
-  The ancestor of the node is that in which the sampled lineage lay before the vent.
+  The ancestor of the new node is that which was parent to the migrating lineage prior to the event.
+  The deme of the lineage changes accordingly.
+\item At a sample-type event, one new sample node is introduced for each sampled lineage.
+  Each one takes the sampled lineage as child.
+  The ancestor of the sample node is that which was parent to the sampled lineage before the vent.
 \item At a neutral-type event, no change is made to the genealogy.
-\item Finally, events of compound type are readily accommodated by combining the rules just stated.
+\item Finally, events of compound type are accommodated by combining the foregoing rules.
 \end{enumerate}
+When an event results in the addition of one or more new nodes to a genealogy, the lineages which are children of that node are said to \emph{emerge} from the event.
+Thus, after a birth-type event, the emerging lineages include all the new offspring as well as their parents.
+Likewise, at pure migration- or sample-type events, each migrating or sampled lineage emerges from the event.
+At pure death-type events, no lineages emerge.
+In general, at an event of mark $u$, there are $r^u_i$ emergent lineages in deme $i$.
 
@@
 
-\paragraph{Lineages, event-counter, deme-residence}
+Although the process just described yields a genealogy that relates all extant members of the population, and all samples, the data we ultimately wish to analyze will contain only the samples.
+We therefore describe how the genealogy process is \emph{pruned} to yield the sample-only genealogy.
+Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Prune_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every internal node without children.
+In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and some of the interior nodes of the genealogy.
+Observe that the pruned genealogy retains information not only about how much ancestry is shared by any pair of sample lineages, but also about where among the demes each lineage was through time.
+Note also that $T(\Prune_t)=T(\Gen_t)=t$.
 
-Let $\Lin=\Lin(\Gen)$ denote the finite set of all samples represented in genealogy $\Gen$.
-Let $\prec2$ be any ordering of $\Lin$ that is compatible with ancestry.
-That is, if $j,j'\in\Lin$ are such that $j$ is ancestral to $j'$, then we must have $j\prec2{j'}$.
+\paragraph{Alternative representation of a pruned genealogy}
+
+Consider the finite set of all samples represented in pruned genealogy $P$.
+Beginning with 1, assign natural numbers to these in such a way that if sample $j$ is ancestral to sample $j'$, then $j<j'$.
+For example, we can order the samples by their times, resolving any ties arbitrarily.
+Let $\Lin(P)$ denote the set of lineage numbers.
 Using this ordering, we can uniquely associate each point on a genealogical tree with the least of those lineages that descend from that point.
-In particular, any lineage $j\in\Lin$, corresponding to a sample taken at time $t^s_j$, can be traced backward from node to node until either it coalesces with some lesser lineage at some time $t^o_j>0$ or a root is reached (in which case, we define $t^o_j=0$).
+In particular, any lineage $j\in\Lin(P)$, corresponding to a sample taken at time $t^s_j$, can be traced backward from node to node until either it coalesces with some lesser (\ie older) lineage at some time $t^o_j>0$ or a root is reached (in which case, we define $t^o_j=0$).
 Each node encountered along the way represents a genealogical event from which $j$ emerges.
@@
 However, for $t\notin\halfopen{t^o_j,t^s_j}$, lineage $j$ does not exist.
-To express this, we again say that lineage $j$ is in the \emph{underdeme}, which we denote using the symbol $\eth$.
-We define $\Demesplus\coloneq\Demes\cup\{\eth\}$.
+To express this, we again say that lineage $j$ is in the underdeme, $\eth$.
 
-It will be useful to define a function that captures all the relevant features of a pruned genealogy.
-Accordingly, let $\Yspace=\Zp\times\Demesplus\times\Lin$ and define $y:\Rp\times\Lin\to\Yspace$ so that, for $j\in\Lin$ and $t\in\Rp$:
+It will be useful to distill the elements that characterize a pruned genealogy.
+Accordingly, given a pruned genealogy, $P$, we make the following definitions.
 \begin{compactenum}[(a)]
-\item $\ct(y_j(t))\in\Zp$ is a counting process which increases by 1 at each event along lineage $j$.
-\item $\deme(y_j(t))\in\Demesplus$ indicates in which deme lineage $j$ lies at time $t$.
-  In particular $\deme(y_j(t))=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
-\item $\anc(y_j(t))\in\Lin$ indicates the lineage in which the ancestors of lineage $j$ are found.
-  In particular, $\anc(y_j(t))=j$ for $t\ge{t^o_j}$, but $\anc(y_j(t))$ is well defined for all $j$ and $t$.
-  \AAK{[Can define this in a way similar to that of the original Kingman's coalescent\dots.]}
+\item Let $\ct^{P}:\Zp\times\Rp\to\Zp$ be such that, for $j\in\Zp$, $\ct^{P}_j(t)$ is the counting process which increases by 1 at each event along lineage $j$.
+\item Let $\anc^{P}:\Zp\times\Rp\to\Zp$ be such that $\anc^{P}_j(t)$ indicates the unique lineage, ancestral of lineage $j$, and alive at time $t$.
+  In particular, we posit that $\anc^{P}_j(t)=j$ for $t\in\halfopen{t^o_j,t^s_j}$ and $\anc^{P}_j(t)=0$ for $t>t^s_j$, so that the function is well defined for all $j$ and $t$.
+  %%  Alternatively, one can define $\anc$ using the original construction of Kingman's coalescent \citep{Kingman1982a}.
+\item Let $\deme^{P}:\Zp\times\Rp\to\Demesplus$ be such that, for $j\in\Zp$, $\deme^{P}_j(t)$ indicates in which deme lineage $j$ lies at time $t$.
+  In particular $\deme^{P}_j(t)=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
+\item Let $\Yspace=(\Zp\times\Zp\times\Demesplus)^{\Zp}$ and define $Y^{P}:\Rp\to\Yspace$ such that, for $j\in\Zp$ and $t\in\Rp$, $Y^{P}_j(t)=\left(\ct^{P}_j(t),\anc^{P}_j(t),\deme^{P}_j(t)\right)$.
 \end{compactenum}
-One can verify that $y$ is \cadlag.
-Define $\yt(t)=\lim_{t'\uparrow{t}}y(t')$.
+For $j>|\Lin(P)|$, we adopt the convention that $t^o_j=t^s_j=\infty$, so that $\ct^P_j(t)=0$, $\anc^{P}_j(t)=0$, and $\deme^P_j(t)=\eth$ for all $t\in\Rp$.
+It is easy to see that $Y^{P}$ is well defined, piecewise constant, and \cadlag, and that the map $P\mapsto{Y^{P}}$ is one-to-one.
+However, not every piecewise constant, \cadlag\ map $Y:\Rp\to\Yspace$ defines a pruned genealogy.
 
-To visualize these functions, one can make a correspondence between demes and colors.
+Given $\eta\in\Yspace$, we will use the notation $\ct(\eta)$, $\anc(\eta)$, and $\deme(\eta)$ to refer to the coordinates of $\eta$.
+
+Since $\Gen_t$ is a stochastic process, both $\Prune_t\colonequals\prune(G_t)$ and $Y^{\Prune_t}$ are stochastic processes as well.
+In fact, the latter two are Markovian, since each contains within itself all of its past history.
+The process $\Gen_t$ is not Markovian, though $(\X_t,\Gen_t)$ is.
+
+To visualize $\Prune_t$, one can make a correspondence between demes and colors.
 Then a pruned genealogy is visualized as a tree with colored branches.
-Knowing the function $\deme(y)$ is equivalent to knowing the coloring, while $\ct(y)$ determines the locations of events in the genealogy and $\anc(y)$ determines the topology.
-Note in particular that $y_j(t)\ne\yt_j(t)$ if and only if $t$ is the time of an event from which lineage $j$ emerges.
+Knowing the function $\deme^{\Prune_t}$ is equivalent to knowing the coloring, while $\ct^{\Prune_t}$ determines the locations of events in the genealogy and $\anc^{\Prune_t}$ determines the topology.
+Note in particular that, if $y=Y^{\Prune_t}$, then $y_j(t')\ne\yt_j(t')$ if and only if $t'$ is the time of an event from which lineage $j$ emerges.
+Note also that, there can be events on a pruned genealogy where the color does not change.
+That is, where $\ct_j$ increments, but neither $\anc_j$ nor $\deme_j$ change.
 
@@
 In the following, we will find that we need to count the deme-specific numbers of lineages present at a given time.
-Accordingly, for any $\Lin'\subseteq{\Lin}$, $\eta\in(\Zp\times\Demesplus\times\Lin')^{\Lin'}$, and $i\in\Demes$, let us define
+Accordingly, for any $\eta\in\Yspace$, and $i\in\Demes$, let us define
 \begin{equation*}
-  \ell_i(\eta)\coloneq{\left\vert\left\{j\in\Lin'\;\vert\;\deme(\eta_j)=i\right\}\right\vert}\in\Zp, \qquad \ell(\eta)\coloneq(\ell_i(\eta))_{i\in\Demes}\in{\Zp^\Demes}.
+  \ell_i(\eta)\colonequals{\left\vert\left\{j\in\Zp\;\vert\;\deme(\eta_j)=i\right\}\right\vert}\in\Zp, \qquad \ell(\eta)\colonequals(\ell_i(\eta))_{i\in\Demes}\in{\Zp^\Demes}.
 \end{equation*}
-Note that lineages $j$ for which $\eta_j=\eth$ are not counted.
-With this definition, it follows that $\ell_i(y(t))$ is the number of lineages in deme $i$ at time $t$.
+Note that lineages $j$ for which $\deme(\eta_j)=\eth$ are not counted.
+With this definition, it follows that, for any pruned genealogy $P$, $\ell_i(Y^{P}(t))$ is the number of lineages in deme $i$ at time $t$ and
+$\ell(Y^{P}(t))\in\Zp^\Demes$ is the non-negative integer vector telling how many lineages lie within each of the demes at time $t$.
 
 We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
-Accordingly, for $\Lin'\subseteq{\Lin}$, $\eta,\eta'\in(\Zp\times\Demesplus\times\Lin')^{\Lin'}$, and $i\in\Demes$, let us define
+Therefore, for $\eta,\eta'\in\Yspace$, and $i\in\Demes$, let us define
 \begin{equation*}
-  s_i(\eta,\eta')\coloneq{\left\vert\left\{j\in{\Lin'}\;\vert\;\deme(\eta_j)=i\ \ \&\ \ \ct(\eta'_j)=\ct(\eta_j)+1\right\}\right\vert}, \qquad s(\eta,\eta')\coloneq(s_i(\eta,\eta'))_{i\in\Demes}\in{\Zp^\Demes}.
+  s_i(\eta,\eta')\colonequals{\left\vert\left\{j\in{\Zp}\;\vert\;\deme(\eta_j)=i\ \ \&\ \ \ct(\eta'_j)\ne\ct(\eta_j)\right\}\right\vert}, \qquad s(\eta,\eta')\colonequals(s_i(\eta,\eta'))_{i\in\Demes}\in{\Zp^\Demes}.
 \end{equation*}
-With this definition, $s_i(\yt(t),y(t))$ is the number of lineages in deme $i$ that emerge from an event at time $t$ and that, if $t$ is not an event time, then $s(\yt(t),y(t))=0$.
+With this definition, for any $P$, if $y=Y^P$, then $s_i(\yt(t),y(t))$ is the number of lineages in deme $i$ that emerge from an event at time $t$ and $s(\yt(t),y(t))$ is the non-negative integer vector telling how many lineages in each deme emerge at time $t$.
+In particular, $s(\eta,\eta)=0$ so that if $t\notin\event(P)$, then $s(\yt(t),y(t))=0$.
+We refer to $s\left(\yt(t),y(t)\right)$ as the \emph{saturation} at time $t$.
 
-\subsection{Obscuration}
+\subsection{Obscured genealogy}
+
+As we have just seen, a pruned genealogy contains information about the full history of each sample lineage, including the times at which it entered or exited any deme, sired offspring, or was sampled.
+The data we seek to analyze will typically lack much of this information.
+Accordingly, we define the \emph{obscured genealogy} to be that obtained by discarding all information about demes and events not visible from the topology of the tree alone.
+In particular, if $\Prune$ is a pruned genealogy and $\Vis=\obs(\Prune)$ is the corresponding obscured genealogy,
+$\Vis$ is uniquely determined by the function $t\mapsto\anc(Y^{\Prune}(t))$.
+
+\paragraph{Binomial ratio}
+
+For $n,r,\ell,n\in{\Zp^\Demes}$, define the \emph{binomial ratio}
+\begin{equation*}
+  \BinRatio{n}{\ell}{r}{s}\colonequals
+  \begin{cases}
+    \frac{\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}{\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
+    & \forall i\ n_i\ge{\{\ell_i,r_i\}}\ge{s_i}\ge{0}\\
+    0, & \text{otherwise}
+  \end{cases}
+\end{equation*}
+Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
+Moreover, in consequence of the Chu-Vandermonde identity, we have
+\begin{equation*}
+  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
+\end{equation*}
+whenever $n_i\ge{\{\ell_i,r_i\}}\ge{0}$ for all $i$.
 
@@
 
-\begin{defn}
-  For $n,r,\ell,n\in{\Zp^\Demes}$, define the multivariable binomial coefficient by
-  \begin{equation*}
-    \binom{n}{r}\coloneq\prod_{i\in\Demes}\binom{n_i}{r_i}.
-  \end{equation*}
-  Using this, we define the \emph{binomial ratio}
-  \begin{equation*}
-    \BoxProb{n}{\ell}{r}{s}\coloneq\frac{\binom{n-\ell}{r-s}}{\binom{n}{r}}\in{[0,1]}.
-  \end{equation*}
-\end{defn}
+\paragraph{Likelihood conditional on history}
 
-In consequence of the Chu-Vandermonde identity, we have
-\begin{equation}
-  \sum_{s\in\Zp^{\Demes}}\BoxProb{n}{\ell}{r}{s}\binom{\ell}{s}=1
-\end{equation}
-
-\paragraph{Likelihood of a pruned genealogy}
-
-Now certain population-process events at certain times are incompatible with any given pruned genealogy.
-Example: a sample event where no sample is seen, or a migration event involving a change of deme not observed in the pruned genealogy.
-Define the function $Q_u(\eta,\eta')=1$ if a change $\eta\to\eta'$ at time $t$ is compatible with an event of type $u$ at that time, and $Q_u=0$ otherwise.
-In other words, $Q_u$ is the indicator function for the condition that there exist $\omega\in\Omega$, $t\in\Rp$, and $k\in\Zp$ such that $t=T_k(\omega)$, $u=U_k(\omega)$, $\yt(t,\omega)=\eta$, and $y(t,\omega)=\eta'$.
+Our first result will be an expression for the likelihood of a given pruned genealogy $P^*$ given the history $\Hist_t$ of the population process up to time $t=T(P^*)$.
+Of course, not every pruned genealogy is compatible with $\Hist_t$.
+In particular, $P^*$ is only compatible with $\Hist_t$ if every event in $P^*$ coincides with an event in $\Hist_t$.
+That is, only if $\event(P^*)\subseteq{\event(\Hist_t)}$.
+Moreover, certain events in $P^*$ may be incompatible with all possible histories.
+For example, if $P^*$ has an event in which a lineage moves from deme $i$ to deme $i'$ but there are no $u\in\Jumps$ for which this is possible, then $P^*$ is incompatible with the population process itself.
+Let us define the function $Q:\Jumps\times\Yspace\times\Yspace\to\{0,1\}$ so that $Q_u(\eta,\eta')=1$ if and only if a change $\eta\to\eta'$ in a pruned genealogy is compatible with an event of mark $u$ at that time.
+More formally, $Q_u(\eta,\eta')=1$ if and only if there are $\omega\in\Omega$ and $t'<t\in\Rp$ such that $\U_{t'}(\omega)=u$, $\leftlim{Y}^{\prune(\Gen_t(\omega))}(t')=\eta$, and $Y^{\prune(\Gen_t(\omega))}(t')=\eta'$.
 Using this, we define
 \begin{equation*}
-  \phi_u(\xi,\eta,\eta')\coloneq\BoxProb{n(\xi)}{\ell(\eta')}{r_u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
+  \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
 \end{equation*}
+for $\xi\in\Xspace$, $\eta,\eta'\in\Yspace$, and $u\in\Jumps$.
+Recall that $n:\Xspace\to\Zp$ is the deme-occupancy function, $r^u$ is the production of a jump of mark $u$, $\ell$ is the lineage-count function, and $s$ is the saturation.
 
-\begin{thm}
+\begin{thm}\label{thm:pruned_lik}
+  Let $P^*$ be a given pruned genealogy, $t=T(P^*)$, and $y^*=Y^{P^*}$.
+  Then
   \begin{equation*}
-    \Prob{\Prune_t\vert\Hist_t}=\prod_{k=1}^{K}{\phi_u\left(\X_{t_k},\yt(t_k),y(t_k)\right)}=\prod_{k=1}^{K}{\sum_{y_k}{\pi(y_{k-1},y_k)\,\frac{\phi_u\left(\X_{t_k},y_{k-1},y_k\right)}{\pi(y_{k-1},y_k)}\,\kappa^{\Prune_t}(t_k,y_{k-1},y_k)}},
+    \Prob{P^*\vert\Hist_t}=\Indicator{\event(P^*)\subseteq{\event(\Hist_t)}}\,\times\,\prod_{e\in\event(\Hist_t)}{\phi_{U_e}\Big(X_{e},\yt^*(e),y^*(e)\Big)}.
   \end{equation*}
-  where $\kappa^{\Prune_t}(t,y,y')=\Indicator{\yt^{\Prune_t}(t)=y, y^{\Prune_t}(t)=y'}$.
-  \AAK{[Note that we can just sum over $\Prune_t$ and interchange sums and products (using conditional independence) to get Thm 2!]}
 \end{thm}
 \begin{proof}
-  Two approaches possible:
-  \begin{enumerate}
-  \item Conditional on $\Hist_t$, the likelihood of $\Prune_t$ is the product of $p_{jk}$, where $j$ ranges over lineages and $k$ over events.
-    Interchanging the order of integration yields the expression given.
-  \item Use the filter equations developed in the Appendix.
-  \end{enumerate}
+  As we have already observed, if $\event(P^*)\nsubseteq\event(\Hist_t)$, then $\Prob{P^*}=0$.
+  Similarly, if there is any event of $P^*$ which is incompatible with the population process, $\Prob{P^*}=0$.
+  Let us therefore suppose that neither of these conditions hold.
+  Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
+  Moreover, at each event $e\in\event(\Hist_t)$, a jump of mark $\U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
+  In $P^*$, at time $e$, there are $\ell_i=\ell_i(y^*(t))$ lineages in deme $i$, of which $s_i=s_i(\yt^*(t),y^*(t))$ are emergent.
+  The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
+  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, one ends up with $s_i$ black balls, namely
+  \begin{equation*}
+    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
+  \end{equation*}
+  Since our genealogies are labelled, we are interested in the probability that the specific lineages are involved.
+  We therefore divide by $\binom{\ell_i}{s_i}$.
+  Since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
+  \begin{equation*}
+    \Prob{P^*\vert\Hist_t}=\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}}.
+  \end{equation*}
+  Returning to the possibility that $P^*$ is incompatible with $\Hist_t$, since $\Prob{P^*}=0$ if either any $Q_u=0$ or of any event of $P^*$ is not an event of $\Hist_t$, we obtain the result.
 \end{proof}
 
-\begin{figure}
-  \begin{center}
-    \input{figs/thm1diag}
-  \end{center}
-  \caption{\label{fig:thm1diag}}
-\end{figure}
+Our second result concerns the likelihood of a given obscured genealogy conditional on the history.
 
-\begin{center}
-  \input{figs/triangle1}
-\end{center}
+\begin{thm}\label{thm:obsc_lik}
+  Let $V^*$ be a given obscured genealogy and $t=T(V^*)$.
+  Let $\pi:\Jumps\times\Rp\times\Xspace^2\times\Yspace^2\to\Rp$ be a probability kernel, \ie $\pi_u(t,\xi,\xi',\eta,\eta')\ge{0}$ and $\sum_{\eta'}{\pi_u(t,\xi,\xi',\eta,\eta')}=1$ for all $u\in\Jumps$, $t\in\Rp$, $\xi,\xi'\in\Xspace$, and $\eta\in\Yspace$.
+  Suppose moreover that $\pi_u(t,\xi,\xi',\eta,\eta')>0$ for all $t,\xi,\xi'$ whenever $Q_u(\eta,\eta')=1$.
+  Let $\rdm{y}_t$ be the Markov jump process generated by the kernel $\pi$, with jump times fixed at $\event(\Hist_t)$.
+  Then
+  \begin{equation*}
+    \Prob{V^*\vert\Hist_t}=\Expect{\prod_{e\in\event(\Hist_t)}{\frac{\phi_{U_e}(X_e,\leftlim{\rdm{y}}_t,\rdm{y}_t)}{\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}},
+  \end{equation*}
+  the expectation being taken over $\rdm{y}_t$.
+\end{thm}
+\begin{proof}
+  First, observe that, since $\obs$ is a deterministic operator,
+  \begin{equation}\label{eq:IS1}
+    \Prob{V^*\vert\Hist_t}=\int{\Indicator{\obs(P)=V^*}\,\Prob{P\vert\Hist_t}\,\dd{P}},
+  \end{equation}
+  the integral being taken over all pruned genealogies.
+  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
+  we will propose pruned genealogies compatible with $V^*$ as sample paths from a Markov process on $\Yspace$ and
+  treat the integral in \cref{eq:IS1} as an expectation over these paths.
+  Conditional on $\Hist_t$, the probability kernel $\pi$ generates a Markov chain, $\rdm{\hat{y}}_k$, on $\Yspace$ such that
+  \begin{equation*}
+    \Prob{\rdm{\hat{y}}_k=\hat{y}_k\vert\rdm{\hat{y}}_{k-1}=\hat{y}_{k-1}}=\pi_{U_{t_k}}(t_k,\leftlim{X}_{t_k},X_{t_k},\hat{y}_{k-1},\hat{y}_k).
+  \end{equation*}
+  $\rdm{\hat{y}}_k$ is the embedded chain of a unique, piecewise-constant, \cadlag\ process $\rdm{y}_t$, with jump times at $t_k$.
+  The conditions on $\pi$ guarantee that for every $P$ such that $\obs(P)=V^*$, then $Y^P$ is a sample path of this process and, conversely, that for every sample path $y$, there is a pruned genealogy $P=P(y)$ such that $\obs(P)=V^*$ and $Y^P(t)=y_t$.
 
+  We can therefore write
   \begin{equation*}
-  \frac{r!}{(r-s)!}\cdot\frac{(n-\ell)!}{n!}\cdot\frac{(n-r)!}{(n-r-\ell+s)!}
-  =\frac{r!(n-r)!}{n!}\cdot\frac{(n-\ell)!}{(n-\ell-r+s)!(r-s)!}
-  =\BoxProb{n}{\ell}{r}{s}
+    \Prob{V^*\vert\Hist_t}=\Expect{\frac{\Prob{P(\rdm{y})\vert\Hist_t}}{\pi(\rdm{y}\vert\Hist_t)}},
   \end{equation*}
+  the expectation being taken with respect to the random process $\rdm{y}_t$.
+  Here, by definition,
+  \begin{equation*}
+    \pi(\rdm{y}\vert\Hist_t)=\prod_{e\in\event(\Hist_t)}{\pi_{U_{e}}(e,\leftlim{X}_{e},X_{e},\leftlim{\rdm{y}}_{e},\rdm{y}_{e})}.
+  \end{equation*}
+  The result follows from \cref{thm:pruned_lik}.
+\end{proof}
+
+\AAK{Two approaches possible:}
+\begin{enumerate}
+\item Use the filter equations developed in the Appendix to compute $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}$.
+  In effect, we compute $\Prob{\Prune_t}$ and then sum over paintings $y$.
+\item Change the order of the summation: $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}=\Expect{\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}\;\vert\;\Hist_t}}$.
+  That is, conditional on $\Hist_t$, the likelihood of $\Vis_t$ is sum over all $\Prune_t$ such that $\obs(\Prune_t)=\Vis_t$.
+  Therefore, any importance sampling scheme for $\Prune_t$ will do.
+  In particular, we choose a scheme that paints the tree forward in time, driven by a mimic of $\X_t$.
+  We show that this is equivalent to the specific filter equation.
+
+  It may be useful in this to first show that $\Prob{\Prune_t|\Hist_t}$ can be expressed as an expectation over $y$, where $y$ is constrained to be equal $y^{\Prune_t}$.
+  Then, show that the sum of the $\Prune_t$-constraint indicator functions is equal to the $\Vis_t$-constraint indicator function.
+\end{enumerate}
+
+%% \begin{figure}
+%%   \begin{center}
+%%     \input{figs/thm1diag}
+%%   \end{center}
+%%   \caption{\label{fig:thm1diag}}
+%% \end{figure}
+
+%% \begin{center}
+%%   \input{figs/triangle1}
+%% \end{center}
+
+%% \begin{equation*}
+%%   \frac{r!}{(r-s)!}\cdot\frac{(n-\ell)!}{n!}\cdot\frac{(n-r)!}{(n-r-\ell+s)!}
+%%   =\frac{r!(n-r)!}{n!}\cdot\frac{(n-\ell)!}{(n-\ell-r+s)!(r-s)!}
+%%   =\BinRatio{n}{\ell}{r}{s}
+%% \end{equation*}
+
+\paragraph{Integrating out the history, filter equations}
 
@@
 \end{equation}
-
-\paragraph{Likelihood of an obscured genealogy}
-
-\begin{thm}
-  State the filter equation definition.
-\end{thm}
-\begin{proof}
-  Two approaches possible:
-  \begin{enumerate}
-  \item Use the filter equations developed in the Appendix to compute $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}$.
-    In effect, we compute $\Prob{\Prune_t}$ and then sum over paintings $y$.
-  \item Change the order of the summation: $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}=\Expect{\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}\;\vert\;\Hist_t}}$.
-    That is, conditional on $\Hist_t$, the likelihood of $\Vis_t$ is sum over all $\Prune_t$ such that $\obs(\Prune_t)=\Vis_t$.
-    Therefore, any importance sampling scheme for $\Prune_t$ will do.
-    In particular, we choose a scheme that paints the tree forward in time, driven by a mimic of $\X_t$.
-    We show that this is equivalent to the specific filter equation.
-
-    It may be useful in this to first show that $\Prob{\Prune_t|\Hist_t}$ can be expressed as an expectation over $y$, where $y$ is constrained to be equal $y^{\Prune_t}$.
-    Then, show that the sum of the $\Prune_t$-constraint indicator functions is equal to the $\Vis_t$-constraint indicator function.
-  \end{enumerate}
-\end{proof}
-
-Let $\kappa^{\Vis}$ be the indicator function for the condition that a pruned genealogy is compatible with a given obscured genealogy.
-In particular, given an obscured genealogy $\Vis$, set $\kappa^{\Vis}_u(t,x,x',\eta,\eta')=\Indicator{\exists\Prune\ \st\ \kappa^{\Prune}_u(t,x,x',y,y')=1\ \&\ \obs(\Prune)=\Vis}$.
 
@@
   $H_{t}(\omega)(t')=X_{t'}(\omega)$.
-  Moreover, there are random variables $K\in\Zp$ and $t_k\in\halfclosed{0,t}$, $k=1,\dots,K$ such that $\predec{X}_t=X_t$ whenever $t\ne{t_k}$ for all $k$.
+  Moreover, there are random variables $K\in\Zp$ and $t_k\in\halfclosed{0,t}$, $k=1,\dots,K$ such that $\leftlim{X}_t=X_t$ whenever $t\ne{t_k}$ for all $k$.
   Suppose $F$ is a real-valued function of $H_t$ such that
   \begin{equation*}
-    F(H_t)=\prod_{k}{Q\Big(t_k,\predec{H}(t_k),H(t_k)\Big)\,B\Big(t_k,\predec{H}(t_{k}),H(t_k)\Big)},
+    F(H_t)=\prod_{k}{Q\Big(t_k,\leftlim{H}(t_k),H(t_k)\Big)\,B\Big(t_k,\leftlim{H}(t_{k}),H(t_k)\Big)},
   \end{equation*}

### 2023-07-13 ms.Rnw
small improvement to thm 2

@@
 More formally, $Q_u(\eta,\eta')=1$ if and only if there are $\omega\in\Omega$ and $t'<t\in\Rp$ such that $\U_{t'}(\omega)=u$, $\leftlim{Y}^{\prune(\Gen_t(\omega))}(t')=\eta$, and $Y^{\prune(\Gen_t(\omega))}(t')=\eta'$.
-Using this, we define
-\begin{equation*}
-  \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
-\end{equation*}
-for $\xi\in\Xspace$, $\eta,\eta'\in\Yspace$, and $u\in\Jumps$.
-Recall that $n:\Xspace\to\Zp$ is the deme-occupancy function, $r^u$ is the production of a jump of mark $u$, $\ell$ is the lineage-count function, and $s$ is the saturation.
+Recall that $n:\Xspace\to\Zp$ denotes the deme-occupancy function, $r^u$, the production of a jump of mark $u$, $\ell$, the lineage-count function, and $s$, the saturation.
 
@@
   \begin{equation*}
-    \Prob{P^*\vert\Hist_t}=\Indicator{\event(P^*)\subseteq{\event(\Hist_t)}}\,\times\,\prod_{e\in\event(\Hist_t)}{\phi_{U_e}\Big(X_{e},\yt^*(e),y^*(e)\Big)}.
+    \Prob{P^*\vert\Hist_t}=\Indicator{\event(P^*)\subseteq{\event(\Hist_t)}}\,\times\,\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y(e))}{r^{U_e}}{s(\yt(e),y(e))}\,Q_{U_e}(\yt(e),y(e))}.
   \end{equation*}
@@
 
+\begin{equation*}
+  \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
+\end{equation*}
+
 Our second result concerns the likelihood of a given obscured genealogy conditional on the history.
@@
   Let $\pi:\Jumps\times\Rp\times\Xspace^2\times\Yspace^2\to\Rp$ be a probability kernel, \ie $\pi_u(t,\xi,\xi',\eta,\eta')\ge{0}$ and $\sum_{\eta'}{\pi_u(t,\xi,\xi',\eta,\eta')}=1$ for all $u\in\Jumps$, $t\in\Rp$, $\xi,\xi'\in\Xspace$, and $\eta\in\Yspace$.
-  Suppose moreover that $\pi_u(t,\xi,\xi',\eta,\eta')>0$ for all $t,\xi,\xi'$ whenever $Q_u(\eta,\eta')=1$.
+  Suppose moreover that $\pi_u(t,\xi,\xi',\eta,\eta')>0$ if and only if $Q_u(\eta,\eta')=1$.
   Let $\rdm{y}_t$ be the Markov jump process generated by the kernel $\pi$, with jump times fixed at $\event(\Hist_t)$.
@@
   \begin{equation*}
-    \Prob{V^*\vert\Hist_t}=\Expect{\prod_{e\in\event(\Hist_t)}{\frac{\phi_{U_e}(X_e,\leftlim{\rdm{y}}_t,\rdm{y}_t)}{\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}},
+    \Prob{V^*\vert\Hist_t}=\Expect{\prod_{e\in\event(\Hist_t)}{\frac{\scriptsize\BinRatio{n(X_e)}{\ell(\rdm{y}_e)}{r^{U_e}}{s(\leftlim{\rdm{y}}_t,\rdm{y}_t)}}{\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}},
   \end{equation*}
@@
   The conditions on $\pi$ guarantee that for every $P$ such that $\obs(P)=V^*$, then $Y^P$ is a sample path of this process and, conversely, that for every sample path $y$, there is a pruned genealogy $P=P(y)$ such that $\obs(P)=V^*$ and $Y^P(t)=y_t$.
+  In particular, they guarantee that $\pi_u(\xi,\xi',\eta,\eta')\propto{Q_u(\eta,\eta')}$.
 

### 2023-07-17 ms.Rnw
progress

@@
 
-\section{From population processes to genealogy processes}
+\section{Mathematical definitions}
 
@@
 \paragraph{Mathematical notation}
+
+We use bold-face to mark random elements and the corresponding regular-face symbol to denote specific values of such elements.
+Given any \cadlag\ process $f_t$, the set of jump-times is denoted $\event(f_t)=\{e\in[0,t]\;\vert\;\leftlim{f}_e\ne{f_e}\}$.
 
@@
 
-\paragraph{Pruning}
+\subsection{The pruned and obscured genealogies}
 
@@
 It is easy to see that $Y^{P}$ is well defined, piecewise constant, and \cadlag, and that the map $P\mapsto{Y^{P}}$ is one-to-one.
-However, not every piecewise constant, \cadlag\ map $Y:\Rp\to\Yspace$ defines a pruned genealogy.
+However, not every piecewise constant, \cadlag\ map $y:\Rp\to\Yspace$ defines a pruned genealogy.
+If $y$ does define a pruned genealogy, we denote that pruned genealogy by $P(y)$.
 
@@
 
-\paragraph{Lineage count, saturation}
+\paragraph{Lineage count and saturation}
 
@@
 
-\subsection{Obscured genealogy}
+\paragraph{Obscured genealogy}
 
@@
   \begin{equation*}
-    \Prob{P^*\vert\Hist_t}=\Indicator{\event(P^*)\subseteq{\event(\Hist_t)}}\,\times\,\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y(e))}{r^{U_e}}{s(\yt(e),y(e))}\,Q_{U_e}(\yt(e),y(e))}.
+    \begin{aligned}
+      \Prob{P^*\vert\Hist_t}=&\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}\,Q_{U_e}(\yt^*(e),y^*(e))}\\
+      &\qquad\times\Indicator{\event(P^*)\subseteq{\event(\Hist_t)}}.
+    \end{aligned}
   \end{equation*}
@@
 \begin{proof}
-  As we have already observed, if $\event(P^*)\nsubseteq\event(\Hist_t)$, then $\Prob{P^*}=0$.
-  Similarly, if there is any event of $P^*$ which is incompatible with the population process, $\Prob{P^*}=0$.
+  As we have already observed, if $\event(P^*)\nsubseteq\event(\Hist_t)$, then $\Prob{P^*\vert\Hist_t}=0$.
+  Similarly, if there is any event of $P^*$ which is incompatible with the population process, $\Prob{P^*\vert\Hist_t}=0$.
   Let us therefore suppose that neither of these conditions hold.
   Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
-  Moreover, at each event $e\in\event(\Hist_t)$, a jump of mark $\U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
+  Moreover, at each event $e\in\event(\Hist_t)$, a jump of mark $U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
   In $P^*$, at time $e$, there are $\ell_i=\ell_i(y^*(t))$ lineages in deme $i$, of which $s_i=s_i(\yt^*(t),y^*(t))$ are emergent.
   The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
-  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, one ends up with $s_i$ black balls, namely
+  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, exactly $s_i$ of the drawn balls are black, namely
   \begin{equation*}
@@
   \end{equation*}
-  Since our genealogies are labelled, we are interested in the probability that the specific lineages are involved.
-  We therefore divide by $\binom{\ell_i}{s_i}$.
-  Since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
+  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $P^*$.
+  Moreover, since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
   \begin{equation*}
@@
 
-\begin{equation*}
-  \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
-\end{equation*}
+%% \begin{equation*}
+%%   \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
+%% \end{equation*}
 
@@
   Let $V^*$ be a given obscured genealogy and $t=T(V^*)$.
-  Let $\pi:\Jumps\times\Rp\times\Xspace^2\times\Yspace^2\to\Rp$ be a probability kernel, \ie $\pi_u(t,\xi,\xi',\eta,\eta')\ge{0}$ and $\sum_{\eta'}{\pi_u(t,\xi,\xi',\eta,\eta')}=1$ for all $u\in\Jumps$, $t\in\Rp$, $\xi,\xi'\in\Xspace$, and $\eta\in\Yspace$.
+  Let $\pi:\Jumps\times\Rp\times\Xspace^2\times\Yspace^2\to\Rp$ be a probability kernel, \ie $\pi_u(t,\xi,\xi',\eta,\eta')\ge{0}$ and $\sum_{\eta'}{\pi_u(t,\xi,\xi',\eta,\eta')}=1$.
   Suppose moreover that $\pi_u(t,\xi,\xi',\eta,\eta')>0$ if and only if $Q_u(\eta,\eta')=1$.
-  Let $\rdm{y}_t$ be the Markov jump process generated by the kernel $\pi$, with jump times fixed at $\event(\Hist_t)$.
+  Let $\rdm{y}_t$ be the Markov jump process on $\Yspace$ generated by the kernel $\pi$, with $\event(\rdm{y})=\event(\Hist_t)$.
   Then
   \begin{equation*}
-    \Prob{V^*\vert\Hist_t}=\Expect{\prod_{e\in\event(\Hist_t)}{\frac{\scriptsize\BinRatio{n(X_e)}{\ell(\rdm{y}_e)}{r^{U_e}}{s(\leftlim{\rdm{y}}_t,\rdm{y}_t)}}{\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}},
+    \Prob{V^*\vert\Hist_t}=\Expect{\left.\prod_{e\in\event(\Hist_t)}{\frac{\scriptsize\BinRatio{n(X_e)}{\ell(\rdm{y}_e)}{r^{U_e}}{s(\leftlim{\rdm{y}}_e,\rdm{y}_e)}}{\scriptsize\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}\times\Indicator{\obs(P(\rdm{y}))=V^*}\;\right\vert\;\Hist_t},
   \end{equation*}
-  the expectation being taken over $\rdm{y}_t$.
+  the expectation being taken over $\rdm{y}$.
 \end{thm}
@@
   \end{equation*}
+  Now let $\event(\Hist_t)=\{t_1,t_2,\dots,t_K\}$.
   $\rdm{\hat{y}}_k$ is the embedded chain of a unique, piecewise-constant, \cadlag\ process $\rdm{y}_t$, with jump times at $t_k$.
-  The conditions on $\pi$ guarantee that for every $P$ such that $\obs(P)=V^*$, then $Y^P$ is a sample path of this process and, conversely, that for every sample path $y$, there is a pruned genealogy $P=P(y)$ such that $\obs(P)=V^*$ and $Y^P(t)=y_t$.
-  In particular, they guarantee that $\pi_u(\xi,\xi',\eta,\eta')\propto{Q_u(\eta,\eta')}$.
+  The conditions on $\pi$ guarantee that, for every pruned genealogy $P$, $Y^P$ is a sample path of this process.
+  In particular, they guarantee that $\pi_u(t,\xi,\xi',\eta,\eta')\;\propto\;{Q_u(\eta,\eta')}$.
+  Clearly, the construction of $\rdm{y}_t$ guarantees that $\event(P(\rdm{y}))\subseteq\event(\Hist_t)$.
 
-  We can therefore write
+  We therefore have that
   \begin{equation*}
-    \Prob{V^*\vert\Hist_t}=\Expect{\frac{\Prob{P(\rdm{y})\vert\Hist_t}}{\pi(\rdm{y}\vert\Hist_t)}},
+    \Prob{V^*\vert\Hist_t}=\Expect{\frac{\Prob{P(\rdm{y})\vert\Hist_t}}{\pi(\rdm{y}\vert\Hist_t)}\cdot\Indicator{\obs(P(\rdm{y}))=V^*}},
   \end{equation*}
@@
   \end{equation*}
-  The result follows from \cref{thm:pruned_lik}.
+  The result then follows from \cref{thm:pruned_lik}.
 \end{proof}
 
-\AAK{Two approaches possible:}
-\begin{enumerate}
-\item Use the filter equations developed in the Appendix to compute $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}$.
-  In effect, we compute $\Prob{\Prune_t}$ and then sum over paintings $y$.
-\item Change the order of the summation: $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}=\Expect{\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}\;\vert\;\Hist_t}}$.
-  That is, conditional on $\Hist_t$, the likelihood of $\Vis_t$ is sum over all $\Prune_t$ such that $\obs(\Prune_t)=\Vis_t$.
-  Therefore, any importance sampling scheme for $\Prune_t$ will do.
-  In particular, we choose a scheme that paints the tree forward in time, driven by a mimic of $\X_t$.
-  We show that this is equivalent to the specific filter equation.
+%% \AAK{Two approaches possible:}
+%% \begin{enumerate}
+%% \item Use the filter equations developed in the Appendix to compute $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}$.
+%%   In effect, we compute $\Prob{\Prune_t}$ and then sum over paintings $y$.
+%% \item Change the order of the summation: $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}=\Expect{\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}\;\vert\;\Hist_t}}$.
+%%   That is, conditional on $\Hist_t$, the likelihood of $\Vis_t$ is sum over all $\Prune_t$ such that $\obs(\Prune_t)=\Vis_t$.
+%%   Therefore, any importance sampling scheme for $\Prune_t$ will do.
+%%   In particular, we choose a scheme that paints the tree forward in time, driven by a mimic of $\X_t$.
+%%   We show that this is equivalent to the specific filter equation.
 
-  It may be useful in this to first show that $\Prob{\Prune_t|\Hist_t}$ can be expressed as an expectation over $y$, where $y$ is constrained to be equal $y^{\Prune_t}$.
-  Then, show that the sum of the $\Prune_t$-constraint indicator functions is equal to the $\Vis_t$-constraint indicator function.
-\end{enumerate}
+%%   It may be useful in this to first show that $\Prob{\Prune_t|\Hist_t}$ can be expressed as an expectation over $y$, where $y$ is constrained to be equal $y^{\Prune_t}$.
+%%   Then, show that the sum of the $\Prune_t$-constraint indicator functions is equal to the $\Vis_t$-constraint indicator function.
+%% \end{enumerate}
 
@@
 
-\paragraph{Integrating out the history, filter equations}
+\paragraph{Integrating out the history: filter equations}
 
@@
 
-\begin{defn}
-  Suppose $X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
+Suppose $\X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
 \begin{equation}\label{eq:kfe2}
@@
 is the \emph{filter equation} with \emph{driver} $X_t$, \emph{boost} $B$, and \emph{decay} $\lambda$.
-\end{defn}
 
 Filter equations afford a convenient means of computing the likelihood of a given sequence of events.
-This is facilitated by the following
+This is facilitated by the following fact.
+
 \begin{lemma}\label{lemma:filtereq}
@@
 
-We recognize in \cref{eq:filterlemma} the KFE of a certain process $(X_t,V_t)$.
+\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(X_t,V_t)$.
 In particular, $X_t$ is the driver with KFE \cref{eq:kfe2}.
-The $V_t$ process has jumps wherever $X_t$ does, such that when $X_t$ jumps from $x$ to $x'$, $V_t$ jumps by the multiplicative factor $B(t,x,x')$.
+The $V_t$ process has jumps wherever $X_t$ does, such that when $X$ jumps at time $t$ from $x$ to $x'$, $V$ jumps by the multiplicative factor $B(t,x,x')>0$.
 Between jumps, $V_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
-If we view $V_t$ as a weight, then \cref{lemma:filtereq} says that the $V_t$-weighted average of $X_t$ evolves according to \cref{eq:filtereq}.
-This motivates the following result, which effectively allows boosts of zero.
+If we view $V_t$ as a weight, then \cref{lemma:filtereq} tells us how the $V_t$-weighted average of $X_t$ evolves in time:
+this average is simply $\int{w(t,x)\,\dd{x}}$.
+
+This motivates the following result, which shows how filter equations allow one to integrate over random histories.
 
 \begin{prop}\label{prop:zeroboost}
-  Suppose $X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE as in \cref{eq:kfe2}.
-  Let $H_t$ be its history process.
-  That is, for $\omega\in\Omega$ and $t\in\Rp$, $H_{t}(\omega):[0,t]\to\Xspace$ such that, for $t'\in{[0,t]}$,
-  $H_{t}(\omega)(t')=X_{t'}(\omega)$.
-  Moreover, there are random variables $K\in\Zp$ and $t_k\in\halfclosed{0,t}$, $k=1,\dots,K$ such that $\leftlim{X}_t=X_t$ whenever $t\ne{t_k}$ for all $k$.
-  Suppose $F$ is a real-valued function of $H_t$ such that
+  Suppose $\X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE as in \cref{eq:kfe2}.
+  %% Suppose that $B(t,x,x')>0$, $Q(t,x,x')\in\{0,1\}$, and $\lambda(t,x)\in\R$ are given measurable functions.
+  %% Consider the equation
+  %% \begin{equation*}
+  %%   \frac{\partial{w}}{\partial{t}}
+  %%   =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,Q(t,x',x)\,\dd{x'}}
+  %%   -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
+  %%   -\lambda(t,x)\,w(t,x)
+  %% \end{equation*}
+  Let $\Hist_t$ be its history process.
+  Suppose $\rdm{R}_t$ is a real-valued random process such that
   \begin{equation*}
-    F(H_t)=\prod_{k}{Q\Big(t_k,\leftlim{H}(t_k),H(t_k)\Big)\,B\Big(t_k,\leftlim{H}(t_{k}),H(t_k)\Big)},
+    \Expect{\rdm{R}_t\;\vert\;\Hist_t}=\prod_{e\in\event(\Hist_t)}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)},
   \end{equation*}
@@
   \end{equation*}
-  Then $\Expect{F(H_t)}=\int{w(t,x)\dd{x}}$.
+  Then $\Expect{\rdm{R}_t}=\int{w(t,x)\dd{x}}$.
 \end{prop}
@@
 \begin{prop}
-  Suppose $X:[0,T]\to\Xspace$ is a deterministic, piecewise constant, \cadlag\ function.
-  Let $E$ be the set of its jump times.
-  Then the KFE for $X_t$ is \cref{eq:kfe2} with $\beta(t,x,x')=\sum_{e\in{E}}{\delta(t,e)\delta(x',X_e)}$.
+  Suppose $X:[0,T]\to\Xspace$ is a deterministic, piecewise constant, \cadlag\ function and let $M=\event(X)$.
+  Then the KFE for $X_t$ is \cref{eq:kfe2} with $\beta(t,x,x')=\sum_{e\in{M}}{\delta(t,e)\delta(x',X_e)}$.
   With this driver, the filter equation (\cref{eq:filtereq}) becomes
@@
       \frac{\partial{w}}{\partial{t}}=
-      -\lambda(t,x)\,w(t,x)\quad\text{for}\quad{t\notin{E}},\\
-      w(e,x)=\delta(x,X_e)\,\int{\wt(e,x')\,B(e,x',X_e)\,\dd{x'}}\quad\text{for}\quad{e\in{E}}.
+      -\lambda(t,x)\,w(t,x)\quad\text{for}\quad{t\in[0,T]\setminus{M}},\\
+      w(t,x)=\delta(x,X_t)\,\int{\wt(t,x')\,B(t,x',X_t)\,\dd{x'}}\quad\text{for}\quad{t\in{M}}.
     \end{gathered}
@@
 \begin{prop}
-  Suppose $\X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE as in \cref{eq:kfe}.
+  Suppose $\X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE
+  \begin{equation*}
+    \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'}.
+  \end{equation*}
   Let $\Hist_t$ be its history process.
-  Suppose $0<t_1<\dots<t_N\le{t}$ are fixed times and set $E=\{t_1,\dots,t_N\}$.
+  Suppose that $0<t_1<\dots<t_N\le{t}$ are fixed times and set $M=\{t_1,\dots,t_N\}$.
   Let $B>{0}$ and $Q\in\{0,1\}$ be given measurable functions.
-  Suppose $F_E$ is an $\Rp$-valued random variable such that
+  Let $\rdm{R}$ be a non-negative real-valued random variable such that
   \begin{equation*}
-    \Expect{F_E\;\vert\;\Hist_t}={\prod_{e\in{E}}{B(e,\Xt_e,\X_e)\,Q(e,\Xt_e,\X_e)}}.
+    \Expect{\rdm{R}\;\vert\;\Hist_t}=\prod_{e\in{\event(\Hist_t)}}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)}.
   \end{equation*}
-  Let $\pi(x,x')$ be a given probability kernel and $Y_t$ be the Markov process generated by $\beta(t,x,x')=\left(\alpha(t,x,x')+\sum_{e\in{E}}{\delta(t,e)\,\pi(x,x')}\right)\,Q(t,x,x')$.
-  Suppose $w$ satisfies the filter equation with driver $Y_t$, boost $B$, and decay $\lambda=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\,\dd{x'}}$.
-  Then $\Expect{F_E}=\int{w(t,x)\,\dd{x}}$.
+  Let $\pi(t,x,x')$ be a given probability kernel and $\Y_t$ be the Markov process generated by \cref{eq:kfe2} where $\beta(t,x,x')=\left[\alpha(t,x,x')+\sum_{e\in{M}}{\delta(t,e)\,\pi(t,x,x')}\right]\,Q(t,x,x')$.
+  Suppose $w$ satisfies the filter equation with driver $\Y_t$, boost
+  \begin{equation*}
+    C(t,x,x')=
+    \begin{cases}
+      B(t,x,x'), &t\notin{M},\\
+      B(t,x,x')\,\displaystyle\frac{\alpha(t,x,x')}{\pi(t,x,x')}, &t\in{M},
+    \end{cases}
+  \end{equation*}
+  and decay $\lambda=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\,\dd{x'}}$.
+  Then
+  \begin{equation*}
+    \Expect{\rdm{R}\;\big\vert\;M\subseteq{\event(\Hist_t)}}=\int{w(t,x)\,\dd{x}}.
+  \end{equation*}
 \end{prop}
@@
 
+\paragraph{More systematically}
+
+Suppose $\X_t$ is a non-explosive pure jump process with KFE
+\begin{equation*}
+  \begin{gathered}
+    \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},\\
+    w(0,x)=p_0(x).
+  \end{gathered}
+\end{equation*}
+We can always take a version of $\X_t$ with \cadlag\ sample paths.
+
+Let $\rdm{\hat{X}}_k$ be the embedded chain of $\X_t$ and let $\rdm{T}^X_k$ be the point process of its jump times.
+It is elementary that
+\begin{equation*}
+  \Prob{\rdm{\hat{X}}_k=\hat{X}_k\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{T}^X_k=T_k}=\frac{\alpha(T_k,\hat{X}_{k-1},\hat{X}_k)}{\int{\alpha(T_k,\hat{X}_{k-1},x')\dd{x'}}}.
+\end{equation*}
+Moreover, if one has the $(\rdm{\hat{X}}_k,\rdm{T}^X_k)$ chain, one can reconstruct the $\X_t$ process.
+Conditional on $\rdm{\hat{X}}$, the jump-time chain $\rdm{T}^X_k$ is a Poisson point process with intensity $\int{\alpha(t,\X_t,x')\dd{x'}}$.
+
+We can rewrite the KFE as
+\begin{equation}\label{eq:poissdriver}
+  \begin{aligned}
+    \frac{\partial{w}}{\partial{t}}(t,n,x)=&\int\!w(t,n-1,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}-\int\!w(t,n,x)\,\alpha(t,x,x')\,\dd{x'}\\
+    =&\int\!w(t,n-1,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\dd{x'}-\int\!w(t,n,x)\,\nu\,\pi(t,x,x')\,\dd{x'}\\
+    &\qquad-\lambda(t,x)\,w(t,n,x)
+  \end{aligned}
+\end{equation}
+where $\nu>0$ is the intensity of homogenous Poisson process and
+\begin{equation*}
+  \begin{gathered}
+    A(t,x)=\int{\alpha(t,x,x')\dd{x'}}, \qquad
+    \pi(t,x,x')=\frac{\alpha(t,x,x')}{A(t,x)},\\
+    B(t,x,x')=\frac{A(t,x)}{\nu}, \qquad
+    \lambda(t,x)=A(t,x)-\nu.
+  \end{gathered}
+\end{equation*}
+Note that $\pi$ is the probability kernel of the embedded chain $\rdm{\hat{X}}$ and $A(t,x)$ is the intensity of the $\rdm{T}^X_k$ process.
+We recognize this equation as the filter equation with the rate-$\nu$ Poisson-process driver, boost $B$, and decay $\lambda$.
+
+By integrating equations of this kind, we effectively integrate over the $(\rdm{\hat{X}},\rdm{T}^X)$ process.
+To marginalize over only some of the $\rdm{T}^X_k$, we reason as follows.
+Fix $0<t_1<\dots<t_K\le{T}$ and $\nu>0$.
+Consider the mixed point process $T_k$ with intensity $\nu+\sum_{k=1}^K\delta(t,t_k)$.
+Using this as the driver in \cref{eq:poissdriver}, we obtain
+\begin{equation*}
+  \begin{aligned}
+    \frac{\partial{w}}{\partial{t}}(t,n,x)=&\int{w(t,n-1,x')\,\nu\,\pi(t,x',x)\,B(t,x',x)\dd{x'}}-\int{w(t,n,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}\\
+    &\quad+\sum_{k}\delta(t,t_k)\,\left\{\int{w(t,n-1,x')\,\pi(t,x',x)\,B(t,x,x')\dd{x'}}-\int{w(t,n,x)\,\pi(t,x,x')\,\dd{x'}}\right\}\\
+    &\qquad-\lambda(t,x)\,w(t,n,x).
+  \end{aligned}
+\end{equation*}
+Equivalently, we write:
+\begin{equation*}
+  \begin{aligned}
+    \frac{\partial{w}}{\partial{t}}(t,n,x)=&\int{w(t,n-1,x')\,\nu\,\pi(t,x',x)\,B(t,x',x)\dd{x'}}-\int{w(t,n,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}\\
+    &\qquad-\lambda(t,x)\,w(t,n,x),\\
+    w(t_k,n,x)=&\int{\leftlim{w}(t_k,n-1,x')\,\pi(t_k,x',x)\,B(t_k,x',x)\,\dd{x'}}\\
+    &\qquad=\tfrac{1}{\nu}\,\int{\leftlim{w}(t_k,n-1,x')\,\alpha(t,x',x)\,\dd{x'}}\\.
+  \end{aligned}
+\end{equation*}
+We then have $\Prob{\X_t\in\mathcal{E}\;\big\vert\;t_1,\dots,t_K\in\event(\Hist_t)}=\int_{\mathcal{E}}{w(t,x)\,\dd{x}}$.
+
+Now suppose $\rdm{R}_k$ is a Markov chain satisfying
+\begin{equation*}
+  \Prob{\rdm{R}_k=R_{k-1}\,B(T_k,\hat{X}_{k-1},\hat{X}_k)\;\Big\vert\;\rdm{\hat{X}}_k=\hat{X}_k,\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{T}^X_k=T_k,\rdm{R}_{k-1}=R_{k-1}}=1,
+\end{equation*}
+for some function $B>0$.
+\AAK{Or perhaps the following is sufficient for our purposes?}
+\begin{equation*}
+  \Expect{\rdm{R}_k\;\Big\vert\;\rdm{\hat{X}}_k=\hat{X}_k,\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{T}^X_k=T_k,\rdm{R}_{k-1}=R_{k-1}}=R_{k-1}\,B(T_k,\hat{X}_{k-1},\hat{X}_k).
+\end{equation*}
+Let $\rdm{R}_t$ be the corresponding continuous-time jump process with jumps at $\rdm{T}^X_k$.
+By \cref{lemma:filtereq}, if $w$ satisfies the boost equation with boost $B$, driver $\X$, and decay $\lambda=0$, then
+\begin{equation*}
+  \Expect{\rdm{R}_t}=\int{w(t,x)\dd{x}}.
+\end{equation*}
+
 \end{document}

### 2023-07-19 ms.Rnw
thm 3, appendix

@@
     \Prob{\N_{t+\Delta}>n+1\;\vert\;\N_{t}=n}=o(\Delta),\\
-    \Prob{\X_{t+\Delta}\in\mathcal{E}\;\vert\;\X_{t}=x, \N_{t+\Delta}-\N_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}+\frac{o(\Delta)}{\Delta}.
+    \Prob{\X_{t+\Delta}\in\mathcal{E}\;\vert\;\X_{t}=x, \N_{t+\Delta}-\N_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
   \end{gathered}
 \end{equation*}
-\AAK{[Do we need the last term?]}
 We will further assume that $\X_t$ is non-explosive, \ie, that $\Prob{\N_t<\infty}=1$ for all $t$.
@@
 The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
-%\begin{equation}\label[pluralequation]{eq:kfe}
 \begin{equation}\label{eq:kfe}
@@
 
-Suppose we have a pruned genealogy $\Prune^*$, defined on the time-interval $[0,T]$, with event times $0=t_0<t_1<\cdots<t_n=T$.
-From the theorem, we have, for $t\notin\event(\Prune^*)$,
+Next, we derive expressions for the likelihood of pruned and obscured genealogies, unconditional on the history.
+For this, we use the filter equation technology developed in the Appendix.
+In particular, the following theorem follows immediately from \cref{prop:fullfilt}.
+
+\begin{thm}
+  Suppose we have a pruned genealogy $P^*_T$, $T>0$, with event times $\event(P^*_T)$.
+  Let $y^*=Y^{\Prune^*_T}$.
+  For $x,x'\in\Xspace$, $y,y'\in\Yspace$, $u\in\Jumps$, define
+  \begin{equation*}
+    \begin{gathered}
+      \phi_u\big(x,y,y')=\BinRatio{n(x)}{\ell(y')}{r^u}{s(y,y')}\qquad
+      \text{and} \qquad
+      \beta_u(t,x,x')=\alpha_u(t,x,x')\,Q_u(y,y'),
+    \end{gathered}
+  \end{equation*}
+  where $Q_u(y,y')$ is as in \cref{thm:pruned_lik}.
+  Fix $\nu>0$ and suppose $w=w(t,x)$ satisfies the filter equation
   \begin{equation}
-  \frac{\partial{w}}{\partial{t}}=\sum_{u}\int\!w(t,x')\,\alpha_u(t,x',x)\,\phi_u\big(x,y(t),y(t)\big)\,\dd{x'}
-  -\sum_{u}\int\!w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}.
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}=
+      &\sum_u{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u\big(x,\yt^*_t,y^*_t\big)\,\dd{x'}}}
+      -\sum_u{\int{w(t,x)\,\beta_u(t,x,x')\,\dd{x'}}}\\
+      &\quad+\sum_{e\in\event(P^*_T)}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\frac{\beta_u(t,x',x)}{\nu}\,\phi_u\big(x,\yt^*_t,y^*_t\big)\,\dd{x'}}}-w(t,x)\right\}},
+    \end{aligned}
   \end{equation}
-At event times, $t\in\event(\Prune^*)$, one has
-\begin{equation}\label{eq:prune_filt_disc}
-  w(t,x)=\sum_{u}\int\!\wt(t,x')\,\frac{\alpha_u(t_k,x',x)}{\mu}\,\phi_u\big(x,\yt(t),y(t)\big)\,\dd{x'}\\,
-\end{equation}
-In addition, there are the initial and boundary conditions
-\begin{equation}
-  \begin{gathered}
-    w(0,x)=p_0(x), \qquad w(t,x)=0\quad\text{whenever}\quad n(x)<\ell(y(t)).
-  \end{gathered}
-\end{equation}
+  with the initial condition $w(0,x)=p_0(x)$.
+  Then the likelihood of $P^*_T$ is
+  \begin{equation*}
+    \lik(P^*_T)=\int{w(T,x)\,\dd{x}}.
+  \end{equation*}
+\end{thm}
 
@@
 \appendix
+\setcounter{thm}{0}
+\setcounter{equation}{0}
+\renewcommand{\theequation}{\thesection\arabic{equation}}
 
@@
 
-Explicit expressions for the probability densities that will arise in the following are not always available.
-Hence, we will develop some technology for manipulating them that avoids the need for explicit expressions.
+Explicit expressions for the quantities that arise in this paper are not always readily available.
+Here, we develop tools for manipulating complex expressions that are otherwise cumbersome.
 
+\begin{defn}
   Suppose $\X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
@@
     =\int{w(t,x')\,\beta(t,x',x)\,\dd{x'}}
-  -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}},
+    -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}.
   \end{equation}
-Suppose that $B(t,x,x')>0$ and $\lambda(t,x)\in\R$ are given functions.
+  Suppose that $B(t,x,x')>0$ and $\lambda(t,x)\in\R$ are given measurable functions.
+  Let $\mathcal{M}\subseteq\Rp$ be countable such that $\mathcal{M}\cap[0,t]$ is finite for all $t$.
+  Let $\nu>0$.
   We say that the equation
   \begin{equation}\label{eq:filtereq}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}
+      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
+      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
+      -\lambda(t,x)\,w(t,x)\\
+      &\quad+\sum_{e\in\mathcal{M}}\left\{\int{w(t,x')\,\frac{\beta(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}.
+    \end{aligned}
+  \end{equation}
+  is the \emph{filter equation} with \emph{driver} $\X_t$, \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$, \emph{relative to} $\nu$.
+\end{defn}
+
+\begin{remark}
+  \Cref{eq:filtereq} is equivalent to
+  \begin{equation*}
+    \begin{gathered}
       \frac{\partial{w}}{\partial{t}}
@@
       -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
-  -\lambda(t,x)\,w(t,x)
-\end{equation}
-is the \emph{filter equation} with \emph{driver} $X_t$, \emph{boost} $B$, and \emph{decay} $\lambda$.
+      -\lambda(t,x)\,w(t,x),\quad t\notin{\mathcal{M}},\\
+      w(t,x)=\int{\leftlim{w}(t,x')\,\frac{\beta(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}},\quad t\in{\mathcal{M}}.
+    \end{gathered}
+  \end{equation*}
+\end{remark}
 
-Filter equations afford a convenient means of computing the likelihood of a given sequence of events.
+Filter equations afford a convenient means of computing expectations and likelihoods for pure jump processes.
 This is facilitated by the following fact.
@@
 \begin{lemma}\label{lemma:filtereq}
-  \Cref{eq:filtereq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
+  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
   \begin{equation}\label{eq:filterlemma}
@@
       &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
-      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right].
+      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right],
     \end{aligned}
   \end{equation}
+  on the space $\Xspace\times(0,\infty)$.
   Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
@@
 
-\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(X_t,V_t)$.
-In particular, $X_t$ is the driver with KFE \cref{eq:kfe2}.
-The $V_t$ process has jumps wherever $X_t$ does, such that when $X$ jumps at time $t$ from $x$ to $x'$, $V$ jumps by the multiplicative factor $B(t,x,x')>0$.
-Between jumps, $V_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
-If we view $V_t$ as a weight, then \cref{lemma:filtereq} tells us how the $V_t$-weighted average of $X_t$ evolves in time:
+\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(\X_t,\rdm{V}_t)$.
+In particular, $\X_t$ is the driver with KFE \eqref{eq:kfe2}.
+The $\rdm{V}_t$ is \emph{directed} by $\X_t$ in the sense that $\rdm{V}$ has jumps wherever $\X$ does:
+when $\X$ jumps at time $t$ from $x$ to $x'$, $\rdm{V}$ jumps by the multiplicative factor $B(t,x,x')>0$.
+Between jumps, $\rdm{V}_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
+If we view $V_t$ as a weight, then \cref{lemma:filtereq} tells us how the $\rdm{V}_t$-weighted average of $\rdm{X}_t$ evolves in time:
 this average is simply $\int{w(t,x)\,\dd{x}}$.
-
 This motivates the following result, which shows how filter equations allow one to integrate over random histories.
@@
 \begin{prop}\label{prop:zeroboost}
-  Suppose $\X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE as in \cref{eq:kfe2}.
-  %% Suppose that $B(t,x,x')>0$, $Q(t,x,x')\in\{0,1\}$, and $\lambda(t,x)\in\R$ are given measurable functions.
-  %% Consider the equation
-  %% \begin{equation*}
-  %%   \frac{\partial{w}}{\partial{t}}
-  %%   =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,Q(t,x',x)\,\dd{x'}}
-  %%   -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
-  %%   -\lambda(t,x)\,w(t,x)
-  %% \end{equation*}
+  Suppose $\X_t$ is a non-explosive pure jump process with KFE
+  \begin{equation*}
+    \begin{gathered}
+      \frac{\partial{w}}{\partial{t}}(t,x)=\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
+      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
+      w(0,x)=p_0(x).
+    \end{gathered}
+  \end{equation*}
   Let $\Hist_t$ be its history process.
-  Suppose $\rdm{R}_t$ is a real-valued random process such that
+  Suppose $\rdm{V}_t$ is a real-valued random process such that
   \begin{equation*}
-    \Expect{\rdm{R}_t\;\vert\;\Hist_t}=\prod_{e\in\event(\Hist_t)}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)},
+    \Expect{\rdm{V}_t\;\vert\;\Hist_t=H_t}=\prod_{e\in\event(H_t)}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)},
   \end{equation*}
   for some given measurable functions $B>{0}$, $Q\in\{0,1\}$.
-  Suppose $w$ satisfies the filter equation
+  If $w$ satisfies the filter equation,
   \begin{equation*}
@@
       \frac{\partial{w}}{\partial{t}}(t,x)
-      =&\int{w(t,x')\,\beta(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\beta(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
-      &\qquad-\int{w(t,x)\,\beta(t,x,x')\,\Big(1-Q(t,x,x')\Big)\,\dd{x'}}.
+      =&\int{w(t,x')\,\alpha(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
+      -\int{w(t,x)\,\alpha(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
+      &\qquad-\int{w(t,x)\,\alpha(t,x,x')\,\Big(1-Q(t,x,x')\Big)\,\dd{x'}},
     \end{aligned}
   \end{equation*}
-  Then $\Expect{\rdm{R}_t}=\int{w(t,x)\dd{x}}$.
+  then $\Expect{\rdm{V}_t}=\int{w(t,x)\,\dd{x}}$.
 \end{prop}
 \begin{proof}
-  Apply \cref{lemma:filtereq} with the driver generated by the rate functions $\beta(t,x,x')\,Q(t,x,x')$.
+  Apply \cref{lemma:filtereq} with $\beta(t,x,x')=\alpha(t,x,x')\,Q(t,x,x')$.
 \end{proof}
 
-An important special case is that of a deterministic driving process.
-The following result is established by routine calculation.
+%% An important special case is that of a deterministic driving process.
+%% The following result is established by routine calculation.
 
-\begin{prop}
-  Suppose $X:[0,T]\to\Xspace$ is a deterministic, piecewise constant, \cadlag\ function and let $M=\event(X)$.
-  Then the KFE for $X_t$ is \cref{eq:kfe2} with $\beta(t,x,x')=\sum_{e\in{M}}{\delta(t,e)\delta(x',X_e)}$.
-  With this driver, the filter equation (\cref{eq:filtereq}) becomes
+%% \begin{prop}
+%%   Suppose $X:[0,T]\to\Xspace$ is a deterministic, piecewise constant, \cadlag\ function and let $M=\event(X)$.
+%%   Then the KFE for $X_t$ is \cref{eq:kfe2} with $\beta(t,x,x')=\sum_{e\in{M}}{\delta(t,e)\delta(x',X_e)}$.
+%%   With this driver, the filter equation (\cref{eq:filtereq}) becomes
+%%   \begin{equation*}
+%%     \begin{gathered}
+%%       \frac{\partial{w}}{\partial{t}}=
+%%       -\lambda(t,x)\,w(t,x)\quad\text{for}\quad{t\in[0,T]\setminus{M}},\\
+%%       w(t,x)=\delta(x,X_t)\,\int{\wt(t,x')\,B(t,x',X_t)\,\dd{x'}}\quad\text{for}\quad{t\in{M}}.
+%%     \end{gathered}
+%%   \end{equation*}
+%% \end{prop}
+
+Filter equations allow us to pass easily between equivalent representations of a process.
+For example, an equivalent way of representing $\X_t$ is in terms of its embedded chain and event times.
+Let $\rdm{\hat{X}}_k$ be the embedded chain of $\X_t$ and let $\rdm{\hat{T}}_k$ be the point process of its event times.
+It is elementary that
+\fontsize{10pt}{12pt}\selectfont
 \begin{equation*}
   \begin{gathered}
-      \frac{\partial{w}}{\partial{t}}=
-      -\lambda(t,x)\,w(t,x)\quad\text{for}\quad{t\in[0,T]\setminus{M}},\\
-      w(t,x)=\delta(x,X_t)\,\int{\wt(t,x')\,B(t,x',X_t)\,\dd{x'}}\quad\text{for}\quad{t\in{M}}.
+    \Prob{\rdm{\hat{X}}_k=\hat{X}_k\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_k=\hat{T}_k}=\frac{\alpha(\hat{T}_k,\hat{X}_{k-1},\hat{X}_k)}{\int{\alpha(\hat{T}_k,\hat{X}_{k-1},x')\dd{x'}}},\\
+    \Prob{\rdm{\hat{T}}_k>\rdm{\hat{T}}_{k-1}+t\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_{k-1}=\hat{T}_{k-1}}=\exp{\left(-\int_{0}^{t}{\int{\alpha(\hat{T}_{k-1}+s,\hat{X}_{k-1},x')\,\dd{x'}}\,\dd{s}}\right)}.
   \end{gathered}
 \end{equation*}
-\end{prop}
-
-The results so far allow us to compute expectations over random jumps and jump times.
-We will have occasion to compute marginal expectations in the situation where some jump times are known.
-The following result handles this case.
-
-\begin{prop}
-  Suppose $\X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'}.
-  \end{equation*}
-  Let $\Hist_t$ be its history process.
-  Suppose that $0<t_1<\dots<t_N\le{t}$ are fixed times and set $M=\{t_1,\dots,t_N\}$.
-  Let $B>{0}$ and $Q\in\{0,1\}$ be given measurable functions.
-  Let $\rdm{R}$ be a non-negative real-valued random variable such that
-  \begin{equation*}
-    \Expect{\rdm{R}\;\vert\;\Hist_t}=\prod_{e\in{\event(\Hist_t)}}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)}.
-  \end{equation*}
-  Let $\pi(t,x,x')$ be a given probability kernel and $\Y_t$ be the Markov process generated by \cref{eq:kfe2} where $\beta(t,x,x')=\left[\alpha(t,x,x')+\sum_{e\in{M}}{\delta(t,e)\,\pi(t,x,x')}\right]\,Q(t,x,x')$.
-  Suppose $w$ satisfies the filter equation with driver $\Y_t$, boost
-  \begin{equation*}
-    C(t,x,x')=
-    \begin{cases}
-      B(t,x,x'), &t\notin{M},\\
-      B(t,x,x')\,\displaystyle\frac{\alpha(t,x,x')}{\pi(t,x,x')}, &t\in{M},
-    \end{cases}
-  \end{equation*}
-  and decay $\lambda=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\,\dd{x'}}$.
-  Then
-  \begin{equation*}
-    \Expect{\rdm{R}\;\big\vert\;M\subseteq{\event(\Hist_t)}}=\int{w(t,x)\,\dd{x}}.
-  \end{equation*}
-\end{prop}
-\begin{proof}
-  Apply \cref{lemma:filtereq}.
-\end{proof}
-
-\paragraph{More systematically}
-
-Suppose $\X_t$ is a non-explosive pure jump process with KFE
-\begin{equation*}
+\normalfont
+Making the definitions,
+\begin{equation}\label[pluralequation]{eq:nudefs}
   \begin{gathered}
-    \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},\\
-    w(0,x)=p_0(x).
-  \end{gathered}
-\end{equation*}
-We can always take a version of $\X_t$ with \cadlag\ sample paths.
-
-Let $\rdm{\hat{X}}_k$ be the embedded chain of $\X_t$ and let $\rdm{T}^X_k$ be the point process of its jump times.
-It is elementary that
-\begin{equation*}
-  \Prob{\rdm{\hat{X}}_k=\hat{X}_k\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{T}^X_k=T_k}=\frac{\alpha(T_k,\hat{X}_{k-1},\hat{X}_k)}{\int{\alpha(T_k,\hat{X}_{k-1},x')\dd{x'}}}.
-\end{equation*}
-Moreover, if one has the $(\rdm{\hat{X}}_k,\rdm{T}^X_k)$ chain, one can reconstruct the $\X_t$ process.
-Conditional on $\rdm{\hat{X}}$, the jump-time chain $\rdm{T}^X_k$ is a Poisson point process with intensity $\int{\alpha(t,\X_t,x')\dd{x'}}$.
-
-We can rewrite the KFE as
-\begin{equation}\label{eq:poissdriver}
-  \begin{aligned}
-    \frac{\partial{w}}{\partial{t}}(t,n,x)=&\int\!w(t,n-1,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}-\int\!w(t,n,x)\,\alpha(t,x,x')\,\dd{x'}\\
-    =&\int\!w(t,n-1,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\dd{x'}-\int\!w(t,n,x)\,\nu\,\pi(t,x,x')\,\dd{x'}\\
-    &\qquad-\lambda(t,x)\,w(t,n,x)
-  \end{aligned}
-\end{equation}
-where $\nu>0$ is the intensity of homogenous Poisson process and
-\begin{equation*}
-  \begin{gathered}
-    A(t,x)=\int{\alpha(t,x,x')\dd{x'}}, \qquad
+    A(t,x)=\int{\alpha(t,x,x')\,\dd{x'}}, \qquad
     \pi(t,x,x')=\frac{\alpha(t,x,x')}{A(t,x)},\\
     B(t,x,x')=\frac{A(t,x)}{\nu}, \qquad
-    \lambda(t,x)=A(t,x)-\nu.
+    \lambda(t,x)=A(t,x)-\nu,
   \end{gathered}
-\end{equation*}
-Note that $\pi$ is the probability kernel of the embedded chain $\rdm{\hat{X}}$ and $A(t,x)$ is the intensity of the $\rdm{T}^X_k$ process.
-We recognize this equation as the filter equation with the rate-$\nu$ Poisson-process driver, boost $B$, and decay $\lambda$.
+\end{equation}
+we can rewrite the KFE as
+\begin{equation}\label{eq:poissdriver}
+  \frac{\partial{w}}{\partial{t}}(t,x)=
+  \int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
+  -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
+  -\lambda(t,x)\,w(t,x).
+\end{equation}
+Here, $\nu>0$ is the intensity of a time-homogenous Poisson process.
+Note that $\pi$ is the probability kernel of the embedded chain $\rdm{\hat{X}}$ and $A(t,x)$ is the intensity of the $\rdm{\hat{T}}_k$ process.
+We recognize this equation as the filter equation with boost $B$, decay $\lambda$, and driver generated by $\nu\,\pi(t,x,x')$.
+It corresponds to the following procedure for simulating $\X_t$:
+\begin{compactenum}[(a)]
+\item Simulate jump times according to the rate-$\nu$ Poisson process.
+\item Simulate the embedded chain $\rdm{\hat{X}}_k$ using the kernel $\pi$.
+\item Weight the realization by the product of the $B$ factors.
+  Note that this makes the appropriate importance-sampling correction.
+\end{compactenum}
 
-By integrating equations of this kind, we effectively integrate over the $(\rdm{\hat{X}},\rdm{T}^X)$ process.
-To marginalize over only some of the $\rdm{T}^X_k$, we reason as follows.
-Fix $0<t_1<\dots<t_K\le{T}$ and $\nu>0$.
-Consider the mixed point process $T_k$ with intensity $\nu+\sum_{k=1}^K\delta(t,t_k)$.
-Using this as the driver in \cref{eq:poissdriver}, we obtain
+By integrating equations of this kind, we effectively compute expectations with respect to the $\X_t$ process.
+In particular, we can use \cref{prop:zeroboost} or \cref{eq:poissdriver} to integrate out the unobserved event times $\rdm{\hat{T}}_k$, as well as the unobserved states $\rdm{\hat{X}}_k$.
+However, we also need to compute likelihoods for data that include observed event times.
+Technically, in such a case, the likelihood is a density on the cone $\mathbb{C}_t=\coprod_{m=0}^{\infty}{\mathbb{C}^m_t}$, where $\mathbb{C}^m_t=\{t\in\halfclosed{0,t}\;\vert\;t_1<t_2<\dots<t_m\}$.
+To avoid having to work with such complex objects, it is convenient to represent the likelihood as a density with respect to a constant-rate Poisson point process.
+This is accomplished as follows.
+
+Suppose the finite set of observed event times $\mathcal{M}\subseteq\halfclosed{0,t}$ is fixed and $|\mathcal{M}|=M$.
+Let $\nu>0$ be a constant rate and consider the mixed point process with intensity $\nu+\sum_{e\in\mathcal{M}}\delta(t,e)$.
+This point process has events at $e\in\mathcal{M}$ with probability 1 as well as additional, Poisson-distributed random event times.
+Using this as the driver in \cref{eq:poissdriver}, we obtain the filter equation
 \begin{equation*}
   \begin{aligned}
-    \frac{\partial{w}}{\partial{t}}(t,n,x)=&\int{w(t,n-1,x')\,\nu\,\pi(t,x',x)\,B(t,x',x)\dd{x'}}-\int{w(t,n,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}\\
-    &\quad+\sum_{k}\delta(t,t_k)\,\left\{\int{w(t,n-1,x')\,\pi(t,x',x)\,B(t,x,x')\dd{x'}}-\int{w(t,n,x)\,\pi(t,x,x')\,\dd{x'}}\right\}\\
-    &\qquad-\lambda(t,x)\,w(t,n,x).
+    \frac{\partial{w}}{\partial{t}}(t,x)=&\int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x',x)\,\dd{x'}}-
+    \int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}\\
+    &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
+    -\int{w(t,x)\,\pi(t,x,x')\,\dd{x'}}\right\}\\
+    &\qquad-\lambda(t,x)\,w(t,x).
   \end{aligned}
@@
 Equivalently, we write:
+\begin{equation}\label[pluralequation]{eq:mixedKFE}
+  \begin{aligned}
+    \frac{\partial{w}}{\partial{t}}(t,x)=&\int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x',x)\,\dd{x'}}
+    -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}\\
+    &\qquad-\lambda(t,x)\,w(t,x)\\
+    =&\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
+    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\quad t\notin{\mathcal{M}},\\
+    w(e,x)=&\int{\leftlim{w}(e,x')\,\pi(e,x',x)\,B(e,x',x)\,\dd{x'}}\\
+    =&\tfrac{1}{\nu}\,\int{\leftlim{w}(e,x')\,\alpha(e,x',x)\,\dd{x'}},\quad e\in\mathcal{M}.
+  \end{aligned}
+\end{equation}
+\Cref{eq:mixedKFE} is a filter equation relative to $\nu$ with observed events $\mathcal{M}$.
+It defines a probability measure $\mu_M$ on $\Xspace\times{\mathbb{C}^M_t}$.
+Specifically, $w(t,x)$ is the Radon-Nikodym derivative of $\mu_M$ with respect to the product of the measure on $\Xspace$ and the probability measure of the rate-$\nu$ Poisson process on $\mathbb{C}^M_t$.
+That is,
+\begin{equation*}
+  \mu^M_t(\dd{x}\,\dd{t_1}\,\cdots\,\dd{t_M})=\nu^{M}\,e^{-\nu\,t}\,w(t,x)\,\dd{x}\,\dd{t_1}\,\cdots\,\dd{t_M}.
+\end{equation*}
+
+The same reasoning shows how to compute expectations of a real-valued process $\rdm{V}_t$ directed by $\X_t$.
+As before, we will express the expectation $\Expect{\rdm{V}_t}$ as a density with respect to the rate-$\nu$ Poisson point process.
+
+\begin{prop}\label{prop:fullfilt}
+  Suppose $\X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE
+  \begin{equation*}
+    \frac{\partial{w}}{\partial{t}}(t,x)=\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
+    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
+  \end{equation*}
+  Let $\Hist_t$ be its history process.
+  Suppose that $\mathcal{M}\subseteq\halfclosed{0,t}$, $|\mathcal{M}|=M$, is a fixed, finite set of observed event times.
+  Let $B>{0}$ and $Q\in\{0,1\}$ be given measurable functions.
+  Let $\rdm{V}_t$ be a non-negative real-valued random process such that
+  \begin{equation*}
+    \Expect{\rdm{V}_t\;\vert\;\Hist_t}=\prod_{e\in{\event(\Hist_t)}}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)}.
+  \end{equation*}
+  Let $\nu>0$ be a fixed rate and let $\X'_t$ be the pure-jump process generated by the rates $\alpha(t,x,x')\,Q(t,x,x')$.
+  Suppose that $w(t,x)$ satisfies the filter equation (relative to $\nu$) with driver $\X'_t$, boost $B'$, observation events $\mathcal{M}$, and decay
+  \begin{equation*}
+    \lambda(t,x)=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\,\dd{x'}}.
+  \end{equation*}
+  Then the measure $\Expect{\rdm{V}_t}$ has density $\int{w(t,x)\,\dd{x}}$ with respect to the rate-$\nu$ Poisson point process.
+\end{prop}
+\begin{proof}
+  We rewrite the filter equation of \cref{prop:zeroboost} using the rate-$\nu$ Poisson driver:
   \begin{equation*}
     \begin{aligned}
-    \frac{\partial{w}}{\partial{t}}(t,n,x)=&\int{w(t,n-1,x')\,\nu\,\pi(t,x',x)\,B(t,x',x)\dd{x'}}-\int{w(t,n,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}\\
-    &\qquad-\lambda(t,x)\,w(t,n,x),\\
-    w(t_k,n,x)=&\int{\leftlim{w}(t_k,n-1,x')\,\pi(t_k,x',x)\,B(t_k,x',x)\,\dd{x'}}\\
-    &\qquad=\tfrac{1}{\nu}\,\int{\leftlim{w}(t_k,n-1,x')\,\alpha(t,x',x)\,\dd{x'}}\\.
+      \frac{\partial{w}}{\partial{t}}(t,x)
+      =&\int{w(t,x')\,\alpha(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
+      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
+      =&\int{w(t,x')\,\nu\,\pi(t,x',x)\,\frac{A(t,x')\,B(t,x',x)}{\nu}\,\dd{x'}}
+      -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}\\
+      &\quad-\lambda(t,x)\,w(t,x),
     \end{aligned}
   \end{equation*}
-We then have $\Prob{\X_t\in\mathcal{E}\;\big\vert\;t_1,\dots,t_K\in\event(\Hist_t)}=\int_{\mathcal{E}}{w(t,x)\,\dd{x}}$.
+  where
+  \begin{equation*}
+    \begin{gathered}
+      A(t,x)=\int{\alpha(t,x,x')\,Q(t,x,x')\dd{x'}}, \qquad
+      \lambda(t,x)=\int{\alpha(t,x,x')\,\dd{x'}}-\nu,\\
+      \pi(t,x,x')=\frac{\alpha(t,x,x')\,Q(t,x,x')}{A(t,x)}.
+    \end{gathered}
+  \end{equation*}
+  Replacing the Poisson driver with $\nu+\sum_{e\in\mathcal{M}}{\delta(t,e)}$ as before, we have
+  \begin{equation*}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}(t,x)
+      =&\int{w(t,x')\,\nu\,\pi(t,x',x)\,C(t,x',x)\,\dd{x'}}
+      -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}-\lambda(t,x)\,w(t,x)\\
+      &\quad+\sum_{e\in\mathcal{M}}{\delta(t,e)\,\left\{\int{w(t,x')\,\pi(t,x',x')\,C(t,x',x)\,\dd{x'}}
+        -\int{w(t,x)\,\pi(t,x,x')\,\dd{x'}}\right\}}\\
+    \end{aligned}
+  \end{equation*}
+  where
+  \begin{equation*}
+    C(t,x,x')=\frac{A(t,x)\,B(t,x,x')}{\nu}.
+  \end{equation*}
+  This simplifies to
+  \begin{equation*}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}(t,x)
+      =&\int{w(t,x')\,\alpha(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
+      -\int{w(t,x)\,\alpha(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
+      &\quad+\int{w(t,x)\,\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\,\dd{x'}}\\
+      &\quad+\sum_{e\in\mathcal{M}}{\delta(t,e)\,\left\{\int{w(t,x')\,\frac{\alpha(t,x',x)\,Q(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}}
+        -w(t,x)\right\}}.\\
+    \end{aligned}
+  \end{equation*}
+  In the latter, we recognize the filter equation of the statement.
+\end{proof}
 
-Now suppose $\rdm{R}_k$ is a Markov chain satisfying
-\begin{equation*}
-  \Prob{\rdm{R}_k=R_{k-1}\,B(T_k,\hat{X}_{k-1},\hat{X}_k)\;\Big\vert\;\rdm{\hat{X}}_k=\hat{X}_k,\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{T}^X_k=T_k,\rdm{R}_{k-1}=R_{k-1}}=1,
-\end{equation*}
-for some function $B>0$.
-\AAK{Or perhaps the following is sufficient for our purposes?}
-\begin{equation*}
-  \Expect{\rdm{R}_k\;\Big\vert\;\rdm{\hat{X}}_k=\hat{X}_k,\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{T}^X_k=T_k,\rdm{R}_{k-1}=R_{k-1}}=R_{k-1}\,B(T_k,\hat{X}_{k-1},\hat{X}_k).
-\end{equation*}
-Let $\rdm{R}_t$ be the corresponding continuous-time jump process with jumps at $\rdm{T}^X_k$.
-By \cref{lemma:filtereq}, if $w$ satisfies the boost equation with boost $B$, driver $\X$, and decay $\lambda=0$, then
-\begin{equation*}
-  \Expect{\rdm{R}_t}=\int{w(t,x)\dd{x}}.
-\end{equation*}
+\begin{remark}
+  Suppose $\X_t$, $\Y_t$ have KFEs
+  \begin{equation}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}(t,x)=
+      &\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
+      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\\
+      \frac{\partial{w}}{\partial{t}}(t,y)=
+      &\int{w(t,y')\,\beta(t,y',y)\,\dd{y'}}
+      -\int{w(t,y)\,\beta(t,y,y')\,\dd{y'}},
+    \end{aligned}
+  \end{equation}
+  where $\beta(t,y,y')=\alpha(t,y',y)\,\left(1+\frac{1}{\nu}\sum_{e\in\mathcal{M}}{\delta(t,e)}\right)$.
+\end{remark}
 

### 2023-07-20 ms.Rnw
some improvements

@@
 
-We use bold-face to mark random elements and the corresponding regular-face symbol to denote specific values of such elements.
-Given any \cadlag\ process $f_t$, the set of jump-times is denoted $\event(f_t)=\{e\in[0,t]\;\vert\;\leftlim{f}_e\ne{f_e}\}$.
-
 Denote the underlying probability space by $(\Omega,\Borel,\mathrm{Prob})$.
-We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\colonequals\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$ and taking values in some space $\Xspace$.
+Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\rdm{X}$), denotes a random element while the same symbol in regular face ($X$) denotes an element in its range.
+We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\colonequals\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$) and taking values in some space $\Xspace$.
 In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
@@
   \begin{gathered}
-    \Prob{\N_{t+\Delta}=n+1\;\vert\;\N_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),
-    \qquad
+    \Prob{\N_{t+\Delta}=n+1\;\vert\;\N_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
     \Prob{\N_{t+\Delta}>n+1\;\vert\;\N_{t}=n}=o(\Delta),\\
@@
 
-Without loss of generality, one can assume, as we do here, that the sample paths $t\mapsto\X_t(\omega)$ for $\omega\in\Omega$ are right-continuous with left limits (\cadlag).
-In fact, all of the processes we will describe in this paper will be taken to be \cadlag, and we will frequently need to refer to the left-limit.
+Without loss of generality, one can assume, as we do here, that the sample paths $t\mapsto\X_t(\omega)$ for $\omega\in\Omega$ are right-continuous with left limits (\ie \cadlag).
+In fact, all of the processes we will describe in this paper will be taken to have \cadlag\ sample paths, and we will frequently need to refer to their left-limits.
 Accordingly, if $\rdm{Z}_t$ is any \cadlag\ random process, we define
@@
 Note that $\leftlim{\rdm{Z}}_t$ is thus left-continuous with right limits.
+Denote the set of event times of $\rdm{Z}$ by $\event{\rdm{Z}}{t}=\{e\in[0,t]\;\vert\;\leftlim{\rdm{Z}}_e\ne{\rdm{Z}_e}\}$.
 
@@
 As usual, we take the sample paths, $t\mapsto{\U_t(\omega)}$ for $\omega\in\Omega$, to be \cadlag.
-Observe that $\U_t$ is a random, but not a Markov, process.
+Observe that $(\X_t,\U_t)$ is a Markov process but that $\U_t$ is not.
 
@@
 
-For $\omega\in\Omega$, $t\mapsto(\X_t(\omega),\U_t(\omega))$ is a \cadlag\ function of time.
-Let the \emph{history process}, $\Hist_t$, be the restriction of this random function to the interval $[0,t]$.
-Note that $\Hist_t$ is a Markov process.
-
-The non-explosiveness assumption implies that a.s., for every $t$, there is a finite, increasing sequence of random event times, which we denote by $\event(\Hist_t)\colonequals\left(\rdm{T}_k\right)_{k=1}^{\rdm{K}_t}$;
-the length, $\rdm{K}_t$, of this sequence is itself random.
-However, conditional on $\Hist_t$, $\rdm{T}_k$ and $\rdm{K}_t$, together with the mark process $\U_t$, and the population process $\X_t$ are deterministic.
-One can write down an explicit probability measure for $\Hist_t$ in terms of these:
-\begin{equation}\label{eq:Hdens}
+Consider the Markov process $(\X_t,\U_t)$.
+We define its \emph{history process}, $\Hist_t$, to be the restriction of the random function $t\mapsto(\X_t,\U_t)$ to the interval $[0,t]$.
+Note that $\Hist_t$ is itself a Markov process.
+%% Conditional on $\Hist_t$, both $\X_t$ and $\U_t$ are deterministic, as of course are the Markov chains of event times $\rdm{\hat{T}}_k$ and the embedded chains $\rdm{\hat{X}}_k=\X_{\rdm{\hat{T}}_k}$ and $\rdm{\hat{U}}_k=\U_{\rdm{\hat{T}}_k}$.
+Conditional on $\Hist_t$, both $\X_t$ and $\U_t$ are deterministic as are the embedded chain, $(\X_k,\U_k)$, and the point process of event times $\rdm{\hat{T}}_k$.
+(Recall that $\rdm{\hat{X}}_k=\rdm{X}_{\rdm{\hat{T}}_k}$.)
+The non-explosiveness assumption implies that, for every $t$, $\rdm{K}_t\coloneq\big\vert\event{\Hist}{t}\cap[0,t]\big\vert<\infty$ a.s.
+The probability measure, $\pi^H$, for $\Hist_t$ can be expressed in terms of these:
+\begin{equation*}%\label{eq:Hdens}
   \begin{aligned}
-    \pi^{\Hist}(\dd{\Hist_t})
-    =&p_{0}(X_{0})\,\dd{X_0}\,\prod_{k=1}^{K_{t}}{\left(\alpha_{U_{T_k}}\left(T_k,\leftlim{X}_{T_k},X_{T_k}\right)\,\dd{X_{T_k}}\,\dd{T_k}\right)}\\
-    &\qquad\times\exp{\left(-\sum_{u}{\int_{0}^{t}{\int{\alpha_u(t',\leftlim{X}_{t'},x')\,\dd{x'}}\,\dd{t'}}}\right)}.
+    \pi^{H}(\dd{H_t})
+    =&p_{0}(\hat{X}_{0})\,\dd{\hat{X}_0}\,
+    \prod_{k=1}^{K_{t}}{\left(\alpha_{\hat{U}_k}\left(\hat{T}_k,\hat{X}_{k-1},\hat{X}_{k}\right)\,\dd{\hat{X}_k}\,\dd{\hat{T}_k}\right)}\\
+    &\qquad\times\exp{\left(-\sum_{k=1}^{K_t}{\int_{\hat{T}_{k-1}}^{\hat{T}_k}{\sum_{u}{\int{\alpha_{u}(t',\hat{X}_{k-1},x')\,\dd{x'}}}}\,\dd{t'}}\right)}.
   \end{aligned}
-\end{equation}
+\end{equation*}
 
@@
 Each $\Node{p}\in\Gen$ has a creation-time, $T(\Node{p})$, a parent, $\Anc(\Node{p})$, and a deme.
-We let $\event(G)$ denote the sequence of node times:
-$\event(G)=\left(\Node{p}_k(G)\right)_{k=1}^{K(G)}$, where $0=T(\Node{p}_1(G)){\le}T(\Node{p}_2(G)){\le}\dots{\le}T(\Node{p}_{K(G)}(G)){\le}T(G)$.
+Moreover, $\event{G}{t}$ is just the sequence of node times:
+$\event{G}{t}=\left(T(\Node{p}_k(G))\right)_{k=1}^{K(G)}$.
 %% The nodes are ordered so that $k<k'$ implies that $T(\Node{p}_k(\Gen))\le{T(\Node{p}_{k'}(\Gen))}$.
@@
   In particular $\deme^{P}_j(t)=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
-\item Let $\Yspace=(\Zp\times\Zp\times\Demesplus)^{\Zp}$ and define $Y^{P}:\Rp\to\Yspace$ such that, for $j\in\Zp$ and $t\in\Rp$, $Y^{P}_j(t)=\left(\ct^{P}_j(t),\anc^{P}_j(t),\deme^{P}_j(t)\right)$.
+\item Let $\Yspace=(\Zp\times\Zp\times\Demesplus)^{\Zp}$ and define $Y^{P}:\Rp\to\Yspace$ such that, for $t\in\Rp$,
+  $Y^{P}(t)=\left(\ct^{P}(t),\anc^{P}(t),\deme^{P}(t)\right)$.
 \end{compactenum}
@@
 If $y$ does define a pruned genealogy, we denote that pruned genealogy by $P(y)$.
-
-Given $\eta\in\Yspace$, we will use the notation $\ct(\eta)$, $\anc(\eta)$, and $\deme(\eta)$ to refer to the coordinates of $\eta$.
+Given $y\in\Yspace$, we will use the notation $\ct(y)$, $\anc(y)$, and $\deme(y)$ to refer to the coordinates of $y$.
 
@@
 We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
-Therefore, for $\eta,\eta'\in\Yspace$, and $i\in\Demes$, let us define
+Therefore, for $y,y'\in\Yspace$, and $i\in\Demes$, let us define
 \begin{equation*}
-  s_i(\eta,\eta')\colonequals{\left\vert\left\{j\in{\Zp}\;\vert\;\deme(\eta_j)=i\ \ \&\ \ \ct(\eta'_j)\ne\ct(\eta_j)\right\}\right\vert}, \qquad s(\eta,\eta')\colonequals(s_i(\eta,\eta'))_{i\in\Demes}\in{\Zp^\Demes}.
+  s_i(y,y')\colonequals{\left\vert\left\{j\in{\Zp}\;\vert\;\deme(y_j)=i\ \ \&\ \ \ct(y'_j)\ne\ct(y_j)\right\}\right\vert}, \qquad s(y,y')\colonequals(s_i(y,y'))_{i\in\Demes}\in{\Zp^\Demes}.
 \end{equation*}
 With this definition, for any $P$, if $y=Y^P$, then $s_i(\yt(t),y(t))$ is the number of lineages in deme $i$ that emerge from an event at time $t$ and $s(\yt(t),y(t))$ is the non-negative integer vector telling how many lineages in each deme emerge at time $t$.
-In particular, $s(\eta,\eta)=0$ so that if $t\notin\event(P)$, then $s(\yt(t),y(t))=0$.
+In particular, $s(y,y)=0$ so that if $e\notin\event{P}{t}$, then $s(\yt(e),y(e))=0$.
 We refer to $s\left(\yt(t),y(t)\right)$ as the \emph{saturation} at time $t$.
@@
 
-For $n,r,\ell,n\in{\Zp^\Demes}$, define the \emph{binomial ratio}
+For $n,r,\ell,s\in{\Zp^\Demes}$, define the \emph{binomial ratio}
 \begin{equation*}
@@
   \begin{cases}
-    \frac{\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}{\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
-    & \forall i\ n_i\ge{\{\ell_i,r_i\}}\ge{s_i}\ge{0}\\
-    0, & \text{otherwise}
+    \frac{\displaystyle\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}%
+         {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
+         & \forall i\ n_i\ge{\{\ell_i,r_i\}}\ge{s_i}\ge{0},\\
+         0, & \text{otherwise}.
   \end{cases}
@@
 
-Our first result will be an expression for the likelihood of a given pruned genealogy $P^*$ given the history $\Hist_t$ of the population process up to time $t=T(P^*)$.
+Our first result will be an expression for the likelihood of a given pruned genealogy $P^*_t$ given the history $\Hist_t$ of the population process to time $t$.
 Of course, not every pruned genealogy is compatible with $\Hist_t$.
-In particular, $P^*$ is only compatible with $\Hist_t$ if every event in $P^*$ coincides with an event in $\Hist_t$.
-That is, only if $\event(P^*)\subseteq{\event(\Hist_t)}$.
-Moreover, certain events in $P^*$ may be incompatible with all possible histories.
-For example, if $P^*$ has an event in which a lineage moves from deme $i$ to deme $i'$ but there are no $u\in\Jumps$ for which this is possible, then $P^*$ is incompatible with the population process itself.
-Let us define the function $Q:\Jumps\times\Yspace\times\Yspace\to\{0,1\}$ so that $Q_u(\eta,\eta')=1$ if and only if a change $\eta\to\eta'$ in a pruned genealogy is compatible with an event of mark $u$ at that time.
-More formally, $Q_u(\eta,\eta')=1$ if and only if there are $\omega\in\Omega$ and $t'<t\in\Rp$ such that $\U_{t'}(\omega)=u$, $\leftlim{Y}^{\prune(\Gen_t(\omega))}(t')=\eta$, and $Y^{\prune(\Gen_t(\omega))}(t')=\eta'$.
-Recall that $n:\Xspace\to\Zp$ denotes the deme-occupancy function, $r^u$, the production of a jump of mark $u$, $\ell$, the lineage-count function, and $s$, the saturation.
+In particular, $P^*_t$ is only compatible with $\Hist_t$ if every event in $P^*_t$ coincides with an event in $\Hist_t$.
+That is, only if $\event{P^*}{t}\subseteq{\event{\Hist}{t}}$.
+Moreover, certain events in $P^*_t$ may be incompatible with all possible histories.
+For example, if $P^*_t$ has an event in which a lineage moves from deme $i$ to deme $i'$ but there are no $u\in\Jumps$ for which this is possible, then $P^*_t$ is incompatible with the population process itself.
+Let us define the function $Q:\Jumps\times\Yspace\times\Yspace\to\{0,1\}$ so that $Q_u(y,y')=1$ if and only if a change $y\to{y'}$ in a pruned genealogy is compatible with an event of mark $u$ at that time.
+Recall that $n:\Xspace\to\Zp^{\Demes}$ denotes the deme-occupancy function, $r^u$, the production of a jump of mark $u$, $\ell$, the lineage-count function, and $s$, the saturation.
 
 \begin{thm}\label{thm:pruned_lik}
-  Let $P^*$ be a given pruned genealogy, $t=T(P^*)$, and $y^*=Y^{P^*}$.
+  Let $P^*_t$ be a given pruned genealogy and $y^*=Y^{P^*_t}$.
   Then
@@
     \begin{aligned}
-      \Prob{P^*\vert\Hist_t}=&\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}\,Q_{U_e}(\yt^*(e),y^*(e))}\\
-      &\qquad\times\Indicator{\event(P^*)\subseteq{\event(\Hist_t)}}.
+      \Prob{P^*_t\vert\Hist_t}=&\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}\,Q_{U_e}(\yt^*(e),y^*(e))}\\
+      &\qquad\times\Indicator{\event{P^*}{t}\subseteq{\event{\Hist}{t}}}.
     \end{aligned}
@@
 \begin{proof}
-  As we have already observed, if $\event(P^*)\nsubseteq\event(\Hist_t)$, then $\Prob{P^*\vert\Hist_t}=0$.
-  Similarly, if there is any event of $P^*$ which is incompatible with the population process, $\Prob{P^*\vert\Hist_t}=0$.
+  As we have already observed, if $\event{P^*}{t}\nsubseteq\event{\Hist}{t}$, then $\Prob{P^*_t\vert\Hist_t}=0$.
+  Similarly, if there is any event of $P^*_t$ which is incompatible with the population process, $\Prob{P^*_t\vert\Hist_t}=0$.
   Let us therefore suppose that neither of these conditions hold.
   Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
-  Moreover, at each event $e\in\event(\Hist_t)$, a jump of mark $U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
-  In $P^*$, at time $e$, there are $\ell_i=\ell_i(y^*(t))$ lineages in deme $i$, of which $s_i=s_i(\yt^*(t),y^*(t))$ are emergent.
+  Moreover, at each event $e\in\event{\Hist}{t}$, a jump of mark $U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
+  In $P^*_t$, at time $e$, there are $\ell_i=\ell_i(y^*(t))$ lineages in deme $i$, of which $s_i=s_i(\yt^*(t),y^*(t))$ are emergent.
   The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
@@
   \end{equation*}
-  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $P^*$.
+  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $P^*_t$.
   Moreover, since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
   \begin{equation*}
-    \Prob{P^*\vert\Hist_t}=\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}}.
+    \Prob{P^*_t\vert\Hist_t}=\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}}.
   \end{equation*}
-  Returning to the possibility that $P^*$ is incompatible with $\Hist_t$, since $\Prob{P^*}=0$ if either any $Q_u=0$ or of any event of $P^*$ is not an event of $\Hist_t$, we obtain the result.
+  Returning to the possibility that $P^*_t$ is incompatible with $\Hist_t$, since $\Prob{P^*_t}=0$ if either any $Q_u=0$ or any event of $P^*_t$ is not an event of $\Hist_t$, we obtain the result.
 \end{proof}
@@
 \begin{thm}\label{thm:obsc_lik}
-  Let $V^*$ be a given obscured genealogy and $t=T(V^*)$.
-  Let $\pi:\Jumps\times\Rp\times\Xspace^2\times\Yspace^2\to\Rp$ be a probability kernel, \ie $\pi_u(t,\xi,\xi',\eta,\eta')\ge{0}$ and $\sum_{\eta'}{\pi_u(t,\xi,\xi',\eta,\eta')}=1$.
-  Suppose moreover that $\pi_u(t,\xi,\xi',\eta,\eta')>0$ if and only if $Q_u(\eta,\eta')=1$.
-  Let $\rdm{y}_t$ be the Markov jump process on $\Yspace$ generated by the kernel $\pi$, with $\event(\rdm{y})=\event(\Hist_t)$.
-  Then
+  Let $V^*_t$ be a given obscured genealogy.
+  Let $\pi:\Jumps\times\Rp\times\Xspace^2\times\Yspace^2\to\Rp$ be a probability kernel, \ie $\pi_u(t,x,x',y,y')\ge{0}$ and $\sum_{y'}{\pi_u(t,x,x',y,y')}=1$.
+  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ if and only if $Q_u(y,y')=1$.
+  Let $\rdm{y}_t$ be the Markov jump process on $\Yspace$ generated by the kernel $\pi$, with $\event{\rdm{y}}{t}=\event{\Hist}{t}$.
+  Then there is a measurable function $\kappa_{V^*_t}:[0,t]\times\Yspace^2\to\{0,1\}$ such that
   \begin{equation*}
-    \Prob{V^*\vert\Hist_t}=\Expect{\left.\prod_{e\in\event(\Hist_t)}{\frac{\scriptsize\BinRatio{n(X_e)}{\ell(\rdm{y}_e)}{r^{U_e}}{s(\leftlim{\rdm{y}}_e,\rdm{y}_e)}}{\scriptsize\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}\times\Indicator{\obs(P(\rdm{y}))=V^*}\;\right\vert\;\Hist_t},
+    \begin{aligned}
+      \Prob{V^*_t\vert\Hist_t}=&\Expect{\left.\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(\rdm{y}_e)}{r^{U_e}}{s(\leftlim{\rdm{y}}_e,\rdm{y}_e)}\,\frac{\scriptsize\kappa_{V^*_t}(e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}{\scriptsize\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}\;\right\vert\;\Hist_t}\\
+      &\qquad\times\Indicator{\event{V^*}{t}\subseteq\event{\Hist}{t}},
+    \end{aligned}
   \end{equation*}
@@
   \begin{equation}\label{eq:IS1}
-    \Prob{V^*\vert\Hist_t}=\int{\Indicator{\obs(P)=V^*}\,\Prob{P\vert\Hist_t}\,\dd{P}},
+    \Prob{V^*_t\vert\Hist_t}=\int{\Indicator{\obs(P_t)=V^*_t}\,\Prob{P_t\vert\Hist_t}\,\dd{P_t}},
   \end{equation}
@@
   Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
-  we will propose pruned genealogies compatible with $V^*$ as sample paths from a Markov process on $\Yspace$ and
+  we will propose pruned genealogies compatible with $V^*_t$ as sample paths from a Markov process on $\Yspace$ and
   treat the integral in \cref{eq:IS1} as an expectation over these paths.
@@
   \end{equation*}
-  Now let $\event(\Hist_t)=\{t_1,t_2,\dots,t_K\}$.
+  Now let $\event{\Hist}{t}=\{t_1,t_2,\dots,t_K\}$.
   $\rdm{\hat{y}}_k$ is the embedded chain of a unique, piecewise-constant, \cadlag\ process $\rdm{y}_t$, with jump times at $t_k$.
-  The conditions on $\pi$ guarantee that, for every pruned genealogy $P$, $Y^P$ is a sample path of this process.
-  In particular, they guarantee that $\pi_u(t,\xi,\xi',\eta,\eta')\;\propto\;{Q_u(\eta,\eta')}$.
-  Clearly, the construction of $\rdm{y}_t$ guarantees that $\event(P(\rdm{y}))\subseteq\event(\Hist_t)$.
+  This construction of $\rdm{y}_t$ obviously guarantees that $\event{P(\rdm{y})}{t}\subseteq\event{\Hist}{t}$.
+  Moreover, the conditions on $\pi$ guarantee that, for every pruned genealogy $P_t$, $Y^{P_t}$ is a sample path of this process.
+  In particular, they guarantee that $\pi_u(t,x,x',y,y')\;\propto\;{Q_u(y,y')}$.
 
@@
   \begin{equation*}
-    \Prob{V^*\vert\Hist_t}=\Expect{\frac{\Prob{P(\rdm{y})\vert\Hist_t}}{\pi(\rdm{y}\vert\Hist_t)}\cdot\Indicator{\obs(P(\rdm{y}))=V^*}},
+    \Prob{V^*_t\vert\Hist_t}=\Expect{\frac{\Prob{P(\rdm{y})\vert\Hist_t}}{\pi(\rdm{y}\vert\Hist_t)}\cdot\Indicator{\obs(P(\rdm{y}))=V^*_t}},
   \end{equation*}
@@
   \begin{equation*}
-    \pi(\rdm{y}\vert\Hist_t)=\prod_{e\in\event(\Hist_t)}{\pi_{U_{e}}(e,\leftlim{X}_{e},X_{e},\leftlim{\rdm{y}}_{e},\rdm{y}_{e})}.
+    \pi(\rdm{y}\vert\Hist_t)=\prod_{e\in\event{\Hist}{t}}{\pi_{U_{e}}(e,\leftlim{X}_{e},X_{e},\leftlim{\rdm{y}}_{e},\rdm{y}_{e})}.
   \end{equation*}
+
+  Finally, observe that the condition $\obs(P(\rdm{y}))=V_t$ can be factorized for all obscured genealogies $V_t$.
+  In particular, there is a function $\kappa_{V_t}:[0,t]\times\Yspace^2\to\{0,1\}$ such that $\kappa_{V_t}(s,\eta,\eta')=1$ if and only if
+  there is $y:\Rp\to\Yspace$ such that $\obs(P(y))=V_t$, $\leftlim{y}_s=\eta$, and $y_s=\eta'$, for all $s\in[0,t]$.
+  Thus, conditional on $\Hist_t$, if $\event{P(y)}{t}\subseteq\event{\Hist}{t}$,
+  \begin{equation*}
+    \Indicator{\obs(P(y))=V^*_t}=\prod_{e\in\event{\Hist}{t}}{\kappa_{V^*_t}(e,\leftlim{y}_e,y_e)}.
+  \end{equation*}
+
   The result then follows from \cref{thm:pruned_lik}.
@@
 \begin{thm}
-  Suppose we have a pruned genealogy $P^*_T$, $T>0$, with event times $\event(P^*_T)$.
-  Let $y^*=Y^{\Prune^*_T}$.
-  For $x,x'\in\Xspace$, $y,y'\in\Yspace$, $u\in\Jumps$, define
+  Suppose that $P^*_T$ is a pruned genealogy, for $T>0$.
+  Let $y^*=Y^{\Prune^*_T}$ and, for $x\in\Xspace$, $u\in\Jumps$, $t\le{T}$, define
   \begin{equation*}
-    \begin{gathered}
-      \phi_u\big(x,y,y')=\BinRatio{n(x)}{\ell(y')}{r^u}{s(y,y')}\qquad
-      \text{and} \qquad
-      \beta_u(t,x,x')=\alpha_u(t,x,x')\,Q_u(y,y'),
-    \end{gathered}
+    \phi_u\big(t,x)=\BinRatio{n(x)}{\ell(y^*_t)}{r^u}{s(\yt^*_t,y^*_t)}\,Q_u(\yt^*_t,y^*_t),
   \end{equation*}
-  where $Q_u(y,y')$ is as in \cref{thm:pruned_lik}.
-  Fix $\nu>0$ and suppose $w=w(t,x)$ satisfies the filter equation
+  where $Q_u$ is as in \cref{thm:pruned_lik}.
+  Suppose $w=w(t,x)$ satisfies the filter equation
   \begin{equation}
@@
       \frac{\partial{w}}{\partial{t}}=
-      &\sum_u{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u\big(x,\yt^*_t,y^*_t\big)\,\dd{x'}}}
-      -\sum_u{\int{w(t,x)\,\beta_u(t,x,x')\,\dd{x'}}}\\
-      &\quad+\sum_{e\in\event(P^*_T)}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\frac{\beta_u(t,x',x)}{\nu}\,\phi_u\big(x,\yt^*_t,y^*_t\big)\,\dd{x'}}}-w(t,x)\right\}},
+      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(t,x)\,\dd{x'}}}
+      -\sum_u{\int{w(t,x)\,\alpha_u(t,x',x)\,\dd{x'}}}\\
+      &\ +\sum_{e\in\event{P^*}{T}}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(t,x)\,\dd{x'}}}-w(t,x)\right\}},
     \end{aligned}
@@
     \lik(P^*_T)=\int{w(T,x)\,\dd{x}}.
+  \end{equation*}
+\end{thm}
+
+The final result shows how to compute the likelihood of an obscured genealogy.
+
+\begin{thm}
+  Let $V^*_T$, $T>0$ be a given obscured genealogy.
+  Let $\pi$, $\kappa_{V^*_T}$, and $\rdm{y}_t$ be as in \cref{thm:obsc_lik}.
+  For $x,x'\in\Xspace$, $y,y'\in\Yspace$, $u\in\Jumps$, define
+  \begin{equation*}
+    \begin{gathered}
+      \phi_u\big(t,x,y,y')=\BinRatio{n(x)}{\ell(y')}{r^u}{s(y,y')}\,\frac{\kappa_{V^*_T}(t,y,y')}{\pi_u(t,x,x',y,y')},\\
+      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y').
+    \end{gathered}
+  \end{equation*}
+  Suppose $w=w(t,x)$ satisfies the filter equation
+  \begin{equation}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}(t,x,y)=
+      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\phi_u(t,x,y',y)\,\dd{x'}}}\\
+      &\ -\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
+      &\ +\sum_{e\in\event{P^*}{T}}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u(t,x,y',y)\,\dd{x'}}}-w(t,x)\right\}},
+    \end{aligned}
+  \end{equation}
+  with the initial condition $w(0,x,y)=p_0(x)\,\pi_0(y)$.
+  Then the likelihood of $V^*_T$ is
+  \begin{equation*}
+    \lik(V^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
   \end{equation*}
@@
 \appendix
-\setcounter{thm}{0}
 \setcounter{equation}{0}
@@
       -\lambda(t,x)\,w(t,x)\\
-      &\quad+\sum_{e\in\mathcal{M}}\left\{\int{w(t,x')\,\frac{\beta(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}.
+      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\frac{\beta(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}
+      +\nu\,w(t,x).
     \end{aligned}
@@
       -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
-      -\lambda(t,x)\,w(t,x),\quad t\notin{\mathcal{M}},\\
+      -(\lambda(t,x)-\nu)\,w(t,x),\quad t\notin{\mathcal{M}},\\
       w(t,x)=\int{\leftlim{w}(t,x')\,\frac{\beta(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}},\quad t\in{\mathcal{M}}.
@@
   \begin{equation*}
-    \Expect{\rdm{V}_t\;\vert\;\Hist_t=H_t}=\prod_{e\in\event(H_t)}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)},
+    \Expect{\rdm{V}_t\;\vert\;\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)},
   \end{equation*}
@@
 In particular, we can use \cref{prop:zeroboost} or \cref{eq:poissdriver} to integrate out the unobserved event times $\rdm{\hat{T}}_k$, as well as the unobserved states $\rdm{\hat{X}}_k$.
-However, we also need to compute likelihoods for data that include observed event times.
+However, we also need to be able to compute likelihoods for data that include observed event times.
 Technically, in such a case, the likelihood is a density on the cone $\mathbb{C}_t=\coprod_{m=0}^{\infty}{\mathbb{C}^m_t}$, where $\mathbb{C}^m_t=\{t\in\halfclosed{0,t}\;\vert\;t_1<t_2<\dots<t_m\}$.
-To avoid having to work with such complex objects, it is convenient to represent the likelihood as a density with respect to a constant-rate Poisson point process.
+To avoid having to work directly with such complex objects, it is convenient to represent the likelihood as a density with respect to a constant-rate Poisson point process.
 This is accomplished as follows.
@@
     \frac{\partial{w}}{\partial{t}}(t,x)=&\int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x',x)\,\dd{x'}}-
-    \int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}\\
+    \int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
+    -\lambda(t,x)\,w(t,x)\\
     &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
-    -\int{w(t,x)\,\pi(t,x,x')\,\dd{x'}}\right\}\\
-    &\qquad-\lambda(t,x)\,w(t,x).
+    -\int{w(t,x)\,\pi(t,x,x')\,\dd{x'}}\right\}.\\
   \end{aligned}
 \end{equation*}
-Equivalently, we write:
+This is equivalent to
 \begin{equation}\label[pluralequation]{eq:mixedKFE}
   \begin{aligned}
-    \frac{\partial{w}}{\partial{t}}(t,x)=&\int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x',x)\,\dd{x'}}
-    -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}\\
-    &\qquad-\lambda(t,x)\,w(t,x)\\
+    \frac{\partial{w}}{\partial{t}}(t,x)
     =&\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
     -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\quad t\notin{\mathcal{M}},\\
-    w(e,x)=&\int{\leftlim{w}(e,x')\,\pi(e,x',x)\,B(e,x',x)\,\dd{x'}}\\
-    =&\tfrac{1}{\nu}\,\int{\leftlim{w}(e,x')\,\alpha(e,x',x)\,\dd{x'}},\quad e\in\mathcal{M}.
+    w(e,x)=&\int{\leftlim{w}(e,x')\,\pi(e,x',x)\,B(e,x',x)\,\dd{x'}}
+    =\tfrac{1}{\nu}\,\int{\leftlim{w}(e,x')\,\alpha(e,x',x)\,\dd{x'}},\quad e\in\mathcal{M}.
   \end{aligned}
@@
 
-The same reasoning shows how to compute expectations of a real-valued process $\rdm{V}_t$ directed by $\X_t$.
+%% We can re-express the history-process density (\cref{eq:Hdens}) using a filter equation as well.
+%% Specifically, one can readily verify that if $H_t$ is a history and $w=w(t,x,u)$ satisfies
+%% \begin{equation}\label{eq:Hdens_filt}
+%%   \begin{aligned}
+%%     \frac{\partial{w}}{\partial{t}}(t,x,u)=&-w(t,x,u)\,\sum_{u'}{\int{\alpha_{u'}(t,x,x')\,\dd{x'}}}\\
+%%     &\quad+\sum_{e\in\event{H}{t}}{\delta(t,e)\,\left\{\sum_{u'}{\int{w(t,x',u')\,\frac{\alpha_{u'}(t,x',x)}{\nu}\,\dd{x'}}}-w(t,x,u)\right\}},
+%%   \end{aligned}
+%% \end{equation}
+%% then $\pi^H(\dd{t_1}\cdots\dd{t_K})=\nu^K\,e^{-\nu\,t}\,\sum_u{\int{w(t,x,u)\,\dd{x'}}}\,\dd{t_1}\cdots\dd{t_K}$,
+%% where $K=\left\vert\event{H}{t}\right\vert$.
+%% \Cref{eq:Hdens_filt} is the filter equation, relative to $\nu$, with null driver, boost $B(t,x,x',u,u')=\alpha_{u'}(t,x,x')$, and decay $\lambda(t,x,u)=\sum_{u'}{\int{\alpha_{u'}(t,x,x')\,\dd{x'}}}$.
+
+The same kind of reasoning shows how to compute expectations of a real-valued process $\rdm{V}_t$ directed by $\X_t$.
 As before, we will express the expectation $\Expect{\rdm{V}_t}$ as a density with respect to the rate-$\nu$ Poisson point process.
@@
   \begin{equation*}
-    \Expect{\rdm{V}_t\;\vert\;\Hist_t}=\prod_{e\in{\event(\Hist_t)}}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)}.
+    \Expect{\rdm{V}_t\;\vert\;\Hist_t}=\prod_{e\in{\event{\Hist}{t}}}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)}.
   \end{equation*}
@@
       &\quad+\sum_{e\in\mathcal{M}}{\delta(t,e)\,\left\{\int{w(t,x')\,\pi(t,x',x')\,C(t,x',x)\,\dd{x'}}
-        -\int{w(t,x)\,\pi(t,x,x')\,\dd{x'}}\right\}}\\
+        -\int{w(t,x)\,\pi(t,x,x')\,\dd{x'}}\right\}},\\
     \end{aligned}
@@
 
-\begin{remark}
-  Suppose $\X_t$, $\Y_t$ have KFEs
-  \begin{equation}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)=
-      &\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\\
-      \frac{\partial{w}}{\partial{t}}(t,y)=
-      &\int{w(t,y')\,\beta(t,y',y)\,\dd{y'}}
-      -\int{w(t,y)\,\beta(t,y,y')\,\dd{y'}},
-    \end{aligned}
-  \end{equation}
-  where $\beta(t,y,y')=\alpha(t,y',y)\,\left(1+\frac{1}{\nu}\sum_{e\in\mathcal{M}}{\delta(t,e)}\right)$.
-\end{remark}
-
 \end{document}

### 2023-07-31 ms.Rnw
good statements and proofs?

@@
 
-\paragraph{Likelihood conditional on history}
+\paragraph{Likelihood for pruned genealogies}
 
@@
 
-Our second result concerns the likelihood of a given obscured genealogy conditional on the history.
+Next, we show how the likelihood of a pruned genealogies, unconditional on the history, can be computed.
+For this, we use the filter equation technology developed in the Appendix.
+In particular, the following theorem follows immediately from \cref{prop:fullfilt}.
+
+\begin{thm}
+  Suppose that $P^*_T$ is a pruned genealogy, for $T>0$.
+  Let $y^*=Y^{\Prune^*_T}$ and, for $x\in\Xspace$, $u\in\Jumps$, $t\le{T}$, define
+  \begin{equation*}
+    \phi_u\big(t,x)=\BinRatio{n(x)}{\ell(y^*_t)}{r^u}{s(\yt^*_t,y^*_t)}\,Q_u(\yt^*_t,y^*_t),
+  \end{equation*}
+  where $Q_u$ is as in \cref{thm:pruned_lik}.
+  Suppose $w=w(t,x)$ satisfies the filter equation
+  \begin{equation}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}=
+      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(t,x)\,\dd{x'}}}
+      -\sum_u{\int{w(t,x)\,\alpha_u(t,x',x)\,\dd{x'}}}\\
+      &\ +\sum_{e\in\event{P^*}{T}}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(t,x)\,\dd{x'}}}-w(t,x)\right\}},
+    \end{aligned}
+  \end{equation}
+  with the initial condition $w(0,x)=p_0(x)$.
+  Then the likelihood of $P^*_T$ is
+  \begin{equation*}
+    \lik(P^*_T)=\int{w(T,x)\,\dd{x}}.
+  \end{equation*}
+\end{thm}
+
+\paragraph{Likelihood for obscured genealogies}
+
+Our next result concerns the likelihood of a given obscured genealogy conditional on the history.
 
@@
 
-%% \AAK{Two approaches possible:}
-%% \begin{enumerate}
-%% \item Use the filter equations developed in the Appendix to compute $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}$.
-%%   In effect, we compute $\Prob{\Prune_t}$ and then sum over paintings $y$.
-%% \item Change the order of the summation: $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}=\Expect{\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}\;\vert\;\Hist_t}}$.
-%%   That is, conditional on $\Hist_t$, the likelihood of $\Vis_t$ is sum over all $\Prune_t$ such that $\obs(\Prune_t)=\Vis_t$.
-%%   Therefore, any importance sampling scheme for $\Prune_t$ will do.
-%%   In particular, we choose a scheme that paints the tree forward in time, driven by a mimic of $\X_t$.
-%%   We show that this is equivalent to the specific filter equation.
-
-%%   It may be useful in this to first show that $\Prob{\Prune_t|\Hist_t}$ can be expressed as an expectation over $y$, where $y$ is constrained to be equal $y^{\Prune_t}$.
-%%   Then, show that the sum of the $\Prune_t$-constraint indicator functions is equal to the $\Vis_t$-constraint indicator function.
-%% \end{enumerate}
-
-%% \begin{figure}
-%%   \begin{center}
-%%     \input{figs/thm1diag}
-%%   \end{center}
-%%   \caption{\label{fig:thm1diag}}
-%% \end{figure}
-
-%% \begin{center}
-%%   \input{figs/triangle1}
-%% \end{center}
-
-%% \begin{equation*}
-%%   \frac{r!}{(r-s)!}\cdot\frac{(n-\ell)!}{n!}\cdot\frac{(n-r)!}{(n-r-\ell+s)!}
-%%   =\frac{r!(n-r)!}{n!}\cdot\frac{(n-\ell)!}{(n-\ell-r+s)!(r-s)!}
-%%   =\BinRatio{n}{\ell}{r}{s}
-%% \end{equation*}
-
-\paragraph{Integrating out the history: filter equations}
-
-Next, we derive expressions for the likelihood of pruned and obscured genealogies, unconditional on the history.
-For this, we use the filter equation technology developed in the Appendix.
-In particular, the following theorem follows immediately from \cref{prop:fullfilt}.
-
-\begin{thm}
-  Suppose that $P^*_T$ is a pruned genealogy, for $T>0$.
-  Let $y^*=Y^{\Prune^*_T}$ and, for $x\in\Xspace$, $u\in\Jumps$, $t\le{T}$, define
-  \begin{equation*}
-    \phi_u\big(t,x)=\BinRatio{n(x)}{\ell(y^*_t)}{r^u}{s(\yt^*_t,y^*_t)}\,Q_u(\yt^*_t,y^*_t),
-  \end{equation*}
-  where $Q_u$ is as in \cref{thm:pruned_lik}.
-  Suppose $w=w(t,x)$ satisfies the filter equation
-  \begin{equation}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}=
-      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(t,x)\,\dd{x'}}}
-      -\sum_u{\int{w(t,x)\,\alpha_u(t,x',x)\,\dd{x'}}}\\
-      &\ +\sum_{e\in\event{P^*}{T}}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(t,x)\,\dd{x'}}}-w(t,x)\right\}},
-    \end{aligned}
-  \end{equation}
-  with the initial condition $w(0,x)=p_0(x)$.
-  Then the likelihood of $P^*_T$ is
-  \begin{equation*}
-    \lik(P^*_T)=\int{w(T,x)\,\dd{x}}.
-  \end{equation*}
-\end{thm}
-
 The final result shows how to compute the likelihood of an obscured genealogy.
+It is an immediate consequence of \cref{thm:obsc_lik,prop:fullfilt}.
 
@@
       &\ -\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
-      &\ +\sum_{e\in\event{P^*}{T}}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u(t,x,y',y)\,\dd{x'}}}-w(t,x)\right\}},
+      &\ +\sum_{e\in\event{V^*}{T}}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u(t,x,y',y)\,\dd{x'}}}-w(t,x)\right\}},
     \end{aligned}
@@
   \end{equation}
-  Suppose that $B(t,x,x')>0$ and $\lambda(t,x)\in\R$ are given measurable functions.
+  Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)\in\R$ are given measurable functions.
   Let $\mathcal{M}\subseteq\Rp$ be countable such that $\mathcal{M}\cap[0,t]$ is finite for all $t$.
-  Let $\nu>0$.
   We say that the equation
@@
       -\lambda(t,x)\,w(t,x)\\
-      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\frac{\beta(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}
-      +\nu\,w(t,x).
+      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}.
     \end{aligned}
   \end{equation}
-  is the \emph{filter equation} with \emph{driver} $\X_t$, \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$, \emph{relative to} $\nu$.
+  is the \emph{filter equation} with \emph{driver} $\X_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$.
 \end{defn}
 
-\begin{remark}
+\begin{remark}\label{rem:equivfilt}
   \Cref{eq:filtereq} is equivalent to
@@
       -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
-      -(\lambda(t,x)-\nu)\,w(t,x),\quad t\notin{\mathcal{M}},\\
-      w(t,x)=\int{\leftlim{w}(t,x')\,\frac{\beta(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}},\quad t\in{\mathcal{M}}.
+      -\lambda(t,x)\,w(t,x),\quad t\notin{\mathcal{M}},\\
+      w(t,x)=\int{\leftlim{w}(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}},\quad t\in{\mathcal{M}}.
     \end{gathered}
@@
 \begin{lemma}\label{lemma:filtereq}
-  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
+  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ and $B>0$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
   \begin{equation}\label{eq:filterlemma}
@@
   Suppose $\rdm{V}_t$ is a real-valued random process such that
-  \begin{equation*}
-    \Expect{\rdm{V}_t\;\vert\;\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)},
-  \end{equation*}
-  for some given measurable functions $B>{0}$, $Q\in\{0,1\}$.
+  \begin{equation}\label{eq:Vprod}
+    \Expect{\rdm{V}_t\;\vert\;\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{B(e,\leftlim{X}_e,X_e)},
+  \end{equation}
+  for some given measurable function $B\ge{0}$.
   If $w$ satisfies the filter equation,
   \begin{equation*}
-    \begin{aligned}
     \frac{\partial{w}}{\partial{t}}(t,x)
-      =&\int{w(t,x')\,\alpha(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\alpha(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
-      &\qquad-\int{w(t,x)\,\alpha(t,x,x')\,\Big(1-Q(t,x,x')\Big)\,\dd{x'}},
-    \end{aligned}
+    =\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
+    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
   \end{equation*}
@@
 \begin{proof}
-  Apply \cref{lemma:filtereq} with $\beta(t,x,x')=\alpha(t,x,x')\,Q(t,x,x')$.
+  Define $Q(t,x,x')=\Indicator{B(t,x,x')>0}$ and apply \cref{lemma:filtereq} with
+  \begin{equation*}
+    \begin{gathered}
+      \beta(t,x,x')=\alpha(t,x,x')\,Q(t,x,x'),
+      \qquad
+      \lambda=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\dd{x'}}.
+    \end{gathered}
+  \end{equation*}
 \end{proof}
@@
 
-Filter equations allow us to pass easily between equivalent representations of a process.
-For example, an equivalent way of representing $\X_t$ is in terms of its embedded chain and event times.
-Let $\rdm{\hat{X}}_k$ be the embedded chain of $\X_t$ and let $\rdm{\hat{T}}_k$ be the point process of its event times.
-It is elementary that
-\fontsize{10pt}{12pt}\selectfont
-\begin{equation*}
-  \begin{gathered}
-    \Prob{\rdm{\hat{X}}_k=\hat{X}_k\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_k=\hat{T}_k}=\frac{\alpha(\hat{T}_k,\hat{X}_{k-1},\hat{X}_k)}{\int{\alpha(\hat{T}_k,\hat{X}_{k-1},x')\dd{x'}}},\\
-    \Prob{\rdm{\hat{T}}_k>\rdm{\hat{T}}_{k-1}+t\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_{k-1}=\hat{T}_{k-1}}=\exp{\left(-\int_{0}^{t}{\int{\alpha(\hat{T}_{k-1}+s,\hat{X}_{k-1},x')\,\dd{x'}}\,\dd{s}}\right)}.
-  \end{gathered}
-\end{equation*}
-\normalfont
-Making the definitions,
-\begin{equation}\label[pluralequation]{eq:nudefs}
-  \begin{gathered}
-    A(t,x)=\int{\alpha(t,x,x')\,\dd{x'}}, \qquad
-    \pi(t,x,x')=\frac{\alpha(t,x,x')}{A(t,x)},\\
-    B(t,x,x')=\frac{A(t,x)}{\nu}, \qquad
-    \lambda(t,x)=A(t,x)-\nu,
-  \end{gathered}
-\end{equation}
-we can rewrite the KFE as
-\begin{equation}\label{eq:poissdriver}
-  \frac{\partial{w}}{\partial{t}}(t,x)=
-  \int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
-  -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
-  -\lambda(t,x)\,w(t,x).
-\end{equation}
-Here, $\nu>0$ is the intensity of a time-homogenous Poisson process.
-Note that $\pi$ is the probability kernel of the embedded chain $\rdm{\hat{X}}$ and $A(t,x)$ is the intensity of the $\rdm{\hat{T}}_k$ process.
-We recognize this equation as the filter equation with boost $B$, decay $\lambda$, and driver generated by $\nu\,\pi(t,x,x')$.
-It corresponds to the following procedure for simulating $\X_t$:
-\begin{compactenum}[(a)]
-\item Simulate jump times according to the rate-$\nu$ Poisson process.
-\item Simulate the embedded chain $\rdm{\hat{X}}_k$ using the kernel $\pi$.
-\item Weight the realization by the product of the $B$ factors.
-  Note that this makes the appropriate importance-sampling correction.
-\end{compactenum}
+The following result extends \cref{prop:zeroboost} to the situation where $\mathcal{M}\ne\emptyset$, \ie where the times of some of the events in the history are observed.
 
-By integrating equations of this kind, we effectively compute expectations with respect to the $\X_t$ process.
-In particular, we can use \cref{prop:zeroboost} or \cref{eq:poissdriver} to integrate out the unobserved event times $\rdm{\hat{T}}_k$, as well as the unobserved states $\rdm{\hat{X}}_k$.
-However, we also need to be able to compute likelihoods for data that include observed event times.
-Technically, in such a case, the likelihood is a density on the cone $\mathbb{C}_t=\coprod_{m=0}^{\infty}{\mathbb{C}^m_t}$, where $\mathbb{C}^m_t=\{t\in\halfclosed{0,t}\;\vert\;t_1<t_2<\dots<t_m\}$.
-To avoid having to work directly with such complex objects, it is convenient to represent the likelihood as a density with respect to a constant-rate Poisson point process.
-This is accomplished as follows.
-
-Suppose the finite set of observed event times $\mathcal{M}\subseteq\halfclosed{0,t}$ is fixed and $|\mathcal{M}|=M$.
-Let $\nu>0$ be a constant rate and consider the mixed point process with intensity $\nu+\sum_{e\in\mathcal{M}}\delta(t,e)$.
-This point process has events at $e\in\mathcal{M}$ with probability 1 as well as additional, Poisson-distributed random event times.
-Using this as the driver in \cref{eq:poissdriver}, we obtain the filter equation
+\begin{prop}\label{prop:fullfilt}
+  Let $\X_t$, $\Hist_t$, $B$, and $\rdm{V}_t$ be as in \cref{prop:zeroboost}.
+  Suppose that $\mathcal{M}\subseteq\halfclosed{0,t}$ is a fixed, finite set of observed event times.
+  Let $\Expect{\rdm{V}_t\;\vert\;\mathcal{M}}$ denote the density of $\rdm{V}_t$ conditional on $\mathcal{M}\subseteq\event{\Hist}{t}$.
+  If $w(t,x)$ satisfies $w(0,x)=p_0(x)$ and the filter equation
   \begin{equation*}
     \begin{aligned}
-    \frac{\partial{w}}{\partial{t}}(t,x)=&\int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x',x)\,\dd{x'}}-
-    \int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
-    -\lambda(t,x)\,w(t,x)\\
-    &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
-    -\int{w(t,x)\,\pi(t,x,x')\,\dd{x'}}\right\}.\\
+      \frac{\partial{w}}{\partial{t}}
+      =&\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
+      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
+      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\},
     \end{aligned}
   \end{equation*}
-This is equivalent to
-\begin{equation}\label[pluralequation]{eq:mixedKFE}
-  \begin{aligned}
-    \frac{\partial{w}}{\partial{t}}(t,x)
-    =&\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
-    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\quad t\notin{\mathcal{M}},\\
-    w(e,x)=&\int{\leftlim{w}(e,x')\,\pi(e,x',x)\,B(e,x',x)\,\dd{x'}}
-    =\tfrac{1}{\nu}\,\int{\leftlim{w}(e,x')\,\alpha(e,x',x)\,\dd{x'}},\quad e\in\mathcal{M}.
-  \end{aligned}
-\end{equation}
-\Cref{eq:mixedKFE} is a filter equation relative to $\nu$ with observed events $\mathcal{M}$.
-It defines a probability measure $\mu_M$ on $\Xspace\times{\mathbb{C}^M_t}$.
-Specifically, $w(t,x)$ is the Radon-Nikodym derivative of $\mu_M$ with respect to the product of the measure on $\Xspace$ and the probability measure of the rate-$\nu$ Poisson process on $\mathbb{C}^M_t$.
-That is,
-\begin{equation*}
-  \mu^M_t(\dd{x}\,\dd{t_1}\,\cdots\,\dd{t_M})=\nu^{M}\,e^{-\nu\,t}\,w(t,x)\,\dd{x}\,\dd{t_1}\,\cdots\,\dd{t_M}.
-\end{equation*}
-
-%% We can re-express the history-process density (\cref{eq:Hdens}) using a filter equation as well.
-%% Specifically, one can readily verify that if $H_t$ is a history and $w=w(t,x,u)$ satisfies
-%% \begin{equation}\label{eq:Hdens_filt}
-%%   \begin{aligned}
-%%     \frac{\partial{w}}{\partial{t}}(t,x,u)=&-w(t,x,u)\,\sum_{u'}{\int{\alpha_{u'}(t,x,x')\,\dd{x'}}}\\
-%%     &\quad+\sum_{e\in\event{H}{t}}{\delta(t,e)\,\left\{\sum_{u'}{\int{w(t,x',u')\,\frac{\alpha_{u'}(t,x',x)}{\nu}\,\dd{x'}}}-w(t,x,u)\right\}},
-%%   \end{aligned}
-%% \end{equation}
-%% then $\pi^H(\dd{t_1}\cdots\dd{t_K})=\nu^K\,e^{-\nu\,t}\,\sum_u{\int{w(t,x,u)\,\dd{x'}}}\,\dd{t_1}\cdots\dd{t_K}$,
-%% where $K=\left\vert\event{H}{t}\right\vert$.
-%% \Cref{eq:Hdens_filt} is the filter equation, relative to $\nu$, with null driver, boost $B(t,x,x',u,u')=\alpha_{u'}(t,x,x')$, and decay $\lambda(t,x,u)=\sum_{u'}{\int{\alpha_{u'}(t,x,x')\,\dd{x'}}}$.
-
-The same kind of reasoning shows how to compute expectations of a real-valued process $\rdm{V}_t$ directed by $\X_t$.
-As before, we will express the expectation $\Expect{\rdm{V}_t}$ as a density with respect to the rate-$\nu$ Poisson point process.
-
-\begin{prop}\label{prop:fullfilt}
-  Suppose $\X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{t}}(t,x)=\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
-    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
-  \end{equation*}
-  Let $\Hist_t$ be its history process.
-  Suppose that $\mathcal{M}\subseteq\halfclosed{0,t}$, $|\mathcal{M}|=M$, is a fixed, finite set of observed event times.
-  Let $B>{0}$ and $Q\in\{0,1\}$ be given measurable functions.
-  Let $\rdm{V}_t$ be a non-negative real-valued random process such that
-  \begin{equation*}
-    \Expect{\rdm{V}_t\;\vert\;\Hist_t}=\prod_{e\in{\event{\Hist}{t}}}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)}.
-  \end{equation*}
-  Let $\nu>0$ be a fixed rate and let $\X'_t$ be the pure-jump process generated by the rates $\alpha(t,x,x')\,Q(t,x,x')$.
-  Suppose that $w(t,x)$ satisfies the filter equation (relative to $\nu$) with driver $\X'_t$, boost $B'$, observation events $\mathcal{M}$, and decay
-  \begin{equation*}
-    \lambda(t,x)=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\,\dd{x'}}.
-  \end{equation*}
-  Then the measure $\Expect{\rdm{V}_t}$ has density $\int{w(t,x)\,\dd{x}}$ with respect to the rate-$\nu$ Poisson point process.
+  then $\Expect{\rdm{V}_t\;\vert\;\mathcal{M}}=\int{w(t,x)\,\dd{x}}$.
 \end{prop}
 \begin{proof}
-  We rewrite the filter equation of \cref{prop:zeroboost} using the rate-$\nu$ Poisson driver:
+  Suppose $s\in[0,t]$ and let
+  \begin{equation*}
+    w(s,x)=\Expect{\prod_{e\in\event{\Hist}{s}}{B\left(e,\Xt_e,\X_e\right)}}.
+  \end{equation*}
+  That is, $w(s,x)$ is the expectation (over histories) of the partial products of \cref{eq:Vprod}.
+  If $s\notin{\mathcal{M}}$, then for $\Delta>0$ sufficiently small,
   \begin{equation*}
     \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)
-      =&\int{w(t,x')\,\alpha(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
-      =&\int{w(t,x')\,\nu\,\pi(t,x',x)\,\frac{A(t,x')\,B(t,x',x)}{\nu}\,\dd{x'}}
-      -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}\\
-      &\quad-\lambda(t,x)\,w(t,x),
+      w(s,x)=&\left(1-\Delta\,\int{\alpha(s-\Delta,x,x')\,\dd{x'}}\right)\,w(s-\Delta,x)\\
+      &\qquad+\int{\alpha(s-\Delta,x',x)\,B(s-\Delta,x',x)\,w(s-\Delta,x')\,\dd{x'}}+o(\Delta).
     \end{aligned}
   \end{equation*}
-  where
+  Taking the limit as $\Delta\downarrow{0}$, we obtain
   \begin{equation*}
-    \begin{gathered}
-      A(t,x)=\int{\alpha(t,x,x')\,Q(t,x,x')\dd{x'}}, \qquad
-      \lambda(t,x)=\int{\alpha(t,x,x')\,\dd{x'}}-\nu,\\
-      \pi(t,x,x')=\frac{\alpha(t,x,x')\,Q(t,x,x')}{A(t,x)}.
-    \end{gathered}
+    \frac{\partial{w}}{\partial{s}}(s,x)
+    =\int{w(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}
+    -\int{w(s,x)\,\alpha(s,x,x')\,\dd{x'}}.
   \end{equation*}
-  Replacing the Poisson driver with $\nu+\sum_{e\in\mathcal{M}}{\delta(t,e)}$ as before, we have
+  If $s\in\mathcal{M}$, then
   \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)
-      =&\int{w(t,x')\,\nu\,\pi(t,x',x)\,C(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}-\lambda(t,x)\,w(t,x)\\
-      &\quad+\sum_{e\in\mathcal{M}}{\delta(t,e)\,\left\{\int{w(t,x')\,\pi(t,x',x')\,C(t,x',x)\,\dd{x'}}
-        -\int{w(t,x)\,\pi(t,x,x')\,\dd{x'}}\right\}},\\
-    \end{aligned}
+    w(s,x)=\int{\wt(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}.
   \end{equation*}
-  where
-  \begin{equation*}
-    C(t,x,x')=\frac{A(t,x)\,B(t,x,x')}{\nu}.
-  \end{equation*}
-  This simplifies to
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)
-      =&\int{w(t,x')\,\alpha(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\alpha(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
-      &\quad+\int{w(t,x)\,\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\,\dd{x'}}\\
-      &\quad+\sum_{e\in\mathcal{M}}{\delta(t,e)\,\left\{\int{w(t,x')\,\frac{\alpha(t,x',x)\,Q(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}}
-        -w(t,x)\right\}}.\\
-    \end{aligned}
-  \end{equation*}
-  In the latter, we recognize the filter equation of the statement.
+  By \cref{rem:equivfilt}, the result follows.
 \end{proof}
 
+%% Filter equations allow us to pass easily between equivalent representations of a process.
+%% For example, an equivalent way of representing $\X_t$ is in terms of its embedded chain and event times.
+%% Let $\rdm{\hat{X}}_k$ be the embedded chain of $\X_t$ and let $\rdm{\hat{T}}_k$ be the point process of its event times.
+%% It is elementary that
+%% \fontsize{10pt}{12pt}\selectfont
+%% \begin{equation*}
+%%   \begin{gathered}
+%%     \Prob{\rdm{\hat{X}}_k=\hat{X}_k\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_k=\hat{T}_k}=\frac{\alpha(\hat{T}_k,\hat{X}_{k-1},\hat{X}_k)}{\int{\alpha(\hat{T}_k,\hat{X}_{k-1},x')\dd{x'}}},\\
+%%     \Prob{\rdm{\hat{T}}_k>\rdm{\hat{T}}_{k-1}+t\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_{k-1}=\hat{T}_{k-1}}=\exp{\left(-\int_{0}^{t}{\int{\alpha(\hat{T}_{k-1}+s,\hat{X}_{k-1},x')\,\dd{x'}}\,\dd{s}}\right)}.
+%%   \end{gathered}
+%% \end{equation*}
+%% \normalfont
+%% Fixing $\nu>0$ and making the definitions,
+%% \begin{equation}\label[pluralequation]{eq:nudefs}
+%%   \begin{gathered}
+%%     A(t,x)=\int{\alpha(t,x,x')\,\dd{x'}}, \qquad
+%%     \pi(t,x,x')=\frac{\alpha(t,x,x')}{A(t,x)},\\
+%%     B(t,x,x')=\frac{A(t,x)}{\nu}, \qquad
+%%     \lambda(t,x)=A(t,x)-\nu,
+%%   \end{gathered}
+%% \end{equation}
+%% we can rewrite the KFE as
+%% \begin{equation}\label{eq:poissdriver}
+%%   \frac{\partial{w}}{\partial{t}}(t,x)=
+%%   \int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
+%%   -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
+%%   -\lambda(t,x)\,w(t,x).
+%% \end{equation}
+%% Here, $\nu$ is the intensity of a time-homogenous Poisson process.
+%% Note that $\pi$ is the probability kernel of the embedded chain $\rdm{\hat{X}}$ and $A(t,x)$ is the intensity of the $\rdm{\hat{T}}_k$ process.
+%% We recognize this equation as the filter equation with boost $B$, decay $\lambda$, and driver generated by $\nu\,\pi(t,x,x')$.
+%% It corresponds to the following procedure for simulating $\X_t$:
+%% \begin{compactenum}[(a)]
+%% \item Simulate jump times according to the rate-$\nu$ Poisson process.
+%% \item Simulate the embedded chain $\rdm{\hat{X}}_k$ using the kernel $\pi$.
+%% \item Weight the realization by the product of the $B$ factors.
+%%   Note that this makes the appropriate importance-sampling correction.
+%% \end{compactenum}
+
 \end{document}

### 2023-08-01 ms.Rnw
edits

@@
 %% Conditional on $\Hist_t$, both $\X_t$ and $\U_t$ are deterministic, as of course are the Markov chains of event times $\rdm{\hat{T}}_k$ and the embedded chains $\rdm{\hat{X}}_k=\X_{\rdm{\hat{T}}_k}$ and $\rdm{\hat{U}}_k=\U_{\rdm{\hat{T}}_k}$.
-Conditional on $\Hist_t$, both $\X_t$ and $\U_t$ are deterministic as are the embedded chain, $(\X_k,\U_k)$, and the point process of event times $\rdm{\hat{T}}_k$.
-(Recall that $\rdm{\hat{X}}_k=\rdm{X}_{\rdm{\hat{T}}_k}$.)
+Conditional on $\Hist_t$, both $\X_t$ and $\U_t$ are deterministic as are the embedded chain, $(\rdm{\hat{X}}_k,\rdm{\hat{U}}_k)$, and the point process of event times $\rdm{\hat{T}}_k$.
+(Recall that $\rdm{\hat{X}}_k=\rdm{X}_{\rdm{\hat{T}}_k}$ and $\rdm{\hat{U}}_k=\rdm{U}_{\rdm{\hat{T}}_k}$.)
 The non-explosiveness assumption implies that, for every $t$, $\rdm{K}_t\coloneq\big\vert\event{\Hist}{t}\cap[0,t]\big\vert<\infty$ a.s.
@@
 \item \emph{tip nodes}, which represent labeled extant lineages;
-\item \emph{internal nodes}, which represent ancestral events;
+\item \emph{branch nodes}, which represent ancestral events;
 \item \emph{sample nodes}, which represent labeled samples.
 \end{inparaenum}
-More formally, we can take a genealogy, $G$, to be a finite sequence of internal nodes, together with a time.
+More formally, we can take a genealogy, $G$, to be a finite sequence of nodes, together with a time.
 Given $G$, the time is denoted $T(G)$; this is the time corresponding to the extant lineages.
@@
 \begin{enumerate}[(a)]
-\item A birth-type event at time $t$ results in the creation of one new internal node for each parent lineage.
+\item A birth-type event at time $t$ results in the creation of one new branch node for each parent lineage.
   In particular, if $j$ is one of the parent lineages, and $\Node{p}$ is the new node, then $T(\Node{p})=t$, $\Anc(\Node{p})$ is the node that was parent to $j$ prior to the event.
@@
 \item In a death-type event, all the lineages $j$ that die are removed.
-  Internal nodes without children are then recursively removed.
+  Branch nodes without children are then recursively removed.
   Sample nodes are never removed.
-\item In a migration-type event, one node is added for each migrating lineage;
+\item In a migration-type event, one branch node is added for each migrating lineage;
   each one takes the migrating lineage as child.
@@
 We therefore describe how the genealogy process is \emph{pruned} to yield the sample-only genealogy.
-Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Prune_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every internal node without children.
-In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and some of the interior nodes of the genealogy.
+Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Prune_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every branch node without children.
+In a pruned genealogy only branch and sample nodes remain, and sample nodes are found at all of the leaves and some of the interior nodes of the genealogy.
 Observe that the pruned genealogy retains information not only about how much ancestry is shared by any pair of sample lineages, but also about where among the demes each lineage was through time.
@@
 Knowing the function $\deme^{\Prune_t}$ is equivalent to knowing the coloring, while $\ct^{\Prune_t}$ determines the locations of events in the genealogy and $\anc^{\Prune_t}$ determines the topology.
-Note in particular that, if $y=Y^{\Prune_t}$, then $y_j(t')\ne\yt_j(t')$ if and only if $t'$ is the time of an event from which lineage $j$ emerges.
+Note in particular that, $Y^{P_t}_j(t')\ne\leftlim{Y}^{P_t}_j(t')$ if and only if $t'$ is the time of an event from which lineage $j$ emerges.
 Note also that, there can be events on a pruned genealogy where the color does not change.
@@
 \end{equation*}
-With this definition, for any $P$, if $y=Y^P$, then $s_i(\yt(t),y(t))$ is the number of lineages in deme $i$ that emerge from an event at time $t$ and $s(\yt(t),y(t))$ is the non-negative integer vector telling how many lineages in each deme emerge at time $t$.
-In particular, $s(y,y)=0$ so that if $e\notin\event{P}{t}$, then $s(\yt(e),y(e))=0$.
-We refer to $s\left(\yt(t),y(t)\right)$ as the \emph{saturation} at time $t$.
+With this definition, for any $P$, if $y=Y^P$, then $s_i(\yt_t,y_t)$ is the number of lineages in deme $i$ that emerge from an event at time $t$ and $s(\yt_t,y_t)$ is the non-negative integer vector telling how many lineages in each deme emerge at time $t$.
+In particular, $s(y,y)=0$ so that if $e\notin\event{P}{t}$, then $s(\yt_e,y_e)=0$.
+We refer to $s(\yt_t,y_t)$ as the \emph{saturation} at time $t$.
 
@@
 For example, if $P^*_t$ has an event in which a lineage moves from deme $i$ to deme $i'$ but there are no $u\in\Jumps$ for which this is possible, then $P^*_t$ is incompatible with the population process itself.
-Let us define the function $Q:\Jumps\times\Yspace\times\Yspace\to\{0,1\}$ so that $Q_u(y,y')=1$ if and only if a change $y\to{y'}$ in a pruned genealogy is compatible with an event of mark $u$ at that time.
+Let us define the function $Q:\Jumps\times\Yspace^2\to\{0,1\}$ so that $Q_u(y,y')=1$ if and only if a change $y\to{y'}$ in a pruned genealogy is compatible with an event of mark $u$ at that time.
 Recall that $n:\Xspace\to\Zp^{\Demes}$ denotes the deme-occupancy function, $r^u$, the production of a jump of mark $u$, $\ell$, the lineage-count function, and $s$, the saturation.
@@
   \begin{equation*}
-    \begin{aligned}
-      \Prob{P^*_t\vert\Hist_t}=&\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}\,Q_{U_e}(\yt^*(e),y^*(e))}\\
-      &\qquad\times\Indicator{\event{P^*}{t}\subseteq{\event{\Hist}{t}}}.
-    \end{aligned}
+    \Prob{P^*_t\vert\Hist_t}=\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(y^*_e)}{r^{U_e}}{s(\yt^*_e,y^*_e)}\,Q_{U_e}(\yt^*_e,y^*_e)}
+    \:\times\:\Indicator{\event{P^*}{t}\subseteq{\event{\Hist}{t}}}.
   \end{equation*}
@@
   Moreover, at each event $e\in\event{\Hist}{t}$, a jump of mark $U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
-  In $P^*_t$, at time $e$, there are $\ell_i=\ell_i(y^*(t))$ lineages in deme $i$, of which $s_i=s_i(\yt^*(t),y^*(t))$ are emergent.
+  In $P^*_t$, at time $e$, there are $\ell_i=\ell_i(y^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,y^*_t)$ are emergent.
   The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
@@
   \begin{equation*}
-    \Prob{P^*_t\vert\Hist_t}=\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}}.
+    \Prob{P^*_t\vert\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{\BinRatio{n(X_e)}{\ell(y^*_e)}{r^{U_e}}{s(\yt^*_e,y^*_e)}}.
   \end{equation*}
@@
     \begin{aligned}
-      \Prob{V^*_t\vert\Hist_t}=&\Expect{\left.\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(\rdm{y}_e)}{r^{U_e}}{s(\leftlim{\rdm{y}}_e,\rdm{y}_e)}\,\frac{\scriptsize\kappa_{V^*_t}(e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}{\scriptsize\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}\;\right\vert\;\Hist_t}\\
-      &\qquad\times\Indicator{\event{V^*}{t}\subseteq\event{\Hist}{t}},
+      \Prob{V^*_t\vert\Hist_t=H_t}=&\Expect{\left.\prod_{e\in\event{H}{t}}{\BinRatio{n(X_e)}{\ell(\rdm{y}_e)}{r^{U_e}}{s(\leftlim{\rdm{y}}_e,\rdm{y}_e)}\,\frac{\scriptsize\kappa_{V^*_t}(e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}{\scriptsize\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}\;\right\vert\;\Hist_t}\\
+      &\qquad\times\Indicator{\event{V^*}{t}\subseteq\event{H}{t}},
     \end{aligned}
@@
   \end{equation*}
-  Now let $\event{\Hist}{t}=\{t_1,t_2,\dots,t_K\}$.
+  Now let $\event{H}{t}=\{t_1,t_2,\dots,t_K\}$.
   $\rdm{\hat{y}}_k$ is the embedded chain of a unique, piecewise-constant, \cadlag\ process $\rdm{y}_t$, with jump times at $t_k$.
-  This construction of $\rdm{y}_t$ obviously guarantees that $\event{P(\rdm{y})}{t}\subseteq\event{\Hist}{t}$.
+  This construction of $\rdm{y}_t$ obviously guarantees that $\event{P(\rdm{y})}{t}\subseteq\event{H}{t}$.
   Moreover, the conditions on $\pi$ guarantee that, for every pruned genealogy $P_t$, $Y^{P_t}$ is a sample path of this process.
@@
 
-  Finally, observe that the condition $\obs(P(\rdm{y}))=V_t$ can be factorized for all obscured genealogies $V_t$.
-  In particular, there is a function $\kappa_{V_t}:[0,t]\times\Yspace^2\to\{0,1\}$ such that $\kappa_{V_t}(s,\eta,\eta')=1$ if and only if
-  there is $y:\Rp\to\Yspace$ such that $\obs(P(y))=V_t$, $\leftlim{y}_s=\eta$, and $y_s=\eta'$, for all $s\in[0,t]$.
+  Finally, observe that the condition $\obs(P(\rdm{y}))=V$ can be factorized for all obscured genealogies $V$.
+  In particular, there is a function $\kappa_{V}:[0,t]\times\Yspace^2\to\{0,1\}$ such that $\kappa_{V}(s,\eta,\eta')=1$ if and only if
+  there is $y:\Rp\to\Yspace$ such that $\obs(P(y))=V$, $\leftlim{y}_s=\eta$, and $y_s=\eta'$, for all $s\in[0,T(V)]$.
   Thus, conditional on $\Hist_t$, if $\event{P(y)}{t}\subseteq\event{\Hist}{t}$,
@@
   \end{equation}
-  with the initial condition $w(0,x,y)=p_0(x)\,\pi_0(y)$.
+  with the initial condition $w(0,x,y)=p_0(x)$.
   Then the likelihood of $V^*_T$ is

### 2023-08-10 ms.Rnw
update

@@
   A.~A.~King,
-  Department of Ecology \& Evolutionary Biology,
-  Center for the Study of Complex Systems, and
+  Department of Ecology \& Evolutionary~Biology,
+  Center for the Study of Complex~Systems, and
   Department of Mathematics,
   University of Michigan,
-  Ann Arbor, MI 48109 USA
+  Ann~Arbor, MI~48109~USA\\
+  Santa~Fe~Institute,
+  1399 Hyde~Park~Road,
+  Santa~Fe, NM~87501~USA
 }
@@
 \urladdr{\href{https://kinglab.eeb.lsa.umich.edu/}{https://kinglab.eeb.lsa.umich.edu/}}
-\author[Lin]{Qianying Lin}
+\author[Lin]{Qianying~Lin}
 \address{
-  Q.-Y. Lin,
+  Q.-Y.~Lin,
   Theoretical Biology and Biophysics,
-  Los Alamos National Laboratory,
-  Los Alamos, NM XXXXX USA
+  Los~Alamos National Laboratory,
+  Los~Alamos, NM~87545~USA
 }
@@
   University of Michigan,
-  Ann Arbor, MI 48109 USA
+  Ann~Arbor, MI~48109~USA
 }

### 2023-08-10 setup.R

@@ -7,7 +7,7 @@ opts_chunk$set(
   echo=FALSE,
   eval=TRUE,
   include=TRUE,
-  dev="cairo_pdf",
+  dev="png",
   dev.args=list(bg='transparent'),
   dpi=300,
   error=FALSE,

### 2023-08-17 ms.Rnw
Ed's comments

@@
 We write $r^u=(r^u_i)_{i\in\Demes}$.
-Note that we lineages that die as a result of an event do not count in the production.
+Note that the lineages that die as a result of an event do not count in the production.
 Also, it is important to note that the parent lineage or lineages, if they survive the event, are always counted in the production.
@@
   In particular, we posit that $\anc^{P}_j(t)=j$ for $t\in\halfopen{t^o_j,t^s_j}$ and $\anc^{P}_j(t)=0$ for $t>t^s_j$, so that the function is well defined for all $j$ and $t$.
+  \ELI{[How is this defined when there is immigration in the system?]}
   %%  Alternatively, one can define $\anc$ using the original construction of Kingman's coalescent \citep{Kingman1982a}.
@@
   \end{equation*}
+  \ELI{[Do we need $Q$, if the binomial ratio is defined to be zero in impossible cases?]}
 \end{thm}
@@
   \end{equation*}
-  Suppose $w=w(t,x)$ satisfies the filter equation
+  Suppose $w=w(t,x,y)$ satisfies the filter equation
   \begin{equation}
@@
   \end{equation*}
+  \ELI{[I was surprised by the assertion of existence of $k_V$ here, and then the proof providing a specific construction---why not assert the exact form?
+      The proof is also unusual---via looking for an importance sampling solution.]}
 \end{thm}
@@
   Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)\in\R$ are given measurable functions.
+  \ELI{[Ranges are defined elementwise, which is not the usual math formalism,
+      e.g., $\lambda:T\times\Xspace\to\mathbb{R}$
+      or just say $\lambda(t,x)$ takes values in $\mathbb{R}$.]}
   Let $\mathcal{M}\subseteq\Rp$ be countable such that $\mathcal{M}\cap[0,t]$ is finite for all $t$.
@@
   is the \emph{filter equation} with \emph{driver} $\X_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$.
+  \ELI{[Why is $B=1$, $\lambda=0$ not equal to Eq.~(A1)?
+      (It is if we have $|M|=0$.)]}
+  \AAK{[The KFE is a special case of a filter equation.]}
 \end{defn}
@@
 \begin{lemma}\label{lemma:filtereq}
-  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ and $B>0$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
+  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ and $B>0$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satisfies the KFE
   \begin{equation}\label{eq:filterlemma}
@@
   Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
+  \ELI{[Note: the last term in eq (A3) is not in the form of $-lambda\,u$ for some $\lambda$.
+      So, Eq (A3) is not a filter equation in the form of Eq (A1).
+  ]}
+  \AAK{[But it is a KFE, for a process with a drift.]}
 \end{lemma}
 \begin{proof}
+  \ELI{[Conditions for differentiating under the integral in the proof?
+      Need $[v^2 \lambda u(t,x,v)]_{v=0}^{\infty} = 0$.
+      That's okay at $v=0$, and a minor constraint on $u$.]}
   \begin{equation*}
@@
   then $\Expect{\rdm{V}_t}=\int{w(t,x)\,\dd{x}}$.
+  \ELI{[$dw/dt$ for $X$ could be $du/dt$ since this is not the same $w$ as the subsequent line.
+      The terminology uses the letter $w$ as a dummy variable, but this makes it harder to talk about the solutions to different KFEs.]}
 \end{prop}
@@
   \end{equation*}
+  \ELI{[Lemma A1 has $B>0$, which only covers the cases when $Q=1$ and $\lambda=0$.
+      But, $B\ge{0}$ seems enough in the proof of lemma A1, as long as one allows a one-sided Dirac $\delta$ at $0$.
+      Maybe the Dirac $\delta$ should all be one-sided in order to integrate
+      correctly on right-continuous functions? ]}
 \end{proof}
@@
   then $\Expect{\rdm{V}_t\;\vert\;\mathcal{M}}=\int{w(t,x)\,\dd{x}}$.
+  \ELI{[(a) line 2: density -> expectation.
+      (b) the term $\delta(t,e)\,w(t,x)$ seems irrelevant:
+      the derivative at a finite set of points only matters when it is a delta function.
+      (c) what is a ``fixed, finite set of event times''?
+      $M_t^H$ is a random variable, so it cannot usually be fixed.
+      $M$ is also treated sometimes as a random variable, even though it is defined to be fixed.
+      I see this means, ``conditional on the event that $M_t^H$ contains $M$'', but we should be explicit, to avoid confusion about whether $M$ is a random variable or not.]}
 \end{prop}
@@
   By \cref{rem:equivfilt}, the result follows.
+  \ELI{[The argument shows that the filter equation satisfies a KFE, but the limit expression seems justified by a similar KFE.
+      Is the point that one KFE implies the other?]}
 \end{proof}
+
+\ELI{[Maybe, Prop A3 could be a version of Prop A2 allowing for discrete transitions at a finite set of times;
+    a combination of cts time and discrete time Markov chains.
+    That is all we need to extend Prop A2 to observation times, which can be viewed just as an extended Markov process including the observation events.
+    That is, to make a distinction between ``observation events'' that give the sequences used to construct the phylogeny, and ``observed events'' which are determined points on the phylogeny.
+    The latter are a realization of a random process, and the former are defined to be fixed a priori.]}
 

### 2024-01-17 ms.Rnw
update
rewrite

--- 2023-08-10 ms.Rnw
@@
-\documentclass[11pt,reqno,final]{amsart}
-\usepackage[round,elide]{natbib}
+\documentclass[10pt,reqno,final]{amsart}
 \input{header}
@@
 <<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
-prefix <- "smgp"
+prefix <- "sgp"
 source("setup.R")
@@
 
+\begin{abstract}
+  We derive an algorithm for the exact computation of the likelihood of an observed transmission tree.
+\end{abstract}
+
 \maketitle
-
-\begin{abstract}
-
-\end{abstract}
 
@@
 
-\section{Mathematical definitions}
+\section{Mathematical preliminaries}
 
-\subsection{Population processes}
+\subsection{Notation}
+
+Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\Xr$), denotes a random element.
+We will be concerned with a variety of stochastic processes, in both discrete and continuous time.
+In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
+In the case of continuous-time processes, we will assume that sample paths are right-continuous with left limits (\ie \cadlag).
+We will frequently need to refer to the left-limit of such a process.
+Accordingly, if $\Zr_t$ is a \cadlag\ random process, we define
+\begin{equation*}
+  \Zrt_t\colonequals
+  \begin{cases}
+    \displaystyle\lim_{t'\,\uparrow\,{t}}\;\Zr_{t'}, & t>0,\\[2ex]
+    \Zr_0, & t=0.\\
+  \end{cases}
+\end{equation*}
+Note that $\Zrt_t$ is thus left-continuous with right limits.
+
+If $Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
+In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
+$\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
+If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its event times in the interval $[0,t]$.
+
+\subsection{Population process}
 
@@
   \begin{center}
-    \resizebox{0.9\linewidth}{!}{\input{figs/model_diagrams}}
+    \resizebox{0.9\linewidth}{!}{
+      \begin{tikzpicture}[scale=1]
+        \tikzstyle{box}=[color=black, draw, fill=white, very thick, minimum size=3em]
+        \tikzstyle{label}=[font=\Large]
+        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
+        \tikzstyle{flow}=[color=black, very thick, >=stealth]
+        \tikzstyle{deme}=[fill=black!20!white]
+        \tikzstyle{modulate}=[color=blue, thick, >=latex]
+        \coordinate (origin) at (0,0);
+        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{A}};
+        \input{figs/seir_diagram}
+        \coordinate (origin) at (9,0);
+        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{B}};
+        \input{figs/siir_diagram}
+        \coordinate (origin) at (0,-4);
+        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{C}};
+        \input{figs/covid_diagram}
+      \end{tikzpicture}
+    }
   \end{center}
   \caption{
-    Examples of compartmental models.
+    Examples of discretely-structured population models.
     Demes are shaded.
+    (A) An SEIRS model.
+    Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
+    Upon recovery ($\lab{R}$), individuals experience immunity from reinfection.
+    If this immunity wanes, they re-enter the susceptible compartment.
+    Pathogen lineages are to be found in hosts within the $\lab{E}$ and $\lab{I}$ compartments only.
+    Accordingly, there are two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
+    If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
     \AAK{[Perhaps another one or two examples here?]}
@@
 
-Another perspective on the Markov processes is to be had from its Markov state transition diagram (\cref{fig:markov_state}).
-
-\begin{figure}
-  \begin{center}
-    \resizebox{0.9\linewidth}{!}{
-      \begin{tikzpicture}[scale=1]
-        \usetikzlibrary{shapes,arrows,positioning}
-        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\large, fill=white, thick, minimum size=5em]
-        \tikzstyle{trans}=[color=black, thick, >=stealth]
-        \node[state] (s0) at (0,0) {$x=(S,E,I,R)$};
-        \node[state] (s1) at (165:7.8) {$x'=(S-1,E+1,I,R)$};
-        \node[state] (s2) at (195:7.1) {$x'=(S,E-1,I+1,R+1)$};
-        \node[state] (s3) at (295:4) {$x'=(S,E,I-1,R+1)$};
-        \node[state] (s4) at (355:9) {$x'=(S+1,E,I,R-1)$};
-        \node[state] (s5) at (18:6) {$x'=(S,E,I,R)$};
-        \draw[trans,->] (s0) -- (s1) node[midway,above,sloped] {$\mathrm{Trans}$};
-        \draw[trans,->] (s0) -- (s2) node[midway,above,sloped] {$\mathrm{Prog}$};
-        \draw[trans,->] (s0) -- (s3) node[midway,above,sloped] {$\mathrm{Recov}$};
-        \draw[trans,->] (s0) -- (s4) node[midway,above,sloped] {$\mathrm{Wane}$};
-        \draw[trans,->] (s0) -- (s5) node[midway,above,sloped] {$\mathrm{Sample}$};
-        \node[font=\Large] (U) at ($(s0)+(-6,-5)$) {$\Jumps=\{\mathrm{Trans},\mathrm{Prog},\mathrm{Recov},\mathrm{Wane},\mathrm{Sample}\}$};
-      \end{tikzpicture}
-    }
-  \end{center}
-  \caption{
-    Markov state transition diagram for an SEIR model.
-    The state is characterized by four numbers, $S$, $E$, $I$, and $R$.
-    From a given state $x$, there are five possible kinds of events $x\mapsto{x'}$ as shown.
-    From the point of view of the induced genealogy process,
-    $\mathrm{Trans}$ (transmission) is of birth type,
-    $\mathrm{Prog}$ (progression) is of migration type,
-    and $\mathrm{Recov}$ (recovery) is of death type,
-    while $\mathrm{Wane}$ (loss or waning of immunity) is of neutral type.
-    Note that, in this formulation, when a $\mathrm{Sample}$ (sampling) event occurs, the state does not change.
-    \label{fig:markov_state}
-  }
-\end{figure}
-
-\paragraph{Mathematical notation}
-
-Denote the underlying probability space by $(\Omega,\Borel,\mathrm{Prob})$.
-Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\rdm{X}$), denotes a random element while the same symbol in regular face ($X$) denotes an element in its range.
-We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\colonequals\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$) and taking values in some space $\Xspace$.
+We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
 In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
@@
 \begin{equation}\label{eq:ic}
-  \Prob{\X_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
+  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
 \end{equation}
@@
 Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
-Let $\N_t$ be the number of jumps that $\X$ has taken by time $t$.
-We assume that $\N_t$ is a simple counting process so that
+Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
+We assume that $\Kr_t$ is a simple counting process so that
 \begin{equation*}
   \begin{gathered}
-    \Prob{\N_{t+\Delta}=n+1\;\vert\;\N_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
-    \Prob{\N_{t+\Delta}>n+1\;\vert\;\N_{t}=n}=o(\Delta),\\
-    \Prob{\X_{t+\Delta}\in\mathcal{E}\;\vert\;\X_{t}=x, \N_{t+\Delta}-\N_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
+    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
+    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
+    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
   \end{gathered}
 \end{equation*}
-We will further assume that $\X_t$ is non-explosive, \ie, that $\Prob{\N_t<\infty}=1$ for all $t$.
-\AAK{[Is this equivalent to non-explosivity? Or merely an implication?]}
+We will further assume that $\Xr_t$ is non-explosive, \ie that $\Prob{\Kr_t<\infty}=1$ for all $t$.
+
+\subsubsection{Kolmogorov forward equation}
 
@@
 and if, moreover, $w(0,x) = p_0(x)$,
-then $\int_{\mathcal{E}}{w(t,x)\,\dd{x}}=\Prob{\X_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
-\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\X_t$.
+then $\int_{\mathcal{E}}\!{w(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
+\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.
 
-Without loss of generality, one can assume, as we do here, that the sample paths $t\mapsto\X_t(\omega)$ for $\omega\in\Omega$ are right-continuous with left limits (\ie \cadlag).
-In fact, all of the processes we will describe in this paper will be taken to have \cadlag\ sample paths, and we will frequently need to refer to their left-limits.
-Accordingly, if $\rdm{Z}_t$ is any \cadlag\ random process, we define
-\begin{equation*}
-  \leftlim{\rdm{Z}}_t\colonequals
-  \begin{cases}
-    \lim_{t'\uparrow{t}}\rdm{Z}_t, & t>0,\\
-    \rdm{Z}_0, & t=0.\\
-  \end{cases}
-\end{equation*}
-Note that $\leftlim{\rdm{Z}}_t$ is thus left-continuous with right limits.
-Denote the set of event times of $\rdm{Z}$ by $\event{\rdm{Z}}{t}=\{e\in[0,t]\;\vert\;\leftlim{\rdm{Z}}_e\ne{\rdm{Z}_e}\}$.
+Another perspective on the Markov processes is to be had from its Markov state transition diagram (\cref{fig:markov_state}).
 
-\paragraph{Structured populations, demes}
+\begin{figure}
+  \begin{center}
+    \resizebox{0.9\linewidth}{!}{
+      \begin{tikzpicture}[scale=1]
+        \usetikzlibrary{shapes,arrows,positioning}
+        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
+        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum size=5em]
+        \tikzstyle{trans}=[color=black, thick, >=stealth]
+        \node[state] (base) at (0,0) {$x=(S,E,I,R)$};
+        \node[state] (trans) at (165:8.4) {$x'=(S-1,E+1,I,R)$};
+        \node[state] (prog) at (195:8) {$x'=(S,E-1,I+1,R+1)$};
+        \node[state] (recov) at (295:4) {$x'=(S,E,I-1,R+1)$};
+        \node[state] (wane) at (355:9) {$x'=(S+1,E,I,R-1)$};
+        \node[state] (sample) at (18:7.5) {$x'=(S,E,I,R)$};
+        \draw[trans,->] (base) -- (trans) node[midway,above,sloped] {$\lab{Trans}$};
+        \draw[trans,->] (base) -- (prog) node[midway,above,sloped] {$\lab{Prog}$};
+        \draw[trans,->] (base) -- (recov) node[midway,above,sloped] {$\lab{Recov}$};
+        \draw[trans,->] (base) -- (wane) node[midway,above,sloped] {$\lab{Wane}$};
+        \draw[trans,->] (base) -- (sample) node[midway,above,sloped] {$\lab{Sample}$};
+        \node[font=\Large] (U) at (95:3.2) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
+      \end{tikzpicture}
+    }
+  \end{center}
+  \caption{
+    Markov state transition diagram for the SEIRS model depicted in \cref{fig:examples}A.
+    The state, $x$, is characterized by four numbers, $S$, $E$, $I$, and $R$.
+    From a given state $x$, there are five possible kinds of jumps $x\mapsto{x'}$.
+    Accordingly, the set, $\Jumps$, of jump marks has five elements.
+    Each of these is of a different type:
+    $\lab{Trans}$ (transmission) is of birth type,
+    $\lab{Prog}$ (progression) is of migration type,
+    $\lab{Recov}$ (recovery) is of death type,
+    $\lab{Sample}$ (sampling) is of sample type,
+    and $\lab{Wane}$ (loss or waning of immunity) is of neutral type.
+    See \cref{sec:jump_types} for a description of these jump types.
+    Note that, in this formulation, when a sampling event occurs, the state does not change.
+    \label{fig:markov_state}
+  }
+\end{figure}
+
+\subsubsection{Structured populations, demes}
 
@@
 
-For any $i\in\Demes$, we let $n_i(\X_t)$ denote the number of lineages present in deme $i$ at time $t$, \ie the \emph{occupancy} of deme $i$.
-Thus $n(\X_t)\in\Zp^{\Demes}$ is the vector of deme occupancies.
+For any $i\in\Demes$, we let $n_i(\Xr_t)$ denote the number of lineages present in deme $i$ at time $t$, \ie the \emph{occupancy} of deme $i$.
+Thus $n(\Xr_t)\in\Zp^{\Demes}$ is the vector of deme occupancies.
 
-\paragraph{Jump marks}
+\subsubsection{Jump marks}
 
@@
 \end{equation*}
-In \cref{fig:markov_state}, we use the marks to distinguish biologically distinct events.
+In \cref{fig:markov_state}, we use the marks to distinguish events of different types.
 Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.
 
-Let us define the \emph{jump mark} process, $\U_t$, to be the mark of the latest jump as of time $t$.
-As usual, we take the sample paths, $t\mapsto{\U_t(\omega)}$ for $\omega\in\Omega$, to be \cadlag.
-Observe that $(\X_t,\U_t)$ is a Markov process but that $\U_t$ is not.
+Let us define the \emph{jump mark} process, $\Ur_t$, to be the mark of the latest jump as of time $t$.
+As usual, we take the sample paths of $\Ur_t$ to be \cadlag.
+Observe that $(\Xr_t,\Ur_t)$ is a Markov process, though $\Ur_t$ is not.
 
@@
 
-\paragraph{SEIRS model}
+\subsubsection{SEIRS model}
 
-\paragraph{SIIR model}
+\subsubsection{SIIR model}
 
-\paragraph{Linear birth-death model}
+\subsubsection{Linear birth-death model}
 
-\paragraph{Moran model and the Kingman coalescent}
+\subsubsection{Moran model and the Kingman coalescent}
 
-\subsection{The history, inventory, and genealogy processes}
+\subsection{History process}
 
-\paragraph{History process}
-
-Consider the Markov process $(\X_t,\U_t)$.
-We define its \emph{history process}, $\Hist_t$, to be the restriction of the random function $t\mapsto(\X_t,\U_t)$ to the interval $[0,t]$.
-Note that $\Hist_t$ is itself a Markov process.
-%% Conditional on $\Hist_t$, both $\X_t$ and $\U_t$ are deterministic, as of course are the Markov chains of event times $\rdm{\hat{T}}_k$ and the embedded chains $\rdm{\hat{X}}_k=\X_{\rdm{\hat{T}}_k}$ and $\rdm{\hat{U}}_k=\U_{\rdm{\hat{T}}_k}$.
-Conditional on $\Hist_t$, both $\X_t$ and $\U_t$ are deterministic as are the embedded chain, $(\rdm{\hat{X}}_k,\rdm{\hat{U}}_k)$, and the point process of event times $\rdm{\hat{T}}_k$.
-(Recall that $\rdm{\hat{X}}_k=\rdm{X}_{\rdm{\hat{T}}_k}$ and $\rdm{\hat{U}}_k=\rdm{U}_{\rdm{\hat{T}}_k}$.)
-The non-explosiveness assumption implies that, for every $t$, $\rdm{K}_t\coloneq\big\vert\event{\Hist}{t}\cap[0,t]\big\vert<\infty$ a.s.
-The probability measure, $\pi^H$, for $\Hist_t$ can be expressed in terms of these:
-\begin{equation*}%\label{eq:Hdens}
+Consider the Markov process $(\Xr_t,\Ur_t)$.
+We define its \emph{history process}, $\Hr_t$, to be the restriction of the random function $s\mapsto(\Xr_s,\Ur_s)$ to the interval $[0,t]$.
+Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.
+Alternatively, one can think of $\Hr_t$ as consisting of the sequence
+$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
+In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
+The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
+\begin{equation*}\label{eq:Hdens}
   \begin{aligned}
-    \pi^{H}(\dd{H_t})
-    =&p_{0}(\hat{X}_{0})\,\dd{\hat{X}_0}\,
-    \prod_{k=1}^{K_{t}}{\left(\alpha_{\hat{U}_k}\left(\hat{T}_k,\hat{X}_{k-1},\hat{X}_{k}\right)\,\dd{\hat{X}_k}\,\dd{\hat{T}_k}\right)}\\
-    &\qquad\times\exp{\left(-\sum_{k=1}^{K_t}{\int_{\hat{T}_{k-1}}^{\hat{T}_k}{\sum_{u}{\int{\alpha_{u}(t',\hat{X}_{k-1},x')\,\dd{x'}}}}\,\dd{t'}}\right)}.
+    \pi^{\mathrm{H}}(\dd{\H_t})
+    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
+    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
+    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
   \end{aligned}
 \end{equation*}
+where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.
 
-\paragraph{Inventories}
+\subsection{Binomial ratio}\label{sec:binomial_ratio}
 
-Our goal in this paper is to probabilistically characterize how the genealogical relationships among lineages evolve through time.
-Accordingly, we develop some notation for this purpose.
-To begin with, we assign to each lineage a unique number $j\in\Zp$.
-This can be done in any fashion, so long as no two lineages ever receive the same number.
-For example, when a new lineage arises, we can assign it the smallest integer that has not yet been assigned.
-We will define the \emph{inventory process}, $\Inv_t$, so that, for every lineage $j\in\Zp$, $\Inv_t(j)$ is the deme in which $j$ is found.
-However, when $t$ is before the birth or after the death of lineage $j$, then clearly $\Inv_t(j)\notin\Demes$.
-We say in this case that lineage $j$ is in the \emph{underdeme}, which we denote using the symbol $\eth$, so that we can write $\Inv_t(j)=\eth$.
-We define $\Demesplus\colonequals\Demes\cup\{\eth\}$ so that $\Inv:\Rp\times\Zp\to\Demesplus$.
+For $n,r,\ell,s\in{\Zp^\Demes}$, define the \emph{binomial ratio}
+\begin{equation*}
+  \BinRatio{n}{\ell}{r}{s}\colonequals
+  \begin{cases}
+    \frac{\displaystyle\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}%
+         {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
+         & \text{if}\ \forall i\ n_i\ge{\Set{\ell_i,r_i}}\ge{s_i}\ge{0},\\
+         0, & \text{otherwise}.
+  \end{cases}
+\end{equation*}
+Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
+Moreover, in consequence of the Chu-Vandermonde identity, we have
+\begin{equation*}
+  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
+\end{equation*}
+whenever $n_i\ge{\Set{\ell_i,r_i}}\ge{0}$ for all $i$.
 
-The birth and death times of lineage $j$ are therefore
-\begin{equation*}
-  \begin{gathered}
-    t^b_j=\min\{t|\Inv_t(j)\ne{\eth}\}\qquad
-    \text{and}\qquad
-    t^d_j=\sup\{t|\Inv_t(j)\ne{\eth}\},
-  \end{gathered}
-\end{equation*}
-respectively.
-Observe that $n_i(\X_t)=\left\vert\left\{j\;\vert\;\Inv_t(j)=i\right\}\right\vert$ for all $t\in\Rp$ and $i\in\Demes$.
-Note also that $n$ does not count the inhabitants of the underdeme.
+\section{The induced genealogy process}
 
-\paragraph{Jump types}
+\subsection{Genealogies}
 
-Different kinds of events that occur for the population process can have different kinds of effects on the inventory process, and indeed not every jump affects $\Inv_t$ at all.
-From the point of view of the inventory process, there are five distinct \emph{pure types} of jumps, which we enumerate here.
-\begin{enumerate}[(a)]
-\item Birth-type events result in the branching of one or more new lineages, each from some existing lineage.
-  If $j$ is one of the new lineages, we use the expression $\Anc(j)$ to refer to its ancestor.
-  Examples of birth-type events include transmission events, speciations, and actual births.
-  It is not assumed that all new lineages arising from a birth event share the same ancestor.
-\item Death-type events result in the extinction of one or more lineages.
-  Examples include recovery, death of a host, and species extinctions.
-\item Migration-type events result in the movement of a lineage from one deme to another.
-  Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
-\item Sample-type events result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
-\item Neutral-type events result in no change to any of the lineages.
-\end{enumerate}
-\Cref{fig:markov_state} depicts an example with all five of the pure types.
-It is not necessary that a jump fall into just one of these types.
-It is allowable, for instance to have compound jumps that fall into more than one category.
-For example, sample/death-type events, in which a lineage is simultaneously sampled and removed, have been used, as have birth/death events in which one lineage reproduces at the same moment that another dies.
-The theory presented here places no restrictions on the complexity of the events that can occur.
-
-However, we do impose the restriction that the \emph{production}, \ie the deme-specific number of lineages emerging from the event, be constant for all jumps of a given mark.
-To be precise, the \emph{production} is defined to be a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
-We write $r^u=(r^u_i)_{i\in\Demes}$.
-Note that we lineages that die as a result of an event do not count in the production.
-Also, it is important to note that the parent lineage or lineages, if they survive the event, are always counted in the production.
-
-Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents and demes, there is implicitly a deterministic indicator function $Q_u$, for $u\in\Jumps$, (described below) that captures these properties.
-
-\paragraph{Inventory process}
-
-The structure of the state space for the inventory process, $\Inv_t$, has already been described.
-It remains to define its stochastic dynamics.
-The $\Inv_t$ process is driven by the population process $\X_t$ in that jumps in $\Inv_t$ do not occur except when jumps in $\X_t$ occur:
-$\leftlim{\Inv}_t\ne\Inv_t$ implies $\leftlim{\N}_t\ne\N_t$.
-
-At jumps of birth type and mark $u$, the appropriate number of random parents are selected from the appropriate deme(s) and each one sires the appropriate number of offspring in each deme.
-At jumps of death type, the appropriate number of lineages are selected from the appropriate demes and moved to the underdeme.
-At jumps of migration type, randomly selected lineages are moved between demes as appropriate.
-At sample-type and neutral-type jumps, no change occurs.
-Jumps of compound type are handled in the obvious way.
-
-Crucially, the assumption that the population process $\X_t$ is Markov empowers us to assume that the individual lineages within each deme are exchangeable with respect to the inventory process.
-Since the law governing $\X_t$ is independent of the identities of the individual lineages, and since the individual lineages are exchangeable by assumption, it follows that, conditional on $\Hist_t$, the jumps of the inventory are independent of one another.
-
-\paragraph{Genealogies}
-
-\citet{King2022}, showed how an unstructured population process induces a process on the space of genealogies.
-Although we now treat a more general case, the construction is much the same, so we abbreviate the presentation.
-Readers wishing for more detail should consult the earlier paper.
-
-For our purposes, a \emph{genealogy} is a labeled, time-calibrated tree.
-Its edges represent relationships of ancestry and descent among its nodes.
+For our purposes, a \emph{genealogy}, $G_{T}$, is a labeled, time-calibrated tree.
+It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
 There are three distinct kinds of nodes:
 \begin{inparaenum}[(i)]
-\item \emph{tip nodes}, which represent labeled extant lineages;
-\item \emph{branch nodes}, which represent ancestral events;
+\item \emph{tip nodes}, which represent labeled living lineages;
+\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
 \item \emph{sample nodes}, which represent labeled samples.
 \end{inparaenum}
-More formally, we can take a genealogy, $G$, to be a finite sequence of nodes, together with a time.
-Given $G$, the time is denoted $T(G)$; this is the time corresponding to the extant lineages.
-The number of nodes is $K(G)$.
-We order the nodes temporally, and denote the $k$-th node $\Node{p}_k(G)$ and we write $\Node{p}\in{G}$ if $\Node{p}$ is one of the nodes of $G$.
-Each $\Node{p}\in\Gen$ has a creation-time, $T(\Node{p})$, a parent, $\Anc(\Node{p})$, and a deme.
-Moreover, $\event{G}{t}$ is just the sequence of node times:
-$\event{G}{t}=\left(T(\Node{p}_k(G))\right)_{k=1}^{K(G)}$.
-%% The nodes are ordered so that $k<k'$ implies that $T(\Node{p}_k(\Gen))\le{T(\Node{p}_{k'}(\Gen))}$.
-Root nodes are distinguished by being their own parents:
-$\Node{p}$ is a root if and only if $\Anc(\Node{p})=\Node{p}$.
-Every node also has one or more descendant nodes called \emph{children}.
+Each node $a$ is associated with a specific time, $t(a)$.
+In particular, if $a$ is a tip node in $G_T$, then $t(a)=T$;
+if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
+Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
+We let $\event{G}{T}$ denote the set of all node-times of the genealogy $G_T$.
+Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
+Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.
 
-\paragraph{Genealogy processes}
+Formally, we define a genealogy, $G_T$, to be an ordered pair $(y,z)$.
+The $z$ element specifies the topological structure, as follows.
+Let $\leaves{G_T}$ be the set of all tip and sample nodes in $G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{G_T}$.
+The partition \emph{fineness} defines a partial order on $\func{P}$.
+Specifically, for $p,p'\in{\func{P}}$, we say $p\preceq{p'}$ if and only if for every $c'\in{p'}$ there is a unique $c\in{p}$ such that $c\supseteq{c'}$.
+The topological structure of $G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
 
-The population and inventory processes together induce a stochastic process, $\Gen_t$, on the space of genealogies.
-In particular, at each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event:
-\begin{enumerate}[(a)]
-\item A birth-type event at time $t$ results in the creation of one new branch node for each parent lineage.
-  In particular, if $j$ is one of the parent lineages, and $\Node{p}$ is the new node, then $T(\Node{p})=t$, $\Anc(\Node{p})$ is the node that was parent to $j$ prior to the event.
-  The children of $\Node{p}$ include one new tip node for each new lineage sired by $j$, as well as $j$ itself.
-\item In a death-type event, all the lineages $j$ that die are removed.
-  Branch nodes without children are then recursively removed.
+The $y$ element of $G_T$ specifies the coloring of branches and location of nodes.
+In particular, for $a\in\leaves{G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
+To make this well-defined, let $d_{t}(a)=\eth$ for $t\in{[t(a),T]}$;
+we call $\eth$ the \emph{underdeme} and define $\Demesplus\coloneq{\Demes\cup\{\eth\}}$.
+Thus $d:[0,T]\times\leaves{G_T}\to\Demesplus$ is a well-defined \cadlag\ function.
+
+Since not every node along a lineage corresponds to a change of deme, we also count the number of nodes along each lineage.
+Let $m:[0,T]\times\leaves{G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
+For $t\in{[t(a),T]}$, let $m_{t}(a)=m_{t(a)}(a)$.
+Clearly, $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
+
+Finally, we define the map $y:[0,T]\times\leaves{G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
+The two \cadlag\ functions $y$ and $z$ specify the genealogy $G_T$ completely and uniquely.
+When the genealogy is not clear from context, we write $y^{G}$ and $z^{G}$ to refer to the respective components of genealogy $G$.
+
+Clearly, the topological structure of a tree imposes constraints on the permissible functions $y$ that describe the painting and node placement along it.
+In particular, $y$ must respect the topological structure of $G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
+In other words, if $a,b\in{c}\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
+Let $\Yspace^z$ denote the space of all functions that respect the topological structure described by $z$.
+Let $\Yspace^\z_t$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.
+
+\subsection{Jump types}\label{sec:jump_types}
+
+We will show how a given population process naturally induces a process in the space of genealogies.
+Because different kinds of jumps have different effects on the genealogy, we need a way of distinguishing them.
+In particular, there are five distinct \emph{pure types} of jumps:
+\begin{compactenum}[(a)]
+\item \emph{Birth-type jumps} result in the branching of one or more new lineages, each from some existing lineages.
+  Examples of birth-type events include transmission events, speciations, and actual births.
+  For simplicity, we assume that all new lineages arising from a birth event share the same parent.
+\item \emph{Death-type jumps} result in the extinction of one or more lineages.
+  Examples include recovery from infection, death of a host, and species extinctions.
+  We allow for the possibility that multiple lineages die simultaneously.
+\item \emph{Migration-type jumps} result in the movement of a lineage from one deme to another.
+  Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
+  We permit multiple movements to occur simultaneously.
+\item \emph{Sample-type jumps} result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
+  We allow for the possibility that multiple samples are collected in a single sample event.
+  In this case, we assume that, in each sample-type event, each extant lineage is sampled at most once.
+\item \emph{Neutral-type jumps} result in no change to any of the lineages.
+\end{compactenum}
+\Cref{fig:markov_state} depicts an example with all five of the pure types.
+It is not necessary that a jump be of a pure type.
+Compound jumps partake of more than one type.
+For example, a sample/death-type event, in which a lineage is simultaneously sampled and removed, has been proposed \citep{Leventhal2014}, as have birth/death events in which one lineage reproduces at the same moment that another dies (\eg the Moran process \citep{Moran1958}).
+The theory presented here places few restrictions on the complexity of the events that can occur.
+
+%% Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents and demes, there is implicitly a deterministic indicator function $Q_u$, for $u\in\Jumps$, (described below) that captures these properties.
+
+\subsection{Genealogy process}
+
+We now show how a given population process induces a stochastic process, $\Gen_t$, on the space of genealogies.
+In the case of unstructured population processes (\ie those having a single deme), \citet{King2022} gave a construction.
+Although we now treat a more general case, the ideas are much the same.
+Readers wishing more detail on the construction here should consult the earlier paper.
+
+At each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event:
+\begin{compactenum}[(a)]
+\item
+  A birth-type event at time $t$ results in the creation of one new internal node, call it $b$.
+  A tip node, $a$, of the appropriate deme is chosen with uniform probability from among those present and $b$ is inserted so that its ancestor is that of $a$, while $a$ takes $b$ as its ancestor.
+  One new tip node, of the appropriate deme, is created for each of the children, all of which take $b$ as their immediate ancestor.
+\item
+  In a death-type event, one or more tip nodes of the appropriate demes are selected with uniform probability from among those present.
+  These are deleted.
+  Next, branch nodes without children are recursively removed.
   Sample nodes are never removed.
-\item In a migration-type event, one branch node is added for each migrating lineage;
-  each one takes the migrating lineage as child.
-  The ancestor of the new node is that which was parent to the migrating lineage prior to the event.
-  The deme of the lineage changes accordingly.
-\item At a sample-type event, one new sample node is introduced for each sampled lineage.
-  Each one takes the sampled lineage as child.
-  The ancestor of the sample node is that which was parent to the sampled lineage before the vent.
-\item At a neutral-type event, no change is made to the genealogy.
-\item Finally, events of compound type are accommodated by combining the foregoing rules.
-\end{enumerate}
-When an event results in the addition of one or more new nodes to a genealogy, the lineages which are children of that node are said to \emph{emerge} from the event.
-Thus, after a birth-type event, the emerging lineages include all the new offspring as well as their parents.
-Likewise, at pure migration- or sample-type events, each migrating or sampled lineage emerges from the event.
-At pure death-type events, no lineages emerge.
-In general, at an event of mark $u$, there are $r^u_i$ emergent lineages in deme $i$.
+\item
+  In a migration-type event, the appropriate number of migrating lineages are selected at random, with uniform probability according to deme, from among those present.
+  For each selected lineage, one new branch node is inserted between the selected tip node and its ancestor.
+  The color of the descendant branch changes accordingly.
+\item
+  At a sample-type event, the appropriate number of sampled lineages are selected at random from among the tip nodes, with uniform probability according to deme.
+  One new sample node is introduced for each selected lineage:
+  each is inserted between a selected tip nodes and its ancestor.
+\item
+  At a neutral-type event, no change is made to the genealogy.
+\item
+  Finally, events of compound type are accommodated by combining the foregoing rules.
+\end{compactenum}
+In each of these events, the new node or nodes that are introduced are situated at time $t$.
 
@@
 
-\subsection{The pruned and obscured genealogies}
+\subsubsection{Emergent lineages and production}
 
-Although the process just described yields a genealogy that relates all extant members of the population, and all samples, the data we ultimately wish to analyze will contain only the samples.
+The lineages which descend from an inserted node are said to \emph{emerge} from the event.
+Thus, after a birth-type event, the emerging lineages include all the new offspring as well as the parent.
+Likewise, at pure migration- or sample-type events, each migrating or sampled lineage emerges from the event.
+At pure death-type events, no lineages emerge.
+In general, at an event of mark $u$, there are $r^u_i$ emergent lineages in deme $i$.
+We require that $r^u_i$ be a constant, for each $u$ and $i$.
+Since one is free to expand the set of jump-marks $\Jumps$ as needed, this is not an important restriction on the models that the theory can accommodate.
+Thus there is a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
+We say $r^u=(r^u_i)_{i\in\Demes}$ is the \emph{production} of an event of mark $u$.
+Note that the lineages that die as a result of an event do not count in the production but that all parent lineages, if they survive the event, are counted in the production.
+
+\subsubsection{Conditional independence and exchangeability}
+
+Application of these rules at each jump of $\Xr_t$ constructs a chain of genealogies $\Grh_k$.
+In particular, at each jump-time $\Trh_k$, the genealogy $\Grh_{k-1}$ is modified according to the jump-mark $\Urh_k$ to yield $\Grh_k$.
+We view $\Grh_k$ as the embedded chain of the continuous-time genealogy process $\Gen_t$.
+It is very important to note that, conditional on $\Urh_k$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of the choices at any other times and independent of $\Xrh_k$ for all $k$.
+Moreover, by construction, any lineage within a deme is as likely as any other lineage in that deme to be selected as a parent or for death, sampling, or migration.
+We refer to this property as the \emph{exchangeability} of lineages within a deme.
+
+\subsection{Pruned and obscured genealogies}
+
+\subsubsection{Pruned genealogy}
+
+Although the process just described yields a genealogy that relates all extant members of the population, and all samples, the data we ultimately wish to analyze will be based only on samples.
 We therefore describe how the genealogy process is \emph{pruned} to yield the sample-only genealogy.
-Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Prune_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every branch node without children.
-In a pruned genealogy only branch and sample nodes remain, and sample nodes are found at all of the leaves and some of the interior nodes of the genealogy.
-Observe that the pruned genealogy retains information not only about how much ancestry is shared by any pair of sample lineages, but also about where among the demes each lineage was through time.
-Note also that $T(\Prune_t)=T(\Gen_t)=t$.
+Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Pr_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every childless internal node.
+In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and possibly some of the interior nodes of the genealogy.
+Observe that a pruned genealogy is a colored genealogy:
+it retains information about where among the demes each of its lineages was through time.
+Note also that a pruned genealogy $\Pr_t$ is characterized by functions $y^{\Pr_t}$ and $z^{\Pr_t}$ just as an un-pruned genealogy is.
 
-\paragraph{Alternative representation of a pruned genealogy}
+\subsubsection{Lineage count and saturation}
 
-Consider the finite set of all samples represented in pruned genealogy $P$.
-Beginning with 1, assign natural numbers to these in such a way that if sample $j$ is ancestral to sample $j'$, then $j<j'$.
-For example, we can order the samples by their times, resolving any ties arbitrarily.
-Let $\Lin(P)$ denote the set of lineage numbers.
-Using this ordering, we can uniquely associate each point on a genealogical tree with the least of those lineages that descend from that point.
-In particular, any lineage $j\in\Lin(P)$, corresponding to a sample taken at time $t^s_j$, can be traced backward from node to node until either it coalesces with some lesser (\ie older) lineage at some time $t^o_j>0$ or a root is reached (in which case, we define $t^o_j=0$).
-Each node encountered along the way represents a genealogical event from which $j$ emerges.
-Moreover, at each time $t\in\halfopen{t^o_j,t^s_j}$, lineage $j$ is in precisely one of the demes $\Demes$.
-However, for $t\notin\halfopen{t^o_j,t^s_j}$, lineage $j$ does not exist.
-To express this, we again say that lineage $j$ is in the underdeme, $\eth$.
-
-It will be useful to distill the elements that characterize a pruned genealogy.
-Accordingly, given a pruned genealogy, $P$, we make the following definitions.
-\begin{compactenum}[(a)]
-\item Let $\ct^{P}:\Zp\times\Rp\to\Zp$ be such that, for $j\in\Zp$, $\ct^{P}_j(t)$ is the counting process which increases by 1 at each event along lineage $j$.
-\item Let $\anc^{P}:\Zp\times\Rp\to\Zp$ be such that $\anc^{P}_j(t)$ indicates the unique lineage, ancestral of lineage $j$, and alive at time $t$.
-  In particular, we posit that $\anc^{P}_j(t)=j$ for $t\in\halfopen{t^o_j,t^s_j}$ and $\anc^{P}_j(t)=0$ for $t>t^s_j$, so that the function is well defined for all $j$ and $t$.
-  %%  Alternatively, one can define $\anc$ using the original construction of Kingman's coalescent \citep{Kingman1982a}.
-\item Let $\deme^{P}:\Zp\times\Rp\to\Demesplus$ be such that, for $j\in\Zp$, $\deme^{P}_j(t)$ indicates in which deme lineage $j$ lies at time $t$.
-  In particular $\deme^{P}_j(t)=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
-\item Let $\Yspace=(\Zp\times\Zp\times\Demesplus)^{\Zp}$ and define $Y^{P}:\Rp\to\Yspace$ such that, for $t\in\Rp$,
-  $Y^{P}(t)=\left(\ct^{P}(t),\anc^{P}(t),\deme^{P}(t)\right)$.
-\end{compactenum}
-For $j>|\Lin(P)|$, we adopt the convention that $t^o_j=t^s_j=\infty$, so that $\ct^P_j(t)=0$, $\anc^{P}_j(t)=0$, and $\deme^P_j(t)=\eth$ for all $t\in\Rp$.
-It is easy to see that $Y^{P}$ is well defined, piecewise constant, and \cadlag, and that the map $P\mapsto{Y^{P}}$ is one-to-one.
-However, not every piecewise constant, \cadlag\ map $y:\Rp\to\Yspace$ defines a pruned genealogy.
-If $y$ does define a pruned genealogy, we denote that pruned genealogy by $P(y)$.
-Given $y\in\Yspace$, we will use the notation $\ct(y)$, $\anc(y)$, and $\deme(y)$ to refer to the coordinates of $y$.
-
-Since $\Gen_t$ is a stochastic process, both $\Prune_t\colonequals\prune(G_t)$ and $Y^{\Prune_t}$ are stochastic processes as well.
-In fact, the latter two are Markovian, since each contains within itself all of its past history.
-The process $\Gen_t$ is not Markovian, though $(\X_t,\Gen_t)$ is.
-
-To visualize $\Prune_t$, one can make a correspondence between demes and colors.
-Then a pruned genealogy is visualized as a tree with colored branches.
-Knowing the function $\deme^{\Prune_t}$ is equivalent to knowing the coloring, while $\ct^{\Prune_t}$ determines the locations of events in the genealogy and $\anc^{\Prune_t}$ determines the topology.
-Note in particular that, $Y^{P_t}_j(t')\ne\leftlim{Y}^{P_t}_j(t')$ if and only if $t'$ is the time of an event from which lineage $j$ emerges.
-Note also that, there can be events on a pruned genealogy where the color does not change.
-That is, where $\ct_j$ increments, but neither $\anc_j$ nor $\deme_j$ change.
-
-\paragraph{Lineage count and saturation}
-
-In the following, we will find that we need to count the deme-specific numbers of lineages present at a given time.
-Accordingly, for any $\eta\in\Yspace$, and $i\in\Demes$, let us define
-\begin{equation*}
-  \ell_i(\eta)\colonequals{\left\vert\left\{j\in\Zp\;\vert\;\deme(\eta_j)=i\right\}\right\vert}\in\Zp, \qquad \ell(\eta)\colonequals(\ell_i(\eta))_{i\in\Demes}\in{\Zp^\Demes}.
-\end{equation*}
-Note that lineages $j$ for which $\deme(\eta_j)=\eth$ are not counted.
-With this definition, it follows that, for any pruned genealogy $P$, $\ell_i(Y^{P}(t))$ is the number of lineages in deme $i$ at time $t$ and
-$\ell(Y^{P}(t))\in\Zp^\Demes$ is the non-negative integer vector telling how many lineages lie within each of the demes at time $t$.
+In the following, we will find that we need to count the deme-specific numbers of lineages present in a given pruned genealogy at a given time.
+Accordingly, suppose $\Pr_T=(y,z)$ is a pruned genealogy and suppose $t\in[0,T]$.
+Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell=(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
+Clearly, $\ell$ depends only on the values, $(y_t,\z_t)$, of $y$ and $z$ at time $t$.
+Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(y,z)$ is a pruned genealogy, $\ell(y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
+We refer to $\ell$ as the \emph{lineage-count} function.
 
 We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
-Therefore, for $y,y'\in\Yspace$, and $i\in\Demes$, let us define
-\begin{equation*}
-  s_i(y,y')\colonequals{\left\vert\left\{j\in{\Zp}\;\vert\;\deme(y_j)=i\ \ \&\ \ \ct(y'_j)\ne\ct(y_j)\right\}\right\vert}, \qquad s(y,y')\colonequals(s_i(y,y'))_{i\in\Demes}\in{\Zp^\Demes}.
-\end{equation*}
-With this definition, for any $P$, if $y=Y^P$, then $s_i(\yt_t,y_t)$ is the number of lineages in deme $i$ that emerge from an event at time $t$ and $s(\yt_t,y_t)$ is the non-negative integer vector telling how many lineages in each deme emerge at time $t$.
-In particular, $s(y,y)=0$ so that if $e\notin\event{P}{t}$, then $s(\yt_e,y_e)=0$.
-We refer to $s(\yt_t,y_t)$ as the \emph{saturation} at time $t$.
+In particular, given a node time $t$ in a pruned genealogy $\Pr_T=(y,z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
+Like the lineage-count function, $s$ depends only on the local structure of $\Pr_T$.
+However, $s$ depends not only on $(\y_t,\z_t)$, but also on $(\yt_t,\zt_t)$.
+Thus, we can define the \emph{saturation} function such that, whenever $\Pr_T=(\y,\z)$ is a pruned genealogy, $s(\yt_t,\y_t,\zt_t,\z_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
 
-\paragraph{Obscured genealogy}
+\begin{figure}
+  \caption{
+    Figure showing lineage count and saturation.
+    \label{fig:ells}
+  }
+\end{figure}
 
-As we have just seen, a pruned genealogy contains information about the full history of each sample lineage, including the times at which it entered or exited any deme, sired offspring, or was sampled.
+\subsubsection{Obscured genealogy}
+
+As we have seen, a pruned genealogy contains information about the full history of each sample lineage, including the times at which it entered or exited any deme, sired offspring, or was sampled.
 The data we seek to analyze will typically lack much of this information.
 Accordingly, we define the \emph{obscured genealogy} to be that obtained by discarding all information about demes and events not visible from the topology of the tree alone.
-In particular, if $\Prune$ is a pruned genealogy and $\Vis=\obs(\Prune)$ is the corresponding obscured genealogy,
-$\Vis$ is uniquely determined by the function $t\mapsto\anc(Y^{\Prune}(t))$.
-
-\paragraph{Binomial ratio}
-
-For $n,r,\ell,s\in{\Zp^\Demes}$, define the \emph{binomial ratio}
-\begin{equation*}
-  \BinRatio{n}{\ell}{r}{s}\colonequals
-  \begin{cases}
-    \frac{\displaystyle\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}%
-         {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
-         & \forall i\ n_i\ge{\{\ell_i,r_i\}}\ge{s_i}\ge{0},\\
-         0, & \text{otherwise}.
-  \end{cases}
-\end{equation*}
-Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
-Moreover, in consequence of the Chu-Vandermonde identity, we have
-\begin{equation*}
-  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
-\end{equation*}
-whenever $n_i\ge{\{\ell_i,r_i\}}\ge{0}$ for all $i$.
+In particular, if $\Pr_T=(y,z)$ is a pruned genealogy, we write $\obs(\Pr_T)=z$ to denote the obscured genealogy.
 
@@
 
-\paragraph{Likelihood for pruned genealogies}
+\subsection{Likelihood for pruned genealogies}
 
-Our first result will be an expression for the likelihood of a given pruned genealogy $P^*_t$ given the history $\Hist_t$ of the population process to time $t$.
-Of course, not every pruned genealogy is compatible with $\Hist_t$.
-In particular, $P^*_t$ is only compatible with $\Hist_t$ if every event in $P^*_t$ coincides with an event in $\Hist_t$.
-That is, only if $\event{P^*}{t}\subseteq{\event{\Hist}{t}}$.
-Moreover, certain events in $P^*_t$ may be incompatible with all possible histories.
-For example, if $P^*_t$ has an event in which a lineage moves from deme $i$ to deme $i'$ but there are no $u\in\Jumps$ for which this is possible, then $P^*_t$ is incompatible with the population process itself.
-Let us define the function $Q:\Jumps\times\Yspace^2\to\{0,1\}$ so that $Q_u(y,y')=1$ if and only if a change $y\to{y'}$ in a pruned genealogy is compatible with an event of mark $u$ at that time.
-Recall that $n:\Xspace\to\Zp^{\Demes}$ denotes the deme-occupancy function, $r^u$, the production of a jump of mark $u$, $\ell$, the lineage-count function, and $s$, the saturation.
+Our first result will be an expression for the likelihood of a given pruned genealogy $\P^*_T$ given the history $\Hr_T$ of the population process to time $T$.
+Of course, not every history is compatible with $\P^*_T$.
+In particular, $\Hr_T$ is only compatible with $\P^*_T$ if every node-time in $\P^*_T$ is an event time of $\Hr_T$, \ie only if $\event{\Hr}{T}\supseteq\event{P}{T}^*$.
+Moreover, even if an event in $\Hr_T$ coincides with one of the node times of $\P^*_T$, it may be incompatible with the local structure of $\P^*_T$.
+For example, if no sample is present in $\P^*_T$ at time $t$ and $\Hr_T$ contains an event of sample type at that time, then incompatibility arises.
+As another example, if $\P^*_T$ has a node at time $t$ at which a lineage moves from deme $i$ to deme $i'$, but there is no $u\in\Jumps$ for which such a move is possible, then incompatibility arises again.
+To keep track of this, let us introduce the indicator function $Q$ such that $Q_u(y,y',z,z')=1$ if and only if there is some pruned genealogy $\P_T$ and $t\in{[0,T]}$,
+such that $\yt^{\P_T}_t=y$, $y^{\P_T}_t=y'$, $\zt^{\P_T}_t=z$, $z^{\P_T}_t=z'$,
+and $\P_T$ is compatible with an event of type $u$ at time $t$;
+$Q=0$ otherwise.
 
 \begin{thm}\label{thm:pruned_lik}
-  Let $P^*_t$ be a given pruned genealogy and $y^*=Y^{P^*_t}$.
+  Let $\P^*_T=(y^*,z^*)$  be a given pruned genealogy and
+  \begin{equation}\label{eq:phidef}
+    \phi_u(x,y,y',z,z')\coloneq\BinRatio{n(x)}{\ell(y',z')}{r^{u}}{s(y,y',z,z')}\,Q_{u}(y,y',z,z')
+  \end{equation}
+  (cf.~\cref{sec:binomial_ratio}).
   Then
   \begin{equation*}
-    \Prob{P^*_t\vert\Hist_t}=\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(y^*_e)}{r^{U_e}}{s(\yt^*_e,y^*_e)}\,Q_{U_e}(\yt^*_e,y^*_e)}
-    \:\times\:\Indicator{\event{P^*}{t}\subseteq{\event{\Hist}{t}}}.
+    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,y^*_t,\zt^*_t,z^*_t)}
+    \:\times\:\Indicator{\event{\H}{T}\supseteq{\event{\P}{T}^*}}.
   \end{equation*}
@@
 \begin{proof}
-  As we have already observed, if $\event{P^*}{t}\nsubseteq\event{\Hist}{t}$, then $\Prob{P^*_t\vert\Hist_t}=0$.
-  Similarly, if there is any event of $P^*_t$ which is incompatible with the population process, $\Prob{P^*_t\vert\Hist_t}=0$.
+  As we have already observed, if $\event{\Hr}{T}\nsupseteq\event{\P}{T}^*$, then $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
+  Similarly, if there is any event of $\Hr_T$ which is incompatible with $\P^*_T$, $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
   Let us therefore suppose that neither of these conditions hold.
-  Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
-  Moreover, at each event $e\in\event{\Hist}{t}$, a jump of mark $U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
-  In $P^*_t$, at time $e$, there are $\ell_i=\ell_i(y^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,y^*_t)$ are emergent.
-  The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
-  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, exactly $s_i$ of the drawn balls are black, namely
+  Conditional on $\Hr_T=\H_T$, at each time $t\in\event{\H}{T}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
+  In $\P^*_T$, at time $t$, there are $\ell_i=\ell_i(y^*_t,z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,y^*_t,\zt^*_t,z^*_t)$ are emergent.
+  By assumption, at each genealogical event, lineages within a deme are exchangeable:
+  each has an identical probability of being involved.
+  This exchangeability implies that each lineage present in a deme at time $t$ was equally likely to have been one of the emergent lineages.
+  In particular, at time $t$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ red balls and $n_i-r_i$ black balls, exactly $s_i$ of the drawn balls are red, namely
   \begin{equation*}
@@
   \end{equation*}
-  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $P^*_t$.
-  Moreover, since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
+  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $\P^*_t$.
+  Moreover, since, again conditional on $\Hr_T$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
   \begin{equation*}
-    \Prob{P^*_t\vert\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{\BinRatio{n(X_e)}{\ell(y^*_e)}{r^{U_e}}{s(\yt^*_e,y^*_e)}}.
+    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\BinRatio{n(\X_t)}{\ell(y^*_t,z^*_t)}{r^{\U_t}}{s(\yt^*_t,y^*_t,\zt^*_t,z^*_t)}}.
   \end{equation*}
-  Returning to the possibility that $P^*_t$ is incompatible with $\Hist_t$, since $\Prob{P^*_t}=0$ if either any $Q_u=0$ or any event of $P^*_t$ is not an event of $\Hist_t$, we obtain the result.
+  Returning to the possibility that $\H_T$ is incompatible with $\P^*_T$, since $\Prob{\Pr_T=\P^*_T}=0$ if either any $Q_u=0$ or any event of $\P^*_T$ is not an event of $\H_T$, we obtain the result.
 \end{proof}
@@
 Next, we show how the likelihood of a pruned genealogies, unconditional on the history, can be computed.
-For this, we use the filter equation technology developed in the Appendix.
+For this, we use the filter equation technology developed in \cref{sec:filter_eqns}.
 In particular, the following theorem follows immediately from \cref{prop:fullfilt}.
@@
 \begin{thm}
-  Suppose that $P^*_T$ is a pruned genealogy, for $T>0$.
-  Let $y^*=Y^{\Prune^*_T}$ and, for $x\in\Xspace$, $u\in\Jumps$, $t\le{T}$, define
-  \begin{equation*}
-    \phi_u\big(t,x)=\BinRatio{n(x)}{\ell(y^*_t)}{r^u}{s(\yt^*_t,y^*_t)}\,Q_u(\yt^*_t,y^*_t),
-  \end{equation*}
-  where $Q_u$ is as in \cref{thm:pruned_lik}.
-  Suppose $w=w(t,x)$ satisfies the filter equation
+  Suppose that $\P^*_T=(y^*,z^*)$ is a pruned genealogy, for $T>0$.
+  If $w=w(t,x)$ satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
   \begin{equation}
     \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}=
-      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(t,x)\,\dd{x'}}}
-      -\sum_u{\int{w(t,x)\,\alpha_u(t,x',x)\,\dd{x'}}}\\
-      &\ +\sum_{e\in\event{P^*}{T}}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(t,x)\,\dd{x'}}}-w(t,x)\right\}},
+      \frac{\partial{w}}{\partial{t}}(t,x)=
+      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}
+      -\sum_u{\int{w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}}}\\
+      &\quad+\sum_{e\in\event{\P}{T}^*}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}-w(t,x)\right\}},
     \end{aligned}
   \end{equation}
-  with the initial condition $w(0,x)=p_0(x)$.
-  Then the likelihood of $P^*_T$ is
+  where $\phi$ is defined in \cref{eq:phidef},
+  then the likelihood of $\P^*_T$ is
   \begin{equation*}
-    \lik(P^*_T)=\int{w(T,x)\,\dd{x}}.
+    \lik(\P^*_T)=\int{w(T,x)\,\dd{x}}.
   \end{equation*}
@@
 
-\paragraph{Likelihood for obscured genealogies}
+\subsection{Likelihood for obscured genealogies}
 
@@
 \begin{thm}\label{thm:obsc_lik}
-  Let $V^*_t$ be a given obscured genealogy.
-  Let $\pi:\Jumps\times\Rp\times\Xspace^2\times\Yspace^2\to\Rp$ be a probability kernel, \ie $\pi_u(t,x,x',y,y')\ge{0}$ and $\sum_{y'}{\pi_u(t,x,x',y,y')}=1$.
-  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ if and only if $Q_u(y,y')=1$.
-  Let $\rdm{y}_t$ be the Markov jump process on $\Yspace$ generated by the kernel $\pi$, with $\event{\rdm{y}}{t}=\event{\Hist}{t}$.
-  Then there is a measurable function $\kappa_{V^*_t}:[0,t]\times\Yspace^2\to\{0,1\}$ such that
+  Let $z^*=\obs(\P^*_T)$ be a given obscured genealogy.
+  Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{z^*}}{q(x,y)}=1$.
+  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,z^*_t)=1$
+  and that $q(x,y)>0$ whenever $\Prob{\Xr_0=x,y^{\Pr_T}_0=y}>0$.
+  Then there is a Markov jump process $\yr_t$ on $\Yspace_{z^*}$ such that
   \begin{equation*}
-    \begin{aligned}
-      \Prob{V^*_t\vert\Hist_t=H_t}=&\Expect{\left.\prod_{e\in\event{H}{t}}{\BinRatio{n(X_e)}{\ell(\rdm{y}_e)}{r^{U_e}}{s(\leftlim{\rdm{y}}_e,\rdm{y}_e)}\,\frac{\scriptsize\kappa_{V^*_t}(e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}{\scriptsize\pi_{U_e}(e,\leftlim{X}_e,X_e,\leftlim{\rdm{y}}_e,\rdm{y}_e)}}\;\right\vert\;\Hist_t}\\
-      &\qquad\times\Indicator{\event{V^*}{t}\subseteq\event{H}{t}},
-    \end{aligned}
+    \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
+    \Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}}
+    \times\Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*},
   \end{equation*}
-  the expectation being taken over $\rdm{y}$.
+  where $\phi$ is defined in \cref{eq:phidef} and
+  the expectation is taken over the sample paths of $\yr_t$, $t\in[0,T]$.
 \end{thm}
@@
   \begin{equation}\label{eq:IS1}
-    \Prob{V^*_t\vert\Hist_t}=\int{\Indicator{\obs(P_t)=V^*_t}\,\Prob{P_t\vert\Hist_t}\,\dd{P_t}},
+    \CondProb{\zr_T=z^*}{\Hr_T}=\CondExpect{\Indicator{\obs(\Pr_T)=z^*}}{\Hr_T}.
   \end{equation}
-  the integral being taken over all pruned genealogies.
   Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
-  we will propose pruned genealogies compatible with $V^*_t$ as sample paths from a Markov process on $\Yspace$ and
-  treat the integral in \cref{eq:IS1} as an expectation over these paths.
-  Conditional on $\Hist_t$, the probability kernel $\pi$ generates a Markov chain, $\rdm{\hat{y}}_k$, on $\Yspace$ such that
+  we will propose pruned genealogies compatible with $z^*_t$ as sample paths from a Markov process on $\Yspace_{z^*}$ and
+  evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
+  Conditional on $\Hr_T$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$, on $\Yspace_{z^*}$ such that
   \begin{equation*}
-    \Prob{\rdm{\hat{y}}_k=\hat{y}_k\vert\rdm{\hat{y}}_{k-1}=\hat{y}_{k-1}}=\pi_{U_{t_k}}(t_k,\leftlim{X}_{t_k},X_{t_k},\hat{y}_{k-1},\hat{y}_k).
+    \begin{gathered}
+      \Prob{\yrh_0}=q(\yrh_0),
+      \qquad
+      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_T=\H_T}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
+    \end{gathered}
   \end{equation*}
-  Now let $\event{H}{t}=\{t_1,t_2,\dots,t_K\}$.
-  $\rdm{\hat{y}}_k$ is the embedded chain of a unique, piecewise-constant, \cadlag\ process $\rdm{y}_t$, with jump times at $t_k$.
-  This construction of $\rdm{y}_t$ obviously guarantees that $\event{P(\rdm{y})}{t}\subseteq\event{H}{t}$.
-  Moreover, the conditions on $\pi$ guarantee that, for every pruned genealogy $P_t$, $Y^{P_t}$ is a sample path of this process.
-  In particular, they guarantee that $\pi_u(t,x,x',y,y')\;\propto\;{Q_u(y,y')}$.
+  The required Markov process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
+  This construction of $\yr_t$ obviously guarantees that $\event{\H}{t}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$.
 
+  Let us define $\Pr(\yr)=(\yr,z^*)$.
+  Then, by construction, for every $\yr$, $\obs(\Pr(\yr))=z^*$ and, conversely, for every pruned genealogy $\Pr_T$ satisfying $\obs(\Pr_T)=z^*$, $y^{\Pr_T}$ is a sample path of this process having nonzero probability.
   We therefore have that
   \begin{equation*}
-    \Prob{V^*_t\vert\Hist_t}=\Expect{\frac{\Prob{P(\rdm{y})\vert\Hist_t}}{\pi(\rdm{y}\vert\Hist_t)}\cdot\Indicator{\obs(P(\rdm{y}))=V^*_t}},
+    \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
+    \Expect{\frac{\CondProb{\Pr_T=P(\yr)}{\Hr_T=\H_T}}{\pi(\yr\vert\H_T)}}
   \end{equation*}
-  the expectation being taken with respect to the random process $\rdm{y}_t$.
+  the expectation being taken with respect to the random process $\yr_t$.
   Here, by definition,
   \begin{equation*}
-    \pi(\rdm{y}\vert\Hist_t)=\prod_{e\in\event{\Hist}{t}}{\pi_{U_{e}}(e,\leftlim{X}_{e},X_{e},\leftlim{\rdm{y}}_{e},\rdm{y}_{e})}.
+    \pi(\yr\vert\H_T)=\prod_{t\in\event{\H}{T}}{\pi_{\U_{t}}(t,\Xt_{t},\X_{t},\yrt_{t},\yr_{t})}.
   \end{equation*}
-
-  Finally, observe that the condition $\obs(P(\rdm{y}))=V$ can be factorized for all obscured genealogies $V$.
-  In particular, there is a function $\kappa_{V}:[0,t]\times\Yspace^2\to\{0,1\}$ such that $\kappa_{V}(s,\eta,\eta')=1$ if and only if
-  there is $y:\Rp\to\Yspace$ such that $\obs(P(y))=V$, $\leftlim{y}_s=\eta$, and $y_s=\eta'$, for all $s\in[0,T(V)]$.
-  Thus, conditional on $\Hist_t$, if $\event{P(y)}{t}\subseteq\event{\Hist}{t}$,
-  \begin{equation*}
-    \Indicator{\obs(P(y))=V^*_t}=\prod_{e\in\event{\Hist}{t}}{\kappa_{V^*_t}(e,\leftlim{y}_e,y_e)}.
-  \end{equation*}
-
   The result then follows from \cref{thm:pruned_lik}.
@@
 \begin{thm}
-  Let $V^*_T$, $T>0$ be a given obscured genealogy.
-  Let $\pi$, $\kappa_{V^*_T}$, and $\rdm{y}_t$ be as in \cref{thm:obsc_lik}.
-  For $x,x'\in\Xspace$, $y,y'\in\Yspace$, $u\in\Jumps$, define
+  Let $z^*$, be a given obscured genealogy.
+  Let $\pi$ and $\yr_t$ be as in \cref{thm:obsc_lik}.
+  For $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $u\in\Jumps$, define
   \begin{equation*}
     \begin{gathered}
-      \phi_u\big(t,x,y,y')=\BinRatio{n(x)}{\ell(y')}{r^u}{s(y,y')}\,\frac{\kappa_{V^*_T}(t,y,y')}{\pi_u(t,x,x',y,y')},\\
-      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y').
+      \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y',\zt^*_t,z^*_t)}{\pi_u(t,x,x',y,y')},\\[2ex]
+      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),
     \end{gathered}
   \end{equation*}
-  Suppose $w=w(t,x)$ satisfies the filter equation
+  where $\phi$ is defined in \cref{eq:phidef}.
+  Suppose $w=w(t,x,y)$ satisfies the filter equation
   \begin{equation}
@@
       \frac{\partial{w}}{\partial{t}}(t,x,y)=
-      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\phi_u(t,x,y',y)\,\dd{x'}}}\\
-      &\ -\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
-      &\ +\sum_{e\in\event{V^*}{T}}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u(t,x,y',y)\,\dd{x'}}}-w(t,x)\right\}},
+      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}}\\
+      &-\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
+      &+\sum_{e\in\event{\z}{}^*}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y,y')\,\psi_u(t,x',x,y',y)\,\dd{x'}}}-w(t,x,y)\right\}},
     \end{aligned}
   \end{equation}
-  with the initial condition $w(0,x,y)=p_0(x)$.
+  with the initial condition $w(0,x,y)=p_0(x)\,q(x,y)$.
   Then the likelihood of $V^*_T$ is
@@
 
-Jumps: $\Jumps=\{\mathrm{Inf},\mathrm{Prog},\mathrm{Recov},\mathrm{Wane},\mathrm{Birth},\mathrm{Death_S},\mathrm{Death_E},\mathrm{Death_I},\mathrm{Death_R},\mathrm{Sample}\}$.
+Jumps: $\Jumps=\Set{\lab{Inf},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Birth},\lab{Death_S},\lab{Death_E},\lab{Death_I},\lab{Death_R},\lab{Sample}}$.
 
-Demes: $\Demes=\{\mathrm{E},\mathrm{I}\}$.
+Demes: $\Demes=\Set{\lab{E},\lab{I}}$.
 
@@
 \begin{itemize}
-\item $\alpha_{\mathrm{Inf}}(t,x,x')=\beta(t)\,\frac{x^{\mathrm{S}}x^{\mathrm{I}}}{N(t)}\,\Indicator{x'=x+(-1,1,0,0)}$
-\item $\alpha_{\mathrm{Prog}}(x,x')=\rho\,x^{\mathrm{E}}\,\Indicator{x'=x+(0,-1,1,0)}$
-\item $\alpha_{\mathrm{Recov}}(x,x')=\gamma\,x^{\mathrm{I}}\,\Indicator{x'=x+(0,0,-1,1)}$
-\item $\alpha_{\mathrm{Wane}}(x,x')=\upsilon\,x^{\mathrm{R}}\,\Indicator{x'=x+(1,0,0,-1)}$
-\item $\alpha_{\mathrm{Sample}}(t,x,x')=\psi\,x^I\,\Indicator{x'=x}$
-\item $\alpha_{\mathrm{Birth}}(t,x,x')=B(t)\,\Indicator{x'=x+(1,0,0,0)}$
-\item $\alpha_{\mathrm{Death}_k}(x,x')=\mu\,x^k\,\Indicator{x'^j=x^j-\delta_{jk}}$, $k\in\{\mathrm{S},\mathrm{E},\mathrm{I},\mathrm{R}\}$
+\item $\alpha_{\lab{Inf}}(t,x,x')=\beta(t)\,\frac{x^{\lab{S}}x^{\lab{I}}}{N(t)}\,\Indicator{x'=x+(-1,1,0,0)}$
+\item $\alpha_{\lab{Prog}}(x,x')=\rho\,x^{\lab{E}}\,\Indicator{x'=x+(0,-1,1,0)}$
+\item $\alpha_{\lab{Recov}}(x,x')=\gamma\,x^{\lab{I}}\,\Indicator{x'=x+(0,0,-1,1)}$
+\item $\alpha_{\lab{Wane}}(x,x')=\upsilon\,x^{\lab{R}}\,\Indicator{x'=x+(1,0,0,-1)}$
+\item $\alpha_{\lab{Sample}}(t,x,x')=\psi\,x^I\,\Indicator{x'=x}$
+\item $\alpha_{\lab{Birth}}(t,x,x')=B(t)\,\Indicator{x'=x+(1,0,0,0)}$
+\item $\alpha_{\lab{Death}_k}(x,x')=\mu\,x^k\,\Indicator{x'^j=x^j-\delta_{jk}}$, $k\in\Set{\lab{S},\lab{E},\lab{I},\lab{R}}$
 \end{itemize}
@@
 %%     \frac{\partial{w}}{\partial{t}}(t,x,y,z)=&B(t)\,\left[\sum_{y'z'}w\big(t,x-(1,0,0,0),y',z'\big)-w\big(t,x,y,z\big)\right]\\
-%%     &\qquad+\sum_{k\in\{\mathrm{S},\mathrm{E},\mathrm{I},\mathrm{R}\}}\mu\,(x^k+1)\,w(t,x+\delta_{k})-\mu\,x^k\,w(t,x)\\
-%%     &\qquad+\upsilon\,x^{\mathrm{R}}\,w(t,x-(1,0,0,-1))\\
+%%     &\qquad+\sum_{k\in\{\lab{S},\lab{E},\lab{I},\lab{R}\}}\mu\,(x^k+1)\,w(t,x+\delta_{k})-\mu\,x^k\,w(t,x)\\
+%%     &\qquad+\upsilon\,x^{\lab{R}}\,w(t,x-(1,0,0,-1))\\
 %%   \end{aligned}
@@
 \setcounter{equation}{0}
+\titleformat{\section}[hang]{\large\bfseries}{Appendix \periodafter\thesection}{2ex}{\periodafter}{}
 \renewcommand{\theequation}{\thesection\arabic{equation}}
@@
 \section{Filter equations}
+\label{sec:filter_eqns}
 
@@
 \begin{defn}
-  Suppose $\X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
+  Suppose $\Xr_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
   \begin{equation}\label{eq:kfe2}
-    \frac{\partial{w}}{\partial{t}}
-    =\int{w(t,x')\,\beta(t,x',x)\,\dd{x'}}
-    -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}.
+    \frac{\partial{u}}{\partial{t}}
+    =\int{u(t,x')\,\beta(t,x',x)\,\dd{x'}}
+    -\int{u(t,x)\,\beta(t,x,x')\,\dd{x'}}.
   \end{equation}
-  Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)\in\R$ are given measurable functions.
+  Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)$ are given, real-valued, measurable functions.
   Let $\mathcal{M}\subseteq\Rp$ be countable such that $\mathcal{M}\cap[0,t]$ is finite for all $t$.
@@
   \end{equation}
-  is the \emph{filter equation} with \emph{driver} $\X_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$.
+  is the \emph{filter equation} with \emph{driver} $\Xr_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$.
 \end{defn}
@@
   \end{equation*}
+  Here $\wt$ refers to the left-limit of $w$.
+\end{remark}
+
+\begin{remark}
+  Trivially, a Kolmogorov forward equation is itself a filter equation with boost $1$, decay $0$, and no observed events ($\mathcal{M}=\emptyset$).
 \end{remark}
@@
 \begin{proof}
+  \ELI{[Conditions for differentiating under the integral in the proof?
+      Need $[v^2 \lambda u(t,x,v)]_{v=0}^{\infty} = 0$.
+      That's okay at $v=0$, and a minor constraint on $u$.]}
   \begin{equation*}
@@
 
-\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(\X_t,\rdm{V}_t)$.
-In particular, $\X_t$ is the driver with KFE \eqref{eq:kfe2}.
-The $\rdm{V}_t$ is \emph{directed} by $\X_t$ in the sense that $\rdm{V}$ has jumps wherever $\X$ does:
-when $\X$ jumps at time $t$ from $x$ to $x'$, $\rdm{V}$ jumps by the multiplicative factor $B(t,x,x')>0$.
-Between jumps, $\rdm{V}_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
-If we view $V_t$ as a weight, then \cref{lemma:filtereq} tells us how the $\rdm{V}_t$-weighted average of $\rdm{X}_t$ evolves in time:
-this average is simply $\int{w(t,x)\,\dd{x}}$.
+\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(\Xr_t,\Vr_t)$.
+In particular, $\Xr_t$ is the driver with KFE \eqref{eq:kfe2}.
+The $\Vr_t$ is \emph{directed} by $\Xr_t$ in the sense that $\Vr$ has jumps wherever $\Xr$ does:
+when $\Xr$ jumps at time $t$ from $x$ to $x'$, $\Vr$ jumps by the multiplicative factor $B(t,x,x')>0$.
+Between jumps, $\Vr_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
+If we view $\Vr_t$ as a weight, then \cref{lemma:filtereq} tells us how the $\Vr_t$-weighted average of $\Xr_t$ evolves in time:
+this average is simply $\int\!{w(t,x)\,\dd{x}}$.
 This motivates the following result, which shows how filter equations allow one to integrate over random histories.
@@
 \begin{prop}\label{prop:zeroboost}
-  Suppose $\X_t$ is a non-explosive pure jump process with KFE
+  Suppose $\Xr_t$ is a non-explosive pure jump process with KFE
   \begin{equation*}
     \begin{gathered}
-      \frac{\partial{w}}{\partial{t}}(t,x)=\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
-      w(0,x)=p_0(x).
+      \frac{\partial{u}}{\partial{t}}(t,x)=\int{u(t,x')\,\alpha(t,x',x)\,\dd{x'}}
+      -\int{u(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
+      u(0,x)=p_0(x).
     \end{gathered}
   \end{equation*}
-  Let $\Hist_t$ be its history process.
-  Suppose $\rdm{V}_t$ is a real-valued random process such that
+  Let $\Hr_t$ be its history process.
+  Suppose $\Vr_t$ is an $\Rp$-valued random process such that
   \begin{equation}\label{eq:Vprod}
-    \Expect{\rdm{V}_t\;\vert\;\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{B(e,\leftlim{X}_e,X_e)},
+    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)},
   \end{equation}
@@
   \end{equation*}
-  then $\Expect{\rdm{V}_t}=\int{w(t,x)\,\dd{x}}$.
+  then $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
 \end{prop}
@@
 \begin{prop}\label{prop:fullfilt}
-  Let $\X_t$, $\Hist_t$, $B$, and $\rdm{V}_t$ be as in \cref{prop:zeroboost}.
+  Let $\Xr_t$, $\Hr_t$, $B$, and $\Vr_t$ be as in \cref{prop:zeroboost}.
   Suppose that $\mathcal{M}\subseteq\halfclosed{0,t}$ is a fixed, finite set of observed event times.
-  Let $\Expect{\rdm{V}_t\;\vert\;\mathcal{M}}$ denote the density of $\rdm{V}_t$ conditional on $\mathcal{M}\subseteq\event{\Hist}{t}$.
+  Let $\CondExpect{\Vr_t}{\event{\Hr}{t}\supseteq\mathcal{M}}$ denote the expectation of $\Vr_t$ conditional on $\event{\Hr}{t}\supseteq\mathcal{M}$.
   If $w(t,x)$ satisfies $w(0,x)=p_0(x)$ and the filter equation
@@
   \end{equation*}
-  then $\Expect{\rdm{V}_t\;\vert\;\mathcal{M}}=\int{w(t,x)\,\dd{x}}$.
+  then $\CondExpect{\Vr_t}{\event{\Hr}{t}\supseteq\mathcal{M}}=\int{w(t,x)\,\dd{x}}$.
+  \ELI{[(a) line 2: density $\rightarrow$ expectation.
+      (b) the term $\delta(t,e)\,w(t,x)$ seems irrelevant:
+      the derivative at a finite set of points only matters when it is a delta function.]}
 \end{prop}
@@
   \begin{equation*}
-    w(s,x)=\Expect{\prod_{e\in\event{\Hist}{s}}{B\left(e,\Xt_e,\X_e\right)}}.
+    w(s,x)=\Expect{\prod_{e\in\event{\Hr}{s}}{B\left(e,\Xrt_e,\Xr_e\right)}}.
   \end{equation*}
@@
   By \cref{rem:equivfilt}, the result follows.
+  \ELI{[The argument shows that the filter equation satisfies a KFE, but the limit expression seems justified by a similar KFE.
+      Is the point that one KFE implies the other?]}
 \end{proof}
 
+\ELI{[Maybe, Prop A3 could be a version of Prop A2 allowing for discrete transitions at a finite set of times;
+    a combination of cts time and discrete time Markov chains.
+    That is all we need to extend Prop A2 to observation times, which can be viewed just as an extended Markov process including the observation events.
+    That is, to make a distinction between ``observation events'' that give the sequences used to construct the phylogeny, and ``observed events'' which are determined points on the phylogeny.
+    The latter are a realization of a random process, and the former are defined to be fixed a priori.]}
+
 %% Filter equations allow us to pass easily between equivalent representations of a process.
-%% For example, an equivalent way of representing $\X_t$ is in terms of its embedded chain and event times.
-%% Let $\rdm{\hat{X}}_k$ be the embedded chain of $\X_t$ and let $\rdm{\hat{T}}_k$ be the point process of its event times.
+%% For example, an equivalent way of representing $\Xr_t$ is in terms of its embedded chain and event times.
+%% Let $\Xrh_k$ be the embedded chain of $\Xr_t$ and let $\Trh_k$ be the point process of its event times.
 %% It is elementary that
@@
 %%   \begin{gathered}
-%%     \Prob{\rdm{\hat{X}}_k=\hat{X}_k\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_k=\hat{T}_k}=\frac{\alpha(\hat{T}_k,\hat{X}_{k-1},\hat{X}_k)}{\int{\alpha(\hat{T}_k,\hat{X}_{k-1},x')\dd{x'}}},\\
-%%     \Prob{\rdm{\hat{T}}_k>\rdm{\hat{T}}_{k-1}+t\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_{k-1}=\hat{T}_{k-1}}=\exp{\left(-\int_{0}^{t}{\int{\alpha(\hat{T}_{k-1}+s,\hat{X}_{k-1},x')\,\dd{x'}}\,\dd{s}}\right)}.
+%%     \Prob{\Xrh_k=\Xh_k\;\Big\vert\;\Xrh_{k-1}=\Xh_{k-1},\Trh_k=\Th_k}=\frac{\alpha(\Th_k,\Xh_{k-1},\Xh_k)}{\int{\alpha(\Th_k,\Xh_{k-1},x')\dd{x'}}},\\
+%%     \Prob{\Trh_k>\Trh_{k-1}+t\;\Big\vert\;\Xrh_{k-1}=\Xh_{k-1},\Trh_{k-1}=\Th_{k-1}}=\exp{\left(-\int_{0}^{t}{\int{\alpha(\Th_{k-1}+s,\Xh_{k-1},x')\,\dd{x'}}\,\dd{s}}\right)}.
 %%   \end{gathered}
@@
 %% Here, $\nu$ is the intensity of a time-homogenous Poisson process.
-%% Note that $\pi$ is the probability kernel of the embedded chain $\rdm{\hat{X}}$ and $A(t,x)$ is the intensity of the $\rdm{\hat{T}}_k$ process.
+%% Note that $\pi$ is the probability kernel of the embedded chain $\Xrh$ and $A(t,x)$ is the intensity of the $\Trh_k$ process.
 %% We recognize this equation as the filter equation with boost $B$, decay $\lambda$, and driver generated by $\nu\,\pi(t,x,x')$.
-%% It corresponds to the following procedure for simulating $\X_t$:
+%% It corresponds to the following procedure for simulating $\Xr_t$:
 %% \begin{compactenum}[(a)]
 %% \item Simulate jump times according to the rate-$\nu$ Poisson process.
-%% \item Simulate the embedded chain $\rdm{\hat{X}}_k$ using the kernel $\pi$.
+%% \item Simulate the embedded chain $\Xrh_k$ using the kernel $\pi$.
 %% \item Weight the realization by the product of the $B$ factors.

### 2024-01-20 ms.Rnw
edits

@@
       \begin{tikzpicture}[scale=1]
-        \tikzstyle{box}=[color=black, draw, fill=white, very thick, minimum size=3em]
+        \colorlet{darkblue}{blue!50!black}
+        \tikzstyle{box}=[draw=black, text=black, fill=white, very thick, minimum size=3em]
+        \tikzstyle{ibox}=[draw=darkblue]
+        \tikzstyle{deme}=[fill=black!20!white]
         \tikzstyle{label}=[font=\Large]
         \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{flow}=[color=black, very thick, >=stealth]
-        \tikzstyle{deme}=[fill=black!20!white]
-        \tikzstyle{modulate}=[color=blue, thick, >=latex]
+        \tikzstyle{flow}=[draw=black, very thick, >=stealth]
+        \tikzstyle{modulate}=[draw=darkblue, >=Circle]
+
+        %% SEIRS model
         \coordinate (origin) at (0,0);
-        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{A}};
-        \input{figs/seir_diagram}
+        \node[label] (lab) at (origin) {\textbf{A}};
+        \node [box] (S) at ($(origin)+(1,-1)$) {$\lab{S}$};
+        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
+        \node [box,ibox,deme] (I) at ($(E)+(2,0)$) {$\lab{I}$};
+        \node [box] (R) at ($(I)+(2,0)$) {$\lab{R}$};
+        \coordinate (overR) at ($(R)+(0,1)$);
+        \coordinate (midSE) at ($(S)!0.5!(E)$);
+        \draw [flow,->] (S) -- (E);
+        \draw [flow,->] (E) -- (I);
+        \draw [flow,->] (I) -- (R);
+        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
+        \draw [modulate,->] (I.north west) .. controls ($(I)+(-1,1)$) and ($(midSE)+(0,1)$) .. (midSE);
+        \node[font=\normalsize] (demes) at ($(E)!0.5!(I)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I}\}$};
+
+        %% SIIR (two-strain) model
+        \coordinate (origin) at (0,-3);
+        \node[label] (lab) at (origin) {\textbf{B}};
+        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
+        \node [box] (R1) at ($(S)+(6,1)$) {$\lab{R}_1$};
+        \node [box] (R2) at ($(S)+(6,-1)$) {$\lab{R}_2$};
+        \coordinate (overR) at ($(R1)+(0,1)$);
+        \coordinate (underR) at ($(R2)-(0,1)$);
+        \node [box,deme] (E1) at ($(S)+(2,1)$) {$\lab{E}_1$};
+        \node [box,deme] (E2) at ($(S)+(2,-1)$) {$\lab{E}_2$};
+        \node [box,ibox,deme] (I1) at ($(E1)+(2,0)$) {$\lab{I}_1$};
+        \node [box,ibox,deme] (I2) at ($(E2)+(2,0)$) {$\lab{I}_2$};
+        \draw [flow,->] (S) -- (E1);
+        \draw [flow,->] (E1) -- (I1);
+        \draw [flow,->] (I1) -- (R1);
+        \draw [flow,->] (S) -- (E2);
+        \draw [flow,->] (E2) -- (I2);
+        \draw [flow,->] (I2) -- (R2);
+        \draw [flow,->] (R1) -- (overR) -- (S |- overR) -- (S);
+        \draw [flow,->] (R2) -- (underR) -- (S |- underR) -- (S);
+        \coordinate (midSE1) at ($(S)!0.5!(E1)$);
+        \coordinate (midSE2) at ($(S)!0.5!(E2)$);
+        \draw [modulate,->] (I1.north west) .. controls ($(I1)+(-1,1)$) and ($(midSE1)+(0,1.8)$) .. (midSE1);
+        \draw [modulate,->] (I2.south west) .. controls ($(I2)+(-1,-1)$) and ($(midSE2)+(0,-1.8)$) .. (midSE2);
+        \node[font=\normalsize] (demes) at ($(E2)!0.5!(I2)+(0,-2)$) {$\Demes=\{\lab{E}_1,\lab{I}_1,\lab{E}_2,\lab{I}_2\}$};
+
+        %% COVID model
         \coordinate (origin) at (9,0);
-        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{B}};
-        \input{figs/siir_diagram}
-        \coordinate (origin) at (0,-4);
-        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{C}};
-        \input{figs/covid_diagram}
+        \node[label] (lab) at (origin) {\textbf{C}};
+        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
+        \node [box] (R) at ($(S)+(7,1)$) {$\lab{R}$};
+        \coordinate (overR) at ($(R)+(0,1)$);
+        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
+        \node [box,ibox,deme] (Ia) at ($(E)+(2,1)$) {$\lab{I_A}$};
+        \node [box,ibox,deme] (Is) at ($(E)+(2,-1)$) {$\lab{I_S}$};
+        \node [box] (H) at ($(Is)+(2,0)$) {$\lab{H}$};
+        \node [box] (D) at ($(H)+(2,0)$) {$\lab{D}$};
+        \draw [flow,->] (S) -- (E);
+        \draw [flow,->] (E) -- (Ia.west);
+        \draw [flow,->] (E) -- (Is.west);
+        \draw [flow,->] (Ia) -- (R);
+        \draw [flow,->] (Is) -- (H);
+        \draw [flow,->] (H) -- (R);
+        \draw [flow,->] (Is) -- (R);
+        \draw [flow,->] (H) -- (D);
+        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
+        \coordinate (midSE) at ($(S)!0.5!(E)$);
+        \draw [modulate,->] (Ia.north west) .. controls ($(Ia.north west)+(165:1)$) and ($(midSE)+(0,2)$) .. (midSE);
+        \draw [modulate,->] (Is.south west) .. controls ($(Is.south west)+(195:1)$) and ($(midSE)+(0,-2)$) .. (midSE);
+        \node[font=\normalsize] (demes) at ($(Is)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I_A},\lab{I_S}\}$};
+
+        %% SI2R (super-spreader) model
+        \coordinate (origin) at (9,-5);
+        \node[label] (lab) at (origin) {\textbf{D}};
+        \node [box] (S) at ($(origin)+(2,-1)$) {$\lab{S}$};
+        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
+        \node [box,ibox,deme] (Il) at ($(E)+(2,0)$) {$\lab{I_L}$};
+        \node [box,ibox,deme] (Ih) at ($(Il)+(0,-2)$) {$\lab{I_H}$};
+        \node [box] (R) at ($(Il)+(2,0)$) {$\lab{R}$};
+        \coordinate (overR) at ($(R)+(0,1)$);
+        \coordinate (midSE) at ($(S)!0.5!(E)$);
+        \draw [flow,->] (S) -- (E);
+        \draw [flow,->] (E) -- (Il);
+        \draw [flow,<->] (Il) -- (Ih);
+        \draw [flow,->] (Il) -- (R);
+        \draw [flow,->] (Ih.east) -- (R |- Ih.east) -- (R);
+        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
+        \draw [modulate,->] (Il.north west) .. controls ($(Il.north west)+(135:0.8)$) and ($(midSE)+(0,1)$) .. (midSE);
+        \draw [modulate,->] (Ih.west) .. controls ($(Ih.west)+(180:0.8)$) and ($(midSE)+(0,-1)$) .. (midSE);
+        \node[font=\normalsize] (demes) at (midSE |- Ih) {$\Demes=\{\lab{E},\lab{I_L},\lab{I_H}\}$};
       \end{tikzpicture}
@@
     Demes are shaded.
-    (A) An SEIRS model.
+    \textbf{(A)} An SEIRS model.
     Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
@@
     If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
-    \AAK{[Perhaps another one or two examples here?]}
     \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
@@
 In particular, we suppose that
-\begin{equation}\label{eq:ic}
+\begin{equation}
+  \label{eq:ic}
   \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
@@
 The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
-\begin{equation}\label{eq:kfe}
+\begin{equation}
+  \label{eq:kfe}
   \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
@@
         \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum size=5em]
-        \tikzstyle{trans}=[color=black, thick, >=stealth]
+        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum height=8em]
+        \tikzstyle{trans}=[color=black, font=\Large, thick, >=stealth]
         \node[state] (base) at (0,0) {$x=(S,E,I,R)$};
-        \node[state] (trans) at (165:8.4) {$x'=(S-1,E+1,I,R)$};
-        \node[state] (prog) at (195:8) {$x'=(S,E-1,I+1,R+1)$};
-        \node[state] (recov) at (295:4) {$x'=(S,E,I-1,R+1)$};
-        \node[state] (wane) at (355:9) {$x'=(S+1,E,I,R-1)$};
+        \node[state] (trans) at (165:8) {$x'=(S-1,E+1,I,R)$};
+        \node[state] (prog) at (205:7) {$x'=(S,E-1,I+1,R+1)$};
+        \node[state] (recov) at (305:5) {$x'=(S,E,I-1,R+1)$};
+        \node[state] (wane) at (350:8) {$x'=(S+1,E,I,R-1)$};
         \node[state] (sample) at (18:7.5) {$x'=(S,E,I,R)$};
@@
         \draw[trans,->] (base) -- (sample) node[midway,above,sloped] {$\lab{Sample}$};
-        \node[font=\Large] (U) at (95:3.2) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
+        \node[font=\LARGE] (U) at (95:3.5) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
       \end{tikzpicture}
@@
 
-\subsubsection{Structured populations, demes}
+\subsubsection{Structured populations, demes, and deme occupancy}
+\label{sec:demes}
 
@@
 \end{equation*}
-In \cref{fig:markov_state}, we use the marks to distinguish events of different types.
+In \cref{fig:markov_state}, we use the marks to distinguish events with distinct effects on genealogical relationships.
 Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.
@@
 The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
-\begin{equation*}\label{eq:Hdens}
+\begin{equation*}
+  \label{eq:Hdens}
   \begin{aligned}
@@
 
-\subsection{Binomial ratio}\label{sec:binomial_ratio}
+\subsection{Binomial ratio}
+\label{sec:binomial_ratio}
 
@@
 
-\subsection{Jump types}\label{sec:jump_types}
+\subsection{Jump types}
+\label{sec:jump_types}
+
+\begin{figure}
+  \begin{center}
+    \resizebox{0.8\linewidth}{!}{
+      \begin{tikzpicture}[scale=2]
+        \usetikzlibrary{shapes,arrows,positioning}
+        \usetikzlibrary{arrows.meta,positioning,decorations}
+        \colorlet{darkblue}{blue!50!black}
+        \colorlet{maize}{yellow!90!black}
+        \colorlet{grey}{black!30!white}
+        \colorlet{darkgreen}{green!40!black}
+        \colorlet{lightblue}{white!30!blue}
+        \tikzstyle{label}=[font=\Large]
+        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
+        \tikzstyle{deme1}=[color=grey, line width=3pt, >={Circle[width=5pt,length=5pt]}]
+        \tikzstyle{deme2}=[color=darkblue, line width=3pt, >={Circle[width=5pt,length=5pt]}]
+        \tikzstyle{deme3}=[color=maize, line width=3pt, >={Circle[width=5pt,length=5pt]}]
+        \tikzstyle{axis}=[color=black, thick, >=stealth]
+        \tikzstyle{timepoint}=[color=red!70!black]
+
+        \def\ysp{0.2} %% spacing between adjacent lineages
+
+        %% Birth-type event
+        \begin{scope}[shift={(0,0)}]
+          \coordinate (origin) at (0,0);
+          \node[label] (lab) at (origin) {\textbf{A}};
+          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
+          \coordinate (right) at ($(left)+(1,0)$);
+          \coordinate (mid) at ($(left)!0.5!(right)$);
+          \coordinate (rlab) at ($(right)+(0.1,0)$);
+          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
+          \coordinate (bright) at (rlab |- bleft);
+          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
+          \coordinate (temp1) at ($(temp)+(0,\ysp)$);
+          \coordinate (temp2) at ($(temp)-(0,\ysp)$);
+          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
+          \draw[deme3] (mid |- temp) -- (mid |- temp2) -- (right |- temp2);
+          \draw[deme1] (left |- temp) -- (right |- temp);
+          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
+          \foreach \y in {1,7} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme1] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {6,8} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme2] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {2,7} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme3] (left |- temp) -- (right |- temp);
+          }
+          \draw[timepoint] (mid) -- (mid |- bleft);
+          \draw[axis,->] (bleft) -- (bright);
+          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
+        \end{scope}
+
+        %% Death-type event
+        \begin{scope}[shift={(2,0)}]
+          \coordinate (origin) at (0,0);
+          \node[label] (lab) at (origin) {\textbf{B}};
+          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
+          \coordinate (right) at ($(left)+(1,0)$);
+          \coordinate (mid) at ($(left)!0.5!(right)$);
+          \coordinate (rlab) at ($(right)+(0.1,0)$);
+          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
+          \coordinate (bright) at (rlab |- bleft);
+          \foreach \y in {1,2,6} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme1] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {3,8} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme2] (left |- temp) -- (right |- temp);
+          }
+          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
+          \draw[deme3,dashed] (left |- temp) -- (mid |- temp);
+          \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
+          \foreach \y in {5,7} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme3] (left |- temp) -- (right |- temp);
+          }
+          \draw[timepoint] (mid) -- (mid |- bleft);
+          \draw[axis,->] (bleft) -- (bright);
+          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
+        \end{scope}
+
+        %% Migration-type event
+        \begin{scope}[shift={(4,0)}]
+          \coordinate (origin) at (0,0);
+          \node[label] (lab) at (origin) {\textbf{C}};
+          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
+          \coordinate (right) at ($(left)+(1,0)$);
+          \coordinate (mid) at ($(left)!0.5!(right)$);
+          \coordinate (rlab) at ($(right)+(0.1,0)$);
+          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
+          \coordinate (bright) at (rlab |- bleft);
+          \foreach \y in {2,3,8} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme1] (left |- temp) -- (right |- temp);
+          }
+          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
+          \draw[deme2] (left |- temp) -- (mid |- temp);
+          \draw[deme3] (mid |- temp) -- (right |- temp);
+          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
+          \foreach \y in {1,5} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme2] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {6,7} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme3] (left |- temp) -- (right |- temp);
+          }
+          \draw[timepoint] (mid) -- (mid |- bleft);
+          \draw[axis,->] (bleft) -- (bright);
+          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
+        \end{scope}
+
+        %% Sample event
+        \begin{scope}[shift={(6,0)}]
+          \coordinate (origin) at (0,0);
+          \node[label] (lab) at (origin) {\textbf{D}};
+          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
+          \coordinate (right) at ($(left)+(1,0)$);
+          \coordinate (mid) at ($(left)!0.5!(right)$);
+          \coordinate (rlab) at ($(right)+(0.1,0)$);
+          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
+          \coordinate (bright) at (rlab |- bleft);
+          \foreach \y in {2,5} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme1] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {1,4,7} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme2] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {3,6,8} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme3] (left |- temp) -- (right |- temp);
+          }
+          \coordinate (temp) at ($(origin)+(0,-2*\ysp)$);
+          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
+          \coordinate (temp) at ($(origin)+(0,-6*\ysp)$);
+          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
+          \draw[timepoint] (mid) -- (mid |- bleft);
+          \draw[axis,->] (bleft) -- (bright);
+          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
+        \end{scope}
+
+        %% Neutral-type event
+        \begin{scope}[shift={(0,-2)}]
+          \coordinate (origin) at (0,0);
+          \node[label] (lab) at (origin) {\textbf{E}};
+          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
+          \coordinate (right) at ($(left)+(1,0)$);
+          \coordinate (mid) at ($(left)!0.5!(right)$);
+          \coordinate (rlab) at ($(right)+(0.1,0)$);
+          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
+          \coordinate (bright) at (rlab |- bleft);
+          \foreach \y in {4,6,7} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme1] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {1,2,5} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme2] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {3,8} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme3] (left |- temp) -- (right |- temp);
+          }
+          \draw[timepoint] (mid) -- (mid |- bleft);
+          \draw[axis,->] (bleft) -- (bright);
+          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
+        \end{scope}
+
+        %% Compound birth/death (Moran) event
+        \begin{scope}[shift={(2,-2)}]
+          \coordinate (origin) at (0,0);
+          \node[label] (lab) at (origin) {\textbf{F}};
+          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
+          \coordinate (right) at ($(left)+(1,0)$);
+          \coordinate (mid) at ($(left)!0.5!(right)$);
+          \coordinate (rlab) at ($(right)+(0.1,0)$);
+          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
+          \coordinate (bright) at (rlab |- bleft);
+          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
+          \coordinate (temp1) at ($(temp)+(0,\ysp)$);
+          \draw[deme1] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
+          \draw[deme1] (left |- temp) -- (right |- temp);
+          \draw[deme3,dashed] (left |- temp1) -- (mid |- temp1);
+          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
+          \node[text=red,font=\Huge] at (mid |- temp1) {$\times$};
+          \foreach \y in {2,5} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme1] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {1,7} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme2] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {6,8} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme3] (left |- temp) -- (right |- temp);
+          }
+          \draw[timepoint] (mid) -- (mid |- bleft);
+          \draw[axis,->] (bleft) -- (bright);
+          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
+        \end{scope}
+
+        %% Compound sample/death event
+        \begin{scope}[shift={(4,-2)}]
+          \coordinate (origin) at (0,0);
+          \node[label] (lab) at (origin) {\textbf{G}};
+          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
+          \coordinate (right) at ($(left)+(1,0)$);
+          \coordinate (mid) at ($(left)!0.5!(right)$);
+          \coordinate (rlab) at ($(right)+(0.1,0)$);
+          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
+          \coordinate (bright) at (rlab |- bleft);
+          \coordinate (temp) at ($(origin)+(0,-5*\ysp)$);
+          \draw[deme3] (left |- temp) -- (mid |- temp);
+          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
+          \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
+          \foreach \y in {1,6} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme1] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {2,3,8} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme2] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {4,7} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme3] (left |- temp) -- (right |- temp);
+          }
+          \draw[timepoint] (mid) -- (mid |- bleft);
+          \draw[axis,->] (bleft) -- (bright);
+          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
+        \end{scope}
+
+        %% Compound event
+        \begin{scope}[shift={(6,-2)}]
+          \coordinate (origin) at (0,0);
+          \node[label] (lab) at (origin) {\textbf{H}};
+          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
+          \coordinate (right) at ($(left)+(1,0)$);
+          \coordinate (mid) at ($(left)!0.5!(right)$);
+          \coordinate (rlab) at ($(right)+(0.1,0)$);
+          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
+          \coordinate (bright) at (rlab |- bleft);
+          \coordinate (temp) at ($(origin)+(0,-3*\ysp)$);
+          \draw[deme3] (left |- temp) -- (mid |- temp);
+          \draw[deme2] (mid |- temp) -- (right |- temp);
+          \coordinate (temp1) at ($(temp)+(0,-\ysp)$);
+          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
+          \coordinate (temp1) at ($(temp1)+(0,-\ysp)$);
+          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
+          \coordinate (temp1) at ($(temp1)+(0,-\ysp)$);
+          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
+          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
+          \foreach \y in {2,7} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme1] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {1} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme2] (left |- temp) -- (right |- temp);
+          }
+          \foreach \y in {8} {
+            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
+            \draw[deme3] (left |- temp) -- (right |- temp);
+          }
+          \draw[timepoint] (mid) -- (mid |- bleft);
+          \draw[axis,->] (bleft) -- (bright);
+          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
+        \end{scope}
+      \end{tikzpicture}
+    }
+  \end{center}
+  \caption{
+    Jump types differ by their effects on the genealogy.
+    \textbf{(A)} A birth-type jump results in the branching of one or more child lineages from the parent.
+    There can be only one parent, though the child lineages may be in demes different from those of their parent.
+    Here, a parent of the grey deme sires one child lineage in each of the blue and yellow demes.
+    The production of an event is an integer vector, with one entry for each deme.
+    The production of this event is therefore $r=(r_{\lab{grey}},r_{\lab{blue}},r_{\lab{yellow}})=(1,1,1)$.
+    \textbf{(B)} A death-type event causes the extinction of a lineage.
+    Since internal node without children are recursively removed, the affected branch disappears.
+    The production of this event is $r=(0,0,0)$.
+    \textbf{(C)} A migration-type event results in the movement of one or more lineages from one deme to another.
+    Here, one lineage moves from the blue to the yellow deme.
+    The production of this event is $r=(0,0,1)$, \ie the production is 0 for the grey and blue demes, and 1 for the yellow.
+    \textbf{(D)} In a sample-type event, one or more sample nodes (blue circles) are inserted.
+    Here, there are two samples, one in each of the grey and yellow demes.
+    Accordingly, its production is $r=(1,0,1)$.
+    \textbf{(E)} A neutral-type event has no effect on the genealogy and zero production in all demes: $r=(0,0,0)$.
+    \textbf{(F)} The theory presented here allows for compound events.
+    As an example, here a birth/death-type event occurs, wherein one yellow lineage is extinguished and a grey lineage simultaneously sires a grey child.
+    The production of this event is $r=(2,0,0)$.
+    \textbf{(G)} Here, a compound sample/death-type event occurs.
+    A yellow lineage is sampled and simultaneously extinguished.
+    Note that the lineage is not recursively removed, since sample nodes are never removed.
+    \textbf{(H)} A compound birth/migration-type event with yellow parent and production $r=c(0,4,0)$.
+    \label{fig:jump_types}
+  }
+\end{figure}
 
@@
 
-\begin{figure}
-  \caption{
-    Illustration of genealogy processes.
-    \AAK{[Similar to that of \citet{King2022} but with multiple demes represented.]}
-  }
-\end{figure}
-
 \subsubsection{Emergent lineages and production}
@@
 
+\begin{figure}
+  \caption{
+    Unpruned, pruned, and obscured enealogies.
+  }
+\end{figure}
+
+The process just described yields a genealogy that relates all extant members of the population, and all samples.
+Moreover, it details each lineage's complete history of movement through the various demes.
+The data we ultimately wish to analyze will be based only on samples, however;
+nor, in general, will the histories of deme occupancy be observable.
+A generative model must account for this loss of information.
+We therefore now describe how genealogies are \emph{pruned} to yield sample-only genealogies and then \emph{obscured} via the erasure of color from their branches.
+
 \subsubsection{Pruned genealogy}
 
-Although the process just described yields a genealogy that relates all extant members of the population, and all samples, the data we ultimately wish to analyze will be based only on samples.
-We therefore describe how the genealogy process is \emph{pruned} to yield the sample-only genealogy.
 Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Pr_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every childless internal node.
@@
 \subsubsection{Lineage count and saturation}
+\label{sec:ells}
 
@@
 
-\subsubsection{Obscured genealogy}
-
-As we have seen, a pruned genealogy contains information about the full history of each sample lineage, including the times at which it entered or exited any deme, sired offspring, or was sampled.
-The data we seek to analyze will typically lack much of this information.
-Accordingly, we define the \emph{obscured genealogy} to be that obtained by discarding all information about demes and events not visible from the topology of the tree alone.
-In particular, if $\Pr_T=(y,z)$ is a pruned genealogy, we write $\obs(\Pr_T)=z$ to denote the obscured genealogy.
-
-\section{Results}
-
-\subsection{Likelihood for pruned genealogies}
-
-Our first result will be an expression for the likelihood of a given pruned genealogy $\P^*_T$ given the history $\Hr_T$ of the population process to time $T$.
-Of course, not every history is compatible with $\P^*_T$.
-In particular, $\Hr_T$ is only compatible with $\P^*_T$ if every node-time in $\P^*_T$ is an event time of $\Hr_T$, \ie only if $\event{\Hr}{T}\supseteq\event{P}{T}^*$.
-Moreover, even if an event in $\Hr_T$ coincides with one of the node times of $\P^*_T$, it may be incompatible with the local structure of $\P^*_T$.
-For example, if no sample is present in $\P^*_T$ at time $t$ and $\Hr_T$ contains an event of sample type at that time, then incompatibility arises.
-As another example, if $\P^*_T$ has a node at time $t$ at which a lineage moves from deme $i$ to deme $i'$, but there is no $u\in\Jumps$ for which such a move is possible, then incompatibility arises again.
-To keep track of this, let us introduce the indicator function $Q$ such that $Q_u(y,y',z,z')=1$ if and only if there is some pruned genealogy $\P_T$ and $t\in{[0,T]}$,
+\subsubsection{Compatibility}
+\label{sec:compatibility}
+
+Suppose $\P_T$ is a pruned genealogy and $t\in\event{P}{T}$.
+The local structure of $\P_T$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
+For example, if the event in $\P_T$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
+Similarly, if the node in $\P_T$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
+To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q_u(y,y',z,z')=1$ if and only if there is some pruned genealogy $\P_T$ and $t\in{[0,T]}$,
 such that $\yt^{\P_T}_t=y$, $y^{\P_T}_t=y'$, $\zt^{\P_T}_t=z$, $z^{\P_T}_t=z'$,
@@
 $Q=0$ otherwise.
+We refer to $Q$ as the \emph{compatibility indicator}.
+
+\subsubsection{Obscured genealogy}
+
+The \emph{obscured genealogy} is obtained by discarding all information about demes and events not visible from the topology of the tree alone.
+In particular, if $\Pr_T=(y,z)$ is a pruned genealogy, we write $\obs(\Pr_T)=z$ to denote the obscured genealogy.
+
+\section{Results}
+
+\subsection{Likelihood for pruned genealogies}
+
+Our first result will be an expression for the likelihood of a given pruned genealogy $\P^*_T$ given the history $\Hr_T$ of the population process to time $T$.
 
@@
   \begin{equation}\label{eq:phidef}
-    \phi_u(x,y,y',z,z')\coloneq\BinRatio{n(x)}{\ell(y',z')}{r^{u}}{s(y,y',z,z')}\,Q_{u}(y,y',z,z')
+    \phi_u(x,y,y',z,z')\coloneq\BinRatio{n(x)}{\ell(y',z')}{r^{u}}{s(y,y',z,z')}\,Q_{u}(y,y',z,z'),
   \end{equation}
-  (cf.~\cref{sec:binomial_ratio}).
+  where $n$ is the deme occupancy (\cref{sec:demes}), $\ell$ and $s$ are the lineage-count and saturation functions, respectively (\cref{sec:ells}), $Q$ is the compatibility indicator (\cref{sec:compatibility}), and the binomial ratio is as defined in \cref{sec:binomial_ratio}.
   Then
@@
 \begin{proof}
-  As we have already observed, if $\event{\Hr}{T}\nsupseteq\event{\P}{T}^*$, then $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
-  Similarly, if there is any event of $\Hr_T$ which is incompatible with $\P^*_T$, $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
+  If $\event{\Hr}{T}\nsupseteq\event{\P}{T}^*$, then $\Hr_T$ and $\P^*_T$ are incompatible and $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
+  Similarly, if any event of $\Hr_T$ is incompatible with the local structure of $\P^*_T$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
   Let us therefore suppose that neither of these conditions hold.
@@
   Suppose $\Xr_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
-  \begin{equation}\label{eq:kfe2}
-    \frac{\partial{u}}{\partial{t}}
+  \begin{equation}
+    \label{eq:kfe2}
+    \frac{\partial{u}}{\partial{t}}(t,x)
     =\int{u(t,x')\,\beta(t,x',x)\,\dd{x'}}
@@
   Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)$ are given, real-valued, measurable functions.
-  Let $\mathcal{M}\subseteq\Rp$ be countable such that $\mathcal{M}\cap[0,t]$ is finite for all $t$.
   We say that the equation
-  \begin{equation}\label{eq:filtereq}
+  \begin{equation}
+    \label{eq:filtereq}
     \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
+      \frac{\partial{w}}{\partial{t}}(t,x)
       =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
       -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
-      -\lambda(t,x)\,w(t,x)\\
-      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}.
+      -\lambda(t,x)\,w(t,x).\\
     \end{aligned}
   \end{equation}
-  is the \emph{filter equation} with \emph{driver} $\Xr_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$.
+  is the \emph{filter equation} with \emph{driver} $\Xr_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, and \emph{decay} $\lambda$.
 \end{defn}
 
-\begin{remark}\label{rem:equivfilt}
-  \Cref{eq:filtereq} is equivalent to
-  \begin{equation*}
-    \begin{gathered}
-      \frac{\partial{w}}{\partial{t}}
-      =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
-      -\lambda(t,x)\,w(t,x),\quad t\notin{\mathcal{M}},\\
-      w(t,x)=\int{\leftlim{w}(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}},\quad t\in{\mathcal{M}}.
-    \end{gathered}
-  \end{equation*}
-  Here $\wt$ refers to the left-limit of $w$.
+\begin{remark}
+  Trivially, a Kolmogorov forward equation is itself a filter equation with boost $1$ and decay $0$.
 \end{remark}
 
-\begin{remark}
-  Trivially, a Kolmogorov forward equation is itself a filter equation with boost $1$, decay $0$, and no observed events ($\mathcal{M}=\emptyset$).
-\end{remark}
+The following results show how filter equations allow one to integrate over random histories.
+
+\begin{lemma}
+  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable.
+  If $\Vr_t$ is an $\Rp$-valued random process satisfying
+  \begin{equation*}
+    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)},
+  \end{equation*}
+  then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued deterministic process on $\Xspace$ with density $w(t,x)$.
+  Moreover, $w$ satisfies the filter equation
+  \begin{equation*}
+    \frac{\partial{w}}{\partial{t}}=\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}
+  \end{equation*}
+  and $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
+\end{lemma}
+\begin{proof}
+  The expectation can be broken into three terms, according to whether $\H_t$ has zero, one, or more than one event in $\halfclosed{t-\Delta,t}$.
+  Accordingly, as $\Delta\downarrow{0}$,
+  \begin{equation*}
+    \begin{aligned}
+      w(t,x)=&\left(1-\Delta\,\int{\alpha(t-\Delta,x,x')\,\dd{x'}}\right)\,w(t-\Delta,x)\\
+      &\qquad+\Delta\,\int{\alpha(t-\Delta,x',x)\,B(t-\Delta,x',x)\,w(t-\Delta,x')\,\dd{x'}}+o(\Delta).
+    \end{aligned}
+  \end{equation*}
+  Taking the limit as $\Delta\downarrow{0}$, we obtain
+  \begin{equation*}
+    \frac{\partial{w}}{\partial{t}}(t,x)
+    =\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
+    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
+  \end{equation*}
+\end{proof}
+
+\begin{lemma}
+  Suppose $\mathcal{M}\subset{\Rp}$ is finite and let $\mathcal{M}_t\coloneq\mathcal{M}\cap{[0,t]}$.
+  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable and
+  $\Vr_t$ is an $\Rp$-valued process satisfying
+  \begin{equation*}
+    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\Indicator{\event{\H}{t}\supseteq\mathcal{M}_t}\,\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)}.
+  \end{equation*}
+  Then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued process on $\Rp^{\vert\mathcal{M}_t\vert}\times\Xspace$ with density $w(t,x)$.
+  Furthermore, $w(t,x)$ satisfies
+  \begin{equation*}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}(t,x)&=
+      \int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
+      &t\notin{\mathcal{M}},\\
+      w(t,x)&=\int{\wt(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
+      &t\in{\mathcal{M}}
+    \end{aligned}
+  \end{equation*}
+  and $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
+\end{lemma}
 
 Filter equations afford a convenient means of computing expectations and likelihoods for pure jump processes.
-This is facilitated by the following fact.
-
-\begin{lemma}\label{lemma:filtereq}
-  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ and $B>0$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satisfies the KFE
-  \begin{equation}\label{eq:filterlemma}
+This is facilitated by the following Lemma, the statement of which relies on a non-standard Dirac delta function.
+Let $\delta(v,v')$ be the right-sided Dirac delta function satisfying $\delta(v,v')=0$ for $v\ne{v'}$ and $\int_a^b{f(v)\,\delta(v,v')\,\dd{v}}=f(v')\,\Indicator{v'\in\halfopen{a,b}}$ for all \cadlag\ $f$.
+
+\begin{lemma}
+  \label{lemma:filter}
+  The filter equation \eqref{eq:filtereq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satsfies the KFE
+  \begin{equation}
+    \label{eq:filterlemma}
     \begin{aligned}
@@
   \end{equation}
-  on the space $\Xspace\times(0,\infty)$.
-  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
+  on the space $\Xspace\times\Rp$.
 \end{lemma}
@@
 The $\Vr_t$ is \emph{directed} by $\Xr_t$ in the sense that $\Vr$ has jumps wherever $\Xr$ does:
-when $\Xr$ jumps at time $t$ from $x$ to $x'$, $\Vr$ jumps by the multiplicative factor $B(t,x,x')>0$.
+when $\Xr$ jumps at time $t$ from $x$ to $x'$, $\Vr$ jumps by the multiplicative factor $B(t,x,x')\ge{0}$.
 Between jumps, $\Vr_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
-If we view $\Vr_t$ as a weight, then \cref{lemma:filtereq} tells us how the $\Vr_t$-weighted average of $\Xr_t$ evolves in time:
+If we view $\Vr_t$ as a weight, then \cref{lemma:filter} tells us how the $\Vr_t$-weighted average of $\Xr_t$ evolves in time:
 this average is simply $\int\!{w(t,x)\,\dd{x}}$.
-This motivates the following result, which shows how filter equations allow one to integrate over random histories.
-
-\begin{prop}\label{prop:zeroboost}
-  Suppose $\Xr_t$ is a non-explosive pure jump process with KFE
-  \begin{equation*}
-    \begin{gathered}
-      \frac{\partial{u}}{\partial{t}}(t,x)=\int{u(t,x')\,\alpha(t,x',x)\,\dd{x'}}
-      -\int{u(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
-      u(0,x)=p_0(x).
-    \end{gathered}
-  \end{equation*}
-  Let $\Hr_t$ be its history process.
-  Suppose $\Vr_t$ is an $\Rp$-valued random process such that
-  \begin{equation}\label{eq:Vprod}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)},
-  \end{equation}
-  for some given measurable function $B\ge{0}$.
-  If $w$ satisfies the filter equation,
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{t}}(t,x)
-    =\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
-    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
-  \end{equation*}
-  then $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
-\end{prop}
-\begin{proof}
-  Define $Q(t,x,x')=\Indicator{B(t,x,x')>0}$ and apply \cref{lemma:filtereq} with
-  \begin{equation*}
-    \begin{gathered}
-      \beta(t,x,x')=\alpha(t,x,x')\,Q(t,x,x'),
-      \qquad
-      \lambda=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\dd{x'}}.
-    \end{gathered}
-  \end{equation*}
-  \ELI{[Lemma A1 has $B>0$, which only covers the cases when $Q=1$ and $\lambda=0$.
-      But, $B\ge{0}$ seems enough in the proof of lemma A1, as long as one allows a one-sided Dirac $\delta$ at $0$.
-      Maybe the Dirac $\delta$ should all be one-sided in order to integrate
-      correctly on right-continuous functions? ]}
-\end{proof}
-
-%% An important special case is that of a deterministic driving process.
-%% The following result is established by routine calculation.
-
-%% \begin{prop}
-%%   Suppose $X:[0,T]\to\Xspace$ is a deterministic, piecewise constant, \cadlag\ function and let $M=\event(X)$.
-%%   Then the KFE for $X_t$ is \cref{eq:kfe2} with $\beta(t,x,x')=\sum_{e\in{M}}{\delta(t,e)\delta(x',X_e)}$.
-%%   With this driver, the filter equation (\cref{eq:filtereq}) becomes
-%%   \begin{equation*}
-%%     \begin{gathered}
-%%       \frac{\partial{w}}{\partial{t}}=
-%%       -\lambda(t,x)\,w(t,x)\quad\text{for}\quad{t\in[0,T]\setminus{M}},\\
-%%       w(t,x)=\delta(x,X_t)\,\int{\wt(t,x')\,B(t,x',X_t)\,\dd{x'}}\quad\text{for}\quad{t\in{M}}.
-%%     \end{gathered}
-%%   \end{equation*}
-%% \end{prop}
-
-The following result extends \cref{prop:zeroboost} to the situation where $\mathcal{M}\ne\emptyset$, \ie where the times of some of the events in the history are observed.
-
-\begin{prop}\label{prop:fullfilt}
-  Let $\Xr_t$, $\Hr_t$, $B$, and $\Vr_t$ be as in \cref{prop:zeroboost}.
-  Suppose that $\mathcal{M}\subseteq\halfclosed{0,t}$ is a fixed, finite set of observed event times.
-  Let $\CondExpect{\Vr_t}{\event{\Hr}{t}\supseteq\mathcal{M}}$ denote the expectation of $\Vr_t$ conditional on $\event{\Hr}{t}\supseteq\mathcal{M}$.
-  If $w(t,x)$ satisfies $w(0,x)=p_0(x)$ and the filter equation
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
-      =&\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
-      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\},
-    \end{aligned}
-  \end{equation*}
-  then $\CondExpect{\Vr_t}{\event{\Hr}{t}\supseteq\mathcal{M}}=\int{w(t,x)\,\dd{x}}$.
-  \ELI{[(a) line 2: density $\rightarrow$ expectation.
-      (b) the term $\delta(t,e)\,w(t,x)$ seems irrelevant:
-      the derivative at a finite set of points only matters when it is a delta function.]}
-\end{prop}
-\begin{proof}
-  Suppose $s\in[0,t]$ and let
-  \begin{equation*}
-    w(s,x)=\Expect{\prod_{e\in\event{\Hr}{s}}{B\left(e,\Xrt_e,\Xr_e\right)}}.
-  \end{equation*}
-  That is, $w(s,x)$ is the expectation (over histories) of the partial products of \cref{eq:Vprod}.
-  If $s\notin{\mathcal{M}}$, then for $\Delta>0$ sufficiently small,
-  \begin{equation*}
-    \begin{aligned}
-      w(s,x)=&\left(1-\Delta\,\int{\alpha(s-\Delta,x,x')\,\dd{x'}}\right)\,w(s-\Delta,x)\\
-      &\qquad+\int{\alpha(s-\Delta,x',x)\,B(s-\Delta,x',x)\,w(s-\Delta,x')\,\dd{x'}}+o(\Delta).
-    \end{aligned}
-  \end{equation*}
-  Taking the limit as $\Delta\downarrow{0}$, we obtain
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{s}}(s,x)
-    =\int{w(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}
-    -\int{w(s,x)\,\alpha(s,x,x')\,\dd{x'}}.
-  \end{equation*}
-  If $s\in\mathcal{M}$, then
-  \begin{equation*}
-    w(s,x)=\int{\wt(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}.
-  \end{equation*}
-  By \cref{rem:equivfilt}, the result follows.
-  \ELI{[The argument shows that the filter equation satisfies a KFE, but the limit expression seems justified by a similar KFE.
-      Is the point that one KFE implies the other?]}
-\end{proof}
-
-\ELI{[Maybe, Prop A3 could be a version of Prop A2 allowing for discrete transitions at a finite set of times;
-    a combination of cts time and discrete time Markov chains.
-    That is all we need to extend Prop A2 to observation times, which can be viewed just as an extended Markov process including the observation events.
-    That is, to make a distinction between ``observation events'' that give the sequences used to construct the phylogeny, and ``observed events'' which are determined points on the phylogeny.
-    The latter are a realization of a random process, and the former are defined to be fixed a priori.]}
 

### 2024-01-21 figs/event_types.tex
more figures

--- 2024-01-20 ms.Rnw
@@
-\documentclass[10pt,reqno,final]{amsart}
-\input{header}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta,positioning,decorations,calc,math}
-
-\title[Exact Phylodynamics]{Exact Phylodynamics via Structured Markov Genealogy Processes}
-\author[King]{Aaron~A.~King}
-\address{
-  A.~A.~King,
-  Department of Ecology \& Evolutionary~Biology,
-  Center for the Study of Complex~Systems, and
-  Department of Mathematics,
-  University of Michigan,
-  Ann~Arbor, MI~48109~USA\\
-  Santa~Fe~Institute,
-  1399 Hyde~Park~Road,
-  Santa~Fe, NM~87501~USA
-}
-\email{kingaa@umich.edu}
-\urladdr{\href{https://kinglab.eeb.lsa.umich.edu/}{https://kinglab.eeb.lsa.umich.edu/}}
-\author[Lin]{Qianying~Lin}
-\address{
-  Q.-Y.~Lin,
-  Theoretical Biology and Biophysics,
-  Los~Alamos National Laboratory,
-  Los~Alamos, NM~87545~USA
-}
-\author[Ionides]{Edward~L.~Ionides}
-\address{
-  E.~L.~Ionides,
-  Department of Statistics
-  University of Michigan,
-  Ann~Arbor, MI~48109~USA
-}
-\date{\today}
-
-\hypersetup{pdftitle={Exact Phylodynamics via Structured Markov Genealogy Processes}}
-\hypersetup{pdfauthor={A.A. King, Q.-Y. Lin, E.L. Ionides}}
-\hypersetup{urlcolor=blue,citecolor=blue,linkcolor=blue,filecolor=blue}
-
-<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
-prefix <- "sgp"
-source("setup.R")
-@
-<<packages,include=FALSE,cache=FALSE>>=
-library(tidyverse)
-library(ggtree)
-library(pomp)
-library(cowplot)
-library(phylopomp)
-stopifnot(getRversion() >= "4.3")
-stopifnot(packageVersion("pomp")>="5.1")
-stopifnot(packageVersion("phylopomp")>="0.9.2")
-theme_set(theme_bw(base_family="serif"))
-set.seed(1159254136)
-@
-
-\begin{document}
-
-\begin{abstract}
-  We derive an algorithm for the exact computation of the likelihood of an observed transmission tree.
-\end{abstract}
-
-\maketitle
-
-\section{Introduction}
-
-Problem of phylodynamics.
-Factorization of problem into two subproblems.
-
-Relation to previous work.
-Existing methods \citep{Volz2009a,Stadler2010}.
-Large-population, small sample-size approximations.
-
-Extension of previous results \citep{King2022}.
-Broader class of state-spaces.
-Accommodating discrete structure.
-
-Classes of Markov processes.
-Utility and flexibility of Markov assumptions.
-
-Population process induces Markov history and genealogy processes.
-Using these, we derive equations for the likelihood of a genealogy conditional on the history.
-We then integrate out the history to obtain nonlinear filtering equations, the solution of which yields the likelihood.
-These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likelihood.
-We demonstrate with several examples.
-
-In the following, we show a Markov population process of the kind that is a staple in epidemiology induces a Markov process on the space of genealogies.
-We then show how one can comput the likelihood of a given genealogy.
-
-\section{Mathematical preliminaries}
-
-\subsection{Notation}
-
-Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\Xr$), denotes a random element.
-We will be concerned with a variety of stochastic processes, in both discrete and continuous time.
-In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
-In the case of continuous-time processes, we will assume that sample paths are right-continuous with left limits (\ie \cadlag).
-We will frequently need to refer to the left-limit of such a process.
-Accordingly, if $\Zr_t$ is a \cadlag\ random process, we define
-\begin{equation*}
-  \Zrt_t\colonequals
-  \begin{cases}
-    \displaystyle\lim_{t'\,\uparrow\,{t}}\;\Zr_{t'}, & t>0,\\[2ex]
-    \Zr_0, & t=0.\\
-  \end{cases}
-\end{equation*}
-Note that $\Zrt_t$ is thus left-continuous with right limits.
-
-If $Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
-In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
-$\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
-If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its event times in the interval $[0,t]$.
-
-\subsection{Population process}
-
-Motivating examples: compartmental models.
-Wide variety of models.
-Linear chain trick.
-Migration, superspreading, competition between strains.
-
-\begin{figure}
-  \begin{center}
-    \resizebox{0.9\linewidth}{!}{
-      \begin{tikzpicture}[scale=1]
-        \colorlet{darkblue}{blue!50!black}
-        \tikzstyle{box}=[draw=black, text=black, fill=white, very thick, minimum size=3em]
-        \tikzstyle{ibox}=[draw=darkblue]
-        \tikzstyle{deme}=[fill=black!20!white]
-        \tikzstyle{label}=[font=\Large]
-        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{flow}=[draw=black, very thick, >=stealth]
-        \tikzstyle{modulate}=[draw=darkblue, >=Circle]
-
-        %% SEIRS model
-        \coordinate (origin) at (0,0);
-        \node[label] (lab) at (origin) {\textbf{A}};
-        \node [box] (S) at ($(origin)+(1,-1)$) {$\lab{S}$};
-        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
-        \node [box,ibox,deme] (I) at ($(E)+(2,0)$) {$\lab{I}$};
-        \node [box] (R) at ($(I)+(2,0)$) {$\lab{R}$};
-        \coordinate (overR) at ($(R)+(0,1)$);
-        \coordinate (midSE) at ($(S)!0.5!(E)$);
-        \draw [flow,->] (S) -- (E);
-        \draw [flow,->] (E) -- (I);
-        \draw [flow,->] (I) -- (R);
-        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
-        \draw [modulate,->] (I.north west) .. controls ($(I)+(-1,1)$) and ($(midSE)+(0,1)$) .. (midSE);
-        \node[font=\normalsize] (demes) at ($(E)!0.5!(I)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I}\}$};
-
-        %% SIIR (two-strain) model
-        \coordinate (origin) at (0,-3);
-        \node[label] (lab) at (origin) {\textbf{B}};
-        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
-        \node [box] (R1) at ($(S)+(6,1)$) {$\lab{R}_1$};
-        \node [box] (R2) at ($(S)+(6,-1)$) {$\lab{R}_2$};
-        \coordinate (overR) at ($(R1)+(0,1)$);
-        \coordinate (underR) at ($(R2)-(0,1)$);
-        \node [box,deme] (E1) at ($(S)+(2,1)$) {$\lab{E}_1$};
-        \node [box,deme] (E2) at ($(S)+(2,-1)$) {$\lab{E}_2$};
-        \node [box,ibox,deme] (I1) at ($(E1)+(2,0)$) {$\lab{I}_1$};
-        \node [box,ibox,deme] (I2) at ($(E2)+(2,0)$) {$\lab{I}_2$};
-        \draw [flow,->] (S) -- (E1);
-        \draw [flow,->] (E1) -- (I1);
-        \draw [flow,->] (I1) -- (R1);
-        \draw [flow,->] (S) -- (E2);
-        \draw [flow,->] (E2) -- (I2);
-        \draw [flow,->] (I2) -- (R2);
-        \draw [flow,->] (R1) -- (overR) -- (S |- overR) -- (S);
-        \draw [flow,->] (R2) -- (underR) -- (S |- underR) -- (S);
-        \coordinate (midSE1) at ($(S)!0.5!(E1)$);
-        \coordinate (midSE2) at ($(S)!0.5!(E2)$);
-        \draw [modulate,->] (I1.north west) .. controls ($(I1)+(-1,1)$) and ($(midSE1)+(0,1.8)$) .. (midSE1);
-        \draw [modulate,->] (I2.south west) .. controls ($(I2)+(-1,-1)$) and ($(midSE2)+(0,-1.8)$) .. (midSE2);
-        \node[font=\normalsize] (demes) at ($(E2)!0.5!(I2)+(0,-2)$) {$\Demes=\{\lab{E}_1,\lab{I}_1,\lab{E}_2,\lab{I}_2\}$};
-
-        %% COVID model
-        \coordinate (origin) at (9,0);
-        \node[label] (lab) at (origin) {\textbf{C}};
-        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
-        \node [box] (R) at ($(S)+(7,1)$) {$\lab{R}$};
-        \coordinate (overR) at ($(R)+(0,1)$);
-        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
-        \node [box,ibox,deme] (Ia) at ($(E)+(2,1)$) {$\lab{I_A}$};
-        \node [box,ibox,deme] (Is) at ($(E)+(2,-1)$) {$\lab{I_S}$};
-        \node [box] (H) at ($(Is)+(2,0)$) {$\lab{H}$};
-        \node [box] (D) at ($(H)+(2,0)$) {$\lab{D}$};
-        \draw [flow,->] (S) -- (E);
-        \draw [flow,->] (E) -- (Ia.west);
-        \draw [flow,->] (E) -- (Is.west);
-        \draw [flow,->] (Ia) -- (R);
-        \draw [flow,->] (Is) -- (H);
-        \draw [flow,->] (H) -- (R);
-        \draw [flow,->] (Is) -- (R);
-        \draw [flow,->] (H) -- (D);
-        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
-        \coordinate (midSE) at ($(S)!0.5!(E)$);
-        \draw [modulate,->] (Ia.north west) .. controls ($(Ia.north west)+(165:1)$) and ($(midSE)+(0,2)$) .. (midSE);
-        \draw [modulate,->] (Is.south west) .. controls ($(Is.south west)+(195:1)$) and ($(midSE)+(0,-2)$) .. (midSE);
-        \node[font=\normalsize] (demes) at ($(Is)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I_A},\lab{I_S}\}$};
-
-        %% SI2R (super-spreader) model
-        \coordinate (origin) at (9,-5);
-        \node[label] (lab) at (origin) {\textbf{D}};
-        \node [box] (S) at ($(origin)+(2,-1)$) {$\lab{S}$};
-        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
-        \node [box,ibox,deme] (Il) at ($(E)+(2,0)$) {$\lab{I_L}$};
-        \node [box,ibox,deme] (Ih) at ($(Il)+(0,-2)$) {$\lab{I_H}$};
-        \node [box] (R) at ($(Il)+(2,0)$) {$\lab{R}$};
-        \coordinate (overR) at ($(R)+(0,1)$);
-        \coordinate (midSE) at ($(S)!0.5!(E)$);
-        \draw [flow,->] (S) -- (E);
-        \draw [flow,->] (E) -- (Il);
-        \draw [flow,<->] (Il) -- (Ih);
-        \draw [flow,->] (Il) -- (R);
-        \draw [flow,->] (Ih.east) -- (R |- Ih.east) -- (R);
-        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
-        \draw [modulate,->] (Il.north west) .. controls ($(Il.north west)+(135:0.8)$) and ($(midSE)+(0,1)$) .. (midSE);
-        \draw [modulate,->] (Ih.west) .. controls ($(Ih.west)+(180:0.8)$) and ($(midSE)+(0,-1)$) .. (midSE);
-        \node[font=\normalsize] (demes) at (midSE |- Ih) {$\Demes=\{\lab{E},\lab{I_L},\lab{I_H}\}$};
-      \end{tikzpicture}
-    }
-  \end{center}
-  \caption{
-    Examples of discretely-structured population models.
-    Demes are shaded.
-    \textbf{(A)} An SEIRS model.
-    Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
-    Upon recovery ($\lab{R}$), individuals experience immunity from reinfection.
-    If this immunity wanes, they re-enter the susceptible compartment.
-    Pathogen lineages are to be found in hosts within the $\lab{E}$ and $\lab{I}$ compartments only.
-    Accordingly, there are two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
-    If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
-    \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
-    \label{fig:examples}
-  }
-\end{figure}
-
-We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
-In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
-The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
-In particular, we suppose that
-\begin{equation}
-  \label{eq:ic}
-  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
-\end{equation}
-for all measurable sets $\mathcal{E}\subseteq\Xspace$.
-For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
-More precisely, the transition rates have the following properties:
-\begin{equation*}
-  \begin{gathered}
-    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
-  \end{gathered}
-\end{equation*}
-for all $t\in\Rp$ and $x,x'\in\Xspace$.
-Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
-Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
-We assume that $\Kr_t$ is a simple counting process so that
-\begin{equation*}
-  \begin{gathered}
-    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
-    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
-    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
-  \end{gathered}
-\end{equation*}
-We will further assume that $\Xr_t$ is non-explosive, \ie that $\Prob{\Kr_t<\infty}=1$ for all $t$.
-
-\subsubsection{Kolmogorov forward equation}
-
-The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
-\begin{equation}
-  \label{eq:kfe}
-  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
-\end{equation}
-and if, moreover, $w(0,x) = p_0(x)$,
-then $\int_{\mathcal{E}}\!{w(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
-\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.
-
-Another perspective on the Markov processes is to be had from its Markov state transition diagram (\cref{fig:markov_state}).
-
-\begin{figure}
-  \begin{center}
-    \resizebox{0.9\linewidth}{!}{
-      \begin{tikzpicture}[scale=1]
-        \usetikzlibrary{shapes,arrows,positioning}
-        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum height=8em]
-        \tikzstyle{trans}=[color=black, font=\Large, thick, >=stealth]
-        \node[state] (base) at (0,0) {$x=(S,E,I,R)$};
-        \node[state] (trans) at (165:8) {$x'=(S-1,E+1,I,R)$};
-        \node[state] (prog) at (205:7) {$x'=(S,E-1,I+1,R+1)$};
-        \node[state] (recov) at (305:5) {$x'=(S,E,I-1,R+1)$};
-        \node[state] (wane) at (350:8) {$x'=(S+1,E,I,R-1)$};
-        \node[state] (sample) at (18:7.5) {$x'=(S,E,I,R)$};
-        \draw[trans,->] (base) -- (trans) node[midway,above,sloped] {$\lab{Trans}$};
-        \draw[trans,->] (base) -- (prog) node[midway,above,sloped] {$\lab{Prog}$};
-        \draw[trans,->] (base) -- (recov) node[midway,above,sloped] {$\lab{Recov}$};
-        \draw[trans,->] (base) -- (wane) node[midway,above,sloped] {$\lab{Wane}$};
-        \draw[trans,->] (base) -- (sample) node[midway,above,sloped] {$\lab{Sample}$};
-        \node[font=\LARGE] (U) at (95:3.5) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
-      \end{tikzpicture}
-    }
-  \end{center}
-  \caption{
-    Markov state transition diagram for the SEIRS model depicted in \cref{fig:examples}A.
-    The state, $x$, is characterized by four numbers, $S$, $E$, $I$, and $R$.
-    From a given state $x$, there are five possible kinds of jumps $x\mapsto{x'}$.
-    Accordingly, the set, $\Jumps$, of jump marks has five elements.
-    Each of these is of a different type:
-    $\lab{Trans}$ (transmission) is of birth type,
-    $\lab{Prog}$ (progression) is of migration type,
-    $\lab{Recov}$ (recovery) is of death type,
-    $\lab{Sample}$ (sampling) is of sample type,
-    and $\lab{Wane}$ (loss or waning of immunity) is of neutral type.
-    See \cref{sec:jump_types} for a description of these jump types.
-    Note that, in this formulation, when a sampling event occurs, the state does not change.
-    \label{fig:markov_state}
-  }
-\end{figure}
-
-\subsubsection{Structured populations, demes, and deme occupancy}
-\label{sec:demes}
-
-In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
-\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
-Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
-In particular, we suppose that there are a countable set of subpopulations that differ in their vital rates, but within each of which, individual lineages are statistically identical.
-We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
-
-For any $i\in\Demes$, we let $n_i(\Xr_t)$ denote the number of lineages present in deme $i$ at time $t$, \ie the \emph{occupancy} of deme $i$.
-Thus $n(\Xr_t)\in\Zp^{\Demes}$ is the vector of deme occupancies.
-
-\subsubsection{Jump marks}
-
-In the following, it will be useful to break the jumps into distinct categories.
-For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
-\begin{equation*}
-  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
-\end{equation*}
-In \cref{fig:markov_state}, we use the marks to distinguish events with distinct effects on genealogical relationships.
-Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.
-
-Let us define the \emph{jump mark} process, $\Ur_t$, to be the mark of the latest jump as of time $t$.
-As usual, we take the sample paths of $\Ur_t$ to be \cadlag.
-Observe that $(\Xr_t,\Ur_t)$ is a Markov process, though $\Ur_t$ is not.
-
-\subsection{Examples}
-
-\subsubsection{SEIRS model}
-
-\subsubsection{SIIR model}
-
-\subsubsection{Linear birth-death model}
-
-\subsubsection{Moran model and the Kingman coalescent}
-
-\subsection{History process}
-
-Consider the Markov process $(\Xr_t,\Ur_t)$.
-We define its \emph{history process}, $\Hr_t$, to be the restriction of the random function $s\mapsto(\Xr_s,\Ur_s)$ to the interval $[0,t]$.
-Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.
-Alternatively, one can think of $\Hr_t$ as consisting of the sequence
-$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
-In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
-The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
-\begin{equation*}
-  \label{eq:Hdens}
-  \begin{aligned}
-    \pi^{\mathrm{H}}(\dd{\H_t})
-    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
-    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
-    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
-  \end{aligned}
-\end{equation*}
-where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.
-
-\subsection{Binomial ratio}
-\label{sec:binomial_ratio}
-
-For $n,r,\ell,s\in{\Zp^\Demes}$, define the \emph{binomial ratio}
-\begin{equation*}
-  \BinRatio{n}{\ell}{r}{s}\colonequals
-  \begin{cases}
-    \frac{\displaystyle\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}%
-         {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
-         & \text{if}\ \forall i\ n_i\ge{\Set{\ell_i,r_i}}\ge{s_i}\ge{0},\\
-         0, & \text{otherwise}.
-  \end{cases}
-\end{equation*}
-Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
-Moreover, in consequence of the Chu-Vandermonde identity, we have
-\begin{equation*}
-  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
-\end{equation*}
-whenever $n_i\ge{\Set{\ell_i,r_i}}\ge{0}$ for all $i$.
-
-\section{The induced genealogy process}
-
-\subsection{Genealogies}
-
-For our purposes, a \emph{genealogy}, $G_{T}$, is a labeled, time-calibrated tree.
-It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
-There are three distinct kinds of nodes:
-\begin{inparaenum}[(i)]
-\item \emph{tip nodes}, which represent labeled living lineages;
-\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
-\item \emph{sample nodes}, which represent labeled samples.
-\end{inparaenum}
-Each node $a$ is associated with a specific time, $t(a)$.
-In particular, if $a$ is a tip node in $G_T$, then $t(a)=T$;
-if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
-Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
-We let $\event{G}{T}$ denote the set of all node-times of the genealogy $G_T$.
-Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
-Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.
-
-Formally, we define a genealogy, $G_T$, to be an ordered pair $(y,z)$.
-The $z$ element specifies the topological structure, as follows.
-Let $\leaves{G_T}$ be the set of all tip and sample nodes in $G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{G_T}$.
-The partition \emph{fineness} defines a partial order on $\func{P}$.
-Specifically, for $p,p'\in{\func{P}}$, we say $p\preceq{p'}$ if and only if for every $c'\in{p'}$ there is a unique $c\in{p}$ such that $c\supseteq{c'}$.
-The topological structure of $G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
-
-The $y$ element of $G_T$ specifies the coloring of branches and location of nodes.
-In particular, for $a\in\leaves{G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
-To make this well-defined, let $d_{t}(a)=\eth$ for $t\in{[t(a),T]}$;
-we call $\eth$ the \emph{underdeme} and define $\Demesplus\coloneq{\Demes\cup\{\eth\}}$.
-Thus $d:[0,T]\times\leaves{G_T}\to\Demesplus$ is a well-defined \cadlag\ function.
-
-Since not every node along a lineage corresponds to a change of deme, we also count the number of nodes along each lineage.
-Let $m:[0,T]\times\leaves{G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
-For $t\in{[t(a),T]}$, let $m_{t}(a)=m_{t(a)}(a)$.
-Clearly, $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
-
-Finally, we define the map $y:[0,T]\times\leaves{G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
-The two \cadlag\ functions $y$ and $z$ specify the genealogy $G_T$ completely and uniquely.
-When the genealogy is not clear from context, we write $y^{G}$ and $z^{G}$ to refer to the respective components of genealogy $G$.
-
-Clearly, the topological structure of a tree imposes constraints on the permissible functions $y$ that describe the painting and node placement along it.
-In particular, $y$ must respect the topological structure of $G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
-In other words, if $a,b\in{c}\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
-Let $\Yspace^z$ denote the space of all functions that respect the topological structure described by $z$.
-Let $\Yspace^\z_t$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.
-
-\subsection{Jump types}
-\label{sec:jump_types}
-
-\begin{figure}
 \begin{center}
@@
       \usetikzlibrary{arrows.meta,positioning,decorations}
-        \colorlet{darkblue}{blue!50!black}
-        \colorlet{maize}{yellow!90!black}
-        \colorlet{grey}{black!30!white}
-        \colorlet{darkgreen}{green!40!black}
-        \colorlet{lightblue}{white!30!blue}
       \tikzstyle{label}=[font=\Large]
       \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{deme1}=[color=grey, line width=3pt, >={Circle[width=5pt,length=5pt]}]
-        \tikzstyle{deme2}=[color=darkblue, line width=3pt, >={Circle[width=5pt,length=5pt]}]
-        \tikzstyle{deme3}=[color=maize, line width=3pt, >={Circle[width=5pt,length=5pt]}]
+      \tikzstyle{deme1}=[color=darkblue, line width=3pt]
+      \tikzstyle{deme1H}=[color=darkblue, line width=3pt, densely dotted]
+      \tikzstyle{deme2}=[color=maize, line width=3pt]
+      \tikzstyle{deme2H}=[color=maize, line width=3pt, densely dotted]
       \tikzstyle{axis}=[color=black, thick, >=stealth]
@@
         \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
-          \draw[deme3] (mid |- temp) -- (mid |- temp2) -- (right |- temp2);
+        \draw[deme1] (mid |- temp) -- (mid |- temp2) -- (right |- temp2);
         \draw[deme1] (left |- temp) -- (right |- temp);
@@
           \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
+          \draw[deme2] (left |- temp) -- (right |- temp);
         }
@@
         \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
-          \draw[deme3,dashed] (left |- temp) -- (mid |- temp);
+        \draw[deme2H] (left |- temp) -- (mid |- temp);
         \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
@@
         \draw[deme2] (left |- temp) -- (mid |- temp);
-          \draw[deme3] (mid |- temp) -- (right |- temp);
+        \draw[deme1] (mid |- temp) -- (right |- temp);
         \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
@@
           \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
+          \draw[deme1] (left |- temp) -- (right |- temp);
         }
@@
           \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
+          \draw[deme2] (left |- temp) -- (right |- temp);
         }
         \coordinate (temp) at ($(origin)+(0,-2*\ysp)$);
-          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
+        \node[text=royalblue,font=\Huge] at (mid |- temp) {$\bullet$};
         \coordinate (temp) at ($(origin)+(0,-6*\ysp)$);
-          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
+        \node[text=royalblue,font=\Huge] at (mid |- temp) {$\bullet$};
         \draw[timepoint] (mid) -- (mid |- bleft);
@@
         \draw[deme1] (left |- temp) -- (right |- temp);
-          \draw[deme3,dashed] (left |- temp1) -- (mid |- temp1);
+        \draw[deme2H] (left |- temp1) -- (mid |- temp1);
         \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
@@
         \coordinate (temp) at ($(origin)+(0,-5*\ysp)$);
-          \draw[deme3] (left |- temp) -- (mid |- temp);
-          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
+        \draw[deme1] (left |- temp) -- (mid |- temp);
+        \node[text=royalblue,font=\Huge] at (mid |- temp) {$\bullet$};
         \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
@@
         \coordinate (temp) at ($(origin)+(0,-3*\ysp)$);
-          \draw[deme3] (left |- temp) -- (mid |- temp);
-          \draw[deme2] (mid |- temp) -- (right |- temp);
+        \draw[deme2] (left |- temp) -- (mid |- temp);
+        \draw[deme1] (mid |- temp) -- (right |- temp);
         \coordinate (temp1) at ($(temp)+(0,-\ysp)$);
-          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
+        \draw[deme1] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
         \coordinate (temp1) at ($(temp1)+(0,-\ysp)$);
-          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
+        \draw[deme1] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
         \coordinate (temp1) at ($(temp1)+(0,-\ysp)$);
-          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
+        \draw[deme1] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
         \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
@@
         }
-          \foreach \y in {1} {
+        \foreach \y in {1,8} {
           \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
           \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
         }
@@
   Jump types differ by their effects on the genealogy.
+  This can be seen by examining the local structure of the genealogy in a neighborhood of a jump event.
   \textbf{(A)} A birth-type jump results in the branching of one or more child lineages from the parent.
   There can be only one parent, though the child lineages may be in demes different from those of their parent.
-    Here, a parent of the grey deme sires one child lineage in each of the blue and yellow demes.
+  Here, a parent of the blue deme sires one child lineage in each of the blue and yellow demes.
   The production of an event is an integer vector, with one entry for each deme.
-    The production of this event is therefore $r=(r_{\lab{grey}},r_{\lab{blue}},r_{\lab{yellow}})=(1,1,1)$.
+  The production of this event is therefore $r=(r_{\lab{blue}},r_{\lab{yellow}})=(2,1)$.
   \textbf{(B)} A death-type event causes the extinction of a lineage.
   Since internal node without children are recursively removed, the affected branch disappears.
-    The production of this event is $r=(0,0,0)$.
+  The production of this event is $r=(0,0)$.
   \textbf{(C)} A migration-type event results in the movement of one or more lineages from one deme to another.
-    Here, one lineage moves from the blue to the yellow deme.
-    The production of this event is $r=(0,0,1)$, \ie the production is 0 for the grey and blue demes, and 1 for the yellow.
+  Here, one lineage moves from the yellow to the blue deme.
+  The production of this event is $r=(1,0)$, \ie the production is 1 for the blue deme and 0 for the yellow.
   \textbf{(D)} In a sample-type event, one or more sample nodes (blue circles) are inserted.
-    Here, there are two samples, one in each of the grey and yellow demes.
-    Accordingly, its production is $r=(1,0,1)$.
-    \textbf{(E)} A neutral-type event has no effect on the genealogy and zero production in all demes: $r=(0,0,0)$.
+  Here, there are two samples, one in each of the blue and yellow demes.
+  Accordingly, its production is $r=(1,1)$.
+  \textbf{(E)} A neutral-type event has no effect on the genealogy and zero production in all demes: $r=(0,0)$.
   \textbf{(F)} The theory presented here allows for compound events.
-    As an example, here a birth/death-type event occurs, wherein one yellow lineage is extinguished and a grey lineage simultaneously sires a grey child.
-    The production of this event is $r=(2,0,0)$.
-    \textbf{(G)} Here, a compound sample/death-type event occurs.
-    A yellow lineage is sampled and simultaneously extinguished.
-    Note that the lineage is not recursively removed, since sample nodes are never removed.
-    \textbf{(H)} A compound birth/migration-type event with yellow parent and production $r=c(0,4,0)$.
-    \label{fig:jump_types}
+  As an example, here a birth/death-type event occurs, wherein one yellow lineage is extinguished and a blue lineage simultaneously sires a blue child.
+  The production of this event is $r=(2,0)$.
+  \textbf{(G)} Here, a compound sample/death-type event with $r=(0,0)$ occurs.
+  A blue lineage is sampled and simultaneously extinguished.
+  Note that recursive removal does not occur, since sample nodes are never removed.
+  \textbf{(H)} A compound birth/migration-type event with $r=(4,0)$.
+  \label{fig:event_types}
 }
-\end{figure}
-
-We will show how a given population process naturally induces a process in the space of genealogies.
-Because different kinds of jumps have different effects on the genealogy, we need a way of distinguishing them.
-In particular, there are five distinct \emph{pure types} of jumps:
-\begin{compactenum}[(a)]
-\item \emph{Birth-type jumps} result in the branching of one or more new lineages, each from some existing lineages.
-  Examples of birth-type events include transmission events, speciations, and actual births.
-  For simplicity, we assume that all new lineages arising from a birth event share the same parent.
-\item \emph{Death-type jumps} result in the extinction of one or more lineages.
-  Examples include recovery from infection, death of a host, and species extinctions.
-  We allow for the possibility that multiple lineages die simultaneously.
-\item \emph{Migration-type jumps} result in the movement of a lineage from one deme to another.
-  Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
-  We permit multiple movements to occur simultaneously.
-\item \emph{Sample-type jumps} result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
-  We allow for the possibility that multiple samples are collected in a single sample event.
-  In this case, we assume that, in each sample-type event, each extant lineage is sampled at most once.
-\item \emph{Neutral-type jumps} result in no change to any of the lineages.
-\end{compactenum}
-\Cref{fig:markov_state} depicts an example with all five of the pure types.
-It is not necessary that a jump be of a pure type.
-Compound jumps partake of more than one type.
-For example, a sample/death-type event, in which a lineage is simultaneously sampled and removed, has been proposed \citep{Leventhal2014}, as have birth/death events in which one lineage reproduces at the same moment that another dies (\eg the Moran process \citep{Moran1958}).
-The theory presented here places few restrictions on the complexity of the events that can occur.
-
-%% Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents and demes, there is implicitly a deterministic indicator function $Q_u$, for $u\in\Jumps$, (described below) that captures these properties.
-
-\subsection{Genealogy process}
-
-We now show how a given population process induces a stochastic process, $\Gen_t$, on the space of genealogies.
-In the case of unstructured population processes (\ie those having a single deme), \citet{King2022} gave a construction.
-Although we now treat a more general case, the ideas are much the same.
-Readers wishing more detail on the construction here should consult the earlier paper.
-
-At each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event:
-\begin{compactenum}[(a)]
-\item
-  A birth-type event at time $t$ results in the creation of one new internal node, call it $b$.
-  A tip node, $a$, of the appropriate deme is chosen with uniform probability from among those present and $b$ is inserted so that its ancestor is that of $a$, while $a$ takes $b$ as its ancestor.
-  One new tip node, of the appropriate deme, is created for each of the children, all of which take $b$ as their immediate ancestor.
-\item
-  In a death-type event, one or more tip nodes of the appropriate demes are selected with uniform probability from among those present.
-  These are deleted.
-  Next, branch nodes without children are recursively removed.
-  Sample nodes are never removed.
-\item
-  In a migration-type event, the appropriate number of migrating lineages are selected at random, with uniform probability according to deme, from among those present.
-  For each selected lineage, one new branch node is inserted between the selected tip node and its ancestor.
-  The color of the descendant branch changes accordingly.
-\item
-  At a sample-type event, the appropriate number of sampled lineages are selected at random from among the tip nodes, with uniform probability according to deme.
-  One new sample node is introduced for each selected lineage:
-  each is inserted between a selected tip nodes and its ancestor.
-\item
-  At a neutral-type event, no change is made to the genealogy.
-\item
-  Finally, events of compound type are accommodated by combining the foregoing rules.
-\end{compactenum}
-In each of these events, the new node or nodes that are introduced are situated at time $t$.
-
-\subsubsection{Emergent lineages and production}
-
-The lineages which descend from an inserted node are said to \emph{emerge} from the event.
-Thus, after a birth-type event, the emerging lineages include all the new offspring as well as the parent.
-Likewise, at pure migration- or sample-type events, each migrating or sampled lineage emerges from the event.
-At pure death-type events, no lineages emerge.
-In general, at an event of mark $u$, there are $r^u_i$ emergent lineages in deme $i$.
-We require that $r^u_i$ be a constant, for each $u$ and $i$.
-Since one is free to expand the set of jump-marks $\Jumps$ as needed, this is not an important restriction on the models that the theory can accommodate.
-Thus there is a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
-We say $r^u=(r^u_i)_{i\in\Demes}$ is the \emph{production} of an event of mark $u$.
-Note that the lineages that die as a result of an event do not count in the production but that all parent lineages, if they survive the event, are counted in the production.
-
-\subsubsection{Conditional independence and exchangeability}
-
-Application of these rules at each jump of $\Xr_t$ constructs a chain of genealogies $\Grh_k$.
-In particular, at each jump-time $\Trh_k$, the genealogy $\Grh_{k-1}$ is modified according to the jump-mark $\Urh_k$ to yield $\Grh_k$.
-We view $\Grh_k$ as the embedded chain of the continuous-time genealogy process $\Gen_t$.
-It is very important to note that, conditional on $\Urh_k$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of the choices at any other times and independent of $\Xrh_k$ for all $k$.
-Moreover, by construction, any lineage within a deme is as likely as any other lineage in that deme to be selected as a parent or for death, sampling, or migration.
-We refer to this property as the \emph{exchangeability} of lineages within a deme.
-
-\subsection{Pruned and obscured genealogies}
-
-\begin{figure}
-  \caption{
-    Unpruned, pruned, and obscured enealogies.
-  }
-\end{figure}
-
-The process just described yields a genealogy that relates all extant members of the population, and all samples.
-Moreover, it details each lineage's complete history of movement through the various demes.
-The data we ultimately wish to analyze will be based only on samples, however;
-nor, in general, will the histories of deme occupancy be observable.
-A generative model must account for this loss of information.
-We therefore now describe how genealogies are \emph{pruned} to yield sample-only genealogies and then \emph{obscured} via the erasure of color from their branches.
-
-\subsubsection{Pruned genealogy}
-
-Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Pr_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every childless internal node.
-In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and possibly some of the interior nodes of the genealogy.
-Observe that a pruned genealogy is a colored genealogy:
-it retains information about where among the demes each of its lineages was through time.
-Note also that a pruned genealogy $\Pr_t$ is characterized by functions $y^{\Pr_t}$ and $z^{\Pr_t}$ just as an un-pruned genealogy is.
-
-\subsubsection{Lineage count and saturation}
-\label{sec:ells}
-
-In the following, we will find that we need to count the deme-specific numbers of lineages present in a given pruned genealogy at a given time.
-Accordingly, suppose $\Pr_T=(y,z)$ is a pruned genealogy and suppose $t\in[0,T]$.
-Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell=(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
-Clearly, $\ell$ depends only on the values, $(y_t,\z_t)$, of $y$ and $z$ at time $t$.
-Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(y,z)$ is a pruned genealogy, $\ell(y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
-We refer to $\ell$ as the \emph{lineage-count} function.
-
-We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
-In particular, given a node time $t$ in a pruned genealogy $\Pr_T=(y,z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
-Like the lineage-count function, $s$ depends only on the local structure of $\Pr_T$.
-However, $s$ depends not only on $(\y_t,\z_t)$, but also on $(\yt_t,\zt_t)$.
-Thus, we can define the \emph{saturation} function such that, whenever $\Pr_T=(\y,\z)$ is a pruned genealogy, $s(\yt_t,\y_t,\zt_t,\z_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
-
-\begin{figure}
-  \caption{
-    Figure showing lineage count and saturation.
-    \label{fig:ells}
-  }
-\end{figure}
-
-\subsubsection{Compatibility}
-\label{sec:compatibility}
-
-Suppose $\P_T$ is a pruned genealogy and $t\in\event{P}{T}$.
-The local structure of $\P_T$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
-For example, if the event in $\P_T$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
-Similarly, if the node in $\P_T$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
-To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q_u(y,y',z,z')=1$ if and only if there is some pruned genealogy $\P_T$ and $t\in{[0,T]}$,
-such that $\yt^{\P_T}_t=y$, $y^{\P_T}_t=y'$, $\zt^{\P_T}_t=z$, $z^{\P_T}_t=z'$,
-and $\P_T$ is compatible with an event of type $u$ at time $t$;
-$Q=0$ otherwise.
-We refer to $Q$ as the \emph{compatibility indicator}.
-
-\subsubsection{Obscured genealogy}
-
-The \emph{obscured genealogy} is obtained by discarding all information about demes and events not visible from the topology of the tree alone.
-In particular, if $\Pr_T=(y,z)$ is a pruned genealogy, we write $\obs(\Pr_T)=z$ to denote the obscured genealogy.
-
-\section{Results}
-
-\subsection{Likelihood for pruned genealogies}
-
-Our first result will be an expression for the likelihood of a given pruned genealogy $\P^*_T$ given the history $\Hr_T$ of the population process to time $T$.
-
-\begin{thm}\label{thm:pruned_lik}
-  Let $\P^*_T=(y^*,z^*)$  be a given pruned genealogy and
-  \begin{equation}\label{eq:phidef}
-    \phi_u(x,y,y',z,z')\coloneq\BinRatio{n(x)}{\ell(y',z')}{r^{u}}{s(y,y',z,z')}\,Q_{u}(y,y',z,z'),
-  \end{equation}
-  where $n$ is the deme occupancy (\cref{sec:demes}), $\ell$ and $s$ are the lineage-count and saturation functions, respectively (\cref{sec:ells}), $Q$ is the compatibility indicator (\cref{sec:compatibility}), and the binomial ratio is as defined in \cref{sec:binomial_ratio}.
-  Then
-  \begin{equation*}
-    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,y^*_t,\zt^*_t,z^*_t)}
-    \:\times\:\Indicator{\event{\H}{T}\supseteq{\event{\P}{T}^*}}.
-  \end{equation*}
-\end{thm}
-\begin{proof}
-  If $\event{\Hr}{T}\nsupseteq\event{\P}{T}^*$, then $\Hr_T$ and $\P^*_T$ are incompatible and $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
-  Similarly, if any event of $\Hr_T$ is incompatible with the local structure of $\P^*_T$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
-  Let us therefore suppose that neither of these conditions hold.
-  Conditional on $\Hr_T=\H_T$, at each time $t\in\event{\H}{T}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
-  In $\P^*_T$, at time $t$, there are $\ell_i=\ell_i(y^*_t,z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,y^*_t,\zt^*_t,z^*_t)$ are emergent.
-  By assumption, at each genealogical event, lineages within a deme are exchangeable:
-  each has an identical probability of being involved.
-  This exchangeability implies that each lineage present in a deme at time $t$ was equally likely to have been one of the emergent lineages.
-  In particular, at time $t$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ red balls and $n_i-r_i$ black balls, exactly $s_i$ of the drawn balls are red, namely
-  \begin{equation*}
-    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
-  \end{equation*}
-  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $\P^*_t$.
-  Moreover, since, again conditional on $\Hr_T$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
-  \begin{equation*}
-    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\BinRatio{n(\X_t)}{\ell(y^*_t,z^*_t)}{r^{\U_t}}{s(\yt^*_t,y^*_t,\zt^*_t,z^*_t)}}.
-  \end{equation*}
-  Returning to the possibility that $\H_T$ is incompatible with $\P^*_T$, since $\Prob{\Pr_T=\P^*_T}=0$ if either any $Q_u=0$ or any event of $\P^*_T$ is not an event of $\H_T$, we obtain the result.
-\end{proof}
-
-%% \begin{equation*}
-%%   \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
-%% \end{equation*}
-
-Next, we show how the likelihood of a pruned genealogies, unconditional on the history, can be computed.
-For this, we use the filter equation technology developed in \cref{sec:filter_eqns}.
-In particular, the following theorem follows immediately from \cref{prop:fullfilt}.
-
-\begin{thm}
-  Suppose that $\P^*_T=(y^*,z^*)$ is a pruned genealogy, for $T>0$.
-  If $w=w(t,x)$ satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
-  \begin{equation}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)=
-      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}
-      -\sum_u{\int{w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}}}\\
-      &\quad+\sum_{e\in\event{\P}{T}^*}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}-w(t,x)\right\}},
-    \end{aligned}
-  \end{equation}
-  where $\phi$ is defined in \cref{eq:phidef},
-  then the likelihood of $\P^*_T$ is
-  \begin{equation*}
-    \lik(\P^*_T)=\int{w(T,x)\,\dd{x}}.
-  \end{equation*}
-\end{thm}
-
-\subsection{Likelihood for obscured genealogies}
-
-Our next result concerns the likelihood of a given obscured genealogy conditional on the history.
-
-\begin{thm}\label{thm:obsc_lik}
-  Let $z^*=\obs(\P^*_T)$ be a given obscured genealogy.
-  Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{z^*}}{q(x,y)}=1$.
-  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,z^*_t)=1$
-  and that $q(x,y)>0$ whenever $\Prob{\Xr_0=x,y^{\Pr_T}_0=y}>0$.
-  Then there is a Markov jump process $\yr_t$ on $\Yspace_{z^*}$ such that
-  \begin{equation*}
-    \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
-    \Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}}
-    \times\Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*},
-  \end{equation*}
-  where $\phi$ is defined in \cref{eq:phidef} and
-  the expectation is taken over the sample paths of $\yr_t$, $t\in[0,T]$.
-\end{thm}
-\begin{proof}
-  First, observe that, since $\obs$ is a deterministic operator,
-  \begin{equation}\label{eq:IS1}
-    \CondProb{\zr_T=z^*}{\Hr_T}=\CondExpect{\Indicator{\obs(\Pr_T)=z^*}}{\Hr_T}.
-  \end{equation}
-  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
-  we will propose pruned genealogies compatible with $z^*_t$ as sample paths from a Markov process on $\Yspace_{z^*}$ and
-  evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
-  Conditional on $\Hr_T$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$, on $\Yspace_{z^*}$ such that
-  \begin{equation*}
-    \begin{gathered}
-      \Prob{\yrh_0}=q(\yrh_0),
-      \qquad
-      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_T=\H_T}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
-    \end{gathered}
-  \end{equation*}
-  The required Markov process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
-  This construction of $\yr_t$ obviously guarantees that $\event{\H}{t}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$.
-
-  Let us define $\Pr(\yr)=(\yr,z^*)$.
-  Then, by construction, for every $\yr$, $\obs(\Pr(\yr))=z^*$ and, conversely, for every pruned genealogy $\Pr_T$ satisfying $\obs(\Pr_T)=z^*$, $y^{\Pr_T}$ is a sample path of this process having nonzero probability.
-  We therefore have that
-  \begin{equation*}
-    \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
-    \Expect{\frac{\CondProb{\Pr_T=P(\yr)}{\Hr_T=\H_T}}{\pi(\yr\vert\H_T)}}
-  \end{equation*}
-  the expectation being taken with respect to the random process $\yr_t$.
-  Here, by definition,
-  \begin{equation*}
-    \pi(\yr\vert\H_T)=\prod_{t\in\event{\H}{T}}{\pi_{\U_{t}}(t,\Xt_{t},\X_{t},\yrt_{t},\yr_{t})}.
-  \end{equation*}
-  The result then follows from \cref{thm:pruned_lik}.
-\end{proof}
-
-The final result shows how to compute the likelihood of an obscured genealogy.
-It is an immediate consequence of \cref{thm:obsc_lik,prop:fullfilt}.
-
-\begin{thm}
-  Let $z^*$, be a given obscured genealogy.
-  Let $\pi$ and $\yr_t$ be as in \cref{thm:obsc_lik}.
-  For $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $u\in\Jumps$, define
-  \begin{equation*}
-    \begin{gathered}
-      \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y',\zt^*_t,z^*_t)}{\pi_u(t,x,x',y,y')},\\[2ex]
-      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),
-    \end{gathered}
-  \end{equation*}
-  where $\phi$ is defined in \cref{eq:phidef}.
-  Suppose $w=w(t,x,y)$ satisfies the filter equation
-  \begin{equation}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x,y)=
-      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}}\\
-      &-\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
-      &+\sum_{e\in\event{\z}{}^*}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y,y')\,\psi_u(t,x',x,y',y)\,\dd{x'}}}-w(t,x,y)\right\}},
-    \end{aligned}
-  \end{equation}
-  with the initial condition $w(0,x,y)=p_0(x)\,q(x,y)$.
-  Then the likelihood of $V^*_T$ is
-  \begin{equation*}
-    \lik(V^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
-  \end{equation*}
-\end{thm}
-
-\section{Examples}
-
-\subsection{SEIRS}
-
-Jumps: $\Jumps=\Set{\lab{Inf},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Birth},\lab{Death_S},\lab{Death_E},\lab{Death_I},\lab{Death_R},\lab{Sample}}$.
-
-Demes: $\Demes=\Set{\lab{E},\lab{I}}$.
-
-Jump rates:
-\begin{itemize}
-\item $\alpha_{\lab{Inf}}(t,x,x')=\beta(t)\,\frac{x^{\lab{S}}x^{\lab{I}}}{N(t)}\,\Indicator{x'=x+(-1,1,0,0)}$
-\item $\alpha_{\lab{Prog}}(x,x')=\rho\,x^{\lab{E}}\,\Indicator{x'=x+(0,-1,1,0)}$
-\item $\alpha_{\lab{Recov}}(x,x')=\gamma\,x^{\lab{I}}\,\Indicator{x'=x+(0,0,-1,1)}$
-\item $\alpha_{\lab{Wane}}(x,x')=\upsilon\,x^{\lab{R}}\,\Indicator{x'=x+(1,0,0,-1)}$
-\item $\alpha_{\lab{Sample}}(t,x,x')=\psi\,x^I\,\Indicator{x'=x}$
-\item $\alpha_{\lab{Birth}}(t,x,x')=B(t)\,\Indicator{x'=x+(1,0,0,0)}$
-\item $\alpha_{\lab{Death}_k}(x,x')=\mu\,x^k\,\Indicator{x'^j=x^j-\delta_{jk}}$, $k\in\Set{\lab{S},\lab{E},\lab{I},\lab{R}}$
-\end{itemize}
-
-%% Nonlinear filtering equation:
-%% \begin{equation}
-%%   \begin{aligned}
-%%     \frac{\partial{w}}{\partial{t}}(t,x,y,z)=&B(t)\,\left[\sum_{y'z'}w\big(t,x-(1,0,0,0),y',z'\big)-w\big(t,x,y,z\big)\right]\\
-%%     &\qquad+\sum_{k\in\{\lab{S},\lab{E},\lab{I},\lab{R}\}}\mu\,(x^k+1)\,w(t,x+\delta_{k})-\mu\,x^k\,w(t,x)\\
-%%     &\qquad+\upsilon\,x^{\lab{R}}\,w(t,x-(1,0,0,-1))\\
-%%   \end{aligned}
-%% \end{equation}
-
-\section{Discussion}
-
-\bibliographystyle{preprint}
-\bibliography{phylopomp}
-
-\appendix
-\setcounter{equation}{0}
-\titleformat{\section}[hang]{\large\bfseries}{Appendix \periodafter\thesection}{2ex}{\periodafter}{}
-\renewcommand{\theequation}{\thesection\arabic{equation}}
-
-\section{Filter equations}
-\label{sec:filter_eqns}
-
-Explicit expressions for the quantities that arise in this paper are not always readily available.
-Here, we develop tools for manipulating complex expressions that are otherwise cumbersome.
-
-\begin{defn}
-  Suppose $\Xr_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
-  \begin{equation}
-    \label{eq:kfe2}
-    \frac{\partial{u}}{\partial{t}}(t,x)
-    =\int{u(t,x')\,\beta(t,x',x)\,\dd{x'}}
-    -\int{u(t,x)\,\beta(t,x,x')\,\dd{x'}}.
-  \end{equation}
-  Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)$ are given, real-valued, measurable functions.
-  We say that the equation
-  \begin{equation}
-    \label{eq:filtereq}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)
-      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
-      -\lambda(t,x)\,w(t,x).\\
-    \end{aligned}
-  \end{equation}
-  is the \emph{filter equation} with \emph{driver} $\Xr_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, and \emph{decay} $\lambda$.
-\end{defn}
-
-\begin{remark}
-  Trivially, a Kolmogorov forward equation is itself a filter equation with boost $1$ and decay $0$.
-\end{remark}
-
-The following results show how filter equations allow one to integrate over random histories.
-
-\begin{lemma}
-  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable.
-  If $\Vr_t$ is an $\Rp$-valued random process satisfying
-  \begin{equation*}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)},
-  \end{equation*}
-  then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued deterministic process on $\Xspace$ with density $w(t,x)$.
-  Moreover, $w$ satisfies the filter equation
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{t}}=\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}
-  \end{equation*}
-  and $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
-\end{lemma}
-\begin{proof}
-  The expectation can be broken into three terms, according to whether $\H_t$ has zero, one, or more than one event in $\halfclosed{t-\Delta,t}$.
-  Accordingly, as $\Delta\downarrow{0}$,
-  \begin{equation*}
-    \begin{aligned}
-      w(t,x)=&\left(1-\Delta\,\int{\alpha(t-\Delta,x,x')\,\dd{x'}}\right)\,w(t-\Delta,x)\\
-      &\qquad+\Delta\,\int{\alpha(t-\Delta,x',x)\,B(t-\Delta,x',x)\,w(t-\Delta,x')\,\dd{x'}}+o(\Delta).
-    \end{aligned}
-  \end{equation*}
-  Taking the limit as $\Delta\downarrow{0}$, we obtain
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{t}}(t,x)
-    =\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
-    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
-  \end{equation*}
-\end{proof}
-
-\begin{lemma}
-  Suppose $\mathcal{M}\subset{\Rp}$ is finite and let $\mathcal{M}_t\coloneq\mathcal{M}\cap{[0,t]}$.
-  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable and
-  $\Vr_t$ is an $\Rp$-valued process satisfying
-  \begin{equation*}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\Indicator{\event{\H}{t}\supseteq\mathcal{M}_t}\,\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)}.
-  \end{equation*}
-  Then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued process on $\Rp^{\vert\mathcal{M}_t\vert}\times\Xspace$ with density $w(t,x)$.
-  Furthermore, $w(t,x)$ satisfies
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)&=
-      \int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
-      &t\notin{\mathcal{M}},\\
-      w(t,x)&=\int{\wt(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
-      &t\in{\mathcal{M}}
-    \end{aligned}
-  \end{equation*}
-  and $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
-\end{lemma}
-
-Filter equations afford a convenient means of computing expectations and likelihoods for pure jump processes.
-This is facilitated by the following Lemma, the statement of which relies on a non-standard Dirac delta function.
-Let $\delta(v,v')$ be the right-sided Dirac delta function satisfying $\delta(v,v')=0$ for $v\ne{v'}$ and $\int_a^b{f(v)\,\delta(v,v')\,\dd{v}}=f(v')\,\Indicator{v'\in\halfopen{a,b}}$ for all \cadlag\ $f$.
-
-\begin{lemma}
-  \label{lemma:filter}
-  The filter equation \eqref{eq:filtereq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satsfies the KFE
-  \begin{equation}
-    \label{eq:filterlemma}
-    \begin{aligned}
-      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
-      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right],
-    \end{aligned}
-  \end{equation}
-  on the space $\Xspace\times\Rp$.
-\end{lemma}
-\begin{proof}
-  \ELI{[Conditions for differentiating under the integral in the proof?
-      Need $[v^2 \lambda u(t,x,v)]_{v=0}^{\infty} = 0$.
-      That's okay at $v=0$, and a minor constraint on $u$.]}
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
-      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
-      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
-      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
-    \end{aligned}
-  \end{equation*}
-  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
-      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
-      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
-      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
-    \end{aligned}
-  \end{equation*}
-  which is simplified to obtain \cref{eq:filtereq}.
-\end{proof}
-
-\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(\Xr_t,\Vr_t)$.
-In particular, $\Xr_t$ is the driver with KFE \eqref{eq:kfe2}.
-The $\Vr_t$ is \emph{directed} by $\Xr_t$ in the sense that $\Vr$ has jumps wherever $\Xr$ does:
-when $\Xr$ jumps at time $t$ from $x$ to $x'$, $\Vr$ jumps by the multiplicative factor $B(t,x,x')\ge{0}$.
-Between jumps, $\Vr_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
-If we view $\Vr_t$ as a weight, then \cref{lemma:filter} tells us how the $\Vr_t$-weighted average of $\Xr_t$ evolves in time:
-this average is simply $\int\!{w(t,x)\,\dd{x}}$.
-
-%% Filter equations allow us to pass easily between equivalent representations of a process.
-%% For example, an equivalent way of representing $\Xr_t$ is in terms of its embedded chain and event times.
-%% Let $\Xrh_k$ be the embedded chain of $\Xr_t$ and let $\Trh_k$ be the point process of its event times.
-%% It is elementary that
-%% \fontsize{10pt}{12pt}\selectfont
-%% \begin{equation*}
-%%   \begin{gathered}
-%%     \Prob{\Xrh_k=\Xh_k\;\Big\vert\;\Xrh_{k-1}=\Xh_{k-1},\Trh_k=\Th_k}=\frac{\alpha(\Th_k,\Xh_{k-1},\Xh_k)}{\int{\alpha(\Th_k,\Xh_{k-1},x')\dd{x'}}},\\
-%%     \Prob{\Trh_k>\Trh_{k-1}+t\;\Big\vert\;\Xrh_{k-1}=\Xh_{k-1},\Trh_{k-1}=\Th_{k-1}}=\exp{\left(-\int_{0}^{t}{\int{\alpha(\Th_{k-1}+s,\Xh_{k-1},x')\,\dd{x'}}\,\dd{s}}\right)}.
-%%   \end{gathered}
-%% \end{equation*}
-%% \normalfont
-%% Fixing $\nu>0$ and making the definitions,
-%% \begin{equation}\label[pluralequation]{eq:nudefs}
-%%   \begin{gathered}
-%%     A(t,x)=\int{\alpha(t,x,x')\,\dd{x'}}, \qquad
-%%     \pi(t,x,x')=\frac{\alpha(t,x,x')}{A(t,x)},\\
-%%     B(t,x,x')=\frac{A(t,x)}{\nu}, \qquad
-%%     \lambda(t,x)=A(t,x)-\nu,
-%%   \end{gathered}
-%% \end{equation}
-%% we can rewrite the KFE as
-%% \begin{equation}\label{eq:poissdriver}
-%%   \frac{\partial{w}}{\partial{t}}(t,x)=
-%%   \int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
-%%   -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
-%%   -\lambda(t,x)\,w(t,x).
-%% \end{equation}
-%% Here, $\nu$ is the intensity of a time-homogenous Poisson process.
-%% Note that $\pi$ is the probability kernel of the embedded chain $\Xrh$ and $A(t,x)$ is the intensity of the $\Trh_k$ process.
-%% We recognize this equation as the filter equation with boost $B$, decay $\lambda$, and driver generated by $\nu\,\pi(t,x,x')$.
-%% It corresponds to the following procedure for simulating $\Xr_t$:
-%% \begin{compactenum}[(a)]
-%% \item Simulate jump times according to the rate-$\nu$ Poisson process.
-%% \item Simulate the embedded chain $\Xrh_k$ using the kernel $\pi$.
-%% \item Weight the realization by the product of the $B$ factors.
-%%   Note that this makes the appropriate importance-sampling correction.
-%% \end{compactenum}
-
-\end{document}

### 2024-01-21 figs/example_models.tex
more figures

--- 2024-01-20 ms.Rnw
@@
-\documentclass[10pt,reqno,final]{amsart}
-\input{header}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta,positioning,decorations,calc,math}
-
-\title[Exact Phylodynamics]{Exact Phylodynamics via Structured Markov Genealogy Processes}
-\author[King]{Aaron~A.~King}
-\address{
-  A.~A.~King,
-  Department of Ecology \& Evolutionary~Biology,
-  Center for the Study of Complex~Systems, and
-  Department of Mathematics,
-  University of Michigan,
-  Ann~Arbor, MI~48109~USA\\
-  Santa~Fe~Institute,
-  1399 Hyde~Park~Road,
-  Santa~Fe, NM~87501~USA
-}
-\email{kingaa@umich.edu}
-\urladdr{\href{https://kinglab.eeb.lsa.umich.edu/}{https://kinglab.eeb.lsa.umich.edu/}}
-\author[Lin]{Qianying~Lin}
-\address{
-  Q.-Y.~Lin,
-  Theoretical Biology and Biophysics,
-  Los~Alamos National Laboratory,
-  Los~Alamos, NM~87545~USA
-}
-\author[Ionides]{Edward~L.~Ionides}
-\address{
-  E.~L.~Ionides,
-  Department of Statistics
-  University of Michigan,
-  Ann~Arbor, MI~48109~USA
-}
-\date{\today}
-
-\hypersetup{pdftitle={Exact Phylodynamics via Structured Markov Genealogy Processes}}
-\hypersetup{pdfauthor={A.A. King, Q.-Y. Lin, E.L. Ionides}}
-\hypersetup{urlcolor=blue,citecolor=blue,linkcolor=blue,filecolor=blue}
-
-<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
-prefix <- "sgp"
-source("setup.R")
-@
-<<packages,include=FALSE,cache=FALSE>>=
-library(tidyverse)
-library(ggtree)
-library(pomp)
-library(cowplot)
-library(phylopomp)
-stopifnot(getRversion() >= "4.3")
-stopifnot(packageVersion("pomp")>="5.1")
-stopifnot(packageVersion("phylopomp")>="0.9.2")
-theme_set(theme_bw(base_family="serif"))
-set.seed(1159254136)
-@
-
-\begin{document}
-
-\begin{abstract}
-  We derive an algorithm for the exact computation of the likelihood of an observed transmission tree.
-\end{abstract}
-
-\maketitle
-
-\section{Introduction}
-
-Problem of phylodynamics.
-Factorization of problem into two subproblems.
-
-Relation to previous work.
-Existing methods \citep{Volz2009a,Stadler2010}.
-Large-population, small sample-size approximations.
-
-Extension of previous results \citep{King2022}.
-Broader class of state-spaces.
-Accommodating discrete structure.
-
-Classes of Markov processes.
-Utility and flexibility of Markov assumptions.
-
-Population process induces Markov history and genealogy processes.
-Using these, we derive equations for the likelihood of a genealogy conditional on the history.
-We then integrate out the history to obtain nonlinear filtering equations, the solution of which yields the likelihood.
-These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likelihood.
-We demonstrate with several examples.
-
-In the following, we show a Markov population process of the kind that is a staple in epidemiology induces a Markov process on the space of genealogies.
-We then show how one can comput the likelihood of a given genealogy.
-
-\section{Mathematical preliminaries}
-
-\subsection{Notation}
-
-Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\Xr$), denotes a random element.
-We will be concerned with a variety of stochastic processes, in both discrete and continuous time.
-In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
-In the case of continuous-time processes, we will assume that sample paths are right-continuous with left limits (\ie \cadlag).
-We will frequently need to refer to the left-limit of such a process.
-Accordingly, if $\Zr_t$ is a \cadlag\ random process, we define
-\begin{equation*}
-  \Zrt_t\colonequals
-  \begin{cases}
-    \displaystyle\lim_{t'\,\uparrow\,{t}}\;\Zr_{t'}, & t>0,\\[2ex]
-    \Zr_0, & t=0.\\
-  \end{cases}
-\end{equation*}
-Note that $\Zrt_t$ is thus left-continuous with right limits.
-
-If $Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
-In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
-$\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
-If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its event times in the interval $[0,t]$.
-
-\subsection{Population process}
-
-Motivating examples: compartmental models.
-Wide variety of models.
-Linear chain trick.
-Migration, superspreading, competition between strains.
-
-\begin{figure}
 \begin{center}
@@
     \begin{tikzpicture}[scale=1]
-        \colorlet{darkblue}{blue!50!black}
       \tikzstyle{box}=[draw=black, text=black, fill=white, very thick, minimum size=3em]
-        \tikzstyle{ibox}=[draw=darkblue]
+      \tikzstyle{ibox}=[draw=darkgreen]
       \tikzstyle{deme}=[fill=black!20!white]
@@
       \tikzstyle{flow}=[draw=black, very thick, >=stealth]
-        \tikzstyle{modulate}=[draw=darkblue, >=Circle]
+      \tikzstyle{modulate}=[draw=darkgreen, >=Circle]
 
@@
   Demes are shaded.
+  Compartments containing infectious hosts are outlined in blue.
+  Curved green lines connect transmission rates with the compartments whose occupancies control their modulation;
+  each such connection gives rise to a nonlinearity in the model.
   \textbf{(A)} An SEIRS model.
@@
   If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
-    \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
-    \label{fig:examples}
+  \textbf{(B)} In this four-deme model, two distinct pathogen strains compete for susceptibles.
+  \textbf{(C)} A three-deme model for SARS-CoV-2 infection.
+  After an incubation period, individuals may develop asymptomatic infection ($\lab{I_A}$).
+  If they do not recover, symptomatically infected individuals ($\lab{I_S}$) can progress to hospitalization ($\lab{H}$) and death ($\lab{D}$).
+  \textbf{(D)} A three-deme model with heterogeneity in transmission behavior.
+  Contagious individuals move randomly between low-transmission ($\lab{I_L}$) and high-transmission ($\lab{I_H}$) behaviors.
+  \label{fig:example_models}
 }
-\end{figure}
-
-We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
-In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
-The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
-In particular, we suppose that
-\begin{equation}
-  \label{eq:ic}
-  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
-\end{equation}
-for all measurable sets $\mathcal{E}\subseteq\Xspace$.
-For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
-More precisely, the transition rates have the following properties:
-\begin{equation*}
-  \begin{gathered}
-    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
-  \end{gathered}
-\end{equation*}
-for all $t\in\Rp$ and $x,x'\in\Xspace$.
-Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
-Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
-We assume that $\Kr_t$ is a simple counting process so that
-\begin{equation*}
-  \begin{gathered}
-    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
-    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
-    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
-  \end{gathered}
-\end{equation*}
-We will further assume that $\Xr_t$ is non-explosive, \ie that $\Prob{\Kr_t<\infty}=1$ for all $t$.
-
-\subsubsection{Kolmogorov forward equation}
-
-The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
-\begin{equation}
-  \label{eq:kfe}
-  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
-\end{equation}
-and if, moreover, $w(0,x) = p_0(x)$,
-then $\int_{\mathcal{E}}\!{w(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
-\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.
-
-Another perspective on the Markov processes is to be had from its Markov state transition diagram (\cref{fig:markov_state}).
-
-\begin{figure}
-  \begin{center}
-    \resizebox{0.9\linewidth}{!}{
-      \begin{tikzpicture}[scale=1]
-        \usetikzlibrary{shapes,arrows,positioning}
-        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum height=8em]
-        \tikzstyle{trans}=[color=black, font=\Large, thick, >=stealth]
-        \node[state] (base) at (0,0) {$x=(S,E,I,R)$};
-        \node[state] (trans) at (165:8) {$x'=(S-1,E+1,I,R)$};
-        \node[state] (prog) at (205:7) {$x'=(S,E-1,I+1,R+1)$};
-        \node[state] (recov) at (305:5) {$x'=(S,E,I-1,R+1)$};
-        \node[state] (wane) at (350:8) {$x'=(S+1,E,I,R-1)$};
-        \node[state] (sample) at (18:7.5) {$x'=(S,E,I,R)$};
-        \draw[trans,->] (base) -- (trans) node[midway,above,sloped] {$\lab{Trans}$};
-        \draw[trans,->] (base) -- (prog) node[midway,above,sloped] {$\lab{Prog}$};
-        \draw[trans,->] (base) -- (recov) node[midway,above,sloped] {$\lab{Recov}$};
-        \draw[trans,->] (base) -- (wane) node[midway,above,sloped] {$\lab{Wane}$};
-        \draw[trans,->] (base) -- (sample) node[midway,above,sloped] {$\lab{Sample}$};
-        \node[font=\LARGE] (U) at (95:3.5) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
-      \end{tikzpicture}
-    }
-  \end{center}
-  \caption{
-    Markov state transition diagram for the SEIRS model depicted in \cref{fig:examples}A.
-    The state, $x$, is characterized by four numbers, $S$, $E$, $I$, and $R$.
-    From a given state $x$, there are five possible kinds of jumps $x\mapsto{x'}$.
-    Accordingly, the set, $\Jumps$, of jump marks has five elements.
-    Each of these is of a different type:
-    $\lab{Trans}$ (transmission) is of birth type,
-    $\lab{Prog}$ (progression) is of migration type,
-    $\lab{Recov}$ (recovery) is of death type,
-    $\lab{Sample}$ (sampling) is of sample type,
-    and $\lab{Wane}$ (loss or waning of immunity) is of neutral type.
-    See \cref{sec:jump_types} for a description of these jump types.
-    Note that, in this formulation, when a sampling event occurs, the state does not change.
-    \label{fig:markov_state}
-  }
-\end{figure}
-
-\subsubsection{Structured populations, demes, and deme occupancy}
-\label{sec:demes}
-
-In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
-\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
-Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
-In particular, we suppose that there are a countable set of subpopulations that differ in their vital rates, but within each of which, individual lineages are statistically identical.
-We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
-
-For any $i\in\Demes$, we let $n_i(\Xr_t)$ denote the number of lineages present in deme $i$ at time $t$, \ie the \emph{occupancy} of deme $i$.
-Thus $n(\Xr_t)\in\Zp^{\Demes}$ is the vector of deme occupancies.
-
-\subsubsection{Jump marks}
-
-In the following, it will be useful to break the jumps into distinct categories.
-For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
-\begin{equation*}
-  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
-\end{equation*}
-In \cref{fig:markov_state}, we use the marks to distinguish events with distinct effects on genealogical relationships.
-Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.
-
-Let us define the \emph{jump mark} process, $\Ur_t$, to be the mark of the latest jump as of time $t$.
-As usual, we take the sample paths of $\Ur_t$ to be \cadlag.
-Observe that $(\Xr_t,\Ur_t)$ is a Markov process, though $\Ur_t$ is not.
-
-\subsection{Examples}
-
-\subsubsection{SEIRS model}
-
-\subsubsection{SIIR model}
-
-\subsubsection{Linear birth-death model}
-
-\subsubsection{Moran model and the Kingman coalescent}
-
-\subsection{History process}
-
-Consider the Markov process $(\Xr_t,\Ur_t)$.
-We define its \emph{history process}, $\Hr_t$, to be the restriction of the random function $s\mapsto(\Xr_s,\Ur_s)$ to the interval $[0,t]$.
-Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.
-Alternatively, one can think of $\Hr_t$ as consisting of the sequence
-$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
-In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
-The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
-\begin{equation*}
-  \label{eq:Hdens}
-  \begin{aligned}
-    \pi^{\mathrm{H}}(\dd{\H_t})
-    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
-    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
-    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
-  \end{aligned}
-\end{equation*}
-where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.
-
-\subsection{Binomial ratio}
-\label{sec:binomial_ratio}
-
-For $n,r,\ell,s\in{\Zp^\Demes}$, define the \emph{binomial ratio}
-\begin{equation*}
-  \BinRatio{n}{\ell}{r}{s}\colonequals
-  \begin{cases}
-    \frac{\displaystyle\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}%
-         {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
-         & \text{if}\ \forall i\ n_i\ge{\Set{\ell_i,r_i}}\ge{s_i}\ge{0},\\
-         0, & \text{otherwise}.
-  \end{cases}
-\end{equation*}
-Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
-Moreover, in consequence of the Chu-Vandermonde identity, we have
-\begin{equation*}
-  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
-\end{equation*}
-whenever $n_i\ge{\Set{\ell_i,r_i}}\ge{0}$ for all $i$.
-
-\section{The induced genealogy process}
-
-\subsection{Genealogies}
-
-For our purposes, a \emph{genealogy}, $G_{T}$, is a labeled, time-calibrated tree.
-It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
-There are three distinct kinds of nodes:
-\begin{inparaenum}[(i)]
-\item \emph{tip nodes}, which represent labeled living lineages;
-\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
-\item \emph{sample nodes}, which represent labeled samples.
-\end{inparaenum}
-Each node $a$ is associated with a specific time, $t(a)$.
-In particular, if $a$ is a tip node in $G_T$, then $t(a)=T$;
-if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
-Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
-We let $\event{G}{T}$ denote the set of all node-times of the genealogy $G_T$.
-Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
-Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.
-
-Formally, we define a genealogy, $G_T$, to be an ordered pair $(y,z)$.
-The $z$ element specifies the topological structure, as follows.
-Let $\leaves{G_T}$ be the set of all tip and sample nodes in $G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{G_T}$.
-The partition \emph{fineness} defines a partial order on $\func{P}$.
-Specifically, for $p,p'\in{\func{P}}$, we say $p\preceq{p'}$ if and only if for every $c'\in{p'}$ there is a unique $c\in{p}$ such that $c\supseteq{c'}$.
-The topological structure of $G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
-
-The $y$ element of $G_T$ specifies the coloring of branches and location of nodes.
-In particular, for $a\in\leaves{G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
-To make this well-defined, let $d_{t}(a)=\eth$ for $t\in{[t(a),T]}$;
-we call $\eth$ the \emph{underdeme} and define $\Demesplus\coloneq{\Demes\cup\{\eth\}}$.
-Thus $d:[0,T]\times\leaves{G_T}\to\Demesplus$ is a well-defined \cadlag\ function.
-
-Since not every node along a lineage corresponds to a change of deme, we also count the number of nodes along each lineage.
-Let $m:[0,T]\times\leaves{G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
-For $t\in{[t(a),T]}$, let $m_{t}(a)=m_{t(a)}(a)$.
-Clearly, $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
-
-Finally, we define the map $y:[0,T]\times\leaves{G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
-The two \cadlag\ functions $y$ and $z$ specify the genealogy $G_T$ completely and uniquely.
-When the genealogy is not clear from context, we write $y^{G}$ and $z^{G}$ to refer to the respective components of genealogy $G$.
-
-Clearly, the topological structure of a tree imposes constraints on the permissible functions $y$ that describe the painting and node placement along it.
-In particular, $y$ must respect the topological structure of $G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
-In other words, if $a,b\in{c}\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
-Let $\Yspace^z$ denote the space of all functions that respect the topological structure described by $z$.
-Let $\Yspace^\z_t$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.
-
-\subsection{Jump types}
-\label{sec:jump_types}
-
-\begin{figure}
-  \begin{center}
-    \resizebox{0.8\linewidth}{!}{
-      \begin{tikzpicture}[scale=2]
-        \usetikzlibrary{shapes,arrows,positioning}
-        \usetikzlibrary{arrows.meta,positioning,decorations}
-        \colorlet{darkblue}{blue!50!black}
-        \colorlet{maize}{yellow!90!black}
-        \colorlet{grey}{black!30!white}
-        \colorlet{darkgreen}{green!40!black}
-        \colorlet{lightblue}{white!30!blue}
-        \tikzstyle{label}=[font=\Large]
-        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{deme1}=[color=grey, line width=3pt, >={Circle[width=5pt,length=5pt]}]
-        \tikzstyle{deme2}=[color=darkblue, line width=3pt, >={Circle[width=5pt,length=5pt]}]
-        \tikzstyle{deme3}=[color=maize, line width=3pt, >={Circle[width=5pt,length=5pt]}]
-        \tikzstyle{axis}=[color=black, thick, >=stealth]
-        \tikzstyle{timepoint}=[color=red!70!black]
-
-        \def\ysp{0.2} %% spacing between adjacent lineages
-
-        %% Birth-type event
-        \begin{scope}[shift={(0,0)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{A}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
-          \coordinate (temp1) at ($(temp)+(0,\ysp)$);
-          \coordinate (temp2) at ($(temp)-(0,\ysp)$);
-          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
-          \draw[deme3] (mid |- temp) -- (mid |- temp2) -- (right |- temp2);
-          \draw[deme1] (left |- temp) -- (right |- temp);
-          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
-          \foreach \y in {1,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {6,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {2,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Death-type event
-        \begin{scope}[shift={(2,0)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{B}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \foreach \y in {1,2,6} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {3,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
-          \draw[deme3,dashed] (left |- temp) -- (mid |- temp);
-          \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
-          \foreach \y in {5,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Migration-type event
-        \begin{scope}[shift={(4,0)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{C}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \foreach \y in {2,3,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
-          \draw[deme2] (left |- temp) -- (mid |- temp);
-          \draw[deme3] (mid |- temp) -- (right |- temp);
-          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
-          \foreach \y in {1,5} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {6,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Sample event
-        \begin{scope}[shift={(6,0)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{D}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \foreach \y in {2,5} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {1,4,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {3,6,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \coordinate (temp) at ($(origin)+(0,-2*\ysp)$);
-          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
-          \coordinate (temp) at ($(origin)+(0,-6*\ysp)$);
-          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Neutral-type event
-        \begin{scope}[shift={(0,-2)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{E}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \foreach \y in {4,6,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {1,2,5} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {3,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Compound birth/death (Moran) event
-        \begin{scope}[shift={(2,-2)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{F}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
-          \coordinate (temp1) at ($(temp)+(0,\ysp)$);
-          \draw[deme1] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
-          \draw[deme1] (left |- temp) -- (right |- temp);
-          \draw[deme3,dashed] (left |- temp1) -- (mid |- temp1);
-          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
-          \node[text=red,font=\Huge] at (mid |- temp1) {$\times$};
-          \foreach \y in {2,5} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {1,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {6,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Compound sample/death event
-        \begin{scope}[shift={(4,-2)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{G}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \coordinate (temp) at ($(origin)+(0,-5*\ysp)$);
-          \draw[deme3] (left |- temp) -- (mid |- temp);
-          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
-          \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
-          \foreach \y in {1,6} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {2,3,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {4,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Compound event
-        \begin{scope}[shift={(6,-2)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{H}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \coordinate (temp) at ($(origin)+(0,-3*\ysp)$);
-          \draw[deme3] (left |- temp) -- (mid |- temp);
-          \draw[deme2] (mid |- temp) -- (right |- temp);
-          \coordinate (temp1) at ($(temp)+(0,-\ysp)$);
-          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
-          \coordinate (temp1) at ($(temp1)+(0,-\ysp)$);
-          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
-          \coordinate (temp1) at ($(temp1)+(0,-\ysp)$);
-          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
-          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
-          \foreach \y in {2,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {1} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-      \end{tikzpicture}
-    }
-  \end{center}
-  \caption{
-    Jump types differ by their effects on the genealogy.
-    \textbf{(A)} A birth-type jump results in the branching of one or more child lineages from the parent.
-    There can be only one parent, though the child lineages may be in demes different from those of their parent.
-    Here, a parent of the grey deme sires one child lineage in each of the blue and yellow demes.
-    The production of an event is an integer vector, with one entry for each deme.
-    The production of this event is therefore $r=(r_{\lab{grey}},r_{\lab{blue}},r_{\lab{yellow}})=(1,1,1)$.
-    \textbf{(B)} A death-type event causes the extinction of a lineage.
-    Since internal node without children are recursively removed, the affected branch disappears.
-    The production of this event is $r=(0,0,0)$.
-    \textbf{(C)} A migration-type event results in the movement of one or more lineages from one deme to another.
-    Here, one lineage moves from the blue to the yellow deme.
-    The production of this event is $r=(0,0,1)$, \ie the production is 0 for the grey and blue demes, and 1 for the yellow.
-    \textbf{(D)} In a sample-type event, one or more sample nodes (blue circles) are inserted.
-    Here, there are two samples, one in each of the grey and yellow demes.
-    Accordingly, its production is $r=(1,0,1)$.
-    \textbf{(E)} A neutral-type event has no effect on the genealogy and zero production in all demes: $r=(0,0,0)$.
-    \textbf{(F)} The theory presented here allows for compound events.
-    As an example, here a birth/death-type event occurs, wherein one yellow lineage is extinguished and a grey lineage simultaneously sires a grey child.
-    The production of this event is $r=(2,0,0)$.
-    \textbf{(G)} Here, a compound sample/death-type event occurs.
-    A yellow lineage is sampled and simultaneously extinguished.
-    Note that the lineage is not recursively removed, since sample nodes are never removed.
-    \textbf{(H)} A compound birth/migration-type event with yellow parent and production $r=c(0,4,0)$.
-    \label{fig:jump_types}
-  }
-\end{figure}
-
-We will show how a given population process naturally induces a process in the space of genealogies.
-Because different kinds of jumps have different effects on the genealogy, we need a way of distinguishing them.
-In particular, there are five distinct \emph{pure types} of jumps:
-\begin{compactenum}[(a)]
-\item \emph{Birth-type jumps} result in the branching of one or more new lineages, each from some existing lineages.
-  Examples of birth-type events include transmission events, speciations, and actual births.
-  For simplicity, we assume that all new lineages arising from a birth event share the same parent.
-\item \emph{Death-type jumps} result in the extinction of one or more lineages.
-  Examples include recovery from infection, death of a host, and species extinctions.
-  We allow for the possibility that multiple lineages die simultaneously.
-\item \emph{Migration-type jumps} result in the movement of a lineage from one deme to another.
-  Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
-  We permit multiple movements to occur simultaneously.
-\item \emph{Sample-type jumps} result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
-  We allow for the possibility that multiple samples are collected in a single sample event.
-  In this case, we assume that, in each sample-type event, each extant lineage is sampled at most once.
-\item \emph{Neutral-type jumps} result in no change to any of the lineages.
-\end{compactenum}
-\Cref{fig:markov_state} depicts an example with all five of the pure types.
-It is not necessary that a jump be of a pure type.
-Compound jumps partake of more than one type.
-For example, a sample/death-type event, in which a lineage is simultaneously sampled and removed, has been proposed \citep{Leventhal2014}, as have birth/death events in which one lineage reproduces at the same moment that another dies (\eg the Moran process \citep{Moran1958}).
-The theory presented here places few restrictions on the complexity of the events that can occur.
-
-%% Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents and demes, there is implicitly a deterministic indicator function $Q_u$, for $u\in\Jumps$, (described below) that captures these properties.
-
-\subsection{Genealogy process}
-
-We now show how a given population process induces a stochastic process, $\Gen_t$, on the space of genealogies.
-In the case of unstructured population processes (\ie those having a single deme), \citet{King2022} gave a construction.
-Although we now treat a more general case, the ideas are much the same.
-Readers wishing more detail on the construction here should consult the earlier paper.
-
-At each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event:
-\begin{compactenum}[(a)]
-\item
-  A birth-type event at time $t$ results in the creation of one new internal node, call it $b$.
-  A tip node, $a$, of the appropriate deme is chosen with uniform probability from among those present and $b$ is inserted so that its ancestor is that of $a$, while $a$ takes $b$ as its ancestor.
-  One new tip node, of the appropriate deme, is created for each of the children, all of which take $b$ as their immediate ancestor.
-\item
-  In a death-type event, one or more tip nodes of the appropriate demes are selected with uniform probability from among those present.
-  These are deleted.
-  Next, branch nodes without children are recursively removed.
-  Sample nodes are never removed.
-\item
-  In a migration-type event, the appropriate number of migrating lineages are selected at random, with uniform probability according to deme, from among those present.
-  For each selected lineage, one new branch node is inserted between the selected tip node and its ancestor.
-  The color of the descendant branch changes accordingly.
-\item
-  At a sample-type event, the appropriate number of sampled lineages are selected at random from among the tip nodes, with uniform probability according to deme.
-  One new sample node is introduced for each selected lineage:
-  each is inserted between a selected tip nodes and its ancestor.
-\item
-  At a neutral-type event, no change is made to the genealogy.
-\item
-  Finally, events of compound type are accommodated by combining the foregoing rules.
-\end{compactenum}
-In each of these events, the new node or nodes that are introduced are situated at time $t$.
-
-\subsubsection{Emergent lineages and production}
-
-The lineages which descend from an inserted node are said to \emph{emerge} from the event.
-Thus, after a birth-type event, the emerging lineages include all the new offspring as well as the parent.
-Likewise, at pure migration- or sample-type events, each migrating or sampled lineage emerges from the event.
-At pure death-type events, no lineages emerge.
-In general, at an event of mark $u$, there are $r^u_i$ emergent lineages in deme $i$.
-We require that $r^u_i$ be a constant, for each $u$ and $i$.
-Since one is free to expand the set of jump-marks $\Jumps$ as needed, this is not an important restriction on the models that the theory can accommodate.
-Thus there is a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
-We say $r^u=(r^u_i)_{i\in\Demes}$ is the \emph{production} of an event of mark $u$.
-Note that the lineages that die as a result of an event do not count in the production but that all parent lineages, if they survive the event, are counted in the production.
-
-\subsubsection{Conditional independence and exchangeability}
-
-Application of these rules at each jump of $\Xr_t$ constructs a chain of genealogies $\Grh_k$.
-In particular, at each jump-time $\Trh_k$, the genealogy $\Grh_{k-1}$ is modified according to the jump-mark $\Urh_k$ to yield $\Grh_k$.
-We view $\Grh_k$ as the embedded chain of the continuous-time genealogy process $\Gen_t$.
-It is very important to note that, conditional on $\Urh_k$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of the choices at any other times and independent of $\Xrh_k$ for all $k$.
-Moreover, by construction, any lineage within a deme is as likely as any other lineage in that deme to be selected as a parent or for death, sampling, or migration.
-We refer to this property as the \emph{exchangeability} of lineages within a deme.
-
-\subsection{Pruned and obscured genealogies}
-
-\begin{figure}
-  \caption{
-    Unpruned, pruned, and obscured enealogies.
-  }
-\end{figure}
-
-The process just described yields a genealogy that relates all extant members of the population, and all samples.
-Moreover, it details each lineage's complete history of movement through the various demes.
-The data we ultimately wish to analyze will be based only on samples, however;
-nor, in general, will the histories of deme occupancy be observable.
-A generative model must account for this loss of information.
-We therefore now describe how genealogies are \emph{pruned} to yield sample-only genealogies and then \emph{obscured} via the erasure of color from their branches.
-
-\subsubsection{Pruned genealogy}
-
-Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Pr_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every childless internal node.
-In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and possibly some of the interior nodes of the genealogy.
-Observe that a pruned genealogy is a colored genealogy:
-it retains information about where among the demes each of its lineages was through time.
-Note also that a pruned genealogy $\Pr_t$ is characterized by functions $y^{\Pr_t}$ and $z^{\Pr_t}$ just as an un-pruned genealogy is.
-
-\subsubsection{Lineage count and saturation}
-\label{sec:ells}
-
-In the following, we will find that we need to count the deme-specific numbers of lineages present in a given pruned genealogy at a given time.
-Accordingly, suppose $\Pr_T=(y,z)$ is a pruned genealogy and suppose $t\in[0,T]$.
-Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell=(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
-Clearly, $\ell$ depends only on the values, $(y_t,\z_t)$, of $y$ and $z$ at time $t$.
-Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(y,z)$ is a pruned genealogy, $\ell(y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
-We refer to $\ell$ as the \emph{lineage-count} function.
-
-We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
-In particular, given a node time $t$ in a pruned genealogy $\Pr_T=(y,z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
-Like the lineage-count function, $s$ depends only on the local structure of $\Pr_T$.
-However, $s$ depends not only on $(\y_t,\z_t)$, but also on $(\yt_t,\zt_t)$.
-Thus, we can define the \emph{saturation} function such that, whenever $\Pr_T=(\y,\z)$ is a pruned genealogy, $s(\yt_t,\y_t,\zt_t,\z_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
-
-\begin{figure}
-  \caption{
-    Figure showing lineage count and saturation.
-    \label{fig:ells}
-  }
-\end{figure}
-
-\subsubsection{Compatibility}
-\label{sec:compatibility}
-
-Suppose $\P_T$ is a pruned genealogy and $t\in\event{P}{T}$.
-The local structure of $\P_T$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
-For example, if the event in $\P_T$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
-Similarly, if the node in $\P_T$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
-To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q_u(y,y',z,z')=1$ if and only if there is some pruned genealogy $\P_T$ and $t\in{[0,T]}$,
-such that $\yt^{\P_T}_t=y$, $y^{\P_T}_t=y'$, $\zt^{\P_T}_t=z$, $z^{\P_T}_t=z'$,
-and $\P_T$ is compatible with an event of type $u$ at time $t$;
-$Q=0$ otherwise.
-We refer to $Q$ as the \emph{compatibility indicator}.
-
-\subsubsection{Obscured genealogy}
-
-The \emph{obscured genealogy} is obtained by discarding all information about demes and events not visible from the topology of the tree alone.
-In particular, if $\Pr_T=(y,z)$ is a pruned genealogy, we write $\obs(\Pr_T)=z$ to denote the obscured genealogy.
-
-\section{Results}
-
-\subsection{Likelihood for pruned genealogies}
-
-Our first result will be an expression for the likelihood of a given pruned genealogy $\P^*_T$ given the history $\Hr_T$ of the population process to time $T$.
-
-\begin{thm}\label{thm:pruned_lik}
-  Let $\P^*_T=(y^*,z^*)$  be a given pruned genealogy and
-  \begin{equation}\label{eq:phidef}
-    \phi_u(x,y,y',z,z')\coloneq\BinRatio{n(x)}{\ell(y',z')}{r^{u}}{s(y,y',z,z')}\,Q_{u}(y,y',z,z'),
-  \end{equation}
-  where $n$ is the deme occupancy (\cref{sec:demes}), $\ell$ and $s$ are the lineage-count and saturation functions, respectively (\cref{sec:ells}), $Q$ is the compatibility indicator (\cref{sec:compatibility}), and the binomial ratio is as defined in \cref{sec:binomial_ratio}.
-  Then
-  \begin{equation*}
-    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,y^*_t,\zt^*_t,z^*_t)}
-    \:\times\:\Indicator{\event{\H}{T}\supseteq{\event{\P}{T}^*}}.
-  \end{equation*}
-\end{thm}
-\begin{proof}
-  If $\event{\Hr}{T}\nsupseteq\event{\P}{T}^*$, then $\Hr_T$ and $\P^*_T$ are incompatible and $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
-  Similarly, if any event of $\Hr_T$ is incompatible with the local structure of $\P^*_T$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
-  Let us therefore suppose that neither of these conditions hold.
-  Conditional on $\Hr_T=\H_T$, at each time $t\in\event{\H}{T}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
-  In $\P^*_T$, at time $t$, there are $\ell_i=\ell_i(y^*_t,z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,y^*_t,\zt^*_t,z^*_t)$ are emergent.
-  By assumption, at each genealogical event, lineages within a deme are exchangeable:
-  each has an identical probability of being involved.
-  This exchangeability implies that each lineage present in a deme at time $t$ was equally likely to have been one of the emergent lineages.
-  In particular, at time $t$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ red balls and $n_i-r_i$ black balls, exactly $s_i$ of the drawn balls are red, namely
-  \begin{equation*}
-    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
-  \end{equation*}
-  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $\P^*_t$.
-  Moreover, since, again conditional on $\Hr_T$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
-  \begin{equation*}
-    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\BinRatio{n(\X_t)}{\ell(y^*_t,z^*_t)}{r^{\U_t}}{s(\yt^*_t,y^*_t,\zt^*_t,z^*_t)}}.
-  \end{equation*}
-  Returning to the possibility that $\H_T$ is incompatible with $\P^*_T$, since $\Prob{\Pr_T=\P^*_T}=0$ if either any $Q_u=0$ or any event of $\P^*_T$ is not an event of $\H_T$, we obtain the result.
-\end{proof}
-
-%% \begin{equation*}
-%%   \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
-%% \end{equation*}
-
-Next, we show how the likelihood of a pruned genealogies, unconditional on the history, can be computed.
-For this, we use the filter equation technology developed in \cref{sec:filter_eqns}.
-In particular, the following theorem follows immediately from \cref{prop:fullfilt}.
-
-\begin{thm}
-  Suppose that $\P^*_T=(y^*,z^*)$ is a pruned genealogy, for $T>0$.
-  If $w=w(t,x)$ satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
-  \begin{equation}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)=
-      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}
-      -\sum_u{\int{w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}}}\\
-      &\quad+\sum_{e\in\event{\P}{T}^*}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}-w(t,x)\right\}},
-    \end{aligned}
-  \end{equation}
-  where $\phi$ is defined in \cref{eq:phidef},
-  then the likelihood of $\P^*_T$ is
-  \begin{equation*}
-    \lik(\P^*_T)=\int{w(T,x)\,\dd{x}}.
-  \end{equation*}
-\end{thm}
-
-\subsection{Likelihood for obscured genealogies}
-
-Our next result concerns the likelihood of a given obscured genealogy conditional on the history.
-
-\begin{thm}\label{thm:obsc_lik}
-  Let $z^*=\obs(\P^*_T)$ be a given obscured genealogy.
-  Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{z^*}}{q(x,y)}=1$.
-  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,z^*_t)=1$
-  and that $q(x,y)>0$ whenever $\Prob{\Xr_0=x,y^{\Pr_T}_0=y}>0$.
-  Then there is a Markov jump process $\yr_t$ on $\Yspace_{z^*}$ such that
-  \begin{equation*}
-    \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
-    \Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}}
-    \times\Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*},
-  \end{equation*}
-  where $\phi$ is defined in \cref{eq:phidef} and
-  the expectation is taken over the sample paths of $\yr_t$, $t\in[0,T]$.
-\end{thm}
-\begin{proof}
-  First, observe that, since $\obs$ is a deterministic operator,
-  \begin{equation}\label{eq:IS1}
-    \CondProb{\zr_T=z^*}{\Hr_T}=\CondExpect{\Indicator{\obs(\Pr_T)=z^*}}{\Hr_T}.
-  \end{equation}
-  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
-  we will propose pruned genealogies compatible with $z^*_t$ as sample paths from a Markov process on $\Yspace_{z^*}$ and
-  evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
-  Conditional on $\Hr_T$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$, on $\Yspace_{z^*}$ such that
-  \begin{equation*}
-    \begin{gathered}
-      \Prob{\yrh_0}=q(\yrh_0),
-      \qquad
-      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_T=\H_T}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
-    \end{gathered}
-  \end{equation*}
-  The required Markov process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
-  This construction of $\yr_t$ obviously guarantees that $\event{\H}{t}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$.
-
-  Let us define $\Pr(\yr)=(\yr,z^*)$.
-  Then, by construction, for every $\yr$, $\obs(\Pr(\yr))=z^*$ and, conversely, for every pruned genealogy $\Pr_T$ satisfying $\obs(\Pr_T)=z^*$, $y^{\Pr_T}$ is a sample path of this process having nonzero probability.
-  We therefore have that
-  \begin{equation*}
-    \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
-    \Expect{\frac{\CondProb{\Pr_T=P(\yr)}{\Hr_T=\H_T}}{\pi(\yr\vert\H_T)}}
-  \end{equation*}
-  the expectation being taken with respect to the random process $\yr_t$.
-  Here, by definition,
-  \begin{equation*}
-    \pi(\yr\vert\H_T)=\prod_{t\in\event{\H}{T}}{\pi_{\U_{t}}(t,\Xt_{t},\X_{t},\yrt_{t},\yr_{t})}.
-  \end{equation*}
-  The result then follows from \cref{thm:pruned_lik}.
-\end{proof}
-
-The final result shows how to compute the likelihood of an obscured genealogy.
-It is an immediate consequence of \cref{thm:obsc_lik,prop:fullfilt}.
-
-\begin{thm}
-  Let $z^*$, be a given obscured genealogy.
-  Let $\pi$ and $\yr_t$ be as in \cref{thm:obsc_lik}.
-  For $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $u\in\Jumps$, define
-  \begin{equation*}
-    \begin{gathered}
-      \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y',\zt^*_t,z^*_t)}{\pi_u(t,x,x',y,y')},\\[2ex]
-      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),
-    \end{gathered}
-  \end{equation*}
-  where $\phi$ is defined in \cref{eq:phidef}.
-  Suppose $w=w(t,x,y)$ satisfies the filter equation
-  \begin{equation}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x,y)=
-      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}}\\
-      &-\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
-      &+\sum_{e\in\event{\z}{}^*}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y,y')\,\psi_u(t,x',x,y',y)\,\dd{x'}}}-w(t,x,y)\right\}},
-    \end{aligned}
-  \end{equation}
-  with the initial condition $w(0,x,y)=p_0(x)\,q(x,y)$.
-  Then the likelihood of $V^*_T$ is
-  \begin{equation*}
-    \lik(V^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
-  \end{equation*}
-\end{thm}
-
-\section{Examples}
-
-\subsection{SEIRS}
-
-Jumps: $\Jumps=\Set{\lab{Inf},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Birth},\lab{Death_S},\lab{Death_E},\lab{Death_I},\lab{Death_R},\lab{Sample}}$.
-
-Demes: $\Demes=\Set{\lab{E},\lab{I}}$.
-
-Jump rates:
-\begin{itemize}
-\item $\alpha_{\lab{Inf}}(t,x,x')=\beta(t)\,\frac{x^{\lab{S}}x^{\lab{I}}}{N(t)}\,\Indicator{x'=x+(-1,1,0,0)}$
-\item $\alpha_{\lab{Prog}}(x,x')=\rho\,x^{\lab{E}}\,\Indicator{x'=x+(0,-1,1,0)}$
-\item $\alpha_{\lab{Recov}}(x,x')=\gamma\,x^{\lab{I}}\,\Indicator{x'=x+(0,0,-1,1)}$
-\item $\alpha_{\lab{Wane}}(x,x')=\upsilon\,x^{\lab{R}}\,\Indicator{x'=x+(1,0,0,-1)}$
-\item $\alpha_{\lab{Sample}}(t,x,x')=\psi\,x^I\,\Indicator{x'=x}$
-\item $\alpha_{\lab{Birth}}(t,x,x')=B(t)\,\Indicator{x'=x+(1,0,0,0)}$
-\item $\alpha_{\lab{Death}_k}(x,x')=\mu\,x^k\,\Indicator{x'^j=x^j-\delta_{jk}}$, $k\in\Set{\lab{S},\lab{E},\lab{I},\lab{R}}$
-\end{itemize}
-
-%% Nonlinear filtering equation:
-%% \begin{equation}
-%%   \begin{aligned}
-%%     \frac{\partial{w}}{\partial{t}}(t,x,y,z)=&B(t)\,\left[\sum_{y'z'}w\big(t,x-(1,0,0,0),y',z'\big)-w\big(t,x,y,z\big)\right]\\
-%%     &\qquad+\sum_{k\in\{\lab{S},\lab{E},\lab{I},\lab{R}\}}\mu\,(x^k+1)\,w(t,x+\delta_{k})-\mu\,x^k\,w(t,x)\\
-%%     &\qquad+\upsilon\,x^{\lab{R}}\,w(t,x-(1,0,0,-1))\\
-%%   \end{aligned}
-%% \end{equation}
-
-\section{Discussion}
-
-\bibliographystyle{preprint}
-\bibliography{phylopomp}
-
-\appendix
-\setcounter{equation}{0}
-\titleformat{\section}[hang]{\large\bfseries}{Appendix \periodafter\thesection}{2ex}{\periodafter}{}
-\renewcommand{\theequation}{\thesection\arabic{equation}}
-
-\section{Filter equations}
-\label{sec:filter_eqns}
-
-Explicit expressions for the quantities that arise in this paper are not always readily available.
-Here, we develop tools for manipulating complex expressions that are otherwise cumbersome.
-
-\begin{defn}
-  Suppose $\Xr_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
-  \begin{equation}
-    \label{eq:kfe2}
-    \frac{\partial{u}}{\partial{t}}(t,x)
-    =\int{u(t,x')\,\beta(t,x',x)\,\dd{x'}}
-    -\int{u(t,x)\,\beta(t,x,x')\,\dd{x'}}.
-  \end{equation}
-  Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)$ are given, real-valued, measurable functions.
-  We say that the equation
-  \begin{equation}
-    \label{eq:filtereq}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)
-      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
-      -\lambda(t,x)\,w(t,x).\\
-    \end{aligned}
-  \end{equation}
-  is the \emph{filter equation} with \emph{driver} $\Xr_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, and \emph{decay} $\lambda$.
-\end{defn}
-
-\begin{remark}
-  Trivially, a Kolmogorov forward equation is itself a filter equation with boost $1$ and decay $0$.
-\end{remark}
-
-The following results show how filter equations allow one to integrate over random histories.
-
-\begin{lemma}
-  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable.
-  If $\Vr_t$ is an $\Rp$-valued random process satisfying
-  \begin{equation*}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)},
-  \end{equation*}
-  then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued deterministic process on $\Xspace$ with density $w(t,x)$.
-  Moreover, $w$ satisfies the filter equation
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{t}}=\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}
-  \end{equation*}
-  and $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
-\end{lemma}
-\begin{proof}
-  The expectation can be broken into three terms, according to whether $\H_t$ has zero, one, or more than one event in $\halfclosed{t-\Delta,t}$.
-  Accordingly, as $\Delta\downarrow{0}$,
-  \begin{equation*}
-    \begin{aligned}
-      w(t,x)=&\left(1-\Delta\,\int{\alpha(t-\Delta,x,x')\,\dd{x'}}\right)\,w(t-\Delta,x)\\
-      &\qquad+\Delta\,\int{\alpha(t-\Delta,x',x)\,B(t-\Delta,x',x)\,w(t-\Delta,x')\,\dd{x'}}+o(\Delta).
-    \end{aligned}
-  \end{equation*}
-  Taking the limit as $\Delta\downarrow{0}$, we obtain
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{t}}(t,x)
-    =\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
-    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
-  \end{equation*}
-\end{proof}
-
-\begin{lemma}
-  Suppose $\mathcal{M}\subset{\Rp}$ is finite and let $\mathcal{M}_t\coloneq\mathcal{M}\cap{[0,t]}$.
-  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable and
-  $\Vr_t$ is an $\Rp$-valued process satisfying
-  \begin{equation*}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\Indicator{\event{\H}{t}\supseteq\mathcal{M}_t}\,\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)}.
-  \end{equation*}
-  Then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued process on $\Rp^{\vert\mathcal{M}_t\vert}\times\Xspace$ with density $w(t,x)$.
-  Furthermore, $w(t,x)$ satisfies
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)&=
-      \int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
-      &t\notin{\mathcal{M}},\\
-      w(t,x)&=\int{\wt(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
-      &t\in{\mathcal{M}}
-    \end{aligned}
-  \end{equation*}
-  and $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
-\end{lemma}
-
-Filter equations afford a convenient means of computing expectations and likelihoods for pure jump processes.
-This is facilitated by the following Lemma, the statement of which relies on a non-standard Dirac delta function.
-Let $\delta(v,v')$ be the right-sided Dirac delta function satisfying $\delta(v,v')=0$ for $v\ne{v'}$ and $\int_a^b{f(v)\,\delta(v,v')\,\dd{v}}=f(v')\,\Indicator{v'\in\halfopen{a,b}}$ for all \cadlag\ $f$.
-
-\begin{lemma}
-  \label{lemma:filter}
-  The filter equation \eqref{eq:filtereq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satsfies the KFE
-  \begin{equation}
-    \label{eq:filterlemma}
-    \begin{aligned}
-      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
-      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right],
-    \end{aligned}
-  \end{equation}
-  on the space $\Xspace\times\Rp$.
-\end{lemma}
-\begin{proof}
-  \ELI{[Conditions for differentiating under the integral in the proof?
-      Need $[v^2 \lambda u(t,x,v)]_{v=0}^{\infty} = 0$.
-      That's okay at $v=0$, and a minor constraint on $u$.]}
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
-      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
-      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
-      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
-    \end{aligned}
-  \end{equation*}
-  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
-      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
-      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
-      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
-    \end{aligned}
-  \end{equation*}
-  which is simplified to obtain \cref{eq:filtereq}.
-\end{proof}
-
-\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(\Xr_t,\Vr_t)$.
-In particular, $\Xr_t$ is the driver with KFE \eqref{eq:kfe2}.
-The $\Vr_t$ is \emph{directed} by $\Xr_t$ in the sense that $\Vr$ has jumps wherever $\Xr$ does:
-when $\Xr$ jumps at time $t$ from $x$ to $x'$, $\Vr$ jumps by the multiplicative factor $B(t,x,x')\ge{0}$.
-Between jumps, $\Vr_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
-If we view $\Vr_t$ as a weight, then \cref{lemma:filter} tells us how the $\Vr_t$-weighted average of $\Xr_t$ evolves in time:
-this average is simply $\int\!{w(t,x)\,\dd{x}}$.
-
-%% Filter equations allow us to pass easily between equivalent representations of a process.
-%% For example, an equivalent way of representing $\Xr_t$ is in terms of its embedded chain and event times.
-%% Let $\Xrh_k$ be the embedded chain of $\Xr_t$ and let $\Trh_k$ be the point process of its event times.
-%% It is elementary that
-%% \fontsize{10pt}{12pt}\selectfont
-%% \begin{equation*}
-%%   \begin{gathered}
-%%     \Prob{\Xrh_k=\Xh_k\;\Big\vert\;\Xrh_{k-1}=\Xh_{k-1},\Trh_k=\Th_k}=\frac{\alpha(\Th_k,\Xh_{k-1},\Xh_k)}{\int{\alpha(\Th_k,\Xh_{k-1},x')\dd{x'}}},\\
-%%     \Prob{\Trh_k>\Trh_{k-1}+t\;\Big\vert\;\Xrh_{k-1}=\Xh_{k-1},\Trh_{k-1}=\Th_{k-1}}=\exp{\left(-\int_{0}^{t}{\int{\alpha(\Th_{k-1}+s,\Xh_{k-1},x')\,\dd{x'}}\,\dd{s}}\right)}.
-%%   \end{gathered}
-%% \end{equation*}
-%% \normalfont
-%% Fixing $\nu>0$ and making the definitions,
-%% \begin{equation}\label[pluralequation]{eq:nudefs}
-%%   \begin{gathered}
-%%     A(t,x)=\int{\alpha(t,x,x')\,\dd{x'}}, \qquad
-%%     \pi(t,x,x')=\frac{\alpha(t,x,x')}{A(t,x)},\\
-%%     B(t,x,x')=\frac{A(t,x)}{\nu}, \qquad
-%%     \lambda(t,x)=A(t,x)-\nu,
-%%   \end{gathered}
-%% \end{equation}
-%% we can rewrite the KFE as
-%% \begin{equation}\label{eq:poissdriver}
-%%   \frac{\partial{w}}{\partial{t}}(t,x)=
-%%   \int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
-%%   -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
-%%   -\lambda(t,x)\,w(t,x).
-%% \end{equation}
-%% Here, $\nu$ is the intensity of a time-homogenous Poisson process.
-%% Note that $\pi$ is the probability kernel of the embedded chain $\Xrh$ and $A(t,x)$ is the intensity of the $\Trh_k$ process.
-%% We recognize this equation as the filter equation with boost $B$, decay $\lambda$, and driver generated by $\nu\,\pi(t,x,x')$.
-%% It corresponds to the following procedure for simulating $\Xr_t$:
-%% \begin{compactenum}[(a)]
-%% \item Simulate jump times according to the rate-$\nu$ Poisson process.
-%% \item Simulate the embedded chain $\Xrh_k$ using the kernel $\pi$.
-%% \item Weight the realization by the product of the $B$ factors.
-%%   Note that this makes the appropriate importance-sampling correction.
-%% \end{compactenum}
-
-\end{document}

### 2024-01-21 figs/markov_diagram.tex
more figures

--- 2024-01-17 ms.Rnw
@@
-\documentclass[10pt,reqno,final]{amsart}
-\input{header}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta,positioning,decorations,calc,math}
-
-\title[Exact Phylodynamics]{Exact Phylodynamics via Structured Markov Genealogy Processes}
-\author[King]{Aaron~A.~King}
-\address{
-  A.~A.~King,
-  Department of Ecology \& Evolutionary~Biology,
-  Center for the Study of Complex~Systems, and
-  Department of Mathematics,
-  University of Michigan,
-  Ann~Arbor, MI~48109~USA\\
-  Santa~Fe~Institute,
-  1399 Hyde~Park~Road,
-  Santa~Fe, NM~87501~USA
-}
-\email{kingaa@umich.edu}
-\urladdr{\href{https://kinglab.eeb.lsa.umich.edu/}{https://kinglab.eeb.lsa.umich.edu/}}
-\author[Lin]{Qianying~Lin}
-\address{
-  Q.-Y.~Lin,
-  Theoretical Biology and Biophysics,
-  Los~Alamos National Laboratory,
-  Los~Alamos, NM~87545~USA
-}
-\author[Ionides]{Edward~L.~Ionides}
-\address{
-  E.~L.~Ionides,
-  Department of Statistics
-  University of Michigan,
-  Ann~Arbor, MI~48109~USA
-}
-\date{\today}
-
-\hypersetup{pdftitle={Exact Phylodynamics via Structured Markov Genealogy Processes}}
-\hypersetup{pdfauthor={A.A. King, Q.-Y. Lin, E.L. Ionides}}
-\hypersetup{urlcolor=blue,citecolor=blue,linkcolor=blue,filecolor=blue}
-
-<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
-prefix <- "sgp"
-source("setup.R")
-@
-<<packages,include=FALSE,cache=FALSE>>=
-library(tidyverse)
-library(ggtree)
-library(pomp)
-library(cowplot)
-library(phylopomp)
-stopifnot(getRversion() >= "4.3")
-stopifnot(packageVersion("pomp")>="5.1")
-stopifnot(packageVersion("phylopomp")>="0.9.2")
-theme_set(theme_bw(base_family="serif"))
-set.seed(1159254136)
-@
-
-\begin{document}
-
-\begin{abstract}
-  We derive an algorithm for the exact computation of the likelihood of an observed transmission tree.
-\end{abstract}
-
-\maketitle
-
-\section{Introduction}
-
-Problem of phylodynamics.
-Factorization of problem into two subproblems.
-
-Relation to previous work.
-Existing methods \citep{Volz2009a,Stadler2010}.
-Large-population, small sample-size approximations.
-
-Extension of previous results \citep{King2022}.
-Broader class of state-spaces.
-Accommodating discrete structure.
-
-Classes of Markov processes.
-Utility and flexibility of Markov assumptions.
-
-Population process induces Markov history and genealogy processes.
-Using these, we derive equations for the likelihood of a genealogy conditional on the history.
-We then integrate out the history to obtain nonlinear filtering equations, the solution of which yields the likelihood.
-These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likelihood.
-We demonstrate with several examples.
-
-In the following, we show a Markov population process of the kind that is a staple in epidemiology induces a Markov process on the space of genealogies.
-We then show how one can comput the likelihood of a given genealogy.
-
-\section{Mathematical preliminaries}
-
-\subsection{Notation}
-
-Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\Xr$), denotes a random element.
-We will be concerned with a variety of stochastic processes, in both discrete and continuous time.
-In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
-In the case of continuous-time processes, we will assume that sample paths are right-continuous with left limits (\ie \cadlag).
-We will frequently need to refer to the left-limit of such a process.
-Accordingly, if $\Zr_t$ is a \cadlag\ random process, we define
-\begin{equation*}
-  \Zrt_t\colonequals
-  \begin{cases}
-    \displaystyle\lim_{t'\,\uparrow\,{t}}\;\Zr_{t'}, & t>0,\\[2ex]
-    \Zr_0, & t=0.\\
-  \end{cases}
-\end{equation*}
-Note that $\Zrt_t$ is thus left-continuous with right limits.
-
-If $Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
-In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
-$\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
-If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its event times in the interval $[0,t]$.
-
-\subsection{Population process}
-
-Motivating examples: compartmental models.
-Wide variety of models.
-Linear chain trick.
-Migration, superspreading, competition between strains.
-
-\begin{figure}
-  \begin{center}
-    \resizebox{0.9\linewidth}{!}{
-      \begin{tikzpicture}[scale=1]
-        \tikzstyle{box}=[color=black, draw, fill=white, very thick, minimum size=3em]
-        \tikzstyle{label}=[font=\Large]
-        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{flow}=[color=black, very thick, >=stealth]
-        \tikzstyle{deme}=[fill=black!20!white]
-        \tikzstyle{modulate}=[color=blue, thick, >=latex]
-        \coordinate (origin) at (0,0);
-        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{A}};
-        \input{figs/seir_diagram}
-        \coordinate (origin) at (9,0);
-        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{B}};
-        \input{figs/siir_diagram}
-        \coordinate (origin) at (0,-4);
-        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{C}};
-        \input{figs/covid_diagram}
-      \end{tikzpicture}
-    }
-  \end{center}
-  \caption{
-    Examples of discretely-structured population models.
-    Demes are shaded.
-    (A) An SEIRS model.
-    Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
-    Upon recovery ($\lab{R}$), individuals experience immunity from reinfection.
-    If this immunity wanes, they re-enter the susceptible compartment.
-    Pathogen lineages are to be found in hosts within the $\lab{E}$ and $\lab{I}$ compartments only.
-    Accordingly, there are two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
-    If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
-    \AAK{[Perhaps another one or two examples here?]}
-    \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
-    \label{fig:examples}
-  }
-\end{figure}
-
-We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
-In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
-The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
-In particular, we suppose that
-\begin{equation}\label{eq:ic}
-  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
-\end{equation}
-for all measurable sets $\mathcal{E}\subseteq\Xspace$.
-For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
-More precisely, the transition rates have the following properties:
-\begin{equation*}
-  \begin{gathered}
-    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
-  \end{gathered}
-\end{equation*}
-for all $t\in\Rp$ and $x,x'\in\Xspace$.
-Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
-Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
-We assume that $\Kr_t$ is a simple counting process so that
-\begin{equation*}
-  \begin{gathered}
-    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
-    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
-    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
-  \end{gathered}
-\end{equation*}
-We will further assume that $\Xr_t$ is non-explosive, \ie that $\Prob{\Kr_t<\infty}=1$ for all $t$.
-
-\subsubsection{Kolmogorov forward equation}
-
-The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
-\begin{equation}\label{eq:kfe}
-  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
-\end{equation}
-and if, moreover, $w(0,x) = p_0(x)$,
-then $\int_{\mathcal{E}}\!{w(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
-\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.
-
-Another perspective on the Markov processes is to be had from its Markov state transition diagram (\cref{fig:markov_state}).
-
-\begin{figure}
 \begin{center}
@@
       \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum size=5em]
-        \tikzstyle{trans}=[color=black, thick, >=stealth]
+      \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum height=8em]
+      \tikzstyle{trans}=[color=black, font=\Large, thick, >=stealth]
       \node[state] (base) at (0,0) {$x=(S,E,I,R)$};
-        \node[state] (trans) at (165:8.4) {$x'=(S-1,E+1,I,R)$};
-        \node[state] (prog) at (195:8) {$x'=(S,E-1,I+1,R+1)$};
-        \node[state] (recov) at (295:4) {$x'=(S,E,I-1,R+1)$};
-        \node[state] (wane) at (355:9) {$x'=(S+1,E,I,R-1)$};
+      \node[state] (trans) at (165:8) {$x'=(S-1,E+1,I,R)$};
+      \node[state] (prog) at (205:7) {$x'=(S,E-1,I+1,R+1)$};
+      \node[state] (recov) at (305:5) {$x'=(S,E,I-1,R+1)$};
+      \node[state] (wane) at (350:8) {$x'=(S+1,E,I,R-1)$};
       \node[state] (sample) at (18:7.5) {$x'=(S,E,I,R)$};
@@
       \draw[trans,->] (base) -- (sample) node[midway,above,sloped] {$\lab{Sample}$};
-        \node[font=\Large] (U) at (95:3.2) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
+      \node[font=\LARGE] (U) at (95:3.5) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
     \end{tikzpicture}
@@
 \caption{
-    Markov state transition diagram for the SEIRS model depicted in \cref{fig:examples}A.
+  Markov state transition diagram for the SEIRS model depicted in \cref{fig:example_models}A.
   The state, $x$, is characterized by four numbers, $S$, $E$, $I$, and $R$.
@@
 }
-\end{figure}
-
-\subsubsection{Structured populations, demes}
-
-In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
-\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
-Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
-In particular, we suppose that there are a countable set of subpopulations that differ in their vital rates, but within each of which, individual lineages are statistically identical.
-We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
-
-For any $i\in\Demes$, we let $n_i(\Xr_t)$ denote the number of lineages present in deme $i$ at time $t$, \ie the \emph{occupancy} of deme $i$.
-Thus $n(\Xr_t)\in\Zp^{\Demes}$ is the vector of deme occupancies.
-
-\subsubsection{Jump marks}
-
-In the following, it will be useful to break the jumps into distinct categories.
-For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
-\begin{equation*}
-  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
-\end{equation*}
-In \cref{fig:markov_state}, we use the marks to distinguish events of different types.
-Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.
-
-Let us define the \emph{jump mark} process, $\Ur_t$, to be the mark of the latest jump as of time $t$.
-As usual, we take the sample paths of $\Ur_t$ to be \cadlag.
-Observe that $(\Xr_t,\Ur_t)$ is a Markov process, though $\Ur_t$ is not.
-
-\subsection{Examples}
-
-\subsubsection{SEIRS model}
-
-\subsubsection{SIIR model}
-
-\subsubsection{Linear birth-death model}
-
-\subsubsection{Moran model and the Kingman coalescent}
-
-\subsection{History process}
-
-Consider the Markov process $(\Xr_t,\Ur_t)$.
-We define its \emph{history process}, $\Hr_t$, to be the restriction of the random function $s\mapsto(\Xr_s,\Ur_s)$ to the interval $[0,t]$.
-Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.
-Alternatively, one can think of $\Hr_t$ as consisting of the sequence
-$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
-In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
-The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
-\begin{equation*}\label{eq:Hdens}
-  \begin{aligned}
-    \pi^{\mathrm{H}}(\dd{\H_t})
-    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
-    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
-    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
-  \end{aligned}
-\end{equation*}
-where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.
-
-\subsection{Binomial ratio}\label{sec:binomial_ratio}
-
-For $n,r,\ell,s\in{\Zp^\Demes}$, define the \emph{binomial ratio}
-\begin{equation*}
-  \BinRatio{n}{\ell}{r}{s}\colonequals
-  \begin{cases}
-    \frac{\displaystyle\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}%
-         {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
-         & \text{if}\ \forall i\ n_i\ge{\Set{\ell_i,r_i}}\ge{s_i}\ge{0},\\
-         0, & \text{otherwise}.
-  \end{cases}
-\end{equation*}
-Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
-Moreover, in consequence of the Chu-Vandermonde identity, we have
-\begin{equation*}
-  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
-\end{equation*}
-whenever $n_i\ge{\Set{\ell_i,r_i}}\ge{0}$ for all $i$.
-
-\section{The induced genealogy process}
-
-\subsection{Genealogies}
-
-For our purposes, a \emph{genealogy}, $G_{T}$, is a labeled, time-calibrated tree.
-It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
-There are three distinct kinds of nodes:
-\begin{inparaenum}[(i)]
-\item \emph{tip nodes}, which represent labeled living lineages;
-\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
-\item \emph{sample nodes}, which represent labeled samples.
-\end{inparaenum}
-Each node $a$ is associated with a specific time, $t(a)$.
-In particular, if $a$ is a tip node in $G_T$, then $t(a)=T$;
-if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
-Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
-We let $\event{G}{T}$ denote the set of all node-times of the genealogy $G_T$.
-Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
-Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.
-
-Formally, we define a genealogy, $G_T$, to be an ordered pair $(y,z)$.
-The $z$ element specifies the topological structure, as follows.
-Let $\leaves{G_T}$ be the set of all tip and sample nodes in $G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{G_T}$.
-The partition \emph{fineness} defines a partial order on $\func{P}$.
-Specifically, for $p,p'\in{\func{P}}$, we say $p\preceq{p'}$ if and only if for every $c'\in{p'}$ there is a unique $c\in{p}$ such that $c\supseteq{c'}$.
-The topological structure of $G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
-
-The $y$ element of $G_T$ specifies the coloring of branches and location of nodes.
-In particular, for $a\in\leaves{G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
-To make this well-defined, let $d_{t}(a)=\eth$ for $t\in{[t(a),T]}$;
-we call $\eth$ the \emph{underdeme} and define $\Demesplus\coloneq{\Demes\cup\{\eth\}}$.
-Thus $d:[0,T]\times\leaves{G_T}\to\Demesplus$ is a well-defined \cadlag\ function.
-
-Since not every node along a lineage corresponds to a change of deme, we also count the number of nodes along each lineage.
-Let $m:[0,T]\times\leaves{G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
-For $t\in{[t(a),T]}$, let $m_{t}(a)=m_{t(a)}(a)$.
-Clearly, $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
-
-Finally, we define the map $y:[0,T]\times\leaves{G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
-The two \cadlag\ functions $y$ and $z$ specify the genealogy $G_T$ completely and uniquely.
-When the genealogy is not clear from context, we write $y^{G}$ and $z^{G}$ to refer to the respective components of genealogy $G$.
-
-Clearly, the topological structure of a tree imposes constraints on the permissible functions $y$ that describe the painting and node placement along it.
-In particular, $y$ must respect the topological structure of $G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
-In other words, if $a,b\in{c}\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
-Let $\Yspace^z$ denote the space of all functions that respect the topological structure described by $z$.
-Let $\Yspace^\z_t$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.
-
-\subsection{Jump types}\label{sec:jump_types}
-
-We will show how a given population process naturally induces a process in the space of genealogies.
-Because different kinds of jumps have different effects on the genealogy, we need a way of distinguishing them.
-In particular, there are five distinct \emph{pure types} of jumps:
-\begin{compactenum}[(a)]
-\item \emph{Birth-type jumps} result in the branching of one or more new lineages, each from some existing lineages.
-  Examples of birth-type events include transmission events, speciations, and actual births.
-  For simplicity, we assume that all new lineages arising from a birth event share the same parent.
-\item \emph{Death-type jumps} result in the extinction of one or more lineages.
-  Examples include recovery from infection, death of a host, and species extinctions.
-  We allow for the possibility that multiple lineages die simultaneously.
-\item \emph{Migration-type jumps} result in the movement of a lineage from one deme to another.
-  Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
-  We permit multiple movements to occur simultaneously.
-\item \emph{Sample-type jumps} result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
-  We allow for the possibility that multiple samples are collected in a single sample event.
-  In this case, we assume that, in each sample-type event, each extant lineage is sampled at most once.
-\item \emph{Neutral-type jumps} result in no change to any of the lineages.
-\end{compactenum}
-\Cref{fig:markov_state} depicts an example with all five of the pure types.
-It is not necessary that a jump be of a pure type.
-Compound jumps partake of more than one type.
-For example, a sample/death-type event, in which a lineage is simultaneously sampled and removed, has been proposed \citep{Leventhal2014}, as have birth/death events in which one lineage reproduces at the same moment that another dies (\eg the Moran process \citep{Moran1958}).
-The theory presented here places few restrictions on the complexity of the events that can occur.
-
-%% Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents and demes, there is implicitly a deterministic indicator function $Q_u$, for $u\in\Jumps$, (described below) that captures these properties.
-
-\subsection{Genealogy process}
-
-We now show how a given population process induces a stochastic process, $\Gen_t$, on the space of genealogies.
-In the case of unstructured population processes (\ie those having a single deme), \citet{King2022} gave a construction.
-Although we now treat a more general case, the ideas are much the same.
-Readers wishing more detail on the construction here should consult the earlier paper.
-
-At each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event:
-\begin{compactenum}[(a)]
-\item
-  A birth-type event at time $t$ results in the creation of one new internal node, call it $b$.
-  A tip node, $a$, of the appropriate deme is chosen with uniform probability from among those present and $b$ is inserted so that its ancestor is that of $a$, while $a$ takes $b$ as its ancestor.
-  One new tip node, of the appropriate deme, is created for each of the children, all of which take $b$ as their immediate ancestor.
-\item
-  In a death-type event, one or more tip nodes of the appropriate demes are selected with uniform probability from among those present.
-  These are deleted.
-  Next, branch nodes without children are recursively removed.
-  Sample nodes are never removed.
-\item
-  In a migration-type event, the appropriate number of migrating lineages are selected at random, with uniform probability according to deme, from among those present.
-  For each selected lineage, one new branch node is inserted between the selected tip node and its ancestor.
-  The color of the descendant branch changes accordingly.
-\item
-  At a sample-type event, the appropriate number of sampled lineages are selected at random from among the tip nodes, with uniform probability according to deme.
-  One new sample node is introduced for each selected lineage:
-  each is inserted between a selected tip nodes and its ancestor.
-\item
-  At a neutral-type event, no change is made to the genealogy.
-\item
-  Finally, events of compound type are accommodated by combining the foregoing rules.
-\end{compactenum}
-In each of these events, the new node or nodes that are introduced are situated at time $t$.
-
-\begin{figure}
-  \caption{
-    Illustration of genealogy processes.
-    \AAK{[Similar to that of \citet{King2022} but with multiple demes represented.]}
-  }
-\end{figure}
-
-\subsubsection{Emergent lineages and production}
-
-The lineages which descend from an inserted node are said to \emph{emerge} from the event.
-Thus, after a birth-type event, the emerging lineages include all the new offspring as well as the parent.
-Likewise, at pure migration- or sample-type events, each migrating or sampled lineage emerges from the event.
-At pure death-type events, no lineages emerge.
-In general, at an event of mark $u$, there are $r^u_i$ emergent lineages in deme $i$.
-We require that $r^u_i$ be a constant, for each $u$ and $i$.
-Since one is free to expand the set of jump-marks $\Jumps$ as needed, this is not an important restriction on the models that the theory can accommodate.
-Thus there is a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
-We say $r^u=(r^u_i)_{i\in\Demes}$ is the \emph{production} of an event of mark $u$.
-Note that the lineages that die as a result of an event do not count in the production but that all parent lineages, if they survive the event, are counted in the production.
-
-\subsubsection{Conditional independence and exchangeability}
-
-Application of these rules at each jump of $\Xr_t$ constructs a chain of genealogies $\Grh_k$.
-In particular, at each jump-time $\Trh_k$, the genealogy $\Grh_{k-1}$ is modified according to the jump-mark $\Urh_k$ to yield $\Grh_k$.
-We view $\Grh_k$ as the embedded chain of the continuous-time genealogy process $\Gen_t$.
-It is very important to note that, conditional on $\Urh_k$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of the choices at any other times and independent of $\Xrh_k$ for all $k$.
-Moreover, by construction, any lineage within a deme is as likely as any other lineage in that deme to be selected as a parent or for death, sampling, or migration.
-We refer to this property as the \emph{exchangeability} of lineages within a deme.
-
-\subsection{Pruned and obscured genealogies}
-
-\subsubsection{Pruned genealogy}
-
-Although the process just described yields a genealogy that relates all extant members of the population, and all samples, the data we ultimately wish to analyze will be based only on samples.
-We therefore describe how the genealogy process is \emph{pruned} to yield the sample-only genealogy.
-Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Pr_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every childless internal node.
-In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and possibly some of the interior nodes of the genealogy.
-Observe that a pruned genealogy is a colored genealogy:
-it retains information about where among the demes each of its lineages was through time.
-Note also that a pruned genealogy $\Pr_t$ is characterized by functions $y^{\Pr_t}$ and $z^{\Pr_t}$ just as an un-pruned genealogy is.
-
-\subsubsection{Lineage count and saturation}
-
-In the following, we will find that we need to count the deme-specific numbers of lineages present in a given pruned genealogy at a given time.
-Accordingly, suppose $\Pr_T=(y,z)$ is a pruned genealogy and suppose $t\in[0,T]$.
-Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell=(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
-Clearly, $\ell$ depends only on the values, $(y_t,\z_t)$, of $y$ and $z$ at time $t$.
-Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(y,z)$ is a pruned genealogy, $\ell(y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
-We refer to $\ell$ as the \emph{lineage-count} function.
-
-We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
-In particular, given a node time $t$ in a pruned genealogy $\Pr_T=(y,z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
-Like the lineage-count function, $s$ depends only on the local structure of $\Pr_T$.
-However, $s$ depends not only on $(\y_t,\z_t)$, but also on $(\yt_t,\zt_t)$.
-Thus, we can define the \emph{saturation} function such that, whenever $\Pr_T=(\y,\z)$ is a pruned genealogy, $s(\yt_t,\y_t,\zt_t,\z_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
-
-\begin{figure}
-  \caption{
-    Figure showing lineage count and saturation.
-    \label{fig:ells}
-  }
-\end{figure}
-
-\subsubsection{Obscured genealogy}
-
-As we have seen, a pruned genealogy contains information about the full history of each sample lineage, including the times at which it entered or exited any deme, sired offspring, or was sampled.
-The data we seek to analyze will typically lack much of this information.
-Accordingly, we define the \emph{obscured genealogy} to be that obtained by discarding all information about demes and events not visible from the topology of the tree alone.
-In particular, if $\Pr_T=(y,z)$ is a pruned genealogy, we write $\obs(\Pr_T)=z$ to denote the obscured genealogy.
-
-\section{Results}
-
-\subsection{Likelihood for pruned genealogies}
-
-Our first result will be an expression for the likelihood of a given pruned genealogy $\P^*_T$ given the history $\Hr_T$ of the population process to time $T$.
-Of course, not every history is compatible with $\P^*_T$.
-In particular, $\Hr_T$ is only compatible with $\P^*_T$ if every node-time in $\P^*_T$ is an event time of $\Hr_T$, \ie only if $\event{\Hr}{T}\supseteq\event{P}{T}^*$.
-Moreover, even if an event in $\Hr_T$ coincides with one of the node times of $\P^*_T$, it may be incompatible with the local structure of $\P^*_T$.
-For example, if no sample is present in $\P^*_T$ at time $t$ and $\Hr_T$ contains an event of sample type at that time, then incompatibility arises.
-As another example, if $\P^*_T$ has a node at time $t$ at which a lineage moves from deme $i$ to deme $i'$, but there is no $u\in\Jumps$ for which such a move is possible, then incompatibility arises again.
-To keep track of this, let us introduce the indicator function $Q$ such that $Q_u(y,y',z,z')=1$ if and only if there is some pruned genealogy $\P_T$ and $t\in{[0,T]}$,
-such that $\yt^{\P_T}_t=y$, $y^{\P_T}_t=y'$, $\zt^{\P_T}_t=z$, $z^{\P_T}_t=z'$,
-and $\P_T$ is compatible with an event of type $u$ at time $t$;
-$Q=0$ otherwise.
-
-\begin{thm}\label{thm:pruned_lik}
-  Let $\P^*_T=(y^*,z^*)$  be a given pruned genealogy and
-  \begin{equation}\label{eq:phidef}
-    \phi_u(x,y,y',z,z')\coloneq\BinRatio{n(x)}{\ell(y',z')}{r^{u}}{s(y,y',z,z')}\,Q_{u}(y,y',z,z')
-  \end{equation}
-  (cf.~\cref{sec:binomial_ratio}).
-  Then
-  \begin{equation*}
-    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,y^*_t,\zt^*_t,z^*_t)}
-    \:\times\:\Indicator{\event{\H}{T}\supseteq{\event{\P}{T}^*}}.
-  \end{equation*}
-\end{thm}
-\begin{proof}
-  As we have already observed, if $\event{\Hr}{T}\nsupseteq\event{\P}{T}^*$, then $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
-  Similarly, if there is any event of $\Hr_T$ which is incompatible with $\P^*_T$, $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
-  Let us therefore suppose that neither of these conditions hold.
-  Conditional on $\Hr_T=\H_T$, at each time $t\in\event{\H}{T}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
-  In $\P^*_T$, at time $t$, there are $\ell_i=\ell_i(y^*_t,z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,y^*_t,\zt^*_t,z^*_t)$ are emergent.
-  By assumption, at each genealogical event, lineages within a deme are exchangeable:
-  each has an identical probability of being involved.
-  This exchangeability implies that each lineage present in a deme at time $t$ was equally likely to have been one of the emergent lineages.
-  In particular, at time $t$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ red balls and $n_i-r_i$ black balls, exactly $s_i$ of the drawn balls are red, namely
-  \begin{equation*}
-    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
-  \end{equation*}
-  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $\P^*_t$.
-  Moreover, since, again conditional on $\Hr_T$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
-  \begin{equation*}
-    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\BinRatio{n(\X_t)}{\ell(y^*_t,z^*_t)}{r^{\U_t}}{s(\yt^*_t,y^*_t,\zt^*_t,z^*_t)}}.
-  \end{equation*}
-  Returning to the possibility that $\H_T$ is incompatible with $\P^*_T$, since $\Prob{\Pr_T=\P^*_T}=0$ if either any $Q_u=0$ or any event of $\P^*_T$ is not an event of $\H_T$, we obtain the result.
-\end{proof}
-
-%% \begin{equation*}
-%%   \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
-%% \end{equation*}
-
-Next, we show how the likelihood of a pruned genealogies, unconditional on the history, can be computed.
-For this, we use the filter equation technology developed in \cref{sec:filter_eqns}.
-In particular, the following theorem follows immediately from \cref{prop:fullfilt}.
-
-\begin{thm}
-  Suppose that $\P^*_T=(y^*,z^*)$ is a pruned genealogy, for $T>0$.
-  If $w=w(t,x)$ satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
-  \begin{equation}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)=
-      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}
-      -\sum_u{\int{w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}}}\\
-      &\quad+\sum_{e\in\event{\P}{T}^*}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}-w(t,x)\right\}},
-    \end{aligned}
-  \end{equation}
-  where $\phi$ is defined in \cref{eq:phidef},
-  then the likelihood of $\P^*_T$ is
-  \begin{equation*}
-    \lik(\P^*_T)=\int{w(T,x)\,\dd{x}}.
-  \end{equation*}
-\end{thm}
-
-\subsection{Likelihood for obscured genealogies}
-
-Our next result concerns the likelihood of a given obscured genealogy conditional on the history.
-
-\begin{thm}\label{thm:obsc_lik}
-  Let $z^*=\obs(\P^*_T)$ be a given obscured genealogy.
-  Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{z^*}}{q(x,y)}=1$.
-  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,z^*_t)=1$
-  and that $q(x,y)>0$ whenever $\Prob{\Xr_0=x,y^{\Pr_T}_0=y}>0$.
-  Then there is a Markov jump process $\yr_t$ on $\Yspace_{z^*}$ such that
-  \begin{equation*}
-    \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
-    \Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}}
-    \times\Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*},
-  \end{equation*}
-  where $\phi$ is defined in \cref{eq:phidef} and
-  the expectation is taken over the sample paths of $\yr_t$, $t\in[0,T]$.
-\end{thm}
-\begin{proof}
-  First, observe that, since $\obs$ is a deterministic operator,
-  \begin{equation}\label{eq:IS1}
-    \CondProb{\zr_T=z^*}{\Hr_T}=\CondExpect{\Indicator{\obs(\Pr_T)=z^*}}{\Hr_T}.
-  \end{equation}
-  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
-  we will propose pruned genealogies compatible with $z^*_t$ as sample paths from a Markov process on $\Yspace_{z^*}$ and
-  evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
-  Conditional on $\Hr_T$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$, on $\Yspace_{z^*}$ such that
-  \begin{equation*}
-    \begin{gathered}
-      \Prob{\yrh_0}=q(\yrh_0),
-      \qquad
-      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_T=\H_T}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
-    \end{gathered}
-  \end{equation*}
-  The required Markov process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
-  This construction of $\yr_t$ obviously guarantees that $\event{\H}{t}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$.
-
-  Let us define $\Pr(\yr)=(\yr,z^*)$.
-  Then, by construction, for every $\yr$, $\obs(\Pr(\yr))=z^*$ and, conversely, for every pruned genealogy $\Pr_T$ satisfying $\obs(\Pr_T)=z^*$, $y^{\Pr_T}$ is a sample path of this process having nonzero probability.
-  We therefore have that
-  \begin{equation*}
-    \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
-    \Expect{\frac{\CondProb{\Pr_T=P(\yr)}{\Hr_T=\H_T}}{\pi(\yr\vert\H_T)}}
-  \end{equation*}
-  the expectation being taken with respect to the random process $\yr_t$.
-  Here, by definition,
-  \begin{equation*}
-    \pi(\yr\vert\H_T)=\prod_{t\in\event{\H}{T}}{\pi_{\U_{t}}(t,\Xt_{t},\X_{t},\yrt_{t},\yr_{t})}.
-  \end{equation*}
-  The result then follows from \cref{thm:pruned_lik}.
-\end{proof}
-
-The final result shows how to compute the likelihood of an obscured genealogy.
-It is an immediate consequence of \cref{thm:obsc_lik,prop:fullfilt}.
-
-\begin{thm}
-  Let $z^*$, be a given obscured genealogy.
-  Let $\pi$ and $\yr_t$ be as in \cref{thm:obsc_lik}.
-  For $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $u\in\Jumps$, define
-  \begin{equation*}
-    \begin{gathered}
-      \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y',\zt^*_t,z^*_t)}{\pi_u(t,x,x',y,y')},\\[2ex]
-      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),
-    \end{gathered}
-  \end{equation*}
-  where $\phi$ is defined in \cref{eq:phidef}.
-  Suppose $w=w(t,x,y)$ satisfies the filter equation
-  \begin{equation}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x,y)=
-      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}}\\
-      &-\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
-      &+\sum_{e\in\event{\z}{}^*}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y,y')\,\psi_u(t,x',x,y',y)\,\dd{x'}}}-w(t,x,y)\right\}},
-    \end{aligned}
-  \end{equation}
-  with the initial condition $w(0,x,y)=p_0(x)\,q(x,y)$.
-  Then the likelihood of $V^*_T$ is
-  \begin{equation*}
-    \lik(V^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
-  \end{equation*}
-\end{thm}
-
-\section{Examples}
-
-\subsection{SEIRS}
-
-Jumps: $\Jumps=\Set{\lab{Inf},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Birth},\lab{Death_S},\lab{Death_E},\lab{Death_I},\lab{Death_R},\lab{Sample}}$.
-
-Demes: $\Demes=\Set{\lab{E},\lab{I}}$.
-
-Jump rates:
-\begin{itemize}
-\item $\alpha_{\lab{Inf}}(t,x,x')=\beta(t)\,\frac{x^{\lab{S}}x^{\lab{I}}}{N(t)}\,\Indicator{x'=x+(-1,1,0,0)}$
-\item $\alpha_{\lab{Prog}}(x,x')=\rho\,x^{\lab{E}}\,\Indicator{x'=x+(0,-1,1,0)}$
-\item $\alpha_{\lab{Recov}}(x,x')=\gamma\,x^{\lab{I}}\,\Indicator{x'=x+(0,0,-1,1)}$
-\item $\alpha_{\lab{Wane}}(x,x')=\upsilon\,x^{\lab{R}}\,\Indicator{x'=x+(1,0,0,-1)}$
-\item $\alpha_{\lab{Sample}}(t,x,x')=\psi\,x^I\,\Indicator{x'=x}$
-\item $\alpha_{\lab{Birth}}(t,x,x')=B(t)\,\Indicator{x'=x+(1,0,0,0)}$
-\item $\alpha_{\lab{Death}_k}(x,x')=\mu\,x^k\,\Indicator{x'^j=x^j-\delta_{jk}}$, $k\in\Set{\lab{S},\lab{E},\lab{I},\lab{R}}$
-\end{itemize}
-
-%% Nonlinear filtering equation:
-%% \begin{equation}
-%%   \begin{aligned}
-%%     \frac{\partial{w}}{\partial{t}}(t,x,y,z)=&B(t)\,\left[\sum_{y'z'}w\big(t,x-(1,0,0,0),y',z'\big)-w\big(t,x,y,z\big)\right]\\
-%%     &\qquad+\sum_{k\in\{\lab{S},\lab{E},\lab{I},\lab{R}\}}\mu\,(x^k+1)\,w(t,x+\delta_{k})-\mu\,x^k\,w(t,x)\\
-%%     &\qquad+\upsilon\,x^{\lab{R}}\,w(t,x-(1,0,0,-1))\\
-%%   \end{aligned}
-%% \end{equation}
-
-\section{Discussion}
-
-\bibliographystyle{preprint}
-\bibliography{phylopomp}
-
-\appendix
-\setcounter{equation}{0}
-\titleformat{\section}[hang]{\large\bfseries}{Appendix \periodafter\thesection}{2ex}{\periodafter}{}
-\renewcommand{\theequation}{\thesection\arabic{equation}}
-
-\section{Filter equations}
-\label{sec:filter_eqns}
-
-Explicit expressions for the quantities that arise in this paper are not always readily available.
-Here, we develop tools for manipulating complex expressions that are otherwise cumbersome.
-
-\begin{defn}
-  Suppose $\Xr_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
-  \begin{equation}\label{eq:kfe2}
-    \frac{\partial{u}}{\partial{t}}
-    =\int{u(t,x')\,\beta(t,x',x)\,\dd{x'}}
-    -\int{u(t,x)\,\beta(t,x,x')\,\dd{x'}}.
-  \end{equation}
-  Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)$ are given, real-valued, measurable functions.
-  Let $\mathcal{M}\subseteq\Rp$ be countable such that $\mathcal{M}\cap[0,t]$ is finite for all $t$.
-  We say that the equation
-  \begin{equation}\label{eq:filtereq}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
-      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
-      -\lambda(t,x)\,w(t,x)\\
-      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}.
-    \end{aligned}
-  \end{equation}
-  is the \emph{filter equation} with \emph{driver} $\Xr_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$.
-\end{defn}
-
-\begin{remark}\label{rem:equivfilt}
-  \Cref{eq:filtereq} is equivalent to
-  \begin{equation*}
-    \begin{gathered}
-      \frac{\partial{w}}{\partial{t}}
-      =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
-      -\lambda(t,x)\,w(t,x),\quad t\notin{\mathcal{M}},\\
-      w(t,x)=\int{\leftlim{w}(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}},\quad t\in{\mathcal{M}}.
-    \end{gathered}
-  \end{equation*}
-  Here $\wt$ refers to the left-limit of $w$.
-\end{remark}
-
-\begin{remark}
-  Trivially, a Kolmogorov forward equation is itself a filter equation with boost $1$, decay $0$, and no observed events ($\mathcal{M}=\emptyset$).
-\end{remark}
-
-Filter equations afford a convenient means of computing expectations and likelihoods for pure jump processes.
-This is facilitated by the following fact.
-
-\begin{lemma}\label{lemma:filtereq}
-  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ and $B>0$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satisfies the KFE
-  \begin{equation}\label{eq:filterlemma}
-    \begin{aligned}
-      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
-      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right],
-    \end{aligned}
-  \end{equation}
-  on the space $\Xspace\times(0,\infty)$.
-  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
-\end{lemma}
-\begin{proof}
-  \ELI{[Conditions for differentiating under the integral in the proof?
-      Need $[v^2 \lambda u(t,x,v)]_{v=0}^{\infty} = 0$.
-      That's okay at $v=0$, and a minor constraint on $u$.]}
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
-      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
-      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
-      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
-    \end{aligned}
-  \end{equation*}
-  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
-      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
-      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
-      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
-    \end{aligned}
-  \end{equation*}
-  which is simplified to obtain \cref{eq:filtereq}.
-\end{proof}
-
-\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(\Xr_t,\Vr_t)$.
-In particular, $\Xr_t$ is the driver with KFE \eqref{eq:kfe2}.
-The $\Vr_t$ is \emph{directed} by $\Xr_t$ in the sense that $\Vr$ has jumps wherever $\Xr$ does:
-when $\Xr$ jumps at time $t$ from $x$ to $x'$, $\Vr$ jumps by the multiplicative factor $B(t,x,x')>0$.
-Between jumps, $\Vr_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
-If we view $\Vr_t$ as a weight, then \cref{lemma:filtereq} tells us how the $\Vr_t$-weighted average of $\Xr_t$ evolves in time:
-this average is simply $\int\!{w(t,x)\,\dd{x}}$.
-This motivates the following result, which shows how filter equations allow one to integrate over random histories.
-
-\begin{prop}\label{prop:zeroboost}
-  Suppose $\Xr_t$ is a non-explosive pure jump process with KFE
-  \begin{equation*}
-    \begin{gathered}
-      \frac{\partial{u}}{\partial{t}}(t,x)=\int{u(t,x')\,\alpha(t,x',x)\,\dd{x'}}
-      -\int{u(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
-      u(0,x)=p_0(x).
-    \end{gathered}
-  \end{equation*}
-  Let $\Hr_t$ be its history process.
-  Suppose $\Vr_t$ is an $\Rp$-valued random process such that
-  \begin{equation}\label{eq:Vprod}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)},
-  \end{equation}
-  for some given measurable function $B\ge{0}$.
-  If $w$ satisfies the filter equation,
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{t}}(t,x)
-    =\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
-    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
-  \end{equation*}
-  then $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
-\end{prop}
-\begin{proof}
-  Define $Q(t,x,x')=\Indicator{B(t,x,x')>0}$ and apply \cref{lemma:filtereq} with
-  \begin{equation*}
-    \begin{gathered}
-      \beta(t,x,x')=\alpha(t,x,x')\,Q(t,x,x'),
-      \qquad
-      \lambda=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\dd{x'}}.
-    \end{gathered}
-  \end{equation*}
-  \ELI{[Lemma A1 has $B>0$, which only covers the cases when $Q=1$ and $\lambda=0$.
-      But, $B\ge{0}$ seems enough in the proof of lemma A1, as long as one allows a one-sided Dirac $\delta$ at $0$.
-      Maybe the Dirac $\delta$ should all be one-sided in order to integrate
-      correctly on right-continuous functions? ]}
-\end{proof}
-
-%% An important special case is that of a deterministic driving process.
-%% The following result is established by routine calculation.
-
-%% \begin{prop}
-%%   Suppose $X:[0,T]\to\Xspace$ is a deterministic, piecewise constant, \cadlag\ function and let $M=\event(X)$.
-%%   Then the KFE for $X_t$ is \cref{eq:kfe2} with $\beta(t,x,x')=\sum_{e\in{M}}{\delta(t,e)\delta(x',X_e)}$.
-%%   With this driver, the filter equation (\cref{eq:filtereq}) becomes
-%%   \begin{equation*}
-%%     \begin{gathered}
-%%       \frac{\partial{w}}{\partial{t}}=
-%%       -\lambda(t,x)\,w(t,x)\quad\text{for}\quad{t\in[0,T]\setminus{M}},\\
-%%       w(t,x)=\delta(x,X_t)\,\int{\wt(t,x')\,B(t,x',X_t)\,\dd{x'}}\quad\text{for}\quad{t\in{M}}.
-%%     \end{gathered}
-%%   \end{equation*}
-%% \end{prop}
-
-The following result extends \cref{prop:zeroboost} to the situation where $\mathcal{M}\ne\emptyset$, \ie where the times of some of the events in the history are observed.
-
-\begin{prop}\label{prop:fullfilt}
-  Let $\Xr_t$, $\Hr_t$, $B$, and $\Vr_t$ be as in \cref{prop:zeroboost}.
-  Suppose that $\mathcal{M}\subseteq\halfclosed{0,t}$ is a fixed, finite set of observed event times.
-  Let $\CondExpect{\Vr_t}{\event{\Hr}{t}\supseteq\mathcal{M}}$ denote the expectation of $\Vr_t$ conditional on $\event{\Hr}{t}\supseteq\mathcal{M}$.
-  If $w(t,x)$ satisfies $w(0,x)=p_0(x)$ and the filter equation
-  \begin{equation*}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
-      =&\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
-      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\},
-    \end{aligned}
-  \end{equation*}
-  then $\CondExpect{\Vr_t}{\event{\Hr}{t}\supseteq\mathcal{M}}=\int{w(t,x)\,\dd{x}}$.
-  \ELI{[(a) line 2: density $\rightarrow$ expectation.
-      (b) the term $\delta(t,e)\,w(t,x)$ seems irrelevant:
-      the derivative at a finite set of points only matters when it is a delta function.]}
-\end{prop}
-\begin{proof}
-  Suppose $s\in[0,t]$ and let
-  \begin{equation*}
-    w(s,x)=\Expect{\prod_{e\in\event{\Hr}{s}}{B\left(e,\Xrt_e,\Xr_e\right)}}.
-  \end{equation*}
-  That is, $w(s,x)$ is the expectation (over histories) of the partial products of \cref{eq:Vprod}.
-  If $s\notin{\mathcal{M}}$, then for $\Delta>0$ sufficiently small,
-  \begin{equation*}
-    \begin{aligned}
-      w(s,x)=&\left(1-\Delta\,\int{\alpha(s-\Delta,x,x')\,\dd{x'}}\right)\,w(s-\Delta,x)\\
-      &\qquad+\int{\alpha(s-\Delta,x',x)\,B(s-\Delta,x',x)\,w(s-\Delta,x')\,\dd{x'}}+o(\Delta).
-    \end{aligned}
-  \end{equation*}
-  Taking the limit as $\Delta\downarrow{0}$, we obtain
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{s}}(s,x)
-    =\int{w(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}
-    -\int{w(s,x)\,\alpha(s,x,x')\,\dd{x'}}.
-  \end{equation*}
-  If $s\in\mathcal{M}$, then
-  \begin{equation*}
-    w(s,x)=\int{\wt(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}.
-  \end{equation*}
-  By \cref{rem:equivfilt}, the result follows.
-  \ELI{[The argument shows that the filter equation satisfies a KFE, but the limit expression seems justified by a similar KFE.
-      Is the point that one KFE implies the other?]}
-\end{proof}
-
-\ELI{[Maybe, Prop A3 could be a version of Prop A2 allowing for discrete transitions at a finite set of times;
-    a combination of cts time and discrete time Markov chains.
-    That is all we need to extend Prop A2 to observation times, which can be viewed just as an extended Markov process including the observation events.
-    That is, to make a distinction between ``observation events'' that give the sequences used to construct the phylogeny, and ``observed events'' which are determined points on the phylogeny.
-    The latter are a realization of a random process, and the former are defined to be fixed a priori.]}
-
-%% Filter equations allow us to pass easily between equivalent representations of a process.
-%% For example, an equivalent way of representing $\Xr_t$ is in terms of its embedded chain and event times.
-%% Let $\Xrh_k$ be the embedded chain of $\Xr_t$ and let $\Trh_k$ be the point process of its event times.
-%% It is elementary that
-%% \fontsize{10pt}{12pt}\selectfont
-%% \begin{equation*}
-%%   \begin{gathered}
-%%     \Prob{\Xrh_k=\Xh_k\;\Big\vert\;\Xrh_{k-1}=\Xh_{k-1},\Trh_k=\Th_k}=\frac{\alpha(\Th_k,\Xh_{k-1},\Xh_k)}{\int{\alpha(\Th_k,\Xh_{k-1},x')\dd{x'}}},\\
-%%     \Prob{\Trh_k>\Trh_{k-1}+t\;\Big\vert\;\Xrh_{k-1}=\Xh_{k-1},\Trh_{k-1}=\Th_{k-1}}=\exp{\left(-\int_{0}^{t}{\int{\alpha(\Th_{k-1}+s,\Xh_{k-1},x')\,\dd{x'}}\,\dd{s}}\right)}.
-%%   \end{gathered}
-%% \end{equation*}
-%% \normalfont
-%% Fixing $\nu>0$ and making the definitions,
-%% \begin{equation}\label[pluralequation]{eq:nudefs}
-%%   \begin{gathered}
-%%     A(t,x)=\int{\alpha(t,x,x')\,\dd{x'}}, \qquad
-%%     \pi(t,x,x')=\frac{\alpha(t,x,x')}{A(t,x)},\\
-%%     B(t,x,x')=\frac{A(t,x)}{\nu}, \qquad
-%%     \lambda(t,x)=A(t,x)-\nu,
-%%   \end{gathered}
-%% \end{equation}
-%% we can rewrite the KFE as
-%% \begin{equation}\label{eq:poissdriver}
-%%   \frac{\partial{w}}{\partial{t}}(t,x)=
-%%   \int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
-%%   -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
-%%   -\lambda(t,x)\,w(t,x).
-%% \end{equation}
-%% Here, $\nu$ is the intensity of a time-homogenous Poisson process.
-%% Note that $\pi$ is the probability kernel of the embedded chain $\Xrh$ and $A(t,x)$ is the intensity of the $\Trh_k$ process.
-%% We recognize this equation as the filter equation with boost $B$, decay $\lambda$, and driver generated by $\nu\,\pi(t,x,x')$.
-%% It corresponds to the following procedure for simulating $\Xr_t$:
-%% \begin{compactenum}[(a)]
-%% \item Simulate jump times according to the rate-$\nu$ Poisson process.
-%% \item Simulate the embedded chain $\Xrh_k$ using the kernel $\pi$.
-%% \item Weight the realization by the product of the $B$ factors.
-%%   Note that this makes the appropriate importance-sampling correction.
-%% \end{compactenum}
-
-\end{document}

### 2024-01-21 ms.Rnw
more figures

@@
 \documentclass[10pt,reqno,final]{amsart}
+
 \input{header}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta,positioning,decorations,calc,math}
 
@@
 library(cowplot)
+library(viridis)
 library(phylopomp)
@@
 stopifnot(packageVersion("pomp")>="5.1")
-stopifnot(packageVersion("phylopomp")>="0.9.2")
+stopifnot(packageVersion("phylopomp")>="0.10.1")
 theme_set(theme_bw(base_family="serif"))
@@
 \begin{figure}
-  \begin{center}
-    \resizebox{0.9\linewidth}{!}{
-      \begin{tikzpicture}[scale=1]
-        \colorlet{darkblue}{blue!50!black}
-        \tikzstyle{box}=[draw=black, text=black, fill=white, very thick, minimum size=3em]
-        \tikzstyle{ibox}=[draw=darkblue]
-        \tikzstyle{deme}=[fill=black!20!white]
-        \tikzstyle{label}=[font=\Large]
-        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{flow}=[draw=black, very thick, >=stealth]
-        \tikzstyle{modulate}=[draw=darkblue, >=Circle]
-
-        %% SEIRS model
-        \coordinate (origin) at (0,0);
-        \node[label] (lab) at (origin) {\textbf{A}};
-        \node [box] (S) at ($(origin)+(1,-1)$) {$\lab{S}$};
-        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
-        \node [box,ibox,deme] (I) at ($(E)+(2,0)$) {$\lab{I}$};
-        \node [box] (R) at ($(I)+(2,0)$) {$\lab{R}$};
-        \coordinate (overR) at ($(R)+(0,1)$);
-        \coordinate (midSE) at ($(S)!0.5!(E)$);
-        \draw [flow,->] (S) -- (E);
-        \draw [flow,->] (E) -- (I);
-        \draw [flow,->] (I) -- (R);
-        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
-        \draw [modulate,->] (I.north west) .. controls ($(I)+(-1,1)$) and ($(midSE)+(0,1)$) .. (midSE);
-        \node[font=\normalsize] (demes) at ($(E)!0.5!(I)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I}\}$};
-
-        %% SIIR (two-strain) model
-        \coordinate (origin) at (0,-3);
-        \node[label] (lab) at (origin) {\textbf{B}};
-        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
-        \node [box] (R1) at ($(S)+(6,1)$) {$\lab{R}_1$};
-        \node [box] (R2) at ($(S)+(6,-1)$) {$\lab{R}_2$};
-        \coordinate (overR) at ($(R1)+(0,1)$);
-        \coordinate (underR) at ($(R2)-(0,1)$);
-        \node [box,deme] (E1) at ($(S)+(2,1)$) {$\lab{E}_1$};
-        \node [box,deme] (E2) at ($(S)+(2,-1)$) {$\lab{E}_2$};
-        \node [box,ibox,deme] (I1) at ($(E1)+(2,0)$) {$\lab{I}_1$};
-        \node [box,ibox,deme] (I2) at ($(E2)+(2,0)$) {$\lab{I}_2$};
-        \draw [flow,->] (S) -- (E1);
-        \draw [flow,->] (E1) -- (I1);
-        \draw [flow,->] (I1) -- (R1);
-        \draw [flow,->] (S) -- (E2);
-        \draw [flow,->] (E2) -- (I2);
-        \draw [flow,->] (I2) -- (R2);
-        \draw [flow,->] (R1) -- (overR) -- (S |- overR) -- (S);
-        \draw [flow,->] (R2) -- (underR) -- (S |- underR) -- (S);
-        \coordinate (midSE1) at ($(S)!0.5!(E1)$);
-        \coordinate (midSE2) at ($(S)!0.5!(E2)$);
-        \draw [modulate,->] (I1.north west) .. controls ($(I1)+(-1,1)$) and ($(midSE1)+(0,1.8)$) .. (midSE1);
-        \draw [modulate,->] (I2.south west) .. controls ($(I2)+(-1,-1)$) and ($(midSE2)+(0,-1.8)$) .. (midSE2);
-        \node[font=\normalsize] (demes) at ($(E2)!0.5!(I2)+(0,-2)$) {$\Demes=\{\lab{E}_1,\lab{I}_1,\lab{E}_2,\lab{I}_2\}$};
-
-        %% COVID model
-        \coordinate (origin) at (9,0);
-        \node[label] (lab) at (origin) {\textbf{C}};
-        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
-        \node [box] (R) at ($(S)+(7,1)$) {$\lab{R}$};
-        \coordinate (overR) at ($(R)+(0,1)$);
-        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
-        \node [box,ibox,deme] (Ia) at ($(E)+(2,1)$) {$\lab{I_A}$};
-        \node [box,ibox,deme] (Is) at ($(E)+(2,-1)$) {$\lab{I_S}$};
-        \node [box] (H) at ($(Is)+(2,0)$) {$\lab{H}$};
-        \node [box] (D) at ($(H)+(2,0)$) {$\lab{D}$};
-        \draw [flow,->] (S) -- (E);
-        \draw [flow,->] (E) -- (Ia.west);
-        \draw [flow,->] (E) -- (Is.west);
-        \draw [flow,->] (Ia) -- (R);
-        \draw [flow,->] (Is) -- (H);
-        \draw [flow,->] (H) -- (R);
-        \draw [flow,->] (Is) -- (R);
-        \draw [flow,->] (H) -- (D);
-        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
-        \coordinate (midSE) at ($(S)!0.5!(E)$);
-        \draw [modulate,->] (Ia.north west) .. controls ($(Ia.north west)+(165:1)$) and ($(midSE)+(0,2)$) .. (midSE);
-        \draw [modulate,->] (Is.south west) .. controls ($(Is.south west)+(195:1)$) and ($(midSE)+(0,-2)$) .. (midSE);
-        \node[font=\normalsize] (demes) at ($(Is)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I_A},\lab{I_S}\}$};
-
-        %% SI2R (super-spreader) model
-        \coordinate (origin) at (9,-5);
-        \node[label] (lab) at (origin) {\textbf{D}};
-        \node [box] (S) at ($(origin)+(2,-1)$) {$\lab{S}$};
-        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
-        \node [box,ibox,deme] (Il) at ($(E)+(2,0)$) {$\lab{I_L}$};
-        \node [box,ibox,deme] (Ih) at ($(Il)+(0,-2)$) {$\lab{I_H}$};
-        \node [box] (R) at ($(Il)+(2,0)$) {$\lab{R}$};
-        \coordinate (overR) at ($(R)+(0,1)$);
-        \coordinate (midSE) at ($(S)!0.5!(E)$);
-        \draw [flow,->] (S) -- (E);
-        \draw [flow,->] (E) -- (Il);
-        \draw [flow,<->] (Il) -- (Ih);
-        \draw [flow,->] (Il) -- (R);
-        \draw [flow,->] (Ih.east) -- (R |- Ih.east) -- (R);
-        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
-        \draw [modulate,->] (Il.north west) .. controls ($(Il.north west)+(135:0.8)$) and ($(midSE)+(0,1)$) .. (midSE);
-        \draw [modulate,->] (Ih.west) .. controls ($(Ih.west)+(180:0.8)$) and ($(midSE)+(0,-1)$) .. (midSE);
-        \node[font=\normalsize] (demes) at (midSE |- Ih) {$\Demes=\{\lab{E},\lab{I_L},\lab{I_H}\}$};
-      \end{tikzpicture}
-    }
-  \end{center}
-  \caption{
-    Examples of discretely-structured population models.
-    Demes are shaded.
-    \textbf{(A)} An SEIRS model.
-    Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
-    Upon recovery ($\lab{R}$), individuals experience immunity from reinfection.
-    If this immunity wanes, they re-enter the susceptible compartment.
-    Pathogen lineages are to be found in hosts within the $\lab{E}$ and $\lab{I}$ compartments only.
-    Accordingly, there are two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
-    If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
-    \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
-    \label{fig:examples}
-  }
+  \input{figs/example_models}
 \end{figure}
@@
 \begin{figure}
-  \begin{center}
-    \resizebox{0.9\linewidth}{!}{
-      \begin{tikzpicture}[scale=1]
-        \usetikzlibrary{shapes,arrows,positioning}
-        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum height=8em]
-        \tikzstyle{trans}=[color=black, font=\Large, thick, >=stealth]
-        \node[state] (base) at (0,0) {$x=(S,E,I,R)$};
-        \node[state] (trans) at (165:8) {$x'=(S-1,E+1,I,R)$};
-        \node[state] (prog) at (205:7) {$x'=(S,E-1,I+1,R+1)$};
-        \node[state] (recov) at (305:5) {$x'=(S,E,I-1,R+1)$};
-        \node[state] (wane) at (350:8) {$x'=(S+1,E,I,R-1)$};
-        \node[state] (sample) at (18:7.5) {$x'=(S,E,I,R)$};
-        \draw[trans,->] (base) -- (trans) node[midway,above,sloped] {$\lab{Trans}$};
-        \draw[trans,->] (base) -- (prog) node[midway,above,sloped] {$\lab{Prog}$};
-        \draw[trans,->] (base) -- (recov) node[midway,above,sloped] {$\lab{Recov}$};
-        \draw[trans,->] (base) -- (wane) node[midway,above,sloped] {$\lab{Wane}$};
-        \draw[trans,->] (base) -- (sample) node[midway,above,sloped] {$\lab{Sample}$};
-        \node[font=\LARGE] (U) at (95:3.5) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
-      \end{tikzpicture}
-    }
-  \end{center}
-  \caption{
-    Markov state transition diagram for the SEIRS model depicted in \cref{fig:examples}A.
-    The state, $x$, is characterized by four numbers, $S$, $E$, $I$, and $R$.
-    From a given state $x$, there are five possible kinds of jumps $x\mapsto{x'}$.
-    Accordingly, the set, $\Jumps$, of jump marks has five elements.
-    Each of these is of a different type:
-    $\lab{Trans}$ (transmission) is of birth type,
-    $\lab{Prog}$ (progression) is of migration type,
-    $\lab{Recov}$ (recovery) is of death type,
-    $\lab{Sample}$ (sampling) is of sample type,
-    and $\lab{Wane}$ (loss or waning of immunity) is of neutral type.
-    See \cref{sec:jump_types} for a description of these jump types.
-    Note that, in this formulation, when a sampling event occurs, the state does not change.
-    \label{fig:markov_state}
-  }
+  \input{figs/markov_diagram}
 \end{figure}
@@
 \subsection{Genealogies}
-
-For our purposes, a \emph{genealogy}, $G_{T}$, is a labeled, time-calibrated tree.
+\label{sec:genealogy}
+
+For our purposes, a \emph{genealogy}, $\G_{T}$, is a labeled, time-calibrated tree.
 It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
@@
 Each node $a$ is associated with a specific time, $t(a)$.
-In particular, if $a$ is a tip node in $G_T$, then $t(a)=T$;
+In particular, if $a$ is a tip node in $\G_T$, then $t(a)=T$;
 if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
 Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
-We let $\event{G}{T}$ denote the set of all node-times of the genealogy $G_T$.
+We let $\event{\G}{T}$ denote the set of all node-times of the genealogy $\G_T$.
 Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
@@
 
-Formally, we define a genealogy, $G_T$, to be an ordered pair $(y,z)$.
+Formally, we define a genealogy, $\G_T$, to be an ordered pair $(y,z)$.
 The $z$ element specifies the topological structure, as follows.
-Let $\leaves{G_T}$ be the set of all tip and sample nodes in $G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{G_T}$.
+Let $\leaves{\G_T}$ be the set of all tip and sample nodes in $\G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{\G_T}$.
 The partition \emph{fineness} defines a partial order on $\func{P}$.
 Specifically, for $p,p'\in{\func{P}}$, we say $p\preceq{p'}$ if and only if for every $c'\in{p'}$ there is a unique $c\in{p}$ such that $c\supseteq{c'}$.
-The topological structure of $G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
-
-The $y$ element of $G_T$ specifies the coloring of branches and location of nodes.
-In particular, for $a\in\leaves{G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
+The topological structure of $\G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
+
+The $y$ element of $\G_T$ specifies the coloring of branches and location of nodes.
+In particular, for $a\in\leaves{\G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
 To make this well-defined, let $d_{t}(a)=\eth$ for $t\in{[t(a),T]}$;
 we call $\eth$ the \emph{underdeme} and define $\Demesplus\coloneq{\Demes\cup\{\eth\}}$.
-Thus $d:[0,T]\times\leaves{G_T}\to\Demesplus$ is a well-defined \cadlag\ function.
+Thus $d:[0,T]\times\leaves{\G_T}\to\Demesplus$ is a well-defined \cadlag\ function.
 
 Since not every node along a lineage corresponds to a change of deme, we also count the number of nodes along each lineage.
-Let $m:[0,T]\times\leaves{G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
+Let $m:[0,T]\times\leaves{\G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
 For $t\in{[t(a),T]}$, let $m_{t}(a)=m_{t(a)}(a)$.
@@
 
-Finally, we define the map $y:[0,T]\times\leaves{G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
-The two \cadlag\ functions $y$ and $z$ specify the genealogy $G_T$ completely and uniquely.
-When the genealogy is not clear from context, we write $y^{G}$ and $z^{G}$ to refer to the respective components of genealogy $G$.
+Finally, we define the map $y:[0,T]\times\leaves{\G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
+The two \cadlag\ functions $y$ and $z$ specify the genealogy $\G_T$ completely and uniquely.
+When the genealogy is not clear from context, we write $y^{\G}$ and $z^{\G}$ to refer to the respective components of genealogy $\G$.
 
 Clearly, the topological structure of a tree imposes constraints on the permissible functions $y$ that describe the painting and node placement along it.
-In particular, $y$ must respect the topological structure of $G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
-In other words, if $a,b\in{c}\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
-Let $\Yspace^z$ denote the space of all functions that respect the topological structure described by $z$.
-Let $\Yspace^\z_t$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.
+In particular, $y$ must respect the topological structure of $\G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
+In other words, if $a,b\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
+Let $\Yspace^z$ denote the space of all functions that obey this constraint at every $t$.
+Let $\Yspace^\z_t\subseteq{\Demesplus\times\Zp}$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.
 
@@
 \label{sec:jump_types}
-
-\begin{figure}
-  \begin{center}
-    \resizebox{0.8\linewidth}{!}{
-      \begin{tikzpicture}[scale=2]
-        \usetikzlibrary{shapes,arrows,positioning}
-        \usetikzlibrary{arrows.meta,positioning,decorations}
-        \colorlet{darkblue}{blue!50!black}
-        \colorlet{maize}{yellow!90!black}
-        \colorlet{grey}{black!30!white}
-        \colorlet{darkgreen}{green!40!black}
-        \colorlet{lightblue}{white!30!blue}
-        \tikzstyle{label}=[font=\Large]
-        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
-        \tikzstyle{deme1}=[color=grey, line width=3pt, >={Circle[width=5pt,length=5pt]}]
-        \tikzstyle{deme2}=[color=darkblue, line width=3pt, >={Circle[width=5pt,length=5pt]}]
-        \tikzstyle{deme3}=[color=maize, line width=3pt, >={Circle[width=5pt,length=5pt]}]
-        \tikzstyle{axis}=[color=black, thick, >=stealth]
-        \tikzstyle{timepoint}=[color=red!70!black]
-
-        \def\ysp{0.2} %% spacing between adjacent lineages
-
-        %% Birth-type event
-        \begin{scope}[shift={(0,0)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{A}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
-          \coordinate (temp1) at ($(temp)+(0,\ysp)$);
-          \coordinate (temp2) at ($(temp)-(0,\ysp)$);
-          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
-          \draw[deme3] (mid |- temp) -- (mid |- temp2) -- (right |- temp2);
-          \draw[deme1] (left |- temp) -- (right |- temp);
-          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
-          \foreach \y in {1,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {6,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {2,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Death-type event
-        \begin{scope}[shift={(2,0)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{B}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \foreach \y in {1,2,6} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {3,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
-          \draw[deme3,dashed] (left |- temp) -- (mid |- temp);
-          \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
-          \foreach \y in {5,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Migration-type event
-        \begin{scope}[shift={(4,0)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{C}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \foreach \y in {2,3,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
-          \draw[deme2] (left |- temp) -- (mid |- temp);
-          \draw[deme3] (mid |- temp) -- (right |- temp);
-          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
-          \foreach \y in {1,5} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {6,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Sample event
-        \begin{scope}[shift={(6,0)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{D}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \foreach \y in {2,5} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {1,4,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {3,6,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \coordinate (temp) at ($(origin)+(0,-2*\ysp)$);
-          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
-          \coordinate (temp) at ($(origin)+(0,-6*\ysp)$);
-          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Neutral-type event
-        \begin{scope}[shift={(0,-2)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{E}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \foreach \y in {4,6,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {1,2,5} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {3,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Compound birth/death (Moran) event
-        \begin{scope}[shift={(2,-2)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{F}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
-          \coordinate (temp1) at ($(temp)+(0,\ysp)$);
-          \draw[deme1] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
-          \draw[deme1] (left |- temp) -- (right |- temp);
-          \draw[deme3,dashed] (left |- temp1) -- (mid |- temp1);
-          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
-          \node[text=red,font=\Huge] at (mid |- temp1) {$\times$};
-          \foreach \y in {2,5} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {1,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {6,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Compound sample/death event
-        \begin{scope}[shift={(4,-2)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{G}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \coordinate (temp) at ($(origin)+(0,-5*\ysp)$);
-          \draw[deme3] (left |- temp) -- (mid |- temp);
-          \node[text=lightblue,font=\Huge] at (mid |- temp) {$\bullet$};
-          \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
-          \foreach \y in {1,6} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {2,3,8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {4,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-
-        %% Compound event
-        \begin{scope}[shift={(6,-2)}]
-          \coordinate (origin) at (0,0);
-          \node[label] (lab) at (origin) {\textbf{H}};
-          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
-          \coordinate (right) at ($(left)+(1,0)$);
-          \coordinate (mid) at ($(left)!0.5!(right)$);
-          \coordinate (rlab) at ($(right)+(0.1,0)$);
-          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
-          \coordinate (bright) at (rlab |- bleft);
-          \coordinate (temp) at ($(origin)+(0,-3*\ysp)$);
-          \draw[deme3] (left |- temp) -- (mid |- temp);
-          \draw[deme2] (mid |- temp) -- (right |- temp);
-          \coordinate (temp1) at ($(temp)+(0,-\ysp)$);
-          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
-          \coordinate (temp1) at ($(temp1)+(0,-\ysp)$);
-          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
-          \coordinate (temp1) at ($(temp1)+(0,-\ysp)$);
-          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
-          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
-          \foreach \y in {2,7} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme1] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {1} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme2] (left |- temp) -- (right |- temp);
-          }
-          \foreach \y in {8} {
-            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
-            \draw[deme3] (left |- temp) -- (right |- temp);
-          }
-          \draw[timepoint] (mid) -- (mid |- bleft);
-          \draw[axis,->] (bleft) -- (bright);
-          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
-        \end{scope}
-      \end{tikzpicture}
-    }
-  \end{center}
-  \caption{
-    Jump types differ by their effects on the genealogy.
-    \textbf{(A)} A birth-type jump results in the branching of one or more child lineages from the parent.
-    There can be only one parent, though the child lineages may be in demes different from those of their parent.
-    Here, a parent of the grey deme sires one child lineage in each of the blue and yellow demes.
-    The production of an event is an integer vector, with one entry for each deme.
-    The production of this event is therefore $r=(r_{\lab{grey}},r_{\lab{blue}},r_{\lab{yellow}})=(1,1,1)$.
-    \textbf{(B)} A death-type event causes the extinction of a lineage.
-    Since internal node without children are recursively removed, the affected branch disappears.
-    The production of this event is $r=(0,0,0)$.
-    \textbf{(C)} A migration-type event results in the movement of one or more lineages from one deme to another.
-    Here, one lineage moves from the blue to the yellow deme.
-    The production of this event is $r=(0,0,1)$, \ie the production is 0 for the grey and blue demes, and 1 for the yellow.
-    \textbf{(D)} In a sample-type event, one or more sample nodes (blue circles) are inserted.
-    Here, there are two samples, one in each of the grey and yellow demes.
-    Accordingly, its production is $r=(1,0,1)$.
-    \textbf{(E)} A neutral-type event has no effect on the genealogy and zero production in all demes: $r=(0,0,0)$.
-    \textbf{(F)} The theory presented here allows for compound events.
-    As an example, here a birth/death-type event occurs, wherein one yellow lineage is extinguished and a grey lineage simultaneously sires a grey child.
-    The production of this event is $r=(2,0,0)$.
-    \textbf{(G)} Here, a compound sample/death-type event occurs.
-    A yellow lineage is sampled and simultaneously extinguished.
-    Note that the lineage is not recursively removed, since sample nodes are never removed.
-    \textbf{(H)} A compound birth/migration-type event with yellow parent and production $r=c(0,4,0)$.
-    \label{fig:jump_types}
-  }
-\end{figure}
 
@@
 
-At each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event:
+\begin{figure}
+  \input{figs/event_types}
+\end{figure}
+
+At each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event (\cref{fig:event_types}):
 \begin{compactenum}[(a)]
 \item
-  A birth-type event at time $t$ results in the creation of one new internal node, call it $b$.
+  A birth-type event at time $t$ (\cref{fig:event_types}A) results in the creation of one new internal node, call it $b$.
   A tip node, $a$, of the appropriate deme is chosen with uniform probability from among those present and $b$ is inserted so that its ancestor is that of $a$, while $a$ takes $b$ as its ancestor.
@@
 \item
-  In a death-type event, one or more tip nodes of the appropriate demes are selected with uniform probability from among those present.
+  In a death-type event (\cref{fig:event_types}B), one or more tip nodes of the appropriate demes are selected with uniform probability from among those present.
   These are deleted.
@@
 \item
-  In a migration-type event, the appropriate number of migrating lineages are selected at random, with uniform probability according to deme, from among those present.
+  In a migration-type event (\cref{fig:event_types}C), the appropriate number of migrating lineages are selected at random, with uniform probability according to deme, from among those present.
   For each selected lineage, one new branch node is inserted between the selected tip node and its ancestor.
@@
 \item
-  At a sample-type event, the appropriate number of sampled lineages are selected at random from among the tip nodes, with uniform probability according to deme.
+  At a sample-type event (\cref{fig:event_types}D), the appropriate number of sampled lineages are selected at random from among the tip nodes, with uniform probability according to deme.
   One new sample node is introduced for each selected lineage:
@@
 \item
-  At a neutral-type event, no change is made to the genealogy.
+  At a neutral-type event (\cref{fig:event_types}E), no change is made to the genealogy.
 \item
-  Finally, events of compound type are accommodated by combining the foregoing rules.
+  Finally, events of compound type (\eg \cref{fig:event_types}F--H) are accommodated by combining the foregoing rules.
 \end{compactenum}
@@
 \begin{figure}
+  <<upo,fig.dim=c(6,8),out.width="60%",results="hide">>=
+  freeze(
+    seed=522390503,
+    simulate(
+      "SEIR",
+      Beta=1,sigma=0.5,gamma=0.1,psi=0.4,delta=0.1,
+      S0=10,E0=1,I0=1,R0=0,
+      time=10
+    )
+  ) -> x
+
+  pal <- c("#00274CFF","#FFCB05FF")
+
+  plot_grid(
+    A=x |>
+      plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
+      lims(x=c(NA,10),y=c(3,NA))+
+      geom_vline(xintercept=10,linewidth=0.2,color="black"),
+    B=x |>
+      plot(points=TRUE,prune=TRUE,obscure=FALSE,palette=pal)+
+      lims(x=c(NA,10),y=c(3,NA))+
+      geom_vline(xintercept=10,linewidth=0.2,color="black"),
+    C=x |>
+      plot(points=TRUE,prune=TRUE,obscure=TRUE,palette="#B3B3B3FF")+
+      lims(x=c(NA,10),y=c(3,NA))+
+      geom_vline(xintercept=10,linewidth=0.2,color="black")+
+      labs(x="time"),
+    ncol=1,
+    align="hv",axis="tblr",
+    labels="AUTO"
+  )
+  @
   \caption{
-    Unpruned, pruned, and obscured enealogies.
+    Unpruned, pruned, and obscured genealogies from a single realization of the genealogy process induced by the SEIRS model depicted in \cref{fig:example_models,fig:markov_state}.
+    \textbf{(A)} The full genealogy at time $t=10$.
+    Tip nodes, corresponding to lineages alive at time $t=10$ are indicated with black points.
+    Blue points represent samples;
+    green points, internal nodes.
+    Branches are colored according to the deme in which the corresponding lineage resided at that point in time:
+    blue denotes $\lab{E}$ and yellow, $\lab{I}$.
+    \textbf{(B)} The full genealogy is \emph{pruned} by removing all tip nodes and then recursively pruning away childless internal nodes.
+    Sample nodes are never removed.
+    \textbf{(C)} A genealogy is \emph{obscured} by removing all deme information from lineage histories:
+    the colors are erased, as are extraneous internal nodes.
+    See the text (\cref{sec:genealogy,sec:pruning,sec:obscuration}) for more detail.
+    \label{fig:upo}
   }
@@
 \subsubsection{Pruned genealogy}
+\label{sec:pruning}
 
@@
 Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(y,z)$ is a pruned genealogy, $\ell(y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
-We refer to $\ell$ as the \emph{lineage-count} function.
+We refer to $\ell$ as the \emph{lineage-count} function (cf.~\cref{fig:ells}).
 
@@
 Thus, we can define the \emph{saturation} function such that, whenever $\Pr_T=(\y,\z)$ is a pruned genealogy, $s(\yt_t,\y_t,\zt_t,\z_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
+\Cref{fig:ells} illustrates.
 
 \begin{figure}
-  \caption{
-    Figure showing lineage count and saturation.
-    \label{fig:ells}
-  }
+  \input{figs/ells}
 \end{figure}
@@
 Similarly, if the node in $\P_T$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
-To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q_u(y,y',z,z')=1$ if and only if there is some pruned genealogy $\P_T$ and $t\in{[0,T]}$,
-such that $\yt^{\P_T}_t=y$, $y^{\P_T}_t=y'$, $\zt^{\P_T}_t=z$, $z^{\P_T}_t=z'$,
-and $\P_T$ is compatible with an event of type $u$ at time $t$;
-$Q=0$ otherwise.
+To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $y$ and $z$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
+That is, $Q_u(y,y',z,z')=1$ if and only if
+there is a feasible genealogy, $\G_T$, and history, $\H_T$,
+and a $t\in{[0,T]}$ such that,
+given $\Gr_T=\G_T$ and $\Hr_T=\H_T$,
+$\U_t=u$
+, $\yt^{\prune(\G_T)}_t=y$, $y^{\prune(\G_T)}_t=y'$,
+$\zt^{\prune(\G_T)}_t=z$, and $z^{\prune(\G_T)}_t=z'$.
 We refer to $Q$ as the \emph{compatibility indicator}.
@@
 \subsubsection{Obscured genealogy}
+\label{sec:obscuration}
 
@@
   \begin{equation*}
-    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,y^*_t,\zt^*_t,z^*_t)}
-    \:\times\:\Indicator{\event{\H}{T}\supseteq{\event{\P}{T}^*}}.
+    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\Indicator{\event{\H}{T}\supseteq{\event{\P}{T}^*}}\,\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,y^*_t,\zt^*_t,z^*_t)}.
   \end{equation*}
@@
 For this, we use the filter equation technology developed in \cref{sec:filter_eqns}.
-In particular, the following theorem follows immediately from \cref{prop:fullfilt}.
+In particular, the following theorem follows immediately from \cref{lemma:sing-filt}.
 
@@
     \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
-    \Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}}
-    \times\Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*},
+    \Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
   \end{equation*}
@@
 The final result shows how to compute the likelihood of an obscured genealogy.
-It is an immediate consequence of \cref{thm:obsc_lik,prop:fullfilt}.
+It is an immediate consequence of \cref{thm:obsc_lik,lemma:sing-filt}.
 
@@
   Let $z^*$, be a given obscured genealogy.
-  Let $\pi$ and $\yr_t$ be as in \cref{thm:obsc_lik}.
+  Let $\pi$, $q$, and $\yr_t$ be as in \cref{thm:obsc_lik}.
   For $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $u\in\Jumps$, define
@@
   \begin{equation}
-    \label{eq:filtereq}
+    \label{eq:reg-filter-eq}
     \begin{aligned}
@@
 \begin{lemma}
+  \label{lemma:reg-filt}
   Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable.
@@
   \end{equation*}
-  then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued deterministic process on $\Xspace$ with density $w(t,x)$.
-  Moreover, $w$ satisfies the filter equation
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{t}}=\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}
-  \end{equation*}
-  and $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
+  then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued deterministic process on $\Xspace$ with density $w(t,x)$ that satisfies the filter equation
+  \begin{equation}
+    \label{eq:reg-filter-eq}
+    \frac{\partial{w}}{\partial{t}}=\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
+  \end{equation}
+  Moreover, $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
 \end{lemma}
@@
   \end{equation*}
-  Taking the limit as $\Delta\downarrow{0}$, we obtain
-  \begin{equation*}
-    \frac{\partial{w}}{\partial{t}}(t,x)
-    =\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
-    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
-  \end{equation*}
+  In the limit, we obtain \cref{eq:reg-filter-eq}.
+  The last statement follows by definition.
 \end{proof}
@@
 \begin{lemma}
+  \label{lemma:sing-filt}
   Suppose $\mathcal{M}\subset{\Rp}$ is finite and let $\mathcal{M}_t\coloneq\mathcal{M}\cap{[0,t]}$.
@@
   \end{equation*}
-  Then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued process on $\Rp^{\vert\mathcal{M}_t\vert}\times\Xspace$ with density $w(t,x)$.
-  Furthermore, $w(t,x)$ satisfies
-  \begin{equation*}
+  Then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued process on $\Rp^{\vert\mathcal{M}_t\vert}\times\Xspace$ with density $w(t,x)$ that satisfies
+  \begin{equation}
+    \label{eq:two-part-filter}
     \begin{aligned}
@@
       w(t,x)&=\int{\wt(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
-      &t\in{\mathcal{M}}
-    \end{aligned}
-  \end{equation*}
-  and $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
+      &t\in{\mathcal{M}}.
+    \end{aligned}
+  \end{equation}
+  Moreover, $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
 \end{lemma}
 
+\begin{lemma}
+  Showing how \cref{eq:two-part-filter} can be stated as a singular filter equation.
+  \begin{equation}
+    \label{eq:sing-filter-eq}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}(t,x)
+      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
+      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}\\
+      &+\sum_e\int{w(t,x')\,\delta(t,e)\,\pi(t,x',x)\,C(t,x',x)\,\dd{x'}}
+      -\sum_e\int{w(t,x)\,\delta(t,e)\,\pi(t,x,x')\,\dd{x'}}\\
+      &-\lambda(t,x)\,w(t,x).\\
+    \end{aligned}
+  \end{equation}
+\end{lemma}
+
 Filter equations afford a convenient means of computing expectations and likelihoods for pure jump processes.
-This is facilitated by the following Lemma, the statement of which relies on a non-standard Dirac delta function.
-Let $\delta(v,v')$ be the right-sided Dirac delta function satisfying $\delta(v,v')=0$ for $v\ne{v'}$ and $\int_a^b{f(v)\,\delta(v,v')\,\dd{v}}=f(v')\,\Indicator{v'\in\halfopen{a,b}}$ for all \cadlag\ $f$.
+This is facilitated by the following Lemma, the statement of which uses a certain non-standard Dirac delta function.
+Specifically, let $\delta(v,v')$ be the right-sided Dirac delta function satisfying
+\begin{equation*}
+  \begin{gathered}
+    \delta(v,v')=0 \qquad \text{for}\ v\ne{v'}\ \text{and}\\
+    \int_a^b{f(v)\,\delta(v,v')\,\dd{v}}=f(v')\,\Indicator{v'\in\halfopen{a,b}},
+    \qquad \text{whenever}\ f \ \text{is \cadlag\ and}\ -\infty\le{a}<{b}\le{\infty}.
+  \end{gathered}
+\end{equation*}
 
@@
   \label{lemma:filter}
-  The filter equation \eqref{eq:filtereq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satsfies the KFE
+  The filter equation \eqref{eq:reg-filter-eq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satsfies the KFE
   \begin{equation}
@@
     \begin{aligned}
-      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
-      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right],
-    \end{aligned}
-  \end{equation}
-  on the space $\Xspace\times\Rp$.
+      \frac{\partial{u}}{\partial{t}}=
+      &\int_0^{\infty}\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
+      &\qquad-\int_0^{\infty}\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
+      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right].
+    \end{aligned}
+  \end{equation}
 \end{lemma}
 \begin{proof}
-  \ELI{[Conditions for differentiating under the integral in the proof?
-      Need $[v^2 \lambda u(t,x,v)]_{v=0}^{\infty} = 0$.
-      That's okay at $v=0$, and a minor constraint on $u$.]}
   \begin{equation*}
@@
       \frac{\partial{w}}{\partial{t}}
-      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
-      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
-      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
-    \end{aligned}
-  \end{equation*}
-  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
+      =&\int_0^{\infty}{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
+      =&\int_0^{\infty}\int\int_0^{\infty}{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
+      &\qquad-\int_0^{\infty}\int\int_0^{\infty}{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
+      &\qquad+\int_0^{\infty}{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
+    \end{aligned}
+  \end{equation*}
+  Here, the non-explosivity assumption guarantees that we can differentiate under the integral sign and exchange the order of integration.
+  Moreover, it ensures that $u\to{0}$ as $v\to{\infty}$.
+  Hence, by evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
   \begin{equation*}
@@
   \end{equation*}
-  which is simplified to obtain \cref{eq:filtereq}.
+  which is simplified to obtain \cref{eq:reg-filter-eq}.
 \end{proof}

### 2024-01-21 ms.R

@@ -3,10 +3,44 @@ library(tidyverse)
 library(ggtree)
 library(pomp)
 library(cowplot)
+library(viridis)
 library(phylopomp)
 stopifnot(getRversion() >= "4.3")
 stopifnot(packageVersion("pomp")>="5.1")
-stopifnot(packageVersion("phylopomp")>="0.9.2")
+stopifnot(packageVersion("phylopomp")>="0.10.1")
 theme_set(theme_bw(base_family="serif"))
 set.seed(1159254136)
+
+
+## ----upo,fig.dim=c(6,8),out.width="60%",results="hide"------------------------
+freeze(
+  seed=522390503,
+  simulate(
+    "SEIR",
+    Beta=1,sigma=0.5,gamma=0.1,psi=0.4,delta=0.1,
+    S0=10,E0=1,I0=1,R0=0,
+    time=10
+  )
+) -> x
+
+pal <- c("#00274CFF","#FFCB05FF")
+
+plot_grid(
+  A=x |>
+    plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
+    lims(x=c(NA,10),y=c(3,NA))+
+    geom_vline(xintercept=10,linewidth=0.2,color="black"),
+  B=x |>
+    plot(points=TRUE,prune=TRUE,obscure=FALSE,palette=pal)+
+    lims(x=c(NA,10),y=c(3,NA))+
+    geom_vline(xintercept=10,linewidth=0.2,color="black"),
+  C=x |>
+    plot(points=TRUE,prune=TRUE,obscure=TRUE,palette="#B3B3B3FF")+
+    lims(x=c(NA,10),y=c(3,NA))+
+    geom_vline(xintercept=10,linewidth=0.2,color="black")+
+    labs(x="time"),
+  ncol=1,
+  align="hv",axis="tblr",
+  labels="AUTO"
+)
 

### 2024-01-21 setup.R

@@ -1,51 +1,59 @@
-## ----setup--------------------------------------------------
 library(knitr)
+if (!exists("prefix")) prefix <- ""
 opts_chunk$set(
-  cache=TRUE,
-  cache.path=paste0("tmp/",prefix,"/"),
-  comment=NA,
-  echo=FALSE,
-  eval=TRUE,
-  include=TRUE,
-  dev="png",
-  dev.args=list(bg='transparent'),
-  dpi=300,
-  error=FALSE,
-  fig.align='center',
-  fig.dim=c(4,6.83),
-  fig.lp="fig:",
-  fig.path=paste0("tmp/",prefix,"/"),
-  fig.pos="h!",
-  fig.show='asis',
-  highlight=TRUE,
-  message=FALSE,
-  progress=TRUE,
-  prompt=FALSE,
-  purl=TRUE,
-  results="markup",
-  size='small',
-  strip.white=TRUE,
-  tidy=FALSE,
-  warning=FALSE
-  )
-
-knit_hooks$set(
-  document = function (x) {
-    sub('\\usepackage[]{color}','\\usepackage{xcolor}',x,fixed=TRUE)
-  }
-)
+             cache=TRUE,
+             cache.extra=rand_seed,
+             cache.path=paste0("tmp/",as.character(prefix),"/cache/"),
+             progress=TRUE,
+             prompt=FALSE,
+             tidy=FALSE,
+             strip.white=TRUE,
+             message=FALSE,
+             warning=FALSE,
+             error=FALSE,
+             echo=FALSE,
+             results="hide",
+             fig.show="asis",
+             size="small",
+             fig.lp="fig:",
+             fig.path=paste0("tmp/",as.character(prefix),"/figure/"),
+             fig.align="center",
+             out.width="80%",
+             fig.dim=c(6.83,4),
+             dpi=300,
+             dev="pdf",
+             dev.args=list(bg="transparent")
+           )
 
 options(
-  width=60,
+  width=150,
   keep.source=TRUE,
   encoding="UTF-8",
-  dplyr.summarise.inform=FALSE
+  stringsAsFactors=FALSE,
+  dplyr.summarise.inform=FALSE,
+  pomp_archive_dir="results"
 )
 
-registerS3method(
-  "knit_print",
-  "data.frame",
-  function (x, ...) {
-    print(x,row.names=FALSE)
+maize <- "#ffcb05"
+blue <- "#00274c"
+
[…]

### 2024-01-24 ms.Rnw
much more in talk

@@
 \begin{thm}\label{thm:pruned_lik}
-  Let $\P^*_T=(y^*,z^*)$  be a given pruned genealogy and
+  Let $\P^*_T=(\y^*,\z^*)$  be a given pruned genealogy and
   \begin{equation}\label{eq:phidef}
@@
   \begin{equation*}
-    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\Indicator{\event{\H}{T}\supseteq{\event{\P}{T}^*}}\,\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,y^*_t,\zt^*_t,z^*_t)}.
+    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\Indicator{\event{\H}{T}\supseteq{\event{\P}{T}^*}}\,\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}.
   \end{equation*}
@@
   Conditional on $\Hr_T=\H_T$, at each time $t\in\event{\H}{T}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
-  In $\P^*_T$, at time $t$, there are $\ell_i=\ell_i(y^*_t,z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,y^*_t,\zt^*_t,z^*_t)$ are emergent.
+  In $\P^*_T$, at time $t$, there are $\ell_i=\ell_i(\y^*_t,\z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)$ are emergent.
   By assumption, at each genealogical event, lineages within a deme are exchangeable:
@@
   \begin{equation*}
-    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\BinRatio{n(\X_t)}{\ell(y^*_t,z^*_t)}{r^{\U_t}}{s(\yt^*_t,y^*_t,\zt^*_t,z^*_t)}}.
+    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\BinRatio{n(\X_t)}{\ell(\y^*_t,\z^*_t)}{r^{\U_t}}{s(\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}}.
   \end{equation*}
@@
       \frac{\partial{w}}{\partial{t}}(t,x)=
-      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}
+      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)\,\dd{x'}}}
       -\sum_u{\int{w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}}}\\
-      &\quad+\sum_{e\in\event{\P}{T}^*}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}-w(t,x)\right\}},
+      &\quad+\sum_{e\in\event{\P}{T}^*}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)\,\dd{x'}}}-w(t,x)\right\}},
     \end{aligned}
@@
 \begin{thm}\label{thm:obsc_lik}
-  Let $z^*=\obs(\P^*_T)$ be a given obscured genealogy.
-  Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{z^*}}{q(x,y)}=1$.
-  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,z^*_t)=1$
-  and that $q(x,y)>0$ whenever $\Prob{\Xr_0=x,y^{\Pr_T}_0=y}>0$.
-  Then there is a Markov jump process $\yr_t$ on $\Yspace_{z^*}$ such that
+  Let $\z^*=\obs(\P^*_T)$ be a given obscured genealogy.
+  Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{\z^*}}{q(x,y)}=1$.
+  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,\z^*_t)=1$
+  and that $q(x,y)>0$ whenever $\Prob{\Xr_0=x,\y^{\Pr_T}_0=y}>0$.
+  Then there is a Markov jump process $\yr_t$ on $\Yspace_{\z^*}$ such that
   \begin{equation*}
-    \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
+    \CondProb{\zr_T=\z^*}{\Hr_T=\H_T}=
     \Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
@@
     \begin{gathered}
-      \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y',\zt^*_t,z^*_t)}{\pi_u(t,x,x',y,y')},\\[2ex]
-      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),
+      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),\\[2ex]
+      \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y',\zt^*_t,\z^*_t)}{\pi_u(t,x,x',y,y')},
     \end{gathered}
@@
   with the initial condition $w(0,x,y)=p_0(x)\,q(x,y)$.
-  Then the likelihood of $V^*_T$ is
+  Then the likelihood of $\z^*_T$ is
   \begin{equation*}
-    \lik(V^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
+    \lik(\z^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
   \end{equation*}

### 2024-01-24 ms.R

@@ -1,46 +1,0 @@
-## ----packages,include=FALSE,cache=FALSE---------------------------------------
-library(tidyverse)
-library(ggtree)
-library(pomp)
-library(cowplot)
-library(viridis)
-library(phylopomp)
-stopifnot(getRversion() >= "4.3")
-stopifnot(packageVersion("pomp")>="5.1")
-stopifnot(packageVersion("phylopomp")>="0.10.1")
-theme_set(theme_bw(base_family="serif"))
-set.seed(1159254136)
-
-
-## ----upo,fig.dim=c(6,8),out.width="60%",results="hide"------------------------
-freeze(
-  seed=522390503,
-  simulate(
-    "SEIR",
-    Beta=1,sigma=0.5,gamma=0.1,psi=0.4,delta=0.1,
-    S0=10,E0=1,I0=1,R0=0,
-    time=10
-  )
-) -> x
-
-pal <- c("#00274CFF","#FFCB05FF")
-
-plot_grid(
-  A=x |>
-    plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
-    lims(x=c(NA,10),y=c(3,NA))+
-    geom_vline(xintercept=10,linewidth=0.2,color="black"),
-  B=x |>
-    plot(points=TRUE,prune=TRUE,obscure=FALSE,palette=pal)+
-    lims(x=c(NA,10),y=c(3,NA))+
-    geom_vline(xintercept=10,linewidth=0.2,color="black"),
-  C=x |>
-    plot(points=TRUE,prune=TRUE,obscure=TRUE,palette="#B3B3B3FF")+
-    lims(x=c(NA,10),y=c(3,NA))+
-    geom_vline(xintercept=10,linewidth=0.2,color="black")+
-    labs(x="time"),
-  ncol=1,
-  align="hv",axis="tblr",
-  labels="AUTO"
-)
-

### 2024-01-27 ms.Rnw
changes to genealogy details

@@
 
-\subsubsection{Jump marks}
+\subsection{Jump marks}
 
-In the following, it will be useful to break the jumps into distinct categories.
+In the following, it will be useful to break the jumps of the population process $\Xr_t$ into distinct categories.
 For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
@@
 As usual, we take the sample paths of $\Ur_t$ to be \cadlag.
-Observe that $(\Xr_t,\Ur_t)$ is a Markov process, though $\Ur_t$ is not.
+Observe that, though $\Xr_t$ and $(\Xr_t,\Ur_t)$ are Markov processes, $\Ur_t$ is not.
 
-\subsection{Examples}
+\subsection{History}
 
-\subsubsection{SEIRS model}
-
-\subsubsection{SIIR model}
-
-\subsubsection{Linear birth-death model}
-
-\subsubsection{Moran model and the Kingman coalescent}
-
-\subsection{History process}
+\AAK{[Maybe it would be better to speak instead of $\Hr_t$ being the $\sigma$-algebra generated by the sequences $(\Trh_k,\Xrh_k,\Urh_k)_{k=1}^{\Kr_t}$.
+    The only thing we do with $\Hr$ is condition on it.
+    One (perhaps minor) problem with this is that we cannot then speak about the measure on histories.
+]}
 
@@
 
+\subsection{Examples}
+
+\subsubsection{SEIRS model}
+
+\subsubsection{SIIR model}
+
+\subsubsection{Linear birth-death model}
+
+\subsubsection{Moran model and the Kingman coalescent}
+
 \section{The induced genealogy process}
@@
 It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
-There are three distinct kinds of nodes:
+A genealogy includes three distinct kinds of nodes:
 \begin{inparaenum}[(i)]
@@
 
-Formally, we define a genealogy, $\G_T$, to be an ordered pair $(y,z)$.
-The $z$ element specifies the topological structure, as follows.
-Let $\leaves{\G_T}$ be the set of all tip and sample nodes in $\G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{\G_T}$.
-The partition \emph{fineness} defines a partial order on $\func{P}$.
-Specifically, for $p,p'\in{\func{P}}$, we say $p\preceq{p'}$ if and only if for every $c'\in{p'}$ there is a unique $c\in{p}$ such that $c\supseteq{c'}$.
-The topological structure of $\G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
+Formally, we define a genealogy, $\G_T$, to be an ordered pair $(\y,\z)$.
+The $\z$ element specifies the genealogy's tree structure, as follows.
+Let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all partitions of subsets of $\leaves$.
+Partition \emph{fineness} defines a partial order on $\func{Z}$.
+Specifically, for $\zeta,\zeta'\in{\func{Z}}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
+The tree structure of $\G_T$ is defined by a \cadlag\ map $\z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
+An element $b\in\z_t$ is a set of labels;
+it represents the branch of the tree that bears the corresponding lineages.
 
-The $y$ element of $\G_T$ specifies the coloring of branches and location of nodes.
-In particular, for $a\in\leaves{\G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
-To make this well-defined, let $d_{t}(a)=\eth$ for $t\in{[t(a),T]}$;
-we call $\eth$ the \emph{underdeme} and define $\Demesplus\coloneq{\Demes\cup\{\eth\}}$.
-Thus $d:[0,T]\times\leaves{\G_T}\to\Demesplus$ is a well-defined \cadlag\ function.
+The $\y$ element of $\G_T$ specifies the coloring of branches and location of nodes.
+In particular, if $a$ is the label of any tip or sample node and $t\in[0,T]$, we define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$ and think of this as the color of the corresponding branch at that time.
+We will also need to count the number of nodes along each lineage:
+let $m_{t}(a)$ be the number of internal or sample nodes encountered as one descends along lineage $a$ from time $t=t(a)$ to $t=0$.
+Observe that $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
+Formally, we define the coloring component $\y$ as follows.
+Let $\Yspace=\Demes\times\Zp$.
+For $\eta\in\Yspace$, write $\eta=(\eta^d,\eta^m)$ to denote the two components.
+Then, for $\zeta\in\func{Z}$, each $\eta\in\Yspace^{\zeta}$ is a map from $\zeta$ into $\Yspace$.
+Now, given a tree $\z:[0,T]\to\func{Z}$ and a map $\y:[0,T]\to{\coprod_{t\in[0,T]}{\Yspace^{\z_t}}}$,
+we say that $\y$ is \emph{compatible} with $\z$ if,
+for every $t\in[0,T]$, $\y_t\in{\Yspace^{\z_t}}$,
+$\y$ is \cadlag\ and,
+whenever $t_1,t_2\in[0,T]$, $b_1\in\z_{t_1}$, $b_2\in\z_{t_2}$
+such that $t_1\le{t_2}$ and $b_1\supseteq{b_2}$,
+we have $y^m_{t_1}\le{y^m_{t_2}}$.
+Let $\func{Y}(\z)$ denote the set of such maps compatible with $\z$.
+\AAK{[The coproduct is a fiber bundle over the space of trees and $\func{Y}(\z)$ is a section of this bundle.]}
 
-Since not every node along a lineage corresponds to a change of deme, we also count the number of nodes along each lineage.
-Let $m:[0,T]\times\leaves{\G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
-For $t\in{[t(a),T]}$, let $m_{t}(a)=m_{t(a)}(a)$.
-Clearly, $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
+The two \cadlag\ functions $\y$ and $\z$ specify a genealogy completely and uniquely.
+When the genealogy is not clear from context, we write $\G^{\y}$ and $\G^{\z}$ to refer to the respective components of genealogy $\G$.
 
-Finally, we define the map $y:[0,T]\times\leaves{\G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
-The two \cadlag\ functions $y$ and $z$ specify the genealogy $\G_T$ completely and uniquely.
-When the genealogy is not clear from context, we write $y^{\G}$ and $z^{\G}$ to refer to the respective components of genealogy $\G$.
-
-Clearly, the topological structure of a tree imposes constraints on the permissible functions $y$ that describe the painting and node placement along it.
-In particular, $y$ must respect the topological structure of $\G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
-In other words, if $a,b\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
-Let $\Yspace^z$ denote the space of all functions that obey this constraint at every $t$.
-Let $\Yspace^\z_t\subseteq{\Demesplus\times\Zp}$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.
-
-\subsection{Jump types}
-\label{sec:jump_types}
+\subsection{Event types}
+\label{sec:event_types}
 
 We will show how a given population process naturally induces a process in the space of genealogies.
-Because different kinds of jumps have different effects on the genealogy, we need a way of distinguishing them.
-In particular, there are five distinct \emph{pure types} of jumps:
+Specifically, we will describe how, at each jump in the population process, a corresponding change occurs in the genealogy, according to lineages branch, die, or move between demes and whether samples are taken.
+For this purpose, there are five distinct \emph{pure types} of events:
 \begin{compactenum}[(a)]
-\item \emph{Birth-type jumps} result in the branching of one or more new lineages, each from some existing lineages.
+\item \emph{Birth-type events} result in the branching of one or more new lineages, each from some existing lineages.
   Examples of birth-type events include transmission events, speciations, and actual births.
   For simplicity, we assume that all new lineages arising from a birth event share the same parent.
-\item \emph{Death-type jumps} result in the extinction of one or more lineages.
+\item \emph{Death-type events} result in the extinction of one or more lineages.
   Examples include recovery from infection, death of a host, and species extinctions.
   We allow for the possibility that multiple lineages die simultaneously.
-\item \emph{Migration-type jumps} result in the movement of a lineage from one deme to another.
+\item \emph{Migration-type events} result in the movement of a lineage from one deme to another.
   Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
   We permit multiple movements to occur simultaneously.
-\item \emph{Sample-type jumps} result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
+\item \emph{Sample-type events} result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
   We allow for the possibility that multiple samples are collected in a single sample event.
   In this case, we assume that, in each sample-type event, each extant lineage is sampled at most once.
-\item \emph{Neutral-type jumps} result in no change to any of the lineages.
+\item \emph{Neutral-type events} result in no change to any of the lineages.
 \end{compactenum}
-\Cref{fig:markov_state} depicts an example with all five of the pure types.
-It is not necessary that a jump be of a pure type.
-Compound jumps partake of more than one type.
-For example, a sample/death-type event, in which a lineage is simultaneously sampled and removed, has been proposed \citep{Leventhal2014}, as have birth/death events in which one lineage reproduces at the same moment that another dies (\eg the Moran process \citep{Moran1958}).
-The theory presented here places few restrictions on the complexity of the events that can occur.
+\Cref{fig:markov_state} depicts an example with jumps of all five pure types.
+It is not necessary that an event be of a pure type;
+\emph{compound events} partake of more than one type.
+For example, a sample/death-type event, in which a lineage is simultaneously sampled and removed, has been proposed \citep{Leventhal2014}, as have birth/death events in which one lineage reproduces at the same moment that another dies (\eg the \citet{Moran1958} process).
+The theory presented here places few restrictions on the complexity of the events that can occur by combining events of the various pure types.
 
@@
 We now show how a given population process induces a stochastic process, $\Gen_t$, on the space of genealogies.
-In the case of unstructured population processes (\ie those having a single deme), \citet{King2022} gave a construction.
-Although we now treat a more general case, the ideas are much the same.
-Readers wishing more detail on the construction here should consult the earlier paper.
+In the case of unstructured population processes (\ie those having a single deme), \citet{King2022} gave a related construction that is equivalent to the one presented here.
 
@@
 
-At each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event (\cref{fig:event_types}):
+At each jump in the population process, a change is made to the genealogy, according to the mark, $u$, of the jump (\cref{fig:event_types}).
+In particular:
 \begin{compactenum}[(a)]
 \item
-  A birth-type event at time $t$ (\cref{fig:event_types}A) results in the creation of one new internal node, call it $b$.
+  If $u$ is of birth-type (\cref{fig:event_types}A), it results in the creation of one new internal node, call it $b$.
   A tip node, $a$, of the appropriate deme is chosen with uniform probability from among those present and $b$ is inserted so that its ancestor is that of $a$, while $a$ takes $b$ as its ancestor.
@@
 \item
-  In a death-type event (\cref{fig:event_types}B), one or more tip nodes of the appropriate demes are selected with uniform probability from among those present.
+  If $u$ is of death-type (\cref{fig:event_types}B), one or more tip nodes of the appropriate demes are selected with uniform probability from among those present.
   These are deleted.
@@
 \item
-  In a migration-type event (\cref{fig:event_types}C), the appropriate number of migrating lineages are selected at random, with uniform probability according to deme, from among those present.
+  At a migration-type event (\cref{fig:event_types}C), the appropriate number of migrating lineages are selected at random with uniform probability, from among those present in the appropriate demes.
   For each selected lineage, one new branch node is inserted between the selected tip node and its ancestor.
@@
 \end{compactenum}
-In each of these events, the new node or nodes that are introduced are situated at time $t$.
+In each of these events, the new node or nodes that are introduced have node-times equal to the time of the jump.
 
@@
 We require that $r^u_i$ be a constant, for each $u$ and $i$.
-Since one is free to expand the set of jump-marks $\Jumps$ as needed, this is not an important restriction on the models that the theory can accommodate.
+Since, in applications, one is free to expand the set of jump-marks $\Jumps$ as needed, this is not an important restriction on the models that the theory can accommodate.
 Thus there is a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
@@
 it retains information about where among the demes each of its lineages was through time.
-Note also that a pruned genealogy $\Pr_t$ is characterized by functions $y^{\Pr_t}$ and $z^{\Pr_t}$ just as an un-pruned genealogy is.
+Note also that a pruned genealogy $\Pr_t$ is characterized by functions $\Pr_t^\y$ and $\Pr_t^\z$ just as an un-pruned genealogy is.
 
@@
 Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell=(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
-Clearly, $\ell$ depends only on the values, $(y_t,\z_t)$, of $y$ and $z$ at time $t$.
-Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(y,z)$ is a pruned genealogy, $\ell(y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
+Clearly, $\ell$ depends only on the values, $(\y_t,\z_t)$, of $y$ and $z$ at time $t$.
+Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(\y,\z)$ is a pruned genealogy, $\ell(\y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
 We refer to $\ell$ as the \emph{lineage-count} function (cf.~\cref{fig:ells}).
@@
 We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
-In particular, given a node time $t$ in a pruned genealogy $\Pr_T=(y,z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
+In particular, given a node time $t$ in a pruned genealogy $\Pr_T=(\y,\z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
 Like the lineage-count function, $s$ depends only on the local structure of $\Pr_T$.
@@
 To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $y$ and $z$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
-That is, $Q_u(y,y',z,z')=1$ if and only if
-there is a feasible genealogy, $\G_T$, and history, $\H_T$,
+That is, $Q_u(\eta,\eta',\zeta,\zeta')=1$ if and only if
+there is a feasible genealogy, $\G_T=(\y,\z)$, and history, $\H_T$,
 and a $t\in{[0,T]}$ such that,
 given $\Gr_T=\G_T$ and $\Hr_T=\H_T$,
-$\U_t=u$
-, $\yt^{\prune(\G_T)}_t=y$, $y^{\prune(\G_T)}_t=y'$,
-$\zt^{\prune(\G_T)}_t=z$, and $z^{\prune(\G_T)}_t=z'$.
+$\U_t=u$,
+$\yt_t=\eta$, $\y_t=\eta'$,
+$\zt_t=\zeta$, and $\z_t=\zeta'$.
 We refer to $Q$ as the \emph{compatibility indicator}.
@@
 The \emph{obscured genealogy} is obtained by discarding all information about demes and events not visible from the topology of the tree alone.
-In particular, if $\Pr_T=(y,z)$ is a pruned genealogy, we write $\obs(\Pr_T)=z$ to denote the obscured genealogy.
+In particular, if $\Pr_T=(\y,\z)$ is a pruned genealogy, we write $\obs(\Pr_T)=\Pr_T^{\z}=\z$ to denote the obscured genealogy.
 
@@
 \begin{thm}\label{thm:obsc_lik}
+  \AAK{[This can be rephrased somewhat more compactly into a result claiming the existence of the Markov process $\y$.]}
   Let $\z^*=\obs(\P^*_T)$ be a given obscured genealogy.
@@
 \begin{thm}
+  \AAK{[This can be rephrased into a claim about the existence of the kernels $\pi$ and $q$.]}
   Let $z^*$, be a given obscured genealogy.

### 2024-01-27 ms.Rnw
more work

@@
 Large-population, small sample-size approximations.
+
+For structured but deterministic models, approach of \citet{Volz2012,Rasmussen2014a}, based on approximation as birth-death processes:
+approximation no longer needed.
 
@@
 Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
-We let $\event{\G}{T}$ denote the set of all node-times of the genealogy $\G_T$.
+We let $\event{\G}{T}$ denote the set of all internal and sample node-times of the genealogy $\G_T$;
+we refer to these as \emph{genealogical event times}.
 Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
@@
 We will show how a given population process naturally induces a process in the space of genealogies.
-Specifically, we will describe how, at each jump in the population process, a corresponding change occurs in the genealogy, according to lineages branch, die, or move between demes and whether samples are taken.
+Specifically, we will describe how, at each jump in the population process, a corresponding change occurs in the genealogy, according to whether lineages branch, die, move between demes, or are sampled.
 For this purpose, there are five distinct \emph{pure types} of events:
@@
   Examples of birth-type events include transmission events, speciations, and actual births.
-  For simplicity, we assume that all new lineages arising from a birth event share the same parent.
+  Importantly, we assume that all new lineages arising from a birth event share the same parent and that  at most one birth event occurs at a time, almost surely.
 \item \emph{Death-type events} result in the extinction of one or more lineages.
@@
   Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
-  We permit multiple movements to occur simultaneously.
+  We permit multiple lineages to move simultaneously.
 \item \emph{Sample-type events} result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
-  We allow for the possibility that multiple samples are collected in a single sample event.
-  In this case, we assume that, in each sample-type event, each extant lineage is sampled at most once.
+  We allow for the possibility that multiple samples are collected simultaneously, though we require that, in this case, each extant lineage is sampled at most once.
 \item \emph{Neutral-type events} result in no change to any of the lineages.
@@
 We say $r^u=(r^u_i)_{i\in\Demes}$ is the \emph{production} of an event of mark $u$.
-Note that the lineages that die as a result of an event do not count in the production but that all parent lineages, if they survive the event, are counted in the production.
+Note that the lineages that die as a result of an event do not count in the production but that a parent lineage that survives the event does count.
 
@@
 We view $\Grh_k$ as the embedded chain of the continuous-time genealogy process $\Gen_t$.
-It is very important to note that, conditional on $\Urh_k$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of the choices at any other times and independent of $\Xrh_k$ for all $k$.
+It is very important to note that, conditional on $\Urh_k$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of the choices at any other times and independent of $(\Trh_k,\Xrh_k)$ for all $k$.
 Moreover, by construction, any lineage within a deme is as likely as any other lineage in that deme to be selected as a parent or for death, sampling, or migration.
@@
 Moreover, it details each lineage's complete history of movement through the various demes.
-The data we ultimately wish to analyze will be based only on samples, however;
-nor, in general, will the histories of deme occupancy be observable.
+The data we ultimately wish to analyze will be based only on samples, however.
+Nor, in general, will the histories of deme occupancy be observable.
 A generative model must account for this loss of information.
@@
 Note also that a pruned genealogy $\Pr_t$ is characterized by functions $\Pr_t^\y$ and $\Pr_t^\z$ just as an un-pruned genealogy is.
+Finally, observe that, since it contains within itself all of its past history, the pruned genealogy process $\Pr_t$ is Markov.
 
@@
 In the following, we will find that we need to count the deme-specific numbers of lineages present in a given pruned genealogy at a given time.
-Accordingly, suppose $\Pr_T=(y,z)$ is a pruned genealogy and suppose $t\in[0,T]$.
+Accordingly, suppose $\P_T=(\y,\z)$ is a pruned genealogy and suppose $t\in[0,T]$.
 Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell=(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
 Clearly, $\ell$ depends only on the values, $(\y_t,\z_t)$, of $y$ and $z$ at time $t$.
-Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(\y,\z)$ is a pruned genealogy, $\ell(\y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
+Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(\yr,\zr)$ is a pruned genealogy, $\ell(\yr_t,\zr_t)$ is the vector of deme-specific lineage counts at time $t$.
 We refer to $\ell$ as the \emph{lineage-count} function (cf.~\cref{fig:ells}).
@@
 We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
-In particular, given a node time $t$ in a pruned genealogy $\Pr_T=(\y,\z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
+In particular, given a node time $t$ in a pruned genealogy $\P_T=(\y,\z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
 Like the lineage-count function, $s$ depends only on the local structure of $\Pr_T$.
 However, $s$ depends not only on $(\y_t,\z_t)$, but also on $(\yt_t,\zt_t)$.
-Thus, we can define the \emph{saturation} function such that, whenever $\Pr_T=(\y,\z)$ is a pruned genealogy, $s(\yt_t,\y_t,\zt_t,\z_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
+Thus, we can define the \emph{saturation} function such that, whenever $\Pr_T=(\yr,\zr)$ is a pruned genealogy, $s(\yrt_t,\yr_t,\zrt_t,\zr_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
 \Cref{fig:ells} illustrates.
@@
 
-Our first result will be an expression for the likelihood of a given pruned genealogy $\P^*_T$ given the history $\Hr_T$ of the population process to time $T$.
+Our first result will be an expression for the likelihood of a given pruned genealogy given the history of the population process.
 
 \begin{thm}\label{thm:pruned_lik}
-  Let $\P^*_T=(\y^*,\z^*)$  be a given pruned genealogy and
+  Suppose $\P^*=(\y^*,\z^*)$ is a given pruned genealogy with $t(\P^*)=T$.
+  Define
   \begin{equation}\label{eq:phidef}
@@
   \begin{equation*}
-    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\Indicator{\event{\H}{T}\supseteq{\event{\P}{T}^*}}\,\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}.
+    \CondProb{\Pr_T=\P^*}{\Hr_T=\H_T}=\Indicator{\event{\H}{T}\supseteq{\event{\P}{}^*}}\,\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}.
   \end{equation*}
@@
 \begin{proof}
-  If $\event{\Hr}{T}\nsupseteq\event{\P}{T}^*$, then $\Hr_T$ and $\P^*_T$ are incompatible and $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
-  Similarly, if any event of $\Hr_T$ is incompatible with the local structure of $\P^*_T$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
+  If $\event{\Hr}{T}\nsupseteq\event{\P}{}^*$, then $\Hr_T$ and $\P^*$ are incompatible and $\CondProb{\Pr_T=\P^*}{\Hr_T}=0$.
+  Similarly, if any event of $\Hr_T$ is incompatible with the local structure of $\P^*$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_T=\P^*}{\Hr_T}=0$.
   Let us therefore suppose that neither of these conditions hold.
-  Conditional on $\Hr_T=\H_T$, at each time $t\in\event{\H}{T}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
-  In $\P^*_T$, at time $t$, there are $\ell_i=\ell_i(\y^*_t,\z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)$ are emergent.
+  Conditional on $\Hr_T=\H_T$, at each time $t\in\event{\H}{T}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
+  In $\P^*$, at time $t$, there are $\ell_i=\ell_i(\y^*_t,\z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)$ are emergent.
   By assumption, at each genealogical event, lineages within a deme are exchangeable:
@@
   \end{equation*}
-  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $\P^*_t$.
+  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct;
+  just one of these is the one present in $\P^*$.
   Moreover, since, again conditional on $\Hr_T$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
@@
   \end{equation*}
-  Returning to the possibility that $\H_T$ is incompatible with $\P^*_T$, since $\Prob{\Pr_T=\P^*_T}=0$ if either any $Q_u=0$ or any event of $\P^*_T$ is not an event of $\H_T$, we obtain the result.
+  Returning to the possibility that $\H_T$ is incompatible with $\P^*$, since $\Prob{\Pr_T=\P^*}=0$ if either any $Q_u=0$ or $\event{\P}{}^*\nsupseteq\event{\H}{T}$, we obtain the result.
 \end{proof}
@@
 \begin{thm}
-  Suppose that $\P^*_T=(y^*,z^*)$ is a pruned genealogy, for $T>0$.
+  Suppose that $\P^*=(y^*,z^*)$ is a pruned genealogy with $T=t(\P^*)>0$.
   If $w=w(t,x)$ satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
@@
   where $\phi$ is defined in \cref{eq:phidef},
-  then the likelihood of $\P^*_T$ is
+  then the likelihood of $\P^*$ is
   \begin{equation*}
-    \lik(\P^*_T)=\int{w(T,x)\,\dd{x}}.
+    \lik(\P^*)=\int{w(T,x)\,\dd{x}}.
   \end{equation*}
@@
   \AAK{[This can be rephrased somewhat more compactly into a result claiming the existence of the Markov process $\y$.]}
-  Let $\z^*=\obs(\P^*_T)$ be a given obscured genealogy.
+  Let $\z^*$ be a given obscured genealogy.
   Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{\z^*}}{q(x,y)}=1$.
   Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,\z^*_t)=1$
-  and that $q(x,y)>0$ whenever $\Prob{\Xr_0=x,\y^{\Pr_T}_0=y}>0$.
-  Then there is a Markov jump process $\yr_t$ on $\Yspace_{\z^*}$ such that
+  and that $q(x,y)>0$ whenever $\CondProb{\y_0=y}{\Xr_0=x}>0$.
+  Then there is a stochastic jump process $\yr_t$ with sample paths in $\func{Y}(\z^*)$ such that $(\Xr_t,\Ur_t,\yr_t)$ is Markov and
   \begin{equation*}
     \CondProb{\zr_T=\z^*}{\Hr_T=\H_T}=
-    \Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
+    \Indicator{\event{\H}{T}\supseteq\event{\z}{}^*}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
   \end{equation*}
@@
   Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
-  we will propose pruned genealogies compatible with $z^*_t$ as sample paths from a Markov process on $\Yspace_{z^*}$ and
+  we will propose pruned genealogies compatible with $z^*_t$ as sample paths from a stochastic process driven by $\Xr_t$ and
   evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
-  Conditional on $\Hr_T$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$, on $\Yspace_{z^*}$ such that
+  Conditional on $\Hr_T$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$ such that
   \begin{equation*}
     \begin{gathered}
-      \Prob{\yrh_0}=q(\yrh_0),
+      \CondProb{\yrh_0}{\Hr_T=\H_T}=q(\X_0,\yrh_0),
       \qquad
@@
   \end{equation*}
-  The required Markov process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
-  This construction of $\yr_t$ obviously guarantees that $\event{\H}{t}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$.
+  The required process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
+  This construction of $\yr_t$ obviously guarantees that $\event{\H}{t}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$ and that $(\Xr_t,\Ur_t,\yr_t)$ is Markov.
 
-  Let us define $\Pr(\yr)=(\yr,z^*)$.
-  Then, by construction, for every $\yr$, $\obs(\Pr(\yr))=z^*$ and, conversely, for every pruned genealogy $\Pr_T$ satisfying $\obs(\Pr_T)=z^*$, $y^{\Pr_T}$ is a sample path of this process having nonzero probability.
+  Let us define $\P(\yr)=(\yr,z^*)$.
+  Then, by construction, for every $\yr\in\func{Y}(\z)$, $\obs(\P(\yr))=z^*$ and, conversely, for every pruned genealogy $\Pr$ satisfying $\obs(\Pr)=z^*$, $\Pr^\y$ is a sample path of this process having nonzero probability.
   We therefore have that
@@
     \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
-    \Expect{\frac{\CondProb{\Pr_T=P(\yr)}{\Hr_T=\H_T}}{\pi(\yr\vert\H_T)}}
+    \Expect{\frac{\CondProb{\Pr_T=\P(\yr)}{\Hr_T=\H_T}}{\pi(\yr\vert\H_T)}}
   \end{equation*}
@@
   \begin{equation*}
-    \pi(\yr\vert\H_T)=\prod_{t\in\event{\H}{T}}{\pi_{\U_{t}}(t,\Xt_{t},\X_{t},\yrt_{t},\yr_{t})}.
+    \pi(\yr\vert\H_T)=q(\X_0,\yr_0)\,\prod_{t\in\event{\H}{T}}{\pi_{\U_{t}}(t,\Xt_{t},\X_{t},\yrt_{t},\yr_{t})}.
   \end{equation*}
@@
   with the initial condition $w(0,x,y)=p_0(x)\,q(x,y)$.
-  Then the likelihood of $\z^*_T$ is
+  Then the likelihood of $\z^*$ is
   \begin{equation*}
-    \lik(\z^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
+    \lik(\z^*)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
   \end{equation*}
@@
 \section{Discussion}
+
+Generalization of coalescent and birth-death process approaches.
+Both Moran model and birth-death processes are special cases.
+
+Allows models with demographic stochasticity.
+Incorporating environmental stochasticity is likely possible:
+extension to stochastic processes with a diffusion component.
+
+Freedom to choose models with many demes.
+Freedom to choose model of sampling.
+
+Price of flexibility is variability in the Monte Carlo estimation.
+Freedom to choose importance sampling distribution.
+Allowable to borrow information from the future.
+
+Filter equation formalism suggests approximations based on discretization of time.
 

### 2024-02-12 ms.Rnw
update

@@
 stopifnot(packageVersion("pomp")>="5.1")
-stopifnot(packageVersion("phylopomp")>="0.10.1")
+stopifnot(packageVersion("phylopomp")>="0.10.2")
 theme_set(theme_bw(base_family="serif"))
@@
       "SEIR",
-      Beta=1,sigma=0.5,gamma=0.1,psi=0.4,delta=0.1,
+      Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
       S0=10,E0=1,I0=1,R0=0,

### 2024-02-14 ms.Rnw
start examples; more edits

@@
 stopifnot(getRversion() >= "4.3")
-stopifnot(packageVersion("pomp")>="5.1")
-stopifnot(packageVersion("phylopomp")>="0.10.2")
+stopifnot(packageVersion("pomp")>="5.5")
+stopifnot(packageVersion("phylopomp")>="0.10.3")
 theme_set(theme_bw(base_family="serif"))
@@
 We will be concerned with a variety of stochastic processes, in both discrete and continuous time.
-In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
-In the case of continuous-time processes, we will assume that sample paths are right-continuous with left limits (\ie \cadlag).
+In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $\Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
+In the case of continuous-time processes, we will assume that sample paths are \cadlag\ \ie, right-continuous with left limits.
 We will frequently need to refer to the left-limit of such a process.
@@
   \begin{cases}
-    \displaystyle\lim_{t'\,\uparrow\,{t}}\;\Zr_{t'}, & t>0,\\[2ex]
+    \displaystyle\lim_{s\,\uparrow\,{t}}\;\Zr_{s}, & t>0,\\[2ex]
     \Zr_0, & t=0.\\
@@
 
-If $Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
+If $\Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
 In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
 $\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
-If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its event times in the interval $[0,t]$.
+If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its jump times in the interval $[0,t]$.
 
@@
 
-Motivating examples: compartmental models.
-Wide variety of models.
+Motivating examples:
+compartmental models in ID epidemiology,
+phylogenetics and systematics.
+Wide variety of models are of interest (\cref{fig:examples}).
 Linear chain trick.
@@
 In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
-The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
+The population process is completely specified by its initial-state density, $p_0$, and its transition rates $\alpha$.
 In particular, we suppose that
@@
 for all $t\in\Rp$ and $x,x'\in\Xspace$.
+\AAK{[We could mention that we use the continuity of $\alpha$ with respect to $t$ in some of the following, but that this condition can be relaxed.]}
 Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
@@
 \end{equation*}
-We will further assume that $\Xr_t$ is non-explosive, \ie that $\Prob{\Kr_t<\infty}=1$ for all $t$.
+We will further assume that $\Xr_t$ is non-explosive, so that $\Prob{\Kr_t<\infty}=1$ for all $t$.
 
@@
 
-The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
+The above may be compactly summarized by stating that if $v(t,x)$ satisfies the Kolmogorov forward equation (KFE),
 \begin{equation}
   \label{eq:kfe}
-  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
+  \frac{\partial{v}}{\partial{t}}(t,x)=\int\!v(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!v(t,x)\,\alpha(t,x,x')\,\dd{x'},
 \end{equation}
-and if, moreover, $w(0,x) = p_0(x)$,
-then $\int_{\mathcal{E}}\!{w(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
+and if, moreover, $v(0,x) = p_0(x)$,
+then $\int_{\mathcal{E}}\!{v(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
 \Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.
@@
 Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
-In particular, we suppose that there are a countable set of subpopulations that differ in their vital rates, but within each of which, individual lineages are statistically identical.
+In particular, we suppose that there are a countable set of subpopulations that may differ in their vital rates, but within each of which, individual lineages are statistically identical.
 We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
-
-For any $i\in\Demes$, we let $n_i(\Xr_t)$ denote the number of lineages present in deme $i$ at time $t$, \ie the \emph{occupancy} of deme $i$.
-Thus $n(\Xr_t)\in\Zp^{\Demes}$ is the vector of deme occupancies.
+We define the \emph{deme occupancy} function $n:\Demes\times\Xspace\to\Zp$ so that
+for $i\in\Demes$, $x\in\Xspace$, $n_i(x)$ is the number of lineages in deme $i$ when the population is in state $x$.
 
@@
 \end{equation*}
-In \cref{fig:markov_state}, we use the marks to distinguish events with distinct effects on genealogical relationships.
+\Cref{fig:markov_state} shows an example with five distinct marks.
 Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.
@@
 Observe that, though $\Xr_t$ and $(\Xr_t,\Ur_t)$ are Markov processes, $\Ur_t$ is not.
+
+\subsection{Examples}
+
+\subsubsection{SIRS model}
+
+\citet{King2022} worked out formulas for the exact likelihood of a genealogy induced by an SIRS model.
+The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.
+Its state vector is $x=(S,I,R)$ and its KFE is
+\begin{equation*}
+  \begin{split}
+    \frac{\partial{v}}{\partial{t}}(t,S,I,R)=
+    &\frac{\beta(t)\,I\,(S+1)}{N}\,v(t,S+1,I-1,R)
+    -\frac{\beta(t)\,I\,S}{N}\,v(t,S,I,R)\\
+    &+\gamma\,(I+1)\,v(t,S,I+1,R-1)
+    -\gamma\,I\,v(t,S,I,R)\\
+    &+\omega\,(R+1)\,v(t,S-1,I,R+1)
+    -\omega\,R\,v(t,S,I,R).
+  \end{split}
+\end{equation*}
+Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.
+
+\subsubsection{SEIRS model}
+
+A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
+The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
+It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
+\begin{equation*}
+  \begin{split}
+    \frac{\partial{v}}{\partial{t}}(t,S,E,I,R)=
+    &\frac{\beta(t)\,I\,(S+1)}{N}\,v(t,S+1,E-1,I,R)
+    -\frac{\beta(t)\,I\,S}{N}\,v(t,S,E,I,R)\\
+    &+\sigma\,(E+1)\,v(t,S,E+1,I-1,R)
+    -\sigma\,E\,v(t,S,E,I,R)\\
+    &+\gamma\,(I+1)\,v(t,S,E,I+1,R-1)
+    -\gamma\,I\,v(t,S,E,I,R)\\
+    &+\omega\,(R+1)\,v(t,S-1,E,I,R+1)
+    -\omega\,R\,v(t,S,E,I,R),
+  \end{split}
+\end{equation*}
+where $N=S+E+I+R$ is the total population size.
+The deme occupancy function in this case is $n(x)=(E,I)$.
+Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.
+
+\subsubsection{Two-strain competition model}
+
+A simple model for the competition of two strains for susceptible hosts is depicted in \cref{fig:example_models}B.
+In this model, the state vector consists of seven numbers: $x=(S,E_1,E_2,I_1,I_2,R_1,R_2)$.
+There are four demes ($\Demes=\Set{\lab{E}_1,\lab{E}_2,\lab{I}_1,\lab{I}_2}$) and the occupancy function is $n(x)=(E_1,E_2,I_1,I_2)$.
+This model has KFE
+\begin{equation*}
+  \begin{split}
+    \frac{\partial{v}}{\partial{t}}=
+    &\frac{\beta_1(t)\,I_1\,(S+1)}{N}\,v(t,S+1,E_1-1,E_2,I_1,I_2,R_1,R_2)
+    -\frac{\beta_1(t)\,I_1\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+    &+\frac{\beta_2(t)\,I_2\,(S+1)}{N}\,v(t,S+1,E_1,E_2-1,I_1,I_2,R_1,R_2)
+    -\frac{\beta_2(t)\,I_2\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+    &+\sigma_1\,(E_1+1)\,v(t,S,E_1+1,E_2,I_1-1,I_2,R_1,R_2)
+    -\sigma_1\,E_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+    &+\sigma_2\,(E_2+1)\,v(t,S,E_1,E_2+1,I_1,I_2-1,R_1,R_2)
+    -\sigma_2\,E_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+    &+\gamma_1\,(I_1+1)\,v(t,S,E_1,E_2,I_1+1,I_2,R_1-1,R_2)
+    -\gamma_1\,I_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+    &+\gamma_2\,(I_2+1)\,v(t,S,E_1,E_2,I_1,I_2+1,R_1,R_2-1)
+    -\gamma_2\,I_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+    &+\omega_1\,(R_1+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1+1,R_2)
+    -\omega_1\,R_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+    &+\omega_2\,(R_2+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1,R_2+1)
+    -\omega_2\,R_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+  \end{split}
+\end{equation*}
+
+\subsubsection{Linear birth-death model}
+
+\begin{equation*}
+  \frac{\partial{v}}{\partial{t}}=
+  \lambda\,(N-1)\,v(t,N-1)
+  -\lambda\,N\,v(t,N)
+  +\mu\,(N+1)\,v(t,N+1)
+  -\mu\,N\,v(t,N)
+\end{equation*}
+
+\subsubsection{Moran model and the Kingman coalescent}
 
@@
 
+\subsection{Genealogies}
+\label{sec:genealogy}
+
+\begin{figure}
+  \begin{center}
+    <<geneal0>>=
+    freeze(
+      seed=522390503,
+      simulate(
+        "SEIR",
+        Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
+        S0=10,E0=1,I0=1,R0=0,
+        time=10
+      )
+    ) -> x
+    @
+    <<geneal1,dependson="geneal0",fig.dim=c(4,2),out.width="50%">>=
+    pal <- c("#00274CFF","#FFCB05FF")
+    x |> plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
+      geom_vline(xintercept=10,linewidth=0.2,color="black")
+    @
+  \end{center}
+  \caption{
+    $\time(\G)=10$.
+    Two demes, $\Demes=\Set{\func{blue},\func{yellow}}$.
+    Extant lineages are denoted with black dots;
+    blue dots indicate samples.
+    \label{fig:geneal}
+  }
+\end{figure}
+
+A \emph{genealogy}, $\G$, encapsulates the relationships of shared ancestry among a set of lineages that are extant at some time $\time(\G)\in\Rp$, and perhaps a set of samples collected at earlier times (\cref{fig:geneal}A).
+A genealogy has a tree- or forest-like structure, with four distinct kinds of nodes:
+\begin{inparaenum}[(i)]
+\item \emph{tip nodes}, which represent labeled living lineages;
+\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another;
+\item \emph{sample nodes}, which represent labeled samples; and
+\item \emph{root nodes}, at the base of each tree.
+\end{inparaenum}
+Each node $a$ is associated with a specific time, $\time(a)$.
+In particular, if $a$ is a tip node in $\G$, then $\time(a)=\time(\G)$;
+if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
+Moreover, if node $a$ is ancestral to node $a'$, then $\time(a)\le{\time(a')}$ and $\time(a')-\time(a)$ is the distance between $a$ and $a'$ along the genealogy.
+Without loss of generality we assume that $\time(a)=0$ for all root nodes $a$.
+We let $\event{\G}{}$ denote the set of all internal and sample node-times of the genealogy $\G$;
+we refer to these as \emph{genealogical event times}.
+
+Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
+Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme (\cref{fig:geneal}A).
+
+Formally, we define a genealogy, $\G$, to be an ordered pair $(\y,\z)$.
+The $\z$ element specifies the genealogy's tree structure, as follows.
+Let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all partitions of subsets of $\leaves$.
+Partition \emph{fineness} defines a partial order on $\func{Z}$.
+Specifically, for $\zeta,\zeta'\in{\func{Z}}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
+If $\time(\G)=T$, the tree structure of $\G$ is defined by a \cadlag\ map $\z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
+An element $b\in\z_t$ is a set of labels;
+it represents the branch of the tree that bears the corresponding lineages.
+
+The $\y$ element of $\G$ specifies the coloring of branches and location of nodes.
+In particular, if $a$ is the label of any tip or sample node and $t\in[0,T]$, we define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$ and think of this as the color of the corresponding branch at that time.
+We will also need to count the number of nodes along each lineage:
+let $m_{t}(a)$ be the number of internal or sample nodes encountered as one descends along lineage $a$ from time $t=\time(a)$ to $t=0$.
+Observe that $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
+Formally, we define the coloring component $\y$ as follows.
+Let $\Yspace=\Demes\times\Zp$.
+For $\eta\in\Yspace$, write $\eta=(\eta^d,\eta^m)$ to denote the two components.
+Then, for $\zeta\in\func{Z}$, each $\eta\in\Yspace^{\zeta}$ is a map from $\zeta$ into $\Yspace$.
+Now, given a tree $\z:[0,T]\to\func{Z}$ and a map $\y:[0,T]\to{\coprod_{t\in[0,T]}{\Yspace^{\z_t}}}$,
+we say that $\y$ is \emph{compatible} with $\z$ if,
+for every $t\in[0,T]$, $\y_t\in{\Yspace^{\z_t}}$,
+$\y$ is \cadlag\ and,
+whenever $t_1,t_2\in[0,T]$, $b_1\in\z_{t_1}$, $b_2\in\z_{t_2}$
+such that $t_1\le{t_2}$ and $b_1\supseteq{b_2}$,
+we have $y^m_{t_1}\le{y^m_{t_2}}$.
+Let $\func{Y}(\z)$ denote the set of such maps compatible with $\z$ and $\func{Y}_t(\z)\coloneq\CondSet{y_t}{y\in\func{Y}(\z)}$.
+\AAK{[The coproduct is a fiber bundle over the space of trees and $\func{Y}(\z)$ is a section of this bundle.]}
+
+The two \cadlag\ functions $\y$ and $\z$ specify a genealogy completely and uniquely.
+When the genealogy is not clear from context, we write $\G^{\y}$ and $\G^{\z}$ to refer to the respective components of genealogy $\G$.
+
 \subsection{Binomial ratio}
@@
 
-\subsection{Examples}
-
-\subsubsection{SEIRS model}
-
-\subsubsection{SIIR model}
-
-\subsubsection{Linear birth-death model}
-
-\subsubsection{Moran model and the Kingman coalescent}
-
 \section{The induced genealogy process}
-
-\subsection{Genealogies}
-\label{sec:genealogy}
-
-For our purposes, a \emph{genealogy}, $\G_{T}$, is a labeled, time-calibrated tree.
-It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
-A genealogy includes three distinct kinds of nodes:
-\begin{inparaenum}[(i)]
-\item \emph{tip nodes}, which represent labeled living lineages;
-\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
-\item \emph{sample nodes}, which represent labeled samples.
-\end{inparaenum}
-Each node $a$ is associated with a specific time, $t(a)$.
-In particular, if $a$ is a tip node in $\G_T$, then $t(a)=T$;
-if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
-Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
-We let $\event{\G}{T}$ denote the set of all internal and sample node-times of the genealogy $\G_T$;
-we refer to these as \emph{genealogical event times}.
-Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
-Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.
-
-Formally, we define a genealogy, $\G_T$, to be an ordered pair $(\y,\z)$.
-The $\z$ element specifies the genealogy's tree structure, as follows.
-Let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all partitions of subsets of $\leaves$.
-Partition \emph{fineness} defines a partial order on $\func{Z}$.
-Specifically, for $\zeta,\zeta'\in{\func{Z}}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
-The tree structure of $\G_T$ is defined by a \cadlag\ map $\z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
-An element $b\in\z_t$ is a set of labels;
-it represents the branch of the tree that bears the corresponding lineages.
-
-The $\y$ element of $\G_T$ specifies the coloring of branches and location of nodes.
-In particular, if $a$ is the label of any tip or sample node and $t\in[0,T]$, we define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$ and think of this as the color of the corresponding branch at that time.
-We will also need to count the number of nodes along each lineage:
-let $m_{t}(a)$ be the number of internal or sample nodes encountered as one descends along lineage $a$ from time $t=t(a)$ to $t=0$.
-Observe that $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
-Formally, we define the coloring component $\y$ as follows.
-Let $\Yspace=\Demes\times\Zp$.
-For $\eta\in\Yspace$, write $\eta=(\eta^d,\eta^m)$ to denote the two components.
-Then, for $\zeta\in\func{Z}$, each $\eta\in\Yspace^{\zeta}$ is a map from $\zeta$ into $\Yspace$.
-Now, given a tree $\z:[0,T]\to\func{Z}$ and a map $\y:[0,T]\to{\coprod_{t\in[0,T]}{\Yspace^{\z_t}}}$,
-we say that $\y$ is \emph{compatible} with $\z$ if,
-for every $t\in[0,T]$, $\y_t\in{\Yspace^{\z_t}}$,
-$\y$ is \cadlag\ and,
-whenever $t_1,t_2\in[0,T]$, $b_1\in\z_{t_1}$, $b_2\in\z_{t_2}$
-such that $t_1\le{t_2}$ and $b_1\supseteq{b_2}$,
-we have $y^m_{t_1}\le{y^m_{t_2}}$.
-Let $\func{Y}(\z)$ denote the set of such maps compatible with $\z$.
-\AAK{[The coproduct is a fiber bundle over the space of trees and $\func{Y}(\z)$ is a section of this bundle.]}
-
-The two \cadlag\ functions $\y$ and $\z$ specify a genealogy completely and uniquely.
-When the genealogy is not clear from context, we write $\G^{\y}$ and $\G^{\z}$ to refer to the respective components of genealogy $\G$.
 
@@
 
-We now show how a given population process induces a stochastic process, $\Gen_t$, on the space of genealogies.
+We now show how a given population process induces a stochastic process, $\Gr_t$, on the space of genealogies.
 In the case of unstructured population processes (\ie those having a single deme), \citet{King2022} gave a related construction that is equivalent to the one presented here.
@@
 In particular, at each jump-time $\Trh_k$, the genealogy $\Grh_{k-1}$ is modified according to the jump-mark $\Urh_k$ to yield $\Grh_k$.
-We view $\Grh_k$ as the embedded chain of the continuous-time genealogy process $\Gen_t$.
+We view $\Grh_k$ as the embedded chain of the continuous-time genealogy process $\Gr_t$.
 It is very important to note that, conditional on $\Urh_k$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of the choices at any other times and independent of $(\Trh_k,\Xrh_k)$ for all $k$.
@@
 \begin{figure}
-  <<upo,fig.dim=c(6,8),out.width="60%",results="hide">>=
-  freeze(
-    seed=522390503,
-    simulate(
-      "SEIR",
-      Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
-      S0=10,E0=1,I0=1,R0=0,
-      time=10
-    )
-  ) -> x
-
+  <<geneal2,dependson="geneal0",fig.dim=c(4,6),out.width="50%">>=
   pal <- c("#00274CFF","#FFCB05FF")
-
   plot_grid(
     A=x |>
-      plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
-      lims(x=c(NA,10),y=c(3,NA))+
+      plot(
+        points=TRUE,prune=FALSE,obscure=FALSE,
+        ladderize=FALSE,palette=pal
+      )+
       geom_vline(xintercept=10,linewidth=0.2,color="black"),
     B=x |>
-      plot(points=TRUE,prune=TRUE,obscure=FALSE,palette=pal)+
-      lims(x=c(NA,10),y=c(3,NA))+
+      plot(
+        points=TRUE,prune=TRUE,obscure=FALSE,
+        ladderize=FALSE,palette=pal
+      )+
       geom_vline(xintercept=10,linewidth=0.2,color="black"),
     C=x |>
-      plot(points=TRUE,prune=TRUE,obscure=TRUE,palette="#B3B3B3FF")+
-      lims(x=c(NA,10),y=c(3,NA))+
-      geom_vline(xintercept=10,linewidth=0.2,color="black")+
-      labs(x="time"),
+      plot(
+        points=TRUE,prune=TRUE,obscure=TRUE,
+        ladderize=FALSE,palette="#B3B3B3FF"
+      )+
+      geom_vline(xintercept=10,linewidth=0.2,color="black"),
     ncol=1,
@@
     Unpruned, pruned, and obscured genealogies from a single realization of the genealogy process induced by the SEIRS model depicted in \cref{fig:example_models,fig:markov_state}.
-    \textbf{(A)} The full genealogy at time $t=10$.
+    \textbf{(A)} A realization of the unpruned genealogy process $\Gr_t$ is shown at $t=10$.
     Tip nodes, corresponding to lineages alive at time $t=10$ are indicated with black points.
@@
     blue denotes $\lab{E}$ and yellow, $\lab{I}$.
-    \textbf{(B)} The full genealogy is \emph{pruned} by removing all tip nodes and then recursively pruning away childless internal nodes.
+    \textbf{(B)} The genealogy is \emph{pruned} by deleting all tip nodes and then recursively pruning away childless internal nodes.
     Sample nodes are never removed.
-    \textbf{(C)} A genealogy is \emph{obscured} by removing all deme information from lineage histories:
+    \textbf{(C)} A genealogy is \emph{obscured} by effacing all deme information from lineage histories:
     the colors are erased, as are extraneous internal nodes.
@@
 
-Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Pr_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every childless internal node.
+Given a genealogy $\G$, one obtains the \emph{pruned genealogy}, $\P=\prune(\G)$ by first dropping every tip node and then recursively dropping every childless internal node (\cref{fig:upo}A--B).
 In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and possibly some of the interior nodes of the genealogy.
 Observe that a pruned genealogy is a colored genealogy:
-it retains information about where among the demes each of its lineages was through time.
-Note also that a pruned genealogy $\Pr_t$ is characterized by functions $\Pr_t^\y$ and $\Pr_t^\z$ just as an un-pruned genealogy is.
-Finally, observe that, since it contains within itself all of its past history, the pruned genealogy process $\Pr_t$ is Markov.
+it retains information about where among the demes each of its lineages was through time (\cref{fig:upo}B).
+Note also that a pruned genealogy $\P$ is characterized by functions $\P^\y$ and $\P^\z$ just as an un-pruned genealogy is.
+Finally, observe that, since it contains within itself all of its past history, the pruned genealogy process $\Pr_t=\prune(\Gr_t)$ is Markov, even though the unpruned genealogy process, $\Gr_t$, is not.
 
@@
 In the following, we will find that we need to count the deme-specific numbers of lineages present in a given pruned genealogy at a given time.
-Accordingly, suppose $\P_T=(\y,\z)$ is a pruned genealogy and suppose $t\in[0,T]$.
+Accordingly, suppose $\P=(\y,\z)$ is a pruned genealogy with $\time(\P)=T$ and suppose $t\in[0,T]$.
 Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell=(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
 Clearly, $\ell$ depends only on the values, $(\y_t,\z_t)$, of $y$ and $z$ at time $t$.
-Therefore, we can define $\ell$ as a function such that, whenever $\Pr_T=(\yr,\zr)$ is a pruned genealogy, $\ell(\yr_t,\zr_t)$ is the vector of deme-specific lineage counts at time $t$.
+Therefore, we can define $\ell$ as a function such that, whenever $\P=(\y,\z)$ is a pruned genealogy, $\ell(\y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
 We refer to $\ell$ as the \emph{lineage-count} function (cf.~\cref{fig:ells}).
@@
 We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
-In particular, given a node time $t$ in a pruned genealogy $\P_T=(\y,\z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
-Like the lineage-count function, $s$ depends only on the local structure of $\Pr_T$.
+In particular, given a node time $t$ in a pruned genealogy $\P=(\y,\z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
+Like the lineage-count function, $s$ depends only on the local structure of $\P$.
 However, $s$ depends not only on $(\y_t,\z_t)$, but also on $(\yt_t,\zt_t)$.
-Thus, we can define the \emph{saturation} function such that, whenever $\Pr_T=(\yr,\zr)$ is a pruned genealogy, $s(\yrt_t,\yr_t,\zrt_t,\zr_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
+Thus, we can define the \emph{saturation} function such that, whenever $\P=(\y,\z)$ is a pruned genealogy, $s(\yrt_t,\yr_t,\zrt_t,\zr_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
 \Cref{fig:ells} illustrates.
@@
 
-Suppose $\P_T$ is a pruned genealogy and $t\in\event{P}{T}$.
-The local structure of $\P_T$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
-For example, if the event in $\P_T$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
-Similarly, if the node in $\P_T$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
+Suppose $\P$ is a pruned genealogy, with $\time(\P)=T$ and $t\in\event{\P}{}$.
+The local structure of $\P$ at $t$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
+For example, if the event in $\P$ at $t$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
+Similarly, if the node in $\P$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
 To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $y$ and $z$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
 That is, $Q_u(\eta,\eta',\zeta,\zeta')=1$ if and only if
-there is a feasible genealogy, $\G_T=(\y,\z)$, and history, $\H_T$,
+there is a feasible genealogy, $\G=(\y,\z)$, and history, $\H$,
 and a $t\in{[0,T]}$ such that,
-given $\Gr_T=\G_T$ and $\Hr_T=\H_T$,
+given $\Gr_T=\G$ and $\Hr_T=\H$,
 $\U_t=u$,
@@
 The \emph{obscured genealogy} is obtained by discarding all information about demes and events not visible from the topology of the tree alone.
-In particular, if $\Pr_T=(\y,\z)$ is a pruned genealogy, we write $\obs(\Pr_T)=\Pr_T^{\z}=\z$ to denote the obscured genealogy.
+In particular, if $\P=(\y,\z)$ is a pruned genealogy, we write $\obs(\P)=\P^{\z}=\z$ to denote the obscured genealogy.
 
@@
 \begin{thm}\label{thm:pruned_lik}
-  Suppose $\P^*=(\y^*,\z^*)$ is a given pruned genealogy with $t(\P^*)=T$.
+  Suppose $\P^*=(\y^*,\z^*)$ is a given pruned genealogy with $\time(\P^*)=T$.
   Define
@@
   \begin{equation*}
-    \CondProb{\Pr_T=\P^*}{\Hr_T=\H_T}=\Indicator{\event{\H}{T}\supseteq{\event{\P}{}^*}}\,\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}.
+    \CondProb{\Pr_T=\P^*}{\Hr_T=\H}=\Indicator{\event{\H}{}\supseteq{\event{\P}{}^*}}\,\prod_{t\in\event{\H}{}}{\phi_{\U_t}(\X_t,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}.
   \end{equation*}
@@
 \begin{proof}
-  If $\event{\Hr}{T}\nsupseteq\event{\P}{}^*$, then $\Hr_T$ and $\P^*$ are incompatible and $\CondProb{\Pr_T=\P^*}{\Hr_T}=0$.
-  Similarly, if any event of $\Hr_T$ is incompatible with the local structure of $\P^*$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_T=\P^*}{\Hr_T}=0$.
+  If $\event{\H}{}\nsupseteq\event{\P}{}^*$, then $\H$ and $\P^*$ are incompatible and $\CondProb{\Pr_T=\P^*}{\Hr_T=\H}=0$.
+  Similarly, if any event of $\H$ is incompatible with the local structure of $\P^*$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_T=\P^*}{\Hr_T=\H}=0$.
   Let us therefore suppose that neither of these conditions hold.
-  Conditional on $\Hr_T=\H_T$, at each time $t\in\event{\H}{T}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
+  Conditional on $\Hr_T=\H$, at each time $t\in\event{\H}{}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
   In $\P^*$, at time $t$, there are $\ell_i=\ell_i(\y^*_t,\z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)$ are emergent.
@@
   just one of these is the one present in $\P^*$.
-  Moreover, since, again conditional on $\Hr_T$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
+  Moreover, since, again conditional on $\Hr_T=\H$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
   \begin{equation*}
-    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\BinRatio{n(\X_t)}{\ell(\y^*_t,\z^*_t)}{r^{\U_t}}{s(\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}}.
+    \CondProb{\Pr_T=\P^*}{\Hr_T=\H}=\prod_{t\in\event{\H}{}}{\BinRatio{n(\X_t)}{\ell(\y^*_t,\z^*_t)}{r^{\U_t}}{s(\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}}.
   \end{equation*}
-  Returning to the possibility that $\H_T$ is incompatible with $\P^*$, since $\Prob{\Pr_T=\P^*}=0$ if either any $Q_u=0$ or $\event{\P}{}^*\nsupseteq\event{\H}{T}$, we obtain the result.
+  Returning to the possibility that $\H$ is incompatible with $\P^*$, since $\Prob{\Pr_T=\P^*}=0$ if either any $Q_u=0$ or $\event{\P}{}^*\nsubseteq\event{\H}{}$, we obtain the result.
 \end{proof}
@@
 \begin{thm}
-  Suppose that $\P^*=(y^*,z^*)$ is a pruned genealogy with $T=t(\P^*)>0$.
-  If $w=w(t,x)$ satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
+  Suppose that $\P^*=(\y^*,\z^*)$ is a pruned genealogy with $\time(\P^*)=T>0$.
+  Suppose that $w=w(t,x)$ satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
   \begin{equation}
@@
       &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)\,\dd{x'}}}
-      -\sum_u{\int{w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}}}\\
-      &\quad+\sum_{e\in\event{\P}{T}^*}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)\,\dd{x'}}}-w(t,x)\right\}},
+      -\sum_u{\int{w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}}},
+      &t\notin{\event{\P}{}^*},\\
+      w(t,x)=&\sum_u{\int{\wt(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)\,\dd{x'}}},
+      &t\in{\event{\P}{}^*},
     \end{aligned}
   \end{equation}
-  where $\phi$ is defined in \cref{eq:phidef},
-  then the likelihood of $\P^*$ is
+  where $\phi$ is defined in \cref{eq:phidef}.
+  Then the likelihood of $\P^*$ is
   \begin{equation*}
@@
 \begin{thm}\label{thm:obsc_lik}
-  \AAK{[This can be rephrased somewhat more compactly into a result claiming the existence of the Markov process $\y$.]}
-  Let $\z^*$ be a given obscured genealogy.
-  Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{\z^*}}{q(x,y)}=1$.
-  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,\z^*_t)=1$
-  and that $q(x,y)>0$ whenever $\CondProb{\y_0=y}{\Xr_0=x}>0$.
+  Suppose $\z^*$ is a given obscured genealogy with $\time(\z^*)=T$.
+  Let $q$ and $\pi$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\func{Y}_t(\z^*)$,
+  \begin{equation*}
+    \begin{gathered}
+      q(x,y)\ge{0},\qquad
+      \sum_{y\in\func{Y}_0(\z^*)}{q(x,y)}=1, \qquad
+      \pi_u(t,x,x',y,y')\ge{0},\qquad
+      \sum_{y'\in\func{Y}_t(\z^*)}{\pi_u(t,x,x',y,y')}=1.
+    \end{gathered}
+  \end{equation*}
+  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $\alpha_u(t,x,x')\,Q_u(y,y',\zt^*_t,\z^*_t)>0$
+  and that $q(x,y)>0$ whenever $\CondProb{\Pr_0^\y=y}{\Xr_0=x}>0$.
   Then there is a stochastic jump process $\yr_t$ with sample paths in $\func{Y}(\z^*)$ such that $(\Xr_t,\Ur_t,\yr_t)$ is Markov and
   \begin{equation*}
-    \CondProb{\zr_T=\z^*}{\Hr_T=\H_T}=
-    \Indicator{\event{\H}{T}\supseteq\event{\z}{}^*}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
+    \CondProb{\Pr^{\z}_T=\z^*}{\Hr_T=\H}=
+    \Indicator{\event{\H}{}\supseteq\event{\z}{}^*}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
   \end{equation*}
   where $\phi$ is defined in \cref{eq:phidef} and
-  the expectation is taken over the sample paths of $\yr_t$, $t\in[0,T]$.
+  the expectation is taken over the sample paths of $\yr_t$.
 \end{thm}
@@
   \begin{equation}\label{eq:IS1}
-    \CondProb{\zr_T=z^*}{\Hr_T}=\CondExpect{\Indicator{\obs(\Pr_T)=z^*}}{\Hr_T}.
+    \CondProb{\zr_T=\z^*}{\Hr_T=\H}=\CondExpect{\Indicator{\Pr^{\z}_T=\z^*}}{\Hr_T=\H}.
   \end{equation}
   Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
-  we will propose pruned genealogies compatible with $z^*_t$ as sample paths from a stochastic process driven by $\Xr_t$ and
+  we will propose pruned genealogies compatible with $\z^*_t$ as sample paths from a stochastic process driven by $\Xr_t$ and
   evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
-  Conditional on $\Hr_T$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$ such that
+  Conditional on $\Hr_T=\H$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$ such that
   \begin{equation*}
     \begin{gathered}
-      \CondProb{\yrh_0}{\Hr_T=\H_T}=q(\X_0,\yrh_0),
+      \CondProb{\yrh_0}{\Hr_T=\H}=q(\X_0,\yrh_0),
       \qquad
-      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_T=\H_T}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
+      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_T=\H}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
     \end{gathered}
@@
   The required process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
-  This construction of $\yr_t$ obviously guarantees that $\event{\H}{t}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$ and that $(\Xr_t,\Ur_t,\yr_t)$ is Markov.
+  This construction of $\yr_t$ obviously guarantees that $\event{\H}{}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$ and that $(\Xr_t,\Ur_t,\yr_t)$ is Markov.
 
-  Let us define $\P(\yr)=(\yr,z^*)$.
-  Then, by construction, for every $\yr\in\func{Y}(\z)$, $\obs(\P(\yr))=z^*$ and, conversely, for every pruned genealogy $\Pr$ satisfying $\obs(\Pr)=z^*$, $\Pr^\y$ is a sample path of this process having nonzero probability.
+  Now, for $\y\in\func{Y}(\z^*)$, let us define $C(\y)=(\y,\z^*)$.
+  Then, by construction, $\obs(C(\y))=\z^*$ and,
+  conversely, for every pruned genealogy $\P$ satisfying $\P^{\z}=\z^*$, $C(\P^{\y})=\P$.
+  Moreover, the conditions on the kernels $q$ and $\pi$ guarantee that, if $\CondProb{\Pr_T=\P}{\Hr_T=\H}>0$ and $\P^{\z}=\z^*$, then $\Prob{\yr=\P^{\y}}>0$.
   We therefore have that
   \begin{equation*}
-    \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
-    \Expect{\frac{\CondProb{\Pr_T=\P(\yr)}{\Hr_T=\H_T}}{\pi(\yr\vert\H_T)}}
+    \CondProb{\zr_T=\z^*}{\Hr_T=\H}=
+    \Expect{\frac{\CondProb{\Pr_T=C(\yr)}{\Hr_T=\H}}{\pi(\yr\vert\H)}},
   \end{equation*}
-  the expectation being taken with respect to the random process $\yr_t$.
+  the expectation being taken with respect to the random process $\yr$.
   Here, by definition,
   \begin{equation*}
-    \pi(\yr\vert\H_T)=q(\X_0,\yr_0)\,\prod_{t\in\event{\H}{T}}{\pi_{\U_{t}}(t,\Xt_{t},\X_{t},\yrt_{t},\yr_{t})}.
+    \pi(\yr\vert\H)=q(\X_0,\yr_0)\,\prod_{t\in\event{\H}{}}{\pi_{\U_{t}}(t,\Xt_{t},\X_{t},\yrt_{t},\yr_{t})}.
   \end{equation*}
@@
 \begin{thm}
-  \AAK{[This can be rephrased into a claim about the existence of the kernels $\pi$ and $q$.]}
-  Let $z^*$, be a given obscured genealogy.
-  Let $\pi$, $q$, and $\yr_t$ be as in \cref{thm:obsc_lik}.
-  For $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $u\in\Jumps$, define
+  Let $\z^*$ be a given obscured genealogy and $T=\time(\z^*)$.
+  Then there are probability kernels $q$ and $\pi$ as in \cref{thm:obsc_lik} such that if
   \begin{equation*}
     \begin{gathered}
-      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),\\[2ex]
+      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),\qquad
       \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y',\zt^*_t,\z^*_t)}{\pi_u(t,x,x',y,y')},
@@
   \end{equation*}
-  where $\phi$ is defined in \cref{eq:phidef}.
-  Suppose $w=w(t,x,y)$ satisfies the filter equation
-  \begin{equation}
+  and if $w=w(t,x,y)$ satisfies
+  the initial condition $w(0,x,y)=p_0(x)\,q(x,y)$
+  and the filter equation
+  \begin{equation*}
     \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x,y)=
-      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}}\\
-      &-\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
-      &+\sum_{e\in\event{\z}{}^*}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y,y')\,\psi_u(t,x',x,y',y)\,\dd{x'}}}-w(t,x,y)\right\}},
+      &\frac{\partial{w}}{\partial{t}}=
+      \sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}}
+      -\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}},
+      &t\in{\event{\z}{}^*},\\
+      &w(t,x,y)=\sum_{uy'}{\int{\wt(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}},
+      &t\in{\event{\z}{}^*},
     \end{aligned}
-  \end{equation}
-  with the initial condition $w(0,x,y)=p_0(x)\,q(x,y)$.
-  Then the likelihood of $\z^*$ is
+  \end{equation*}
+  then the likelihood of $\z^*$ is
   \begin{equation*}
@@
   \begin{equation*}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)},
+    \CondExpect{\Vr_t}{\Hr_t=\H}=\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)},
   \end{equation*}
@@
   \begin{equation}
-    \label{eq:reg-filter-eq}
+    \label{eq:reg-filter-eq2}
     \frac{\partial{w}}{\partial{t}}=\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
@@
 Filter equations afford a convenient means of computing expectations and likelihoods for pure jump processes.
-This is facilitated by the following Lemma, the statement of which uses a certain non-standard Dirac delta function.
-Specifically, let $\delta(v,v')$ be the right-sided Dirac delta function satisfying
+This is facilitated by the following Lemma, the statement of which uses a one-sdied Dirac delta function.
+Specifically, let $\delta(v,v')$ be the right-sided Dirac delta function satisfying $\delta(v,v')=0$ for $v\ne{v'}$ and
 \begin{equation*}
-  \begin{gathered}
-    \delta(v,v')=0 \qquad \text{for}\ v\ne{v'}\ \text{and}\\
   \int_a^b{f(v)\,\delta(v,v')\,\dd{v}}=f(v')\,\Indicator{v'\in\halfopen{a,b}},
-    \qquad \text{whenever}\ f \ \text{is \cadlag\ and}\ -\infty\le{a}<{b}\le{\infty}.
-  \end{gathered}
 \end{equation*}
+whenever $f$ is \cadlag\ and $-\infty\le{a}<{b}\le{\infty}$.
 
@@
   \label{lemma:filter}
-  The filter equation \eqref{eq:reg-filter-eq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satsfies the KFE
+  The filter equation \eqref{eq:reg-filter-eq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satisfies the KFE
   \begin{equation}

### 2024-02-15 ms.Rnw
update

@@
 In the following, we show a Markov population process of the kind that is a staple in epidemiology induces a Markov process on the space of genealogies.
-We then show how one can comput the likelihood of a given genealogy.
+We then show how one can compute the likelihood of a given genealogy.
 
@@
 $\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
-If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its jump times in the interval $[0,t]$.
 
@@
 phylogenetics and systematics.
-Wide variety of models are of interest (\cref{fig:examples}).
+Wide variety of models are of interest (\cref{fig:example_models}).
 Linear chain trick.
@@
 
-In the following, it will be useful to break the jumps of the population process $\Xr_t$ into distinct categories.
+In the following, it will be useful to divide the jumps of the population process $\Xr_t$ into distinct categories.
 For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
@@
 
+\subsubsection{Superspreading model}
+
+\Cref{fig:example_models}D depicts a model of superspreading.
+There are three demes ($\Demes=\Set{\lab{E},\lab{I_L},\lab{I_H}}$).
+The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
+\begin{equation*}
+  \begin{split}
+    \frac{\partial{v}}{\partial{t}}=
+    &\frac{\beta(t)\,(I_L+\theta\,I_H)\,(S+1)}{N}\,v(t,S+1,E-1,I_L,I_H,R)
+    -\frac{\beta(t)\,(I_L+\theta\,I_H)\,S}{N}\,v(t,S,E,I_L,I_H,R)\\
+    &\sigma\,(E+1)\,v(t,S,E+1,I_L-1,I_H,R)
+    -\sigma\,E\,v(t,S,E,I_L,I_H,R)\\
+    &\eps_{LH}\,(I_L+1)\,v(t,S,E,I_L+1,I_H-1,R)
+    -\eps_{LH}\,I_L\,v(t,S,E,I_L,I_H,R)\\
+    &\eps_{HL}\,(I_H+1)\,v(t,S,E,I_L-1,I_H+1,R)
+    -\eps_{HL}\,I_H\,v(t,S,E,I_L,I_H,R)\\
+    &\gamma\,(I_L+1)\,v(t,S,E,I_L+1,I_H,R-1)
+    -\gamma\,I_L\,v(t,S,E,I_L,I_H,R)\\
+    &\gamma\,(I_H+1)\,v(t,S,E,I_L,I_H+1,R-1)
+    -\gamma\,I_H\,v(t,S,E,I_L,I_H,R)\\
+    &\omega\,(R+1)\,v(t,S-1,E,I_L,I_H,R+1)
+    -\omega\,R\,v(t,S,E,I_L,I_H,R).\\
+  \end{split}
+\end{equation*}
+
 \subsubsection{Linear birth-death model}
@@
 
-\AAK{[Maybe it would be better to speak instead of $\Hr_t$ being the $\sigma$-algebra generated by the sequences $(\Trh_k,\Xrh_k,\Urh_k)_{k=1}^{\Kr_t}$.
+\AAK{[Maybe it would be better to speak instead of $\Hr_t$ being the random $\sigma$-algebra generated by the sequences $(\Trh_k,\Xrh_k,\Urh_k)_{k=1}^{\Kr_t}$.
     The only thing we do with $\Hr$ is condition on it.
@@
 Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.
+
 Alternatively, one can think of $\Hr_t$ as consisting of the sequence
@@
 
+If $\H$ is such a history, we define $\time(\H)$ to be the right endpoint of its domain and use the notation $\event{\H}=\Set{\Th_1,\dots,\Th_{\K_t}}\subset{[0,\time(\H)]}$ to denote the set of its jump times.
+
 \subsection{Genealogies}
@@
   \begin{center}
-    <<geneal0>>=
+    <<geneal,fig.dim=c(4,2),out.width="50%">>=
     freeze(
-      seed=522390503,
+      seed=382490723,
       simulate(
         "SEIR",
-        Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
-        S0=10,E0=1,I0=1,R0=0,
+        Beta=3,sigma=0.5,gamma=0.2,psi=0.3,omega=0.5,
+        S0=15,E0=1,I0=2,R0=0,
         time=10
@@
     ) -> x
-    @
-    <<geneal1,dependson="geneal0",fig.dim=c(4,2),out.width="50%">>=
+
     pal <- c("#00274CFF","#FFCB05FF")
+
     x |> plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
@@
     Two demes, $\Demes=\Set{\func{blue},\func{yellow}}$.
-    Extant lineages are denoted with black dots;
-    blue dots indicate samples.
+    Tip nodes, denoting extant lineages, are shown in black;
+    sample nodes are shown in blue;
+    internal nodes are indicated using green dots.
+    Note that internal nodes occur both at branch-points and inline.
+    Wherever a lineage moves from one deme (color) to another, an internal node occurs, but not conversely.
     \label{fig:geneal}
@@
 In particular, if $a$ is a tip node in $\G$, then $\time(a)=\time(\G)$;
-if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
+if $a$ is a sample node, then $\time(a)$ is the time at which the sample was taken.
 Moreover, if node $a$ is ancestral to node $a'$, then $\time(a)\le{\time(a')}$ and $\time(a')-\time(a)$ is the distance between $a$ and $a'$ along the genealogy.
@@
 Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
-Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme (\cref{fig:geneal}A).
+Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted with a distinct color for each deme (\cref{fig:geneal}A).
+Note that a genealogy will in general have \emph{branch-point nodes}, \ie internal nodes with more than one descendant, but may also have internal nodes with only one descendant.
+These occur whenever the color changes along a branch, but can also occur without a color-change.
+We refer to such nodes as \emph{inline nodes}.
 
@@
 it represents the branch of the tree that bears the corresponding lineages.
-
-The $\y$ element of $\G$ specifies the coloring of branches and location of nodes.
+We use the notation $\event{\z}$ to denote the set of all event-times in the tree $\z$.
+Note that $\event{\z}$ includes the times of all tip, sample, and branch-point nodes, but excludes inline and root nodes.
+Therefore, $\event{\z}\subseteq{\event{\G}}$.
+
+The $\y$ element of $\G$ specifies the coloring of branches and locations of tip, sample, and internal nodes (including inline nodes).
 In particular, if $a$ is the label of any tip or sample node and $t\in[0,T]$, we define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$ and think of this as the color of the corresponding branch at that time.
@@
 Formally, we define the coloring component $\y$ as follows.
-Let $\Yspace=\Demes\times\Zp$.
-For $\eta\in\Yspace$, write $\eta=(\eta^d,\eta^m)$ to denote the two components.
-Then, for $\zeta\in\func{Z}$, each $\eta\in\Yspace^{\zeta}$ is a map from $\zeta$ into $\Yspace$.
-Now, given a tree $\z:[0,T]\to\func{Z}$ and a map $\y:[0,T]\to{\coprod_{t\in[0,T]}{\Yspace^{\z_t}}}$,
-we say that $\y$ is \emph{compatible} with $\z$ if,
-for every $t\in[0,T]$, $\y_t\in{\Yspace^{\z_t}}$,
-$\y$ is \cadlag\ and,
-whenever $t_1,t_2\in[0,T]$, $b_1\in\z_{t_1}$, $b_2\in\z_{t_2}$
-such that $t_1\le{t_2}$ and $b_1\supseteq{b_2}$,
-we have $y^m_{t_1}\le{y^m_{t_2}}$.
+Let $\Yspace=\Demes\times\Zp$ and write $\upsilon\in\Yspace$ component-wise as $\upsilon=(\upsilon^d,\upsilon^m)$.
+For any $\zeta\in\func{Z}$, an element $\eta\in\Yspace^{\zeta}$ is a map from the partition $\zeta$ into $\Yspace$.
+Now, suppose we are given a tree $\z:[0,T]\to\func{Z}$ and a \cadlag\ map $\y:[0,T]\to{\coprod_{t\in[0,T]}{\Yspace^{\z_t}}}$.
+Let $\event{\y}$ denote the set of discontinuities of $\y$.
+We say that $\y$ is \emph{compatible} with $\z$ if it is a valid coloring of the tree $\z$.
+Formally, compatibility requires the following:
+\begin{inparaenum}[(i)]
+\item
+  $\event{\y}\supseteq\event{\z}$;
+\item
+  for every $t\in[0,T]$, $\y_t\in{\Yspace^{\z_t}}$;
+\item
+  if $0\le{t_1}\le{t_2}\le{T}$, $b_1\in\z_{t_1}$, $b_2\in\z_{t_2}$, and $b_1\supseteq{b_2}$,
+  then $y^m_{t_1}\le{y^m_{t_2}}$;
+\item $y^m_{t_1}=y^m_{t_2}$ implies $y^d_{t_1}=y^d_{t_2}$.
+\end{inparaenum}
 Let $\func{Y}(\z)$ denote the set of such maps compatible with $\z$ and $\func{Y}_t(\z)\coloneq\CondSet{y_t}{y\in\func{Y}(\z)}$.
-\AAK{[The coproduct is a fiber bundle over the space of trees and $\func{Y}(\z)$ is a section of this bundle.]}
+\AAK{[These conditions are necessary but not sufficient.]}
+\AAK{[The coproduct is a fiber bundle over the $\z$ and $\func{Y}(\z)$ is a section of this bundle.]}
 
@@
 \begin{figure}
-  <<geneal2,dependson="geneal0",fig.dim=c(4,6),out.width="50%">>=
+  <<upo,fig.dim=c(4,6),out.width="50%">>=
+  freeze(
+    seed=522390503,
+    simulate(
+      "SEIR",
+      Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
+      S0=10,E0=1,I0=1,R0=0,
+      time=10
+    )
+  ) -> x
+
   pal <- c("#00274CFF","#FFCB05FF")
+
   plot_grid(
@@
     \textbf{(C)} A genealogy is \emph{obscured} by effacing all deme information from lineage histories:
-    the colors are erased, as are extraneous internal nodes.
+    the colors are erased, as are all inline nodes.
     See the text (\cref{sec:genealogy,sec:pruning,sec:obscuration}) for more detail.

### 2024-02-18 ms.Rnw
reworking

@@
 
+Advent of routine molecular surveillance.
+With certain pathogens, with matching of time-scales, information on transmission is possible.
+With sufficiently dense sampling, it is sometimes possible to reconstruct WAIFW.
+More typically, one wishes to gain insight into the structure and parameterization of the transmission process.
+This is accomplished by formalizing a transmission model or suite of transmission models, estimating parameters, and comparing their ability to explain the data.
+
+Because there is commonly broad uncertainty regarding the structure of the transmission process, for example due to heterogeneities in transmission rates and susceptibilities, complex behavior patterns, etc., methods that have the plug-and-play property are particularly useful.
+Such techniques for inference based on time series are well understood and widely used.
+When the data lie in some Euclidean space, they can commonly be modeled as draws from some error distribution conditional on the latent state of the transmission process.
+However, when the data are genealogies (also called phylogenies) representing the relationships of shared ancestry among genomic samples, the data-space is non-Euclidean and the appropriate models connecting the data to the latent state are non-trivial.
+Here, we derive expressions for the exact likelihood of a genealogy generated through sampling genomes from an arbitrary discretely-structured, Markov latent state process.
+
 Problem of phylodynamics.
@@
 We will frequently need to refer to the left-limit of such a process.
-Accordingly, if $\Zr_t$ is a \cadlag\ random process, we define
-\begin{equation*}
-  \Zrt_t\colonequals
+Accordingly, if $\Phir_t$ is a \cadlag\ random process, we define
+\begin{equation*}
+  \Phirt_t\colonequals
   \begin{cases}
-    \displaystyle\lim_{s\,\uparrow\,{t}}\;\Zr_{s}, & t>0,\\[2ex]
-    \Zr_0, & t=0.\\
+    \displaystyle\lim_{s\,\uparrow\,{t}}\;\Phir_{s}, & t>0,\\[2ex]
+    \Phir_0, & t=0.\\
   \end{cases}
 \end{equation*}
-Note that $\Zrt_t$ is thus left-continuous with right limits.
-
-If $\Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
+Note that $\Phirt_t$ is thus left-continuous with right limits.
+
+If $\Phir_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Phirh_k\coloneq{\Phir_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
 In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
-$\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
+$\Phir_t=\Phirh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
 
@@
 We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
-In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
+In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\mathbb{Z}^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
 The population process is completely specified by its initial-state density, $p_0$, and its transition rates $\alpha$.
@@
 \end{equation*}
-We will further assume that $\Xr_t$ is non-explosive, so that $\Prob{\Kr_t<\infty}=1$ for all $t$.
+We will further assume that the number of jumps that occur in a finite time-interval is finite: $\Prob{\Kr_t<\infty}=1$ for all $t$.
 
@@
   \caption{
-    $\time(\G)=10$.
-    Two demes, $\Demes=\Set{\func{blue},\func{yellow}}$.
-    Tip nodes, denoting extant lineages, are shown in black;
-    sample nodes are shown in blue;
-    internal nodes are indicated using green dots.
-    Note that internal nodes occur both at branch-points and inline.
-    Wherever a lineage moves from one deme (color) to another, an internal node occurs, but not conversely.
+    A genealogy, $G$, specifies the relationships of shared ancestry (via its tree-structure) and deme occupancy histories (via the coloring of its branches) of a set of lineages extant at some time $\time(G)$, as well as some samples gathered at earlier times.
+    Here, $\time(G)=10$ and there are two demes, $\Demes=\Set{\func{blue},\func{yellow}}$.
+    Tip nodes, denoting extant lineages, are shown as black dots;
+    sample nodes are shown as blue dots;
+    internal nodes are indicated in green.
+    Note that internal nodes occur both at branch-points and inline (\ie along branches).
+    Wherever a lineage moves from one deme (color) to another, an internal node occurs;
+    the converse does not necessarily hold.
     \label{fig:geneal}
@@
 
-A \emph{genealogy}, $\G$, encapsulates the relationships of shared ancestry among a set of lineages that are extant at some time $\time(\G)\in\Rp$, and perhaps a set of samples collected at earlier times (\cref{fig:geneal}A).
+A \emph{genealogy}, $G$, encapsulates the relationships of shared ancestry among a set of lineages that are extant at some time $\time(G)\in\Rp$, and perhaps a set of samples collected at earlier times (\cref{fig:geneal}).
 A genealogy has a tree- or forest-like structure, with four distinct kinds of nodes:
 \begin{inparaenum}[(i)]
-\item \emph{tip nodes}, which represent labeled living lineages;
+\item \emph{tip nodes}, which represent labeled extant lineages;
 \item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another;
@@
 Each node $a$ is associated with a specific time, $\time(a)$.
-In particular, if $a$ is a tip node in $\G$, then $\time(a)=\time(\G)$;
+In particular, if $a$ is a tip node in $G$, then $\time(a)=\time(G)$;
 if $a$ is a sample node, then $\time(a)$ is the time at which the sample was taken.
@@
 Without loss of generality we assume that $\time(a)=0$ for all root nodes $a$.
-We let $\event{\G}{}$ denote the set of all internal and sample node-times of the genealogy $\G$;
+We let $\event{G}$ denote the set of all internal and sample node-times of the genealogy $G$;
 we refer to these as \emph{genealogical event times}.
@@
 Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
-Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted with a distinct color for each deme (\cref{fig:geneal}A).
+Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted with a distinct color for each deme (\cref{fig:geneal}).
 Note that a genealogy will in general have \emph{branch-point nodes}, \ie internal nodes with more than one descendant, but may also have internal nodes with only one descendant.
@@
 
-Formally, we define a genealogy, $\G$, to be an ordered pair $(\y,\z)$.
-The $\z$ element specifies the genealogy's tree structure, as follows.
-Let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all partitions of subsets of $\leaves$.
+Formally, we define a genealogy, $G$, to be a triple, $(T,Z,Y)$, where $T=\time(G)\in\Rp$ is the \emph{genealogy time}, $Z$ specifies the genealogy's \emph{tree structure}, and $Y$ gives the \emph{coloring}.
+In particular, let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all collections of finite, mutually-disjoint subsets of $\leaves$.
+That is, an element $\zeta\in{\func{Z}}$ is a partition of the finite set $\bigcup\zeta\subseteq\leaves$.
 Partition \emph{fineness} defines a partial order on $\func{Z}$.
 Specifically, for $\zeta,\zeta'\in{\func{Z}}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
-If $\time(\G)=T$, the tree structure of $\G$ is defined by a \cadlag\ map $\z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $\z_{t_1}\preceq{\z_{t_2}}$.
-An element $b\in\z_t$ is a set of labels;
+The tree structure of $G$ is defined by a \cadlag\ map $Z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $Z_{t_1}\preceq{Z_{t_2}}$.
+An element $b\in{Z_t}$ is a set of labels;
 it represents the branch of the tree that bears the corresponding lineages.
-We use the notation $\event{\z}$ to denote the set of all event-times in the tree $\z$.
-Note that $\event{\z}$ includes the times of all tip, sample, and branch-point nodes, but excludes inline and root nodes.
-Therefore, $\event{\z}\subseteq{\event{\G}}$.
-
-The $\y$ element of $\G$ specifies the coloring of branches and locations of tip, sample, and internal nodes (including inline nodes).
-In particular, if $a$ is the label of any tip or sample node and $t\in[0,T]$, we define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$ and think of this as the color of the corresponding branch at that time.
-We will also need to count the number of nodes along each lineage:
-let $m_{t}(a)$ be the number of internal or sample nodes encountered as one descends along lineage $a$ from time $t=\time(a)$ to $t=0$.
-Observe that $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.
-Formally, we define the coloring component $\y$ as follows.
-Let $\Yspace=\Demes\times\Zp$ and write $\upsilon\in\Yspace$ component-wise as $\upsilon=(\upsilon^d,\upsilon^m)$.
-For any $\zeta\in\func{Z}$, an element $\eta\in\Yspace^{\zeta}$ is a map from the partition $\zeta$ into $\Yspace$.
-Now, suppose we are given a tree $\z:[0,T]\to\func{Z}$ and a \cadlag\ map $\y:[0,T]\to{\coprod_{t\in[0,T]}{\Yspace^{\z_t}}}$.
-Let $\event{\y}$ denote the set of discontinuities of $\y$.
-We say that $\y$ is \emph{compatible} with $\z$ if it is a valid coloring of the tree $\z$.
-Formally, compatibility requires the following:
-\begin{inparaenum}[(i)]
-\item
-  $\event{\y}\supseteq\event{\z}$;
-\item
-  for every $t\in[0,T]$, $\y_t\in{\Yspace^{\z_t}}$;
-\item
-  if $0\le{t_1}\le{t_2}\le{T}$, $b_1\in\z_{t_1}$, $b_2\in\z_{t_2}$, and $b_1\supseteq{b_2}$,
-  then $y^m_{t_1}\le{y^m_{t_2}}$;
-\item $y^m_{t_1}=y^m_{t_2}$ implies $y^d_{t_1}=y^d_{t_2}$.
-\end{inparaenum}
-Let $\func{Y}(\z)$ denote the set of such maps compatible with $\z$ and $\func{Y}_t(\z)\coloneq\CondSet{y_t}{y\in\func{Y}(\z)}$.
-\AAK{[These conditions are necessary but not sufficient.]}
-\AAK{[The coproduct is a fiber bundle over the $\z$ and $\func{Y}(\z)$ is a section of this bundle.]}
-
-The two \cadlag\ functions $\y$ and $\z$ specify a genealogy completely and uniquely.
-When the genealogy is not clear from context, we write $\G^{\y}$ and $\G^{\z}$ to refer to the respective components of genealogy $\G$.
+We use the notation $\event{Z}$ to denote the set of times at which $Z$ is discontinuous.
+Note that $\event{Z}$ includes the times of all tip, sample, and branch-point nodes, but excludes inline and root nodes.
+Therefore, $\event{Z}\subseteq{\event{G}}$.
+
+The third element of $G$ specifies the coloring of branches and locations of tip, sample, and internal nodes (including inline nodes).
+Mathematically, if $G=(T,Z,Y)$, then $Y$ is a \cadlag\ function that maps each point on the genealogy to a deme and a non-negative integer.
+In particular, if $t\in[0,T]$ and $a$ is the label of any tip or sample node,
+$Y_t(a)=(Y_t^{\lab{d}}(a),Y_t^{\lab{m}}(a))\in\Demes\times\Zp$, where $Y_t^{\lab{d}}(a)$ is the deme in which the lineage of $a$ is located at time $t$ and $Y_t^{\lab{m}}(a)$ is the number of internal or sample nodes encountered along the lineage of $a$ in going from time $0$ to time $t$.
+In particular, $Y_t^{\lab{m}}(a)$ is a simple counting process, with $Y_0^{\lab{m}}(a)=0$ for all $a$.
+Since $a,a'\in{b}\in{Z_t}$ implies $Y_t(a)=Y_t(a')$, one can equally well think of $Y_t$ as a map $Z_t\to\Demes\times\Zp$.
+Given a tree $Z$, we let $\func{Y}(Z)$ denote the set of colorings $Y$ that are compatible with $Z$ and $\func{Y}_t(Z)\coloneq\CondSet{Y_t}{Y\in{\func{Y}(Z)}}$.
+
+\AAK{[Formally speaking, the set of possible colorings is a section of a fiber bundle over $Z$, where each fiber is the coloring at a given time.]}
+
+It will sometimes be convenient to make use of notation whereby a genealogy $G=(\time(G),G^{\lab{Z}},G^{\lab{Y}})$.
 
@@
 
-Given a genealogy $\G$, one obtains the \emph{pruned genealogy}, $\P=\prune(\G)$ by first dropping every tip node and then recursively dropping every childless internal node (\cref{fig:upo}A--B).
+Given a genealogy $G$, one obtains the \emph{pruned genealogy}, $P=\prune(G)$ by first dropping every tip node and then recursively dropping every childless internal node (\cref{fig:upo}A--B).
 In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and possibly some of the interior nodes of the genealogy.
@@
 it retains information about where among the demes each of its lineages was through time (\cref{fig:upo}B).
-Note also that a pruned genealogy $\P$ is characterized by functions $\P^\y$ and $\P^\z$ just as an un-pruned genealogy is.
+Note also that a pruned genealogy $P$ is characterized by its time, $\time(P)$ and the functions $P^{\lab{Y}}$ and $P^{\lab{Z}}$ just as an un-pruned genealogy is.
 Finally, observe that, since it contains within itself all of its past history, the pruned genealogy process $\Pr_t=\prune(\Gr_t)$ is Markov, even though the unpruned genealogy process, $\Gr_t$, is not.
@@
 In the following, we will find that we need to count the deme-specific numbers of lineages present in a given pruned genealogy at a given time.
-Accordingly, suppose $\P=(\y,\z)$ is a pruned genealogy with $\time(\P)=T$ and suppose $t\in[0,T]$.
+Accordingly, suppose $P=(T,Z,Y)$ is a pruned genealogy and suppose $t\in[0,T]$.
 Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell=(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
-Clearly, $\ell$ depends only on the values, $(\y_t,\z_t)$, of $y$ and $z$ at time $t$.
-Therefore, we can define $\ell$ as a function such that, whenever $\P=(\y,\z)$ is a pruned genealogy, $\ell(\y_t,\z_t)$ is the vector of deme-specific lineage counts at time $t$.
+Clearly, $\ell$ depends only $Y_t$.
+Therefore, we can define $\ell$ as a function such that, whenever $P=(T,Z,Y)$ is a pruned genealogy, $\ell(Y_t)$ is the vector of deme-specific lineage counts at time $t$.
 We refer to $\ell$ as the \emph{lineage-count} function (cf.~\cref{fig:ells}).
@@
 We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
-In particular, given a node time $t$ in a pruned genealogy $\P=(\y,\z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
+In particular, given a node time $t$ in a pruned genealogy $P=(T,Y,Z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
 Like the lineage-count function, $s$ depends only on the local structure of $\P$.
-However, $s$ depends not only on $(\y_t,\z_t)$, but also on $(\yt_t,\zt_t)$.
-Thus, we can define the \emph{saturation} function such that, whenever $\P=(\y,\z)$ is a pruned genealogy, $s(\yrt_t,\yr_t,\zrt_t,\zr_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
+However, $s$ depends not only on $Y_t$, but also on $\leftlim{Y}_t$.
+Thus, we can define the \emph{saturation} function such that, whenever $P=(T,Y,Z)$ is a pruned genealogy, $s(\leftlim{Y}_t,Y_t)$ is the integer vector of deme-specific numbers of emerging lineages at time $t$.
 \Cref{fig:ells} illustrates.
@@
 
-Suppose $\P$ is a pruned genealogy, with $\time(\P)=T$ and $t\in\event{\P}{}$.
-The local structure of $\P$ at $t$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
-For example, if the event in $\P$ at $t$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
-Similarly, if the node in $\P$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
-To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $y$ and $z$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
-That is, $Q_u(\eta,\eta',\zeta,\zeta')=1$ if and only if
-there is a feasible genealogy, $\G=(\y,\z)$, and history, $\H$,
-and a $t\in{[0,T]}$ such that,
-given $\Gr_T=\G$ and $\Hr_T=\H$,
-$\U_t=u$,
-$\yt_t=\eta$, $\y_t=\eta'$,
-$\zt_t=\zeta$, and $\z_t=\zeta'$.
+Suppose $P$ is a pruned genealogy, with $\time(P)=T$ and $t\in\event{P}$.
+The local structure of $P$ at $t$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
+For example, if the event in $P$ at $t$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
+Similarly, if the node in $P$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
+To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $P^{\lab{Y}}$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
+That is, $Q_u(\eta,\eta')=1$ if and only if
+there is a feasible genealogy, $\G=(\T,\Z,\Y)$, and history, $\H$,
+and a $t\in{[0,\T]}$ such that,
+given $\Gr_\T=\G$ and $\Hr_\T=\H$,
+$\U_t=u$, $\Yt_t=\eta$, and $\Y_t=\eta'$.
 We refer to $Q$ as the \emph{compatibility indicator}.
@@
 
-The \emph{obscured genealogy} is obtained by discarding all information about demes and events not visible from the topology of the tree alone.
-In particular, if $\P=(\y,\z)$ is a pruned genealogy, we write $\obs(\P)=\P^{\z}=\z$ to denote the obscured genealogy.
+The \emph{obscured genealogy} is obtained by discarding all information about demes and events not visible from the topology of the tree alone (\cref{fig:upo}B--C).
+In particular, if $P=(T,Z,Y)$ is a pruned genealogy, we write $\obs(P)=(T,Z)$ to denote the obscured genealogy.
 
@@
 \begin{thm}\label{thm:pruned_lik}
-  Suppose $\P^*=(\y^*,\z^*)$ is a given pruned genealogy with $\time(\P^*)=T$.
+  Suppose $\P=(\T,\Z,\Y)$ is a given pruned genealogy.
   Define
   \begin{equation}\label{eq:phidef}
-    \phi_u(x,y,y',z,z')\coloneq\BinRatio{n(x)}{\ell(y',z')}{r^{u}}{s(y,y',z,z')}\,Q_{u}(y,y',z,z'),
+    \phi_u(x,y,y')\coloneq\BinRatio{n(x)}{\ell(y')}{r^{u}}{s(y,y')}\,Q_{u}(y,y'),
   \end{equation}
@@
   \begin{equation*}
-    \CondProb{\Pr_T=\P^*}{\Hr_T=\H}=\Indicator{\event{\H}{}\supseteq{\event{\P}{}^*}}\,\prod_{t\in\event{\H}{}}{\phi_{\U_t}(\X_t,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}.
+    \CondProb{\Pr_\T=\P}{\Hr_\T=\H}=\Indicator{\event{\H}\supseteq{\event{\P}}}\,\prod_{t\in\event{\H}}{\phi_{\U_t}(\X_t,\Yt_t,\Y_t)}.
   \end{equation*}
@@
 \begin{proof}
-  If $\event{\H}{}\nsupseteq\event{\P}{}^*$, then $\H$ and $\P^*$ are incompatible and $\CondProb{\Pr_T=\P^*}{\Hr_T=\H}=0$.
-  Similarly, if any event of $\H$ is incompatible with the local structure of $\P^*$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_T=\P^*}{\Hr_T=\H}=0$.
+  If $\event{\H}\nsupseteq\event{\P}$, then $\H$ and $\P$ are incompatible and $\CondProb{\Pr_\T=\P}{\Hr_\T=\H}=0$.
+  Similarly, if any event of $\H$ is incompatible with the local structure of $\P$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_\T=\P}{\Hr_\T=\H}=0$.
   Let us therefore suppose that neither of these conditions hold.
-  Conditional on $\Hr_T=\H$, at each time $t\in\event{\H}{}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
-  In $\P^*$, at time $t$, there are $\ell_i=\ell_i(\y^*_t,\z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)$ are emergent.
+  Conditional on $\Hr_\T=\H$, at each time $t\in\event{\H}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
+  In $\P$, at time $t$, there are $\ell_i=\ell_i(\Y_t)$ lineages in deme $i$, of which $s_i=s_i(\Yt_t,\Y_t)$ are emergent.
   By assumption, at each genealogical event, lineages within a deme are exchangeable:
@@
   Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct;
-  just one of these is the one present in $\P^*$.
-  Moreover, since, again conditional on $\Hr_T=\H$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
-  \begin{equation*}
-    \CondProb{\Pr_T=\P^*}{\Hr_T=\H}=\prod_{t\in\event{\H}{}}{\BinRatio{n(\X_t)}{\ell(\y^*_t,\z^*_t)}{r^{\U_t}}{s(\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}}.
-  \end{equation*}
-  Returning to the possibility that $\H$ is incompatible with $\P^*$, since $\Prob{\Pr_T=\P^*}=0$ if either any $Q_u=0$ or $\event{\P}{}^*\nsubseteq\event{\H}{}$, we obtain the result.
+  just one of these is the one present in $\P$.
+  Moreover, since, again conditional on $\Hr_\T=\H$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
+  \begin{equation*}
+    \CondProb{\Pr_\T=\P}{\Hr_\T=\H}=\prod_{t\in\event{\H}}{\BinRatio{n(\X_t)}{\ell(\Y_t)}{r^{\U_t}}{s(\Yt_t,\Y_t)}}.
+  \end{equation*}
+  Returning to the possibility that $\H$ is incompatible with $\P$, since $\Prob{\Pr_\T=\P}=0$ if either any $Q_u=0$ or $\event{\P}\nsubseteq\event{\H}$, we obtain the result.
 \end{proof}
@@
 
-\begin{thm}
-  Suppose that $\P^*=(\y^*,\z^*)$ is a pruned genealogy with $\time(\P^*)=T>0$.
+\begin{corol}
+  Suppose that $\P=(\T,\Z,\Y)$ is a given pruned genealogy.
   Suppose that $w=w(t,x)$ satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
@@
       \frac{\partial{w}}{\partial{t}}(t,x)=
-      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)\,\dd{x'}}}
+      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\Yt_t,\Y_t)\,\dd{x'}}}
       -\sum_u{\int{w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}}},
-      &t\notin{\event{\P}{}^*},\\
-      w(t,x)=&\sum_u{\int{\wt(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)\,\dd{x'}}},
-      &t\in{\event{\P}{}^*},
+      &t\notin{\event{\P}},\\
+      w(t,x)=&\sum_u{\int{\wt(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\Yt_t,\Y_t)\,\dd{x'}}},
+      &t\in{\event{\P}},
     \end{aligned}
@@
   where $\phi$ is defined in \cref{eq:phidef}.
-  Then the likelihood of $\P^*$ is
-  \begin{equation*}
-    \lik(\P^*)=\int{w(T,x)\,\dd{x}}.
-  \end{equation*}
-\end{thm}
+  Then the likelihood of $\P$ is
+  \begin{equation*}
+    \lik(\P)=\int{w(T,x)\,\dd{x}}.
+  \end{equation*}
+\end{corol}
 
@@
 \begin{thm}\label{thm:obsc_lik}
-  Suppose $\z^*$ is a given obscured genealogy with $\time(\z^*)=T$.
-  Let $q$ and $\pi$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\func{Y}_t(\z^*)$,
+  Suppose that $(\T,\Z)$ is a given obscured genealogy.
+  Let $q$ and $\pi$ be probability kernels, such that
+  for all $x\in\Xspace$ and $y\in\func{Y}_0(\Z)$,
   \begin{equation*}
@@
       q(x,y)\ge{0},\qquad
-      \sum_{y\in\func{Y}_0(\z^*)}{q(x,y)}=1, \qquad
+      \sum_{y\in\func{Y}_0(\Z)}{q(x,y)}=1,
+    \end{gathered}
+  \end{equation*}
+  and, for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\func{Y}_t(\Z)$,
+  \begin{equation*}
+    \begin{gathered}
       \pi_u(t,x,x',y,y')\ge{0},\qquad
-      \sum_{y'\in\func{Y}_t(\z^*)}{\pi_u(t,x,x',y,y')}=1.
+      \sum_{y'\in\func{Y}_t(\Z)}{\pi_u(t,x,x',y,y')}=1.
     \end{gathered}
   \end{equation*}
-  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $\alpha_u(t,x,x')\,Q_u(y,y',\zt^*_t,\z^*_t)>0$
-  and that $q(x,y)>0$ whenever $\CondProb{\Pr_0^\y=y}{\Xr_0=x}>0$.
-  Then there is a stochastic jump process $\yr_t$ with sample paths in $\func{Y}(\z^*)$ such that $(\Xr_t,\Ur_t,\yr_t)$ is Markov and
-  \begin{equation*}
-    \CondProb{\Pr^{\z}_T=\z^*}{\Hr_T=\H}=
-    \Indicator{\event{\H}{}\supseteq\event{\z}{}^*}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
+  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $\alpha_u(t,x,x')\,Q_u(y,y')>0$
+  and that $q(x,y)>0$ whenever $\CondProb{\Pr_0^{\lab{Y}}=y}{\Xr_0=x}>0$.
+  Then there is a stochastic jump process $\yr_t$ with sample paths in $\func{Y}(\Z)$ such that $(\Xr_t,\Ur_t,\yr_t)$ is Markov and
+  \begin{equation*}
+    \CondProb{\Pr^{\lab{Z}}_\T=\Z}{\Hr_\T=\H}=
+    \Indicator{\event{\H}\supseteq\event{\Z}}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
   \end{equation*}
@@
   \begin{equation}\label{eq:IS1}
-    \CondProb{\zr_T=\z^*}{\Hr_T=\H}=\CondExpect{\Indicator{\Pr^{\z}_T=\z^*}}{\Hr_T=\H}.
+    \CondProb{\Pr^{\lab{Z}}_\T=\Z}{\Hr_\T=\H}=\CondExpect{\Indicator{\Pr^{\lab{Z}}_\T=\Z}}{\Hr_\T=\H}.
   \end{equation}
   Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
-  we will propose pruned genealogies compatible with $\z^*_t$ as sample paths from a stochastic process driven by $\Xr_t$ and
+  we will propose pruned genealogies compatible with $\Z$ as sample paths from a stochastic process driven by $\Xr_t$ and
   evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
-  Conditional on $\Hr_T=\H$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$ such that
+  Conditional on $\Hr_\T=\H$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$ such that
   \begin{equation*}
     \begin{gathered}
-      \CondProb{\yrh_0}{\Hr_T=\H}=q(\X_0,\yrh_0),
+      \CondProb{\yrh_0}{\Hr_\T=\H}=q(\X_0,\yrh_0),
       \qquad
-      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_T=\H}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
+      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_\T=\H}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
     \end{gathered}
   \end{equation*}
-  The required process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
-  This construction of $\yr_t$ obviously guarantees that $\event{\H}{}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$ and that $(\Xr_t,\Ur_t,\yr_t)$ is Markov.
-
-  Now, for $\y\in\func{Y}(\z^*)$, let us define $C(\y)=(\y,\z^*)$.
-  Then, by construction, $\obs(C(\y))=\z^*$ and,
-  conversely, for every pruned genealogy $\P$ satisfying $\P^{\z}=\z^*$, $C(\P^{\y})=\P$.
-  Moreover, the conditions on the kernels $q$ and $\pi$ guarantee that, if $\CondProb{\Pr_T=\P}{\Hr_T=\H}>0$ and $\P^{\z}=\z^*$, then $\Prob{\yr=\P^{\y}}>0$.
+  The required process $\yr_t$ is the unique \cadlag\ process with event times $\Th_k$ and $\yrh_k$ as its embedded chain.
+  This construction of $\yr_t$ obviously guarantees that $\event{\H}\supseteq\event{\yr}\supseteq\event{\Z}$ and that $(\Xr_t,\Ur_t,\yr_t)$ is Markov.
+
+  Now, for $\y\in\func{Y}(\Z)$, let us define $C(\y)=(\T,\Z,\y)$.
+  Then, by construction, $\obs(C(\y))=(\T,\Z)$ and,
+  conversely, for every pruned genealogy $\P$ satisfying $\time(\P)=\T$ and $\P^{\lab{Z}}=\Z$, $C(\P^{\lab{Y}})=\P$.
+  Moreover, the conditions on the kernels $q$ and $\pi$ guarantee that, if $\CondProb{\Pr_\T=\P}{\Hr_\T=\H}>0$ and $\P^{\lab{Z}}=\Z$, then $\CondProb{\yr=\P^{\lab{Y}}}{\Hr_\T=\H}>0$.
   We therefore have that
   \begin{equation*}
-    \CondProb{\zr_T=\z^*}{\Hr_T=\H}=
-    \Expect{\frac{\CondProb{\Pr_T=C(\yr)}{\Hr_T=\H}}{\pi(\yr\vert\H)}},
+    \CondProb{\Pr^{\lab{Z}}_\T=\Z}{\Hr_\T=\H}=
+    \Expect{\frac{\CondProb{\Pr_\T=C(\yr)}{\Hr_\T=\H}}{\pi(\yr\vert\H)}},
   \end{equation*}
@@
   \begin{equation*}
-    \pi(\yr\vert\H)=q(\X_0,\yr_0)\,\prod_{t\in\event{\H}{}}{\pi_{\U_{t}}(t,\Xt_{t},\X_{t},\yrt_{t},\yr_{t})}.
+    \pi(\yr\vert\H)=q(\X_0,\yr_0)\,\prod_{t\in\event{\H}}{\pi_{\U_{t}}(t,\Xt_{t},\X_{t},\yrt_{t},\yr_{t})}.
   \end{equation*}
@@
 
-\begin{thm}
-  Let $\z^*$ be a given obscured genealogy and $T=\time(\z^*)$.
+\begin{corol}
+  Let $V=(\T,\Z)$ be a given obscured genealogy.
   Then there are probability kernels $q$ and $\pi$ as in \cref{thm:obsc_lik} such that if
@@
       \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),\qquad
-      \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y',\zt^*_t,\z^*_t)}{\pi_u(t,x,x',y,y')},
+      \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y')}{\pi_u(t,x,x',y,y')},
     \end{gathered}
@@
   and if $w=w(t,x,y)$ satisfies
-  the initial condition $w(0,x,y)=p_0(x)\,q(x,y)$
+  the initial condition $w(0,x,y)=p_0(x)\,\Indicator{q(x,y)>0}$
   and the filter equation
@@
       -\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}},
-      &t\in{\event{\z}{}^*},\\
+      &t\in{\event{\Z}},\\
       &w(t,x,y)=\sum_{uy'}{\int{\wt(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}},
-      &t\in{\event{\z}{}^*},
+      &t\in{\event{\Z}},
     \end{aligned}
   \end{equation*}
-  then the likelihood of $\z^*$ is
-  \begin{equation*}
-    \lik(\z^*)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
-  \end{equation*}
-\end{thm}
+  then the likelihood of $V$ is
+  \begin{equation*}
+    \lik(V)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
+  \end{equation*}
+\end{corol}
+
+Lemmas in the appendix show how this can be computed via Sequential Monte Carlo.
 
@@
 
+\subsection{SIRS}
+
+Was shown in an earlier paper.
+[Display filter equation and plot a curve.]
+
 \subsection{SEIRS}
 
-Jumps: $\Jumps=\Set{\lab{Inf},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Birth},\lab{Death_S},\lab{Death_E},\lab{Death_I},\lab{Death_R},\lab{Sample}}$.
-
-Demes: $\Demes=\Set{\lab{E},\lab{I}}$.
-
-Jump rates:
-\begin{itemize}
-\item $\alpha_{\lab{Inf}}(t,x,x')=\beta(t)\,\frac{x^{\lab{S}}x^{\lab{I}}}{N(t)}\,\Indicator{x'=x+(-1,1,0,0)}$
-\item $\alpha_{\lab{Prog}}(x,x')=\rho\,x^{\lab{E}}\,\Indicator{x'=x+(0,-1,1,0)}$
-\item $\alpha_{\lab{Recov}}(x,x')=\gamma\,x^{\lab{I}}\,\Indicator{x'=x+(0,0,-1,1)}$
-\item $\alpha_{\lab{Wane}}(x,x')=\upsilon\,x^{\lab{R}}\,\Indicator{x'=x+(1,0,0,-1)}$
-\item $\alpha_{\lab{Sample}}(t,x,x')=\psi\,x^I\,\Indicator{x'=x}$
-\item $\alpha_{\lab{Birth}}(t,x,x')=B(t)\,\Indicator{x'=x+(1,0,0,0)}$
-\item $\alpha_{\lab{Death}_k}(x,x')=\mu\,x^k\,\Indicator{x'^j=x^j-\delta_{jk}}$, $k\in\Set{\lab{S},\lab{E},\lab{I},\lab{R}}$
-\end{itemize}
-
-%% Nonlinear filtering equation:
-%% \begin{equation}
-%%   \begin{aligned}
-%%     \frac{\partial{w}}{\partial{t}}(t,x,y,z)=&B(t)\,\left[\sum_{y'z'}w\big(t,x-(1,0,0,0),y',z'\big)-w\big(t,x,y,z\big)\right]\\
-%%     &\qquad+\sum_{k\in\{\lab{S},\lab{E},\lab{I},\lab{R}\}}\mu\,(x^k+1)\,w(t,x+\delta_{k})-\mu\,x^k\,w(t,x)\\
-%%     &\qquad+\upsilon\,x^{\lab{R}}\,w(t,x-(1,0,0,-1))\\
-%%   \end{aligned}
-%% \end{equation}
+Display filter equation.
+Discuss the choice of $\pi$ and $q$.
+Present a likelihood curve.
+
+\subsection{Two-strain competition}
+
+Display filter equation.
+Discuss the choice of $\pi$ and $q$.
+Present a likelihood curve.
+
+\subsection{Superspreading model}
+
+Display filter equation.
+Discuss the choice of $\pi$ and $q$.
+Present a likelihood curve.
+
+\subsection{Linear birth-death and Moran models}
+
+Exact solution of the filtering equations is possible in the unstructured case, yielding precisely the likelihoods for these two models.
+This establishes that the present theory is a strict generalization of these cases.
+It does away with the need for large sample-size and small sample-fraction approximations, the necessity of assuming slow changes in effective population size, and the need for linearization.
 
@@
 Freedom to choose importance sampling distribution.
-Allowable to borrow information from the future.
+It is permissible to borrow information from the future.
 
@@
   \begin{equation*}
-    \CondExpect{\Vr_t}{\Hr_t=\H}=\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)},
+    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H}_t}{B(e,\Xt_e,\X_e)},
   \end{equation*}
@@
   \begin{equation*}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\Indicator{\event{\H}{t}\supseteq\mathcal{M}_t}\,\prod_{e\in\event{\H}{t}}{B(e,\Xt_e,\X_e)}.
+    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\Indicator{\event{\H}_{t}\supseteq\mathcal{M}_t}\,\prod_{e\in\event{\H}_{t}}{B(e,\Xt_e,\X_e)}.
   \end{equation*}
@@
 \begin{lemma}
-  Showing how \cref{eq:two-part-filter} can be stated as a singular filter equation.
-  \begin{equation}
-    \label{eq:sing-filter-eq}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)
-      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
-      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}\\
-      &+\sum_e\int{w(t,x')\,\delta(t,e)\,\pi(t,x',x)\,C(t,x',x)\,\dd{x'}}
-      -\sum_e\int{w(t,x)\,\delta(t,e)\,\pi(t,x,x')\,\dd{x'}}\\
-      &-\lambda(t,x)\,w(t,x).\\
-    \end{aligned}
-  \end{equation}
+  Showing how \cref{eq:two-part-filter} can be stated as a singular filter equation, \ie with driver $\beta$ and boost $C$, where
+  \begin{equation*}
+    \begin{gathered}
+      A(t,x)=\sum_{x'}{\alpha(t,x,x')}, \qquad
+      \beta(t,x,x')=\alpha(t,x,x')+\sum_{e\in\mathcal{M}}{\delta(t,e)\,\frac{\alpha(t,x,x')}{A(t,x)}}\,,\\
+      C(t,x,x')=B(t,x,x')+\sum_{e\in\mathcal{M}}{\delta(t,e)\,A(t,x)}.
+    \end{gathered}
+  \end{equation*}
+  %% \begin{equation}
+  %%   \label{eq:sing-filter-eq}
+  %%   \begin{aligned}
+  %%     \frac{\partial{w}}{\partial{t}}(t,x)
+  %%     =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
+  %%     -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}\\
+  %%     &+\sum_e\int{w(t,x')\,\delta(t,e)\,\pi(t,x',x)\,C(t,x',x)\,\dd{x'}}
+  %%     -\sum_e\int{w(t,x)\,\delta(t,e)\,\pi(t,x,x')\,\dd{x'}}\\
+  %%     &-\lambda(t,x)\,w(t,x).\\
+  %%   \end{aligned}
+  %% \end{equation}
 \end{lemma}

### 2024-02-22 ms.Rnw
corrections

@@
 We view $\Grh_k$ as the embedded chain of the continuous-time genealogy process $\Gr_t$.
-It is very important to note that, conditional on $\Urh_k$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of the choices at any other times and independent of $(\Trh_k,\Xrh_k)$ for all $k$.
+It is very important to note that, conditional on $(\Xrh_k,\Urh_k)$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of these choices at any other times and independent of $(\Xrh_j,\Urh_j)$ for all $j\ne{k}$.
 Moreover, by construction, any lineage within a deme is as likely as any other lineage in that deme to be selected as a parent or for death, sampling, or migration.
 We refer to this property as the \emph{exchangeability} of lineages within a deme.
+Finally, note that $\Gr_t$ does not have the Markov property, though $(\Xr_t,\Ur_t,\Gr_t)$ and $(\Xr_t,\Gr_t)$ do.
 
@@
 
+\begin{remark}
+  Note that, since $\func{Y}_t(\Z)$ is finite, it is permissible to choose $q$ and $\pi$ to be uniform.
+\end{remark}
+
 Lemmas in the appendix show how this can be computed via Sequential Monte Carlo.
@@
 
+\AAK{[We should define the singular filter equations (\ie those with a discrete portion) also.]}
+
 The following results show how filter equations allow one to integrate over random histories.
@@
   \label{lemma:sing-filt}
-  Suppose $\mathcal{M}\subset{\Rp}$ is finite and let $\mathcal{M}_t\coloneq\mathcal{M}\cap{[0,t]}$.
+  Suppose $\mathcal{M}\subset{\Rp}$ is countable such that $\mathcal{M}_t\coloneq\mathcal{M}\cap{[0,t]}$ is finite for all $t$.
   Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable and

### 2024-03-22 figs/event_types.tex
start examples; more edits
AIM seminar

@@
 \begin{center}
-  \resizebox{0.8\linewidth}{!}{
+  \resizebox{0.7\linewidth}{!}{
     \begin{tikzpicture}[scale=2]
@@
   \textbf{(A)} A birth-type jump results in the branching of one or more child lineages from the parent.
-  There can be only one parent, though the child lineages may be in demes different from those of their parent.
+  There can be only one parent, though the demes of the child lineages may differ from that of their parent.
   Here, a parent of the blue deme sires one child lineage in each of the blue and yellow demes.
-  The production of an event is an integer vector, with one entry for each deme.
+  The \emph{production} of an event is an integer vector, with one entry for each deme.
   The production of this event is therefore $r=(r_{\lab{blue}},r_{\lab{yellow}})=(2,1)$.
+  The \emph{deme occupancy} of an event is the number of lineages in each deme just to the right of the event.
+  The deme occupancy at this event is therefore $n=(n_{\lab{blue}},n_{\lab{yellow}})=(3,5)$.
   \textbf{(B)} A death-type event causes the extinction of a lineage.
-  Since internal node without children are recursively removed, the affected branch disappears.
-  The production of this event is $r=(0,0)$.
+  Since internal nodes without children are recursively removed, the affected branch is dropped.
+  The production of this event is $r=(0,0)$ and the deme occupancy is $n=(3,4)$.
   \textbf{(C)} A migration-type event results in the movement of one or more lineages from one deme to another.
@@
   The production of this event is $r=(1,0)$, \ie the production is 1 for the blue deme and 0 for the yellow.
+  The deme occupancy is $n=(6,2)$.
   \textbf{(D)} In a sample-type event, one or more sample nodes (blue circles) are inserted.
   Here, there are two samples, one in each of the blue and yellow demes.
-  Accordingly, its production is $r=(1,1)$.
-  \textbf{(E)} A neutral-type event has no effect on the genealogy and zero production in all demes: $r=(0,0)$.
+  Accordingly, $r=(1,1)$ and $n=(2,6)$.
+  \textbf{(E)} A neutral-type event has no effect on the genealogy and zero production in all demes: $r=(0,0)$, $n=(5,3)$.
   \textbf{(F)} The theory presented here allows for compound events.
   As an example, here a birth/death-type event occurs, wherein one yellow lineage is extinguished and a blue lineage simultaneously sires a blue child.
-  The production of this event is $r=(2,0)$.
-  \textbf{(G)} Here, a compound sample/death-type event with $r=(0,0)$ occurs.
+  For this event, we have $r=(2,0)$ and $n=(6,2)$.
+  \textbf{(G)} Here, a compound sample/death-type event with $r=(0,0)$ and $n=(2,5)$ occurs.
   A blue lineage is sampled and simultaneously extinguished.
   Note that recursive removal does not occur, since sample nodes are never removed.
-  \textbf{(H)} A compound birth/migration-type event with $r=(4,0)$.
+  \textbf{(H)} A compound birth/migration-type event with $r=(4,0)$ and $n=(6,2)$.
   \label{fig:event_types}

### 2024-03-22 figs/twodudes.tex
AIM seminar

\begin{tikzpicture}[scale=0.6]
  \usetikzlibrary{calc}
  \tikzstyle{branch}=[color=black, line width=4pt, line join=miter]
  \coordinate (origin) at (0,0);
  \coordinate (GB) at (-0.5,0);
  \coordinate (bl) at (-6,0);
  \coordinate (S1) at (5,9);
  \coordinate (S2) at (5,6);
  \coordinate (B1) at (0,4);
  \coordinate (B2) at (0,1);
  \node[anchor=west,outer sep=1ex] (Seq1) at (S1.east) {{ACCGTATCGCCA}};
  \node[anchor=west,outer sep=1ex] (Seq2) at (S2.east) {{ACCG\textcolor{red}{GC}TCG\textcolor{red}{G}CA}};
  \coordinate (R1) at (S1 |- B1);
  \coordinate (R2) at (S2 |- B2);
  \coordinate (Bmid) at ($(B1)!0.5!(B2)$);
  \coordinate (Smid) at ($(S1)!0.5!(S2)$);
  \coordinate (L) at (bl |- Bmid);
  \node[anchor=west] (R1i) at (R1.east) {\includegraphics[height=36pt]{infected}};
  \node[anchor=west] (R2i) at (R2.east) {\includegraphics[height=36pt]{infected}};
  \draw[branch] (R1i.west) -- (B1) -- (Bmid) -- (L) (R2i.west) -- (B2) -- (Bmid);
  \node[anchor=east] (Li) at (L.west) {\includegraphics[height=36pt]{infected}};
  \draw[branch,color=blue] (S1) -- (GB |- S1) -- (GB |- S2) -- (S2);
  \node[anchor=east,outer sep=1em] (lab1) at (Li.west) {\parbox[t]{72pt}{\raggedright Transmission tree}};
  \node[outer sep=1em] (lab2) at (lab1 |- Smid) {\parbox[t]{72pt}{\raggedright Pathogen genealogy}};
  \coordinate (tc) at ($(origin|-Seq1.north)+(0,1em)$);
  \draw[very thin,color=gray] (origin) -- (tc);
  \node (tlab) at (Seq1|-origin) {$t$};
  \draw[thin,->,>=latex] (lab1 |- bl) -- (tlab);
\end{tikzpicture}

### 2024-03-22 ms.Rnw
minor edits from ELI
AIM seminar

@@
 
-Advent of routine molecular surveillance.
-With certain pathogens, with matching of time-scales, information on transmission is possible.
-With sufficiently dense sampling, it is sometimes possible to reconstruct WAIFW.
-More typically, one wishes to gain insight into the structure and parameterization of the transmission process.
-This is accomplished by formalizing a transmission model or suite of transmission models, estimating parameters, and comparing their ability to explain the data.
+With the advent of inexpensive genome sequencing technologies, routine molecular surveillance has become an important tool in the epidemiologist's kit.
+When an infectious agent evolves quickly enough so that information on its history of transmission accumulates in its genome, yet not so quickly that it is dissipated, it is possible to extract this information to gain insight into the determinants of transmission.
+With sufficiently dense sampling, it can be possible to reconstruct the specific history of who has infected whom, which can aid identification of the correlates of transmission and immunity and the features of infections.
+More typically, one can attempt to learn about these features via the estimation of a model of the transmission process on the basis of relatively sparse samples.
+In particular, one can formalize one or more mathematical models of transmission, estimate their parameters, and comparing their ability to explain data, following a standard statistical paradigm.
+For the purposes of this paper, we refer to the latter program as \emph{phylodynamics}.
 
@@
 Factorization of problem into two subproblems.
+Problems with this approach \citep{Smith2017}.
 
@@
 approximation no longer needed.
+
+Modeling of the sampling process.
 
@@
 
-If $\Phir_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Phirh_k\coloneq{\Phir_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
+If $\Phir_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Phirh^{}_k\coloneq{\Phir_{\Trh^{}_k}}$, $k=0,\dots,\Kr^{}_t$.
 In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
@@
 Its state vector is $x=(S,I,R)$ and its KFE is
+\fontsize{10pt}{11pt}\selectfont
 \begin{equation*}
@@
 \end{equation*}
+\normalfont
 Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.
@@
 It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
+\fontsize{10pt}{11pt}\selectfont
 \begin{equation*}
@@
 \end{equation*}
+\normalfont
 where $N=S+E+I+R$ is the total population size.
@@
 This model has KFE
+\fontsize{10pt}{11pt}\selectfont
 \begin{equation*}
@@
 \end{equation*}
+\normalfont
 
@@
 The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
+\fontsize{10pt}{11pt}\selectfont
 \begin{equation*}
@@
 
+\fontsize{10pt}{11pt}\selectfont
 \begin{equation*}
@@
 \subsection{History}
-
-\AAK{[Maybe it would be better to speak instead of $\Hr_t$ being the random $\sigma$-algebra generated by the sequences $(\Trh_k,\Xrh_k,\Urh_k)_{k=1}^{\Kr_t}$.
-    The only thing we do with $\Hr$ is condition on it.
-    One (perhaps minor) problem with this is that we cannot then speak about the measure on histories.
-]}
 
@@
 $\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
-In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
+In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic, as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
 The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
+\fontsize{10pt}{11pt}\selectfont
 \begin{equation*}
   \label{eq:Hdens}
-  \begin{aligned}
   \pi^{\mathrm{H}}(\dd{\H_t})
-    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
-    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
-    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
-  \end{aligned}
-\end{equation*}
-where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.
-
-If $\H$ is such a history, we define $\time(\H)$ to be the right endpoint of its domain and use the notation $\event{\H}=\Set{\Th_1,\dots,\Th_{\K_t}}\subset{[0,\time(\H)]}$ to denote the set of its jump times.
+  =p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
+  \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}
+  \,\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
+\end{equation*}
+\normalfont
+where again, by convention, $\Th^{}_0=0$ and $\Th^{}_{K^{}_t+1}=t$.
+
+If $\H$ is such a history, we define $\time(\H)$ to be the right endpoint of its domain and use the notation $\event{\H}\coloneq\Set{\Th^{}_1,\dots,\Th^{}_{\K_t}}\subset{[0,\time(\H)]}$ to denote the set of its jump times.
 
@@
 Note that a genealogy will in general have \emph{branch-point nodes}, \ie internal nodes with more than one descendant, but may also have internal nodes with only one descendant.
+We refer to such nodes as \emph{inline nodes}.
 These occur whenever the color changes along a branch, but can also occur without a color-change.
-We refer to such nodes as \emph{inline nodes}.
 
 Formally, we define a genealogy, $G$, to be a triple, $(T,Z,Y)$, where $T=\time(G)\in\Rp$ is the \emph{genealogy time}, $Z$ specifies the genealogy's \emph{tree structure}, and $Y$ gives the \emph{coloring}.
-In particular, let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all collections of finite, mutually-disjoint subsets of $\leaves$.
-That is, an element $\zeta\in{\func{Z}}$ is a partition of the finite set $\bigcup\zeta\subseteq\leaves$.
-Partition \emph{fineness} defines a partial order on $\func{Z}$.
-Specifically, for $\zeta,\zeta'\in{\func{Z}}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
-The tree structure of $G$ is defined by a \cadlag\ map $Z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $Z_{t_1}\preceq{Z_{t_2}}$.
+In particular, let $\leaves$ be a countable set of labels and let $\part(\leaves)$ be the set of all collections of finite, mutually-disjoint subsets of $\leaves$.
+That is, an element $\zeta\in{\part(\leaves)}$ is a partition of the finite set $\bigcup\zeta\subseteq\leaves$.
+Partition \emph{fineness} defines a partial order on $\part(\leaves)$.
+Specifically, for $\zeta,\zeta'\in{\part(\leaves)}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
+The tree structure of $G$ is defined by a \cadlag\ map $Z:[0,T]\to\part(\leaves)$ that is monotone in the sense that $t_1\le{t_2}$ implies $Z_{t_1}\preceq{Z_{t_2}}$.
 An element $b\in{Z_t}$ is a set of labels;
@@
 A generative model must account for this loss of information.
-We therefore now describe how genealogies are \emph{pruned} to yield sample-only genealogies and then \emph{obscured} via the erasure of color from their branches.
+We therefore now describe how genealogies are \emph{pruned} to yield sample-only genealogies and then \emph{obscured} via the erasure of color from their branches (\cref{fig:upo}).
 
@@
 Similarly, if the node in $P$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
-To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $P^{\lab{Y}}$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
+To succinctly accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $P^{\lab{Y}}$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
 That is, $Q_u(\eta,\eta')=1$ if and only if
@@
 given $\Gr_\T=\G$ and $\Hr_\T=\H$,
-$\U_t=u$, $\Yt_t=\eta$, and $\Y_t=\eta'$.
+we have $\U_t=u$, $\Yt_t=\eta$, and $\Y_t=\eta'$.
 We refer to $Q$ as the \emph{compatibility indicator}.
@@
   \begin{equation*}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H}_t}{B(e,\Xt_e,\X_e)},
+    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H_t}}{B(e,\Xt_e,\X_e)},
   \end{equation*}
@@
   \begin{equation*}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\Indicator{\event{\H}_{t}\supseteq\mathcal{M}_t}\,\prod_{e\in\event{\H}_{t}}{B(e,\Xt_e,\X_e)}.
+    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\Indicator{\event{\H_{t}}\supseteq\mathcal{M}_t}\,\prod_{e\in\event{\H_{t}}}{B(e,\Xt_e,\X_e)}.
   \end{equation*}

### 2024-03-28 ms.Rnw
Lemma A2?

@@
 
+Relations to lookdown construction of \citet{Etheridge2019}.
+
 Classes of Markov processes.
@@
     \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
-    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
+    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}+o(\Delta).
   \end{gathered}
@@
 Its state vector is $x=(S,I,R)$ and its KFE is
+\begin{equation*}
   \fontsize{10pt}{11pt}\selectfont
-\begin{equation*}
   \begin{split}
@@
   \end{split}
-\end{equation*}
   \normalfont
+\end{equation*}
 Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.
@@
 It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
+\begin{equation*}
   \fontsize{10pt}{11pt}\selectfont
-\begin{equation*}
   \begin{split}
@@
   \end{split}
-\end{equation*}
   \normalfont
+\end{equation*}
 where $N=S+E+I+R$ is the total population size.
@@
 This model has KFE
+\begin{equation*}
   \fontsize{10pt}{11pt}\selectfont
-\begin{equation*}
   \begin{split}
@@
   \end{split}
-\end{equation*}
   \normalfont
+\end{equation*}
 
@@
 The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
+\begin{equation*}
   \fontsize{10pt}{11pt}\selectfont
-\begin{equation*}
   \begin{split}
@@
 The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
-\fontsize{10pt}{11pt}\selectfont
-\begin{equation*}
+\begin{equation}
   \label{eq:Hdens}
+  \fontsize{9pt}{11pt}\selectfont
   \pi^{\mathrm{H}}(\dd{\H_t})
@@
   \,\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
-\end{equation*}
   \normalfont
+\end{equation}
 where again, by convention, $\Th^{}_0=0$ and $\Th^{}_{K^{}_t+1}=t$.
@@
 \begin{corol}
+  \label{thm:pruned_uncond}
   Suppose that $\P=(\T,\Z,\Y)$ is a given pruned genealogy.
@@
 
+Note that, since $\func{Y}_t(\Z)$ is finite, it is permissible, for example, to choose $q$ and $\pi$ to be uniform.
+
 The final result shows how to compute the likelihood of an obscured genealogy.
@@
 \begin{corol}
+  \label{thm:obsc_uncond}
   Let $V=(\T,\Z)$ be a given obscured genealogy.
@@
 
-\begin{remark}
-  Note that, since $\func{Y}_t(\Z)$ is finite, it is permissible to choose $q$ and $\pi$ to be uniform.
-\end{remark}
-
 Lemmas in the appendix show how this can be computed via Sequential Monte Carlo.
@@
 
+Relations to filtering of continuous-time Markov chains by \citet{Giesecke2018}.
+
 \begin{defn}
-  Suppose $\Xr_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
+  Let $\Xr_t$ be a continuous-time Markov process with KFE
   \begin{equation}
+    \fontsize{9pt}{11pt}\selectfont
     \label{eq:kfe2}
@@
     -\int{u(t,x)\,\beta(t,x,x')\,\dd{x'}}.
+    \normalfont
   \end{equation}
-  Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)$ are given, real-valued, measurable functions.
-  We say that the equation
+  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ and $\lambda:\Rp\times\Xspace\to\mathbb{R}$ are are given measurable functions.
+  Let $S\subset\Rp$ be countable and locally finite (\ie $S\cap{[0,t]}$ is finite for all $t>0$).
+  Then the system of equations
   \begin{equation}
-    \label{eq:reg-filter-eq}
+    \fontsize{9pt}{11pt}\selectfont
+    \label[pluralequation]{eq:filter-eq-defn}
     \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)
-      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
+      \frac{\partial{w}}{\partial{t}}(t,x)&=
+      \int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
       -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
-      -\lambda(t,x)\,w(t,x).\\
+      -\lambda(t,x)\,w(t,x),
+      \qquad
+      &t\notin{S},\\
+      w(t,x)&=\int{\wt(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
+      &t\in{S},
     \end{aligned}
+    \normalfont
   \end{equation}
-  is the \emph{filter equation} with \emph{driver} $\Xr_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, and \emph{decay} $\lambda$.
+  is called the \emph{filter equation} \emph{generated by} $\beta$, with \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observation times} $S$.
+  The process $\Xr_t$ is said to be the \emph{driver} of the filter equation.
+  The first of \cref{eq:filter-eq-defn} is the \emph{regular part} of the filter equation;
+  the second is known as the \emph{singular part}.
 \end{defn}
@@
 \begin{remark}
-  Trivially, a Kolmogorov forward equation is itself a filter equation with boost $1$ and decay $0$.
+  Trivially, a Kolmogorov forward equation is itself a filter equation with boost $1$, decay $0$, and $S=\emptyset$.
 \end{remark}
 
-\AAK{[We should define the singular filter equations (\ie those with a discrete portion) also.]}
-
 The following results show how filter equations allow one to integrate over random histories.
+First, \cref{lemma:reg-filt} shows how one integrates over the full space of histories using a regular filter equation.
+\Cref{lemma:sing-filt} builds on this when the set of histories is restricted.
 
@@
   Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable.
-  If $\Vr_t$ is an $\Rp$-valued random process satisfying
-  \begin{equation*}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H_t}}{B(e,\Xt_e,\X_e)},
-  \end{equation*}
-  then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued deterministic process on $\Xspace$ with density $w(t,x)$ that satisfies the filter equation
+  Let $\Vr_t$ be an $\Rp$-valued random process satisfying
+  \begin{equation*}
+    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{\mathclap{e\;\in\;\event{\H_t}}}{B(e,\Xt_e,\X_e)}.
+  \end{equation*}
+  Let the family of measures $\lambda_t$ on $\Xspace$ be defined by
+  \begin{equation*}
+    \lambda_t(\mathcal{E})=\Expect{\Vr_t\cdot\Indicator{\Xr_t\in{\mathcal{E}}}},
+  \end{equation*}
+  for measurable $\mathcal{E}$, and let $w(t,x)$ be the density of $\lambda_t$,
+  \ie $\lambda_t(\dd{x})=w(t,x)\,\dd{x}$.
+  In particular, $\Expect{\Vr_t}=\lambda_t(\Xspace)=\int{w(t,x)\,\dd{x}}$.
+  Then $w$ satisfies the initial condition $w(0,x)=p_0(x)$ and the regular filter equation,
   \begin{equation}
@@
   \end{equation}
-  Moreover, $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
 \end{lemma}
 \begin{proof}
-  The expectation can be broken into three terms, according to whether $\H_t$ has zero, one, or more than one event in $\halfclosed{t-\Delta,t}$.
+  Since $\Prob{\Vr_0=1}=1$, $\lambda_0(\mathcal{E})=\Prob{\Xr_0\in\mathcal{E}}$, which implies that $w(0,x)=p_0(x)$.
+  For $t>0$ and $\Delta>0$ sufficiently small, the expectation can be broken into three terms, according to whether $\H_t$ has zero, one, or more than one event in $\halfclosed{t-\Delta,t}$.
   Accordingly, as $\Delta\downarrow{0}$,
@@
   \end{equation*}
-  In the limit, we obtain \cref{eq:reg-filter-eq}.
-  The last statement follows by definition.
+  In the limit, we obtain \cref{eq:reg-filter-eq2}, the regular filtering equation generated by $\alpha$, with boost $B$ and zero decay.
+  \AAK{[This depends on $\alpha$ and $B$ being continuous in their first arguments.]}
 \end{proof}
+
+When events are known to have occurred at particular times, it is of interest to integrate over those histories that include an event at each of these times.
+This leads to singular filter equations, as the next lemma shows.
+Before we state the lemma, some terminology is needed.
+Let $\mathbb{S}$ be the space of increasing, locally finite sequences in $\Rp$, with the topology induced by the Skorokhod metric and Lebesgue measure.
+For $t\in\Rp$ and $s\in\mathbb{S}$, let $s^{}_t\coloneq{s\cap{[0,t]}}$.
+Thus if $s\in\mathbb{S}$ and $s^{}_t=(\hat{s}_1,\dots,\hat{s}_K)$, then the infinitesimal element of Lebesgue measure at $s^{}_t$ is $\dd{s^{}_t}=\prod_{n=1}^{K}\dd{\hat{s}_n}$.
 
@@
   \label{lemma:sing-filt}
-  Suppose $\mathcal{M}\subset{\Rp}$ is countable such that $\mathcal{M}_t\coloneq\mathcal{M}\cap{[0,t]}$ is finite for all $t$.
   Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable and
-  $\Vr_t$ is an $\Rp$-valued process satisfying
-  \begin{equation*}
-    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\Indicator{\event{\H_{t}}\supseteq\mathcal{M}_t}\,\prod_{e\in\event{\H_{t}}}{B(e,\Xt_e,\X_e)}.
-  \end{equation*}
-  Then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued process on $\Rp^{\vert\mathcal{M}_t\vert}\times\Xspace$ with density $w(t,x)$ that satisfies
+  $\Vr_t$ is an $\Rp$-valued random process satisfying
+  \begin{equation*}
+    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{\mathclap{e\;\in\;\event{\H_{t}}}}{B(e,\Xt_e,\X_e)}.
+  \end{equation*}
+  Let $\lambda_t$ be a family measures on $\Xspace\times\mathbb{S}$ defined by
+  \begin{equation*}
+    \lambda_t(\mathcal{E},\mathcal{S})=\Expect{\Vr_t\cdot\Indicator{\Xr_t\in{\mathcal{E}}}\cdot\Indicator{\exists{s\in\mathcal{S}}\:\text{s.t.}\:\event{\Hr_t}\supseteq{s^{}_t}}},
+  \end{equation*}
+  whenever $\mathcal{E}\subseteq\Xspace$ and $\mathcal{S}\subseteq\mathbb{S}$ are measurable.
+  Let $w(t,x,s)$ be the density of this measure, \ie
+  \begin{equation*}
+    \lambda_t(\dd{x}\,\dd{s})=w(t,x,s)\,\dd{x}\,\dd{s^{}_t}.
+  \end{equation*}
+  Then $w$ satisfies
   \begin{equation}
@@
     \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,x)&=
-      \int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
-      &t\notin{\mathcal{M}},\\
-      w(t,x)&=\int{\wt(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
-      &t\in{\mathcal{M}}.
+      \frac{\partial{w}}{\partial{t}}(t,x,s)&=
+      \int{w(t,x',s)\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
+      -\int{w(t,x,s)\,\alpha(t,x,x')\,\dd{x'}},\qquad
+      &t\notin{s},\\
+      w(t,x,s)&=\int{\wt(t,x',s)\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
+      &t\in{s}.
     \end{aligned}
   \end{equation}
-  Moreover, $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
 \end{lemma}
+\begin{proof}
+  The proof proceeds by induction on the cardinality of $s_t$.
+  The base case, for which $s_t=\emptyset$, follows immediately from \cref{lemma:reg-filt}.
+  Assuming that it holds for $|s^{}_t|<K$, one has only to verify the singular part of \cref{eq:two-part-filter}.
+  One can integrate \cref{eq:Hdens} directly to accomplish this.
+\end{proof}
 

### 2024-03-29 ms.Rnw
tweaks; mathsize environment defined

@@
-\documentclass[10pt,reqno,final]{amsart}
+\documentclass[10pt,leqno,final]{amsart}
 
@@
 
-The above may be compactly summarized by stating that if $v(t,x)$ satisfies the Kolmogorov forward equation (KFE),
+The above may be compactly summarized by stating that if $u(t,x)$ satisfies the Kolmogorov forward equation (KFE),
 \begin{equation}
   \label{eq:kfe}
-  \frac{\partial{v}}{\partial{t}}(t,x)=\int\!v(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!v(t,x)\,\alpha(t,x,x')\,\dd{x'},
+  \frac{\partial{u}}{\partial{t}}(t,x)
+  =\int\!{u(t,x')\,\alpha(t,x',x)\,\dd{x'}}
+  -\int\!{u(t,x)\,\alpha(t,x,x')\,\dd{x'}},
 \end{equation}
@@
 Its state vector is $x=(S,I,R)$ and its KFE is
-\begin{equation*}
-  \fontsize{10pt}{11pt}\selectfont
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation*}
     \begin{split}
@@
     \end{split}
-  \normalfont
-\end{equation*}
+  \end{equation*}
+\end{mathsize}
 Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.
@@
 It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
-\begin{equation*}
-  \fontsize{10pt}{11pt}\selectfont
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation*}
     \begin{split}
@@
     \end{split}
-  \normalfont
-\end{equation*}
+  \end{equation*}
+\end{mathsize}
 where $N=S+E+I+R$ is the total population size.
@@
 This model has KFE
-\begin{equation*}
-  \fontsize{10pt}{11pt}\selectfont
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation*}
     \begin{split}
@@
     \end{split}
-  \normalfont
-\end{equation*}
+  \end{equation*}
+\end{mathsize}
 
@@
 The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
-\begin{equation*}
-  \fontsize{10pt}{11pt}\selectfont
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation*}
     \begin{split}
@@
 
-\fontsize{10pt}{11pt}\selectfont
+\begin{mathsize}{10pt}{11pt}
   \begin{equation*}
@@
   \end{equation*}
-\normalfont
+\end{mathsize}
 
@@
 The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
+\begin{mathsize}{9pt}{10pt}
   \begin{equation}
     \label{eq:Hdens}
-  \fontsize{9pt}{11pt}\selectfont
     \pi^{\mathrm{H}}(\dd{\H_t})
@@
     \,\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
-  \normalfont
   \end{equation}
+\end{mathsize}
 where again, by convention, $\Th^{}_0=0$ and $\Th^{}_{K^{}_t+1}=t$.
@@
 
-\begin{corol}
+\begin{thm}
   \label{thm:pruned_uncond}
@@
   \end{equation*}
-\end{corol}
+\end{thm}
 
@@
 
-\begin{corol}
+\begin{thm}
   \label{thm:obsc_uncond}
@@
   \end{equation*}
-\end{corol}
-
-Lemmas in the appendix show how this can be computed via Sequential Monte Carlo.
+\end{thm}
+
+\Cref{lemma:monte-carlo} shows how this can be computed via Sequential Monte Carlo.
 
@@
   Let $\Xr_t$ be a continuous-time Markov process with KFE
+  \begin{mathsize}{9pt}{10pt}
     \begin{equation}
-    \fontsize{9pt}{11pt}\selectfont
       \label{eq:kfe2}
@@
       -\int{u(t,x)\,\beta(t,x,x')\,\dd{x'}}.
-    \normalfont
     \end{equation}
+  \end{mathsize}
   Suppose that $B:\Rp\times\Xspace^2\to\Rp$ and $\lambda:\Rp\times\Xspace\to\mathbb{R}$ are are given measurable functions.
@@
   Then the system of equations
-  \begin{equation}
-    \fontsize{9pt}{11pt}\selectfont
-    \label[pluralequation]{eq:filter-eq-defn}
-    \begin{aligned}
+  \begin{mathsize}{9pt}{11pt}
+    \begin{align}
       \frac{\partial{w}}{\partial{t}}(t,x)&=
@@
       \qquad
-      &t\notin{S},\\
+      &t\notin{S},
+      \label{eq:filter-eq-defn-reg}\\
       w(t,x)&=\int{\wt(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
       &t\in{S},
-    \end{aligned}
+      \label{eq:filter-eq-defn-sing}
+    \end{align}
     \normalfont
-  \end{equation}
+  \end{mathsize}
   is called the \emph{filter equation} \emph{generated by} $\beta$, with \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observation times} $S$.
   The process $\Xr_t$ is said to be the \emph{driver} of the filter equation.
-  The first of \cref{eq:filter-eq-defn} is the \emph{regular part} of the filter equation;
-  the second is known as the \emph{singular part}.
+  \Cref{eq:filter-eq-defn-reg} is the \emph{regular part} of the filter equation;
+  \Cref{eq:filter-eq-defn-sing} is known as the \emph{singular part}.
 \end{defn}
@@
   Then $w$ satisfies
-  \begin{equation}
-    \label{eq:two-part-filter}
-    \begin{aligned}
+  \begin{align}
+    \label{eq:two-part-filter-reg}
     \frac{\partial{w}}{\partial{t}}(t,x,s)&=
@@
     &t\notin{s},\\
+    \label{eq:two-part-filter-sing}
     w(t,x,s)&=\int{\wt(t,x',s)\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
     &t\in{s}.
-    \end{aligned}
-  \end{equation}
+  \end{align}
 \end{lemma}
@@
   The base case, for which $s_t=\emptyset$, follows immediately from \cref{lemma:reg-filt}.
-  Assuming that it holds for $|s^{}_t|<K$, one has only to verify the singular part of \cref{eq:two-part-filter}.
-  One can integrate \cref{eq:Hdens} directly to accomplish this.
+  Assuming that it holds for $|s^{}_t|<K$, one has only to verify \cref{eq:two-part-filter-sing}.
+  This can be accomplished by integrating \cref{eq:Hdens} directly.
 \end{proof}
@@
 \begin{lemma}
-  Showing how \cref{eq:two-part-filter} can be stated as a singular filter equation, \ie with driver $\beta$ and boost $C$, where
+  Showing how \cref{eq:two-part-filter-reg,eq:two-part-filter-sing} can be stated as a singular filter equation, \ie with driver $\beta$ and boost $C$, where
   \begin{equation*}
@@
   \label{lemma:filter}
-  The filter equation \eqref{eq:reg-filter-eq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satisfies the KFE
+  The filter equation \eqref{eq:reg-filter-eq-reg} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satisfies the KFE
   \begin{equation}

### 2024-04-03 ms.Rnw
more work on ms

@@
 We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
-In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\mathbb{Z}^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
+In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\mathbb{Z}^d$, but here we assume only that $\Xspace$ is a complete metric measure space with a countable dense subset.
 The population process is completely specified by its initial-state density, $p_0$, and its transition rates $\alpha$.
@@
 for all $t\in\Rp$ and $x,x'\in\Xspace$.
-\AAK{[We could mention that we use the continuity of $\alpha$ with respect to $t$ in some of the following, but that this condition can be relaxed.]}
+We further assume that $\alpha(t,x,x')$ is \cadlag\ as a function of time for all $x,x'\in{\Xspace}$.
 Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
@@
 
-The above may be compactly summarized by stating that if $u(t,x)$ satisfies the Kolmogorov forward equation (KFE),
+The above may be compactly summarized by stating that if $v(t,x)$ satisfies the Kolmogorov forward equation (KFE),
 \begin{equation}
   \label{eq:kfe}
-  \frac{\partial{u}}{\partial{t}}(t,x)
-  =\int\!{u(t,x')\,\alpha(t,x',x)\,\dd{x'}}
-  -\int\!{u(t,x)\,\alpha(t,x,x')\,\dd{x'}},
+  \frac{\partial{v}}{\partial{t}}(t,x)
+  =\int\!{v(t,x')\,\alpha(t,x',x)\,\dd{x'}}
+  -\int\!{v(t,x)\,\alpha(t,x,x')\,\dd{x'}},
 \end{equation}
@@
 \Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.
+
+\AAK{[
+    We can also let $\alpha$ have some singular components (i.e., $\delta$ functions).
+]}
 
@@
 Its state vector is $x=(S,I,R)$ and its KFE is
-\begin{mathsize}{10pt}{11pt}
-  \begin{equation*}
-    \begin{split}
-      \frac{\partial{v}}{\partial{t}}(t,S,I,R)=
-      &\frac{\beta(t)\,I\,(S+1)}{N}\,v(t,S+1,I-1,R)
-      -\frac{\beta(t)\,I\,S}{N}\,v(t,S,I,R)\\
-      &+\gamma\,(I+1)\,v(t,S,I+1,R-1)
-      -\gamma\,I\,v(t,S,I,R)\\
-      &+\omega\,(R+1)\,v(t,S-1,I,R+1)
-      -\omega\,R\,v(t,S,I,R).
-    \end{split}
-  \end{equation*}
-\end{mathsize}
 Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.
@@
 The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
-It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
-\begin{mathsize}{10pt}{11pt}
-  \begin{equation*}
-    \begin{split}
-      \frac{\partial{v}}{\partial{t}}(t,S,E,I,R)=
-      &\frac{\beta(t)\,I\,(S+1)}{N}\,v(t,S+1,E-1,I,R)
-      -\frac{\beta(t)\,I\,S}{N}\,v(t,S,E,I,R)\\
-      &+\sigma\,(E+1)\,v(t,S,E+1,I-1,R)
-      -\sigma\,E\,v(t,S,E,I,R)\\
-      &+\gamma\,(I+1)\,v(t,S,E,I+1,R-1)
-      -\gamma\,I\,v(t,S,E,I,R)\\
-      &+\omega\,(R+1)\,v(t,S-1,E,I,R+1)
-      -\omega\,R\,v(t,S,E,I,R),
-    \end{split}
-  \end{equation*}
-\end{mathsize}
-where $N=S+E+I+R$ is the total population size.
+It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
 The deme occupancy function in this case is $n(x)=(E,I)$.
@@
 There are four demes ($\Demes=\Set{\lab{E}_1,\lab{E}_2,\lab{I}_1,\lab{I}_2}$) and the occupancy function is $n(x)=(E_1,E_2,I_1,I_2)$.
-This model has KFE
-\begin{mathsize}{10pt}{11pt}
-  \begin{equation*}
-    \begin{split}
-      \frac{\partial{v}}{\partial{t}}=
-      &\frac{\beta_1(t)\,I_1\,(S+1)}{N}\,v(t,S+1,E_1-1,E_2,I_1,I_2,R_1,R_2)
-      -\frac{\beta_1(t)\,I_1\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\frac{\beta_2(t)\,I_2\,(S+1)}{N}\,v(t,S+1,E_1,E_2-1,I_1,I_2,R_1,R_2)
-      -\frac{\beta_2(t)\,I_2\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\sigma_1\,(E_1+1)\,v(t,S,E_1+1,E_2,I_1-1,I_2,R_1,R_2)
-      -\sigma_1\,E_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\sigma_2\,(E_2+1)\,v(t,S,E_1,E_2+1,I_1,I_2-1,R_1,R_2)
-      -\sigma_2\,E_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\gamma_1\,(I_1+1)\,v(t,S,E_1,E_2,I_1+1,I_2,R_1-1,R_2)
-      -\gamma_1\,I_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\gamma_2\,(I_2+1)\,v(t,S,E_1,E_2,I_1,I_2+1,R_1,R_2-1)
-      -\gamma_2\,I_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\omega_1\,(R_1+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1+1,R_2)
-      -\omega_1\,R_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\omega_2\,(R_2+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1,R_2+1)
-      -\omega_2\,R_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-    \end{split}
-  \end{equation*}
-\end{mathsize}
 
@@
 There are three demes ($\Demes=\Set{\lab{E},\lab{I_L},\lab{I_H}}$).
-The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
-\begin{mathsize}{10pt}{11pt}
-  \begin{equation*}
-    \begin{split}
-      \frac{\partial{v}}{\partial{t}}=
-      &\frac{\beta(t)\,(I_L+\theta\,I_H)\,(S+1)}{N}\,v(t,S+1,E-1,I_L,I_H,R)
-      -\frac{\beta(t)\,(I_L+\theta\,I_H)\,S}{N}\,v(t,S,E,I_L,I_H,R)\\
-      &\sigma\,(E+1)\,v(t,S,E+1,I_L-1,I_H,R)
-      -\sigma\,E\,v(t,S,E,I_L,I_H,R)\\
-      &\eps_{LH}\,(I_L+1)\,v(t,S,E,I_L+1,I_H-1,R)
-      -\eps_{LH}\,I_L\,v(t,S,E,I_L,I_H,R)\\
-      &\eps_{HL}\,(I_H+1)\,v(t,S,E,I_L-1,I_H+1,R)
-      -\eps_{HL}\,I_H\,v(t,S,E,I_L,I_H,R)\\
-      &\gamma\,(I_L+1)\,v(t,S,E,I_L+1,I_H,R-1)
-      -\gamma\,I_L\,v(t,S,E,I_L,I_H,R)\\
-      &\gamma\,(I_H+1)\,v(t,S,E,I_L,I_H+1,R-1)
-      -\gamma\,I_H\,v(t,S,E,I_L,I_H,R)\\
-      &\omega\,(R+1)\,v(t,S-1,E,I_L,I_H,R+1)
-      -\omega\,R\,v(t,S,E,I_L,I_H,R).\\
-    \end{split}
-  \end{equation*}
-\end{mathsize}
 
 \subsubsection{Linear birth-death model}
-
-\begin{mathsize}{10pt}{11pt}
-  \begin{equation*}
-    \frac{\partial{v}}{\partial{t}}=
-    \lambda\,(N-1)\,v(t,N-1)
-    -\lambda\,N\,v(t,N)
-    +\mu\,(N+1)\,v(t,N+1)
-    -\mu\,N\,v(t,N)
-  \end{equation*}
-\end{mathsize}
 
@@
 
-\Cref{lemma:monte-carlo} shows how this can be computed via Sequential Monte Carlo.
+\Cref{lemma:filter} shows how this can be computed via Sequential Monte Carlo.
 
@@
 
+\section{Examples}
+
+\subsection{SIRS model}
+
+\citet{King2022} worked out formulas for the exact likelihood of a genealogy induced by an SIRS model.
+The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.
+Its state vector is $x=(S,I,R)$ and its KFE is
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation*}
+    \begin{split}
+      \frac{\partial{v}}{\partial{t}}(t,S,I,R)=
+      &\frac{\beta(t)\,(S+1)\,(I-1)}{N}\,v(t,S+1,I-1,R)
+      -\frac{\beta(t)\,S\,I}{N}\,v(t,S,I,R)\\
+      &+\gamma\,(I+1)\,v(t,S,I+1,R-1)
+      -\gamma\,I\,v(t,S,I,R)\\
+      &+\omega\,(R+1)\,v(t,S-1,I,R+1)
+      -\omega\,R\,v(t,S,I,R).
+    \end{split}
+  \end{equation*}
+\end{mathsize}%
+Here $N=S+I+R$ is the host population size.
+Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.
+
+\subsection{SEIRS model}
+
+A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
+The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
+It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation*}
+    \begin{split}
+      \frac{\partial{v}}{\partial{t}}(t,S,E,I,R)=
+      &\frac{\beta(t)\,(S+1)\,I}{N}\,v(t,S+1,E-1,I,R)
+      -\frac{\beta(t)\,S\,I}{N}\,v(t,S,E,I,R)\\
+      &+\sigma\,(E+1)\,v(t,S,E+1,I-1,R)
+      -\sigma\,E\,v(t,S,E,I,R)\\
+      &+\gamma\,(I+1)\,v(t,S,E,I+1,R-1)
+      -\gamma\,I\,v(t,S,E,I,R)\\
+      &+\omega\,(R+1)\,v(t,S-1,E,I,R+1)
+      -\omega\,R\,v(t,S,E,I,R),
+    \end{split}
+  \end{equation*}
+\end{mathsize}%
+where $N=S+E+I+R$ is the total population size.
+The deme occupancy function in this case is $n(x)=(E,I)$.
+Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.
+
+\subsection{Two-strain competition model}
+
+A simple model for the competition of two strains for susceptible hosts is depicted in \cref{fig:example_models}B.
+In this model, the state vector consists of seven numbers: $x=(S,E_1,E_2,I_1,I_2,R_1,R_2)$.
+There are four demes ($\Demes=\Set{\lab{E}_1,\lab{E}_2,\lab{I}_1,\lab{I}_2}$) and the occupancy function is $n(x)=(E_1,E_2,I_1,I_2)$.
+This model has KFE
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation*}
+    \begin{split}
+      \frac{\partial{v}}{\partial{t}}=
+      &\frac{\beta_1(t)\,I_1\,(S+1)}{N}\,v(t,S+1,E_1-1,E_2,I_1,I_2,R_1,R_2)
+      -\frac{\beta_1(t)\,I_1\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+      &+\frac{\beta_2(t)\,I_2\,(S+1)}{N}\,v(t,S+1,E_1,E_2-1,I_1,I_2,R_1,R_2)
+      -\frac{\beta_2(t)\,I_2\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+      &+\sigma_1\,(E_1+1)\,v(t,S,E_1+1,E_2,I_1-1,I_2,R_1,R_2)
+      -\sigma_1\,E_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+      &+\sigma_2\,(E_2+1)\,v(t,S,E_1,E_2+1,I_1,I_2-1,R_1,R_2)
+      -\sigma_2\,E_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+      &+\gamma_1\,(I_1+1)\,v(t,S,E_1,E_2,I_1+1,I_2,R_1-1,R_2)
+      -\gamma_1\,I_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+      &+\gamma_2\,(I_2+1)\,v(t,S,E_1,E_2,I_1,I_2+1,R_1,R_2-1)
+      -\gamma_2\,I_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+      &+\omega_1\,(R_1+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1+1,R_2)
+      -\omega_1\,R_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+      &+\omega_2\,(R_2+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1,R_2+1)
+      -\omega_2\,R_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
+    \end{split}
+  \end{equation*}
+\end{mathsize}%
+
+\subsection{Superspreading model}
+
+\Cref{fig:example_models}D depicts a model of superspreading.
+There are three demes ($\Demes=\Set{\lab{E},\lab{I_L},\lab{I_H}}$).
+The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation*}
+    \begin{split}
+      \frac{\partial{v}}{\partial{t}}=
+      &\frac{\beta(t)\,(I_L+\theta\,I_H)\,(S+1)}{N}\,v(t,S+1,E-1,I_L,I_H,R)
+      -\frac{\beta(t)\,(I_L+\theta\,I_H)\,S}{N}\,v(t,S,E,I_L,I_H,R)\\
+      &\sigma\,(E+1)\,v(t,S,E+1,I_L-1,I_H,R)
+      -\sigma\,E\,v(t,S,E,I_L,I_H,R)\\
+      &\eps_{LH}\,(I_L+1)\,v(t,S,E,I_L+1,I_H-1,R)
+      -\eps_{LH}\,I_L\,v(t,S,E,I_L,I_H,R)\\
+      &\eps_{HL}\,(I_H+1)\,v(t,S,E,I_L-1,I_H+1,R)
+      -\eps_{HL}\,I_H\,v(t,S,E,I_L,I_H,R)\\
+      &\gamma\,(I_L+1)\,v(t,S,E,I_L+1,I_H,R-1)
+      -\gamma\,I_L\,v(t,S,E,I_L,I_H,R)\\
+      &\gamma\,(I_H+1)\,v(t,S,E,I_L,I_H+1,R-1)
+      -\gamma\,I_H\,v(t,S,E,I_L,I_H,R)\\
+      &\omega\,(R+1)\,v(t,S-1,E,I_L,I_H,R+1)
+      -\omega\,R\,v(t,S,E,I_L,I_H,R).\\
+    \end{split}
+  \end{equation*}
+\end{mathsize}%
+
+\subsection{Linear birth-death model}
+
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation*}
+    \frac{\partial{v}}{\partial{t}}=
+    \lambda\,(N-1)\,v(t,N-1)
+    -\lambda\,N\,v(t,N)
+    +\mu\,(N+1)\,v(t,N+1)
+    -\mu\,N\,v(t,N)
+  \end{equation*}
+\end{mathsize}%
+
+\subsection{Moran model and the Kingman coalescent}
+
+In the Moran model, events occur according to a rate-$\mu$ Poisson process.
+At each event, a compound birth-death jump (cf.~\cref{fig:event_types}F) occurs so that the population size, $n$, remains constant.
+If we let $X_t$ be the number of events that have occurred by time $t$, then $X_t$ is a simple counting process, which we can use to define the state of the population process.
+Its KFE is then
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation*}
+    \begin{gathered}
+      \frac{\partial{v}}{\partial{t}}=
+      \mu\,(x-1)\,v(t,x-1)
+      -\mu\,x\,v(t,x),
+      \qquad
+      v(0,x) =
+      \begin{cases}
+        1, & x=0,\\
+        0, & x>0.
+      \end{cases}
+    \end{gathered}
+  \end{equation*}
+\end{mathsize}%
+
+Since there is only a single deme, and since nothing depends on the state, in writing the corresponding filter equation, we can take $w$ to be independent of both $x$ and $y$.
+If only $m$ samples are taken simultaneously at a single time, $T$, the filter equation thus reads
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation}
+    \label{eq:moran-filter}
+    \begin{gathered}
+      w(0) = 1,
+      \qquad
+      \frac{\partial{w}}{\partial{t}}=
+      \mu\,w(t)\,\left(1-\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\right)
+      -\mu\,w(t),\quad t\notin{S},
+      \qquad
+      w(t) = \mu\,\frac{\wt(t)}{\tbinom{n}{2}},\quad t\in{S}.
+    \end{gathered}
+  \end{equation}
+\end{mathsize}%
+Integrating \cref{eq:moran-filter} and taking logarithms yields
+\begin{equation}
+  \label{eq:kingman}
+  \log{w(T)}=k\,\log{\frac{\mu}{\tbinom{n}{2}}}-\frac{\mu}{\tbinom{n}{2}}\,\sum_{i=m-k}^{m}\!{\tbinom{i}{2}\,s^{}_i},
+\end{equation}
+where $k$ is the number of branch-points and the $s^{}_{i}\coloneq\int{\Indicator{\ell(t)=i}\,\dd{t}}$ are the durations of the \emph{coalescent intervals}, \ie intervals between successive branch-points.
+We recognize \cref{eq:kingman} as the expression for the \citet{Kingman1982a} coalescent \citep[\eg][]{Wakeley2008}.
+
 \end{document}

### 2024-04-04 ms.Rnw
progress

@@
   University of Michigan,
-  Ann~Arbor, MI~48109~USA\\
-  Santa~Fe~Institute,
+  Ann~Arbor, MI~48109~USA
+  \emph{and} Santa~Fe~Institute,
   1399 Hyde~Park~Road,
@@
   E.~L.~Ionides,
-  Department of Statistics
+  Department of Statistics,
   University of Michigan,
@@
 
+In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
+\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
+Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
+
 Relations to lookdown construction of \citet{Etheridge2019}.
@@
 for all $t\in\Rp$ and $x,x'\in\Xspace$.
-We further assume that $\alpha(t,x,x')$ is \cadlag\ as a function of time for all $x,x'\in{\Xspace}$.
 Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
@@
 \end{equation*}
-We will further assume that the number of jumps that occur in a finite time-interval is finite: $\Prob{\Kr_t<\infty}=1$ for all $t$.
-
-\subsubsection{Kolmogorov forward equation}
+We further assume that $\alpha(t,x,x')$ is \cadlag\ as a function of time for all $x,x'\in{\Xspace}$ and that the number of jumps that occur in a finite time-interval is finite: $\Prob{\Kr_t<\infty}=1$ for all $t$.
+
+\subsection{Kolmogorov forward equation}
 
@@
 
-\AAK{[
-    We can also let $\alpha$ have some singular components (i.e., $\delta$ functions).
-]}
-
-Another perspective on the Markov processes is to be had from its Markov state transition diagram (\cref{fig:markov_state}).
+\subsection{Inclusion of jumps at deterministic times}
+
+For modeling purposes, it is sometimes desirable to insist that certain events occur at specific, known times.
+For example, if samples are collected at specific times in such a way that the timing itself conveys no information about the process, one might wish to condition on the sampling time.
+We can expand the class of population models to allow for this as follows.
+Suppose that $S=\Set{s_1,s_2,\dots,}\subset\Zp$ is a sequence of event times.
+If we postulate that, at time $s_i$, an event occurred at which $\Xr_t$ jumped from state $x$ to state $x'$ with probability $\pi(s_i,x,x')$, the KFE for the process becomes
+\begin{align}
+  \label{eq:kfe-reg}
+  \frac{\partial{v}}{\partial{t}}(t,x)
+  &=\int\!{v(t,x')\,\alpha(t,x',x)\,\dd{x'}}
+  -\int\!{v(t,x)\,\alpha(t,x,x')\,\dd{x'}},
+  &t\notin{S},\\
+  \label{eq:kfe-sing}
+  v(t,x)
+  &=\int\!{\leftlim{v}(t,x')\,\pi(t,x',x)\,\dd{x'}},
+  &t\in{S}.
+\end{align}
+Note that the \cref{eq:kfe-reg} is identical to \cref{eq:kfe};
+we call this the \emph{regular part} of the KFE.
+We refer to \cref{eq:kfe-sing} as the \emph{singular part} of the KFE.
+
+Using a Dirac delta notation, one can represent \cref{eq:kfe-reg,eq:kfe-sing} as a single equation in the form of \cref{eq:kfe}.
+In particular, if in \cref{eq:kfe} we make the substitution
+\begin{equation*}
+  \alpha(t,x,x')\mapsto\alpha(t,x,x')+\sum_{s\in{S}}{\delta(t,s)\,\pi(t,x,x')},
+\end{equation*}
+we obtain an equation which we can view as shorthand for \cref{eq:kfe-reg,eq:kfe-sing}.
+%% To be even more parsimonious, we can refer to \cref{eq:kfe-reg,eq:kfe-sing} as the $(\alpha,\pi,S)$ KFE.
+
+\subsection{Jump marks}
+
+%% Another perspective on the Markov processes is to be had from its Markov state transition diagram (\cref{fig:markov_state}).
 
@@
 
-\subsubsection{Structured populations, demes, and deme occupancy}
-\label{sec:demes}
-
-In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
-\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
-Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
-In particular, we suppose that there are a countable set of subpopulations that may differ in their vital rates, but within each of which, individual lineages are statistically identical.
-We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
-We define the \emph{deme occupancy} function $n:\Demes\times\Xspace\to\Zp$ so that
-for $i\in\Demes$, $x\in\Xspace$, $n_i(x)$ is the number of lineages in deme $i$ when the population is in state $x$.
-
-\subsection{Jump marks}
-
-In the following, it will be useful to divide the jumps of the population process $\Xr_t$ into distinct categories.
+It will be useful to divide the jumps of the population process $\Xr_t$ into distinct categories.
 For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
@@
 
+\subsection{Demes and deme occupancy}
+\label{sec:demes}
+
+Our first goal in this paper is to show how a given population process induces a unique stochastic process on the space of genealogies.
+At each time, this genealogy will represent the relationships of shared ancestry among a population of lineages extant at that time.
+To accommodate the structure of the population, this population of lineages will itself be subdivided into discrete categories.
+In particular, we suppose that there are a countable set of subpopulations, within each of which individual lineages are statistically identical.
+We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
+Demes can be used to introduce spatial, behavioral, physiological, and life-history structure into a model.
+The flexibility this allows can also be used, via the linear chain trick, to introduce delays into the dynamics.
+\Cref{fig:example_models} illustrates this concept in the context of several compartmental models.
+
+We define the \emph{deme occupancy} function $n:\Demes\times\Xspace\to\Zp$ so that
+for $i\in\Demes$, $x\in\Xspace$, $n_i(x)$ is the number of lineages in deme $i$ when the population is in state $x$.
+
 \subsection{Examples}
+
+The class of population models to which the theory presented here applies is very broad indeed.
+In particular, it encompasses the entire class of compartmental models with time-dependent flow rates.
+Here, to give a sense of this breadth, we briefly describe a few models of interest.
+\Cref{appendix:examples} works out the theory for each of these examples.
 
@@
 The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.
-Its state vector is $x=(S,I,R)$ and its KFE is
-Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.
 
@@
 A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
-The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
-It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
+The state space is $\Zp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
+It has two demes: $\Demes=\Set{\lab{E},\lab{I}}$.
 The deme occupancy function in this case is $n(x)=(E,I)$.
@@
 
+The linear birth-death process, a mainstay of existing phylodynamic methods, is a special case of the theory presented here.
+For this process, we have $\Xspace=\Zp$ and there is a single deme.
+$\Xr_t$ represents the size of a population and $n(X_t)=X_t$.
+
 \subsubsection{Moran model and the Kingman coalescent}
+
+The Kingman coalescent is another workhorse in existing phylodynamic approaches.
+It is the ancestral process for the Moran model, in which a fixed population of $n$ lineages experiences events at times distributed according to a rate-$\mu$ Poisson process.
+At each such event, an individual lineage selected uniformly at random dies and is replaced by the offspring of a second randomly selected lineage.
 
@@
 $\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
-In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic, as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
+In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic, as are $\Kr_t$, the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
 The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
@@
   \end{equation}
-\end{mathsize}
+\end{mathsize}%
 where again, by convention, $\Th^{}_0=0$ and $\Th^{}_{K^{}_t+1}=t$.
@@
     \end{equation}
-  \end{mathsize}
+  \end{mathsize}%
   Suppose that $B:\Rp\times\Xspace^2\to\Rp$ and $\lambda:\Rp\times\Xspace\to\mathbb{R}$ are are given measurable functions.
@@
     \end{align}
-    \normalfont
-  \end{mathsize}
+  \end{mathsize}%
   is called the \emph{filter equation} \emph{generated by} $\beta$, with \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observation times} $S$.
@@
 Since there is only a single deme, and since nothing depends on the state, in writing the corresponding filter equation, we can take $w$ to be independent of both $x$ and $y$.
-If only $m$ samples are taken simultaneously at a single time, $T$, the filter equation thus reads
+
+In the classical \citet{Kingman1982a} case, $m$ samples are taken simultaneously at a single time, $T$.
+For this case, if $B$ is the set of branch-times and $\ell(t)$ is the number of lineages in the genealogy at time $t$, the filter equation reads
 \begin{mathsize}{10pt}{11pt}
   \begin{equation}
-    \label{eq:moran-filter}
+    \label[pluralequation]{eq:moran-filter}
     \begin{gathered}
@@
       \mu\,w(t)\,\left(1-\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\right)
-      -\mu\,w(t),\quad t\notin{S},
+      -\mu\,w(t),\quad t\notin{B},
       \qquad
-      w(t) = \mu\,\frac{\wt(t)}{\tbinom{n}{2}},\quad t\in{S}.
+      w(t) = \frac{\mu}{\tbinom{n}{2}}\,\wt(t),\quad t\in{B}.
     \end{gathered}
@@
 \end{equation}
-where $k$ is the number of branch-points and the $s^{}_{i}\coloneq\int{\Indicator{\ell(t)=i}\,\dd{t}}$ are the durations of the \emph{coalescent intervals}, \ie intervals between successive branch-points.
+where $k=|B|$ is the number of branch-points in $[0,T]$ and the $s^{}_{i}\coloneq\int{\Indicator{\ell(t)=i}\,\dd{t}}$ are the durations of the \emph{coalescent intervals}, \ie intervals between successive branch-points.
 We recognize \cref{eq:kingman} as the expression for the \citet{Kingman1982a} coalescent \citep[\eg][]{Wakeley2008}.
 
+More generally, if in addition samples are taken through time at a per-capita rate $\psi(t)$ such that the set of sample-times in the genealogy is $S=S_0\cup{S_1}$, where $S_0$, $S_1$ are the sets of times of terminal and inline samples, respectively, then the filter equation reads
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation}
+    \label[pluralequation]{eq:moran-filter2}
+    \begin{gathered}
+      w(0) = 1,
+      \qquad
+      \frac{\partial{w}}{\partial{t}}=
+      -\mu\,\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\,w(t),
+      \quad t\notin{S\cup{B}},
+      \qquad
+      w(t) = \frac{\mu}{\tbinom{n}{2}}\,\wt(t),\quad t\in{B},\\[8pt]
+      w(t) = \psi(t)\,\left(1-\frac{\ell(t)}{n}\right)\,\wt(t),\quad t\in{S_0},
+      \qquad
+      w(t) = \frac{\psi(t)}{n}\,\wt(t),\quad t\in{S_1}.
+    \end{gathered}
+  \end{equation}
+\end{mathsize}%
+Integrating \cref{eq:moran-filter2} yields
+\begin{equation}
+  \label{eq:mgp}
+  \log{w(T)}=
+  |B|\,\log{\frac{\mu}{\tbinom{n}{2}}}
+  +\sum_{t\in{S_0}}\log{\left[\psi(t)\,\left(1-\frac{\ell(t)}{n}\right)\right]}
+  +\sum_{t\in{S_1}}\log{\left[\frac{\psi(t)}{n}\right]}
+  -\frac{\mu}{\tbinom{n}{2}}\,\sum_{i=1}^{\infty}\!{\tbinom{i}{2}\,s^{}_i}.
+\end{equation}
+
 \end{document}

### 2024-04-04 ms.Rnw
progress

@@
 
+In this model, the state variable is the size, $N_t$, of a population at time $t$.
+All individuals face the same per-capita birth and death rates, which are $\lambda$ and $\mu$, respectively.
+The KFE is
 \begin{mathsize}{10pt}{11pt}
@@
     \frac{\partial{v}}{\partial{t}}=
-    \lambda\,(N-1)\,v(t,N-1)
-    -\lambda\,N\,v(t,N)
-    +\mu\,(N+1)\,v(t,N+1)
-    -\mu\,N\,v(t,N)
+    \lambda\,(n-1)\,v(t,n-1)
+    -\lambda\,n\,v(t,n)
+    +\mu\,(n+1)\,v(t,n+1)
+    -\mu\,n\,v(t,n)
   \end{equation*}
 \end{mathsize}%
+\citet{Stadler2010} considered the case where samples are taken through time at a uniform per-capita rate $\psi$.
+In this case, the filter equation becomes
 
@@
 
-In the classical \citet{Kingman1982a} case, $m$ samples are taken simultaneously at a single time, $T$.
-For this case, if $B$ is the set of branch-times and $\ell(t)$ is the number of lineages in the genealogy at time $t$, the filter equation reads
+In the classical case \citep{Kingman1982a}, $m$ samples are taken simultaneously at a single time, $T$.
+Then, if $B$ is the set of branch-times and $\ell(t)$ is the number of lineages in the genealogy at time $t$, the filter equation reads
 \begin{mathsize}{10pt}{11pt}
@@
 
-More generally, if in addition samples are taken through time at a per-capita rate $\psi(t)$ such that the set of sample-times in the genealogy is $S=S_0\cup{S_1}$, where $S_0$, $S_1$ are the sets of times of terminal and inline samples, respectively, then the filter equation reads
+More generally, if in addition samples are taken according to a rate-$\nu$ Poisson process such that the set of sample-times in the genealogy is $S=S_0\cup{S_1}$, where $S_0$, $S_1$ are the sets of times of terminal and inline samples, respectively, then the filter equation reads
 \begin{mathsize}{10pt}{11pt}
@@
       w(t) = \frac{\mu}{\tbinom{n}{2}}\,\wt(t),\quad t\in{B},\\[8pt]
-      w(t) = \psi(t)\,\left(1-\frac{\ell(t)}{n}\right)\,\wt(t),\quad t\in{S_0},
+      w(t) = \nu\,\left(1-\frac{\ell(t)}{n}\right)\,\wt(t),\quad t\in{S_0},
       \qquad
-      w(t) = \frac{\psi(t)}{n}\,\wt(t),\quad t\in{S_1}.
+      w(t) = \frac{\nu}{n}\,\wt(t),\quad t\in{S_1}.
     \end{gathered}
@@
   \label{eq:mgp}
-  \log{w(T)}=
-  |B|\,\log{\frac{\mu}{\tbinom{n}{2}}}
-  +\sum_{t\in{S_0}}\log{\left[\psi(t)\,\left(1-\frac{\ell(t)}{n}\right)\right]}
-  +\sum_{t\in{S_1}}\log{\left[\frac{\psi(t)}{n}\right]}
+  \log{w(T)}-|S|\,\log{\nu}
+  =\sum_{t\in{S_0}}{\log{\left(1-\frac{\ell(t)}{n}\right)}}
+  -|S_1|\,\log{n}
+  +|B|\,\log{\frac{\mu}{\tbinom{n}{2}}}
   -\frac{\mu}{\tbinom{n}{2}}\,\sum_{i=1}^{\infty}\!{\tbinom{i}{2}\,s^{}_i}.

### 2024-05-03 figs/example_models.tex
much more in talk
progress

@@
       \draw [modulate,->] (I2.south west) .. controls ($(I2)+(-1,-1)$) and ($(midSE2)+(0,-1.8)$) .. (midSE2);
-      \node[font=\normalsize] (demes) at ($(E2)!0.5!(I2)+(0,-2)$) {$\Demes=\{\lab{E}_1,\lab{I}_1,\lab{E}_2,\lab{I}_2\}$};
+      \node[font=\normalsize] (demes) at ($(E2)!0.5!(I2)+(0,-2)$) {$\Demes=\{\lab{E}_1,\lab{E}_2,\lab{I}_1,\lab{I}_2\}$};
 
@@
   Demes are shaded.
-  Compartments containing infectious hosts are outlined in blue.
+  Compartments containing infectious hosts are outlined in green.
   Curved green lines connect transmission rates with the compartments whose occupancies control their modulation;
@@
   \textbf{(B)} In this four-deme model, two distinct pathogen strains compete for susceptibles.
-  \textbf{(C)} A three-deme model for SARS-CoV-2 infection.
-  After an incubation period, individuals may develop asymptomatic infection ($\lab{I_A}$).
-  If they do not recover, symptomatically infected individuals ($\lab{I_S}$) can progress to hospitalization ($\lab{H}$) and death ($\lab{D}$).
+  \textbf{(C)} A three-deme model according to which, after an incubation period, hosts may develop asymptomatic infection ($\lab{I_A}$).
+  If they do not recover, symptomatically infected hosts ($\lab{I_S}$) can progress to hospitalization ($\lab{H}$) and death ($\lab{D}$).
   \textbf{(D)} A three-deme model with heterogeneity in transmission behavior.

### 2024-05-03 ms.Rnw
progress

@@
 
-With the advent of inexpensive genome sequencing technologies, routine molecular surveillance has become an important tool in the epidemiologist's kit.
-When an infectious agent evolves quickly enough so that information on its history of transmission accumulates in its genome, yet not so quickly that it is dissipated, it is possible to extract this information to gain insight into the determinants of transmission.
-With sufficiently dense sampling, it can be possible to reconstruct the specific history of who has infected whom, which can aid identification of the correlates of transmission and immunity and the features of infections.
-More typically, one can attempt to learn about these features via the estimation of a model of the transmission process on the basis of relatively sparse samples.
-In particular, one can formalize one or more mathematical models of transmission, estimate their parameters, and comparing their ability to explain data, following a standard statistical paradigm.
-For the purposes of this paper, we refer to the latter program as \emph{phylodynamics}.
-
-Because there is commonly broad uncertainty regarding the structure of the transmission process, for example due to heterogeneities in transmission rates and susceptibilities, complex behavior patterns, etc., methods that have the plug-and-play property are particularly useful.
-Such techniques for inference based on time series are well understood and widely used.
-When the data lie in some Euclidean space, they can commonly be modeled as draws from some error distribution conditional on the latent state of the transmission process.
-However, when the data are genealogies (also called phylogenies) representing the relationships of shared ancestry among genomic samples, the data-space is non-Euclidean and the appropriate models connecting the data to the latent state are non-trivial.
-Here, we derive expressions for the exact likelihood of a genealogy generated through sampling genomes from an arbitrary discretely-structured, Markov latent state process.
-
-Problem of phylodynamics.
-Factorization of problem into two subproblems.
-Problems with this approach \citep{Smith2017}.
-
-Relation to previous work.
-Existing methods \citep{Volz2009a,Stadler2010}.
-Large-population, small sample-size approximations.
-
-For structured but deterministic models, approach of \citet{Volz2012,Rasmussen2014a}, based on approximation as birth-death processes:
-approximation no longer needed.
-
-Modeling of the sampling process.
-
-Extension of previous results \citep{King2022}.
-Broader class of state-spaces.
-Accommodating discrete structure.
-
-In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
-\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
-Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
-
-Relations to lookdown construction of \citet{Etheridge2019}.
-
-Classes of Markov processes.
-Utility and flexibility of Markov assumptions.
-
-Population process induces Markov history and genealogy processes.
-Using these, we derive equations for the likelihood of a genealogy conditional on the history.
-We then integrate out the history to obtain nonlinear filtering equations, the solution of which yields the likelihood.
-These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likelihood.
-We demonstrate with several examples.
-
-In the following, we show a Markov population process of the kind that is a staple in epidemiology induces a Markov process on the space of genealogies.
-We then show how one can compute the likelihood of a given genealogy.
+When an infectious agent mutates quickly enough so that information on the history of its passage through hosts accumulates in its genome, yet not so quickly that this information is effaced, it is possible to extract this information from sampled genomes to gain insight into key determinants of infectious disease transmission.
+With sufficiently dense sampling, it can be possible to reconstruct the detailed history of who infected whom, which can aid identification of the correlates of transmission and immunity and the properties of infections.
+More typically, one can use relatively sparse samples to learn about these features via the estimation of a parsimonious model of the transmission process.
+In particular, one can formalize one or more mathematical models of transmission, estimate their parameters, and compare their ability to explain data, following standard statistical paradigms.
+This is known as \emph{phylodynamic} inference, the term being due to \citet{Grenfell2004}.
+
+The most common approach to phylodynamic inference rests upon a mathematical linkage between the tree-like \emph{genealogy} or \emph{phylogeny} that expresses the relationships of shared ancestry among sampled genomes and a model of the dynamics of the transmission system.
+Various linkages are possible, but because it is maximally efficient (\ie loses the least information), it is desirable to be able to compute the likelihood function for models of interest.
+This is simply the probability density of a given genealogy conditional on a given model, viewed as a function of the parameters of that model.
+To date, however, computation of this likelihood function has been out of reach, except in several special cases.
+
+%% Phylodynamic inference rests upon mathematical linkage between mathematical models of disease transmission at the population level and genome-sequence data collected from individual infected hosts.
+%% The tightest possible such linkage is effected by a computable likelihood function.
+%% The tree-like \emph{genealogy} or \emph{phylogeny} that expresses the relationships of shared ancestry among the sequences.
+%% In particular, if one can determine the likelihood of any given genealogy given a model and also the likelihood of the data given a genealogy, then the likelihood of the data given the model is obtained by summing over the space of genealogies.
+
+In particular, existing approaches to likelihood-based phylodynamic inference have been based on one of two mathematical idealizations.
+The first is the \citet{Kingman1982a} coalescent, by which likelihood of a given genealogy is computed via a reverse-time argument.
+This computation is exact when the population size is constant.
+Extensions of this approach develop approximate likelihoods for the case when the population size varies slowly \citep{Drummond2005} or when the sample fraction is negligible \citep{Volz2009a}.
+The second idealization is the linear birth-death process, for which exact expressions for the likelihood are available \citep{Stadler2010}.
+Linearity in this context amounts to the assumption that distinct lineages do not interact:
+it is the resulting self-similarity of genealogies that renders the likelihood analytically tractable.
+Extensions of this approach develop approximations via linearization of nonlinear processes or restriction to scenarios in which population growth is nearly linear \citep{MacPherson2021}.
+Although the tractability of these approaches are attractive, concern naturally arises as to validity of the approximations, the biases introduced by them, and the amount of information in data left unutilized by these approximate methods.
+
+%% Factorization of problem into two subproblems.
+%% Problems with this approach \citep{Smith2017}.
+
+What would an improved phylodynamic inference approach look like?
+First, it would afford exact computation of the phylodynamic likelihood, so that comparisons among parameterizations and models could be made on a sound basis.
+Second, because nonlinearity, noise, and measurement error are so prominent and ubiquitous in epidemiology, it would accommodate nonlinear, stochastic transmission models.
+Third, because many of the most important uncertainties concern heterogeneities in transmission rates and the susceptibility, behavior, age, and location of hosts, it would accommodate host populations structured by these factors.
+While some structuring factors (\eg age, spatial location) are most naturally expressed in terms of continuous variables, discretely structured models have repeatedly proved their value in epidemiology.
+In particular, compartmental models are extremely flexible and have often been used as approximations when continuous structure leads to uncomfortably high model dimension.
+Finally, because there is typically uncertainty not only in the parameters, but also in the structure, of a host-pathogen system, an improved phylodynamic inference methodology would place minimal restrictions on the form of the models that it can accommodate.
+This paper demonstrates how these desiderata can be achieved---at least for models with discrete structure---including arbitrary nonlinear compartmental models with a countable number of compartments.
+
+To connect a model at the level of a population with genealogies based on samples taken from individual hosts, it is necessary to make assumptions about the individuals in the population.
+The simplest such assumption is that the individuals that are identical with respect to the population dynamics are indeed statistically identical.
+That is, that they are \emph{exchangeable}.
+For example, in a compartmental model, one can assume that individual hosts within a compartment are exchangeable.
+This implies that their residence times within a compartment are identically distributed, though not independent.
+Although this is indeed an additional assumption, it is so natural that it is frequently unrecognized as such, and one often reads statements to the effect that exchangeability of individuals is a consequence of the Markovian assumption about the discretely structured population.
+Nonetheless, since it adds minimum additional structure, it is the natural assumption, and the one we will make in this paper.
+
+In the following, we take as our starting point a transmission model in the form of a discretely structured, Markov process.
+We show how such a process uniquely induces each of several stochastic processes in the space of genealogies.
+We go on to derive expressions for the exact likelihoods of these genealogies.
+
+%% In particular, while the constructions of \citet{Etheridge2019} can in principle accommodate.
+%% For structured but deterministic models, approach of \citet{Volz2012,Rasmussen2014a}, based on approximation as birth-death processes:
+%% approximation no longer needed.
+
+%% Because there is commonly broad uncertainty regarding the structure of the transmission process, for example due to heterogeneities in transmission rates and susceptibilities, complex behavior patterns, etc., methods that have the plug-and-play property are particularly useful.
+%% Such techniques for inference based on time series are well understood and widely used.
+%% When the data lie in some Euclidean space, they can commonly be modeled as draws from some error distribution conditional on the latent state of the transmission process.
+%% However, when the data are genealogies (also called phylogenies) representing the relationships of shared ancestry among genomic samples, the data-space is non-Euclidean and the appropriate models connecting the data to the latent state are non-trivial.
+
+%% Modeling of the sampling process.
+
+%% Extension of previous results \citep{King2022}.
+%% Broader class of state-spaces.
+%% Accommodating discrete structure.
+
+%% In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
+%% \citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
+%% Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
+
+%% Classes of Markov processes.
+%% Utility and flexibility of Markov assumptions.
+
+%% Population process induces Markov history and genealogy processes.
+%% Using these, we derive equations for the likelihood of a genealogy conditional on the history.
+%% We then integrate out the history to obtain nonlinear filtering equations, the solution of which yields the likelihood.
+%% These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likelihood.
+%% We demonstrate with several examples.
 
@@
 In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $\Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
-In the case of continuous-time processes, we will assume that sample paths are \cadlag\ \ie, right-continuous with left limits.
+In the case of continuous-time processes, we will assume that sample paths are \cadlag\ \ie right-continuous with left limits.
 We will frequently need to refer to the left-limit of such a process.
@@
 
-Motivating examples:
-compartmental models in ID epidemiology,
-phylogenetics and systematics.
-Wide variety of models are of interest (\cref{fig:example_models}).
-Linear chain trick.
-Migration, superspreading, competition between strains.
+We are motivated by the desire for exact phylodynamic inference methods for as wide a class of epidemiological models as possible.
+In particular, we would like to be able to formulate and parameterize an arbitrary compartmental model and to quantify its ability to explain data using likelihood.
+\Cref{fig:example_models} depicts a few such models in order to give a sense of the kinds of complexities that can arise.
+Of course, with the ability to entertain models with countably many compartments, much greater complexity is possible.
+In particular, one can model not only complex infection progression, but also strain structure, behavioral structure (\eg superspreading), age structure, and spatial structure using compartmental models.
+As is well known, one can discretize continuous structure-variables and employ the linear chain trick to accommodate non-exponential residence times.
+While the utility of these approximations will vary, a very wide range of model assumptions lie within the scope of the theory presented here.
 
@@
 Suppose that $S=\Set{s_1,s_2,\dots,}\subset\Zp$ is a sequence of event times.
-If we postulate that, at time $s_i$, an event occurred at which $\Xr_t$ jumped from state $x$ to state $x'$ with probability $\pi(s_i,x,x')$, the KFE for the process becomes
+Let us postulate that, at each of these times, an event occurs at which $\Xr_t$ jumps according to a given probability kernel $\pi$.
+In particular, for any state $x\in\Xspace$ and measurable $\mathcal{E}\subset\Xspace$,
+$\pi(s_i,x,\mathcal{E})$ is the probability that the jump at time $s_i$ is to $\mathcal{E}$, conditional on the state just before the jump being $x$.
+With this notation, the KFE for the process becomes
 \begin{align}
@@
   v(t,x)
-  &=\int\!{\leftlim{v}(t,x')\,\pi(t,x',x)\,\dd{x'}},
+  &=\int\!{\leftlim{v}(t,x')\,\pi(t,x',\dd{x'})},
   &t\in{S}.
@@
 \begin{equation*}
-  \alpha(t,x,x')\mapsto\alpha(t,x,x')+\sum_{s\in{S}}{\delta(t,s)\,\pi(t,x,x')},
+  \alpha(t,x,x')\mapsto\alpha(t,x,x')+\sum_{s\in{S}}{\delta(t,s)\,\frac{\dd{\pi}}{\dd{x}}(t,x,x')},
 \end{equation*}
 we obtain an equation which we can view as shorthand for \cref{eq:kfe-reg,eq:kfe-sing}.
+Here, $\dd{\pi}/\dd{x}$ denotes the density (\ie Radon-Nikodym derivative) of $\pi$ with respect to the measure on $\Xspace$.
 %% To be even more parsimonious, we can refer to \cref{eq:kfe-reg,eq:kfe-sing} as the $(\alpha,\pi,S)$ KFE.
@@
 \begin{equation*}
-  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
+  \alpha(t,x,x')=\sum_{u\in\Jumps}{\alpha_u(t,x,x')}.
 \end{equation*}
 \Cref{fig:markov_state} shows an example with five distinct marks.
-Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.
+In the following, sums over $u$ are to be taken over the whole of $\Jumps$ unless otherwise indicated.
 
@@
 To accommodate the structure of the population, this population of lineages will itself be subdivided into discrete categories.
-In particular, we suppose that there are a countable set of subpopulations, within each of which individual lineages are statistically identical.
+In particular, we suppose that there are a countable set of subpopulations, within each of which individual lineages are exchangeable.
 We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
@@
 Here, to give a sense of this breadth, we briefly describe a few models of interest.
-\Cref{appendix:examples} works out the theory for each of these examples.
-
-\subsubsection{SIRS model}
+\Cref{sec:app-examples} works out the theory for each of these examples.
+
+\paragraph{SIRS model}
 
@@
 
-\subsubsection{SEIRS model}
+\paragraph{SEIRS model}
 
@@
 
-\subsubsection{Two-strain competition model}
+\paragraph{Two-strain competition model}
 
@@
 
-\subsubsection{Superspreading model}
+\paragraph{Superspreading model}
 
@@
 
-\subsubsection{Linear birth-death model}
+\paragraph{Linear birth-death model}
 
@@
 
-\subsubsection{Moran model and the Kingman coalescent}
-
-The Kingman coalescent is another workhorse in existing phylodynamic approaches.
+\paragraph{Moran model and the Kingman coalescent}
+
+The \citep{Kingman1982a} coalescent is another workhorse in existing phylodynamic approaches.
 It is the ancestral process for the Moran model, in which a fixed population of $n$ lineages experiences events at times distributed according to a rate-$\mu$ Poisson process.
@@
     internal nodes are indicated in green.
-    Note that internal nodes occur both at branch-points and inline (\ie along branches).
+    Note that internal nodes occur not only at branch-points, but also inline (\ie along branches).
     Wherever a lineage moves from one deme (color) to another, an internal node occurs;
@@
 
-A \emph{genealogy}, $G$, encapsulates the relationships of shared ancestry among a set of lineages that are extant at some time $\time(G)\in\Rp$, and perhaps a set of samples collected at earlier times (\cref{fig:geneal}).
+A \emph{genealogy}, $G$, encapsulates the relationships of shared ancestry among a set of lineages that are extant at some time $\time(G)\in\Rp$ and perhaps a set of samples collected at earlier times (\cref{fig:geneal}).
 A genealogy has a tree- or forest-like structure, with four distinct kinds of nodes:
@@
 In particular, if $a$ is a tip node in $G$, then $\time(a)=\time(G)$;
-if $a$ is a sample node, then $\time(a)$ is the time at which the sample was taken.
+if $a$ is a sample node, then $\time(a)\le{\time(G)}$ is the time at which the sample was taken.
 Moreover, if node $a$ is ancestral to node $a'$, then $\time(a)\le{\time(a')}$ and $\time(a')-\time(a)$ is the distance between $a$ and $a'$ along the genealogy.
@@
 Given a tree $Z$, we let $\func{Y}(Z)$ denote the set of colorings $Y$ that are compatible with $Z$ and $\func{Y}_t(Z)\coloneq\CondSet{Y_t}{Y\in{\func{Y}(Z)}}$.
-
-\AAK{[Formally speaking, the set of possible colorings is a section of a fiber bundle over $Z$, where each fiber is the coloring at a given time.]}
+Formally speaking, $\func{Y}(Z)$ is a fiber bundle over $Z$, each $\func{Y}_t(Z)$ being a fiber.
 
@@
          {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
-         & \text{if}\ \forall i\ n_i\ge{\Set{\ell_i,r_i}}\ge{s_i}\ge{0},\\
+         & \text{if}\ \forall i\ n_i\ge{\Set{\ell_i,r_i}}\ge{s_i}\ge{0},\\[7ex]
          0, & \text{otherwise}.
@@
 
-We will show how a given population process naturally induces a process in the space of genealogies.
-Specifically, we will describe how, at each jump in the population process, a corresponding change occurs in the genealogy, according to whether lineages branch, die, move between demes, or are sampled.
+We now show how a given population process naturally induces a process in the space of genealogies.
+Specifically, at each jump in the population process, a corresponding change occurs in the genealogy, according to whether lineages branch, die, move between demes, or are sampled.
 For this purpose, there are five distinct \emph{pure types} of events:
 \begin{compactenum}[(a)]
-\item \emph{Birth-type events} result in the branching of one or more new lineages, each from some existing lineages.
+\item \emph{Birth-type events} result in the branching of one or more new lineages, each from some existing lineage.
   Examples of birth-type events include transmission events, speciations, and actual births.
@@
 \emph{compound events} partake of more than one type.
-For example, a sample/death-type event, in which a lineage is simultaneously sampled and removed, has been proposed \citep{Leventhal2014}, as have birth/death events in which one lineage reproduces at the same moment that another dies (\eg the \citet{Moran1958} process).
+For example, a sample/death-type event, in which a lineage is simultaneously sampled and removed, has been employed \citep{Leventhal2014}, as have birth/death events in which one lineage reproduces at the same moment that another dies (\eg the \citet{Moran1958} process).
 The theory presented here places few restrictions on the complexity of the events that can occur by combining events of the various pure types.
@@
 Thus there is a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
-We say $r^u=(r^u_i)_{i\in\Demes}$ is the \emph{production} of an event of mark $u$.
+We say $r^u\coloneq{(r^u_i)_{i\in\Demes}}$ is the \emph{production} of an event of mark $u$.
 Note that the lineages that die as a result of an event do not count in the production but that a parent lineage that survives the event does count.
@@
 It is very important to note that, conditional on $(\Xrh_k,\Urh_k)$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of these choices at any other times and independent of $(\Xrh_j,\Urh_j)$ for all $j\ne{k}$.
-Moreover, by construction, any lineage within a deme is as likely as any other lineage in that deme to be selected as a parent or for death, sampling, or migration.
-We refer to this property as the \emph{exchangeability} of lineages within a deme.
+Moreover, by construction, the lineages within each deme are exchangeable:
+any lineage within a deme is as likely as any other lineage in that deme to be selected as a parent or for death, sampling, or migration.
 Finally, note that $\Gr_t$ does not have the Markov property, though $(\Xr_t,\Ur_t,\Gr_t)$ and $(\Xr_t,\Gr_t)$ do.
@@
 Accordingly, suppose $P=(T,Z,Y)$ is a pruned genealogy and suppose $t\in[0,T]$.
-Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell=(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
+Let $\ell_i$ denote the number of lineages in deme $i$ at time $t$ and $\ell\coloneq(\ell_i)_{i\in\Demes}\in\Zp^{\Demes}$.
 Clearly, $\ell$ depends only $Y_t$.
@@
 We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
-In particular, given a node time $t$ in a pruned genealogy $P=(T,Y,Z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s=\left(s_i\right)_{i\in\Demes}$.
+In particular, given a node time $t$ in a pruned genealogy $P=(T,Y,Z)$, the number $s_i$ of lineages of deme $i$ emerging from all nodes with time $t$ is well defined and we can write $s\coloneq\left(s_i\right)_{i\in\Demes}$.
 Like the lineage-count function, $s$ depends only on the local structure of $\P$.
@@
   \end{equation*}
-  Returning to the possibility that $\H$ is incompatible with $\P$, since $\Prob{\Pr_\T=\P}=0$ if either any $Q_u=0$ or $\event{\P}\nsubseteq\event{\H}$, we obtain the result.
+  Returning to the possibility that $\H$ is incompatible with $\P$, since $\Prob{\Pr_\T=\P}=0$ if either any $Q^{}_{\U_t}=0$ or $\event{\P}\nsubseteq\event{\H}$, we obtain the result.
 \end{proof}
@@
 
-\section{Examples}
-
-\subsection{SIRS}
-
-Was shown in an earlier paper.
-[Display filter equation and plot a curve.]
-
-\subsection{SEIRS}
-
-Display filter equation.
-Discuss the choice of $\pi$ and $q$.
-Present a likelihood curve.
-
-\subsection{Two-strain competition}
-
-Display filter equation.
-Discuss the choice of $\pi$ and $q$.
-Present a likelihood curve.
-
-\subsection{Superspreading model}
-
-Display filter equation.
-Discuss the choice of $\pi$ and $q$.
-Present a likelihood curve.
-
-\subsection{Linear birth-death and Moran models}
-
-Exact solution of the filtering equations is possible in the unstructured case, yielding precisely the likelihoods for these two models.
-This establishes that the present theory is a strict generalization of these cases.
-It does away with the need for large sample-size and small sample-fraction approximations, the necessity of assuming slow changes in effective population size, and the need for linearization.
+%% \section{Examples}
+
+%% \subsection{SIRS}
+
+%% Was shown in an earlier paper.
+%% [Display filter equation and plot a curve.]
+
+%% \subsection{SEIRS}
+
+%% Display filter equation.
+%% Discuss the choice of $\pi$ and $q$.
+%% Present a likelihood curve.
+
+%% \subsection{Two-strain competition}
+
+%% Display filter equation.
+%% Discuss the choice of $\pi$ and $q$.
+%% Present a likelihood curve.
+
+%% \subsection{Superspreading model}
+
+%% Display filter equation.
+%% Discuss the choice of $\pi$ and $q$.
+%% Present a likelihood curve.
+
+%% \subsection{Linear birth-death and Moran models}
+
+%% Exact solution of the filtering equations is possible in the unstructured case, yielding precisely the likelihoods for these two models.
+%% This establishes that the present theory is a strict generalization of these cases.
+%% It does away with the need for large sample-size and small sample-fraction approximations, the necessity of assuming slow changes in effective population size, and the need for linearization.
 
 \section{Discussion}
+
+This approach gives us the freedom to choose models with many demes.
+Moreover, there is freedom to choose the sampling model.
 
@@
 
-Freedom to choose models with many demes.
-Freedom to choose model of sampling.
-
-Price of flexibility is variability in the Monte Carlo estimation.
-Freedom to choose importance sampling distribution.
-It is permissible to borrow information from the future.
+The price for this flexibility is variability in the Monte Carlo estimation.
+However, there is substantial freedom in the choice of the importance sampling distribution $\pi$.
+Moreover, since it is permissible to borrow information from the future, one can envision importance sampling distributions tailored to any given genealogy.
 
@@
 \section{Examples}
+\label{sec:app-examples}
 
@@
 \citet{Stadler2010} considered the case where samples are taken through time at a uniform per-capita rate $\psi$.
-In this case, the filter equation becomes
+In this case, since there is only one deme, in the filter equation, $w$ can be taken to be independent of $y$.
+If $B$ is the set of branch-times and $S_0$, $S_1$ are the sets of terminal and inline samples, respectively, then the regular part of the filter equation is
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation}
+    \label{eq:filter-lbdp-reg}
+    \begin{aligned}
+      \frac{\partial{w}}{\partial{t}}(t,n)=
+      &\lambda\,(n-1)\,\left(1-\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\right)\,w(t,n-1)
+      -\lambda\,n\,w(t,n)
+      \\
+      &\qquad+\mu\,(n+1)\,w(t,n+1)-\mu\,n\,w(t,n)-\psi\,n\,w(t,n),
+      &&n\ge\ell(t),\quad t\notin{B\cup{S_0}\cup{S_1}}
+    \end{aligned}
+  \end{equation}
+\end{mathsize}%
+and the singular part is
+\begin{mathsize}{10pt}{11pt}
+  \begin{equation}
+    \label{eq:filter-lbdp-sing}
+    \begin{gathered}
+      w(t,n)=
+      \frac{\lambda\,(n-1)}{\tbinom{n}{2}}\,\wt(t,n-1),
+      \quad t\in{B},
+      \\
+      w(t,n)=
+      \psi\,n\,\left(1-\frac{\ell(t)}{n}\right)\,\wt(t,n),
+      \quad t\in{S_0},
+      \qquad
+      w(t,n)=
+      \psi\,\wt(t,n),
+      \quad t\in{S_1}.
+    \end{gathered}
+  \end{equation}
+\end{mathsize}
+\Cref{eq:filter-lbdp-reg,eq:filter-lbdp-sing} are supplemented by the ancillary condition $w(t,n)=0$ for $n<\ell(t)$.
 

### 2024-05-07 ms.Rnw
more work

@@
 it is the resulting self-similarity of genealogies that renders the likelihood analytically tractable.
-Extensions of this approach develop approximations via linearization of nonlinear processes or restriction to scenarios in which population growth is nearly linear \citep{MacPherson2021}.
+Extensions of this approach develop approximations via linearization of nonlinear processes or restriction to scenarios in which population growth is nearly linear \citep[\eg]{MacPherson2021}.
 Although the tractability of these approaches are attractive, concern naturally arises as to validity of the approximations, the biases introduced by them, and the amount of information in data left unutilized by these approximate methods.
@@
 
+\citep{Alizon2024}
+
 This approach gives us the freedom to choose models with many demes.
@@
 \bibliographystyle{preprint}
-\bibliography{phylopomp}
+\bibliography{ms}
 

### 2024-05-07 ms.Rnw
more work

@@
 This is known as \emph{phylodynamic} inference, the term being due to \citet{Grenfell2004}.
+\citet{Alizon2024} gives a good review of the history of the subject.
 
@@
 This is simply the probability density of a given genealogy conditional on a given model, viewed as a function of the parameters of that model.
-To date, however, computation of this likelihood function has been out of reach, except in several special cases.
+In particular, if $M$ represents a dynamic transmission model, $S$ a set of genome sequences, $\Phi$ a genealogical tree relating these sequences, and $E$ a model of sequence evolution along a tree,
+then the likelihood of $S$ is a function
+\begin{equation*}
+  f(S|M,E) = \int{f(S|\Phi,E)\,f(\Phi|M)\,\dd{\Phi}},
+\end{equation*}
+where we somewhat loosely use the functions $f$ to refer to distinct probability densities, the nature of each of which is clear from its arguments and the integral is taken over all possible genealogies relating the sequences $S$.
+In this expression, $f(S|\Phi,E)$ is the \citet{Felsenstein2004} likelihood.
+The function $f(\Phi|M)$, which links the phylogeny $\Phi$ to the dynamic model $M$, may be termed the \emph{phylodynamic likelihood}.
+In the Bayesian context, this same function is sometimes referred to as a \emph{tree prior} \citep{Moeller2018,Volz2018}.
+The computation of this function has remained out of reach, except in several special cases.
 
@@
 
-In particular, existing approaches to likelihood-based phylodynamic inference have been based on one of two mathematical idealizations.
+In particular, existing approaches to the phylodynamic likelihood have been based on one of two mathematical idealizations.
 The first is the \citet{Kingman1982a} coalescent, by which likelihood of a given genealogy is computed via a reverse-time argument.
@@
 it is the resulting self-similarity of genealogies that renders the likelihood analytically tractable.
-Extensions of this approach develop approximations via linearization of nonlinear processes or restriction to scenarios in which population growth is nearly linear \citep[\eg]{MacPherson2021}.
+Extensions of this approach develop approximations via linearization of nonlinear processes or restriction to scenarios in which population growth is nearly linear \citep[\eg][]{MacPherson2021}.
 Although the tractability of these approaches are attractive, concern naturally arises as to validity of the approximations, the biases introduced by them, and the amount of information in data left unutilized by these approximate methods.
@@
 
-\citep{Alizon2024}
-
 This approach gives us the freedom to choose models with many demes.

### 2024-05-10 figs/event_types.tex
more

--- 2024-01-21 figs/event_types.tex
@@
 \end{center}
-\caption{
-  Jump types differ by their effects on the genealogy.
-  This can be seen by examining the local structure of the genealogy in a neighborhood of a jump event.
-  \textbf{(A)} A birth-type jump results in the branching of one or more child lineages from the parent.
-  There can be only one parent, though the child lineages may be in demes different from those of their parent.
-  Here, a parent of the blue deme sires one child lineage in each of the blue and yellow demes.
-  The production of an event is an integer vector, with one entry for each deme.
-  The production of this event is therefore $r=(r_{\lab{blue}},r_{\lab{yellow}})=(2,1)$.
-  \textbf{(B)} A death-type event causes the extinction of a lineage.
-  Since internal node without children are recursively removed, the affected branch disappears.
-  The production of this event is $r=(0,0)$.
-  \textbf{(C)} A migration-type event results in the movement of one or more lineages from one deme to another.
-  Here, one lineage moves from the yellow to the blue deme.
-  The production of this event is $r=(1,0)$, \ie the production is 1 for the blue deme and 0 for the yellow.
-  \textbf{(D)} In a sample-type event, one or more sample nodes (blue circles) are inserted.
-  Here, there are two samples, one in each of the blue and yellow demes.
-  Accordingly, its production is $r=(1,1)$.
-  \textbf{(E)} A neutral-type event has no effect on the genealogy and zero production in all demes: $r=(0,0)$.
-  \textbf{(F)} The theory presented here allows for compound events.
-  As an example, here a birth/death-type event occurs, wherein one yellow lineage is extinguished and a blue lineage simultaneously sires a blue child.
-  The production of this event is $r=(2,0)$.
-  \textbf{(G)} Here, a compound sample/death-type event with $r=(0,0)$ occurs.
-  A blue lineage is sampled and simultaneously extinguished.
-  Note that recursive removal does not occur, since sample nodes are never removed.
-  \textbf{(H)} A compound birth/migration-type event with $r=(4,0)$.
-  \label{fig:event_types}
-}

### 2024-05-10 figs/example_models.tex
more

@@
 \end{center}
-\caption{
-  Examples of discretely-structured population models.
-  Demes are shaded.
-  Compartments containing infectious hosts are outlined in green.
-  Curved green lines connect transmission rates with the compartments whose occupancies control their modulation;
-  each such connection gives rise to a nonlinearity in the model.
-  \textbf{(A)} An SEIRS model.
-  Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
-  Upon recovery ($\lab{R}$), individuals experience immunity from reinfection.
-  If this immunity wanes, they re-enter the susceptible compartment.
-  Pathogen lineages are to be found in hosts within the $\lab{E}$ and $\lab{I}$ compartments only.
-  Accordingly, there are two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
-  If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
-  \textbf{(B)} In this four-deme model, two distinct pathogen strains compete for susceptibles.
-  \textbf{(C)} A three-deme model according to which, after an incubation period, hosts may develop asymptomatic infection ($\lab{I_A}$).
-  If they do not recover, symptomatically infected hosts ($\lab{I_S}$) can progress to hospitalization ($\lab{H}$) and death ($\lab{D}$).
-  \textbf{(D)} A three-deme model with heterogeneity in transmission behavior.
-  Contagious individuals move randomly between low-transmission ($\lab{I_L}$) and high-transmission ($\lab{I_H}$) behaviors.
-  \label{fig:example_models}
-}

### 2024-05-10 ms.Rnw
more

@@
 \begin{abstract}
-  We derive an algorithm for the exact computation of the likelihood of an observed transmission tree.
+  We show how each member of a broad class of Markov population models uniquely induces a stochastic process on the space of genealogies.
+  We then derive exact expressions for the likelihood of an observed genealogy in terms of filtering equations, which can be integrated using standard Monte Carlo methods.
+  The class of population models that can be accommodated includes essentially all compartmental models.
 \end{abstract}
@@
 
-When an infectious agent mutates quickly enough so that information on the history of its passage through hosts accumulates in its genome, yet not so quickly that this information is effaced, it is possible to extract this information from sampled genomes to gain insight into key determinants of infectious disease transmission.
+When an infectious agent mutates quickly enough so that information on the history of its passage through hosts accumulates in its genome, yet not so quickly that this information is dissipated, it is possible to extract this information from sampled genomes to gain insight into key determinants of infectious disease transmission.
 With sufficiently dense sampling, it can be possible to reconstruct the detailed history of who infected whom, which can aid identification of the correlates of transmission and immunity and the properties of infections.
@@
 This is simply the probability density of a given genealogy conditional on a given model, viewed as a function of the parameters of that model.
-In particular, if $M$ represents a dynamic transmission model, $S$ a set of genome sequences, $\Phi$ a genealogical tree relating these sequences, and $E$ a model of sequence evolution along a tree,
-then the likelihood of $S$ is a function
-\begin{equation*}
-  f(S|M,E) = \int{f(S|\Phi,E)\,f(\Phi|M)\,\dd{\Phi}},
-\end{equation*}
-where we somewhat loosely use the functions $f$ to refer to distinct probability densities, the nature of each of which is clear from its arguments and the integral is taken over all possible genealogies relating the sequences $S$.
-In this expression, $f(S|\Phi,E)$ is the \citet{Felsenstein2004} likelihood.
-The function $f(\Phi|M)$, which links the phylogeny $\Phi$ to the dynamic model $M$, may be termed the \emph{phylodynamic likelihood}.
+In particular, if $S$ a set of genome sequences, $\Phi$ a genealogical tree relating these sequences, $E$ a model of sequence evolution, and $D$ a dynamic transmission model, then the likelihood of $S$ is
+\begin{equation*}
+  f(S|D,E) = \int{f(S|\Phi,E)\,f(\Phi|D)\,\dd{\Phi}},
+\end{equation*}
+where the integral is taken over all possible genealogies and we somewhat loosely use the functions $f$ to refer to distinct probability densities, the nature of each of which is clear from its arguments.
+In this expression, $f(S|\Phi,E)$ is the \citet{Felsenstein2004} phylodynamic likelihood.
+The function $f(\Phi|D)$, which links the phylogeny to the dynamic model, may be termed the \emph{phylodynamic likelihood}.
 In the Bayesian context, this same function is sometimes referred to as a \emph{tree prior} \citep{Moeller2018,Volz2018}.
 The computation of this function has remained out of reach, except in several special cases.
+This paper presents theory that enables its computation for a very broad range of dynamic models.
 
@@
 This computation is exact when the population size is constant.
-Extensions of this approach develop approximate likelihoods for the case when the population size varies slowly \citep{Drummond2005} or when the sample fraction is negligible \citep{Volz2009a}.
+Extensions of this approach develop approximate likelihoods for the case when the population size varies as a function of time \citep{Griffiths1994,Drummond2005} or according to an SIR process \citep{Volz2009a,Rasmussen2011}, as long as the population size is large and the sample-fraction remains negligible.
 The second idealization is the linear birth-death process, for which exact expressions for the likelihood are available \citep{Stadler2010}.
@@
 Extensions of this approach develop approximations via linearization of nonlinear processes or restriction to scenarios in which population growth is nearly linear \citep[\eg][]{MacPherson2021}.
-Although the tractability of these approaches are attractive, concern naturally arises as to validity of the approximations, the biases introduced by them, and the amount of information in data left unutilized by these approximate methods.
+Although the tractability of these approaches makes them attractive, concern naturally arises as to validity of the approximations in specific cases, the biases introduced by them, and the amount of information in data left unutilized by these approximate methods.
+For this reason, there is interest in improved phylodynamic inference techniques.
 
@@
 
-What would an improved phylodynamic inference approach look like?
+What would an ideal phylodynamic inference method look like?
 First, it would afford exact computation of the phylodynamic likelihood, so that comparisons among parameterizations and models could be made on a sound basis.
-Second, because nonlinearity, noise, and measurement error are so prominent and ubiquitous in epidemiology, it would accommodate nonlinear, stochastic transmission models.
+Second, because nonlinearity, nonstationarity, noise, and measurement error are so prominent and ubiquitous in epidemiology, it would accommodate nonlinear, time-inhomogeneous, stochastic transmission models.
 Third, because many of the most important uncertainties concern heterogeneities in transmission rates and the susceptibility, behavior, age, and location of hosts, it would accommodate host populations structured by these factors.
@@
 That is, that they are \emph{exchangeable}.
-For example, in a compartmental model, one can assume that individual hosts within a compartment are exchangeable.
-This implies that their residence times within a compartment are identically distributed, though not independent.
-Although this is indeed an additional assumption, it is so natural that it is frequently unrecognized as such, and one often reads statements to the effect that exchangeability of individuals is a consequence of the Markovian assumption about the discretely structured population.
-Nonetheless, since it adds minimum additional structure, it is the natural assumption, and the one we will make in this paper.
+In a compartmental model, this is tantamount to the assumption that the residence times of the individuals within each compartment are identically distributed, though not independent.
+Although exchangeability is indeed an additional assumption, it is so natural that it is frequently unrecognized as such, and one often reads statements to the effect that exchangeability of individuals is a consequence of the Markovian assumption.
+Nonetheless, since it adds minimal additional structure, it is the natural assumption, and the one we will make in this paper.
 
@@
 Of course, with the ability to entertain models with countably many compartments, much greater complexity is possible.
-In particular, one can model not only complex infection progression, but also strain structure, behavioral structure (\eg superspreading), age structure, and spatial structure using compartmental models.
+In particular, one can model not only complex infection progression, but also strain structure, behavioral structure, age structure, and spatial structure using compartmental models.
 As is well known, one can discretize continuous structure-variables and employ the linear chain trick to accommodate non-exponential residence times.
@@
   \input{figs/example_models}
+  \caption{
+    Examples of discretely-structured population models.
+    Demes are shaded.
+    Compartments containing infectious hosts are outlined in green.
+    Curved green lines connect transmission rates with the compartments whose occupancies control their modulation;
+    each such connection gives rise to a nonlinearity in the model.
+    \textbf{(A)} An SEIRS model.
+    Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
+    Upon recovery ($\lab{R}$), individuals experience immunity from reinfection.
+    If this immunity wanes, they re-enter the susceptible compartment.
+    Pathogen lineages are to be found in hosts within the $\lab{E}$ and $\lab{I}$ compartments only.
+    Accordingly, there are two demes: $\Demes=\Set{\lab{E},\lab{I}}$.
+    If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
+    See \cref{sec:demes} for definition and discussion of demes and deme occupancy.
+    \textbf{(B)} In this four-deme model, two distinct pathogen strains compete for susceptibles.
+    \textbf{(C)} A three-deme model according to which, after an incubation period, hosts may develop asymptomatic infection ($\lab{I_A}$).
+    If they do not recover, symptomatically infected hosts ($\lab{I_S}$) can progress to hospitalization ($\lab{H}$) and death ($\lab{D}$).
+    \textbf{(D)} A three-deme model with heterogeneity in transmission behavior.
+    Contagious individuals move randomly between low-transmission ($\lab{I_L}$) and high-transmission ($\lab{I_H}$) behaviors.
+    \label{fig:example_models}
+  }
 \end{figure}
@@
 \end{equation*}
-for all $t\in\Rp$ and $x,x'\in\Xspace$.
+for all $t\in\Rp$ and $x,x'\in\Xspace$ and that, as a function of time, $\alpha$ is continuous almost everywhere.
 Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
@@
 \end{equation*}
-We further assume that $\alpha(t,x,x')$ is \cadlag\ as a function of time for all $x,x'\in{\Xspace}$ and that the number of jumps that occur in a finite time-interval is finite: $\Prob{\Kr_t<\infty}=1$ for all $t$.
+We further assume that $\alpha(t,x,x')$ is \cadlag\ as a function of time for all $x,x'\in{\Xspace}$ and that the number of jumps that occur in a finite time-interval is finite, \ie $\Prob{\Kr_t<\infty}=1$ for all $t$.
 
@@
 
-For modeling purposes, it is sometimes desirable to insist that certain events occur at specific, known times.
+For modeling purposes, it is sometimes desirable to insist that certain events occur at known times.
 For example, if samples are collected at specific times in such a way that the timing itself conveys no information about the process, one might wish to condition on the sampling time.
@@
   \label{eq:kfe-sing}
-  v(t,x)
-  &=\int\!{\leftlim{v}(t,x')\,\pi(t,x',\dd{x'})},
+  v(t,x)\,\dd{x}
+  &=\int\!{\leftlim{v}(t,x')\,\pi(t,x',\dd{x})\,\dd{x'}},
   &t\in{S}.
@@
 
-Using a Dirac delta notation, one can represent \cref{eq:kfe-reg,eq:kfe-sing} as a single equation in the form of \cref{eq:kfe}.
+As a matter of notation, one can represent \cref{eq:kfe-reg,eq:kfe-sing} as a single equation in the form of \cref{eq:kfe}.
 In particular, if in \cref{eq:kfe} we make the substitution
 \begin{equation*}
-  \alpha(t,x,x')\mapsto\alpha(t,x,x')+\sum_{s\in{S}}{\delta(t,s)\,\frac{\dd{\pi}}{\dd{x}}(t,x,x')},
+  \alpha(t,x,x')\mapsto\alpha(t,x,x')+\sum_{s\in{S}}{\delta(t,s)\,\frac{\dd{\pi}}{\dd{x'}}(t,x,x')},
 \end{equation*}
 we obtain an equation which we can view as shorthand for \cref{eq:kfe-reg,eq:kfe-sing}.
-Here, $\dd{\pi}/\dd{x}$ denotes the density (\ie Radon-Nikodym derivative) of $\pi$ with respect to the measure on $\Xspace$.
+Here, $\delta(t,s)$ is a Dirac delta function and $\dd{\pi}/\dd{x'}$ denotes the density (\ie Radon-Nikodym derivative) of $\pi$ with respect to the measure on $\Xspace$.
 %% To be even more parsimonious, we can refer to \cref{eq:kfe-reg,eq:kfe-sing} as the $(\alpha,\pi,S)$ KFE.
@@
 
-It will be useful to divide the jumps of the population process $\Xr_t$ into distinct categories.
+It will be useful to divide the jumps of the population process $\Xr_t$ into distinct categories, which differ with respect to the changes they induce in a genealogy.
 For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
@@
 \end{equation*}
-\Cref{fig:markov_state} shows an example with five distinct marks.
+\Cref{fig:markov_state} shows an example for which $\Jumps$ has five elements.
 In the following, sums over $u$ are to be taken over the whole of $\Jumps$ unless otherwise indicated.
@@
 We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
-Demes can be used to introduce spatial, behavioral, physiological, and life-history structure into a model.
-The flexibility this allows can also be used, via the linear chain trick, to introduce delays into the dynamics.
 \Cref{fig:example_models} illustrates this concept in the context of several compartmental models.
@@
 In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic, as are $\Kr_t$, the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
-The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
+The probability measure on the space of histories can be expressed in terms of these:
 \begin{mathsize}{9pt}{10pt}
@@
     \label{eq:Hdens}
-    \pi^{\mathrm{H}}(\dd{\H_t})
+    \Prob{\dd{\H_t}}
     =p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
@@
 Since $a,a'\in{b}\in{Z_t}$ implies $Y_t(a)=Y_t(a')$, one can equally well think of $Y_t$ as a map $Z_t\to\Demes\times\Zp$.
-Given a tree $Z$, we let $\func{Y}(Z)$ denote the set of colorings $Y$ that are compatible with $Z$ and $\func{Y}_t(Z)\coloneq\CondSet{Y_t}{Y\in{\func{Y}(Z)}}$.
+Given a tree $Z$, we let $\func{Y}(Z)$ denote the set of colorings $Y$ that are compatible with $Z$.
+We moreover define $\func{Y}_t(Z)\coloneq\CondSet{Y_t}{Y\in{\func{Y}(Z)}}$.
 Formally speaking, $\func{Y}(Z)$ is a fiber bundle over $Z$, each $\func{Y}_t(Z)$ being a fiber.
@@
   We permit multiple lineages to move simultaneously.
-\item \emph{Sample-type events} result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
+\item \emph{Sample-type events} result in the collection of a sample from a lineage.
   We allow for the possibility that multiple samples are collected simultaneously, though we require that, in this case, each extant lineage is sampled at most once.
@@
   \input{figs/event_types}
+  \caption{
+    Event types differ by their effects on the genealogy.
+    This can be seen by examining the local structure of the genealogy in the neighborhood of a jump.
+    \textbf{(A)} A birth-type jump results in the branching of one or more child lineages from the parent.
+    There can be only one parent, though the demes of the child lineages may differ from that of their parent.
+    Here, a parent of the blue deme sires one child lineage in each of the blue and yellow demes.
+    The \emph{production} of an event is an integer vector, with one entry for each deme.
+    The production of this event is therefore $r=(r_{\lab{blue}},r_{\lab{yellow}})=(2,1)$.
+    The \emph{deme occupancy} of an event is the number of lineages in each deme just to the right of the event.
+    The deme occupancy at this event is therefore $n=(n_{\lab{blue}},n_{\lab{yellow}})=(3,5)$.
+    \textbf{(B)} A death-type event causes the extinction of a lineage.
+    Since internal nodes without children are recursively removed, the affected branch is dropped.
+    The production of this event is $r=(0,0)$ and the deme occupancy is $n=(3,4)$.
+    \textbf{(C)} A migration-type event results in the movement of one or more lineages from one deme to another.
+    Here, one lineage moves from the yellow to the blue deme.
+    The production of this event is $r=(1,0)$, \ie the production is 1 for the blue deme and 0 for the yellow.
+    The deme occupancy is $n=(6,2)$.
+    \textbf{(D)} In a sample-type event, one or more sample nodes (blue circles) are inserted.
+    Here, there are two samples, one in each of the blue and yellow demes.
+    Accordingly, $r=(1,1)$ and $n=(2,6)$.
+    \textbf{(E)} A neutral-type event has no effect on the genealogy and zero production in all demes: $r=(0,0)$, $n=(5,3)$.
+    \textbf{(F)} The theory presented here allows for compound events.
+    As an example, here a birth/death-type event occurs, wherein one yellow lineage is extinguished and a blue lineage simultaneously sires a blue child.
+    For this event, we have $r=(2,0)$ and $n=(6,2)$.
+    \textbf{(G)} Here, a compound sample/death-type event with $r=(0,0)$ and $n=(2,5)$ occurs.
+    A blue lineage is sampled and simultaneously extinguished.
+    Note that recursive removal does not occur, since sample nodes are never removed.
+    \textbf{(H)} A compound birth/migration-type event with $r=(4,0)$ and $n=(6,2)$.
+    \label{fig:event_types}
+  }
 \end{figure}
@@
   These are deleted.
-  Next, branch nodes without children are recursively removed.
+  Next, internal nodes without children are recursively removed.
   Sample nodes are never removed.
@@
 \subsubsection{Emergent lineages and production}
+\label{sec:production}
 
@@
 We require that $r^u_i$ be a constant, for each $u$ and $i$.
-Since, in applications, one is free to expand the set of jump-marks $\Jumps$ as needed, this is not an important restriction on the models that the theory can accommodate.
 Thus there is a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
+Since, in applications, one is free to expand the set of jump-marks $\Jumps$ as needed, this is not a restriction on the models that the theory can accommodate.
 We say $r^u\coloneq{(r^u_i)_{i\in\Demes}}$ is the \emph{production} of an event of mark $u$.
@@
 It is very important to note that, conditional on $(\Xrh_k,\Urh_k)$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of these choices at any other times and independent of $(\Xrh_j,\Urh_j)$ for all $j\ne{k}$.
-Moreover, by construction, the lineages within each deme are exchangeable:
+Moreover, by assumption, the lineages within each deme are exchangeable:
 any lineage within a deme is as likely as any other lineage in that deme to be selected as a parent or for death, sampling, or migration.
@@
     Sample nodes are never removed.
-    \textbf{(C)} A genealogy is \emph{obscured} by effacing all deme information from lineage histories:
+    \textbf{(C)} A pruned genealogy is \emph{obscured} by effacing all deme information from lineage histories:
     the colors are erased, as are all inline nodes.
@@
 Moreover, it details each lineage's complete history of movement through the various demes.
-The data we ultimately wish to analyze will be based only on samples, however.
+However, the data we ultimately wish to analyze will be based only on samples.
 Nor, in general, will the histories of deme occupancy be observable.
@@
   \input{figs/ells}
+  \caption{
+    \textbf{Lineage count and saturation.}
+    Each panel shows the neighborhood of a single event in the unpruned genealogy (top row) and the corresponding pruned genealogy (bottom row).
+    Pruning consists of the removal of all branches that are not ancestral to some sample.
+    In the bottom row of panels, pruned branches are indicated using broken lines.
+    \textbf{(A)} A birth-type event with production $r=(r_{\lab{blue}},r_{\lab{yellow}})=(1,1)$ occurs.
+    \textbf{(B)} Suppose that pruning results in the removal of the dashed lineages.
+    Then the lineage count at this event-time is $\ell=(\ell_{\lab{blue}},\ell_{\lab{yellow}})=(2,2)$.
+    The saturation is $s=(0,1)$ since only a single, yellow lineage emerges from the event.
+    \textbf{(C)} A migration-type event with production $r=(0,1)$ occurs.
+    \textbf{(D)} After pruning, $\ell=(2,2)$ and $s=(0,1)$.
+    \textbf{(E)} A sample-type event occurs in which two blue lineages are sampled (production $r=(2,0)$).
+    \textbf{(F)} After pruning, $\ell=(2,2)$ and $s=(1,0)$.
+    Observe that in panels B and D, the local structures of the pruned genealogies are identical, though they arise from events of different type.
+    \label{fig:ells}
+  }
 \end{figure}
@@
   \end{equation}
-  where $n$ is the deme occupancy (\cref{sec:demes}), $\ell$ and $s$ are the lineage-count and saturation functions, respectively (\cref{sec:ells}), $Q$ is the compatibility indicator (\cref{sec:compatibility}), and the binomial ratio is as defined in \cref{sec:binomial_ratio}.
+  where $n$ is the deme occupancy (\cref{sec:demes}), $r^u$ is the production (\cref{sec:production}), $\ell$ and $s$ are the lineage-count and saturation functions, respectively (\cref{sec:ells}), $Q$ is the compatibility indicator (\cref{sec:compatibility}), and the binomial ratio is as defined in \cref{sec:binomial_ratio}.
   Then
@@
   This exchangeability implies that each lineage present in a deme at time $t$ was equally likely to have been one of the emergent lineages.
-  In particular, at time $t$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ red balls and $n_i-r_i$ black balls, exactly $s_i$ of the drawn balls are red, namely
+  In particular, at time $t$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the unpruned genealogy process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ red balls and $n_i-r_i$ black balls, exactly $s_i$ of the drawn balls are red, namely
   \begin{equation*}
@@
   \end{equation*}
-  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct;
+  Because our lineages are labelled, each of the $\tbinom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct;
   just one of these is the one present in $\P$.
@@
 
-\Cref{lemma:filter} shows how this can be computed via Sequential Monte Carlo.
+\Cref{lemma:monte_carlo} shows how this can be computed via Sequential Monte Carlo.
 
@@
 
-This approach gives us the freedom to choose models with many demes.
-Moreover, there is freedom to choose the sampling model.
-
-Generalization of coalescent and birth-death process approaches.
-Both Moran model and birth-death processes are special cases.
-
-Allows models with demographic stochasticity.
-Incorporating environmental stochasticity is likely possible:
-extension to stochastic processes with a diffusion component.
-
-The price for this flexibility is variability in the Monte Carlo estimation.
-However, there is substantial freedom in the choice of the importance sampling distribution $\pi$.
-Moreover, since it is permissible to borrow information from the future, one can envision importance sampling distributions tailored to any given genealogy.
-
-Filter equation formalism suggests approximations based on discretization of time.
+Importantly, because the theory presented here allows computation of the likelihood via strictly forward-in-time computations, it permits consideration of models for which time-reversal arguments are not available.
+Moreover, inasmuch as the formulae of \cref{thm:obsc_uncond} can be efficiently computed via sequential Monte Carlo, explicit expressions for transition probabilities are not needed:
+it is sufficient to be able to simulate from the population process.
+This feature of the algorithms---known as the \emph{plug-and-play property}---expands the set of population models that can be confronted with data still further.
+
+In particular, the theory gives us the freedom to choose models with many demes.
+For deterministic population models, \citet{Volz2012} and \citet{Rasmussen2014a} showed how one could accommodate discrete population structure.
+Their procedures involve solving a large number of differential equations backward in time, relying on the time-reversibility of deterministic dynamics.
+In general, this time-reversibility is not a property of stochastic processes.
+
+Some existing methods put rather severe limits on the form of the sampling model and, as \citet{Volz2014a} pointed out, misspecification of the sampling model can lead to large inferential biases.
+With the theory presented here, essentially arbitrary specification of the sampling model is possible.
+In particular, one can posit sampling at a rate which is an arbitrary function of time and state and include discrete sampling events as well.
+It is also possible to condition on the existence of samples.
+
+If Sequential Monte Carlo algorithms are used to compute the likelihoods of \cref{thm:obsc_uncond}, then it is straightforward to simultaneously assimilate information from both time-series and genealogical data.
+One can therefore supplement traditional incidence, disease, or mortality time series with genealogical data in an inferential exercise.
+
+The theory presented here represents a strict generalization of the existing coalescent and birth-death process approaches to phylodynamic inference.
+Both of the latter processes represent special cases of the genealogical processes constructed here.
+In \cref{sec:app-examples}, this is demonstrated.
+
+A limitation of the theory presented here is that the population models are assumed to be pure jump processes, which allows consideration of demographic stochasticity and some kinds of environmental stochasticity.
+It should be possible to incorporate of the full range of Markovian environmental stochasticity via extension of this theory to population models containing both diffusion and jump components.
+
+The price of the theory's flexibility is primarily computational:
+Monte Carlo variability in the likelihood estimate depends on computational effort in the usual way.
+However, the substantial freedom one has in the choice of the importance-sampling distribution $\pi$ (\cref{thm:obsc_uncond}) can be exploited to diminish Monte Carlo error.
+In particular, since it is permissible to ``borrow information'' from the future by means of the importance sampling, there is hope for highly efficient algorithmic computation.
+
+\section*{Acknowledgments}
+
+This work was supported by grants from
+the U.S. National Institutes of Health, (Grant \#1R01AI143852 to AAK, \#1U54GM111274 to AAK and ELI)
+and a grant from the Interface program, jointly operated by the U.S. National Science Foundation and the National Institutes of Health (Grant \#1761603 to ELI and AAK).
+QL acknowledges the support of the Michigan Institute for Data Science.
 
 \bibliographystyle{preprint}
-\bibliography{ms}
+\bibliography{phylopomp}
 
@@
 
-Explicit expressions for the quantities that arise in this paper are not always readily available.
-Here, we develop tools for manipulating complex expressions that are otherwise cumbersome.
-
-Relations to filtering of continuous-time Markov chains by \citet{Giesecke2018}.
+The likelihoods that appear in \cref{thm:pruned_uncond,thm:obsc_uncond} are integrals over large sets of histories.
+As such, explicit expressions for them are not available, and we require mathematical tools to allow us to manipulate these quantities and devise algorithms for their numerical solution.
+The \emph{filtering equations} we introduce here are suitable for these purposes, and we devote this appendix to exposing their essential properties.
+This extremely convenient formalism has, to our knowledge, not been thoroughly exploited, though we note their resemblance to the constructions of \citet{Ogata1978}, \citet{Puri1986}, \citet{Kliemann1990}, and \citet{Giesecke2018}.
 
@@
   In the limit, we obtain \cref{eq:reg-filter-eq2}, the regular filtering equation generated by $\alpha$, with boost $B$ and zero decay.
-  \AAK{[This depends on $\alpha$ and $B$ being continuous in their first arguments.]}
+  %%  \AAK{[This depends on $\alpha$ and $B$ being continuous in their first arguments. We have assumed they are continuous almost everywhere.]}
 \end{proof}
@@
 
+\begin{remark}
+  In the same way that \cref{eq:kfe-reg,eq:kfe-sing} can be represented as a single equation by means of a Dirac delta notation, the \cref{eq:two-part-filter-reg,eq:two-part-filter-sing} can be collapsed into a more compact form if $\beta$ is allowed to have atoms at a countable set of time-points and the boost $B$ is adjusted appropriately.
+\end{remark}
+
+Filter equations afford a convenient means of computing expectations and likelihoods for pure jump processes.
+This is facilitated by the following Lemma, the statement of which uses a one-sided Dirac delta function.
+Specifically, let $\delta(v,v')$ be the right-sided Dirac delta function satisfying $\delta(v,v')=0$ for $v\ne{v'}$ and
+\begin{equation*}
+  \int_a^b{f(v)\,\delta(v,v')\,\dd{v}}=f(v')\,\Indicator{v'\in\halfopen{a,b}},
+\end{equation*}
+whenever $f$ is \cadlag\ and $-\infty\le{a}<{b}\le{\infty}$.
+
 \begin{lemma}
-  Showing how \cref{eq:two-part-filter-reg,eq:two-part-filter-sing} can be stated as a singular filter equation, \ie with driver $\beta$ and boost $C$, where
-  \begin{equation*}
-    \begin{gathered}
-      A(t,x)=\sum_{x'}{\alpha(t,x,x')}, \qquad
-      \beta(t,x,x')=\alpha(t,x,x')+\sum_{e\in\mathcal{M}}{\delta(t,e)\,\frac{\alpha(t,x,x')}{A(t,x)}}\,,\\
-      C(t,x,x')=B(t,x,x')+\sum_{e\in\mathcal{M}}{\delta(t,e)\,A(t,x)}.
-    \end{gathered}
-  \end{equation*}
-  %% \begin{equation}
-  %%   \label{eq:sing-filter-eq}
-  %%   \begin{aligned}
-  %%     \frac{\partial{w}}{\partial{t}}(t,x)
-  %%     =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
-  %%     -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}\\
-  %%     &+\sum_e\int{w(t,x')\,\delta(t,e)\,\pi(t,x',x)\,C(t,x',x)\,\dd{x'}}
-  %%     -\sum_e\int{w(t,x)\,\delta(t,e)\,\pi(t,x,x')\,\dd{x'}}\\
-  %%     &-\lambda(t,x)\,w(t,x).\\
-  %%   \end{aligned}
-  %% \end{equation}
-\end{lemma}
-
-Filter equations afford a convenient means of computing expectations and likelihoods for pure jump processes.
-This is facilitated by the following Lemma, the statement of which uses a one-sdied Dirac delta function.
-Specifically, let $\delta(v,v')$ be the right-sided Dirac delta function satisfying $\delta(v,v')=0$ for $v\ne{v'}$ and
-\begin{equation*}
-  \int_a^b{f(v)\,\delta(v,v')\,\dd{v}}=f(v')\,\Indicator{v'\in\halfopen{a,b}},
-\end{equation*}
-whenever $f$ is \cadlag\ and $-\infty\le{a}<{b}\le{\infty}$.
-
-\begin{lemma}
-  \label{lemma:filter}
-  The filter equation \eqref{eq:reg-filter-eq-reg} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satisfies the KFE
+  \label{lemma:monte_carlo}
+  \Cref{eq:filter-eq-defn-reg,eq:filter-eq-defn-sing} are satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satisfies the KFE
+  \begin{mathsize}{9pt}{10pt}
     \begin{equation}
-    \label{eq:filterlemma}
       \begin{aligned}
-      \frac{\partial{u}}{\partial{t}}=
-      &\int_0^{\infty}\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int_0^{\infty}\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
-      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right].
+        \label[pluralequation]{eq:filterlemma}
+        \frac{\partial{u}}{\partial{t}}(t,x,v)=
+        &\frac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]
+        +\int_0^{\infty}\int{u(t,x',v')\,\beta(t,x',x)\,\delta\!\left(v,B(t,x',x)\,v'\right)\,\dd{x'}\,\dd{v'}}\\
+        &\quad-\int_0^{\infty}\int{u(t,x,v)\,\beta(t,x,x')\,\delta\!\left(v',B(t,x,x')\,v\right)\,\dd{x'}\,\dd{v'}},
+        &t\notin{S},\\
+        u(t,x,v)\,\dd{x}=
+        &\int_0^{\infty}\int{\ut(t,x',v')\,\pi(t,x',\dd{x})\,\delta\!\left(v,A(t,x')\,B(t,x',x)\,v'\right)\,\dd{x'}\,\dd{v'}},&t\in{S}.
       \end{aligned}
     \end{equation}
+  \end{mathsize}%
+  Here, $A(t,x)\coloneq{\int{\beta(t,x,x')\,\dd{x'}}}$
+  and $\pi(t,x,\dd{x'})\coloneq\beta(t,x,x')\,\dd{x'}/A(t,x)$.
 \end{lemma}
 \begin{proof}
+  For each $t\notin{S}$, we have
+  \begin{mathsize}{9pt}{10pt}
     \begin{equation*}
       \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
+        \frac{\partial{w}}{\partial{t}}(t,x)
         =&\int_0^{\infty}{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
-      =&\int_0^{\infty}\int\int_0^{\infty}{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
-      &\qquad-\int_0^{\infty}\int\int_0^{\infty}{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
+        =&\int_0^{\infty}\int\int_0^{\infty}{v\,u(t,x',v')\,\beta(t,x',x)\,\delta\!\left(v,B(t,x',x)\,v'\right)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
+        &\qquad-\int_0^{\infty}\int\int_0^{\infty}{v\,u(t,x,v)\,\beta(t,x,x')\,\delta\!\left(v',B(t,x,x')\,v\right)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
         &\qquad+\int_0^{\infty}{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
@@
     \end{equation*}
+  \end{mathsize}%
   Here, the non-explosivity assumption guarantees that we can differentiate under the integral sign and exchange the order of integration.
@@
   Hence, by evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
+  \begin{mathsize}{9pt}{10pt}
     \begin{equation*}
       \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}
+        \frac{\partial{w}}{\partial{t}}(t,x)
         =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
@@
     \end{equation*}
-  which is simplified to obtain \cref{eq:reg-filter-eq}.
+  \end{mathsize}%
+  which is simplified to obtain \cref{eq:filter-eq-defn-reg}.
+  Similarly, at each $t\in{S}$, we have
+  \begin{mathsize}{9pt}{10pt}
+    \begin{equation*}
+      \begin{aligned}
+        w(t,x)\,\dd{x}
+        =&\int_0^{\infty}\int\int_0^{\infty}{v\,\ut(t,x',v')\,\pi(t,x',\dd{x})\,\delta\!\left(v,A(t,x')\,B(t,x',x)\,v'\right)\,\dd{x'}\,\dd{v'}\,\dd{v}}\\
+        =&\int_0^{\infty}\int{v'\,\ut(t,x',v')\,\pi(t,x',\dd{x})\,A(t,x')\,B(t,x',x)\,\dd{x'}\,\dd{v'}}\\
+        =&\int{\wt(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}\,\dd{x}.
+      \end{aligned}
+    \end{equation*}
+  \end{mathsize}%
+  which is equivalent to \cref{eq:filter-eq-defn-sing}
 \end{proof}
 
-\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(\Xr_t,\Vr_t)$.
-In particular, $\Xr_t$ is the driver with KFE \eqref{eq:kfe2}.
-The $\Vr_t$ is \emph{directed} by $\Xr_t$ in the sense that $\Vr$ has jumps wherever $\Xr$ does:
+\begin{remark}
+  \Cref{eq:filterlemma} are recognizable as the KFE of a certain process $(\Xr_t,\Vr_t)$.
+  In particular, the driver $\Xr_t$ has KFE \cref{eq:kfe2}.
+  $\Vr_t$ is \emph{directed} by $\Xr_t$ in the sense that $\Vr$ has jumps wherever $\Xr$ does:
   when $\Xr$ jumps at time $t$ from $x$ to $x'$, $\Vr$ jumps by the multiplicative factor $B(t,x,x')\ge{0}$.
   Between jumps, $\Vr_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
-If we view $\Vr_t$ as a weight, then \cref{lemma:filter} tells us how the $\Vr_t$-weighted average of $\Xr_t$ evolves in time:
-this average is simply $\int\!{w(t,x)\,\dd{x}}$.
+  At the known times in $S$, $\Xr$ jumps according to the probability kernel $\pi$ and, $\Vr$ jumps by the factor $A(t,x)\,B(t,x,x')$.
+  If we view $\Vr_t$ as a weight, then \cref{lemma:monte_carlo} tells us how the $\Vr_t$-weighted average of $\Xr_t$ evolves in time:
+  this average is simply $\int{w(t,x)\,\dd{x}}$.
+  Thus, \cref{lemma:monte_carlo} shows how to integrate \cref{eq:two-part-filter-reg,eq:two-part-filter-sing} in the Monte Carlo sense.
+\end{remark}
 
@@
 Its state vector is $x=(S,I,R)$ and its KFE is
-\begin{mathsize}{10pt}{11pt}
+\begin{mathsize}{9pt}{10pt}
   \begin{equation*}
@@
 It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
-\begin{mathsize}{10pt}{11pt}
+\begin{mathsize}{9pt}{10pt}
   \begin{equation*}
@@
 This model has KFE
-\begin{mathsize}{10pt}{11pt}
+\begin{mathsize}{9pt}{10pt}
   \begin{equation*}
@@
 The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
-\begin{mathsize}{10pt}{11pt}
+\begin{mathsize}{9pt}{10pt}
   \begin{equation*}
@@
 The KFE is
-\begin{mathsize}{10pt}{11pt}
+\begin{mathsize}{9pt}{10pt}
   \begin{equation*}
@@
 If $B$ is the set of branch-times and $S_0$, $S_1$ are the sets of terminal and inline samples, respectively, then the regular part of the filter equation is
-\begin{mathsize}{10pt}{11pt}
+\begin{mathsize}{9pt}{10pt}
   \begin{equation}
@@
 and the singular part is
-\begin{mathsize}{10pt}{11pt}
+\begin{mathsize}{9pt}{10pt}
   \begin{equation}
@@
   \end{equation}
-\end{mathsize}
+\end{mathsize}%
 \Cref{eq:filter-lbdp-reg,eq:filter-lbdp-sing} are supplemented by the ancillary condition $w(t,n)=0$ for $n<\ell(t)$.
@@
 Its KFE is then
-\begin{mathsize}{10pt}{11pt}
+\begin{mathsize}{9pt}{10pt}
   \begin{equation*}
@@
 Then, if $B$ is the set of branch-times and $\ell(t)$ is the number of lineages in the genealogy at time $t$, the filter equation reads
-\begin{mathsize}{10pt}{11pt}
+\begin{mathsize}{9pt}{10pt}
   \begin{equation}
@@
 where $k=|B|$ is the number of branch-points in $[0,T]$ and the $s^{}_{i}\coloneq\int{\Indicator{\ell(t)=i}\,\dd{t}}$ are the durations of the \emph{coalescent intervals}, \ie intervals between successive branch-points.
-We recognize \cref{eq:kingman} as the expression for the \citet{Kingman1982a} coalescent \citep[\eg][]{Wakeley2008}.
+We recognize \cref{eq:kingman} as the expression for the \citet{Kingman1982a} coalescent \citep[\eg][]{Wakeley2009}.
 
 More generally, if in addition samples are taken according to a rate-$\nu$ Poisson process such that the set of sample-times in the genealogy is $S=S_0\cup{S_1}$, where $S_0$, $S_1$ are the sets of times of terminal and inline samples, respectively, then the filter equation reads
-\begin{mathsize}{10pt}{11pt}
+\begin{mathsize}{9pt}{10pt}
   \begin{equation}

### 2024-05-16 ms.Rnw
postpone examples
figures for examples

@@
     \begin{split}
-      \frac{\partial{v}}{\partial{t}}(t,S,I,R)=
-      &\frac{\beta(t)\,(S+1)\,(I-1)}{N}\,v(t,S+1,I-1,R)
-      -\frac{\beta(t)\,S\,I}{N}\,v(t,S,I,R)\\
+      \frac{\partial{v}}{\partial{t}}(S,I,R)=
+      &\frac{\beta\,(S+1)\,(I-1)}{N}\,v(t,S+1,I-1,R)
+      -\frac{\beta\,S\,I}{N}\,v(t,S,I,R)\\
       &+\gamma\,(I+1)\,v(t,S,I+1,R-1)
@@
 Here $N=S+I+R$ is the host population size.
-Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.
+Note that, though the theory allows for time-dependent event rates, this example is time-homogenous.
+This model has one deme, $\Demes=\Set{\lab{I}}$, with occupancy function $n(x)=I$.
+There are four kinds of jumps:
+transmission, recovery, waning of immunity, and sampling.
+Accordingly, the marks are $\Jumps=\Set{\lab{Trans},\lab{Recov},\lab{Wane},\lab{Sample}}$.
+\Cref{tab:sirs_model_elements} gives $\alpha_u$, $r_u$, and the jump-type for each of these marks.
+
+\begin{table}
+  \caption{
+    \label{tab:sirs_model_elements}
+    Elements of the SIRS model pertinent to the genealogy process.
+  }
+  \renewcommand{\arraystretch}{1.2}
+  \begin{tabular}{cclc}
+    \hline\hline
+    $u$ & $\alpha_u$ & Event type & $r_u$ \\
+    \hline
+    $\lab{Trans}$  & ${\beta\,S\,I}/{N}$  & pure birth  & 2 \\
+    $\lab{Recov}$  & $\gamma\,I$          & pure death  & 0 \\
+    $\lab{Wane}$   & $\omega\,R$          & neutral     & 0 \\
+    $\lab{Sample}$ & $\psi\,I$            & pure sample & 1 \\
+    \hline\hline
+  \end{tabular}
+\end{table}
+
+Given an obscured genealogy $Z$, let $\event{Z}=B\cup{S_0}\cup{S_1}$, where $B$ is the set of branch-times, and $S_0$, $S_1$ are the sets of sample-times with saturations $0$ and $1$, respectively.
+Since there is only one deme, paintings of $Z$ can differ only in number and position of inline, internal nodes along branches.
+Each of these can only correspond to $u=\lab{Trans}$ with $s=1$.
+For $t\notin{\event{Z}}$, we can take the importance sampling distribution to be
+\begin{equation*}
+  \pi_u=\begin{cases}
+  c,                & u=\lab{Trans}, s=0, t\notin{\event{Z}},\\
+  \frac{1-c}{\ell}, & u=\lab{Trans}, s=1, t\notin{\event{Z}},\\
+  0,                & \text{otherwise}.
+  \end{cases}
+\end{equation*}
+Here $c$ is an arbitrary probability that does not affect the computation.
+The relevant binomial ratios are
+\begin{equation*}
+  \BinRatio{n}{\ell}{r}{s}=
+  \begin{cases}
+    \frac{(I-\ell)\,(I-\ell-1)}{I\,(I-1)}, & u=\lab{Trans}, s=0, \\[1ex]
+    \frac{2\,(I-\ell)}{I\,(I-1)}, & u=\lab{Trans}, s=1, \\[1ex]
+    \frac{2}{I\,(I-1)}, & u=\lab{Trans}, s=2, \\[1ex]
+    %% 1, & u=\lab{Recov}, I\ge{\ell}, \\[1ex]
+    %% 0, & u=\lab{Recov}, I=\ell, \\[1ex]
+    %% 1, & u=\lab{Wane}, \\[1ex]
+    \frac{I-\ell}{I}, & u=\lab{Sample}, s=0, \\[1ex]
+    \frac{1}{I}, & u=\lab{Sample}, s=1. \\[1ex]
+  \end{cases}
+\end{equation*}
+
+This leads, for $t\notin{\event{Z}}$, to the following regular part of the filter equation:
+\begin{mathsize}{9pt}{10pt}
+  \begin{equation*}
+    \begin{split}
+      \frac{\partial{w}}{\partial{t}}=
+      &\frac{\beta\,(S+1)\,(I-1)}{N}\,\left(1-\frac{\tbinom{\ell(t)}{2}}{\tbinom{I}{2}}\right)\,w(t,S+1,I-1,R)
+      -\frac{\beta\,S\,I}{N}\,w(t,S,I,R)\\
+      &+\gamma\,(I+1)\,w(t,S,I+1,R-1)
+      -\gamma\,I\,w(t,S,I,R)\\
+      &+\omega\,(R+1)\,w(t,S-1,I,R+1)
+      -\omega\,R\,w(t,S,I,R)\\
+      &-\psi\,I\,w(t,S,I,R).
+    \end{split}
+  \end{equation*}
+\end{mathsize}%
+Here, we have summed over the various paintings for $u=\lab{Trans}$, $s<2$.
+Note the presence of the decay term proportional to $\psi$.
+At event-times $t\in\event{Z}$, the singular part of the filter equation reads
+\begin{mathsize}{9pt}{10pt}
+  \begin{equation*}
+    w(t,S,I,R)=
+    \begin{cases}
+      \wt(t,S+1,I-1,R)\,\frac{2\,\beta\,(S+1)}{N\,I}, & t\in{B}, \\[1ex]
+      \wt(t,S,I,R)\,\psi\,(I-\ell(t)), & t\in{S_0}, \\[1ex]
+      \wt(t,S,I,R)\,\psi, & t\in{S_1}. \\[1ex]
+    \end{cases}
+  \end{equation*}
+\end{mathsize}%
+Finally, note that $w(t,S,I,R)=0$ for all $I<\ell(t)$.
+
+<<sirs3>>=
+bake(
+  file="sirs3a.rds",
+  seed=328168304L,
+  {
+    library(phylopomp)
+    runSIRS(
+      Beta=4,gamma=2,psi=1,omega=1,
+      S0=97,I0=3,R0=0,t0=0,time=40
+    )
+  }
+) -> x
+
+bake(
+  file="sirs3b.rds",
+  seed=621400057L,
+  {
+    expand_grid(
+      Beta=4,
+      ##      gamma=2,
+      gamma=seq(1.7,2.3,by=0.02),
+      psi=1,
+      omega=1,
+      ##      omega=seq(0.5,2,by=0.05),
+      S0=97,I0=3,R0=0,
+      t0=0,
+      rep=1:16,
+      Np=5000,
+      ) -> params
+
+    library(iterators)
+    library(doFuture)
+    plan(multicore)
+
+    foreach (
+      p=iter(params,"row"),
+      .combine=bind_rows,
+      .options.future=list(seed=TRUE)
+    ) %dofuture% {
+      library(pomp)
+      library(phylopomp)
+      p |>
+        with({
+          x |>
+            sirs_pomp(
+              Beta=Beta,gamma=gamma,psi=psi,omega=omega,
+              S0=S0,I0=I0,R0=R0,t0=0
+            ) |>
+            pfilter(Np=Np) |>
+            logLik()
+        }) -> ll
+      bind_cols(p,logLik=ll)
+    } -> params
+  }
+) -> params
+@
+
+\begin{figure}
+  \begin{center}
+    <<sirs3_plot,dependson="sirs1",fig.dim=c(8,2.8),out.width="100%">>=
+    plot_grid(
+      A=x |>
+        plot(points=FALSE,palette=c("#000000"))+
+        labs(x="time"),
+      B=params |>
+        group_by(Beta,gamma,psi,omega,S0,I0,R0,Np) |>
+        reframe(
+          name=c("logLik","logLik_se"),
+          value=logmeanexp(logLik,se=TRUE)
+        ) |>
+        ungroup() |>
+        pivot_wider() |>
+        mutate(
+          y=logLik,
+          ymax=logLik+2*logLik_se,
+          ymin=logLik-2*logLik_se
+        ) |>
+        filter(round(Beta,2)==4) |>
+        filter(logLik>max(logLik)-16) |>
+        ggplot(
+          aes(
+            x=gamma,
+            group=Np,
+            y=y,
+            ymin=ymin,
+            ymax=ymax
+          )
+        )+
+        geom_errorbar(position="dodge")+
+        geom_vline(xintercept=2,color="red")+
+        labs(
+          color=character(0),
+          y="log likelihood",
+          x=expression(gamma)
+        )+
+        theme_classic()+
+        theme(
+          legend.position=c(0.5,0.27)
+        ),
+      labels="AUTO",
+      nrow=1,
+      rel_widths=c(1,1)
+    )
+    @
+  \end{center}
+\end{figure}
 
@@
 The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
-It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
+The KFE for the population process is
 \begin{mathsize}{9pt}{10pt}
@@
 where $N=S+E+I+R$ is the total population size.
+This model has two demes: $\Demes=\Set{\lab{E},\lab{I}}$ and four
 The deme occupancy function in this case is $n(x)=(E,I)$.
@@
 A simple model for the competition of two strains for susceptible hosts is depicted in \cref{fig:example_models}B.
-In this model, the state vector consists of seven numbers: $x=(S,E_1,E_2,I_1,I_2,R_1,R_2)$.
-There are four demes ($\Demes=\Set{\lab{E}_1,\lab{E}_2,\lab{I}_1,\lab{I}_2}$) and the occupancy function is $n(x)=(E_1,E_2,I_1,I_2)$.
-This model has KFE
-\begin{mathsize}{9pt}{10pt}
-  \begin{equation*}
-    \begin{split}
-      \frac{\partial{v}}{\partial{t}}=
-      &\frac{\beta_1(t)\,I_1\,(S+1)}{N}\,v(t,S+1,E_1-1,E_2,I_1,I_2,R_1,R_2)
-      -\frac{\beta_1(t)\,I_1\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\frac{\beta_2(t)\,I_2\,(S+1)}{N}\,v(t,S+1,E_1,E_2-1,I_1,I_2,R_1,R_2)
-      -\frac{\beta_2(t)\,I_2\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\sigma_1\,(E_1+1)\,v(t,S,E_1+1,E_2,I_1-1,I_2,R_1,R_2)
-      -\sigma_1\,E_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\sigma_2\,(E_2+1)\,v(t,S,E_1,E_2+1,I_1,I_2-1,R_1,R_2)
-      -\sigma_2\,E_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\gamma_1\,(I_1+1)\,v(t,S,E_1,E_2,I_1+1,I_2,R_1-1,R_2)
-      -\gamma_1\,I_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\gamma_2\,(I_2+1)\,v(t,S,E_1,E_2,I_1,I_2+1,R_1,R_2-1)
-      -\gamma_2\,I_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\omega_1\,(R_1+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1+1,R_2)
-      -\omega_1\,R_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-      &+\omega_2\,(R_2+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1,R_2+1)
-      -\omega_2\,R_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
-    \end{split}
-  \end{equation*}
-\end{mathsize}%
+This example will be completed in a forthcoming draft.
 
@@
 \Cref{fig:example_models}D depicts a model of superspreading.
-There are three demes ($\Demes=\Set{\lab{E},\lab{I_L},\lab{I_H}}$).
-The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
-\begin{mathsize}{9pt}{10pt}
-  \begin{equation*}
-    \begin{split}
-      \frac{\partial{v}}{\partial{t}}=
-      &\frac{\beta(t)\,(I_L+\theta\,I_H)\,(S+1)}{N}\,v(t,S+1,E-1,I_L,I_H,R)
-      -\frac{\beta(t)\,(I_L+\theta\,I_H)\,S}{N}\,v(t,S,E,I_L,I_H,R)\\
-      &\sigma\,(E+1)\,v(t,S,E+1,I_L-1,I_H,R)
-      -\sigma\,E\,v(t,S,E,I_L,I_H,R)\\
-      &\eps_{LH}\,(I_L+1)\,v(t,S,E,I_L+1,I_H-1,R)
-      -\eps_{LH}\,I_L\,v(t,S,E,I_L,I_H,R)\\
-      &\eps_{HL}\,(I_H+1)\,v(t,S,E,I_L-1,I_H+1,R)
-      -\eps_{HL}\,I_H\,v(t,S,E,I_L,I_H,R)\\
-      &\gamma\,(I_L+1)\,v(t,S,E,I_L+1,I_H,R-1)
-      -\gamma\,I_L\,v(t,S,E,I_L,I_H,R)\\
-      &\gamma\,(I_H+1)\,v(t,S,E,I_L,I_H+1,R-1)
-      -\gamma\,I_H\,v(t,S,E,I_L,I_H,R)\\
-      &\omega\,(R+1)\,v(t,S-1,E,I_L,I_H,R+1)
-      -\omega\,R\,v(t,S,E,I_L,I_H,R).\\
-    \end{split}
-  \end{equation*}
-\end{mathsize}%
+This example will be completed in a forthcoming draft.
 
@@
 
+<<lbdp3>>=
+true.params <- data.frame(lambda=1.2,mu=0.8,psi=1,n0=5)
+
+bake(
+  file="lbdp3a.rds",
+  seed=505641575,  
+  true.params |>
+    with(
+      runLBDP(lambda=lambda,mu=mu,psi=psi,n0=n0,time=5)
+    ) -> x
+) -> x
+
+bake(
+  file="lbdp3b.rds",
+  dependson=x,
+  seed=715604304,
+  {
+    expand_grid(
+      rep=1:10,
+      lambda=1.2,
+      mu=seq(0.3,2,by=0.1),
+      ## mu=0.8,
+      ## psi=seq(0.2,2.5,by=0.1),
+      psi=1,
+      time=3,
+      n0=5,
+      Np=c(1000,2000,4000,8000,16000,32000,64000,128000)
+    ) -> params
+
+    library(iterators)
+    library(doFuture)
+    plan(multicore)
+
+    foreach (
+      p=iter(params,"row"),
+      .combine=bind_rows,
+      .options.future=list(seed=TRUE)
+    ) %dofuture% {
+      p |>
+        with(
+          x |>
+            lbdp_exact(lambda=lambda,mu=mu,psi=psi,n0=n0)
+        ) -> ll1
+      p |>
+        with(
+          x |>
+            lbdp_pomp(lambda=lambda,mu=mu,psi=psi,n0=n0) |>
+            pfilter(Np=Np) |>
+            logLik()
+        ) -> ll2
+      bind_cols(p,exact=ll1,pf=ll2)
+    }-> params
+  }
+) -> params
+
+params |>
+  mutate(
+    diff=(pf-exact)/exact
+  ) |>
+  group_by(Np) |>
+  summarize(
+    rmse=sqrt(mean(diff*diff)),
+    sd=sd(diff),
+    bias=abs(mean(pf-exact))
+  ) |>
+  ungroup() -> stats
+@
+
+\begin{figure}
+  \begin{center}
+    <<lbdp3_plot,fig.dim=c(8,5),out.width="100%">>=
+    pal <- c(viridis_pal(option="H",begin=0.1,end=0.8)(8),"#000000")
+    names(pal) <- c("1k","2k","4k","8k","16k","32k","64k","128k","exact")
+
+    plot_grid(
+      params |>
+        pivot_longer(c(exact,pf)) |>
+        unite(name,name,Np) |>
+        mutate(
+          name=if_else(grepl("exact",name),"exact",name),
+          name=gsub("pf_","",name),
+          name=gsub("000","k",name),
+          name=ordered(name,levels=names(pal))
+        ) |>
+        group_by(lambda,mu,psi,time,n0,name) |>
+        reframe(
+          type=c("logLik","logLik_se"),
+          value=logmeanexp(value,se=TRUE)
+        ) |>
+        ungroup() |>
+        pivot_wider(names_from=type) |>
+        mutate(
+          y=logLik,
+          ymax=logLik+2*logLik_se,
+          ymin=logLik-2*logLik_se
+        ) |>
+        filter(logLik>max(logLik)-16) |>
+        ggplot(aes(x=mu,group=name,color=name,
+          y=y,ymin=ymin,ymax=ymax))+
+        geom_errorbar(
+          position="dodge"
+        )+
+        geom_vline(xintercept=true.params$mu,linetype=2)+
+        geom_hline(
+          yintercept=max(params$exact)-
+            c(0,0.5*qchisq(p=0.95,df=1)),
+          linetype=2
+        )+
+        scale_color_manual(values=pal)+
+        labs(
+          color="effort",
+          y="log likelihood",
+          x=expression(mu)
+        )+
+        theme(
+          legend.position="inside",
+          legend.position.inside=c(0.5,0.4),
+          legend.background=element_rect(fill="white")
+        ),
+      plot_grid(
+        ##        x |> plot(points=TRUE),
+        stats |>
+          ggplot(aes(x=Np,y=rmse))+
+          geom_smooth(formula=y~x,method="lm")+
+          geom_point()+
+          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
+          scale_y_log10()+
+          coord_fixed(ratio=1)+
+          labs(x="effort",y="RMSE"),
+        stats |>
+          ggplot(aes(x=Np,y=sd))+
+          geom_smooth(formula=y~x,method="lm")+
+          geom_point()+
+          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
+          scale_y_log10()+
+          coord_fixed(ratio=1)+
+          labs(x="effort",y="SD"),
+        stats |>
+          ggplot(aes(x=Np,y=bias))+
+          geom_smooth(formula=y~x,method="lm")+
+          geom_point()+
+          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
+          scale_y_log10()+
+          coord_fixed(ratio=1)+
+          labs(x="effort",y="bias"),
+        ncol=1,
+        align="h",
+        axis="btlr",
+        rel_heights=c(1,1,1)
+      ),
+      nrow=1,
+      rel_widths=c(2,1)
+    )
+    @
+  \end{center}
+\end{figure}
+
 \subsection{Moran model and the Kingman coalescent}

### 2024-05-18 ms.Rnw
more work on examples

@@
 theme_set(theme_bw(base_family="serif"))
+options(
+  width=150,
+  keep.source=TRUE,
+  encoding="UTF-8",
+  stringsAsFactors=FALSE,
+  dplyr.summarise.inform=FALSE,
+  pomp_archive_dir="results"
+)
 set.seed(1159254136)
@@
 This is simply the probability density of a given genealogy conditional on a given model, viewed as a function of the parameters of that model.
-In particular, if $S$ a set of genome sequences, $\Phi$ a genealogical tree relating these sequences, $E$ a model of sequence evolution, and $D$ a dynamic transmission model, then the likelihood of $S$ is
-\begin{equation*}
-  f(S|D,E) = \int{f(S|\Phi,E)\,f(\Phi|D)\,\dd{\Phi}},
-\end{equation*}
-where the integral is taken over all possible genealogies and we somewhat loosely use the functions $f$ to refer to distinct probability densities, the nature of each of which is clear from its arguments.
-In this expression, $f(S|\Phi,E)$ is the \citet{Felsenstein2004} phylodynamic likelihood.
+In particular, if $S$ is a set of genome sequences, $\Phi$ a genealogical tree relating these sequences, $E$ a model of sequence evolution, and $D$ a dynamic transmission model, then the likelihood is
+\begin{equation*}
+  \lik(D,E) = f(S|D,E) = \int{f(S|\Phi,E)\,f(\Phi|D)\,\dd{\Phi}},
+\end{equation*}
+where the integral is taken over all possible genealogies and we somewhat loosely use the symbol $f$ for the various distinct probability densities, the nature of each of which is clear from its arguments.
+In this expression, $f(S|\Phi,E)$ is typically the \citet{Felsenstein2004} phylogenetic likelihood.
 The function $f(\Phi|D)$, which links the phylogeny to the dynamic model, may be termed the \emph{phylodynamic likelihood}.
@@
 We go on to derive expressions for the exact likelihoods of these genealogies.
+
+Code sufficient for the reproduction of all the results presented in this paper are freely available for download at \url{https://github.com/kingaa/phylopomp-project}.
+An archival version of these will be stored on Zotero upon publication of this paper.
+The open-source \pkg{R} package \pkg{phylopomp} (\url{https://github.com/kingaa/phylopomp}) implements the simulation and likelihood-computation algorithms employed here.
 
@@
 \setcounter{equation}{0}
+\setcounter{figure}{0}
+\setcounter{table}{0}
 \titleformat{\section}[hang]{\large\bfseries}{Appendix \periodafter\thesection}{2ex}{\periodafter}{}
 \renewcommand{\theequation}{\thesection\arabic{equation}}
+\renewcommand{\thefigure}{\thesection\arabic{figure}}
+\renewcommand{\thetable}{\thesection\arabic{table}}
 
@@
 
-\subsection{SIRS model}
+\subsection{SIRS model} %% SIRS
 
@@
 Note that, though the theory allows for time-dependent event rates, this example is time-homogenous.
-This model has one deme, $\Demes=\Set{\lab{I}}$, with occupancy function $n(x)=I$.
+This model has one deme and occupancy function $n(x)=I$.
 There are four kinds of jumps:
@@
 Accordingly, the marks are $\Jumps=\Set{\lab{Trans},\lab{Recov},\lab{Wane},\lab{Sample}}$.
-\Cref{tab:sirs_model_elements} gives $\alpha_u$, $r_u$, and the jump-type for each of these marks.
-
-\begin{table}
+\Cref{tab:sirs_model_elements} gives $\alpha_u$, $r_u$, and the event type for each of these marks.
+
+\begin{table}[h!]
   \caption{
@@
     \hline
-    $\lab{Trans}$  & ${\beta\,S\,I}/{N}$  & pure birth  & 2 \\
-    $\lab{Recov}$  & $\gamma\,I$          & pure death  & 0 \\
-    $\lab{Wane}$   & $\omega\,R$          & neutral     & 0 \\
-    $\lab{Sample}$ & $\psi\,I$            & pure sample & 1 \\
+    $\lab{Trans}$  & ${\beta S I}/{N}$ & pure birth  & 2 \\
+    $\lab{Recov}$  & $\gamma I$        & pure death  & 0 \\
+    $\lab{Wane}$   & $\omega R$        & neutral     & 0 \\
+    $\lab{Sample}$ & $\psi I$          & pure sample & 1 \\
     \hline\hline
@@
 <<sirs3>>=
+data.frame(
+  Beta=4,gamma=2,psi=1,omega=1,
+  S0=97,I0=3,R0=0,t0=0,time=40
+) -> sirs.params
+
 bake(
@@
   seed=328168304L,
+  dependson=sirs.params,
   {
     library(phylopomp)
+    sirs.params |>
+      with(
         runSIRS(
-      Beta=4,gamma=2,psi=1,omega=1,
-      S0=97,I0=3,R0=0,t0=0,time=40
+          Beta=Beta,gamma=gamma,psi=psi,omega=omega,
+          S0=S0,I0=I0,R0=R0,t0=t0,time=time
+        )
       )
   }
-) -> x
+) -> sirs_tree
 
@@
   seed=621400057L,
+  dependson=list(sirs_tree,sirs.params),
   {
+    sirs.params |>
+      with(
         expand_grid(
-      Beta=4,
-      ##      gamma=2,
+          Beta=Beta,
+          ##      gamma=gamma,
           gamma=seq(1.7,2.3,by=0.02),
-      psi=1,
-      omega=1,
+          psi=psi,
+          omega=omega,
           ##      omega=seq(0.5,2,by=0.05),
-      S0=97,I0=3,R0=0,
-      t0=0,
+          S0=S0,I0=I0,R0=R0,
+          t0=t0,
           rep=1:16,
           Np=5000,
+          )
       ) -> params
@@
         with({
-          x |>
+          sirs_tree |>
             sirs_pomp(
@@
     plot_grid(
-      A=x |>
+      A=sirs_tree |>
         plot(points=FALSE,palette=c("#000000"))+
@@
         geom_errorbar(position="dodge")+
-        geom_vline(xintercept=2,color="red")+
+        geom_vline(xintercept=sirs.params$gamma,color="red")+
         labs(
@@
   \end{center}
+  \caption{
+    \label{fig:sirs_example}
+    Likelihood computation for the SIRS model by sequential Monte Carlo.
+    (A)~A~simulated genealogy for $\beta=\Sexpr{sirs.params$Beta}$, $\gamma=\Sexpr{sirs.params$gamma}$, $\omega=\Sexpr{sirs.params$omega}$, $\psi=\Sexpr{sirs.params$psi}$, $(S_0,I_0,R_0)=(\Sexpr{with(sirs.params,c(S0,I0,R0))})$.
+    (B)~A~slice through the likelihood surface at the true parameters in the $\gamma$-direction.
+    Error bars show $\pm{2}$ s.d.\ of the Monte Carlo estimates.
+  }
 \end{figure}
 
-\subsection{SEIRS model}
+\subsection{SEIRS model} %% SEIRS
 
@@
 where $N=S+E+I+R$ is the total population size.
-This model has two demes: $\Demes=\Set{\lab{E},\lab{I}}$ and four
-The deme occupancy function in this case is $n(x)=(E,I)$.
 Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.
 
+This model has two demes: $\Demes=\Set{\lab{E},\lab{I}}$;
+its deme occupancy function is $n(x)=(E,I)$.
+There are five kinds of jumps:
+transmission, progression, recovery, waning of immunity, and sampling.
+The corresponding marks are $\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$.
+\Cref{tab:seirs_model_elements} gives $\alpha_u$, $r_u$, and the event type for each of these marks.
+
+\begin{table}[h!]
+  \caption{
+    \label{tab:seirs_model_elements}
+    Elements of the SEIRS model pertinent to the genealogy process.
+  }
+  \renewcommand{\arraystretch}{1.2}
+  \begin{tabular}{cclc}
+    \hline\hline
+    $u$ & $\alpha_u$ & Event type & $r_u$ \\
+    \hline
+    $\lab{Trans}$  & ${\beta S I}/{N}$ & pure birth     & $(1,1)$ \\
+    $\lab{Prog}$   & $\sigma E$        & pure migration & $(0,1)$ \\
+    $\lab{Recov}$  & $\gamma I$        & pure death     & $(0,0)$ \\
+    $\lab{Wane}$   & $\omega R$        & neutral        & $(0,0)$ \\
+    $\lab{Sample}$ & $\psi I$          & pure sample    & $(0,1)$ \\
+    \hline\hline
+  \end{tabular}
+\end{table}
+
+<<seirs3>>=
+seirs.params <- data.frame(
+  Beta=4,sigma=1,gamma=1,psi=1,omega=1,
+  S0=200,E0=3,I0=5,R0=100,
+  time=3
+)
+
+bake(
+  file="seirs3a.rds",
+  seed=831282841L,
+  dependson=seirs.params,
+  seirs.params |>
+    with(
+      runSEIR(
+        Beta=Beta,sigma=sigma,gamma=gamma,psi=psi,omega=omega,
+        S0=S0,E0=E0,I0=I0,R0=R0,
+        time=time
+      )
+    )
+) -> seirs_tree
+
+bake(
+  file="seirs3b.rds",
+  seed=831282841L,
+  dependson=list(seirs.params,seirs_tree),
+  {
+    library(phylopomp)
+    library(circumstance)
+    library(doFuture)
+    plan(multicore)
+
+    seirs.params |>
+      with(
+        seirs_tree |>
+          seirs_pomp(
+            Beta=Beta,sigma=sigma,gamma=gamma,psi=psi,omega=omega,
+            S0=S0,E0=E0,I0=I0,R0=R0
+          )
+      ) -> po
+
+    seq(0.2,2.5,by=0.1) |>
+      lapply(
+        \(s) {
+          x <- po
+          coef(x,"sigma") <- s
+          x
+        }
+      ) |>
+      concat() |>
+      pfilter(Np=20000,Nrep=10)
+  }
+) -> pfs
+
+left_join(
+  pfs |> coef() |> melt() |> pivot_wider(),
+  pfs |> logLik() |> melt() |> rename(logLik=value),
+  by=c(".id"="name")
+) -> params
+
+params |>
+  group_by(sigma) |>
+  summarize(
+    sd=sd(logLik),
+    logLik=logmeanexp(logLik),
+    hi=logLik+2*sd,
+    lo=logLik-2*sd
+  ) |>
+  ungroup() -> stats
+@
+
+\begin{figure}
+  \begin{center}
+    <<seirs3_plot,fig.dim=c(8,2.8),out.width="100%">>=
+    plot_grid(
+      A=seirs_tree |>
+        plot(points=TRUE,palette="#000000")+
+        labs(x="time"),
+      B=params |>
+        filter(logLik>max(logLik-16)) |>
+        ggplot(aes(x=sigma,y=logLik))+
+        geom_point()+
+        geom_smooth()+
+        geom_vline(xintercept=seirs.params$sigma,color="red")+
+        geom_hline(
+          yintercept=max(stats$logLik)-
+            c(0,0.5*qchisq(p=0.95,df=1)),
+          linetype=2
+        )+
+        labs(
+          color=character(0),
+          y="log likelihood",
+          x=expression(sigma)
+        )+
+        theme_classic(),
+      labels="AUTO",
+      nrow=1,
+      rel_widths=c(3,4)
+    )
+    @
+  \end{center}
+  \caption{
+    \label{fig:seirs_example}
+    Likelihood computation for the SEIRS model by sequential Monte Carlo.
+    (A)~Simulated genealogy for $\beta=\Sexpr{seirs.params$Beta}$,
+    $\sigma=\Sexpr{seirs.params$sigma}$,
+    $\gamma=\Sexpr{seirs.params$gamma}$,
+    $\psi=\Sexpr{seirs.params$psi}$,
+    $\omega=\Sexpr{seirs.params$omega}$,
+    $(S_0,E_0,I_0,R_0)=(\Sexpr{with(seirs.params,c(S0,E0,I0,R0))})$.
+    (B)~Likelihood slice in the $\sigma$-direction.
+    Each point represents the estimate of a distinct sequential Monte Carlo computation.
+  }
+\end{figure}
+
 \subsection{Two-strain competition model}
@@
 A simple model for the competition of two strains for susceptible hosts is depicted in \cref{fig:example_models}B.
-This example will be completed in a forthcoming draft.
+This example will be included in a forthcoming draft.
 
@@
 \Cref{fig:example_models}D depicts a model of superspreading.
-This example will be completed in a forthcoming draft.
-
-\subsection{Linear birth-death model}
+This example will be included in a forthcoming draft.
+
+\subsection{Linear birth-death model} %% LBDP
 
@@
 
+\begin{table}[h!]
+  \caption{
+    \label{tab:lbdp_model_elements}
+    Elements of the linear birth-death-sampling model pertinent to the genealogy process.
+  }
+  \renewcommand{\arraystretch}{1.2}
+  \begin{tabular}{cclc}
+    \hline\hline
+    $u$ & $\alpha_u$ & Event type & $r_u$ \\
+    \hline
+    $\lab{Birth}$  & $\lambda N$  & pure birth  & 2 \\
+    $\lab{Death}$  & $\gamma N$   & pure death  & 0 \\
+    $\lab{Sample}$ & $\psi N$     & pure sample & 1 \\
+    \hline\hline
+  \end{tabular}
+\end{table}
+
 <<lbdp3>>=
-true.params <- data.frame(lambda=1.2,mu=0.8,psi=1,n0=5)
+lbdp.params <- data.frame(
+  lambda=1.2,mu=0.8,psi=1,n0=5,
+  time=10
+)
 
@@
   file="lbdp3a.rds",
-  seed=505641575,  
-  true.params |>
+  seed=915645370,
+  dependson=lbdp.params,
+  lbdp.params |>
     with(
-      runLBDP(lambda=lambda,mu=mu,psi=psi,n0=n0,time=5)
-    ) -> x
-) -> x
+      runLBDP(lambda=lambda,mu=mu,psi=psi,n0=n0,time=time)
+    )
+) -> lbdp_tree
 
@@
   file="lbdp3b.rds",
-  dependson=x,
-  seed=715604304,
+  seed=712604404,
+  dependson=list(lbdp_tree,lbdp.params),
   {
+    lbdp.params |>
+      with(
         expand_grid(
           rep=1:10,
-      lambda=1.2,
-      mu=seq(0.3,2,by=0.1),
-      ## mu=0.8,
-      ## psi=seq(0.2,2.5,by=0.1),
-      psi=1,
-      time=3,
-      n0=5,
-      Np=c(1000,2000,4000,8000,16000,32000,64000,128000)
+          lambda=lambda,
+          mu=seq(0.5,1.1,by=0.05),
+          psi=psi,
+          n0=n0,
+          Np=1000*2^seq(0,7)
+        )
       ) -> params
@@
         with(
-          x |>
+          lbdp_tree |>
             lbdp_exact(lambda=lambda,mu=mu,psi=psi,n0=n0)
@@
         with(
-          x |>
+          lbdp_tree |>
             lbdp_pomp(lambda=lambda,mu=mu,psi=psi,n0=n0) |>
@@
   mutate(
-    diff=(pf-exact)/exact
+    diff=pf-exact
   ) |>
@@
     rmse=sqrt(mean(diff*diff)),
-    sd=sd(diff),
-    bias=abs(mean(pf-exact))
+    bias=abs(mean(diff)),
+    prec=sqrt(rmse^2-bias^2)
   ) |>
@@
   \begin{center}
-    <<lbdp3_plot,fig.dim=c(8,5),out.width="100%">>=
+    <<lbdp3_plot,fig.dim=c(9,7),out.width="100%">>=
     pal <- c(viridis_pal(option="H",begin=0.1,end=0.8)(8),"#000000")
@@
     plot_grid(
-      params |>
+      ncol=1,
+      rel_heights=c(3,2),
+      AB=plot_grid(
+        labels=c("A","B"),
+        nrow=1,
+        rel_widths=c(3,5),
+        A=lbdp_tree |>
+          plot(points=FALSE,palette="#000000")+
+          labs(x="time"),
+        B=params |>
           pivot_longer(c(exact,pf)) |>
@@
           ) |>
-        group_by(lambda,mu,psi,time,n0,name) |>
+          group_by(lambda,mu,psi,n0,name) |>
           reframe(
@@
           filter(logLik>max(logLik)-16) |>
-        ggplot(aes(x=mu,group=name,color=name,
-          y=y,ymin=ymin,ymax=ymax))+
+          ggplot(
+            aes(
+              x=mu,group=name,color=name,
+              y=y,ymin=ymin,ymax=ymax
+            )
+          )+
           geom_errorbar(
@@
           )+
-        geom_vline(xintercept=true.params$mu,linetype=2)+
+          geom_vline(xintercept=lbdp.params$mu,color="red")+
           geom_hline(
@@
           theme(
-          legend.position="inside",
-          legend.position.inside=c(0.5,0.4),
-          legend.background=element_rect(fill="white")
+            ## legend.position="inside",
+            ## legend.position.inside=c(0.5,0.4),
+            ## legend.background=element_rect(fill="white")
+          )
       ),
-      plot_grid(
-        ##        x |> plot(points=TRUE),
-        stats |>
+      CDE=plot_grid(
+        labels=c("C","D","E"),
+        nrow=1,
+        rel_widths=c(24,24,11),
+        C=stats |>
           ggplot(aes(x=Np,y=rmse))+
@@
           labs(x="effort",y="RMSE"),
-        stats |>
-          ggplot(aes(x=Np,y=sd))+
+        D=stats |>
+          ggplot(aes(x=Np,y=prec))+
           geom_smooth(formula=y~x,method="lm")+
@@
           labs(x="effort",y="SD"),
-        stats |>
+        E=stats |>
           ggplot(aes(x=Np,y=bias))+
@@
           scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
-          scale_y_log10()+
+          scale_y_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
           coord_fixed(ratio=1)+
-          labs(x="effort",y="bias"),
-        ncol=1,
-        align="h",
-        axis="btlr",
-        rel_heights=c(1,1,1)
-      ),
-      nrow=1,
-      rel_widths=c(2,1)
+          labs(x="effort",y=expression(group("|",bias,"|")))
+      )
     )
@@
   \end{center}
+  \caption{
+    \label{fig:lbdp_example}
+    Likelihood computation for the constant-parameter, linear birth-death-sampling model, according to \cref{thm:obsc_uncond} via sequential Monte Carlo.
+    Panel (A) shows the genealogy, simulated for $\lambda=\Sexpr{lbdp.params$lambda}$, $\mu=\Sexpr{lbdp.params$mu}$, $\psi=\Sexpr{lbdp.params$psi}$, $N_0=\Sexpr{lbdp.params$mu}$.
+    Panel (B) shows a likelihood slice, through the true parameters in the $\mu$ direction.
+    As computational effort (\ie number of particles) increases, the Monte Carlo estimates converge on the exact values, for which an explicit formula exists in this case.
+    The dashed horizontal lines show the approximate maximized likelihood and the 95\% critical value (under the likelihood-ratio test).
+    Panels (C--E) are log-log plots that show how the root-mean-square error (RMSE), imprecision (SD), and bias decrease with effort.
+    Note that the bias and the SD are roughly inversely proportional to the effort and its square-root, respectively, as expected.
+  }
 \end{figure}

### 2024-05-19 ms.Rnw
use MCAP

@@
 In particular, let $\leaves$ be a countable set of labels and let $\part(\leaves)$ be the set of all collections of finite, mutually-disjoint subsets of $\leaves$.
-That is, an element $\zeta\in{\part(\leaves)}$ is a partition of the finite set $\bigcup\zeta\subseteq\leaves$.
+That is, an element $z\in{\part(\leaves)}$ is a partition of the finite set $\bigcup{z}\subseteq\leaves$.
 Partition \emph{fineness} defines a partial order on $\part(\leaves)$.
-Specifically, for $\zeta,\zeta'\in{\part(\leaves)}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
+Specifically, for $z,z'\in{\part(\leaves)}$, we say $z\preceq{z'}$ if and only if for every $b'\in{z'}$ there is $b\in{z}$ such that $b\supseteq{b'}$.
 The tree structure of $G$ is defined by a \cadlag\ map $Z:[0,T]\to\part(\leaves)$ that is monotone in the sense that $t_1\le{t_2}$ implies $Z_{t_1}\preceq{Z_{t_2}}$.
@@
 To succinctly accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $P^{\lab{Y}}$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
-That is, $Q_u(\eta,\eta')=1$ if and only if
+That is, $Q_u(y,y')=1$ if and only if
 there is a feasible genealogy, $\G=(\T,\Z,\Y)$, and history, $\H$,
@@
 given $\Gr_\T=\G$ and $\Hr_\T=\H$,
-we have $\U_t=u$, $\Yt_t=\eta$, and $\Y_t=\eta'$.
+we have $\U_t=u$, $\Yt_t=y$, and $\Y_t=y'$.
 We refer to $Q$ as the \emph{compatibility indicator}.
@@
 %% \begin{equation*}
-%%   \phi_u(\xi,\eta,\eta')\colonequals\BinRatio{n(\xi)}{\ell(\eta')}{r^u}{s(\eta,\eta')}\,Q_u(\eta,\eta').
+%%   \phi_u(\xi,y,y')\colonequals\BinRatio{n(\xi)}{\ell(y')}{r^u}{s(y,y')}\,Q_u(y,y').
 %% \end{equation*}
@@
 ) -> params
+
+params |>
+  with(
+    mcap(logLik,gamma)
+  ) -> mcap
 @
@@
       B=params |>
-        group_by(Beta,gamma,psi,omega,S0,I0,R0,Np) |>
-        reframe(
-          name=c("logLik","logLik_se"),
-          value=logmeanexp(logLik,se=TRUE)
-        ) |>
-        ungroup() |>
-        pivot_wider() |>
-        mutate(
-          y=logLik,
-          ymax=logLik+2*logLik_se,
-          ymin=logLik-2*logLik_se
-        ) |>
-        filter(round(Beta,2)==4) |>
-        filter(logLik>max(logLik)-16) |>
-        ggplot(
-          aes(
-            x=gamma,
-            group=Np,
-            y=y,
-            ymin=ymin,
-            ymax=ymax
-          )
+        ggplot(aes(x=gamma,y=logLik))+
+        geom_point(alpha=0.2)+
+        geom_line(data=mcap$fit,aes(x=parameter,y=smoothed),color="blue")+
+        geom_vline(xintercept=sirs.params$gamma,color="red")+
+        geom_vline(xintercept=mcap$ci,linetype=2)+
+        geom_hline(
+          yintercept=max(mcap$fit$smoothed)-c(0,mcap$delta),
+          linetype=2
         )+
-        geom_errorbar(position="dodge")+
-        geom_vline(xintercept=sirs.params$gamma,color="red")+
         labs(
@@
         )+
+        ## lims(y=c(max(params$logLik)-16,NA))+
         theme_classic()+
@@
     (B)~A~slice through the likelihood surface at the true parameters in the $\gamma$-direction.
-    Error bars show $\pm{2}$ s.d.\ of the Monte Carlo estimates.
+    Each point is a distinct Monte Carlo estimate.
+    The blue curve is a LOESS smooth; the dashed lines show Monte Carlo-adjusted 95\% confidence intervals \citep{Ionides2017}.
   }
@@
 
-This model has two demes: $\Demes=\Set{\lab{E},\lab{I}}$;
-its deme occupancy function is $n(x)=(E,I)$.
+This model has two demes: $\Demes=\Set{\lab{E},\lab{I}}$.
+Its deme occupancy function is $n(x)=(E,I)$.
 There are five kinds of jumps:
@@
 params |>
-  group_by(sigma) |>
-  summarize(
-    sd=sd(logLik),
-    logLik=logmeanexp(logLik),
-    hi=logLik+2*sd,
-    lo=logLik-2*sd
-  ) |>
-  ungroup() -> stats
+  with(
+    mcap(logLik,sigma,span=0.25)
+  ) -> mcap
 @
@@
       B=params |>
-        filter(logLik>max(logLik-16)) |>
-        ggplot(aes(x=sigma,y=logLik))+
-        geom_point()+
-        geom_smooth()+
+        ggplot()+
+        geom_point(aes(x=sigma,y=logLik))+
+        geom_line(data=mcap$fit,aes(x=parameter,y=smoothed),color="blue")+
         geom_vline(xintercept=seirs.params$sigma,color="red")+
+        geom_vline(xintercept=mcap$ci,linetype=2)+
         geom_hline(
-          yintercept=max(stats$logLik)-
-            c(0,0.5*qchisq(p=0.95,df=1)),
+          yintercept=max(mcap$fit$smoothed)-c(0,mcap$delta),
           linetype=2
         )+
+        ##        lims(y=c(max(params$logLik)-12,NA))+
         labs(
@@
     Each point represents the estimate of a distinct sequential Monte Carlo computation.
+    The blue curve shows a LOESS smooth;
+    the dashed vertical lines enclose the Monte Carlo-adjusted 95\% confidence interval \citep{Ionides2017}.
   }

### 2024-05-19 figs/seirs_events.tex
SEIRS table

\def\ysp{0.3}

\def\unbroken{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-0.5ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \draw[deme1] (0,\ysp) -- (2,\ysp);
      \draw[deme2] (0,0) -- (2,0);
    \end{tikzpicture}%
  }%
}

\def\maizemaize{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-0.5ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \tikzstyle{event}=[text=darkgreen, font=\Huge]
      \draw[deme2] (0,\ysp) -- (2,\ysp);
      \draw[deme1] (0,0) -- (2,0);
      \node[event] at (1,\ysp) {$\bullet$};
    \end{tikzpicture}%
  }%
}

\def\maizeblue{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-0.5ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \tikzstyle{event}=[text=darkgreen, font=\Huge]
      \draw[deme2] (0,\ysp) -- (1,\ysp);
      \draw[deme1] (1,\ysp) -- (2,\ysp);
      \draw[deme1] (0,0) -- (2,0);
      \node[event] at (1,\ysp) {$\bullet$};
    \end{tikzpicture}%
  }%
}

\def\bluemaize{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-0.5ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \tikzstyle{event}=[text=darkgreen, font=\Huge]
      \draw[deme1] (0,\ysp) -- (1,\ysp);
      \draw[deme2] (1,\ysp) -- (2,\ysp);
      \draw[deme2] (0,0) -- (2,0);
      \node[event] at (1,\ysp) {$\bullet$};
    \end{tikzpicture}%
  }%
}

\def\samplezero{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-0.5ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \tikzstyle{event}=[text=darkgreen, font=\Huge]
      \tikzstyle{sample}=[text=royalblue, font=\Huge]
      \draw[deme1] (0,\ysp) -- (2,\ysp);
      \draw[deme2] (0,0) -- (1,0);
      \node[sample] at (1,0) {$\bullet$};
    \end{tikzpicture}%
  }%
}

\def\samplezeronix{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-0.5ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \tikzstyle{event}=[text=darkgreen, font=\Huge]
      \tikzstyle{sample}=[text=royalblue, font=\Huge]
      \draw[deme1] (0,\ysp) -- (1,\ysp);
      \draw[deme2] (0,0) -- (2,0);
      \node[sample] at (1,\ysp) {$\bullet$};
    \end{tikzpicture}%
  }%
}

\def\sampleone{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-0.5ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \tikzstyle{event}=[text=darkgreen, font=\Huge]
      \tikzstyle{sample}=[text=royalblue, font=\Huge]
      \draw[deme1] (0,\ysp) -- (2,\ysp);
      \draw[deme2] (0,0) -- (2,0);
      \node[sample] at (1,0) {$\bullet$};
    \end{tikzpicture}%
  }%
}

\def\sampleonenix{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-0.5ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \tikzstyle{event}=[text=darkgreen, font=\Huge]
      \tikzstyle{sample}=[text=royalblue, font=\Huge]
      \draw[deme1] (0,\ysp) -- (2,\ysp);
      \draw[deme2] (0,0) -- (2,0);
      \node[sample] at (1,\ysp) {$\bullet$};
    \end{tikzpicture}%
  }%
}

\def\branchup{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-1ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \tikzstyle{event}=[text=darkgreen, font=\Huge]
      \tikzstyle{sample}=[text=royalblue, font=\Huge]
      \draw[deme2] (0,0) -- (1,0);
      \draw[deme1] (1,0) -- (1,\ysp) -- (2,\ysp);
      \draw[deme2] (1,0) -- (1,-\ysp) -- (2,-\ysp);
      \node[event] at (1,0) {$\bullet$};
    \end{tikzpicture}%
  }%
}

\def\branchupnix{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-1ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \tikzstyle{event}=[text=darkgreen, font=\Huge]
      \tikzstyle{sample}=[text=royalblue, font=\Huge]
      \draw[deme1] (0,0) -- (1,0);
      \draw[deme1] (1,0) -- (1,\ysp) -- (2,\ysp);
      \draw[deme2] (1,0) -- (1,-\ysp) -- (2,-\ysp);
      \node[event] at (1,0) {$\bullet$};
    \end{tikzpicture}%
  }%
}

\def\branchdown{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-1ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \tikzstyle{event}=[text=darkgreen, font=\Huge]
      \tikzstyle{sample}=[text=royalblue, font=\Huge]
      \draw[deme2] (0,0) -- (1,0);
      \draw[deme2] (1,0) -- (1,\ysp) -- (2,\ysp);
      \draw[deme1] (1,0) -- (1,-\ysp) -- (2,-\ysp);
      \node[event] at (1,0) {$\bullet$};
    \end{tikzpicture}%
  }%
}

\def\branchdownnix{%
  \resizebox{3em}{!}{%
    \begin{tikzpicture}[baseline=-1ex]
      \tikzstyle{deme1}=[color=darkblue, line width=2pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=2pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=2pt]
      \tikzstyle{deme2H}=[color=maize, line width=2pt, densely dotted]
      \tikzstyle{event}=[text=darkgreen, font=\Huge]
      \tikzstyle{sample}=[text=royalblue, font=\Huge]
      \draw[deme1] (0,0) -- (1,0);
      \draw[deme2] (1,0) -- (1,\ysp) -- (2,\ysp);
      \draw[deme1] (1,0) -- (1,-\ysp) -- (2,-\ysp);
      \node[event] at (1,0) {$\bullet$};
    \end{tikzpicture}%
  }%
}

### 2024-05-19 ms.Rnw
SEIRS table

@@
 
-%% \section{Examples}
-
-%% \subsection{SIRS}
-
-%% Was shown in an earlier paper.
-%% [Display filter equation and plot a curve.]
-
-%% \subsection{SEIRS}
-
-%% Display filter equation.
-%% Discuss the choice of $\pi$ and $q$.
-%% Present a likelihood curve.
-
-%% \subsection{Two-strain competition}
-
-%% Display filter equation.
-%% Discuss the choice of $\pi$ and $q$.
-%% Present a likelihood curve.
-
-%% \subsection{Superspreading model}
-
-%% Display filter equation.
-%% Discuss the choice of $\pi$ and $q$.
-%% Present a likelihood curve.
-
-%% \subsection{Linear birth-death and Moran models}
-
-%% Exact solution of the filtering equations is possible in the unstructured case, yielding precisely the likelihoods for these two models.
-%% This establishes that the present theory is a strict generalization of these cases.
-%% It does away with the need for large sample-size and small sample-fraction approximations, the necessity of assuming slow changes in effective population size, and the need for linearization.
-
 \section{Discussion}
@@
   }
-  \renewcommand{\arraystretch}{1.2}
+  %% \renewcommand{\arraystretch}{1.2}
   \begin{tabular}{cclc}
@@
     \hline
-    $\lab{Trans}$  & ${\beta S I}/{N}$ & pure birth  & 2 \\
-    $\lab{Recov}$  & $\gamma I$        & pure death  & 0 \\
-    $\lab{Wane}$   & $\omega R$        & neutral     & 0 \\
-    $\lab{Sample}$ & $\psi I$          & pure sample & 1 \\
+    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & pure birth  & 2 \\
+    $\lab{Recov}$  & $\gamma I$        \bigstrut & pure death  & 0 \\
+    $\lab{Wane}$   & $\omega R$        \bigstrut & neutral     & 0 \\
+    $\lab{Sample}$ & $\psi I$          \bigstrut & pure sample & 1 \\
     \hline\hline
@@
       \frac{\partial{v}}{\partial{t}}(t,S,E,I,R)=
-      &\frac{\beta(t)\,(S+1)\,I}{N}\,v(t,S+1,E-1,I,R)
-      -\frac{\beta(t)\,S\,I}{N}\,v(t,S,E,I,R)\\
+      &\frac{\beta\,(S+1)\,I}{N}\,v(t,S+1,E-1,I,R)
+      -\frac{\beta\,S\,I}{N}\,v(t,S,E,I,R)\\
       &+\sigma\,(E+1)\,v(t,S,E+1,I-1,R)
@@
 
+\input{figs/seirs_events}
+
 \begin{table}[h!]
@@
     \hline
-    $\lab{Trans}$  & ${\beta S I}/{N}$ & pure birth     & $(1,1)$ \\
-    $\lab{Prog}$   & $\sigma E$        & pure migration & $(0,1)$ \\
-    $\lab{Recov}$  & $\gamma I$        & pure death     & $(0,0)$ \\
-    $\lab{Wane}$   & $\omega R$        & neutral        & $(0,0)$ \\
-    $\lab{Sample}$ & $\psi I$          & pure sample    & $(0,1)$ \\
+    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & pure birth     & $(1,1)$ \\
+    $\lab{Prog}$   & $\sigma E$        \bigstrut & pure migration & $(0,1)$ \\
+    $\lab{Recov}$  & $\gamma I$        \bigstrut & pure death     & $(0,0)$ \\
+    $\lab{Wane}$   & $\omega R$        \bigstrut & neutral        & $(0,0)$ \\
+    $\lab{Sample}$ & $\psi I$          \bigstrut & pure sample    & $(0,1)$ \\
     \hline\hline
@@
 \end{table}
+
+\begin{table}[h!]
+  \caption{
+    \label{tab:seirs_integ}
+    Elements of a scheme for computing the likelihood under the SEIRS model.
+    The $y\mapsto{y'}$ column depicts the proposed painting schematically.
+    Blue is used for the $\lab{E}$ deme and yellow for the $\lab{I}$ deme.
+  }
+  %% \renewcommand{\arraystretch}{1.2}
+  \begin{tabular}{c c c c c c c | c}
+    \hline\hline
+    & $u$ & $s$ & $\BinRatio{n}{\ell}{r}{s}$ & $y\mapsto{y'}$ & $\pi$ & $Q$ & Line\\[2ex]
+    \hline
+    \multirow[c]{10}{*}[-1ex]{\rotatebox{90}{$t\notin{{B}\cup{S_0}\cup{S_1}}$}}
+    & \multirow{4}{*}{$\lab{Trans}$}
+    & $(0,0)$ & $\left(\frac{E-\ell_E}{E}\right)\left(\frac{I-\ell_I}{I}\right)$ \bigstrut & \unbroken \bigstrut & $\frac{I-\ell_I}{I}$ & $1$ & 1 \\
+    \cline{3-8}
+    && $(1,0)$ & $\frac{I-\ell_I}{E I}$ \bigstrut & \maizeblue \bigstrut & $\frac{1}{2 I}$ & $1$ & 2 \\
+    \cline{3-8}
+    && $(0,1)$ & $\frac{E-\ell_E}{E I}$ \bigstrut & \maizemaize \bigstrut & $\frac{1}{2 I}$ & $1$ & 3 \\
+    \cline{3-8}
+    && $(1,1)$ & --- & --- & 0 & $0$ & 4 \\
+    \cline{2-8}
+    &\multirow{2}{*}{$\lab{Prog}$}
+    & $(0,0)$ & $\frac{I-\ell_I}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \unbroken \bigstrut & $\frac{E-\ell_E}{E}\,\Indicator{E>\ell_E}$ \bigstrut & $1$ & 5 \\
+    \cline{3-8}
+    && $(0,1)$ & $\frac{1}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \bluemaize \bigstrut & $\frac{1}{E}\,\Indicator{E>\ell_E}$ \bigstrut & $1$ & 6 \\
+    \cline{2-8}
+    & $\lab{Recov}$ & $(0,0)$ & $\Indicator{I\ge\ell_I}$ \bigstrut & \unbroken \bigstrut & $\Indicator{I>\ell_I}$ \bigstrut & $1$ & 7 \\
+    \cline{2-8}
+    & $\lab{Wane}$ & $(0,0)$ & $1$ \bigstrut & \unbroken \bigstrut & $1$ & $1$ & 8 \\
+    \cline{2-8}
+    & $\lab{Sample}$ & $\ast$ & --- & --- & 0 & 0 & 9 \\
+    \hline
+    \multirow[c]{4}{*}[-3ex]{\rotatebox{90}{$t\in{B}$}}
+    & \multirow{4}{*}{$\lab{Trans}$}
+    & \multirow{4}{*}{$(1,1)$}
+    & $\frac{1}{E I}$ \bigstrut & \branchup \bigstrut & $\frac{1}{2}$ \bigstrut & $1$ & 10 \\
+    \cline{4-8}
+    &&& $\frac{1}{E I}$ \bigstrut & \branchdown \bigstrut & $\frac{1}{2}$ \bigstrut & $1$ & 11 \\
+    \cline{4-8}
+    &&& --- \bigstrut & \branchupnix \bigstrut & 0 & $0$ & 12 \\
+    \cline{4-8}
+    &&& --- \bigstrut & \branchdownnix \bigstrut & 0 & $0$ & 13 \\
+    \cline{2-8}
+    & $\ast$ & $\ast$ & --- \bigstrut & --- & 0 & $0$ & 14 \\
+    \hline
+    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_0}$} \bigstrut}
+    & \multirow{2}{*}{$\lab{Sample}$}
+    & \multirow{2}{*}{$(0,0)$}
+    & $\frac{I-\ell_I}{I}$ \bigstrut & \samplezero & $1$ & $1$ & 15 \\
+    \cline{4-8}
+    &&& --- & \samplezeronix \bigstrut & 0 & 0 & 16 \\
+    \cline{2-8}
+    & $\ast$ & $\ast$ & --- & --- & 0 & $0$ & 19 \\
+    \hline
+    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_1}$}}
+    & \multirow{2}{*}{$\lab{Sample}$}
+    & \multirow{2}{*}{$(0,1)$}
+    & $\frac{1}{I}$ \bigstrut & \sampleone \bigstrut & $1$ & $1$ & 17 \\
+    \cline{4-8}
+    &&& --- & \sampleonenix \bigstrut & 0 & 0 & 18 \\
+    \cline{2-8}
+    & $\ast$ & $\ast$ & --- & --- & 0 & $0$ & 19 \\
+    \hline\hline
+  \end{tabular}
+\end{table}
+
+A variety of importance-sampling kernels are permissible under the terms of \cref{thm:obsc_uncond}.
+Given a particular choice of importance-sampling kernel, the filtering equation follows.
+For example, with the choices as in \cref{tab:seirs_integ}, the filtering equation is as follows.
+As previously, let $B$ be the set of branch times, $S_0$ be the set of tip-sample times, and $S_1$ be the set of inline sample times.
+Then for $t\notin{{B}\cup{S_0}\cup{S_1}}$, we have
+\begin{mathsize}{9pt}{10pt}
+  \begin{equation*}
+    \begin{split}
+      \frac{\partial{w}}{\partial{t}}(t,S,E,I,R)=
+      &\frac{\beta\,(S+1)\,I}{N}\,w(t,S+1,E-1,I,R)
+      -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R)\\
+      &+\sigma\,(E+1)\,w(t,S,E+1,I-1,R)
+      -\sigma\,E\,w(t,S,E,I,R)\\
+      &+\gamma\,(I+1)\,w(t,S,E,I+1,R-1)
+      -\gamma\,I\,w(t,S,E,I,R)\\
+      &+\omega\,(R+1)\,w(t,S-1,E,I,R+1)
+      -\omega\,R\,w(t,S,E,I,R),
+    \end{split}
+  \end{equation*}
+\end{mathsize}%
 

### 2024-05-21 ms.Rnw
new phylopomp version

--- 2024-05-19 ms.Rnw
@@
 stopifnot(getRversion() >= "4.3")
-stopifnot(packageVersion("pomp")>="5.5")
-stopifnot(packageVersion("phylopomp")>="0.10.3")
+stopifnot(packageVersion("pomp")>="5.8")
+stopifnot(packageVersion("phylopomp")>="0.11")
 theme_set(theme_bw(base_family="serif"))
@@
   }
-  \renewcommand{\arraystretch}{1.2}
-  \begin{tabular}{cclc}
+  %% \renewcommand{\arraystretch}{1.2}
+  \begin{tabular}{cccl}
     \hline\hline
-    $u$ & $\alpha_u$ & Event type & $r_u$ \\
+    $u$ & $\alpha_u$ & $r_u$ & Event type \\
     \hline
-    $\lab{Trans}$  & ${\beta S I}/{N}$ & pure birth  & 2 \\
-    $\lab{Recov}$  & $\gamma I$        & pure death  & 0 \\
-    $\lab{Wane}$   & $\omega R$        & neutral     & 0 \\
-    $\lab{Sample}$ & $\psi I$          & pure sample & 1 \\
+    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & 2 & pure birth \\
+    $\lab{Recov}$  & $\gamma I$ \bigstrut & 0 & pure death \\
+    $\lab{Wane}$   & $\omega R$ \bigstrut & 0 & neutral \\
+    $\lab{Sample}$ & $\psi I$ \bigstrut & 1 & pure sample \\
     \hline\hline
@@
   }
-  \renewcommand{\arraystretch}{1.2}
-  \begin{tabular}{cclc}
+  %% \renewcommand{\arraystretch}{1.2}
+  \begin{tabular}{ccccl}
     \hline\hline
-    $u$ & $\alpha_u$ & Event type & $r_u$ \\
+    $u$ & $\alpha_u$ & $x\mapsto{x'}$ & $r_u$ & Event type \\
     \hline
-    $\lab{Trans}$  & ${\beta S I}/{N}$ & pure birth     & $(1,1)$ \\
-    $\lab{Prog}$   & $\sigma E$        & pure migration & $(0,1)$ \\
-    $\lab{Recov}$  & $\gamma I$        & pure death     & $(0,0)$ \\
-    $\lab{Wane}$   & $\omega R$        & neutral        & $(0,0)$ \\
-    $\lab{Sample}$ & $\psi I$          & pure sample    & $(0,1)$ \\
+    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & $(S,E)\mapsto(S-1,E+1)$  \bigstrut & $(1,1)$ & pure birth \\
+    $\lab{Prog}$   & $\sigma E$ \bigstrut & $(E,I)\mapsto(E-1,I+1)$ \bigstrut & $(0,1)$ & pure migration \\
+    $\lab{Recov}$  & $\gamma I$ \bigstrut & $(I,R)\mapsto(I-1,R+1)$ \bigstrut & $(0,0)$ & pure death \\
+    $\lab{Wane}$   & $\omega R$ \bigstrut & $(S,R)\mapsto(S+1,R-1)$ \bigstrut & $(0,0)$ & neutral \\
+    $\lab{Sample}$ & $\psi I$ \bigstrut & $x\mapsto{x}$ \bigstrut & $(0,1)$ & pure sample \\
     \hline\hline
@@
 \end{table}
+
+\begin{table}[h!]
+  \caption{
+    \label{tab:seirs_integ}
+    Elements of a scheme for computing the likelihood under the SEIRS model.
+    The $y\mapsto{y'}$ column depicts the proposed painting schematically.
+    Blue is used for the $\lab{E}$ deme and yellow for the $\lab{I}$ deme.
+  }
+  %% \renewcommand{\arraystretch}{1.2}
+  \begin{tabular}{c c c c c c c c | c}
+    \hline\hline
+    & $u$ & $s$ & $\BinRatio{n}{\ell}{r}{s}$ & $y\mapsto{y'}$ & $\pi$ & n & $Q$ & Line\\[2ex]
+    \hline
+    \multirow[c]{10}{*}[-1ex]{\rotatebox{90}{$t\notin{{B}\cup{S_0}\cup{S_1}}$}}
+    & \multirow{4}{*}{$\lab{Trans}$}
+    & $(0,0)$ & $\left(\frac{E-\ell_E}{E}\right)\left(\frac{I-\ell_I}{I}\right)$ \bigstrut & \unbroken \bigstrut & $\frac{I+1-\ell_I}{I+1}$ & 1 & 1 & 1 \\
+    \cline{3-8}
+    && $(1,0)$ & $\frac{I-\ell_I}{E I}$ \bigstrut & \maizeblue \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 1 & 2 \\
+    \cline{3-8}
+    && $(0,1)$ & $\frac{E-\ell_E}{E I}$ \bigstrut & \maizemaize \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 1 & 3 \\
+    \cline{3-8}
+    && $(1,1)$ & --- & --- & 0 & 0 & 0 & 4 \\
+    \cline{2-8}
+    &\multirow{2}{*}{$\lab{Prog}$}
+    & $(0,0)$ & $\frac{I-\ell_I}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \unbroken \bigstrut & $\frac{E+1-\ell_E}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & 1 & 1 & 5 \\
+    \cline{3-8}
+    && $(0,1)$ & $\frac{1}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \bluemaize \bigstrut & $\frac{1}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & $\ell_E$ & 1 & 6 \\
+    \cline{2-8}
+    & $\lab{Recov}$ & $(0,0)$ & $\Indicator{I\ge\ell_I}$ \bigstrut & \unbroken \bigstrut & $\Indicator{I>\ell_I}$ \bigstrut & 1 & 1 & 7 \\
+    \cline{2-8}
+    & $\lab{Wane}$ & $(0,0)$ & 1 \bigstrut & \unbroken \bigstrut & 1 & 1 & 1 & 8 \\
+    \cline{2-8}
+    & $\lab{Sample}$ & $\ast$ & --- & --- & 1 & 0 & 0 & 9 \\
+    \cline{1-8}
+    \multirow[c]{4}{*}[-3ex]{\rotatebox{90}{$t\in{B}$}}
+    & \multirow{4}{*}{$\lab{Trans}$}
+    & \multirow{4}{*}{$(1,1)$}
+    & $\frac{1}{E I}$ \bigstrut & \branchup \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 1 & 10 \\
+    \cline{4-8}
+    &&& $\frac{1}{E I}$ \bigstrut & \branchdown \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 1 & 11 \\
+    \cline{4-8}
+    &&& --- \bigstrut & \branchupnix \bigstrut & 0 & 0 & 0 & 12 \\
+    \cline{4-8}
+    &&& --- \bigstrut & \branchdownnix \bigstrut & 0 & 0 & 0 & 13 \\
+    \cline{2-8}
+    & $\ast$ & $\ast$ & --- \bigstrut & --- & 0 & 0 & 0 & 14 \\
+    \cline{1-8}
+    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_0}$} \bigstrut}
+    & \multirow{2}{*}{$\lab{Sample}$}
+    & \multirow{2}{*}{$(0,0)$}
+    & $\frac{I-\ell_I}{I}$ \bigstrut & \samplezero & 1 & 1 & 1 & 15 \\
+    \cline{4-8}
+    &&& --- & \samplezeronix \bigstrut & 0 & 0 & 0 & 16 \\
+    \cline{2-8}
+    & $\ast$ & $\ast$ & --- & --- & 0 & 0 & 0 & 17 \\
+    \cline{1-8}
+    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_1}$}}
+    & \multirow{2}{*}{$\lab{Sample}$}
+    & \multirow{2}{*}{$(0,1)$}
+    & $\frac{1}{I}$ \bigstrut & \sampleone \bigstrut & 1 & 1 & 1 & 18 \\
+    \cline{4-8}
+    &&& --- & \sampleonenix \bigstrut & 0 & 0 & 0 & 19 \\
+    \cline{2-8}
+    & $\ast$ & $\ast$ & --- & --- & 0 & 0 & 0 & 20 \\
+    \hline\hline
+  \end{tabular}
+\end{table}
+
+A variety of importance-sampling kernels are permissible under the terms of \cref{thm:obsc_uncond}.
+Given a particular choice of importance-sampling kernel, the filtering equation follows.
+For example, with the choices as in \cref{tab:seirs_integ}, the filtering equation is as follows.
+As previously, let $B$ be the set of branch times, $S_0$ be the set of tip-sample times, and $S_1$ be the set of inline sample times.
+Then for $t\notin{{B}\cup{S_0}\cup{S_1}}$, we have
+\begin{mathsize}{8pt}{9pt}
+  \begin{equation*}
+    \begin{split}
+      \frac{\partial{w}}{\partial{t}}(t,S,E,I,R,y)=
+      %% &\frac{\beta\,(S+1)\,I}{N}\,\frac{I-\ell_I}{I}\,\frac{E-\ell_E}{E}\,w(t,S+1,E-1,I,R,y)\\
+      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\frac{I-\ell_I}{E}\,w(t,S+1,E-1,I,R,\leftlim{y}_2)\\
+      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\frac{E-\ell_E}{E}\,w(t,S+1,E-1,I,R,\leftlim{y}_3)
+      %% -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
+      %% &+\sigma\,(E+1)\frac{E+1-\ell_E}{E+1}\,\Indicator{E\ge{\ell_E}}\,\frac{I-\ell_I}{I}\,\frac{E+1}{E+1-\ell_E}\,w(t,S,E+1,I-1,R,y)\\
+      %% &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\frac{1}{E+1}\,\Indicator{E\ge{\ell_E}}\,\frac{E+1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6)
+      %% -\sigma\,E\,w(t,S,E,I,R,y)\\
+      %% &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
+      %% -\gamma\,I\,w(t,S,E,I,R,y)\\
+      %% &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
+      %% -\omega\,R\,w(t,S,E,I,R,y)
+      %% -\psi\,I\,w(t,S,E,I,R,y).
+      %%
+      &\frac{\beta\,(S+1)\,I}{N}\,\left(1-\frac{\ell_I}{I}\right)\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,y)\\
+      &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\,\frac{1}{E}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_2^k)\\
+      &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_3^k)
+      -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
+      &+\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S,E+1,I-1,R,y)\\
+      &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\frac{1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6^k)
+      -\sigma\,E\,w(t,S,E,I,R,y)\\
+      &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
+      -\gamma\,I\,w(t,S,E,I,R,y)\\
+      &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
+      -\omega\,R\,w(t,S,E,I,R,y)
+      -\psi\,I\,w(t,S,E,I,R,y).
+      %%
+      %% &\frac{\beta\,(S+1)\,I}{N}\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,y)\\
+      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\,\frac{1}{E}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_2^k)
+      %% -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
+      %% &+\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S,E+1,I-1,R,y)\\
+      %% &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\frac{1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6^k)
+      %% -\sigma\,E\,w(t,S,E,I,R,y)\\
+      %% &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
+      %% -\gamma\,I\,w(t,S,E,I,R,y)\\
+      %% &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
+      %% -\omega\,R\,w(t,S,E,I,R,y)
+      %% -\psi\,I\,w(t,S,E,I,R,y).
+    \end{split}
+  \end{equation*}
+\end{mathsize}%
 
@@
   with(
-    mcap(logLik,sigma,span=0.25)
+    mcap(logLik,sigma,span=0.5)
   ) -> mcap
@@
   \begin{center}
-    <<seirs3_plot,fig.dim=c(8,2.8),out.width="100%">>=
+    <<seirs3_plot,fig.dim=c(8,2.8),out.width="100%",dependon="seirs3">>=
     plot_grid(
@@
   \renewcommand{\arraystretch}{1.2}
-  \begin{tabular}{cclc}
+  \begin{tabular}{cccl}
     \hline\hline
-    $u$ & $\alpha_u$ & Event type & $r_u$ \\
+    $u$ & $\alpha_u$ & $r_u$ & Event type \\
     \hline
-    $\lab{Birth}$  & $\lambda N$  & pure birth  & 2 \\
-    $\lab{Death}$  & $\gamma N$   & pure death  & 0 \\
-    $\lab{Sample}$ & $\psi N$     & pure sample & 1 \\
+    $\lab{Birth}$  & $\lambda N$ & 2 & pure birth \\
+    $\lab{Death}$  & $\gamma N$ & 0 & pure death \\
+    $\lab{Sample}$ & $\psi N$ & 1 & pure sample \\
     \hline\hline

### 2024-05-21 ms.Rnw
blue box

@@
   We show how each member of a broad class of Markov population models uniquely induces a stochastic process on the space of genealogies.
-  We then derive exact expressions for the likelihood of an observed genealogy in terms of filtering equations, which can be integrated using standard Monte Carlo methods.
+  We then derive exact expressions for the likelihood of an observed genealogy in terms of filter equations, which can be integrated using standard Monte Carlo methods.
   The class of population models that can be accommodated includes essentially all compartmental models.
@@
 %% Using these, we derive equations for the likelihood of a genealogy conditional on the history.
-%% We then integrate out the history to obtain nonlinear filtering equations, the solution of which yields the likelihood.
+%% We then integrate out the history to obtain nonlinear filter equations, the solution of which yields the likelihood.
 %% These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likelihood.
@@
 \setcounter{table}{0}
+%%\setcounter{tcbcounter}{0}
+
 \titleformat{\section}[hang]{\large\bfseries}{Appendix \periodafter\thesection}{2ex}{\periodafter}{}
@@
 \renewcommand{\thetable}{\thesection\arabic{table}}
-
-\section{Filter equations}
+%%\renewcommand{\thetcbcounter}{\thesection\arabic{tcbcounter}}
+
+\section{filter equations}
 \label{sec:filter_eqns}
@@
 As such, explicit expressions for them are not available, and we require mathematical tools to allow us to manipulate these quantities and devise algorithms for their numerical solution.
-The \emph{filtering equations} we introduce here are suitable for these purposes, and we devote this appendix to exposing their essential properties.
+The \emph{filter equations} we introduce here are suitable for these purposes, and we devote this appendix to exposing their essential properties.
 This extremely convenient formalism has, to our knowledge, not been thoroughly exploited, though we note their resemblance to the constructions of \citet{Ogata1978}, \citet{Puri1986}, \citet{Kliemann1990}, and \citet{Giesecke2018}.
@@
   \end{equation*}
-  In the limit, we obtain \cref{eq:reg-filter-eq2}, the regular filtering equation generated by $\alpha$, with boost $B$ and zero decay.
+  In the limit, we obtain \cref{eq:reg-filter-eq2}, the regular filter equation generated by $\alpha$, with boost $B$ and zero decay.
   %%  \AAK{[This depends on $\alpha$ and $B$ being continuous in their first arguments. We have assumed they are continuous almost everywhere.]}
@@
     Elements of the SIRS model pertinent to the genealogy process.
+    The table shows the rate ($\alpha_u$), jump ($x\mapsto{x'}$), production ($r_u$), and event type for each of the model's four marks ($u$).
   }
   %% \renewcommand{\arraystretch}{1.2}
-  \begin{tabular}{cccl}
+  \begin{tabular}{ccccl}
     \hline\hline
-    $u$ & $\alpha_u$ & $r_u$ & Event type \\
+    $u$ & $\alpha_u$ \bigstrut & $x\mapsto{x'}$ \bigstrut & $r_u$ \bigstrut & Event type \\
     \hline
-    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & 2 & pure birth \\
-    $\lab{Recov}$  & $\gamma I$ \bigstrut & 0 & pure death \\
-    $\lab{Wane}$   & $\omega R$ \bigstrut & 0 & neutral \\
-    $\lab{Sample}$ & $\psi I$ \bigstrut & 1 & pure sample \\
+    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & $(S,I)\mapsto(S-1,I+1)$ \bigstrut & 2 & pure birth \\
+    $\lab{Recov}$  & $\gamma I$ \bigstrut & $(I,R)\mapsto(I-1,R+1)$ \bigstrut & 0 & pure death \\
+    $\lab{Wane}$   & $\omega R$ \bigstrut & $(S,R)\mapsto(S+1,R-1)$ \bigstrut & 0 & neutral \\
+    $\lab{Sample}$ & $\psi I$ \bigstrut & $x\mapsto{x}$ \bigstrut & 1 & pure sample \\
     \hline\hline
@@
     Elements of the SEIRS model pertinent to the genealogy process.
+    The table shows the rate ($\alpha_u$), jump ($x\mapsto{x'}$), production ($r_u$), and event type for each of the model's five marks ($u$).
   }
@@
 
-\begin{table}[h!]
+\begin{table}
   \caption{
     \label{tab:seirs_integ}
-    Elements of a scheme for computing the likelihood under the SEIRS model.
-    The $y\mapsto{y'}$ column depicts the proposed painting schematically.
+    Elements of a scheme for numerically computing the likelihood under the SEIRS model.
+    The regular portion of the filter equation holds in between genealogical events (\ie for $t\notin{{B}\cup{S_0}\cup{S_1}}$);
+    the singular portion describes the effect of these events ($t\in{{B}\cup{S_0}\cup{S_1}}$).
+    Each line corresponds to a potential event, but only those for which $Q=1$ appear in the equation.
+    An event with mark $u$ and saturation $s$ has the boost given by the binomial ratio shown (third column).
+    The $y\mapsto{y'}$ column depicts the proposed painting schematically, and $\pi$ is the probability of that proposal.
     Blue is used for the $\lab{E}$ deme and yellow for the $\lab{I}$ deme.
+    The filter equation contains $n$ terms of each type.
+    An asterisk ($\ast$) stands for cases not explicitly mentioned.
   }
@@
 
-A variety of importance-sampling kernels are permissible under the terms of \cref{thm:obsc_uncond}.
-Given a particular choice of importance-sampling kernel, the filtering equation follows.
-For example, with the choices as in \cref{tab:seirs_integ}, the filtering equation is as follows.
-As previously, let $B$ be the set of branch times, $S_0$ be the set of tip-sample times, and $S_1$ be the set of inline sample times.
-Then for $t\notin{{B}\cup{S_0}\cup{S_1}}$, we have
-\begin{mathsize}{8pt}{9pt}
+The filter equation corresponding to the scheme of \cref{tab:seirs_integ} is presented in \Cref{box:SEIRS}.
+Some numerical results are presented in \cref{fig:seirs_example}.
+
+\clearpage
+\newpage
+
+\begin{bluebox}[label={box:SEIRS}]{Filter equation for the SEIRS model}
+  As previously, given an obscured genealogy, let $B$ be the set of its branch times, $S_0$ be the set of tip-sample times, and $S_1$ be the set of inline sample times.
+  Then for $t\notin{{B}\cup{S_0}\cup{S_1}}$, the filter equation reads:
+  \begin{mathsize}{9pt}{10pt}
     \begin{equation*}
       \begin{split}
-      \frac{\partial{w}}{\partial{t}}(t,S,E,I,R,y)=
-      %% &\frac{\beta\,(S+1)\,I}{N}\,\frac{I-\ell_I}{I}\,\frac{E-\ell_E}{E}\,w(t,S+1,E-1,I,R,y)\\
-      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\frac{I-\ell_I}{E}\,w(t,S+1,E-1,I,R,\leftlim{y}_2)\\
-      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\frac{E-\ell_E}{E}\,w(t,S+1,E-1,I,R,\leftlim{y}_3)
-      %% -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
-      %% &+\sigma\,(E+1)\frac{E+1-\ell_E}{E+1}\,\Indicator{E\ge{\ell_E}}\,\frac{I-\ell_I}{I}\,\frac{E+1}{E+1-\ell_E}\,w(t,S,E+1,I-1,R,y)\\
-      %% &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\frac{1}{E+1}\,\Indicator{E\ge{\ell_E}}\,\frac{E+1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6)
-      %% -\sigma\,E\,w(t,S,E,I,R,y)\\
-      %% &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
-      %% -\gamma\,I\,w(t,S,E,I,R,y)\\
-      %% &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
-      %% -\omega\,R\,w(t,S,E,I,R,y)
-      %% -\psi\,I\,w(t,S,E,I,R,y).
-      %%
-      &\frac{\beta\,(S+1)\,I}{N}\,\left(1-\frac{\ell_I}{I}\right)\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,y)\\
+        \frac{\partial{w}}{\partial{t}}(t,S,E,I,R,y)=&
+        \frac{\beta\,(S+1)\,I}{N}\,\left(1-\frac{\ell_E}{E}\right)\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,y)\\
         &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\,\frac{1}{E}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_2^k)\\
@@
         -\psi\,I\,w(t,S,E,I,R,y).
-      %%
-      %% &\frac{\beta\,(S+1)\,I}{N}\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,y)\\
-      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\,\frac{1}{E}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_2^k)
-      %% -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
-      %% &+\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S,E+1,I-1,R,y)\\
-      %% &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\frac{1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6^k)
-      %% -\sigma\,E\,w(t,S,E,I,R,y)\\
-      %% &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
-      %% -\gamma\,I\,w(t,S,E,I,R,y)\\
-      %% &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
-      %% -\omega\,R\,w(t,S,E,I,R,y)
-      %% -\psi\,I\,w(t,S,E,I,R,y).
       \end{split}
@@
   \end{mathsize}%
+  Here, $\leftlim{y}_j^k$ refers to the coloring of the tree immediately preceding the proposal indicated on line $j$ of \cref{tab:seirs_integ}.
+  The integer $k$ specifies the particular branch on which the change occurs.
+
+  The singular portion of the filter equation has one component for each distinct type of genealogical event:
+  \begin{mathsize}{9pt}{10pt}
+    \begin{equation*}
+      w(t,S,E,I,R,y)=
+      \begin{cases}
+        \frac{\beta\,(S+1)\,I}{N}\,\frac{1}{E I}\,\wt(t,S+1,E-1,I,R,\leftlim{y}_{10})\\[2ex]
+        \qquad+\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{E I}\,\wt(t,S+1,E-1,I,R,\leftlim{y}_{11}), & t\in{B},\\[2ex]
+        \psi\,\left(I-\ell_I\right)\,\wt(t,S,E,I,R,\leftlim{y}_{15}), & t\in{S_0},\\[2ex]
+        \psi\,\wt(t,S,E,I,R,\leftlim{y}_{18}), & t\in{S_1}.
+      \end{cases}
+    \end{equation*}
+  \end{mathsize}%
+
+  A variety of importance-sampling kernels are permissible under the terms of \cref{thm:obsc_uncond}.
+  With a particular choice of importance-sampling kernel, the filter equation uniquely specifies a sequential Monte Carlo algorithm for estimating the likelihood.
+  The choices made in \cref{tab:seirs_integ} underlie the results displayed in \cref{fig:seirs_example}.
+\end{bluebox}
 
@@
     \label{fig:seirs_example}
-    Likelihood computation for the SEIRS model by sequential Monte Carlo.
+    Likelihood computation for the SEIRS model by sequential Monte Carlo, using the scheme of \cref{box:SEIRS}.
     (A)~Simulated genealogy for $\beta=\Sexpr{seirs.params$Beta}$,
@@
     (B)~Likelihood slice in the $\sigma$-direction.
-    Each point represents the estimate of a distinct sequential Monte Carlo computation.
+    Each point represents the estimate of an independent sequential Monte Carlo computation.
     The blue curve shows a LOESS smooth;

### 2024-05-22 ms.Rnw
small tweaks

@@
   \begin{equation}\label{eq:phidef}
-    \phi_u(x,y,y')\coloneq\BinRatio{n(x)}{\ell(y')}{r^{u}}{s(y,y')}\,Q_{u}(y,y'),
+    \phi^{}_u(x,y,y')\coloneq\BinRatio{n(x)}{\ell(y')}{r^{u}}{s(y,y')}\,Q_{u}(y,y'),
   \end{equation}
@@
   \begin{equation*}
-    \CondProb{\Pr_\T=\P}{\Hr_\T=\H}=\Indicator{\event{\H}\supseteq{\event{\P}}}\,\prod_{t\in\event{\H}}{\phi_{\U_t}(\X_t,\Yt_t,\Y_t)}.
+    \CondProb{\Pr_\T=\P}{\Hr_\T=\H}=\Indicator{\event{\H}\supseteq{\event{\P}}}\,\prod_{t\in\event{\H}}{\phi^{}_{\U_t}\!(\X_t,\Yt_t,\Y_t)}.
   \end{equation*}
@@
 
-%% \begin{equation*}
-%%   \phi_u(\xi,y,y')\colonequals\BinRatio{n(\xi)}{\ell(y')}{r^u}{s(y,y')}\,Q_u(y,y').
-%% \end{equation*}
-
 Next, we show how the likelihood of a pruned genealogies, unconditional on the history, can be computed.
@@
       \frac{\partial{w}}{\partial{t}}(t,x)=
-      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\Yt_t,\Y_t)\,\dd{x'}}}
+      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi^{}_u(x,\Yt_t,\Y_t)\,\dd{x'}}}
       -\sum_u{\int{w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}}},
       &t\notin{\event{\P}},\\
-      w(t,x)=&\sum_u{\int{\wt(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\Yt_t,\Y_t)\,\dd{x'}}},
+      w(t,x)=&\sum_u{\int{\wt(t,x')\,\alpha_u(t,x',x)\,\phi^{}_u(x,\Yt_t,\Y_t)\,\dd{x'}}},
       &t\in{\event{\P}},
@@
     \CondProb{\Pr^{\lab{Z}}_\T=\Z}{\Hr_\T=\H}=
-    \Indicator{\event{\H}\supseteq\event{\Z}}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
+    \Indicator{\event{\H}\supseteq\event{\Z}}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}}{\frac{\phi^{}_{\U_t}(\X_t,\yrt_t,\yr_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
   \end{equation*}
@@
       \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),\qquad
-      \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y')}{\pi_u(t,x,x',y,y')},
+      \varPsi_u(t,x,x',y,y')=\frac{\phi^{}_u(x',y,y')}{\pi_u(t,x,x',y,y')},
     \end{gathered}
@@
       &\frac{\partial{w}}{\partial{t}}=
-      \sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}}
+      \sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\varPsi_u(t,x',x,y',y)\,\dd{x'}}}
       -\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}},
       &t\in{\event{\Z}},\\
-      &w(t,x,y)=\sum_{uy'}{\int{\wt(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}},
+      &w(t,x,y)=\sum_{uy'}{\int{\wt(t,x',y')\,\beta_u(t,x',x,y',y)\,\varPsi_u(t,x',x,y',y)\,\dd{x'}}},
       &t\in{\event{\Z}},
@@
 
-\section{filter equations}
+\section{Filter equations}
 \label{sec:filter_eqns}
@@
     Blue is used for the $\lab{E}$ deme and yellow for the $\lab{I}$ deme.
-    The filter equation contains $n$ terms of each type.
+    For each line, the filter equation contains $m$ terms.
     An asterisk ($\ast$) stands for cases not explicitly mentioned.
@@
     \hline\hline
-    & $u$ & $s$ & $\BinRatio{n}{\ell}{r}{s}$ & $y\mapsto{y'}$ & $\pi$ & n & $Q$ & Line\\[2ex]
+    & $u$ & $s$ & $Q$ & $\BinRatio{n}{\ell}{r}{s}$ & $y\mapsto{y'}$ & $\pi$ & $m$ & Line\\[2ex]
     \hline
@@
     & \multirow{4}{*}{$\lab{Trans}$}
-    & $(0,0)$ & $\left(\frac{E-\ell_E}{E}\right)\left(\frac{I-\ell_I}{I}\right)$ \bigstrut & \unbroken \bigstrut & $\frac{I+1-\ell_I}{I+1}$ & 1 & 1 & 1 \\
+    & $(0,0)$ & 1 & $\left(\frac{E-\ell_E}{E}\right)\left(\frac{I-\ell_I}{I}\right)$ \bigstrut & \unbroken \bigstrut & $\frac{I+1-\ell_I}{I+1}$ & 1 & 1 \\
     \cline{3-8}
-    && $(1,0)$ & $\frac{I-\ell_I}{E I}$ \bigstrut & \maizeblue \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 1 & 2 \\
+    && $(1,0)$ & 1 & $\frac{I-\ell_I}{E I}$ \bigstrut & \maizeblue \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 2 \\
     \cline{3-8}
-    && $(0,1)$ & $\frac{E-\ell_E}{E I}$ \bigstrut & \maizemaize \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 1 & 3 \\
+    && $(0,1)$ & 1 & $\frac{E-\ell_E}{E I}$ \bigstrut & \maizemaize \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 3 \\
     \cline{3-8}
-    && $(1,1)$ & --- & --- & 0 & 0 & 0 & 4 \\
+    && $(1,1)$ & 0 & & & & 0 & 4 \\
     \cline{2-8}
     &\multirow{2}{*}{$\lab{Prog}$}
-    & $(0,0)$ & $\frac{I-\ell_I}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \unbroken \bigstrut & $\frac{E+1-\ell_E}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & 1 & 1 & 5 \\
+    & $(0,0)$ & 1 & $\frac{I-\ell_I}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \unbroken \bigstrut & $\frac{E+1-\ell_E}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & 1 & 5 \\
     \cline{3-8}
-    && $(0,1)$ & $\frac{1}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \bluemaize \bigstrut & $\frac{1}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & $\ell_E$ & 1 & 6 \\
+    && $(0,1)$ & 1 & $\frac{1}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \bluemaize \bigstrut & $\frac{1}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & $\ell_E$ & 6 \\
     \cline{2-8}
-    & $\lab{Recov}$ & $(0,0)$ & $\Indicator{I\ge\ell_I}$ \bigstrut & \unbroken \bigstrut & $\Indicator{I>\ell_I}$ \bigstrut & 1 & 1 & 7 \\
+    & $\lab{Recov}$ & $(0,0)$ & 1 & $\Indicator{I\ge\ell_I}$ \bigstrut & \unbroken \bigstrut & $\Indicator{I>\ell_I}$ \bigstrut & 1 & 7 \\
     \cline{2-8}
-    & $\lab{Wane}$ & $(0,0)$ & 1 \bigstrut & \unbroken \bigstrut & 1 & 1 & 1 & 8 \\
+    & $\lab{Wane}$ & $(0,0)$ & 1 & 1 \bigstrut & \unbroken \bigstrut & 1 & 1 & 8 \\
     \cline{2-8}
-    & $\lab{Sample}$ & $\ast$ & --- & --- & 1 & 0 & 0 & 9 \\
+    & $\lab{Sample}$ & $\ast$ & 0 &  &  & & 0 & 9 \\
     \cline{1-8}
@@
     & \multirow{4}{*}{$(1,1)$}
-    & $\frac{1}{E I}$ \bigstrut & \branchup \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 1 & 10 \\
+    & 1 & $\frac{1}{E I}$ \bigstrut & \branchup \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 10 \\
     \cline{4-8}
-    &&& $\frac{1}{E I}$ \bigstrut & \branchdown \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 1 & 11 \\
+    &&& 1 & $\frac{1}{E I}$ \bigstrut & \branchdown \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 11 \\
     \cline{4-8}
-    &&& --- \bigstrut & \branchupnix \bigstrut & 0 & 0 & 0 & 12 \\
+    &&& 0 &  \bigstrut & \branchupnix \bigstrut & & 0 & 12 \\
     \cline{4-8}
-    &&& --- \bigstrut & \branchdownnix \bigstrut & 0 & 0 & 0 & 13 \\
+    &&& 0 &  \bigstrut & \branchdownnix \bigstrut & & 0 & 13 \\
     \cline{2-8}
-    & $\ast$ & $\ast$ & --- \bigstrut & --- & 0 & 0 & 0 & 14 \\
+    & $\ast$ & $\ast$ & 0 &  \bigstrut &  & & 0 & 14 \\
     \cline{1-8}
@@
     & \multirow{2}{*}{$(0,0)$}
-    & $\frac{I-\ell_I}{I}$ \bigstrut & \samplezero & 1 & 1 & 1 & 15 \\
+    & 1 & $\frac{I-\ell_I}{I}$ \bigstrut & \samplezero & 1 & 1 & 15 \\
     \cline{4-8}
-    &&& --- & \samplezeronix \bigstrut & 0 & 0 & 0 & 16 \\
+    &&& 0 &  & \samplezeronix \bigstrut & & 0 & 16 \\
     \cline{2-8}
-    & $\ast$ & $\ast$ & --- & --- & 0 & 0 & 0 & 17 \\
+    & $\ast$ & $\ast$ & 0 &  &  & & 0 & 17 \\
     \cline{1-8}
@@
     & \multirow{2}{*}{$(0,1)$}
-    & $\frac{1}{I}$ \bigstrut & \sampleone \bigstrut & 1 & 1 & 1 & 18 \\
+    & 1 & $\frac{1}{I}$ \bigstrut & \sampleone \bigstrut & 1 & 1 & 18 \\
     \cline{4-8}
-    &&& --- & \sampleonenix \bigstrut & 0 & 0 & 0 & 19 \\
+    &&& 0 &  & \sampleonenix \bigstrut & & 0 & 19 \\
     \cline{2-8}
-    & $\ast$ & $\ast$ & --- & --- & 0 & 0 & 0 & 20 \\
+    & $\ast$ & $\ast$ & 0 &  &  & & 0 & 20 \\
     \hline\hline
@@
   \begin{mathsize}{9pt}{10pt}
-    \begin{equation*}
+    \begin{equation}
+      \label{eq:seirs_filter_reg}
       \begin{split}
-        \frac{\partial{w}}{\partial{t}}(t,S,E,I,R,y)=&
+        \frac{\partial{w}}{\partial{t}}(t,S,&E,I,R,y)=
         \frac{\beta\,(S+1)\,I}{N}\,\left(1-\frac{\ell_E}{E}\right)\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,y)\\
@@
       \end{split}
-    \end{equation*}
+    \end{equation}
   \end{mathsize}%
@@
   \begin{mathsize}{9pt}{10pt}
-    \begin{equation*}
+    \begin{equation}
+      \label{eq:seirs_filter_sing}
       w(t,S,E,I,R,y)=
@@
       \end{cases}
-    \end{equation*}
+    \end{equation}
   \end{mathsize}%
+
+  In addition to \cref{eq:seirs_filter_reg,eq:seirs_filter_sing}, the quantity $w$ should satisfy the condition $w(t,S,E,I,R,y)=0$ whenever $E<\ell_E$ or $I<\ell_I$.
 
@@
     Likelihood computation for the constant-parameter, linear birth-death-sampling model, according to \cref{thm:obsc_uncond} via sequential Monte Carlo.
-    Panel (A) shows the genealogy, simulated for $\lambda=\Sexpr{lbdp.params$lambda}$, $\mu=\Sexpr{lbdp.params$mu}$, $\psi=\Sexpr{lbdp.params$psi}$, $N_0=\Sexpr{lbdp.params$mu}$.
+    Panel (A) shows the genealogy, simulated for $\lambda=\Sexpr{lbdp.params$lambda}$, $\mu=\Sexpr{lbdp.params$mu}$, $\psi=\Sexpr{lbdp.params$psi}$, $N_0=\Sexpr{lbdp.params$n0}$.
     Panel (B) shows a likelihood slice, through the true parameters in the $\mu$ direction.

### 2024-05-24 ms.Rnw
edits in response to comments by ELI

@@
 
-\title[Exact Phylodynamics]{Exact Phylodynamics via Structured Markov Genealogy Processes}
+\title[Exact Phylodynamic Likelihood]{Exact Phylodynamic Likelihood\\ via Structured Markov Genealogy Processes}
 \author[King]{Aaron~A.~King}
@@
 
-\hypersetup{pdftitle={Exact Phylodynamics via Structured Markov Genealogy Processes}}
+\hypersetup{pdftitle={Exact Phylodynamic Likelihood via Structured Markov Genealogy Processes}}
 \hypersetup{pdfauthor={A.A. King, Q.-Y. Lin, E.L. Ionides}}
@@
 \begin{abstract}
-  We show how each member of a broad class of Markov population models uniquely induces a stochastic process on the space of genealogies.
-  We then derive exact expressions for the likelihood of an observed genealogy in terms of filter equations, which can be integrated using standard Monte Carlo methods.
-  The class of population models that can be accommodated includes essentially all compartmental models.
+  We consider genealogies arising from a Markov population process in which individuals are categorized into a discrete collection of compartments, with the requirement that individuals within the same compartment are statistically exchangeable.
+  When equipped with a sampling process, each such population process induces a time-evolving tree-valued process defined as the genealogy of all sampled individuals.
+  We provide a construction of this genealogy process and derive exact expressions for the likelihood of an observed genealogy in terms of filter equations.
+  These filter equations can be numerically solved using standard Monte Carlo integration methods.
+  Thus, we obtain statistically efficient likelihood-based inference for essentially arbitrary compartment models based on an observed genealogy of individuals sampled from the population.
 \end{abstract}
@@
 
-When an infectious agent mutates quickly enough so that information on the history of its passage through hosts accumulates in its genome, yet not so quickly that this information is dissipated, it is possible to extract this information from sampled genomes to gain insight into key determinants of infectious disease transmission.
-With sufficiently dense sampling, it can be possible to reconstruct the detailed history of who infected whom, which can aid identification of the correlates of transmission and immunity and the properties of infections.
-More typically, one can use relatively sparse samples to learn about these features via the estimation of a parsimonious model of the transmission process.
-In particular, one can formalize one or more mathematical models of transmission, estimate their parameters, and compare their ability to explain data, following standard statistical paradigms.
-This is known as \emph{phylodynamic} inference, the term being due to \citet{Grenfell2004}.
+When the genome of an infectious agent accumulates mutations on timescales similar to those of transmission and infection progression, the resulting pattern of differences among genomes contains information on the history of the pathogen's passage through individual hosts and the host population.
+As \citet{Grenfell2004} observed, one can extract this information to gain insight into the structure and dynamics of the host-pathogen system.
+In particular, one can formalize mathematical models of transmission, estimate their parameters, and compare their ability to explain data, following standard statistical paradigms.
+This is known as \emph{phylodynamic} inference.
 \citet{Alizon2024} gives a good review of the history of the subject.
@@
 
-In particular, existing approaches to the phylodynamic likelihood have been based on one of two mathematical idealizations.
-The first is the \citet{Kingman1982a} coalescent, by which likelihood of a given genealogy is computed via a reverse-time argument.
-This computation is exact when the population size is constant.
+Existing approaches to the phylodynamic likelihood have been based on one of two mathematical idealizations.
+The first is the \citet{Kingman1982a} coalescent, by which likelihood of a given genealogy is computed using a reverse-time argument.
+Thiis computation provides the exact likelihood for a genealogy resulting from a particular, constant population-size, dynamic model \citep[the Moran model, \eg][]{Moran1958,Kingman1982b,Moehle2000}.
 Extensions of this approach develop approximate likelihoods for the case when the population size varies as a function of time \citep{Griffiths1994,Drummond2005} or according to an SIR process \citep{Volz2009a,Rasmussen2011}, as long as the population size is large and the sample-fraction remains negligible.
@@
 First, it would afford exact computation of the phylodynamic likelihood, so that comparisons among parameterizations and models could be made on a sound basis.
-Second, because nonlinearity, nonstationarity, noise, and measurement error are so prominent and ubiquitous in epidemiology, it would accommodate nonlinear, time-inhomogeneous, stochastic transmission models.
-Third, because many of the most important uncertainties concern heterogeneities in transmission rates and the susceptibility, behavior, age, and location of hosts, it would accommodate host populations structured by these factors.
+Second, because nonlinearity, nonstationarity, noise, and measurement error are prominent and ubiquitous in epidemiology, it would accommodate nonlinear, time-inhomogeneous, stochastic transmission models.
+Third, because many of the most scientifically important uncertainties concern heterogeneities in transmission rates and the susceptibility, behavior, age, and location of hosts, it would accommodate host populations structured by these factors.
 While some structuring factors (\eg age, spatial location) are most naturally expressed in terms of continuous variables, discretely structured models have repeatedly proved their value in epidemiology.
@@
 This paper demonstrates how these desiderata can be achieved---at least for models with discrete structure---including arbitrary nonlinear compartmental models with a countable number of compartments.
+%% \ELI{[For the claim of generality, should we give a disclaimer that model complexity is limited in practice by both the limited amount of information in data to fit the model, and the numerical tractability of the integration.
+%%     Some idea of the largest size which has been tested to date (say, 5 compartments and 1000 sequences?) would be helpful to the reader.
+%%     Readers may naturally be skeptical about the practical scope of the methods.
+%%     For example, familiar Monte Carlo methods such as BEAST have theoretical guarantees going far beyond their practical applicability.]}
 
@@
 
-Code sufficient for the reproduction of all the results presented in this paper are freely available for download at \url{https://github.com/kingaa/phylopomp-project}.
+Code sufficient for the reproduction of all the results presented in this paper are freely available for download at \url{https://github.com/kingaa/structured-genealogy-process-paper}.
 An archival version of these will be stored on Zotero upon publication of this paper.
@@
 %% We then integrate out the history to obtain nonlinear filter equations, the solution of which yields the likelihood.
-%% These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likelihood.
+%% These readily lend themselves to a family of Sequential Monte Carlo algorithms for computing the likelihood.
 %% We demonstrate with several examples.
@@
 
-The \citep{Kingman1982a} coalescent is another workhorse in existing phylodynamic approaches.
+The \citet{Kingman1982a} coalescent is another workhorse in existing phylodynamic approaches.
 It is the ancestral process for the Moran model, in which a fixed population of $n$ lineages experiences events at times distributed according to a rate-$\mu$ Poisson process.
@@
 
-Importantly, because the theory presented here allows computation of the likelihood via strictly forward-in-time computations, it permits consideration of models for which time-reversal arguments are not available.
+The theory presented here represents a strict generalization of the existing coalescent and birth-death process approaches to phylodynamic inference.
+In \cref{sec:app-examples}, we demonstrate that both of the latter processes are special cases of the genealogical processes constructed here.
+Importantly, because the theory allows computation of the likelihood via strictly forward-in-time computations, it permits consideration of models for which time-reversal arguments are not available.
 Moreover, inasmuch as the formulae of \cref{thm:obsc_uncond} can be efficiently computed via sequential Monte Carlo, explicit expressions for transition probabilities are not needed:
 it is sufficient to be able to simulate from the population process.
-This feature of the algorithms---known as the \emph{plug-and-play property}---expands the set of population models that can be confronted with data still further.
+This feature of the algorithms---known as the \emph{plug-and-play property} \citep{He2010}---further expands the class of population models that can be confronted with data.
 
@@
 
-The theory presented here represents a strict generalization of the existing coalescent and birth-death process approaches to phylodynamic inference.
-Both of the latter processes represent special cases of the genealogical processes constructed here.
-In \cref{sec:app-examples}, this is demonstrated.
-
-A limitation of the theory presented here is that the population models are assumed to be pure jump processes, which allows consideration of demographic stochasticity and some kinds of environmental stochasticity.
+A limitation of the theory is that the population models are assumed to be pure jump processes, which allows consideration of demographic stochasticity and environmental stochasticity modeled by jumps involving multiple individuals \citep{Breto2011}, but disallows stochastic processes with a diffusive component.
 It should be possible to incorporate of the full range of Markovian environmental stochasticity via extension of this theory to population models containing both diffusion and jump components.
 
-The price of the theory's flexibility is primarily computational:
-Monte Carlo variability in the likelihood estimate depends on computational effort in the usual way.
-However, the substantial freedom one has in the choice of the importance-sampling distribution $\pi$ (\cref{thm:obsc_uncond}) can be exploited to diminish Monte Carlo error.
+The price of the theory's flexibility is primarily computational.
+When Sequential Monte Carlo is used to evaluate the likelihood in \cref{thm:obsc_uncond}, the computational effort scales linearly with the number of samples.
+In its most straightforward implementation---using an event-driven algorithm \citep[\eg][]{Gillespie1977a}---it scales nonlinearly with population size in general.
+However, stochastic simulation schemes are available that scale independently of population size \citep{Higham2008}.
+On the other hand, the importance sampling underlying \cref{thm:obsc_uncond} will in general require effort that is exponential in the number of demes.
+For models with many demes, therefore, approaches for ameliorating or circumventing this curse of dimensionality may be necessary.
+Critically, the substantial freedom one has in the choice of the importance-sampling distribution $\pi$ can be exploited for this purpose.
 In particular, since it is permissible to ``borrow information'' from the future by means of the importance sampling, there is hope for highly efficient algorithmic computation.
@@
     \label{fig:sirs_example}
-    Likelihood computation for the SIRS model by sequential Monte Carlo.
+    Likelihood computation for the SIRS model by Sequential Monte Carlo.
     (A)~A~simulated genealogy for $\beta=\Sexpr{sirs.params$Beta}$, $\gamma=\Sexpr{sirs.params$gamma}$, $\omega=\Sexpr{sirs.params$omega}$, $\psi=\Sexpr{sirs.params$psi}$, $(S_0,I_0,R_0)=(\Sexpr{with(sirs.params,c(S0,I0,R0))})$.
@@
     Each point is a distinct Monte Carlo estimate.
-    The blue curve is a LOESS smooth; the dashed lines show Monte Carlo-adjusted 95\% confidence intervals \citep{Ionides2017}.
+    The blue curve is a LOESS smooth;
+    the dashed lines bound the Monte Carlo-adjusted 95\% confidence interval \citep{Ionides2017}.
   }
@@
   A variety of importance-sampling kernels are permissible under the terms of \cref{thm:obsc_uncond}.
-  With a particular choice of importance-sampling kernel, the filter equation uniquely specifies a sequential Monte Carlo algorithm for estimating the likelihood.
+  With a particular choice of importance-sampling kernel, the filter equation uniquely specifies a Sequential Monte Carlo algorithm for estimating the likelihood.
   The choices made in \cref{tab:seirs_integ} underlie the results displayed in \cref{fig:seirs_example}.
@@
     \label{fig:seirs_example}
-    Likelihood computation for the SEIRS model by sequential Monte Carlo, using the scheme of \cref{box:SEIRS}.
+    Likelihood computation for the SEIRS model by Sequential Monte Carlo, using the scheme of \cref{box:SEIRS}.
     (A)~Simulated genealogy for $\beta=\Sexpr{seirs.params$Beta}$,
@@
     (B)~Likelihood slice in the $\sigma$-direction.
-    Each point represents the estimate of an independent sequential Monte Carlo computation.
+    Each point represents the estimate of an independent Sequential Monte Carlo computation.
     The blue curve shows a LOESS smooth;
@@
         nrow=1,
-        rel_widths=c(24,24,11),
+        rel_widths=c(24,24,17),
         C=stats |>
@@
     \label{fig:lbdp_example}
-    Likelihood computation for the constant-parameter, linear birth-death-sampling model, according to \cref{thm:obsc_uncond} via sequential Monte Carlo.
+    Likelihood computation for the constant-parameter, linear birth-death-sampling model, according to \cref{thm:obsc_uncond} via Sequential Monte Carlo.
     Panel (A) shows the genealogy, simulated for $\lambda=\Sexpr{lbdp.params$lambda}$, $\mu=\Sexpr{lbdp.params$mu}$, $\psi=\Sexpr{lbdp.params$psi}$, $N_0=\Sexpr{lbdp.params$n0}$.

### 2024-05-25 ms.R
add ms.R script

@@ -1,0 +1,442 @@
+## ----packages,include=FALSE,cache=FALSE---------------------------------------
+library(tidyverse)
+library(ggtree)
+library(pomp)
+library(cowplot)
+library(viridis)
+library(phylopomp)
+stopifnot(getRversion() >= "4.3")
+stopifnot(packageVersion("pomp")>="5.8")
+stopifnot(packageVersion("phylopomp")>="0.11")
+theme_set(theme_bw(base_family="serif"))
+options(
+  width=150,
+  keep.source=TRUE,
+  encoding="UTF-8",
+  stringsAsFactors=FALSE,
+  dplyr.summarise.inform=FALSE,
+  pomp_archive_dir="results"
+)
+set.seed(1159254136)
+
+
+## ----geneal,fig.dim=c(4,2),out.width="50%"------------------------------------
+freeze(
+  seed=382490723,
+  simulate(
+    "SEIR",
+    Beta=3,sigma=0.5,gamma=0.2,psi=0.3,omega=0.5,
+    S0=15,E0=1,I0=2,R0=0,
+    time=10
+  )
+) -> x
+
+pal <- c("#00274CFF","#FFCB05FF")
+
+x |> plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
+  geom_vline(xintercept=10,linewidth=0.2,color="black")
+
+
+## ----upo,fig.dim=c(4,6),out.width="50%"---------------------------------------
+freeze(
+  seed=522390503,
+  simulate(
+    "SEIR",
+    Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
+    S0=10,E0=1,I0=1,R0=0,
+    time=10
+  )
+) -> x
+
+pal <- c("#00274CFF","#FFCB05FF")
+
+plot_grid(
+  A=x |>
+    plot(
+      points=TRUE,prune=FALSE,obscure=FALSE,
+      ladderize=FALSE,palette=pal
+    )+
+    geom_vline(xintercept=10,linewidth=0.2,color="black"),
+  B=x |>
+    plot(
+      points=TRUE,prune=TRUE,obscure=FALSE,
+      ladderize=FALSE,palette=pal
+    )+
+    geom_vline(xintercept=10,linewidth=0.2,color="black"),
+  C=x |>
+    plot(
+      points=TRUE,prune=TRUE,obscure=TRUE,
[…]

### 2024-05-25 ms.Rnw
recompute to verify reproducibility

@@
 stopifnot(packageVersion("pomp")>="5.8")
-stopifnot(packageVersion("phylopomp")>="0.11")
+stopifnot(packageVersion("phylopomp")>="0.12.0.1")
 theme_set(theme_bw(base_family="serif"))
@@
 \end{document}
+
+<<sessioninfo,include=FALSE,purl=TRUE>>=
+sessionInfo()
+@

### 2024-05-25 ms.R

@@ -7,7 +7,7 @@ library(viridis)
 library(phylopomp)
 stopifnot(getRversion() >= "4.3")
 stopifnot(packageVersion("pomp")>="5.8")
-stopifnot(packageVersion("phylopomp")>="0.11")
+stopifnot(packageVersion("phylopomp")>="0.12.0.1")
 theme_set(theme_bw(base_family="serif"))
 options(
   width=150,
@@ -439,4 +439,8 @@ plot_grid(
       labs(x="effort",y=expression(group("|",bias,"|")))
   )
 )
+
+
+## ----sessioninfo,include=FALSE,purl=TRUE--------------------------------------
+sessionInfo()
 

### 2024-05-27 ms.Rnw
fix typos

@@
 The first is the \citet{Kingman1982a} coalescent, by which likelihood of a given genealogy is computed using a reverse-time argument.
-Thiis computation provides the exact likelihood for a genealogy resulting from a particular, constant population-size, dynamic model \citep[the Moran model, \eg][]{Moran1958,Kingman1982b,Moehle2000}.
+This computation provides the exact likelihood for a genealogy resulting from a particular, constant population-size, dynamic model \citep[the Moran model, \eg][]{Moran1958,Kingman1982b,Moehle2000}.
 Extensions of this approach develop approximate likelihoods for the case when the population size varies as a function of time \citep{Griffiths1994,Drummond2005} or according to an SIR process \citep{Volz2009a,Rasmussen2011}, as long as the population size is large and the sample-fraction remains negligible.
@@
 it retains information about where among the demes each of its lineages was through time (\cref{fig:upo}B).
-Note also that a pruned genealogy $P$ is characterized by its time, $\time(P)$ and the functions $P^{\lab{Y}}$ and $P^{\lab{Z}}$ just as an un-pruned genealogy is.
+Note also that a pruned genealogy $P$ is characterized by its time, $\time(P)$ and the functions $P^{\lab{Y}}$ and $P^{\lab{Z}}$ just as an unpruned genealogy is.
 Finally, observe that, since it contains within itself all of its past history, the pruned genealogy process $\Pr_t=\prune(\Gr_t)$ is Markov, even though the unpruned genealogy process, $\Gr_t$, is not.
@@
   Let us therefore suppose that neither of these conditions hold.
-  Conditional on $\Hr_\T=\H$, at each time $t\in\event{\H}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
+  Conditional on $\Hr_\T=\H$, at each time $t\in\event{\H}$, a jump of mark $\U_t$ occurred, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
   In $\P$, at time $t$, there are $\ell_i=\ell_i(\Y_t)$ lineages in deme $i$, of which $s_i=s_i(\Yt_t,\Y_t)$ are emergent.
@@
   \end{equation*}
-  Because our lineages are labelled, each of the $\tbinom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct;
+  Because our lineages are labeled, each of the $\tbinom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct;
   just one of these is the one present in $\P$.
@@
 Here $N=S+I+R$ is the host population size.
-Note that, though the theory allows for time-dependent event rates, this example is time-homogenous.
+Note that, though the theory allows for time-dependent event rates, this example is time-homogeneous.
 This model has one deme and occupancy function $n(x)=I$.

### 2024-05-28 ms.Rnw
more specificity on software versions

@@
 stopifnot(getRversion() >= "4.3")
-stopifnot(packageVersion("pomp")>="5.8")
-stopifnot(packageVersion("phylopomp")>="0.12.0.1")
+stopifnot(packageVersion("pomp")>="5.8.4")
+stopifnot(packageVersion("phylopomp")=="0.12.0.1")
 theme_set(theme_bw(base_family="serif"))

### 2024-05-28 ms.R

@@ -6,8 +6,8 @@ library(cowplot)
 library(viridis)
 library(phylopomp)
 stopifnot(getRversion() >= "4.3")
-stopifnot(packageVersion("pomp")>="5.8")
-stopifnot(packageVersion("phylopomp")>="0.12.0.1")
+stopifnot(packageVersion("pomp")>="5.8.4")
+stopifnot(packageVersion("phylopomp")=="0.12.0.1")
 theme_set(theme_bw(base_family="serif"))
 options(
   width=150,

### 2024-05-29 ms.Rnw
clean up some ggplot things

@@
         ## lims(y=c(max(params$logLik)-16,NA))+
-        theme_classic()+
-        theme(
-          legend.position=c(0.5,0.27)
-        ),
+        theme_classic(),
       labels="AUTO",
@@
             x=expression(mu)
-          )+
-          theme(
-            ## legend.position="inside",
-            ## legend.position.inside=c(0.5,0.4),
-            ## legend.background=element_rect(fill="white")
           )

### 2024-05-29 ms.R

@@ -171,10 +171,7 @@ plot_grid(
       x=expression(gamma)
     )+
     ## lims(y=c(max(params$logLik)-16,NA))+
-    theme_classic()+
-    theme(
-      legend.position=c(0.5,0.27)
-    ),
+    theme_classic(),
   labels="AUTO",
   nrow=1,
   rel_widths=c(1,1)
@@ -402,11 +399,6 @@ plot_grid(
         color="effort",
         y="log likelihood",
         x=expression(mu)
-      )+
-      theme(
-        ## legend.position="inside",
-        ## legend.position.inside=c(0.5,0.4),
-        ## legend.background=element_rect(fill="white")
       )
   ),
   CDE=plot_grid(

### 2024-06-06 ms.Rnw
update for new phylopomp version

@@
 library(phylopomp)
-stopifnot(getRversion() >= "4.3")
-stopifnot(packageVersion("pomp")>="5.8.4")
-stopifnot(packageVersion("phylopomp")=="0.12.0.1")
+stopifnot(getRversion() >= "4.4")
+stopifnot(packageVersion("pomp")>="5.9")
+stopifnot(packageVersion("phylopomp")>="0.13.2")
 theme_set(theme_bw(base_family="serif"))

### 2024-06-06 ms.R

@@ -5,9 +5,9 @@ library(pomp)
 library(cowplot)
 library(viridis)
 library(phylopomp)
-stopifnot(getRversion() >= "4.3")
-stopifnot(packageVersion("pomp")>="5.8.4")
-stopifnot(packageVersion("phylopomp")=="0.12.0.1")
+stopifnot(getRversion() >= "4.4")
+stopifnot(packageVersion("pomp")>="5.9")
+stopifnot(packageVersion("phylopomp")>="0.13.2")
 theme_set(theme_bw(base_family="serif"))
 options(
   width=150,

### 2024-06-07 ms.Rnw
improvements to examples

@@
   S0=97,I0=3,R0=0,t0=0,time=40
-) -> sirs.params
+) -> sirs_params
 
@@
   seed=328168304L,
-  dependson=sirs.params,
+  dependson=sirs_params,
   {
     library(phylopomp)
-    sirs.params |>
+    sirs_params |>
       with(
@@
 
+sirs_params |>
+  with(
+    expand_grid(
+      Beta=Beta,
+      gamma=seq(1.7,2.3,by=0.02),
+      psi=psi,
+      omega=omega,
+      S0=S0,I0=I0,R0=R0,
+      t0=t0,
+      rep=seq_len(8),
+      Np=10000,
+      )
+  ) -> params
+
 bake(
@@
   seed=621400057L,
-  dependson=list(sirs_tree,sirs.params),
+  dependson=list(sirs_tree,sirs_params,params),
   {
-    sirs.params |>
-      with(
-        expand_grid(
-          Beta=Beta,
-          ##      gamma=gamma,
-          gamma=seq(1.7,2.3,by=0.02),
-          psi=psi,
-          omega=omega,
-          ##      omega=seq(0.5,2,by=0.05),
-          S0=S0,I0=I0,R0=R0,
-          t0=t0,
-          rep=1:16,
-          Np=5000,
-          )
-      ) -> params
-
     library(iterators)
@@
     foreach (
-      p=iter(params,"row"),
-      .combine=bind_rows,
-      .options.future=list(seed=TRUE)
+      p=iter(params,"row")
     ) %dofuture% {
@@
             ) |>
-            pfilter(Np=Np) |>
-            logLik()
-        }) -> ll
-      bind_cols(p,logLik=ll)
-    } -> params
-  }
+            pfilter(Np=Np)
+        })
+    } %seed% TRUE |>
+      concat()
+  }) -> pfs
+
+left_join(
+  pfs |> coef() |> melt() |> pivot_wider(),
+  pfs |> logLik() |> melt() |> rename(logLik=value),
+  by=c(".id"="name")
 ) -> params
@@
         ggplot(aes(x=gamma,y=logLik))+
-        geom_point(alpha=0.2)+
+        geom_point(alpha=0.4)+
         geom_line(data=mcap$fit,aes(x=parameter,y=smoothed),color="blue")+
-        geom_vline(xintercept=sirs.params$gamma,color="red")+
+        geom_vline(xintercept=sirs_params$gamma,color="red")+
         geom_vline(xintercept=mcap$ci,linetype=2)+
@@
         )+
-        ## lims(y=c(max(params$logLik)-16,NA))+
+        lims(y=c(max(params$logLik)-12,NA))+
         theme_classic(),
@@
     Likelihood computation for the SIRS model by Sequential Monte Carlo.
-    (A)~A~simulated genealogy for $\beta=\Sexpr{sirs.params$Beta}$, $\gamma=\Sexpr{sirs.params$gamma}$, $\omega=\Sexpr{sirs.params$omega}$, $\psi=\Sexpr{sirs.params$psi}$, $(S_0,I_0,R_0)=(\Sexpr{with(sirs.params,c(S0,I0,R0))})$.
+    (A)~A~simulated genealogy for $\beta=\Sexpr{sirs_params$Beta}$, $\gamma=\Sexpr{sirs_params$gamma}$, $\omega=\Sexpr{sirs_params$omega}$, $\psi=\Sexpr{sirs_params$psi}$, $(S_0,I_0,R_0)=(\Sexpr{with(sirs_params,c(S0,I0,R0))})$.
     (B)~A~slice through the likelihood surface at the true parameters in the $\gamma$-direction.
@@
     & \multirow{4}{*}{$\lab{Trans}$}
-    & $(0,0)$ & 1 & $\left(\frac{E-\ell_E}{E}\right)\left(\frac{I-\ell_I}{I}\right)$ \bigstrut & \unbroken \bigstrut & $\frac{I+1-\ell_I}{I+1}$ & 1 & 1 \\
+    & $(0,0)$ & 1 & $\left(\frac{E-\ell_E}{E}\right)\left(\frac{I-\ell_I}{I}\right)$ \bigstrut & \unbroken \bigstrut & $\frac{I-\ell_I}{I}$ & 1 & 1 \\
     \cline{3-8}
-    && $(1,0)$ & 1 & $\frac{I-\ell_I}{E I}$ \bigstrut & \maizeblue \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 2 \\
+    && $(1,0)$ & 1 & $\frac{I-\ell_I}{E I}$ \bigstrut & \maizeblue \bigstrut & $\frac{1}{2I}$ & $\ell_I$ & 2 \\
     \cline{3-8}
-    && $(0,1)$ & 1 & $\frac{E-\ell_E}{E I}$ \bigstrut & \maizemaize \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 3 \\
+    && $(0,1)$ & 1 & $\frac{E-\ell_E}{E I}$ \bigstrut & \maizemaize \bigstrut & $\frac{1}{2I}$ & $\ell_I$ & 3 \\
     \cline{3-8}
@@
     &\multirow{2}{*}{$\lab{Prog}$}
-    & $(0,0)$ & 1 & $\frac{I-\ell_I}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \unbroken \bigstrut & $\frac{E+1-\ell_E}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & 1 & 5 \\
+    & $(0,0)$ & 1 & $\frac{I-\ell_I}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \unbroken \bigstrut & $\frac{E-\ell_E}{E}\,\Indicator{E>\ell_E}$ \bigstrut & 1 & 5 \\
     \cline{3-8}
-    && $(0,1)$ & 1 & $\frac{1}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \bluemaize \bigstrut & $\frac{1}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & $\ell_E$ & 6 \\
+    && $(0,1)$ & 1 & $\frac{1}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \bluemaize \bigstrut & $\frac{1}{E}\,\Indicator{E>\ell_E}$ \bigstrut & $\ell_E$ & 6 \\
     \cline{2-8}
@@
 <<seirs3>>=
-seirs.params <- data.frame(
-  Beta=4,sigma=1,gamma=1,psi=1,omega=1,
-  S0=200,E0=3,I0=5,R0=100,
-  time=3
+seirs_params <- data.frame(
+  Beta=3,sigma=1,gamma=0.5,psi=0.02,omega=0.08,
+  S0=70,E0=1,I0=0,R0=50,
+  time=400
 )
@@
   file="seirs3a.rds",
-  seed=831282841L,
-  dependson=seirs.params,
-  seirs.params |>
+  dependson=seirs_params,
+  seed=509673338,
+  seirs_params |>
     with(
@@
     )
-) -> seirs_tree
+)-> seirs_tree
+
+seirs_params |>
+  select(-sigma,-time) |>
+  expand_grid(
+    sigma=seq(0.4,2.4,length.out=25),
+    rep=seq_len(8)
+  ) |>
+  mutate(N=S0+E0+I0+R0) |>
+  collect() -> params
 
@@
   file="seirs3b.rds",
-  seed=831282841L,
-  dependson=list(seirs.params,seirs_tree),
+  dependson=list(params,seirs_params,seirs_tree),
+  seed=751601556,
   {
-    library(phylopomp)
-    library(circumstance)
-    library(doFuture)
-    plan(multicore)
-
-    seirs.params |>
+    seirs_params |>
       with(
@@
 
-    seq(0.2,2.5,by=0.1) |>
-      lapply(
-        \(s) {
-          x <- po
-          coef(x,"sigma") <- s
-          x
-        }
-      ) |>
-      concat() |>
-      pfilter(Np=20000,Nrep=10)
+    library(iterators)
+    library(doFuture)
+    plan(multicore)
+    foreach (
+      p=iter(params,"row")
+    ) %dofuture% {
+      library(phylopomp)
+      po |>
+        pfilter(params=p,Np=1e4)
+    } %seed% TRUE |>
+      concat()
   }
@@
       A=seirs_tree |>
-        plot(points=TRUE,palette="#000000")+
+        plot(points=FALSE,palette="#000000")+
         labs(x="time"),
@@
         geom_line(data=mcap$fit,aes(x=parameter,y=smoothed),color="blue")+
-        geom_vline(xintercept=seirs.params$sigma,color="red")+
+        geom_vline(xintercept=seirs_params$sigma,color="red")+
         geom_vline(xintercept=mcap$ci,linetype=2)+
         geom_hline(
-          yintercept=max(mcap$fit$smoothed)-c(0,mcap$delta),
+          yintercept=with(mcap,max(fit$smoothed)-c(0,delta)),
           linetype=2
         )+
-        ##        lims(y=c(max(params$logLik)-12,NA))+
+        lims(y=c(max(params$logLik)-12,NA))+
         labs(
@@
     Likelihood computation for the SEIRS model by Sequential Monte Carlo, using the scheme of \cref{box:SEIRS}.
-    (A)~Simulated genealogy for $\beta=\Sexpr{seirs.params$Beta}$,
-    $\sigma=\Sexpr{seirs.params$sigma}$,
-    $\gamma=\Sexpr{seirs.params$gamma}$,
-    $\psi=\Sexpr{seirs.params$psi}$,
-    $\omega=\Sexpr{seirs.params$omega}$,
-    $(S_0,E_0,I_0,R_0)=(\Sexpr{with(seirs.params,c(S0,E0,I0,R0))})$.
+    (A)~Simulated genealogy for $\beta=\Sexpr{seirs_params$Beta}$,
+    $\sigma=\Sexpr{seirs_params$sigma}$,
+    $\gamma=\Sexpr{seirs_params$gamma}$,
+    $\psi=\Sexpr{seirs_params$psi}$,
+    $\omega=\Sexpr{seirs_params$omega}$,
+    $(S_0,E_0,I_0,R_0)=(\Sexpr{with(seirs_params,c(S0,E0,I0,R0))})$.
     (B)~Likelihood slice in the $\sigma$-direction.

### 2024-06-07 ms.R

@@ -79,15 +79,15 @@ plot_grid(
 data.frame(
   Beta=4,gamma=2,psi=1,omega=1,
   S0=97,I0=3,R0=0,t0=0,time=40
-) -> sirs.params
+) -> sirs_params
 
 bake(
   file="sirs3a.rds",
   seed=328168304L,
-  dependson=sirs.params,
+  dependson=sirs_params,
   {
     library(phylopomp)
-    sirs.params |>
+    sirs_params |>
       with(
         runSIRS(
           Beta=Beta,gamma=gamma,psi=psi,omega=omega,
@@ -97,35 +97,31 @@ bake(
   }
 ) -> sirs_tree
 
+sirs_params |>
+  with(
+    expand_grid(
+      Beta=Beta,
+      gamma=seq(1.7,2.3,by=0.02),
+      psi=psi,
+      omega=omega,
+      S0=S0,I0=I0,R0=R0,
+      t0=t0,
+      rep=seq_len(8),
+      Np=10000,
+      )
+  ) -> params
+
 bake(
   file="sirs3b.rds",
   seed=621400057L,
-  dependson=list(sirs_tree,sirs.params),
+  dependson=list(sirs_tree,sirs_params,params),
   {
-    sirs.params |>
-      with(
-        expand_grid(
-          Beta=Beta,
-          ##      gamma=gamma,
-          gamma=seq(1.7,2.3,by=0.02),
-          psi=psi,
-          omega=omega,
-          ##      omega=seq(0.5,2,by=0.05),
-          S0=S0,I0=I0,R0=R0,
-          t0=t0,
-          rep=1:16,
-          Np=5000,
-          )
-      ) -> params
-
     library(iterators)
     library(doFuture)
     plan(multicore)
 
     foreach (
-      p=iter(params,"row"),
-      .combine=bind_rows,
-      .options.future=list(seed=TRUE)
+      p=iter(params,"row")
     ) %dofuture% {
       library(pomp)
       library(phylopomp)
@@ -136,12 +132,16 @@ bake(
               Beta=Beta,gamma=gamma,psi=psi,omega=omega,
               S0=S0,I0=I0,R0=R0,t0=0
             ) |>
-            pfilter(Np=Np) |>
-            logLik()
-        }) -> ll
-      bind_cols(p,logLik=ll)
-    } -> params
-  }
+            pfilter(Np=Np)
+        })
+    } %seed% TRUE |>
[…]

### 2024-06-12 ms.Rnw
phylopomp 0.14.2.3

@@
 stopifnot(getRversion() >= "4.4")
-stopifnot(packageVersion("pomp")>="5.9")
-stopifnot(packageVersion("phylopomp")>="0.13.2")
+stopifnot(packageVersion("pomp")>="5.9.1")
+stopifnot(packageVersion("phylopomp")>="0.14.2.3")
 theme_set(theme_bw(base_family="serif"))
@@
     <<geneal,fig.dim=c(4,2),out.width="50%">>=
-    freeze(
-      seed=382490723,
     simulate(
@@
       time=10
-      )
-    ) -> x
+    ) |>
+      freeze(seed=382490723) -> x
 
@@
   <<upo,fig.dim=c(4,6),out.width="50%">>=
-  freeze(
-    seed=522390503,
   simulate(
@@
     time=10
-    )
-  ) -> x
+  ) |>
+    freeze(seed=522390503) -> x
 

### 2024-06-12 ms.R

@@ -6,8 +6,8 @@ library(cowplot)
 library(viridis)
 library(phylopomp)
 stopifnot(getRversion() >= "4.4")
-stopifnot(packageVersion("pomp")>="5.9")
-stopifnot(packageVersion("phylopomp")>="0.13.2")
+stopifnot(packageVersion("pomp")>="5.9.1")
+stopifnot(packageVersion("phylopomp")>="0.14.2.3")
 theme_set(theme_bw(base_family="serif"))
 options(
   width=150,
@@ -21,15 +21,13 @@ set.seed(1159254136)
 
 
 ## ----geneal,fig.dim=c(4,2),out.width="50%"------------------------------------
-freeze(
-  seed=382490723,
-  simulate(
-    "SEIR",
-    Beta=3,sigma=0.5,gamma=0.2,psi=0.3,omega=0.5,
-    S0=15,E0=1,I0=2,R0=0,
-    time=10
-  )
-) -> x
+simulate(
+  "SEIR",
+  Beta=3,sigma=0.5,gamma=0.2,psi=0.3,omega=0.5,
+  S0=15,E0=1,I0=2,R0=0,
+  time=10
+) |>
+  freeze(seed=382490723) -> x
 
 pal <- c("#00274CFF","#FFCB05FF")
 
@@ -38,15 +36,13 @@ x |> plot(points=TRUE,prune=FALSE,obscure=FALSE,palette=pal)+
 
 
 ## ----upo,fig.dim=c(4,6),out.width="50%"---------------------------------------
-freeze(
-  seed=522390503,
-  simulate(
-    "SEIR",
-    Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
-    S0=10,E0=1,I0=1,R0=0,
-    time=10
-  )
-) -> x
+simulate(
+  "SEIR",
+  Beta=1,sigma=0.5,gamma=0.1,psi=0.4,omega=0.1,
+  S0=10,E0=1,I0=1,R0=0,
+  time=10
+) |>
+  freeze(seed=522390503) -> x
 
 pal <- c("#00274CFF","#FFCB05FF")
 

### 2024-11-12 ms.Rnw
some edits to introduction

@@
 This is simply the probability density of a given genealogy conditional on a given model, viewed as a function of the parameters of that model.
+%% Include other data, $Y$, in the expression?
 In particular, if $S$ is a set of genome sequences, $\Phi$ a genealogical tree relating these sequences, $E$ a model of sequence evolution, and $D$ a dynamic transmission model, then the likelihood is
@@
 Finally, because there is typically uncertainty not only in the parameters, but also in the structure, of a host-pathogen system, an improved phylodynamic inference methodology would place minimal restrictions on the form of the models that it can accommodate.
-This paper demonstrates how these desiderata can be achieved---at least for models with discrete structure---including arbitrary nonlinear compartmental models with a countable number of compartments.
-%% \ELI{[For the claim of generality, should we give a disclaimer that model complexity is limited in practice by both the limited amount of information in data to fit the model, and the numerical tractability of the integration.
-%%     Some idea of the largest size which has been tested to date (say, 5 compartments and 1000 sequences?) would be helpful to the reader.
-%%     Readers may naturally be skeptical about the practical scope of the methods.
-%%     For example, familiar Monte Carlo methods such as BEAST have theoretical guarantees going far beyond their practical applicability.]}
+This paper demonstrates how these desiderata can be achieved---at least for models with discrete structure---including arbitrary nonlinear compartmental models.
+
+Of course, practical considerations play an important role as well.
+In practice, data availability typically places strong limits on the degree of model complexity that can be supported by data.
+In addition, computational expense typically grows with model complexity and this can also limit the utility of otherwise attractive models and inference methods.
+Nevertheless, in the present paper we confine ourselves to theoretical considerations.
+The results we present could form the basis for a variety of distinct algorithms the relative value of which will depend on the questions asked, models proposed, and data available, and in any case remains to be seen.
+Moreover, although the theory we present is valid for models with even a countably infinite number of compartments, lack of data and computational resources will in practice require that the models that can effectively be employed may be much simpler than desired.
 
@@
 Code sufficient for the reproduction of all the results presented in this paper are freely available for download at \url{https://github.com/kingaa/structured-genealogy-process-paper}.
-An archival version of these will be stored on Zotero upon publication of this paper.
+An archival version of these will be stored on Zotero upon publication of a peer-reviewed version of this paper.
 The open-source \pkg{R} package \pkg{phylopomp} (\url{https://github.com/kingaa/phylopomp}) implements the simulation and likelihood-computation algorithms employed here.
@@
 Finally, note that $\Gr_t$ does not have the Markov property, though $(\Xr_t,\Ur_t,\Gr_t)$ and $(\Xr_t,\Gr_t)$ do.
+Observe in passing that, if instead of dropping tip nodes at death events we were to retain them as we do samples, the resulting genealogy---which we might call the ``complete'' genealogy---would have the Markov property.
 

### 2025-01-15 ms.Rnw
correct a typo; improve bibliography style re-run computations

@@
 
-\input{header}
+\input{ms_header}
 
@@
 stopifnot(getRversion() >= "4.4")
-stopifnot(packageVersion("pomp")>="5.9.1")
-stopifnot(packageVersion("phylopomp")>="0.14.2.3")
+stopifnot(packageVersion("pomp")>="6.1")
+stopifnot(packageVersion("phylopomp")>="0.14.8")
 theme_set(theme_bw(base_family="serif"))
@@
       -\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}},
-      &t\in{\event{\Z}},\\
+      &t\notin{\event{\Z}},\\
       &w(t,x,y)=\sum_{uy'}{\int{\wt(t,x',y')\,\beta_u(t,x',x,y',y)\,\varPsi_u(t,x',x,y',y)\,\dd{x'}}},

### 2025-01-15 ms.R

@@ -6,8 +6,8 @@ library(cowplot)
 library(viridis)
 library(phylopomp)
 stopifnot(getRversion() >= "4.4")
-stopifnot(packageVersion("pomp")>="5.9.1")
-stopifnot(packageVersion("phylopomp")>="0.14.2.3")
+stopifnot(packageVersion("pomp")>="6.1")
+stopifnot(packageVersion("phylopomp")>="0.14.8")
 theme_set(theme_bw(base_family="serif"))
 options(
   width=150,

### 2025-07-08 figs/markov_diagram.tex
start examples; more edits
arXiv revision 3

@@
       \node[state] (trans) at (165:8) {$x'=(S-1,E+1,I,R)$};
-      \node[state] (prog) at (205:7) {$x'=(S,E-1,I+1,R+1)$};
+      \node[state] (prog) at (205:7) {$x'=(S,E-1,I+1,R)$};
       \node[state] (recov) at (305:5) {$x'=(S,E,I-1,R+1)$};
@@
   Accordingly, the set, $\Jumps$, of jump marks has five elements.
-  Each of these is of a different type:
+  In the terminology of \cref{sec:event_types}, each of these is of a different type:
   $\lab{Trans}$ (transmission) is of birth type,
@@
   and $\lab{Wane}$ (loss or waning of immunity) is of neutral type.
-  See \cref{sec:jump_types} for a description of these jump types.
   Note that, in this formulation, when a sampling event occurs, the state does not change.

### 2025-07-08 ms.Rnw
arXiv revision 3

@@
 \begin{abstract}
-  We consider genealogies arising from a Markov population process in which individuals are categorized into a discrete collection of compartments, with the requirement that individuals within the same compartment are statistically exchangeable.
-  When equipped with a sampling process, each such population process induces a time-evolving tree-valued process defined as the genealogy of all sampled individuals.
-  We provide a construction of this genealogy process and derive exact expressions for the likelihood of an observed genealogy in terms of filter equations.
-  These filter equations can be numerically solved using standard Monte Carlo integration methods.
-  Thus, we obtain statistically efficient likelihood-based inference for essentially arbitrary compartment models based on an observed genealogy of individuals sampled from the population.
+  We show that each member of a broad class of Markovian population models induces a unique stochastic process on the space of genealogies.
+  We construct this genealogy process and derive exact expressions for the likelihood of an observed genealogy in terms of a filter equation, the structure of which is completely determined by the population model.
+  We show that existing phylodynamic methods based on either the coalescent or the linear birth-death processes are special cases.
+  We derive some properties of filter equations and describe a class of algorithms that can be used to numerically solve them.
+  Our results open the door to statistically efficient likelihood-based phylodynamic inference for a much wider class of models than is currently possible.
 \end{abstract}
@@
 In particular, one can formalize mathematical models of transmission, estimate their parameters, and compare their ability to explain data, following standard statistical paradigms.
-This is known as \emph{phylodynamic} inference.
-\citet{Alizon2024} gives a good review of the history of the subject.
-
-The most common approach to phylodynamic inference rests upon a mathematical linkage between the tree-like \emph{genealogy} or \emph{phylogeny} that expresses the relationships of shared ancestry among sampled genomes and a model of the dynamics of the transmission system.
-Various linkages are possible, but because it is maximally efficient (\ie loses the least information), it is desirable to be able to compute the likelihood function for models of interest.
-This is simply the probability density of a given genealogy conditional on a given model, viewed as a function of the parameters of that model.
-%% Include other data, $Y$, in the expression?
-In particular, if $S$ is a set of genome sequences, $\Phi$ a genealogical tree relating these sequences, $E$ a model of sequence evolution, and $D$ a dynamic transmission model, then the likelihood is
+This collection of tasks is known as \emph{phylodynamic} inference;
+\citet{Alizon2024} provides a recent review.
+
+A common approach to phylodynamic inference involves building a mathematical linkage between the tree-like \emph{genealogy} or \emph{phylogeny} that expresses the relationships of shared ancestry among sampled genomes and a model of the dynamics of the transmission system.
+Various linkages are possible, but, to attain maximal statistical efficiency (\ie lose the least information), it is desirable to be able to compute the likelihood function for models of interest.
+The likelihood function is the probability density of a given set of data, conditional on a given model, viewed as a function of the parameters of that model.
+While various kinds of data may be available, we focus here on the case where the data are a set of genomic sequences.
+Notationally, if the data, $S$, is a set of genome sequences, $\Phi$ a genealogical tree relating these sequences, $E$ a model of sequence evolution, and $D$ a dynamic transmission model, then the full likelihood is
 \begin{equation*}
@@
 In this expression, $f(S|\Phi,E)$ is typically the \citet{Felsenstein2004} phylogenetic likelihood.
-The function $f(\Phi|D)$, which links the phylogeny to the dynamic model, may be termed the \emph{phylodynamic likelihood}.
-In the Bayesian context, this same function is sometimes referred to as a \emph{tree prior} \citep{Moeller2018,Volz2018}.
-The computation of this function has remained out of reach, except in several special cases.
-This paper presents theory that enables its computation for a very broad range of dynamic models.
+The function $f(\Phi|D)$, which links the phylogeny to the dynamic model, is the \emph{phylodynamic likelihood}.
+%% Although we suppress the dependencies here, the likelihoods $\lik$, $f(S|\Phi,E)$, and $f(\Phi|D)$ are to be viewed as functions of the parameters that specify the models $D$ and $E$.
+In the Bayesian context, the phylodynamic likelihood is sometimes referred to as a \emph{tree prior} \citep{Moeller2018,Volz2018}.
+Its computation has remained out of reach, except in several special cases.
+This paper presents theory that facilitates its computation for a very broad range of dynamic models.
 
@@
 
-Existing approaches to the phylodynamic likelihood have been based on one of two mathematical idealizations.
-The first is the \citet{Kingman1982a} coalescent, by which likelihood of a given genealogy is computed using a reverse-time argument.
-This computation provides the exact likelihood for a genealogy resulting from a particular, constant population-size, dynamic model \citep[the Moran model, \eg][]{Moran1958,Kingman1982b,Moehle2000}.
+With few exceptions, existing approaches to the phylodynamic likelihood have been based on one of two mathematical idealizations.
+The first is the \citet{Kingman1982a} coalescent, by which the likelihood of a given genealogy is computed using a reverse-time argument.
+This computation provides the exact likelihood for a genealogy resulting from a particular, constant population-size, dynamic model \citep[the Moran model, \eg][]{Moran1958,Kingman1982b,Moehle2000,King2020,King2022}.
 Extensions of this approach develop approximate likelihoods for the case when the population size varies as a function of time \citep{Griffiths1994,Drummond2005} or according to an SIR process \citep{Volz2009a,Rasmussen2011}, as long as the population size is large and the sample-fraction remains negligible.
-The second idealization is the linear birth-death process, for which exact expressions for the likelihood are available \citep{Stadler2010}.
+The second idealization is the linear birth-death process, for which the likelihood is available in closed form \citep{Stadler2010}.
 Linearity in this context amounts to the assumption that distinct lineages do not interact:
@@
 Extensions of this approach develop approximations via linearization of nonlinear processes or restriction to scenarios in which population growth is nearly linear \citep[\eg][]{MacPherson2021}.
-Although the tractability of these approaches makes them attractive, concern naturally arises as to validity of the approximations in specific cases, the biases introduced by them, and the amount of information in data left unutilized by these approximate methods.
+Although the tractability of these approaches is appealing, concern naturally arises as to validity of the approximations in specific cases, the biases introduced by them, and the amount of information in data left uncaptured by these approximate methods.
 For this reason, there is interest in improved phylodynamic inference techniques.
 
-%% Factorization of problem into two subproblems.
-%% Problems with this approach \citep{Smith2017}.
-
 What would an ideal phylodynamic inference method look like?
-First, it would afford exact computation of the phylodynamic likelihood, so that comparisons among parameterizations and models could be made on a sound basis.
-Second, because nonlinearity, nonstationarity, noise, and measurement error are prominent and ubiquitous in epidemiology, it would accommodate nonlinear, time-inhomogeneous, stochastic transmission models.
-Third, because many of the most scientifically important uncertainties concern heterogeneities in transmission rates and the susceptibility, behavior, age, and location of hosts, it would accommodate host populations structured by these factors.
-While some structuring factors (\eg age, spatial location) are most naturally expressed in terms of continuous variables, discretely structured models have repeatedly proved their value in epidemiology.
-In particular, compartmental models are extremely flexible and have often been used as approximations when continuous structure leads to uncomfortably high model dimension.
-Finally, because there is typically uncertainty not only in the parameters, but also in the structure, of a host-pathogen system, an improved phylodynamic inference methodology would place minimal restrictions on the form of the models that it can accommodate.
-This paper demonstrates how these desiderata can be achieved---at least for models with discrete structure---including arbitrary nonlinear compartmental models.
-
-Of course, practical considerations play an important role as well.
-In practice, data availability typically places strong limits on the degree of model complexity that can be supported by data.
-In addition, computational expense typically grows with model complexity and this can also limit the utility of otherwise attractive models and inference methods.
-Nevertheless, in the present paper we confine ourselves to theoretical considerations.
-The results we present could form the basis for a variety of distinct algorithms the relative value of which will depend on the questions asked, models proposed, and data available, and in any case remains to be seen.
-Moreover, although the theory we present is valid for models with even a countably infinite number of compartments, lack of data and computational resources will in practice require that the models that can effectively be employed may be much simpler than desired.
+First, it would allow us to ask the questions we wish to ask.
+Since the set of scientifically interesting hypotheses is not contained within the set of statistically convenient models, an ideal phylodynamic inference methodology would put no restrictions on the form of the models that it can accommodate.
+%% Since the sets of scientific method can lead us to hypothesize models , an ideal phylodynamic inference methodology would put no restrictions on the form of the models that it can accommodate.
+In particular, since nonlinearity, nonstationarity, noise, and measurement error are prominent and ubiquitous in epidemiology, such a method would accommodate nonlinear, time-inhomogeneous, stochastic transmission models.
+Moreover, because many of the most scientifically and practically important uncertainties concern heterogeneities in transmission rates and the susceptibility, behavior, age, and location of hosts, it would accommodate host populations structured by these and other factors.
+Second, it would be both statistically efficient and robust to model misspecification.
+In particular, it would achieve maximal statistical efficiency by being based on the exact phylodynamic likelihood, which would also facilitate objective comparison among parameterizations and models.
+At the same time, model predictions would be robust to small deviations from the model assumptions.
+Finally, an ideal inference method would be computationally efficient.
+If numerical computations were absolutely required for its use, these would scale well with model complexity and data volume.
+
+Of course, there is no hope for the realization of such an ideal method.
+In particular, in selecting methodology, one must navigate trade-offs between statistical efficiency and robustness, and between computational expense and fidelity of model to question.
+With respect to these trade-offs, far more attention has been paid to date on relatively inexpensive approaches that are restricted to models that can only approximate the motivating questions to a greater or lesser extent.
+The present paper offers a complementary perspective:
+we develop mathematical expressions, and corresponding numerical algorithms, for the phylodynamic likelihood of almost arbitrarily complex models.
+Our focus on the exact likelihood offers maximum information capture, even as the computations it involves are somewhat delicate.
+Numerical examples are postponed to \cref{sec:app-examples}, and computational scalability is discussed in \cref{sec:discussion}.
+
+While the theory presented here greatly expands the class of models for which exact likelihood is computable, it is limited to models with \emph{discrete structure}.
+Thus, while structuring factors such as age, stage of infection progression, and spatial location are most naturally expressed in terms of continuous variables, to apply the results presented here, one would have to discretize these factors.
+We suspect that this limitation may often be relatively painless in practice, since discretely structured models have repeatedly proved their value in epidemiology.
+In particular, compartmental models offer great flexibility and have often been used as approximations when continuous structure leads to uncomfortably high model dimension.
 
 To connect a model at the level of a population with genealogies based on samples taken from individual hosts, it is necessary to make assumptions about the individuals in the population.
-The simplest such assumption is that the individuals that are identical with respect to the population dynamics are indeed statistically identical.
+The simplest such assumption is that the individuals that are identical with respect to the population dynamics are fully statistically identical.
 That is, that they are \emph{exchangeable}.
@@
 
-In the following, we take as our starting point a transmission model in the form of a discretely structured, Markov process.
-We show how such a process uniquely induces each of several stochastic processes in the space of genealogies.
-We go on to derive expressions for the exact likelihoods of these genealogies.
-
-Code sufficient for the reproduction of all the results presented in this paper are freely available for download at \url{https://github.com/kingaa/structured-genealogy-process-paper}.
+From the mathematical point of view, the essential difficulty in determining expressions for the likelihood of a genealogy generated by a given population process is the fact that a genealogy's structure is determined by events occurring at widely separated times.
+Existing approaches to phylodynamic inference employ one of a few tricks to get around this non-locality.
+In particular, both coalescent- and linear birth-death process-based approaches construct arguments in reverse time that depend on restrictions on the form of the population process.
+The main mathematical contribution of the present paper is to show how one can reformulate the phylodynamic likelihood in such a way that all needed computations are time-localized without placing restrictive assumptions on the population process.
+In particular, models for which forward simulation is possible but reversal of transition densities is intractable can be treated in this way.
+
+In the following, we take as our starting point a population model in the form of a discretely structured continuous-time Markov process.
+We show how such a process uniquely induces several stochastic processes in the space of genealogies.
+We go on to show that exact likelihoods for these genealogies can be computed by solving a filter equation.
+In particular, for each given population-level model there is a specific filter equation.
+We then show that our approach generalizes both the coalescent- and linear birth-death-process-based approaches.
+Indeed, the filter equations for the models on which these approaches are based can be solved to obtain the familiar likelihood formulae.
+In appendices, we develop the properties of filter equations generally, provide a sequential Monte Carlo algorithm that can be used when numerical integration is needed for their solution, and work out the filter equations and their solutions in two examples.
+
+Code sufficient for the reproduction of all the figures in this paper are freely available for download at \url{https://github.com/kingaa/structured-genealogy-process-paper}.
 An archival version of these will be stored on Zotero upon publication of a peer-reviewed version of this paper.
 The open-source \pkg{R} package \pkg{phylopomp} (\url{https://github.com/kingaa/phylopomp}) implements the simulation and likelihood-computation algorithms employed here.
-
-%% In particular, while the constructions of \citet{Etheridge2019} can in principle accommodate.
-%% For structured but deterministic models, approach of \citet{Volz2012,Rasmussen2014a}, based on approximation as birth-death processes:
-%% approximation no longer needed.
-
-%% Because there is commonly broad uncertainty regarding the structure of the transmission process, for example due to heterogeneities in transmission rates and susceptibilities, complex behavior patterns, etc., methods that have the plug-and-play property are particularly useful.
-%% Such techniques for inference based on time series are well understood and widely used.
-%% When the data lie in some Euclidean space, they can commonly be modeled as draws from some error distribution conditional on the latent state of the transmission process.
-%% However, when the data are genealogies (also called phylogenies) representing the relationships of shared ancestry among genomic samples, the data-space is non-Euclidean and the appropriate models connecting the data to the latent state are non-trivial.
-
-%% Modeling of the sampling process.
-
-%% Extension of previous results \citep{King2022}.
-%% Broader class of state-spaces.
-%% Accommodating discrete structure.
-
-%% In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
-%% \citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
-%% Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
-
-%% Classes of Markov processes.
-%% Utility and flexibility of Markov assumptions.
-
-%% Population process induces Markov history and genealogy processes.
-%% Using these, we derive equations for the likelihood of a genealogy conditional on the history.
-%% We then integrate out the history to obtain nonlinear filter equations, the solution of which yields the likelihood.
-%% These readily lend themselves to a family of Sequential Monte Carlo algorithms for computing the likelihood.
-%% We demonstrate with several examples.
 
@@
 In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $\Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
-In the case of continuous-time processes, we will assume that sample paths are \cadlag\ \ie right-continuous with left limits.
-We will frequently need to refer to the left-limit of such a process.
-Accordingly, if $\Phir_t$ is a \cadlag\ random process, we define
-\begin{equation*}
-  \Phirt_t\colonequals
-  \begin{cases}
-    \displaystyle\lim_{s\,\uparrow\,{t}}\;\Phir_{s}, & t>0,\\[2ex]
-    \Phir_0, & t=0.\\
-  \end{cases}
-\end{equation*}
-Note that $\Phirt_t$ is thus left-continuous with right limits.
+In the case of continuous-time processes, we will find ourselves needing to refer to their left- or right-limits.
+Accordingly, if $\Phir_t$ is any random process, let
+\begin{equation*}
+  \begin{gathered}
+    \Phirt_t\coloneq
+    \displaystyle\lim_{s\,\uparrow\,{t}}\;\Phir_{s}
+    \qquad\text{and}\qquad
+    \Philt_t\coloneq
+    \displaystyle\lim_{s\,\downarrow\,{t}}\;\Phir_{s}.
+  \end{gathered}
+\end{equation*}
+We will typically assume that the random processes treated here are right-continuous with left limits (\ie \cadlag, so that $\Philt_t=\Phir_t$).
+Note that, in this case, $\Phirt_t$ is left-continuous with right limits (\caglad).
 
@@
 In particular, we would like to be able to formulate and parameterize an arbitrary compartmental model and to quantify its ability to explain data using likelihood.
-\Cref{fig:example_models} depicts a few such models in order to give a sense of the kinds of complexities that can arise.
+\Cref{fig:example_models} depicts a few of the simplest such models in order to give a sense of the kinds of complexities that can arise.
 Of course, with the ability to entertain models with countably many compartments, much greater complexity is possible.
@@
 \begin{figure}
-  \input{figs/example_models}
+  \input{example_models}
   \caption{
@@
 
-We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
+We will refer to the stochastic process defined by the given model as the \emph{population process}.
+We will assume that the population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
 In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\mathbb{Z}^d$, but here we assume only that $\Xspace$ is a complete metric measure space with a countable dense subset.
@@
 \end{equation*}
-for all $t\in\Rp$ and $x,x'\in\Xspace$ and that, as a function of time, $\alpha$ is continuous almost everywhere.
+for all $t\in\Rp$ and $x,x'\in\Xspace$ and that, as a function of time, $\alpha$ is \cadlag\ and continuous almost everywhere.
 Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
@@
     \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
-    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}+o(\Delta).
+    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}\setminus\Set{x}}{\Xr_{t}=x}=\Delta\,\int_{\mathcal{E}\setminus\Set{x}}{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
+    %% \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}+o(\Delta).
+    \CondProb{\Xr_{t+\Delta}=x}{\Xr_{t}=x}=1-\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta).
   \end{gathered}
 \end{equation*}
-We further assume that $\alpha(t,x,x')$ is \cadlag\ as a function of time for all $x,x'\in{\Xspace}$ and that the number of jumps that occur in a finite time-interval is finite, \ie $\Prob{\Kr_t<\infty}=1$ for all $t$.
-
-\subsection{Kolmogorov forward equation}
+We further assume that $\Xr_t$ is \emph{non-explosive}, \ie that $\Prob{\Kr_t<\infty}=1$ for all $t<\infty$.
+
+\subsection{Kolmogorov equations}
 
@@
   \label{eq:kfe}
+  \begin{gathered}
     \frac{\partial{v}}{\partial{t}}(t,x)
@@
     -\int\!{v(t,x)\,\alpha(t,x,x')\,\dd{x'}},
-\end{equation}
-and if, moreover, $v(0,x) = p_0(x)$,
-then $\int_{\mathcal{E}}\!{v(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
+    \qquad
+    v(0,x) = p_0(x),
+  \end{gathered}
+\end{equation}
+for $t\in[0,T]$, then $\int_{\mathcal{E}}\!{v(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every $t\in[0,T]$ and measurable $\mathcal{E}\subseteq{\Xspace}$.
 \Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.
+The adjoint form of the KFE is the Kolmogorov backward equation,
+\begin{equation}
+  \label{eq:kbe}
+  \begin{gathered}
+    -\frac{\partial{F}}{\partial{s}}(s,x)
+    =\int\!{\alpha(t,x,x')\,\left[F(s,x')-F(s,x)\right]\dd{x'}},
+    \qquad
+    F(T,x) = f(x).
+  \end{gathered}
+\end{equation}
+If $F$ satisfies \cref{eq:kbe} for $s\in[0,T]$ and $x\in\Xspace$, then $F(s,x)=\CondExpect{f(\Xr_T)}{\Xr_s=x}$ for all $t\in[0,T]$ and $x\in\Xspace$.
 
@@
 
-For modeling purposes, it is sometimes desirable to insist that certain events occur at known times.
+For modeling purposes, it is sometimes desirable to insist that certain events occur at specified times.
 For example, if samples are collected at specific times in such a way that the timing itself conveys no information about the process, one might wish to condition on the sampling time.
 We can expand the class of population models to allow for this as follows.
-Suppose that $S=\Set{s_1,s_2,\dots,}\subset\Zp$ is a sequence of event times.
+Suppose that $S=\Set{s_1,s_2,\dots,}\subset\Rp$ is a sequence of times.
 Let us postulate that, at each of these times, an event occurs at which $\Xr_t$ jumps according to a given probability kernel $\pi$.
 In particular, for any state $x\in\Xspace$ and measurable $\mathcal{E}\subset\Xspace$,
-$\pi(s_i,x,\mathcal{E})$ is the probability that the jump at time $s_i$ is to $\mathcal{E}$, conditional on the state just before the jump being $x$.
+$\pi(s_i,x,\mathcal{E})$ is the probability that the jump at time $s_i$ is to $\mathcal{E}$, conditional on $\Xrt_{s_i}=x$.
 With this notation, the KFE for the process becomes
@@
 \end{align}
-Note that the \cref{eq:kfe-reg} is identical to \cref{eq:kfe};
+Note that \cref{eq:kfe-reg} is identical to \cref{eq:kfe};
 we call this the \emph{regular part} of the KFE.
 We refer to \cref{eq:kfe-sing} as the \emph{singular part} of the KFE.
+In this notation, $\pi(t,x',\dd{x})/\dd{x}$ is the density (\ie Radon-Nikodym derivative) of $\pi$ with respect to the base measure on $\Xspace$.
 
@@
 \begin{equation*}
-  \alpha(t,x,x')\mapsto\alpha(t,x,x')+\sum_{s\in{S}}{\delta(t,s)\,\frac{\dd{\pi}}{\dd{x'}}(t,x,x')},
+  \alpha(t,x,x')\,\dd{x'}\mapsto\alpha(t,x,x')\,\dd{x'}+\sum_{s\in{S}}{\delta(t,s)\,\pi(t,x,\dd{x'})},
 \end{equation*}
 we obtain an equation which we can view as shorthand for \cref{eq:kfe-reg,eq:kfe-sing}.
-Here, $\delta(t,s)$ is a Dirac delta function and $\dd{\pi}/\dd{x'}$ denotes the density (\ie Radon-Nikodym derivative) of $\pi$ with respect to the measure on $\Xspace$.
+Here, $\delta(t,s)$ is a one-sided Dirac delta function.
 %% To be even more parsimonious, we can refer to \cref{eq:kfe-reg,eq:kfe-sing} as the $(\alpha,\pi,S)$ KFE.
@@
 \begin{figure}
-  \input{figs/markov_diagram}
+  \input{markov_diagram}
 \end{figure}
 
-It will be useful to divide the jumps of the population process $\Xr_t$ into distinct categories, which differ with respect to the changes they induce in a genealogy.
+It will be useful to divide the jumps of the population process $\Xr_t$ into distinct categories which differ with respect to the changes they induce in a genealogy.
 For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
@@
 
-Let us define the \emph{jump mark} process, $\Ur_t$, to be the mark of the latest jump as of time $t$.
+Let us define the \emph{jump mark process}, $\Ur_t$, to be the mark of the latest jump as of time $t$.
 As usual, we take the sample paths of $\Ur_t$ to be \cadlag.
-Observe that, though $\Xr_t$ and $(\Xr_t,\Ur_t)$ are Markov processes, $\Ur_t$ is not.
+Observe that $\Xr_t$ and $(\Xr_t,\Ur_t)$ are Markov processes, but that $\Ur_t$ is not.
 
@@
 
-Our first goal in this paper is to show how a given population process induces a unique stochastic process on the space of genealogies.
+Our first goal in this paper is to show how a given population process induces a stochastic process on the space of genealogies.
 At each time, this genealogy will represent the relationships of shared ancestry among a population of lineages extant at that time.
@@
 In particular, we suppose that there are a countable set of subpopulations, within each of which individual lineages are exchangeable.
-We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.
-\Cref{fig:example_models} illustrates this concept in the context of several compartmental models.
+We call these subpopulations \emph{demes} and let $\Demes$ be an index set for them.
+We note that other authors have used different terminology for the same concept, including ``colony'', ``type'', ``state'', and ``population'' \citep{Takahata1988,Notohara1990,Boskova2014,Volz2018,BaridoSottani2020,Seidel2024,Vaughan2025}.
+\Cref{fig:example_models} illustrates this concept in the context of several specific models.
 
@@
 
-The class of population models to which the theory presented here applies is very broad indeed.
+The class of population models to which the theory presented here applies is very broad.
 In particular, it encompasses the entire class of compartmental models with time-dependent flow rates.
 Here, to give a sense of this breadth, we briefly describe a few models of interest.
-\Cref{sec:app-examples} works out the theory for each of these examples.
 
@@
 
-\citet{King2022} worked out formulas for the exact likelihood of a genealogy induced by an SIRS model.
-The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.
+\citet{King2022} developed formulae for the exact likelihood of a genealogy induced by an SIRS model.
+The theory developed in this paper applies, but since there is only one deme in this model, it is a simple case.
+In \cref{sec:app-examples}, we illustrate the theory by working it out in detail for this model.
 
@@
 
-A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
-The state space is $\Zp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
-It has two demes: $\Demes=\Set{\lab{E},\lab{I}}$.
+Modifying the SIRS model by incorporating an incubation period results in the SEIRS model, which has two demes (\cref{fig:example_models}A):
+$\Demes=\Set{\lab{E},\lab{I}}$.
+We can take the state space to be $\Zp^4$:
+the state $x=(S,E,I,R)$ is defined by the numbers of hosts in each of the four compartments.
 The deme occupancy function in this case is $n(x)=(E,I)$.
-Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.
+%% Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.
+In \cref{sec:app-examples}, we work out the theory for this model in detail.
 
@@
 
-A simple model for the competition of two strains for susceptible hosts is depicted in \cref{fig:example_models}B.
+A simple model for the competition of two pathogen strains is depicted in \cref{fig:example_models}B.
 In this model, the state vector consists of seven numbers: $x=(S,E_1,E_2,I_1,I_2,R_1,R_2)$.
@@
 
+\paragraph{Complex infection progression}
+
+\Cref{fig:example_models}C depicts a model in which infections are heterogeneous with respect to the severity of the disease they engender.
+There are three demes ($\Demes=\Set{\lab{E},\lab{I_A},\lab{I_S}}$).
+Asymptomatic infections ($\lab{I_A}$) are unobservable, while symptomatic infections ($\lab{I_S}$) may develop into hospitalized cases ($\lab{H}$) and deaths ($\lab{D}$).
+
 \paragraph{Superspreading model}
 
-\Cref{fig:example_models}D depicts a model of superspreading.
+Compartmental models can be used to capture heterogeneities of various kinds.
+\Cref{fig:example_models}D depicts a model with behavioral heterogeneity.
 There are three demes ($\Demes=\Set{\lab{E},\lab{I_L},\lab{I_H}}$).
-
-\paragraph{Linear birth-death model}
-
-The linear birth-death process, a mainstay of existing phylodynamic methods, is a special case of the theory presented here.
-For this process, we have $\Xspace=\Zp$ and there is a single deme.
-$\Xr_t$ represents the size of a population and $n(X_t)=X_t$.
-
-\paragraph{Moran model and the Kingman coalescent}
-
-The \citet{Kingman1982a} coalescent is another workhorse in existing phylodynamic approaches.
+Hosts in the low-transmitting $\lab{I_L}$ compartment have fewer contacts than those in the high-transmitting $\lab{I_H}$ compartment, and individuals move between these compartments as they engage in bouts of risky or cautious behavior.
+
+\paragraph{Kingman coalescent and Moran model}
+
+The \citet{Kingman1982a} coalescent is one of the two foundations upon which most existing phylodynamic approaches have been constructed.
 It is the ancestral process for the Moran model, in which a fixed population of $n$ lineages experiences events at times distributed according to a rate-$\mu$ Poisson process.
 At each such event, an individual lineage selected uniformly at random dies and is replaced by the offspring of a second randomly selected lineage.
-
-\subsection{History}
+In \cref{sec:kingman}, we show that the Kingman coalescent likelihood is a special case of the theory presented in this paper.
+
+\paragraph{Linear birth-death model}
+
+The linear birth-death process, the second basis for widely-used phylodynamic methods, is also a special case of the theory presented here.
+For this process, we have $\Xspace=\Zp$ and there is a single deme.
+$\Xr_t$ represents the size of a population and $n(x)=x$.
+In \cref{sec:lbdp}, we derive exact expressions for the likelihood under this model.
+
+\subsection{History process}
 
@@
 Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.
-
-Alternatively, one can think of $\Hr_t$ as consisting of the sequence
+Alternatively, one can identify $\Hr_t$ with the sequence
 $\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
@@
 
-If $\H$ is such a history, we define $\time(\H)$ to be the right endpoint of its domain and use the notation $\event{\H}\coloneq\Set{\Th^{}_1,\dots,\Th^{}_{\K_t}}\subset{[0,\time(\H)]}$ to denote the set of its jump times.
+If $\H$ is a potential value of a history process, $\Hr_t$, we define $\time(\H)$ to be the right endpoint of its domain and use the notation $\event{\H}\coloneq\Set{\Th^{}_1,\dots,\Th^{}_{\K_t}}\subset{[0,\time(\H)]}$ to denote the set of its jump times.
+
+\section{Genealogy processes}
 
@@
 
-\begin{figure}
+A \emph{genealogy}, $G$, encapsulates the relationships of shared ancestry among a set of lineages that are extant at some time $\time(G)\in\Rp$ and perhaps a set of samples collected at earlier times (\cref{fig:geneal}).
+A genealogy has a tree- or forest-like structure, with four distinct kinds of nodes:
+\begin{inparaenum}[(i)]
+\item \emph{tip nodes}, which represent labeled extant lineages;
+\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another;
+\item \emph{sample nodes}, which represent labeled samples; and
+\item \emph{root nodes}, at the base of each tree.
+\end{inparaenum}
+Each node $a$ is associated with a specific time, $\time(a)$.
+In particular, if $a$ is a tip node in $G$, then $\time(a)=\time(G)$;
+if $a$ is a sample node, then $\time(a)\le{\time(G)}$ is the time at which the sample was taken.
+Moreover, if node $a$ is ancestral to node $a'$, then $\time(a)\le{\time(a')}$ and $\time(a')-\time(a)$ is the distance between $a$ and $a'$ along the genealogy.
+Without loss of generality we assume that $\time(a)=0$ for all root nodes $a$.
+We let $\event{G}$ denote the set of all internal and sample node-times of the genealogy $G$;
+we refer to these as \emph{genealogical event times}.
+
+Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
+Accordingly, we can visualize a genealogy as a graph, the nodes and edges of which are painted with a distinct color for each deme (\cref{fig:geneal}).
+Note that a genealogy will in general have \emph{branch-point nodes}, \ie internal nodes with more than one descendant, but may also have internal nodes with only one descendant.
+We refer to such nodes as \emph{inline nodes}.
+These occur whenever the color changes along a branch, but can also occur without a color-change.
+
+\begin{figure}[t]
   \begin{center}
@@
 
-A \emph{genealogy}, $G$, encapsulates the relationships of shared ancestry among a set of lineages that are extant at some time $\time(G)\in\Rp$ and perhaps a set of samples collected at earlier times (\cref{fig:geneal}).
-A genealogy has a tree- or forest-like structure, with four distinct kinds of nodes:
-\begin{inparaenum}[(i)]
-\item \emph{tip nodes}, which represent labeled extant lineages;
-\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another;
-\item \emph{sample nodes}, which represent labeled samples; and
-\item \emph{root nodes}, at the base of each tree.
-\end{inparaenum}
-Each node $a$ is associated with a specific time, $\time(a)$.
-In particular, if $a$ is a tip node in $G$, then $\time(a)=\time(G)$;
-if $a$ is a sample node, then $\time(a)\le{\time(G)}$ is the time at which the sample was taken.
-Moreover, if node $a$ is ancestral to node $a'$, then $\time(a)\le{\time(a')}$ and $\time(a')-\time(a)$ is the distance between $a$ and $a'$ along the genealogy.
-Without loss of generality we assume that $\time(a)=0$ for all root nodes $a$.
-We let $\event{G}$ denote the set of all internal and sample node-times of the genealogy $G$;
-we refer to these as \emph{genealogical event times}.
-
-Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
-Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted with a distinct color for each deme (\cref{fig:geneal}).
-Note that a genealogy will in general have \emph{branch-point nodes}, \ie internal nodes with more than one descendant, but may also have internal nodes with only one descendant.
-We refer to such nodes as \emph{inline nodes}.
-These occur whenever the color changes along a branch, but can also occur without a color-change.
-
 Formally, we define a genealogy, $G$, to be a triple, $(T,Z,Y)$, where $T=\time(G)\in\Rp$ is the \emph{genealogy time}, $Z$ specifies the genealogy's \emph{tree structure}, and $Y$ gives the \emph{coloring}.
-In particular, let $\leaves$ be a countable set of labels and let $\part(\leaves)$ be the set of all collections of finite, mutually-disjoint subsets of $\leaves$.
-That is, an element $z\in{\part(\leaves)}$ is a partition of the finite set $\bigcup{z}\subseteq\leaves$.
-Partition \emph{fineness} defines a partial order on $\part(\leaves)$.
-Specifically, for $z,z'\in{\part(\leaves)}$, we say $z\preceq{z'}$ if and only if for every $b'\in{z'}$ there is $b\in{z}$ such that $b\supseteq{b'}$.
-The tree structure of $G$ is defined by a \cadlag\ map $Z:[0,T]\to\part(\leaves)$ that is monotone in the sense that $t_1\le{t_2}$ implies $Z_{t_1}\preceq{Z_{t_2}}$.
+In particular, let $\leaves$ be a countable set of labels---referring to samples and/or extant lineages---and let $\partit(\leaves)$ be the set of all collections of finite, mutually-disjoint subsets of $\leaves$.
+That is, an element $z\in{\partit(\leaves)}$ is a partition of the finite set $\bigcup{z}\subseteq\leaves$.
+Partition \emph{fineness} defines a partial order on $\partit(\leaves)$.
+Specifically, for $z,z'\in{\partit(\leaves)}$, we say $z\preceq{z'}$ if and only if for every $b'\in{z'}$ there is $b\in{z}$ such that $b\supseteq{b'}$.
+The tree structure of $G$ is a \cadlag\ map $Z:[0,T]\to\partit(\leaves)$ that is monotone in the sense that $t_1\le{t_2}$ implies $Z_{t_1}\preceq{Z_{t_2}}$.
 An element $b\in{Z_t}$ is a set of labels;
@@
 We use the notation $\event{Z}$ to denote the set of times at which $Z$ is discontinuous.
-Note that $\event{Z}$ includes the times of all tip, sample, and branch-point nodes, but excludes inline and root nodes.
+Note that $\event{Z}$ includes the times of all tip, sample, and branch-point nodes, but excludes root and non-sample inline nodes.
 Therefore, $\event{Z}\subseteq{\event{G}}$.
@@
 
-It will sometimes be convenient to make use of notation whereby a genealogy $G=(\time(G),G^{\lab{Z}},G^{\lab{Y}})$.
-
-\subsection{Binomial ratio}
-\label{sec:binomial_ratio}
-
-For $n,r,\ell,s\in{\Zp^\Demes}$, define the \emph{binomial ratio}
-\begin{equation*}
-  \BinRatio{n}{\ell}{r}{s}\colonequals
-  \begin{cases}
-    \frac{\displaystyle\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}%
-         {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
-         & \text{if}\ \forall i\ n_i\ge{\Set{\ell_i,r_i}}\ge{s_i}\ge{0},\\[7ex]
-         0, & \text{otherwise}.
-  \end{cases}
-\end{equation*}
-Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
-Moreover, in consequence of the Chu-Vandermonde identity, we have
-\begin{equation*}
-  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
-\end{equation*}
-whenever $n_i\ge{\Set{\ell_i,r_i}}\ge{0}$ for all $i$.
-
-\section{The induced genealogy process}
+It will sometimes be convenient to make use of notation whereby the tree structure of genealogy $G$ is $G^{\lab{Z}}$ and the coloring is $G^{\lab{Y}}$, so that $G=(\time(G),G^{\lab{Z}},G^{\lab{Y}})$.
 
@@
 
-We now show how a given population process naturally induces a process in the space of genealogies.
-Specifically, at each jump in the population process, a corresponding change occurs in the genealogy, according to whether lineages branch, die, move between demes, or are sampled.
+To see how events in the population process leave a trace in the genealogy, we assume that, at each jump in the population process, a corresponding change occurs in the genealogy, according to whether lineages branch, die, move between demes, or are sampled.
 For this purpose, there are five distinct \emph{pure types} of events:
 \begin{compactenum}[(a)]
-\item \emph{Birth-type events} result in the branching of one or more new lineages, each from some existing lineage.
+\item \emph{Birth-type events} result in the branching of one or more new lineages from an existing lineage.
   Examples of birth-type events include transmission events, speciations, and actual births.
   Importantly, we assume that all new lineages arising from a birth event share the same parent and that at most one birth event occurs at a time, almost surely.
+  While it is possible to relax this assumption, the resulting notational complexities are nontrivial, so we postpone consideration of this more general case.
 \item \emph{Death-type events} result in the extinction of one or more lineages.
   Examples include recovery from infection, death of a host, and species extinctions.
-  We allow for the possibility that multiple lineages die simultaneously.
+  We allow for the possibility that multiple lineages, potentially in multiple demes, die simultaneously.
 \item \emph{Migration-type events} result in the movement of a lineage from one deme to another.
   Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
-  We permit multiple lineages to move simultaneously.
+  In a migration-type event, one or more lineages within exactly one deme may move simultaneously, and they may move to different demes.
+  Again, while it possible to relax the assumption that exactly one deme is the source of all migrating lineages within any one event, we postpone consideration of this case for the present.
 \item \emph{Sample-type events} result in the collection of a sample from a lineage.
-  We allow for the possibility that multiple samples are collected simultaneously, though we require that, in this case, each extant lineage is sampled at most once.
+  We allow for the possibility that multiple samples are collected simultaneously, though we require that, in this case, each extant lineage is sampled at most once and that all sampled lineages be in the same deme.
+  It is possible to allow for simultaneous sampling of multiple demes but, again, we postpone consideration of this case to ease the exposition.
 \item \emph{Neutral-type events} result in no change to any of the lineages.
@@
 For example, a sample/death-type event, in which a lineage is simultaneously sampled and removed, has been employed \citep{Leventhal2014}, as have birth/death events in which one lineage reproduces at the same moment that another dies (\eg the \citet{Moran1958} process).
-The theory presented here places few restrictions on the complexity of the events that can occur by combining events of the various pure types.
-
-%% Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents and demes, there is implicitly a deterministic indicator function $Q_u$, for $u\in\Jumps$, (described below) that captures these properties.
+%% There are restrictions on the parentage within an event.
+%% In particular, in Theorem 1, we assume that, given $u$, the parent deme is deterministic.
+In combining the pure types, the main restriction is that, at any compound event, all parent lineages must be drawn from the same deme.
+Beyond this, the theory presented here places few restrictions on the complexity of the events that can occur by combining events of the various pure types.
 
@@
 \begin{figure}
-  \input{figs/event_types}
+  \input{event_types}
   \caption{
@@
   These are deleted.
-  Next, internal nodes without children are recursively removed.
+  Next, non-sample internal nodes without children are recursively removed.
   Sample nodes are never removed.
 \item
-  At a migration-type event (\cref{fig:event_types}C), the appropriate number of migrating lineages are selected at random with uniform probability, from among those present in the appropriate demes.
+  At a migration-type event (\cref{fig:event_types}C), the appropriate number of migrating lineages are selected at random with uniform probability, from among those present in the appropriate deme.
   For each selected lineage, one new branch node is inserted between the selected tip node and its ancestor.
@@
   One new sample node is introduced for each selected lineage:
-  each is inserted between a selected tip nodes and its ancestor.
+  each is inserted between a selected tip node and its ancestor.
 \item
@@
 
-The lineages which descend from an inserted node are said to \emph{emerge} from the event.
+At each event, the lineages which descend from a node inserted at that event are said to \emph{emerge} from the event.
 Thus, after a birth-type event, the emerging lineages include all the new offspring as well as the parent.
@@
 In particular, at each jump-time $\Trh_k$, the genealogy $\Grh_{k-1}$ is modified according to the jump-mark $\Urh_k$ to yield $\Grh_k$.
-We view $\Grh_k$ as the embedded chain of the continuous-time genealogy process $\Gr_t$.
+We view $\Grh_k$, $k=0,1,2,\dots$, as the embedded chain of the continuous-time genealogy process $\Gr_t$.
 It is very important to note that, conditional on $(\Xrh_k,\Urh_k)$, the number of parents and number of offspring in each deme is determined and the random choice of which lineages die, migrate, are sampled, or sire offspring is independent of these choices at any other times and independent of $(\Xrh_j,\Urh_j)$ for all $j\ne{k}$.
@@
 Finally, note that $\Gr_t$ does not have the Markov property, though $(\Xr_t,\Ur_t,\Gr_t)$ and $(\Xr_t,\Gr_t)$ do.
-Observe in passing that, if instead of dropping tip nodes at death events we were to retain them as we do samples, the resulting genealogy---which we might call the ``complete'' genealogy---would have the Markov property.
+In passing, we observe that, if instead of dropping tip nodes at death events we were to retain them as we do samples, the resulting genealogy---which we might call the \emph{complete genealogy}---would have the Markov property.
 
@@
 Note also that a pruned genealogy $P$ is characterized by its time, $\time(P)$ and the functions $P^{\lab{Y}}$ and $P^{\lab{Z}}$ just as an unpruned genealogy is.
-Finally, observe that, since it contains within itself all of its past history, the pruned genealogy process $\Pr_t=\prune(\Gr_t)$ is Markov, even though the unpruned genealogy process, $\Gr_t$, is not.
+Crucially, since it contains within itself all of its past history, the pruned genealogy process $\Pr_t=\prune(\Gr_t)$ is Markov, even though the unpruned genealogy process, $\Gr_t$, is not.
 
@@
 \begin{figure}
-  \input{figs/ells}
+  \input{ells}
   \caption{
@@
 
-Suppose $P$ is a pruned genealogy, with $\time(P)=T$ and $t\in\event{P}$.
+Suppose $P$ is a pruned genealogy, with $\time(P)=T$ and $t\in{[0,T]}$.
 The local structure of $P$ at $t$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
-For example, if the event in $P$ at $t$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
-Similarly, if the node in $P$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
-To succinctly accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $P^{\lab{Y}}$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
+For example, if there is a branch node or a sample node at $t$, then it is compatible only with birth-type or sample-type jumps, respectively.
+Similarly, if a lineage moves from deme $i$ to deme $i'$ at $t$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
+To succinctly accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $P^{\lab{Y}}$ just before and at $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
 That is, $Q_u(y,y')=1$ if and only if
@@
 
+\subsection{Binomial ratio}
+\label{sec:binomial_ratio}
+
+The statement of the theorems to come is made easier with the following definition.
+For $n,r,\ell,s\in{\Zp^\Demes}$,
+we define the \emph{binomial ratio}
+\begin{equation*}
+  \BinRatio{n}{\ell}{r}{s}\coloneq
+  \begin{cases}
+    \frac{\displaystyle\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}%
+         {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
+         & \text{if}\ \forall i\ n_i\ge{\Set{\ell_i,r_i}}\ge{s_i}\ge{0},\\[7ex]
+         0, & \text{otherwise}.
+  \end{cases}
+\end{equation*}
+where $\tbinom{a}{b}=\tfrac{a!}{b!\,(a-b)!}$ is the binomial coefficient.
+Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
+Moreover, in consequence of the Chu-Vandermonde identity, we have
+\begin{equation*}
+  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
+\end{equation*}
+whenever $n_i\ge{\Set{\ell_i,r_i}}\ge{0}$ for all $i$.
+
 \section{Results}
@@
 
-\begin{thm}\label{thm:pruned_lik}
+\begin{thm}
+  \label{thm:pruned_lik}
   Suppose $\P=(\T,\Z,\Y)$ is a given pruned genealogy.
@@
 \end{thm}
-\begin{proof}
-  If $\event{\H}\nsupseteq\event{\P}$, then $\H$ and $\P$ are incompatible and $\CondProb{\Pr_\T=\P}{\Hr_\T=\H}=0$.
-  Similarly, if any event of $\H$ is incompatible with the local structure of $\P$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_\T=\P}{\Hr_\T=\H}=0$.
-  Let us therefore suppose that neither of these conditions hold.
-  Conditional on $\Hr_\T=\H$, at each time $t\in\event{\H}$, a jump of mark $\U_t$ occurred, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
-  In $\P$, at time $t$, there are $\ell_i=\ell_i(\Y_t)$ lineages in deme $i$, of which $s_i=s_i(\Yt_t,\Y_t)$ are emergent.
-  By assumption, at each genealogical event, lineages within a deme are exchangeable:
-  each has an identical probability of being involved.
-  This exchangeability implies that each lineage present in a deme at time $t$ was equally likely to have been one of the emergent lineages.
-  In particular, at time $t$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the unpruned genealogy process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ red balls and $n_i-r_i$ black balls, exactly $s_i$ of the drawn balls are red, namely
-  \begin{equation*}
-    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
-  \end{equation*}
-  Because our lineages are labeled, each of the $\tbinom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct;
-  just one of these is the one present in $\P$.
-  Moreover, since, again conditional on $\Hr_\T=\H$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
-  \begin{equation*}
-    \CondProb{\Pr_\T=\P}{\Hr_\T=\H}=\prod_{t\in\event{\H}}{\BinRatio{n(\X_t)}{\ell(\Y_t)}{r^{\U_t}}{s(\Yt_t,\Y_t)}}.
-  \end{equation*}
-  Returning to the possibility that $\H$ is incompatible with $\P$, since $\Prob{\Pr_\T=\P}=0$ if either any $Q^{}_{\U_t}=0$ or $\event{\P}\nsubseteq\event{\H}$, we obtain the result.
-\end{proof}
-
-Next, we show how the likelihood of a pruned genealogies, unconditional on the history, can be computed.
+
+The proof is given in \cref{sec:proof}.
+
+Next, we show how the likelihood of a pruned genealogy, unconditional on the history, can be computed.
 For this, we use the filter equation technology developed in \cref{sec:filter_eqns}.
-In particular, the following theorem follows immediately from \cref{lemma:sing-filt}.
+In particular, the following theorem follows immediately from \cref{thm:pruned_lik,lemma:sing-filt}.
 
 \begin{thm}
-  \label{thm:pruned_uncond}
+  \label{thm:pruned_filter}
   Suppose that $\P=(\T,\Z,\Y)$ is a given pruned genealogy.
-  Suppose that $w=w(t,x)$ satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
+  Suppose that $w=w(t,x)$ is \cadlag\ in $t$ and satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
   \begin{equation}
+    \label{eq:pruned_filter_eqn}
     \begin{aligned}
@@
   \begin{equation*}
-    \lik(\P)=\int{w(T,x)\,\dd{x}}.
+    \lik=\int{w(T,x)\,\dd{x}}.
   \end{equation*}
@@
 
+Alternatively, one can express the likelihood of a pruned genealogy using the adjoint form of the filter equation (see \cref{sec:adjoint_filter}).
+
+\begin{corol}
+  \label[corol]{corol:pruned_filter}
+  If $\P=(\T,\Z,\Y)$ is a given pruned genealogy and $F=F(s,x)$ is \caglad\ in $s$ and satisfies the final condition $\rightlim{F}(\T,x)=1$ and the adjoint filter equation
+  \begin{equation}
+    \label{eq:pruned_adjoint_eqn}
+    \begin{aligned}
+      -\frac{\partial{F}}{\partial{s}}(s,x)=
+      &\sum_u{\int{\leftlim{\alpha}_u(s,x,x')\,\left[\phi^{}_u(x',\Yt_s,\Y_s)\,F(s,x')-F(s,x)\right]\,\dd{x'}}},
+      &s\notin{\event{\P}},\\
+      F(s,x)=&\sum_u{\int{\alpha_u(s,x,x')\,\phi^{}_u(x',\Yt_s,\Y_s)\,\rightlim{F}(s,x')\dd{x'}}},
+      &s\in{\event{\P}},
+    \end{aligned}
+  \end{equation}
+  where $0\le{s}\le{T}$ and $\phi$ is defined in \cref{eq:phidef},
+  then the likelihood of $\P$ is
+  \begin{equation*}
+    \lik=\int{F(0,x)\,p_0(x)\,\dd{x}}.
+  \end{equation*}
+\end{corol}
+
+\begin{remark}
+  \label{rem:ordering}
+  Note that the likelihood given by the foregoing expressions is the likelihood of a given genealogy \emph{in which the samples are ordered} and that, while this order should be compatible with time-ordering, it may not be completely determined by it.
+  In particular, if $n$ samples are taken at one time, there are $n!$ possible orderings of the samples, each of which has an identical likelihood.
+  In consequence, the likelihood given above will differ by a factor of $n!$ from the likelihood computed ignoring ordering of samples.
+  From the standpoint of parameter estimation, this is immaterial, since factors that depend only on the data drop out of the score function.
+  This consideration is relevant, however, in the context of model comparison.
+  %% In effect, we are assuming that the ordering of samples is part of the data.
+\end{remark}
+
 \subsection{Likelihood for obscured genealogies}
@@
 
-\begin{thm}\label{thm:obsc_lik}
+\begin{thm}
+  \label{thm:obsc_lik}
   Suppose that $(\T,\Z)$ is a given obscured genealogy.
@@
   \end{equation}
-  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
+  Our strategy will be to evaluate \cref{eq:IS1} using a change of measure:
   we will propose pruned genealogies compatible with $\Z$ as sample paths from a stochastic process driven by $\Xr_t$ and
-  evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
+  evaluate the the expectation in \cref{eq:IS1} by a suitably weighted expectation over these paths.
   Conditional on $\Hr_\T=\H$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$ such that
@@
 \begin{thm}
-  \label{thm:obsc_uncond}
+  \label{thm:obsc_filter}
   Let $V=(\T,\Z)$ be a given obscured genealogy.
@@
   \begin{equation*}
-    \lik(V)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
+    \lik=\sum_{y\in\func{Y}_T(\Z)}{\int{w(T,x,y)\,\dd{x}}}.
   \end{equation*}
@@
 
-\Cref{lemma:monte_carlo} shows how this can be computed via Sequential Monte Carlo.
-
-\section{Discussion}
-
-The theory presented here represents a strict generalization of the existing coalescent and birth-death process approaches to phylodynamic inference.
-In \cref{sec:app-examples}, we demonstrate that both of the latter processes are special cases of the genealogical processes constructed here.
-Importantly, because the theory allows computation of the likelihood via strictly forward-in-time computations, it permits consideration of models for which time-reversal arguments are not available.
-Moreover, inasmuch as the formulae of \cref{thm:obsc_uncond} can be efficiently computed via sequential Monte Carlo, explicit expressions for transition probabilities are not needed:
+\Cref{lemma:monte_carlo} shows how this can be computed via sequential Monte Carlo.
+As before, there is also an adjoint form of \cref{thm:obsc_filter}, to wit:
+
+\begin{corol}
+  \label[corol]{corol:obsc_filter}
+  Let $V=(\T,\Z)$ be a given obscured genealogy and let $\beta$ and $\varPsi$ be defined as in \cref{thm:obsc_filter}.
+  Suppose $F$ is \caglad\ and satisfies the final condition
+  $\rightlim{F}(T,x,y)=1$ for all $x\in\Xspace$, $y\in\func{Y}_T(\Z)$ and also the adjoint filter equation
+  \begin{equation*}
+    \begin{aligned}
+      -&\frac{\partial{F}}{\partial{s}}=
+      \sum_{uy'}{\int{\leftlim{\beta}_u(s,x,x',y,y')\,\left[\leftlim{\varPsi}_u(s,x,x',y,y')\,F(s,x',y')-F(s,x,y)\right]\,\dd{x'}}},
+      &s\notin{\event{\Z}},\\
+      &F(s,x,y)=\sum_{uy'}{\int{\beta_u(s,x,x',y,y')\,\varPsi_u(s,x,x',y,y')\,\rightlim{F}(s,x',y')\,\dd{x'}}},
+      &s\in{\event{\Z}},
+    \end{aligned}
+  \end{equation*}
+  for $0\le{s}\le{T}$.
+  Then the likelihood of $V$ is
+  \begin{equation*}
+    \lik=\sum_{y\in\func{Y}_0(\Z)}{\int{F(0,x,y)\,p_0(x)\,\dd{x}}}.
+  \end{equation*}
+\end{corol}
+
+\newpage
+\subsection{Special cases}
+\label{sec:special_cases}
+
+\subsubsection{Moran model and the Kingman coalescent}
+\label{sec:kingman}
+
+In the Moran model \citep{Moran1958,King2020}, events occur according to a rate-$\mu$ Poisson process.
+At each event, a compound birth-death jump occurs so that the population size, $n$, remains constant (cf.~\cref{fig:event_types}F).
+If we let $\Xr_t$ be the number of events that have occurred by time $t$, then $\Xr_t$ is a simple counting process, which we can use to define the state of the population process.
+Its KFE is then
+\begin{mathsize}{9pt}{10pt}
+  \begin{equation*}
+    \begin{gathered}
+      \frac{\partial{v}}{\partial{t}}=
+      \mu\,v(t,x-1)
+      -\mu\,v(t,x),
+      \qquad
+      v(0,x) =
+      \begin{cases}
+        1, & x=0,\\
+        0, & x>0.
+      \end{cases}
+    \end{gathered}
+  \end{equation*}
+\end{mathsize}%
+
+Since there is only a single deme, and since nothing depends on the state, in writing the corresponding filter equation (\cref{thm:obsc_filter}), we can take $w$ to be independent of both $x$ and $y$.
+
+In the classical case \citep{Kingman1982a}, $m$ samples are taken simultaneously at a single time, $T$.
+Then, if $B$ is the set of branch-times and $\ell(t)$ is the number of lineages in the genealogy at time $t$, upon summing over $x$ and $y$, the filter equation reads
+\begin{mathsize}{9pt}{10pt}
+  \begin{equation}
+    \label[pluralequation]{eq:moran-filter}
+    \begin{gathered}
+      w(0) = 1,
+      \qquad
+      \frac{\partial{w}}{\partial{t}}=
+      \mu\,w(t)\,\left(1-\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\right)
+      -\mu\,w(t),\quad t\notin{B},
+      \qquad
+      w(t) = \frac{\mu}{\tbinom{n}{2}}\,\wt(t),\quad t\in{B}.
+    \end{gathered}
+  \end{equation}
+\end{mathsize}%
+Integrating \cref{eq:moran-filter} and taking logarithms yields
+\begin{equation}
+  \label{eq:kingman}
+  \log{\lik}=|B|\,\log{\frac{\mu}{\tbinom{n}{2}}}-\frac{\mu}{\tbinom{n}{2}}\,\sum_{i=1}^{m}\!{\tbinom{i}{2}\,s^{}_i},
+\end{equation}
+where the $s^{}_{i}\coloneq\int{\Indicator{\ell(t)=i}\,\dd{t}}$ are the durations of the \emph{coalescent intervals}, \ie intervals between successive branch-points.
+We recognize \cref{eq:kingman} as the expression for the \citet{Kingman1982a} coalescent likelihood \citep[\eg][]{Wakeley2009}.
+
+More generally, if in addition samples are taken according to a rate-$\psi$ Poisson process such that the set of sample-times in the genealogy is $S=S_0\cup{S_1}$, where $S_0$, $S_1$ are the sets of times of terminal and inline samples, respectively, then the filter equation reads
+\begin{mathsize}{9pt}{10pt}
+  \begin{equation}
+    \label[pluralequation]{eq:moran-filter2}
+    \begin{gathered}
+      w(0) = 1,
+      \qquad
+      \frac{\partial{w}}{\partial{t}}=
+      -\mu\,\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\,w(t),
+      \quad t\notin{S\cup{B}},
+      \qquad
+      \frac{w(t)}{\wt(t)} =
+      \begin{cases}
+        \frac{\mu}{\tbinom{n}{2}}, & t\in{B},\\[8pt]
+        \psi\,\left(1-\frac{\ell(t)}{n}\right), & t\in{S_0},\\
+        \frac{\psi}{n}, & t\in{S_1}.
+      \end{cases}
+    \end{gathered}
+  \end{equation}
+\end{mathsize}%
+Integrating \cref{eq:moran-filter2} yields
+\begin{equation}
+  \label{eq:mgp}
+  \log{\lik}-|S|\,\log{\psi}
+  =\sum_{t\in{S_0}}{\log{\left(1-\frac{\ell(t)}{n}\right)}}
+  -|S_1|\,\log{n}
+  +|B|\,\log{\frac{\mu}{\tbinom{n}{2}}}
+  -\frac{\mu}{\tbinom{n}{2}}\,\sum_{i=1}^{\infty}\!{\tbinom{i}{2}\,s^{}_i},
+\end{equation}
+in agreement with the result of the very different derivation of \citet{King2020}.
+
+\subsubsection{Linear birth-death model} %% LBDP
+\label{sec:lbdp}
+
+In this model, the state variable, $\Xr_t$, is the size of a population at time $t$.
+All individuals face the same per-capita birth and death rates, which are $\lambda$ and $\mu$, respectively.
+Accordingly, the Kolmogorov backward equation for this model is
+\begin{mathsize}{9pt}{10pt}
+  \begin{equation*}
+    -\frac{\partial{F}}{\partial{s}}=
+    \lambda\,x\,\left[F(s,x+1)-F(s,x)\right]
+    +\mu\,x\,\left[F(s,x-1)-F(s,x)\right].
+  \end{equation*}
+\end{mathsize}%
+\citet{Stadler2010} considered the case where samples are taken through time at a uniform per-capita rate $\psi$ and lineages extant at time $T$ are sampled with probability $\rho$.
+Let $B$ denote the set of branch-times in an observed genealogy, let $S_0$ and $S_1$ be, respectively, the sets of terminal and inline sample times prior to time $T$, and suppose $n$ lineages are sampled at time $T$.
+Since there is only a single deme, and since nothing depends on the genealogy coloring, in writing the corresponding filter equation, we can apply \cref{corol:obsc_filter} and take $F$ to be independent of $y$.
+After summing over $y$, the regular part of the adjoint filter equation then reads
+\begin{mathsize}{9pt}{10pt}
+  \begin{equation}
+    \label{eq:filter-lbdp-reg}
+    -\frac{\partial{F}}{\partial{s}}(s,x)=
+    \lambda\,x\,\left(1-\frac{\tbinom{\ell(s)}{2}}{\tbinom{x+1}{2}}\right)\,F(s,x+1)
+    +\mu\,x\,F(s,x-1)\\
+    -\left(\lambda+\mu+\psi\right)\,x\,F(s,x),
+  \end{equation}
+\end{mathsize}%
+which holds for $x\ge\ell(s)$ and $s\notin{B\cup{S_0}\cup{S_1}\cup\Set{T}}$.
+The singular part is
+\begin{mathsize}{9pt}{10pt}
+  \begin{equation}
+    \label{eq:filter-lbdp-sing}
+    F(s,x)=\begin{cases}
+    \frac{2\lambda}{x+1}\,\rightlim{F}(s,x+1),
+    &s\in{B},
+    \\[2ex]
+    \psi\left(x-\ell(s)\right)\,\rightlim{F}(s,x),
+    &s\in{S_0},
+    \\[2ex]
+    \psi\,\rightlim{F}(s,x),
+    &s\in{S_1},
+    \\[2ex]
+    \tbinom{x}{n}\,(1-\rho)^{x-n}\,\rho^n,
+    &s=T.
+    \end{cases}
+  \end{equation}
+\end{mathsize}%
+We note that $F(s,x)=0$ when $x<\ell(s)$.
+Let us make the ansatz
+\begin{equation}
+  F(s,x)=C(s)\,G(s)^{x-\leftlim{\ell}(s)}\,H(s)^{\leftlim{\ell}(s)}\,K(x,\leftlim{\ell}(s)),
+\end{equation}
+where $G$, $H$ are smooth functions and $C$ is \caglad\ and piecewise constant, with discontinuities only at genealogical events, so that
+\begin{equation*}
+  \rightlim{F}(s,x)=\rightlim{C}(s)\,G(s)^{x-{\ell}(s)}\,H(s)^{{\ell}(s)}\,K(x,{\ell}(s)).
+\end{equation*}
+Putting these into the first three cases of \cref{eq:filter-lbdp-sing} yields conditions on $C$ and $K$ that are satisfied when
+\begin{equation*}
+  \begin{gathered}
+    K(x,\ell)=\frac{x!}{(x-\ell)!}
+    \qquad\text{and}\qquad
+    \frac{C(s)}{\rightlim{C}(s)}=\begin{cases}
+         {2\lambda}\,H(s),
+         &s\in{B},
+         \\[2ex]
+         \psi\,G(s)\,H(s)^{-1},
+         &s\in{S_0},
+         \\[2ex]
+         \psi,
+         &s\in{S_1}.
+    \end{cases}
+  \end{gathered}
+\end{equation*}
+To satisfy the last case of \cref{eq:filter-lbdp-sing}, we note that $\leftlim{\ell}(T)=n$ and therefore impose the conditions
+\begin{equation}
+  \label[pluralequation]{eq:FGfc}
+  G(T)=1-\rho, \qquad H(T)=1, \qquad C(T)=\frac{\rho^n}{n!}.
+\end{equation}
+Finally, upon substituting the ansatz into \cref{eq:filter-lbdp-reg}, we obtain for $s\notin{B\cup{S_0}\cup{S_1}\cup{\Set{T}}}$,
+\begin{equation*}
+  \begin{split}
+    -\frac{1}{F}\frac{\partial{F}}{\partial{s}}
+    &=
+    -(x-\ell)\,\frac{G'}{G}
+    -\ell\,\frac{H'}{H}\\
+    &=\lambda\,x\,\left(1-\frac{\tbinom{\ell}{2}}{\tbinom{x+1}{2}}\right)\,\frac{x+1}{x+1-\ell}\,G
+    +\mu\,x\,\frac{x-\ell}{x}\,G^{-1}
+    -\left(\lambda+\mu+\psi\right)\,x\\
+    &=\lambda\,(x+\ell)\,G
+    +\mu\,(x-\ell)\,G^{-1}
+    -\left(\lambda+\mu+\psi\right)\,x.\\
+  \end{split}
+\end{equation*}
+If this is to hold for all $x$ and $\ell$, we must have
+\begin{equation}
+  \label[pluralequation]{eq:FGeq}
+  \begin{gathered}
+    -\frac{G'}{G}
+    =\lambda\,G+\mu\,G^{-1}
+    -\left(\lambda+\mu+\psi\right),
+    \qquad
+    -\frac{H'}{H}
+    =2\,\lambda\,G-(\lambda+\mu+\psi).
+  \end{gathered}
+\end{equation}
+In passing, we note that \cref{eq:FGeq} are central to the original derivation of \citet{Stadler2010}.
+\Cref{eq:FGfc,eq:FGeq} form a final-value problem that can be solved by elementary means to obtain closed-form expressions for $G(s)$ and $H(s)$.
+%% \begin{equation*}
+%%   \begin{gathered}
+%%     G = \frac{\delta\,(1-\rho)\,\cosh\left(\tfrac{\delta}{2}\,(s-T)\right)+\left(\lambda-\mu+\psi-\rho\,(\lambda+\mu+\psi)\right)\,\sinh\left(\tfrac{\delta}{2}\,(s-T)\right)}{\delta\,\cosh\left(\tfrac{\delta}{2}\,(s-T)\right)+(\lambda-\mu-\psi-2\,\lambda\,\rho)\,\sinh\left(\tfrac{\delta}{2}\,(s-T)\right)}\\
+%%     H = \left(\cosh\left(\tfrac{\delta}{2}\,(s-T)\right)+\frac{\lambda-\mu-\psi-2\,\lambda\,\rho}{\delta}\,\sinh\left(\tfrac{\delta}{2}\,(s-T)\right)\right)^{-2}\\
+%%     \delta=\sqrt{(\lambda+\mu+\psi)^2-4\,\lambda\,\mu}
+%%   \end{gathered}
+%% \end{equation*}
+\Cref{corol:obsc_filter} then states that $\lik=\int{F(0,x)\,p_0(x)\,\dd{x}}$.
+In particular, conditional on $\Xr_0=x$, we have
+\begin{equation}
+  \lik=\frac{x!}{(x-\ell(0))!}\,\frac{\rho^n}{n!}\,G(0)^{x-{\ell}(0)}\,H(0)^{{\ell}(0)}\,\psi^{|S_1|}\,\prod_{\mathclap{e\in{B}}}{\left(2\,\lambda\,H(e)\right)}\,\prod_{\mathclap{e\in{S_0}}}{\left(\psi\,{G(e)}{H(e)^{-1}}\right)}.
+\end{equation}
+\citet[Thm.~3.5]{Stadler2010} derives an expression for this quantity in the special case $x=\ell(0)=1$, which is equivalent up to the factor of $n!$ explained in \cref{rem:ordering}.
+%% (our $G$, $H$ are related to the $p_0$, $q$ of that paper by the relations $G(s)=p_0(T-s)$, $H(s)=4/q(T-s)$).
+
+The foregoing represents an independent verification of the correctness of the theory in the case of the simple linear birth-death process with constant sampling through time and one bulk-sampling event.
+It should be clear, however, that even within the context of the unstructured, linear, birth-death process, the theory presented here readily accommodates complexities such as time-varying sampling rates, multiple bulk-sampling events at specified or random times, and time-varying birth and death rates.
+
+\section{Discussion}\label{sec:discussion}
+
+The theory presented here represents a generalization of the existing coalescent and birth-death-process approaches to phylodynamic inference.
+Importantly, because it allows computation of the likelihood via strictly forward-in-time computations, it permits consideration of models for which time-reversal arguments are not available.
+Moreover, inasmuch as the formulae of \cref{thm:obsc_filter} can be efficiently computed via sequential Monte Carlo, explicit expressions for transition probabilities are not needed:
 it is sufficient to be able to simulate from the population process.
@@
 Their procedures involve solving a large number of differential equations backward in time, relying on the time-reversibility of deterministic dynamics.
-In general, this time-reversibility is not a property of stochastic processes.
+In general, this time-reversibility is not available for stochastic processes.
 
@@
 
-If Sequential Monte Carlo algorithms are used to compute the likelihoods of \cref{thm:obsc_uncond}, then it is straightforward to simultaneously assimilate information from both time-series and genealogical data.
-One can therefore supplement traditional incidence, disease, or mortality time series with genealogical data in an inferential exercise.
+If sequential Monte Carlo algorithms are used to compute the likelihoods of \cref{thm:obsc_filter}, then it is straightforward to simultaneously assimilate information from both time-series and genealogical data.
+One can therefore supplement traditional incidence, disease, or mortality time series with genealogical data to improve inference.
 
 A limitation of the theory is that the population models are assumed to be pure jump processes, which allows consideration of demographic stochasticity and environmental stochasticity modeled by jumps involving multiple individuals \citep{Breto2011}, but disallows stochastic processes with a diffusive component.
-It should be possible to incorporate of the full range of Markovian environmental stochasticity via extension of this theory to population models containing both diffusion and jump components.
+However, we anticipate that it will be possible to incorporate the full range of Markovian environmental stochasticity via extension of this theory to population models containing both diffusion and jump components.
+
+Similarly, the theory presented here assumes that at most one birth event can occur at a time and that at most one deme can migrate or be sampled at a time (though multiple migrators or samples are permitted).
+These restrictions are not essential, and the proof of \cref{thm:pruned_lik} can be adapted to accommodate relaxations of these assumptions.
+This will be developed in a sequel.
 
 The price of the theory's flexibility is primarily computational.
-When Sequential Monte Carlo is used to evaluate the likelihood in \cref{thm:obsc_uncond}, the computational effort scales linearly with the number of samples.
+When sequential Monte Carlo is used to evaluate the likelihood in \cref{thm:obsc_filter}, the computational effort scales linearly with the number of samples.
 In its most straightforward implementation---using an event-driven algorithm \citep[\eg][]{Gillespie1977a}---it scales nonlinearly with population size in general.
 However, stochastic simulation schemes are available that scale independently of population size \citep{Higham2008}.
-On the other hand, the importance sampling underlying \cref{thm:obsc_uncond} will in general require effort that is exponential in the number of demes.
+On the other hand, the importance sampling underlying \cref{thm:obsc_filter} will in general require effort that is exponential in the number of demes.
 For models with many demes, therefore, approaches for ameliorating or circumventing this curse of dimensionality may be necessary.
-Critically, the substantial freedom one has in the choice of the importance-sampling distribution $\pi$ can be exploited for this purpose.
+Critically, the substantial freedom one has in the choice of the importance-sampling kernel $\pi$ can be exploited for this purpose.
 In particular, since it is permissible to ``borrow information'' from the future by means of the importance sampling, there is hope for highly efficient algorithmic computation.
 
+\phantomsection
+\addcontentsline{toc}{section}{Acknowledgments}
 \section*{Acknowledgments}
@@
 
+\phantomsection
 \bibliographystyle{preprint}
 \bibliography{phylopomp}
-
+\addcontentsline{toc}{section}{References}
+
+\clearpage
 \appendix
-\setcounter{equation}{0}
-\setcounter{figure}{0}
-\setcounter{table}{0}
-%%\setcounter{tcbcounter}{0}
+\numberwithin{equation}{section}
+\numberwithin{table}{section}
+\numberwithin{figure}{section}
 
@@
 \renewcommand{\thetable}{\thesection\arabic{table}}
-%%\renewcommand{\thetcbcounter}{\thesection\arabic{tcbcounter}}
+\renewcommand{\thealgorithm}{\thesection\arabic{algorithm}}
+
+\section{Proof of Theorem 1}
+\label[appendix]{sec:proof}
+
+We are given a pruned genealogy $\P=(\T,\Z,\Y)$ and a history $\H$ and wish to compute $\CondProb{\Pr_\T=\P}{\Hr_\T=\H}$.
+First, note that if $\event{\H}\nsupseteq\event{\P}$, then $\H$ and $\P$ are incompatible and $\CondProb{\Pr_\T=\P}{\Hr_\T=\H}=0$.
+Similarly, if any event of $\H$ is incompatible with the local structure of $\P$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_\T=\P}{\Hr_\T=\H}=0$.
+In either case, the conclusion of the theorem follows.
+Let us therefore suppose that neither of these conditions hold.
+
+We can always construct a time-respecting, but otherwise arbitrary ordering of the sample nodes.
+Let $\P_j$ represent the genealogy spanned by the first $j$ samples in this ordering.
+Since each $\P_j$ contains $\P_{j'}$ for all $j'<j$, $\P_j$ is a Markov chain.
+Therefore, we have the factorization
+\begin{equation*}
+  \CondProb{\P}{\H}=\prod_{j}{\CondProb{\P_j}{\P_{j-1},\H}}.
+\end{equation*}
+Each factor, $\CondProb{\P_j}{\P_{j-1},\H}$, can itself be factorized.
+In particular, suppose $\H$ and $\P_{j-1}$ are given and suppose that it is the $K$-th event in $\H$ that introduces the $j$-th sample.
+The $j$-th lineage extends backward in time until it either coalesces with $\P_{j-1}$ or reaches $t=0$ (\cref{fig:embedded_chain}).
+Moreover, since $\P$ is compatible with $\H$, nothing can happen to this lineage between events of $\H$.
+Therefore, we proceed backward, event by event, as follows.
+Since the type of event $K$ is given, the color of the $j$-th lineage just prior to that event is known.
+At each subsequent event, the lineage either emerges from the event or does not.
+Moreover, if the lineage does emerge from the event, the color prior to the event is determined and the color remains unchanged if it does not.
+Suppose that lineage $j$ is in deme $i$ just to the right of event $k<K$ and that the deme-$i$ occupancy and production are $n_i$, $r_i$, respectively.
+Let $\ell_{ij}$, $s_{ij}$ be the lineage count and saturation at this event in $\P_{j-1}$.
+Then the new lineage can be any one of the $n_i-\ell_i$ lineages as yet unaccounted for.
+Of these, $r_i-s_{ij}$ emerge from the event.
+Therefore, the probability that lineage $j$ emerges from the event is $(r_i-s_{ij})/(n_i-\ell_{ij})$ and the probability that it does not is $(n_i-r_i-\ell_{ij}+s_{ij})/(n_i-\ell_{ij})$.
+Let $q_{jk}$ be the former if lineage $j$ does in fact emerge from the event and the latter if it does not.
+Then $\CondProb{\P_j}{\P_{j-1},\H}=\prod_{k}{q_{jk}}$ and $\CondProb{\P}{\H}=\prod_{jk}{q_{jk}}$.
+
+\begin{figure}[b]
+  <<upo2b,results="hide">>=
+  plot_grid(
+    x |>
+      curtail(time=7.7,prune=FALSE,obscure=FALSE) |>
+      plot(points=TRUE,prune=TRUE,obscure=FALSE,ladderize=FALSE,palette=pal)+
+      expand_limits(x=9,y=3)+
+      theme(axis.line=element_line(color=grey(0.8)))+
+      labs(title=expression(P[8])),
+    x |>
+      curtail(time=8,prune=FALSE,obscure=FALSE) |>
+      plot(points=TRUE,prune=TRUE,obscure=FALSE,ladderize=FALSE,palette=pal)+
+      expand_limits(x=9,y=3)+
+      theme(axis.line=element_line(color=grey(0.8)))+
+      labs(title=expression(P[9])),
+    x |>
+      curtail(time=8.8,prune=FALSE,obscure=FALSE) |>
+      plot(points=TRUE,prune=TRUE,obscure=FALSE,ladderize=FALSE,palette=pal)+
+      expand_limits(x=9,y=3)+
+      theme(axis.line=element_line(color=grey(0.8)))+
+      labs(title=expression(P[10])),
+    ncol=1
+  )
+  @
+  \caption{
+    \textbf{Embedded chain of the pruned genealogy process.}
+    Three successive states of the embedded chain of the pruned genealogy process are shown.
+    At each step, a new lineage is added in a sampling event.
+    This lineage extends backward until it either coalesces with the pre-existing genealogy or reaches $t=0$.
+    \Cref{thm:pruned_lik} is derived by computing the probability that it coalesces at each possible point, conditional on the history of the population process.
+    \label{fig:embedded_chain}
+  }
+\end{figure}
+
+Now fix $k$ and consider the cumulative product $\prod_{m=1}^{j}{q_{mk}}$.
+Clearly $\ell_{1,k}=s_{1,k}=0$ and, if lineage $j$ has not already coalesced to the right of $k$, then $\ell_{j+1,k}=\ell_{j,k}+1$.
+Moreover, if $j$ emerges from $k$, then $s_{j+1,k}=s_{j,k}+1$ and $s_{j+1,k}=s_{j,k}$ if it does not.
+Therefore, as we accumulate each successive lineage $j$, we move one step to the right in the following diagram.
+
+\input{triangle3}
+
+\noindent
+At each step, the lineage count increases by 1 and the saturation either increases by a unit or stays the same.
+The probability of each option is indicated on the corresponding arrow.
+Although there are multiple paths by which the process may arrive at the final lineage count and saturation, the product of the probabilities along each path is the same:
+\begin{equation*}
+  \frac{\tbinom{n_i-\ell_i}{r_i-s_i}}{\tbinom{n_i}{r_i}}.
+\end{equation*}
+Moreover, since each of the options is independent of the others,
+\begin{equation*}
+  \prod_{j}{q_{jk}}=\BinRatio{n}{\ell}{r}{s},
+\end{equation*}
+the latter being the binomial ratio defined in \cref{sec:binomial_ratio}.
+
+Alternatively, one can reason as follows.
+Conditional on $\Hr_\T=\H$, at each time $t\in\event{\H}$, a jump of mark $\U_t$ occurred, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
+In $\P$, at time $t$, there are $\ell_i=\ell_i(\Y_t)$ lineages in deme $i$, of which $s_i=s_i(\Yt_t,\Y_t)$ are emergent.
+By assumption, at each genealogical event, lineages within a deme are exchangeable:
+each has an identical probability of being involved.
+This exchangeability implies that each lineage present in a deme at time $t$ was equally likely to have been one of the emergent lineages.
+In particular, at time $t$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the unpruned genealogy process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ red balls and $n_i-r_i$ black balls, exactly $s_i$ of the drawn balls are red, namely
+\begin{equation*}
+  \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
+\end{equation*}
+Because the lineages are labeled, each of the $\tbinom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct;
+just one of these is the one present in $\P$.
+Moreover, since, again conditional on $\Hr_\T=\H$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
+\begin{equation*}
+  \CondProb{\Pr_\T=\P}{\Hr_\T=\H}=\prod_{t\in\event{\H}}{\BinRatio{n(\X_t)}{\ell(\Y_t)}{r^{\U_t}}{s(\Yt_t,\Y_t)}}.
+\end{equation*}
+
+Returning to the possibility that $\H$ is incompatible with $\P$, since $\Prob{\Pr_\T=\P}=0$ if either any $Q^{}_{\U_t}=0$ or $\event{\P}\nsubseteq\event{\H}$, we obtain the result.
+\hfill\qedsymbol
 
 \section{Filter equations}
-\label{sec:filter_eqns}
-
-The likelihoods that appear in \cref{thm:pruned_uncond,thm:obsc_uncond} are integrals over large sets of histories.
+\label[appendix]{sec:filter_eqns}
+
+\subsection{Definition}
+
+The likelihoods that appear in \cref{thm:pruned_filter,thm:obsc_filter} are integrals over large sets of histories.
 As such, explicit expressions for them are not available, and we require mathematical tools to allow us to manipulate these quantities and devise algorithms for their numerical solution.
 The \emph{filter equations} we introduce here are suitable for these purposes, and we devote this appendix to exposing their essential properties.
-This extremely convenient formalism has, to our knowledge, not been thoroughly exploited, though we note their resemblance to the constructions of \citet{Ogata1978}, \citet{Puri1986}, \citet{Kliemann1990}, and \citet{Giesecke2018}.
+This extremely convenient formalism has, to our knowledge, not been thoroughly exploited in the context we introduce here, though we note their resemblance to the constructions of \citet{Ogata1978}, \citet{Puri1986}, \citet{Kliemann1990}, and \citet{Giesecke2018}.
 
@@
   Suppose that $B:\Rp\times\Xspace^2\to\Rp$ and $\lambda:\Rp\times\Xspace\to\mathbb{R}$ are are given measurable functions.
-  Let $S\subset\Rp$ be countable and locally finite (\ie $S\cap{[0,t]}$ is finite for all $t>0$).
+  Let $S\subset\Rp$ be locally finite (\ie $S\cap{[0,t]}$ is finite for all $t>0$).
   Then the system of equations
@@
     \begin{align}
-      \frac{\partial{w}}{\partial{t}}(t,x)&=
-      \int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
+      \frac{\partial{w}}{\partial{t}}(t,x)
+      &=\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
       -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
@@
   \end{mathsize}%
-  is called the \emph{filter equation} \emph{generated by} $\beta$, with \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observation times} $S$.
+  is called the \emph{filter equation} \emph{generated by} $\beta$, with \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed event times} $S$.
   The process $\Xr_t$ is said to be the \emph{driver} of the filter equation.
@@
 \end{remark}
+
+\subsection{Properties}
 
@@
   Since $\Prob{\Vr_0=1}=1$, $\lambda_0(\mathcal{E})=\Prob{\Xr_0\in\mathcal{E}}$, which implies that $w(0,x)=p_0(x)$.
-  For $t>0$ and $\Delta>0$ sufficiently small, the expectation can be broken into three terms, according to whether $\H_t$ has zero, one, or more than one event in $\halfclosed{t-\Delta,t}$.
+  For $t>0$ and $\Delta>0$, the expectation can be broken into three terms, according to whether $\H_t$ has zero, one, or more than one event in $\halfclosed{t,t+\Delta}$.
   Accordingly, as $\Delta\downarrow{0}$,
   \begin{equation*}
-    \begin{aligned}
-      w(t,x)=&\left(1-\Delta\,\int{\alpha(t-\Delta,x,x')\,\dd{x'}}\right)\,w(t-\Delta,x)\\
-      &\qquad+\Delta\,\int{\alpha(t-\Delta,x',x)\,B(t-\Delta,x',x)\,w(t-\Delta,x')\,\dd{x'}}+o(\Delta).
-    \end{aligned}
-  \end{equation*}
-  In the limit, we obtain \cref{eq:reg-filter-eq2}, the regular filter equation generated by $\alpha$, with boost $B$ and zero decay.
-  %%  \AAK{[This depends on $\alpha$ and $B$ being continuous in their first arguments. We have assumed they are continuous almost everywhere.]}
+    w(t+\Delta,x)=\left(1-\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}\right)\,w(t,x)+\Delta\,\int{\alpha(t,x',x)\,B(t,x',x)\,w(t,x')\,\dd{x'}}+o(\Delta).
+  \end{equation*}
+  Forming a difference quotient and taking the limit, we obtain \cref{eq:reg-filter-eq2}.
+  Note that here we use the assumption that $\alpha$ and $B$ are \cadlag\ as functions of $t$.
 \end{proof}
@@
 Let $\mathbb{S}$ be the space of increasing, locally finite sequences in $\Rp$, with the topology induced by the Skorokhod metric and Lebesgue measure.
-For $t\in\Rp$ and $s\in\mathbb{S}$, let $s^{}_t\coloneq{s\cap{[0,t]}}$.
-Thus if $s\in\mathbb{S}$ and $s^{}_t=(\hat{s}_1,\dots,\hat{s}_K)$, then the infinitesimal element of Lebesgue measure at $s^{}_t$ is $\dd{s^{}_t}=\prod_{n=1}^{K}\dd{\hat{s}_n}$.
+For $t\in\Rp$ and $S\in\mathbb{S}$, let $S^{}_t\coloneq{S\cap{[0,t]}}$.
+Thus if $S\in\mathbb{S}$ and $S^{}_t=(\hat{s}_1,\dots,\hat{s}_K)$, then the infinitesimal element of Lebesgue measure at $S^{}_t$ is $\dd{S^{}_t}=\prod_{n=1}^{K}\dd{\hat{s}_n}$.
 
@@
   \end{equation*}
-  Let $\lambda_t$ be a family measures on $\Xspace\times\mathbb{S}$ defined by
-  \begin{equation*}
-    \lambda_t(\mathcal{E},\mathcal{S})=\Expect{\Vr_t\cdot\Indicator{\Xr_t\in{\mathcal{E}}}\cdot\Indicator{\exists{s\in\mathcal{S}}\:\text{s.t.}\:\event{\Hr_t}\supseteq{s^{}_t}}},
+  Let $\lambda_t$ be a family of measures on $\Xspace\times\mathbb{S}$ defined by
+  \begin{equation*}
+    \lambda_t(\mathcal{E},\mathcal{S})=\Expect{\Vr_t\cdot\Indicator{\Xr_t\in{\mathcal{E}}}\cdot\Indicator{\exists{S\in\mathcal{S}}\:\text{s.t.}\:\event{\Hr_t}\supseteq{S^{}_t}}},
   \end{equation*}
   whenever $\mathcal{E}\subseteq\Xspace$ and $\mathcal{S}\subseteq\mathbb{S}$ are measurable.
-  Let $w(t,x,s)$ be the density of this measure, \ie
-  \begin{equation*}
-    \lambda_t(\dd{x}\,\dd{s})=w(t,x,s)\,\dd{x}\,\dd{s^{}_t}.
+  Let $w(t,x,S)$ be the density of this measure, \ie
+  \begin{equation*}
+    \lambda_t(\dd{x}\,\dd{S})=w(t,x,S)\,\dd{x}\,\dd{S^{}_t}.
   \end{equation*}
@@
     \label{eq:two-part-filter-reg}
-    \frac{\partial{w}}{\partial{t}}(t,x,s)&=
-    \int{w(t,x',s)\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
-    -\int{w(t,x,s)\,\alpha(t,x,x')\,\dd{x'}},\qquad
-    &t\notin{s},\\
+    \frac{\partial{w}}{\partial{t}}(t,x,S)
+    &=\int{w(t,x',S)\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
+    -\int{w(t,x,S)\,\alpha(t,x,x')\,\dd{x'}},\qquad
+    &t\notin{S},\\
     \label{eq:two-part-filter-sing}
-    w(t,x,s)&=\int{\wt(t,x',s)\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
-    &t\in{s}.
+    w(t,x,S)&=\int{\wt(t,x',S)\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
+    &t\in{S}.
   \end{align}
@@
 \begin{proof}
-  The proof proceeds by induction on the cardinality of $s_t$.
-  The base case, for which $s_t=\emptyset$, follows immediately from \cref{lemma:reg-filt}.
-  Assuming that it holds for $|s^{}_t|<K$, one has only to verify \cref{eq:two-part-filter-sing}.
+  The proof proceeds by induction on the cardinality of $S_t$.
+  The base case, for which $S_t=\emptyset$, follows immediately from \cref{lemma:reg-filt}.
+  Assuming that it holds for $|S^{}_t|<K$, one has only to verify \cref{eq:two-part-filter-sing}.
   This can be accomplished by integrating \cref{eq:Hdens} directly.
@@
 \end{remark}
+
+\subsection{Adjoint filter equation}
+\label{sec:adjoint_filter}
+
+Given driver, boost, and decay functions, a filter equation determines a family of measures on $\Xspace$.
+Specifically, suppose that, for $s\le{t}$ and $S\subset{\R}$ locally finite, $w(s,x,t,x')$ satisfies the filter equation
+\begin{align}
+  \label{eq:gen_filt1}
+  w(s,x,s,x') = &\delta(x,x')\\
+  \label{eq:gen_filt2}
+  \frac{\partial{w}}{\partial{t}}(s,x,t,x')=
+  &\int{w(s,x,t,\xi)\,\beta(t,\xi,x')\,B(t,\xi,x')\,\dd{\xi}}\\
+  &\qquad -\int{w(s,x,t,x')\,\beta(t,x',\xi)\,\dd{\xi}}
+  -\lambda(t,x')\,w(s,x,t,x'),
+  \qquad
+  &t\notin{S},\notag\\
+  \label{eq:gen_filt3}
+  w(s,x,t,x')=&\int{\wt(s,x,t,\xi)\,\beta(t,\xi,x')\,B(t,\xi,x')\,\dd{\xi}},\qquad
+  &t\in{S}.
+\end{align}
+Let $\omega$ be the measure whose density is $w$:
+\begin{equation*}
+  \omega(s,x,t,\mathcal{E})\coloneq\int_{\mathcal{E}}{w(s,x,t,x')\,\dd{x'}}.
+\end{equation*}
+If $f:\Xspace\to\R$ is a measurable function, then one defines its integral with respect to this measure in the usual way:
+\begin{equation*}
+  \omega(s,x,t,f) = \int{\omega(s,x,t,\dd{x'})\,f(x')} = \int{w(s,x,t,x')\,f(x')\,\dd{x'}}.
+\end{equation*}
+In particular, $\omega(s,x,t,f)$ is itself a measurable function of $x$ and we have the Chapman-Kolmogorov-like relation
+\begin{equation}\label{eq:chapkolm}
+  \omega(s,x,t,f) = \int{\omega(s,x,\tau,\omega(\tau,\xi,t,f))\,\dd{\xi}} = \int{\omega(s,x,\tau,\dd{\xi})\,\omega(\tau,\xi,t,f)},
+\end{equation}
+whenever $s\le{\tau}\le{t}$.
+Let us now fix $t$ and $f$ and write $F(s,x)=\omega(s,x,t,f)$.
+Let us take $\beta$, $B$, and $\lambda$ to be \cadlag\ in $t$ and $F$ to be \caglad\ in $s$.
+Applying \cref{eq:chapkolm} we obtain, for $\Delta>0$ and $s\notin{S}$,
+\begin{equation*}
+  \begin{split}
+    F(s-\Delta,x) = &\left(1-\Delta\,\int{\beta(s-\Delta,x,\xi)\,\dd{\xi}}-\Delta\,\lambda(s-\Delta,x)\right)\,F(s,x)\\
+    &\qquad +\Delta\,\int{\beta(s-\Delta,x,\xi)\,B(s-\Delta,x,\xi)\,F(s,\xi)\,\dd{\xi}}+o(\Delta).
+  \end{split}
+\end{equation*}
+Taking $\Delta\downarrow{0}$ gives
+\begin{equation}\label{eq:adjfilt1}
+  \begin{split}
+    -\frac{\partial{F}}{\partial{s}}
+    %%    &= \int{\leftlim{\beta}(s,x,\xi)\,\leftlim{B}(s,x,\xi)\,F(s,\xi)\,\dd{\xi}}-\int{\leftlim{\beta}(s,x,\xi)\,F(s,x)\,\dd{\xi}}-\leftlim{\lambda}(s,x)\,F(s,x)\\
+    &= \int{\leftlim{\beta}(s,x,\xi)\,\left[\leftlim{B}(s,x,\xi)\,F(s,\xi)-F(s,x)\right]\,\dd{\xi}}-\leftlim{\lambda}(s,x)\,F(s,x).
+  \end{split}
+\end{equation}
+For $s\in{S}$, we have
+\begin{equation}\label{eq:adjfilt2}
+  F(s,x)
+  =\int{\omega(s,x,s,\dd{\xi})\,\rightlim{F}(s,\xi)}
+  =\int{\beta(s,x,\xi)\,B(s,x,\xi)\,\rightlim{F}(s,\xi)\,\dd{\xi}},
+\end{equation}
+where $\rightlim{F}$ denotes the right-limit of $F$.
+\Cref{eq:adjfilt1,eq:adjfilt2} constitute the \emph{adjoint form} of \cref{eq:gen_filt1,eq:gen_filt2,eq:gen_filt3}.
+Together with the final condition $F(t,x)=f(x)$, these determine $\omega(s,x,t,f)$ for all $s\le{t}$.
+Note that, when $B=1$ and $\lambda=0$, the adjoint filter equation becomes the familiar Kolmogorov backward equation (\cref{eq:kbe}) for the driving process.
+
+\subsection{Solving filter equations by sequential Monte Carlo}
 
@@
     \begin{equation*}
-      \begin{aligned}
+      %% \begin{aligned}
+      %%   \frac{\partial{w}}{\partial{t}}(t,x)
+      %%   =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
+      %%   -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
+      %%   &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
+      %% \end{aligned}
       \frac{\partial{w}}{\partial{t}}(t,x)
-        =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
-        -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
-        &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
-      \end{aligned}
+      =\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
+      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}
+      -\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
     \end{equation*}
@@
 
-\begin{remark}
 \Cref{eq:filterlemma} are recognizable as the KFE of a certain process $(\Xr_t,\Vr_t)$.
@@
 this average is simply $\int{w(t,x)\,\dd{x}}$.
-  Thus, \cref{lemma:monte_carlo} shows how to integrate \cref{eq:two-part-filter-reg,eq:two-part-filter-sing} in the Monte Carlo sense.
-\end{remark}
+Thus, \cref{lemma:monte_carlo} gives a recipe for the integration of \cref{eq:two-part-filter-reg,eq:two-part-filter-sing} in the Monte Carlo sense.
+A pseudocode representation of this procedure is given in \cref{alg:smc}.
+This is a simple sequential importance sampling algorithm and as such can be improved upon in many ways \citep{Doucet2001,Arulampalam2002,Wills2023}.
+Our purpose here is merely to illustrate the implications of \cref{lemma:monte_carlo} by showing how the features of \cref{eq:filter-eq-defn-reg,eq:filter-eq-defn-sing} translate into computations, in the simplest possible context.
 
@@
 
+\begin{algorithm}[t]
+  \caption{
+    \textbf{Sequential Monte Carlo filter.}
+    The following is a simple Monte Carlo scheme for the filter given by \cref{eq:filter-eq-defn-reg,eq:filter-eq-defn-sing}.
+    It takes as input an initial time $t_0$, a final time $T$, and a sequence of observed event times $S=\Set{t_1,\dots,t_K}$.
+    We order the latter so that $t_0\le{t_1}<t_2<\cdots<t_K\le{t_{K+1}}\coloneq{T}$.
+    In addition, we are given event rates $\beta$, a boost function $B$, a decay function $\lambda$, and the initial distribution of latent states $p_0$.
+    As in \cref{lemma:monte_carlo}, we let
+    \begin{equation*}
+      \begin{gathered}
+        A(t,x)\coloneq{\int{\beta(t,x,x')\,\dd{x'}}},
+        \qquad
+        \pi(t,x,\dd{x'})\coloneq\frac{\beta(t,x,x')}{A(t,x)}\,\dd{x'}.
+      \end{gathered}
+    \end{equation*}
+    For a given, fixed ensemble size, $J\in\Zp$, the algorithm returns an unbiased Monte Carlo estimate, $\rdm{\hat{\lik}}$, of the likelihood.
+  }
+  \label{alg:smc}
+  \begin{algorithmic}[1]
+    %% \Require{
+    %%   times, $S$;
+    %%   initial latent-state distribution, $p_0$;
+    %%   event rates, $\beta$;
+    %%   boost function, $B$;
+    %%   decay function, $\lambda$;
+    %%   ensemble size, $J\in\Zp$.
+    %% }
+    %% \Ensure{
+    %%   unbiased Monte Carlo likelihood estimate, $\rdm{\lik}$
+    %% }
+    \Procedure{SMCFilter}{$t_0,T,S,\beta,B,\lambda,p_0,J$}
+    \Statex Initialize an ensemble of latent states and weights:
+    \State
+    $x_j\sim{p_0(\cdot)}$, for $j=1,\dots,J$
+    \State
+    $w_j\gets{1}$, for $j=1,\dots,J$
+    \For{$k=0,\dots,K$}
+    \Comment loop over observed event times
+    \For{$j=1,\dots,J$}
+    \Comment loop over the ensemble
+    \State $t\gets{t_{k}}$
+    \If{$k>0$}
+    \Comment singular part (\cref{eq:filter-eq-defn-sing})
+    \State $x'_j\sim\pi(t,x_j,\cdot)$
+    \State $w_j\gets{w_j\,A(t,x_j)\,B(t,x_j,x'_j)}$
+    \State $x_j\gets{x'_j}$
+    \EndIf
+    \While{$t<t_{k+1}$}
+    \Comment regular part (\cref{eq:filter-eq-defn-reg})
+    \State $\Delta\sim{\mathrm{Exp}(A(t,x_j))}$
+    \Comment Sojourn times are exponentially distributed.
+    \If{$t+\Delta<{t_{k+1}}$}
+    \State $x'_j\sim{\pi(t,x_j,\cdot)}$
+    \State $t'\gets{t+\Delta}$
+    \State $w_j\gets{w_j\,B(t',x_j,x'_j)\,\exp{(-\lambda(t,x_j)\,\Delta)}}$
+    \State $x_j\gets{x'_j}$
+    \State $t\gets{t'}$
+    \Else
+    \State $\Delta\gets{t_{k+1}-t}$
+    \State $w_j\gets{w_j\,\exp{(-\lambda(t,x_j)\,\Delta)}}$
+    \State $t\gets{t_{k+1}}$
+    \EndIf
+    \EndWhile
+    \EndFor
+    \EndFor
+    \State $\rdm{\hat{\lik}}\gets\tfrac{1}{J}\displaystyle\sum_{j}{w_j}$
+    \EndProcedure
+  \end{algorithmic}
+\end{algorithm}
+
 \section{Examples}
-\label{sec:app-examples}
+\label[appendix]{sec:app-examples}
 
@@
 
-\citet{King2022} worked out formulas for the exact likelihood of a genealogy induced by an SIRS model.
+\citet{King2022} worked out formulae for the exact likelihood of a genealogy induced by an SIRS model.
 The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.
@@
 Accordingly, the marks are $\Jumps=\Set{\lab{Trans},\lab{Recov},\lab{Wane},\lab{Sample}}$.
-\Cref{tab:sirs_model_elements} gives $\alpha_u$, $r_u$, and the event type for each of these marks.
+\Cref{tab:sirs_model_elements} gives $\alpha_u$, $r^u$, and the event type for each of these marks.
 
@@
     Elements of the SIRS model pertinent to the genealogy process.
-    The table shows the rate ($\alpha_u$), jump ($x\mapsto{x'}$), production ($r_u$), and event type for each of the model's four marks ($u$).
+    The table shows the rate ($\alpha_u$), jump ($x\mapsto{x'}$), production ($r^u$), and event type for each of the model's four marks ($u$).
   }
@@
     \hline\hline
-    $u$ & $\alpha_u$ \bigstrut & $x\mapsto{x'}$ \bigstrut & $r_u$ \bigstrut & Event type \\
+    $u$ & $\alpha_u$ \bigstrut & $x\mapsto{x'}$ \bigstrut & $r^u$ \bigstrut & Event type \\
     \hline
@@
     \label{fig:sirs_example}
-    Likelihood computation for the SIRS model by Sequential Monte Carlo.
+    Likelihood computation for the SIRS model by sequential Monte Carlo.
     (A)~A~simulated genealogy for $\beta=\Sexpr{sirs_params$Beta}$, $\gamma=\Sexpr{sirs_params$gamma}$, $\omega=\Sexpr{sirs_params$omega}$, $\psi=\Sexpr{sirs_params$psi}$, $(S_0,I_0,R_0)=(\Sexpr{with(sirs_params,c(S0,I0,R0))})$.
@@
 where $N=S+E+I+R$ is the total population size.
-Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.
+Note that the terms associated with sampling cancel each other in the KFE, since, with the state space defined as above, sampling has no effect on the state.
 
@@
 The corresponding marks are $\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$.
-\Cref{tab:seirs_model_elements} gives $\alpha_u$, $r_u$, and the event type for each of these marks.
-
-\input{figs/seirs_events}
+\Cref{tab:seirs_model_elements} gives $\alpha_u$, $r^u$, and the event type for each of these marks.
+
+\input{seirs_events}
 
@@
     Elements of the SEIRS model pertinent to the genealogy process.
-    The table shows the rate ($\alpha_u$), jump ($x\mapsto{x'}$), production ($r_u$), and event type for each of the model's five marks ($u$).
+    The table shows the rate ($\alpha_u$), jump ($x\mapsto{x'}$), production ($r^u$), and event type for each of the model's five marks ($u$).
   }
@@
     \hline\hline
-    $u$ & $\alpha_u$ & $x\mapsto{x'}$ & $r_u$ & Event type \\
+    $u$ & $\alpha_u$ & $x\mapsto{x'}$ & $r^u$ & Event type \\
     \hline
@@
     the singular portion describes the effect of these events ($t\in{{B}\cup{S_0}\cup{S_1}}$).
-    Each line corresponds to a potential event, but only those for which $Q=1$ appear in the equation.
+    Each line corresponds to a potential event.
     An event with mark $u$ and saturation $s$ has the boost given by the binomial ratio shown (third column).
@@
 
-The filter equation corresponding to the scheme of \cref{tab:seirs_integ} is presented in \Cref{box:SEIRS}.
-Some numerical results are presented in \cref{fig:seirs_example}.
-
-\clearpage
-\newpage
-
-\begin{bluebox}[label={box:SEIRS}]{Filter equation for the SEIRS model}
+In the application of \cref{thm:obsc_filter,corol:obsc_filter}, one has wide latitude in the selection of the kernels $\pi$ and $q$.
+\Cref{tab:seirs_integ} gives a particular set of choices that are \emph{na\"ive} in the sense that they refrain from borrowing information from the future.
+Such a na\"ive filter is always available, though it will rarely be optimal.
+
+Using the choices of \cref{tab:seirs_integ}, we proceed to write down a filter equation for the SEIRS model.
 As previously, given an obscured genealogy, let $B$ be the set of its branch times, $S_0$ be the set of tip-sample times, and $S_1$ be the set of inline sample times.
@@
 
-  A variety of importance-sampling kernels are permissible under the terms of \cref{thm:obsc_uncond}.
-  With a particular choice of importance-sampling kernel, the filter equation uniquely specifies a Sequential Monte Carlo algorithm for estimating the likelihood.
-  The choices made in \cref{tab:seirs_integ} underlie the results displayed in \cref{fig:seirs_example}.
-\end{bluebox}
+A variety of importance-sampling kernels $q$, $\pi$ are permissible under the terms of \cref{thm:obsc_filter}.
+With a particular choice of importance-sampling kernel, the filter equation uniquely specifies a sequential Monte Carlo algorithm for estimating the likelihood.
+The specific choices made in \cref{tab:seirs_integ} underlie the results displayed in \cref{fig:seirs_example}.
+Some numerical results obtained using this scheme are presented in \cref{fig:seirs_example}.
 
@@
     \label{fig:seirs_example}
-    Likelihood computation for the SEIRS model by Sequential Monte Carlo, using the scheme of \cref{box:SEIRS}.
+    Likelihood computation for the SEIRS model by sequential Monte Carlo, using the scheme of \cref{tab:seirs_integ}.
     (A)~Simulated genealogy for $\beta=\Sexpr{seirs_params$Beta}$,
@@
     (B)~Likelihood slice in the $\sigma$-direction.
-    Each point represents the estimate of an independent Sequential Monte Carlo computation.
+    Each point represents the estimate of an independent sequential Monte Carlo computation.
     The blue curve shows a LOESS smooth;
@@
 
-\subsection{Two-strain competition model}
-
-A simple model for the competition of two strains for susceptible hosts is depicted in \cref{fig:example_models}B.
-This example will be included in a forthcoming draft.
-
-\subsection{Superspreading model}
-
-\Cref{fig:example_models}D depicts a model of superspreading.
-This example will be included in a forthcoming draft.
-
-\subsection{Linear birth-death model} %% LBDP
-
-In this model, the state variable is the size, $N_t$, of a population at time $t$.
-All individuals face the same per-capita birth and death rates, which are $\lambda$ and $\mu$, respectively.
-The KFE is
-\begin{mathsize}{9pt}{10pt}
-  \begin{equation*}
-    \frac{\partial{v}}{\partial{t}}=
-    \lambda\,(n-1)\,v(t,n-1)
-    -\lambda\,n\,v(t,n)
-    +\mu\,(n+1)\,v(t,n+1)
-    -\mu\,n\,v(t,n)
-  \end{equation*}
-\end{mathsize}%
-\citet{Stadler2010} considered the case where samples are taken through time at a uniform per-capita rate $\psi$.
-In this case, since there is only one deme, in the filter equation, $w$ can be taken to be independent of $y$.
-If $B$ is the set of branch-times and $S_0$, $S_1$ are the sets of terminal and inline samples, respectively, then the regular part of the filter equation is
-\begin{mathsize}{9pt}{10pt}
-  \begin{equation}
-    \label{eq:filter-lbdp-reg}
-    \begin{aligned}
-      \frac{\partial{w}}{\partial{t}}(t,n)=
-      &\lambda\,(n-1)\,\left(1-\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\right)\,w(t,n-1)
-      -\lambda\,n\,w(t,n)
-      \\
-      &\qquad+\mu\,(n+1)\,w(t,n+1)-\mu\,n\,w(t,n)-\psi\,n\,w(t,n),
-      &&n\ge\ell(t),\quad t\notin{B\cup{S_0}\cup{S_1}}
-    \end{aligned}
-  \end{equation}
-\end{mathsize}%
-and the singular part is
-\begin{mathsize}{9pt}{10pt}
-  \begin{equation}
-    \label{eq:filter-lbdp-sing}
-    \begin{gathered}
-      w(t,n)=
-      \frac{\lambda\,(n-1)}{\tbinom{n}{2}}\,\wt(t,n-1),
-      \quad t\in{B},
-      \\
-      w(t,n)=
-      \psi\,n\,\left(1-\frac{\ell(t)}{n}\right)\,\wt(t,n),
-      \quad t\in{S_0},
-      \qquad
-      w(t,n)=
-      \psi\,\wt(t,n),
-      \quad t\in{S_1}.
-    \end{gathered}
-  \end{equation}
-\end{mathsize}%
-\Cref{eq:filter-lbdp-reg,eq:filter-lbdp-sing} are supplemented by the ancillary condition $w(t,n)=0$ for $n<\ell(t)$.
-
-\begin{table}[h!]
-  \caption{
-    \label{tab:lbdp_model_elements}
-    Elements of the linear birth-death-sampling model pertinent to the genealogy process.
-  }
-  \renewcommand{\arraystretch}{1.2}
-  \begin{tabular}{cccl}
-    \hline\hline
-    $u$ & $\alpha_u$ & $r_u$ & Event type \\
-    \hline
-    $\lab{Birth}$  & $\lambda N$ & 2 & pure birth \\
-    $\lab{Death}$  & $\gamma N$ & 0 & pure death \\
-    $\lab{Sample}$ & $\psi N$ & 1 & pure sample \\
-    \hline\hline
-  \end{tabular}
-\end{table}
-
-<<lbdp3>>=
+%% \begin{table}[h!]
+%%   \caption{
+%%     \label{tab:lbdp_model_elements}
+%%     Elements of the linear birth-death-sampling model pertinent to the genealogy process.
+%%   }
+%%   \renewcommand{\arraystretch}{1.2}
+%%   \begin{tabular}{cccl}
+%%     \hline\hline
+%%     $u$ & $\alpha_u$ & $r^u$ & Event type \\
+%%     \hline
+%%     $\lab{Birth}$  & $\lambda N$ & 2 & pure birth \\
+%%     $\lab{Death}$  & $\gamma N$ & 0 & pure death \\
+%%     $\lab{Sample}$ & $\psi N$ & 1 & pure sample \\
+%%     \hline\hline
+%%   \end{tabular}
+%% \end{table}
+
+<<lbdp3,eval=FALSE>>=
 lbdp.params <- data.frame(
@@
 
-\begin{figure}
-  \begin{center}
-    <<lbdp3_plot,fig.dim=c(9,7),out.width="100%">>=
+%% \begin{figure}
+%%   \begin{center}
+<<lbdp3_plot,fig.dim=c(9,7),out.width="100%",eval=FALSE>>=
 pal <- c(viridis_pal(option="H",begin=0.1,end=0.8)(8),"#000000")
@@
 @
-  \end{center}
-  \caption{
-    \label{fig:lbdp_example}
-    Likelihood computation for the constant-parameter, linear birth-death-sampling model, according to \cref{thm:obsc_uncond} via Sequential Monte Carlo.
-    Panel (A) shows the genealogy, simulated for $\lambda=\Sexpr{lbdp.params$lambda}$, $\mu=\Sexpr{lbdp.params$mu}$, $\psi=\Sexpr{lbdp.params$psi}$, $N_0=\Sexpr{lbdp.params$n0}$.
-    Panel (B) shows a likelihood slice, through the true parameters in the $\mu$ direction.
-    As computational effort (\ie number of particles) increases, the Monte Carlo estimates converge on the exact values, for which an explicit formula exists in this case.
-    The dashed horizontal lines show the approximate maximized likelihood and the 95\% critical value (under the likelihood-ratio test).
-    Panels (C--E) are log-log plots that show how the root-mean-square error (RMSE), imprecision (SD), and bias decrease with effort.
-    Note that the bias and the SD are roughly inversely proportional to the effort and its square-root, respectively, as expected.
-  }
-\end{figure}
-
-\subsection{Moran model and the Kingman coalescent}
-
-In the Moran model, events occur according to a rate-$\mu$ Poisson process.
-At each event, a compound birth-death jump (cf.~\cref{fig:event_types}F) occurs so that the population size, $n$, remains constant.
-If we let $X_t$ be the number of events that have occurred by time $t$, then $X_t$ is a simple counting process, which we can use to define the state of the population process.
-Its KFE is then
-\begin{mathsize}{9pt}{10pt}
-  \begin{equation*}
-    \begin{gathered}
-      \frac{\partial{v}}{\partial{t}}=
-      \mu\,(x-1)\,v(t,x-1)
-      -\mu\,x\,v(t,x),
-      \qquad
-      v(0,x) =
-      \begin{cases}
-        1, & x=0,\\
-        0, & x>0.
-      \end{cases}
-    \end{gathered}
-  \end{equation*}
-\end{mathsize}%
-
-Since there is only a single deme, and since nothing depends on the state, in writing the corresponding filter equation, we can take $w$ to be independent of both $x$ and $y$.
-
-In the classical case \citep{Kingman1982a}, $m$ samples are taken simultaneously at a single time, $T$.
-Then, if $B$ is the set of branch-times and $\ell(t)$ is the number of lineages in the genealogy at time $t$, the filter equation reads
-\begin{mathsize}{9pt}{10pt}
-  \begin{equation}
-    \label[pluralequation]{eq:moran-filter}
-    \begin{gathered}
-      w(0) = 1,
-      \qquad
-      \frac{\partial{w}}{\partial{t}}=
-      \mu\,w(t)\,\left(1-\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\right)
-      -\mu\,w(t),\quad t\notin{B},
-      \qquad
-      w(t) = \frac{\mu}{\tbinom{n}{2}}\,\wt(t),\quad t\in{B}.
-    \end{gathered}
-  \end{equation}
-\end{mathsize}%
-Integrating \cref{eq:moran-filter} and taking logarithms yields
-\begin{equation}
-  \label{eq:kingman}
-  \log{w(T)}=k\,\log{\frac{\mu}{\tbinom{n}{2}}}-\frac{\mu}{\tbinom{n}{2}}\,\sum_{i=m-k}^{m}\!{\tbinom{i}{2}\,s^{}_i},
-\end{equation}
-where $k=|B|$ is the number of branch-points in $[0,T]$ and the $s^{}_{i}\coloneq\int{\Indicator{\ell(t)=i}\,\dd{t}}$ are the durations of the \emph{coalescent intervals}, \ie intervals between successive branch-points.
-We recognize \cref{eq:kingman} as the expression for the \citet{Kingman1982a} coalescent \citep[\eg][]{Wakeley2009}.
-
-More generally, if in addition samples are taken according to a rate-$\nu$ Poisson process such that the set of sample-times in the genealogy is $S=S_0\cup{S_1}$, where $S_0$, $S_1$ are the sets of times of terminal and inline samples, respectively, then the filter equation reads
-\begin{mathsize}{9pt}{10pt}
-  \begin{equation}
-    \label[pluralequation]{eq:moran-filter2}
-    \begin{gathered}
-      w(0) = 1,
-      \qquad
-      \frac{\partial{w}}{\partial{t}}=
-      -\mu\,\frac{\tbinom{\ell(t)}{2}}{\tbinom{n}{2}}\,w(t),
-      \quad t\notin{S\cup{B}},
-      \qquad
-      w(t) = \frac{\mu}{\tbinom{n}{2}}\,\wt(t),\quad t\in{B},\\[8pt]
-      w(t) = \nu\,\left(1-\frac{\ell(t)}{n}\right)\,\wt(t),\quad t\in{S_0},
-      \qquad
-      w(t) = \frac{\nu}{n}\,\wt(t),\quad t\in{S_1}.
-    \end{gathered}
-  \end{equation}
-\end{mathsize}%
-Integrating \cref{eq:moran-filter2} yields
-\begin{equation}
-  \label{eq:mgp}
-  \log{w(T)}-|S|\,\log{\nu}
-  =\sum_{t\in{S_0}}{\log{\left(1-\frac{\ell(t)}{n}\right)}}
-  -|S_1|\,\log{n}
-  +|B|\,\log{\frac{\mu}{\tbinom{n}{2}}}
-  -\frac{\mu}{\tbinom{n}{2}}\,\sum_{i=1}^{\infty}\!{\tbinom{i}{2}\,s^{}_i}.
-\end{equation}
+%%   \end{center}
+%%   \caption{
+%%     \label{fig:lbdp_example}
+%%     Likelihood computation for the constant-parameter, linear birth-death-sampling model, according to \cref{thm:obsc_filter} via sequential Monte Carlo.
+%%     Panel (A) shows the genealogy, simulated for $\lambda=\Sexpr{lbdp.params$lambda}$, $\mu=\Sexpr{lbdp.params$mu}$, $\psi=\Sexpr{lbdp.params$psi}$, $N_0=\Sexpr{lbdp.params$n0}$.
+%%     Panel (B) shows a likelihood slice, through the true parameters in the $\mu$ direction.
+%%     As computational effort (\ie number of particles) increases, the Monte Carlo estimates converge on the exact values, for which an explicit formula exists in this case.
+%%     The dashed horizontal lines show the approximate maximized likelihood and the 95\% critical value (under the likelihood-ratio test).
+%%     Panels (C--E) are log-log plots that show how the root-mean-square error (RMSE), imprecision (SD), and bias decrease with effort.
+%%     Note that the bias and the SD are roughly inversely proportional to the effort and its square-root, respectively, as expected.
+%%   }
+%% \end{figure}
 

### 2025-07-08 ms.R

@@ -71,6 +71,30 @@ plot_grid(
 )
 
 
+## ----upo2b,results="hide"-----------------------------------------------------
+plot_grid(
+  x |>
+    curtail(time=7.7,prune=FALSE,obscure=FALSE) |>
+    plot(points=TRUE,prune=TRUE,obscure=FALSE,ladderize=FALSE,palette=pal)+
+    expand_limits(x=9,y=3)+
+    theme(axis.line=element_line(color=grey(0.8)))+
+    labs(title=expression(P[8])),
+  x |>
+    curtail(time=8,prune=FALSE,obscure=FALSE) |>
+    plot(points=TRUE,prune=TRUE,obscure=FALSE,ladderize=FALSE,palette=pal)+
+    expand_limits(x=9,y=3)+
+    theme(axis.line=element_line(color=grey(0.8)))+
+    labs(title=expression(P[9])),
+  x |>
+    curtail(time=8.8,prune=FALSE,obscure=FALSE) |>
+    plot(points=TRUE,prune=TRUE,obscure=FALSE,ladderize=FALSE,palette=pal)+
+    expand_limits(x=9,y=3)+
+    theme(axis.line=element_line(color=grey(0.8)))+
+    labs(title=expression(P[10])),
+  ncol=1
+)
+
+
 ## ----sirs3--------------------------------------------------------------------
 data.frame(
   Beta=4,gamma=2,psi=1,omega=1,
@@ -272,166 +296,166 @@ plot_grid(
 )
 
 
-## ----lbdp3--------------------------------------------------------------------
-lbdp.params <- data.frame(
-  lambda=1.2,mu=0.8,psi=1,n0=5,
-  time=10
-)
+## ----lbdp3,eval=FALSE---------------------------------------------------------
+# lbdp.params <- data.frame(
+#   lambda=1.2,mu=0.8,psi=1,n0=5,
+#   time=10
+# )
+# 
+# bake(
+#   file="lbdp3a.rds",
+#   seed=915645370,
+#   dependson=lbdp.params,
+#   lbdp.params |>
+#     with(
+#       runLBDP(lambda=lambda,mu=mu,psi=psi,n0=n0,time=time)
+#     )
+# ) -> lbdp_tree
+# 
+# bake(
+#   file="lbdp3b.rds",
+#   seed=712604404,
+#   dependson=list(lbdp_tree,lbdp.params),
+#   {
+#     lbdp.params |>
+#       with(
+#         expand_grid(
+#           rep=1:10,
[…]

### 2025-07-15 ms.Rnw
fix typo

@@
 
-Code sufficient for the reproduction of all the figures in this paper are freely available for download at \url{https://github.com/kingaa/structured-genealogy-process-paper}.
+Codes sufficient for the reproduction of all the figures in this paper are freely available for download at \url{https://github.com/kingaa/structured-genealogy-process-paper}.
 An archival version of these will be stored on Zotero upon publication of a peer-reviewed version of this paper.
